### 2019-12-06 authors.tex
Add skeleton for building paper

\author[1]{Jane Doe}
\affil[1]{Berkeley Institute for Data Science, University of California, Berkeley, CA 94720, USA}

### 2019-12-06 paper.tex
Add skeleton for building paper

\documentclass[fleqn,10pt]{wlscirep}

% Packages
\usepackage{pifont}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{float}
% \usepackage[multiple]{footmisc}
\usepackage{booktabs}
\usepackage{csquotes}
\usepackage{listings, textcomp}

\lstset{ %
  basicstyle=\ttfamily\footnotesize,  % size of fonts used for the code
  breaklines=true,   % automatic line breaking only at whitespace
  captionpos=b,   % sets the caption-position to bottom
  commentstyle=\color{gray},  % comment style
  keywordstyle=\color{blue},  % keyword style
  stringstyle=\color{red},  % string literal style
  upquote=true  %straight single quotes (requires textcomp)
}

% New Commands
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\fixme}[1]{\textcolor{red}{{#1}}}
\newcommand{\inlinecite}[1]{\footnotesize\citen{#1}}

\usepackage{etoolbox}
\makeatletter
\patchcmd{\@maketitle}
 {\noindent
{\parbox{\dimexpr\linewidth-2\fboxsep\relax}{\color{color1}\large\sffamily\textbf{ABSTRACT}}}}
 {\noindent
{\parbox{\dimexpr\linewidth-2\fboxsep\relax}{\color{color1}\large\sffamily\textbf{ABSTRACT}}}}
 {}{}
\makeatother

\title{NumPy---Array Computation for Python}

\input{authors}

\keywords{Scientific computing, Python, Mathematics}

\begin{abstract}
  NumPy is the fundamental array computing library for Python.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section*{Introduction}

\section*{Project organization and community}

\subsection*{Governance}

\subsection*{Maintainers and contributors}

\subsection*{Funding}

\section*{Discussion}

% The Discussion should be succinct and must not contain subheadings.

\section*{Data Availability}
% The statement should be provided as a separate section (titled 'Data Availability') at the end of the main text, before the 'References' section.
All NumPy source code and most data generated for the current study are available in the NumPy GitHub repository, \url{https://github.com/numpy}. Some supporting code and data have also been stored in other public repositories cited by this manuscript.

\bibliography{references}

\section*{Acknowledgements}

\section*{Author Contributions Statement}

%Must include all authors, identified by initials, for example: A.A.
%conceived the experiment(s),  A.A. and B.A. conducted the experiment(s), C.A.
%and D.A. analysed the results.  All authors reviewed the manuscript.

\section*{Competing Interests}

The authors declare no competing interests.

%To include, in this order: \textbf{Accession codes} (where applicable);
%\textbf{Competing financial interests} (mandatory statement).

%The corresponding author is responsible for submitting a
%\href{http://www.nature.com/srep/policies/index.html#competing}{competing
%financial interests statement} on behalf of all authors of the paper. This
%statement must be included in the submitted article file.

\section*{Consortium}
\subsection*{NumPy Contributors}
\input{consortium}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2019-12-10 paper.tex
Add abstract

@@
   NumPy is the fundamental array computing library for Python.
+  It has been actively developed for over two decades and has
+  more than 850 contributors, nearly 300,000 dependent repositories
+  on GitHub, and millions of downloads per year.
+  It is extensively used in research and teaching.
+  According to a 2018 survey, NumPy is the second most commonly used machine
+  learning tool among enterprises \cite{451report2018}.
+  In this article, we provide an overview of the capabilities and development
+  practices of the NumPy library and highlight some recent technical
+  developments.
 \end{abstract}
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 

### 2019-12-10 paper.tex
Add intro and history first draft

@@
 
+NumPy is the fundamental array computation library for the Python
+ecosystem.  It underpins almost every library that does numerical
+computation, including SciPy.  NumPy consists of the `ndarray`-object
+along with utility functions to operate on such arrays.  Because of
+its inherent simplicity—being a pointer to memory with some
+associated meta-data about shape, data-type, and so forth—the NumPy
+array has become the {\it de facto} exchange format for array data in
+Python.  That said, NumPy has now seen such widespread adoption that
+not only the array object but also its {\it Application Programming
+  Interface} (API) has become ubiquitous, and is imitated by, e.g.,
+popular tensor computation libraries.
+
+Over the past few years, with external financial support, NumPy has
+seen renewed developer interest.  It is actively developed to meet the
+needs of its community, including the addition of some fairly large
+changes as outlined in Section XXX.
+
+\section*{History}
+
+NumPy was, at its inception, the unifier of the Python array
+processing world.  Before NumPy, two packages—Numeric and
+numarray—existed.  The first was developed by Paul DuBois and
+colleagues at [the lab], and was aimed at fast manipulation of many
+small arrays.  In contrast, to handle large astronomical images,
+NumArray was developed [by STSCI?]. In other words, these array
+objects had different strengths and weaknesses, and therefore tended
+to divide the already small user community.  In 200x, Travis Oliphant
+wrote NumPy, a ``best of both worlds'' unification of the two
+projects, based on Numeric.  The arrival of stable version 1.0 in 200x unified the scientific Python
+community, and heralded the era of popular scientific computing in
+Python that has led to formidable successes such as [cite some impacts here].
+
+While, by 2006, NumPy was already eminently usable, it lacked good
+documentation. In order to use NumPy in his classes, Joe Harrington
+from the University of South Florida sponsored several rounds of the
+NumPy Documentation Marathon [citations].  During these intensive
+documentation improvement sessions, a Wikipedia-like system was
+developed to allow community members to contribute so-called
+``docstrings''—text describing each NumPy function. A community
+documentation standard was published so that these pieces of
+documentation would all provide similar information, and conform to a
+basic standard of quality. Experienced community members reviewed and
+integrated contributed docstrings, and soon NumPy had thorough
+documentation coverage.
+
+For many years, NumPy operated without external funding, development
+time mostly being contributed freely by students and researchers in
+their spare time.  In-person meetings were sponsored off of
+individual's research grants, and while industry sponsored
+contributions were made, notably those by Mark Wiebe when he was
+sponsored by Travis Oliphant at X, mostly the package had rapid growth
+and adoption without external investment.  In 2014, NumPy received its
+first large grants from the Moore and Sloan foundations, totaling 1.3M
+dollars.  This allowed for hiring several full time programmers and,
+with the support of community members, mobilized a project that had
+become stagnant for the several years prior.
+
+NumPy is now actively developed, with weekly online developer
+meetings, a formalized NumPy enhancement proposal process, and
+multiple active grants (but still almost no industry support).
+
+For a more detailed history of the scientific Python ecosystem, please
+refer to [SciPy arxiv / nature paper].
+
 \section*{Project organization and community}

### 2019-12-10 paper.tex
Add outline

@@
 
-NumPy is the fundamental array computation library for the Python
-ecosystem.  It underpins almost every library that does numerical
+NumPy\cite{vanderwalt2011numpy} is the fundamental array computation library for the Python
+ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
+It underpins almost every library that does numerical
 computation, including SciPy.  NumPy consists of the `ndarray`-object
@@
 from the University of South Florida sponsored several rounds of the
-NumPy Documentation Marathon [citations].  During these intensive
+NumPy Documentation Marathon \cite{harrington2008scipy,vanderwalt2008scipy}.
+During these intensive
 documentation improvement sessions, a Wikipedia-like system was
@@
 For a more detailed history of the scientific Python ecosystem, please
-refer to [SciPy arxiv / nature paper].
+refer to ``SciPy 1.0---Fundamental Algorithms for Scientific Computing
+in Python'' \cite{virtanen2019scipy}.
+
+\section*{Architecture and implementation choices}
+
+\subsection*{Library organisation}
+
+\subsection*{Common infrastructure}
+
+\subsection*{Language choices}
+
+\section*{Recent technical improvements}
+
+\subsection*{NEPS revived and roadmap}
+
+\subsection*{Random}
+
+\subsection*{\_\_array\_func\_\_}
+
+\subsection*{Testing on multiple architectures}
 

### 2019-12-10 paper.tex
Start NEP section

@@
 
+Following Python's practice of using Python Enhancement Proposals (PEPs) for
+``proposing major new features, for collecting community input on an
+issue, and for documenting the design decisions that have gone into Python,''
+we began using NumPy Enhancement Proposals (NEPs) in 2007.
+This process was informal and only occasionally used until 2017 when
+we adopted a more formal process \footnote{\url{https://numpy.org/neps/nep-0000.html}}.
+Since then there have been 19 proposed NEPS---6 have been implemented,
+4 have been accepted and are being implemented, 4 are under consideration,
+3 have deferred or superseded, and 2 have been rejected or withdrawn.
+We highlight a few of the implemented NEPs below.
+
 \subsection*{Random}

### 2019-12-10 paper.tex
Add impact now

@@
 \usepackage{listings, textcomp}
+\lstset{language=Python}
 
@@
 
+\subsection*{Conclusion}
+
+\subsubsection*{Impact Now}
+
+NumPy, while having been wildly successful as a library and array
+exchange project, has now also become a standard language for expressing array
+and tensor computations.  For example, the NumPy API is implemented by
+several popular tensor computation libraries including CuPy
+(https://cupy.chainer.org/), Jax
+(https://jax.readthedocs.io/en/latest/jax.numpy.html), PyTorch
+(https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html),
+and Apache MXNet
+(https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html),
+and to a lesser extent TensorFlow
+(https://www.tensorflow.org/api\_docs/python/tf/Tensor).
+
+Another notable example is Dask, a library for parallel computing in
+Python.  Dask adopts the NumPy API and therefore presents a familiar
+interface to existing NumPy users, while adding powerful abilities to
+parallelize and distribute tasks.
+
+In order to aid better interaction between NumPy and these libraries,
+an ``array function'' protocol was added to NumPy in 2018 (see Section
+X). The new protocol allows for external array objects (sometimes
+called ``duck arrays'') to pass through NumPy:
+
+\begin{lstlisting}
+  import numpy as np
+  import cupy as cp
+
+  x_gpu = cp.array([1, 2, 3])
+  y = np.sum(x_gpu)  # This works, and returns a GPU array
+\end{lstlisting}
+
+\subsubsection*{Future Development}
+
 \section*{Discussion}

### 2019-12-11 paper.tex
First shot at 'Library Organization'

--- 2019-12-10 paper.tex
@@
 \subsection*{Library organisation}
+Broadly, the NumPy library consists of the following parts:
+the NumPy array data structure; the so-called \emph{universal functions};
+a set of library functions for manipulating arrays and doing scientific
+computation; infrastructure libraries for unit tests and Python package
+building; and the program \code{f2py} for wrapping Fortran code in Python.
+The \code{ndarray} and the universal functions are generally considered
+the core of the library.
+In the following, we give a brief summary of these components of the
+library.
+
+\paragraph{\emph{Core.}}  The \code{ndarray} data structure and the
+universal functions make up the core of NumPy.
+
+The \code{ndarray} is the data structure at the heart of NumPy.
+The data structure stores regularly strided homogeneous data types
+in a contiguous block memory, allowing for the efficient representation
+of $n$-dimensional data. \fixme{[How much detail do we want here?]}
+More details about the data structure are given in ``The NumPy array:
+a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.
+
+The \emph{universal functions}, or more concisely, \emph{ufuncs},
+are functions written in C that implement efficient looping over
+NumPy arrays. An important feature of ufuncs is the built-in
+implementation of \emph{broadcasting}.  For example, the function
+\code{arctan2(x, y)} is a ufunc that accepts two values and computes
+$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
+the ufunc will take care of looping over the dimensions of the inputs
+in such a way that if, say, \code{x} is a 1-d array with length 3, and
+\code{y} is a 2-d array with shape $2 \times 1$, the output will be
+an array with shape $2 \times 3$.  The ufunc machinery takes care
+of calling the function with all the appropriate combinations of
+input array elements to complete the output array.
+The elementary arithmetic operations of addition, multiplication, etc.,
+are implemented as ufuncs, so broadcasting applies even in expressions
+such as \code{x + y * z}.
+
+\paragraph{\emph{Computing libraries.}}
+NumPy provides a large library of functions for array manipulation
+and scientific computing, including functions for: creating, reshaping,
+concatenating, and padding arrays; searching, sorting and counting data
+in arrays; computing elementary statistics, such as the mean, median,
+variance, and standard deviation; file I/O; and more.
+
+A suite of functions for computing the \emph{fast Fourier transform (FFT)}
+and its inverse is provided.
+
+NumPy's linear algebra library includes functions for: solving linear
+systems of equations; computing various functions of a matrix, including
+the determinant, the norm, the inverse, and the pseudo-inverse;
+computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
+and more.
+
+The random number generator library in NumPy provides several alternative
+\emph{bit stream generators} that provide the core function of generating
+random integers.
+A higher-level generator class that implements an assortment of
+probability distributions is provided. It includes the beta, gamma
+and Weibull distributions, the univariate and multivariate normal
+distributions, and more.
+
+A \emph{masked array} class, built on top of the \code{ndarray}, is
+provided.  This class allows elements of an array to be masked,
+indicating that the data is missing or unknown.
+
+There is also a \code{matrix} class the implements 2-d arrays with
+specialized behavior.  We mention it here for completeness, but the
+use of \code{matrix} is discouraged.  The quirks of the \code{matrix}
+class have proven to be troublesome, and with the addition of the
+infix \code{@} operator to Python for matrix multiplication, there was no
+longer a need for a separate \code{matrix} class, since one can
+use a 2-d array instead.
+
+\paragraph{\emph{Infrastructure libraries.}} NumPy provides utilities
+for writing unit tests and for building Python packages.
+
+The \code{testing} subpackage provides functions such as
+\code{assert\_allclose(actual, desired)} that may be used in unit
+test suites for code that uses NumPy arrays.
+
+NumPy provides a subpackage includes functions and classes
+that can be used by libraries that have NumPy as dependency,
+to be used in the library's configuration code that installs the
+library and that builds a package that can be used, for example,
+on the PyPI web site.
+
+\paragraph{\emph{F2PY.}}  The program \code{f2py} is a tool for
+building NumPy-aware Python wrappers of Fortran functions.
+NumPy itself does not use any Fortran code;  F2PY is part of NumPy
+for historical reasons.
 

### 2019-12-11 paper.tex

@@
 \subsection*{Funding}
+
+\subsection*{Conclusion}
+
+\subsubsection*{Impact Now}
+
+NumPy, while having been wildly successful as a library and array
+exchange project, has now also become a standard language for expressing array
+and tensor computations.  For example, the NumPy API is implemented by
+several popular tensor computation libraries including CuPy
+(https://cupy.chainer.org/), Jax
+(https://jax.readthedocs.io/en/latest/jax.numpy.html), PyTorch
+(https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html),
+and Apache MXNet
+(https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html),
+and to a lesser extent TensorFlow
+(https://www.tensorflow.org/api\_docs/python/tf/Tensor).
+
+Another notable example is Dask, a library for parallel computing in
+Python.  Dask adopts the NumPy API and therefore presents a familiar
+interface to existing NumPy users, while adding powerful abilities to
+parallelize and distribute tasks.
+
+In order to aid better interaction between NumPy and these libraries,
+an ``array function'' protocol was added to NumPy in 2018 (see Section
+X). The new protocol allows for external array objects (sometimes
+called ``duck arrays'') to pass through NumPy:
+
+\begin{lstlisting}
+  import numpy as np
+  import cupy as cp
+
+  x_gpu = cp.array([1, 2, 3])
+  y = np.sum(x_gpu)  # This works, and returns a GPU array
+\end{lstlisting}
+
+\subsubsection*{Future Development}
 

### 2019-12-12 paper.tex
Minor edits

@@
 
-NumPy\cite{vanderwalt2011numpy} is the fundamental array computation library for the Python
+NumPy is the fundamental array computation library for the Python
 ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
@@
 needs of its community, including the addition of some fairly large
-changes as outlined in Section XXX.
+changes as outlined in Section~\ref{improvements}.
 
-\section*{History}
+\subsection*{History}
 
@@
 processing world.  Before NumPy, two packages—Numeric and
-numarray—existed.  The first was developed by Paul DuBois and
-colleagues at [the lab], and was aimed at fast manipulation of many
+numarray—existed.  The first was developed by Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
+Hinsen, Travis Oliphant, and others, and was aimed at fast manipulation of many
 small arrays.  In contrast, to handle large astronomical images,
-NumArray was developed [by STSCI?]. In other words, these array
+NumArray was developed at the Space Telescope Science Institute (STScI).
+In other words, these array
 objects had different strengths and weaknesses, and therefore tended
-to divide the already small user community.  In 200x, Travis Oliphant
-wrote NumPy, a ``best of both worlds'' unification of the two
-projects, based on Numeric.  The arrival of stable version 1.0 in 200x unified the scientific Python
+to divide the already small user community.  In 2005, Travis Oliphant
+started NumPy, a ``best of both worlds'' unification of the two
+projects, based on Numeric.  The arrival of stable version 1.0 in October
+2006 unified the scientific Python
 community, and heralded the era of popular scientific computing in
@@
 contributions were made, notably those by Mark Wiebe when he was
-sponsored by Travis Oliphant at X, mostly the package had rapid growth
+sponsored by Travis Oliphant at Continuum Analytics, mostly the package had rapid growth
 and adoption without external investment.  In 2014, NumPy received its
@@
 
+\subsection*{Project goals / scope of library}
+
+\subsection*{Current status (maturity, users)}
+
+\subsection*{Enabled ecosystem }
+
+% NEP 29 (which releases are supported in rolling window)
+
+% NumPyDoc standard, recent improvements
+
 \section*{Architecture and implementation choices}
 
-\subsection*{Library organisation}
+\subsection*{Library organization}
 Broadly, the NumPy library consists of the following parts:
@@
 computation; infrastructure libraries for unit tests and Python package
-building; and the program \code{f2py} for wrapping Fortran code in Python.
+building; and the program \code{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
 The \code{ndarray} and the universal functions are generally considered
@@
 in a contiguous block memory, allowing for the efficient representation
-of $n$-dimensional data. \fixme{[How much detail do we want here?]}
+of $n$-dimensional data.
 More details about the data structure are given in ``The NumPy array:
@@
 
+% Wheels build system
+
 \subsection*{Language choices}
 
-\section*{Recent technical improvements}
+\section*{Recent technical improvements}\label{improvements}
 
@@
 
-\subsection*{Conclusion}
+\section*{Discussion}
 
-\subsubsection*{Impact Now}
+% The Discussion should be succinct and must not contain subheadings.
+
+% \subsubsection*{Impact Now}
 
@@
 and tensor computations.  For example, the NumPy API is implemented by
-several popular tensor computation libraries including CuPy
-(https://cupy.chainer.org/), Jax
-(https://jax.readthedocs.io/en/latest/jax.numpy.html), PyTorch
-(https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html),
-and Apache MXNet
-(https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html),
-and to a lesser extent TensorFlow
-(https://www.tensorflow.org/api\_docs/python/tf/Tensor).
+several popular tensor computation libraries including
+CuPy\footnote{\url{https://cupy.chainer.org/}},
+Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
+Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
+and to a lesser extent
+TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.
 
@@
 
-\subsubsection*{Future Development}
-
-\section*{Discussion}
-
-% The Discussion should be succinct and must not contain subheadings.
+% \subsubsection*{Future Development}
 
@@
 % The statement should be provided as a separate section (titled 'Data Availability') at the end of the main text, before the 'References' section.
-All NumPy source code and most data generated for the current study are available in the NumPy GitHub repository, \url{https://github.com/numpy}.
-Some supporting code and data have also been stored in other public repositories cited by this manuscript.
+
+All NumPy source code and most data generated for the current study are
+available in the NumPy GitHub repository, \url{https://github.com/numpy}. Some
+supporting code and data have also been stored in other public repositories
+cited by this manuscript.
 

### 2019-12-12 paper.tex
Fix reference to improvement section

@@
 seen renewed developer interest.  It is actively developed to meet the
-needs of its community, including the addition of some fairly large
-changes as outlined in Section~\ref{improvements}.
+needs of its community, including the addition of some fairly significant
+changes as outlined in the Recent technical improvements section below.
 
@@
 
-\section*{Recent technical improvements}\label{improvements}
+\section*{Recent technical improvements}
 

### 2019-12-13 paper.tex
Add missing sections from outline
ENH: draft CI architectures section
* first draft of section on testing multiple
CPU architectures in CI

@@
 
+At the time of writing the two fastest supercomputers in the
+world, Summit and Sierra, both have IBM POWER9 architectures
+\cite{top500nov2019}. In late 2018, Astra, the first ARM-based
+supercomputer to enter the TOP500 list, went into production\cite{
+astra-wiki}. Furthermore, over 100 billion ARM processors have been
+produced as of 2017\cite{arm-architecture}, making it the most 
+widely used instruction set architecture in the world.
+
+Clearly there are motivations for a large scientific computing
+software library to support POWER and ARM architectures. We've extended
+our continuous integration (CI) testing to include \texttt{ppc64le}
+(POWER8 on Travis CI) and ARMv8 (on Shippable service). We also test
+with the s390x architecture (IBM Z CPUs on Travis CI) so that we
+can probe the behavior of our library on a big-endian machine.
+This satisfies one of the major components of
+improved CI testing laid out in a version of our roadmap
+\cite{numpy-roadmap}---specifically, ``CI for more exotic
+platforms."
+
+PEP 599\cite{PEP599} lays out a plan for new Python binary wheel
+distribution support, \texttt{manylinux2014}, that adds
+support for a number of architectures supported by the CentOS
+Alternative Architecture Special Interest Group, including
+ARMv8, ppc64le, as well as s390x. We are thus well-positioned
+for a future where provision of binaries on these architectures
+will be expected for a library at the base of the ecosystem.
+
 \section*{Project organization and community}
@@
 \subsection*{Funding}
+
+% BIDS -- UCB, CZI -- NumFOCUS/QuanSight, Tidelift, ...
+
+\subsection*{GSoD}
+
+\subsection*{Maintainer overview}
+
+%  nr of core devs / commit rights) & contributors
+
+\subsection*{Industry collaboration}
+
+% Intel, QuanSight, ...
+
+\subsection*{Website}
 

### 2019-12-16 paper.tex
NEW: Draft of section for `__array_function__`.

@@
 
+For a more detailed history of the scientific Python ecosystem, please
+refer to ``SciPy 1.0---Fundamental Algorithms for Scientific Computing
+in Python'' \cite{virtanen2019scipy}.
+
+\subsection*{Funding}
+
 For many years, NumPy operated without external funding, development
@@
 sponsored by Travis Oliphant at Continuum Analytics, mostly the package had rapid growth
-and adoption without external investment.  In 2014, NumPy received its
-first large grants from the Moore and Sloan foundations, totaling 1.3M
-dollars.  This allowed for hiring several full time programmers and,
-with the support of community members, mobilized a project that had
-become stagnant for the several years prior.
+and adoption without external investment.
+
+% In 2014, NumPy received its
+% first large grants from the Moore and Sloan foundations, totaling 1.3M
+% dollars.  This allowed for hiring several full time programmers and,
+% with the support of community members, mobilized a project that had
+% become stagnant for the several years prior.
+
+% BIDS -- UCB
+% https://www.moore.org/grant-detail?grantId=GBMF5447
+% $645,020 in 2016
+% https://sloan.org/grant-detail/8222
+% $659,359 in 2017
+% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
+
+The Berkeley Institute for Data Science (BIDS) received two grants to support
+the development of NumPy.
+At the end of 2016, the Gordon and Betty Moore Foundation awarded them
+\$645,020 for improving NumPy for better data science.
+In 2017, the Alfred P. Sloan Foundation complemented the funding with an
+additional \$659,359.
+Stéfan van der Walt, a senior research data scientist at BIDS and a member of
+the NumPy steering committee, is the PI of both grants and manages a team of
+four programmers working on the project.
+
+% how much do we say about who these developers are and
+% lessons learned?
+
+% technical focus 
+% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
+% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
+
+% CZI -- NumFOCUS/QuanSight
+% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
+% NumPy and OpenBLAS received $195,000 in 2019
+% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
+
+NumFOCUS received a grant for \$195,000 in 2019 for the development of NumPy
+and OpenBLAS from the Chan Zuckerberg Initiative. 
+NumPy depends on OpenBLAS to provide its underlying accelerated linear algebra implementation.
+Ralf Gommers, the Director of Quansight Labs and a member of
+the NumPy steering committee, is the PI.
+
+% focus on better serve NumPy's large number of beginner to intermediate level users
+% focusing all our time on growing the team and better serving new users and contributors
+
+% Tidelift, ...
+% https://tidelift.com/subscription/pkg/pypi-numpy
+% $10,000 over 24 months according to 
+% https://mail.python.org/pipermail/numpy-discussion/2018-September/078736.html
+
+% Intel, QuanSight, ...
+
+\subsection*{Current status}
 
@@
 multiple active grants (but still almost no industry support).
+% check almost no industry
 
-For a more detailed history of the scientific Python ecosystem, please
-refer to ``SciPy 1.0---Fundamental Algorithms for Scientific Computing
-in Python'' \cite{virtanen2019scipy}.
+% (maturity, users)
 
-\subsection*{Project goals / scope of library}
+\subsubsection*{Project goals / scope of library}
 
-\subsection*{Current status (maturity, users)}
+% https://numpy.org/neps/scope.html
 
-\subsection*{Enabled ecosystem }
+\subsubsection*{NEPS revived and roadmap}
 
-% NEP 29 (which releases are supported in rolling window)
+Following Python's practice of using Python Enhancement Proposals (PEPs) for
+``proposing major new features, for collecting community input on an
+issue, and for documenting the design decisions that have gone into Python,''
+we began using NumPy Enhancement Proposals (NEPs) in 2007.
+This process was informal and only occasionally used until 2017 when
+we adopted a more formal process \footnote{\url{https://numpy.org/neps/nep-0000.html}}.
+Since then there have been 19 proposed NEPS---6 have been implemented,
+4 have been accepted and are being implemented, 4 are under consideration,
+3 have deferred or superseded, and 2 have been rejected or withdrawn.
+We highlight a few of the implemented NEPs below.
 
-% NumPyDoc standard, recent improvements
+% https://numpy.org/neps/roadmap.html
 
@@
 
-% Wheels build system
+\subsubsection*{Wheels build system}
+
+A Python \emph{wheel}\cite{PEP427} is a standard file format for
+distributing Python libraries.  In addition to Python code, a
+wheel may include compiled C extensions and other binary data.
+This is important, because many libraries, including NumPy,
+require a C compiler and other build tools to build the software
+from the source code, making it difficult for many users to install
+the software on their own.  The introduction of wheels to the Python
+packaging system has made it much easier for users to install
+precompiled libraries.
+
+A GitHub repository containing scripts to build NumPy wheels has
+been configured so that a simple commit to the repository triggers
+an automated build system that creates NumPy wheels for several
+computer platforms, including Windows, Mac OSX and Linux.  The wheels
+are uploaded to a public server and made available for anyone to use.
+This system makes it easy for users to install precompiled versions
+of NumPy on these platforms.
+
+The technology that is used to build the wheels evolves continually.
+At the time this paper is being written, a key component is the
+\code{multibuild} suite of tools developed by Matthew Brett and
+other developers\cite{multibuild}.  Currently, scripts using
+\code{multibuild} are written for the continuous integration
+platforms Travis-CI (for Linux and Mac OSX) and Appveyor
+(for Windows).
+
+\subsubsection*{NEP 29}
+
+% https://numpy.org/neps/nep-0029-deprecation_policy.html
+In NEP 29, we recommend, along with leaders from various other
+projects, that all projects across the Scientific Python ecosystem
+adopt a common ``time window-based'' policy for support of Python and
+NumPy versions. This standard will simplify downstream project and
+release planning.
+
+\subsubsection*{NumPyDoc standard}
+
+% recent improvements
 
@@
 
-\subsection*{NEPS revived and roadmap}
-
-Following Python's practice of using Python Enhancement Proposals (PEPs) for
-``proposing major new features, for collecting community input on an
-issue, and for documenting the design decisions that have gone into Python,''
-we began using NumPy Enhancement Proposals (NEPs) in 2007.
-This process was informal and only occasionally used until 2017 when
-we adopted a more formal process \footnote{\url{https://numpy.org/neps/nep-0000.html}}.
-Since then there have been 19 proposed NEPS---6 have been implemented,
-4 have been accepted and are being implemented, 4 are under consideration,
-3 have deferred or superseded, and 2 have been rejected or withdrawn.
-We highlight a few of the implemented NEPs below.
-
 \subsection*{Random}
 
-\subsection*{\_\_array\_func\_\_}
+\subsection*{The \_\_array\_function\_\_ Protocol}
+
+The central position which NumPy takes within the python ecosystem with
+respect to scientific computing and numerical works,
+means that a vast number of projects are build on it.
+These projects are consumers of the NumPy API.
+
+On the other side, projects are providers of a
+\emph{NumPy-like API} and array objects targeting audiences with specialized needs beyond
+NumPy's capabilities. Some of these projects provide:
+
+% TODO: A list prints too large currently, so probably should make this
+%       text (although depends a bit on the final layout maybe)
+\begin{description}
+    \item[SciPy's \code{scipy.sparse} and \code{pydata.sparse}:] Sparse arrays
+    \item[CuPy:] Computation on the GPU
+    \item[Dask:] Parallel, distributed computing
+    \item[PyTorch and TensorFlow:] Deep learning frameworks.
+\end{description}
+
+The multitude of specialized projects creates the difficulty that consumers
+of these NumPy-like APIs write code specific to a single project and do not support
+all of the above array providers.
+This is a burden for users relying on the specialized Array-like, since
+a tool they need may not work for them.
+Additionally, a user who has the need of transitioning from NumPy to a more
+specialized array, is faced with a challenge.
+These issue can lead into a fragmentation of the user bases into consumers
+of different NumPy-like APIs.
+
+To address these issues NumPy has the goal of providing the fundamental
+API for \emph{interoperability} between the various NumPy-like APIs.
+An earlier step in this direction was the implementation of the
+\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
+for most mathematical functions.\cite{NEP13}
+In 2019 this was expanded more generally with the inclusion of the
+\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
+These two protocols allows providers of Array-like APIs to be interoperable
+with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
+For the users relying on specialized array projects it means that even though
+much code is written specifically for NumPy arrays and uses the NumPy API as
+\code{import numpy as np}, it can nevertheless work for them.
+A brief example of such a function which is written using NumPy
+but can now be used more generally is:
+\begin{lstlisting}
+def f(x):
+    y = np.tensordot(x, x.T)
+    return np.mean(np.exp(y))
+\end{lstlisting}
+For example, a user relying on this function and normally using it with a
+NumPy array, may now use a distributed Dask array when challenged
+by a very large dataset, without any modification to \code{f(x)}.
+
+While, as of writing this, the interface is still considered experimental
+and may be modified, it has been implemented by multiple projects already
+and proven useful.
+% TODO: Should likely have citation/example if I claim it was useful (based
+%       mainly on Matt's statement).
 
@@
 
+% incl. CoC
+
 \subsection*{Maintainers and contributors}
-
-\subsection*{Funding}
-
-% BIDS -- UCB, CZI -- NumFOCUS/QuanSight, Tidelift, ...
-
-\subsection*{GSoD}
-
-\subsection*{Maintainer overview}
 
@@
 
-\subsection*{Industry collaboration}
+% may be useful:
+% - https://github.com/docathon/watchtower
+% - https://github.com/machine-shop/oss_devkit/blob/master/plot_pr.py % (PR over time plot, there may be other better versions out there)
 
-% Intel, QuanSight, ...
+\subsection*{GSoD}
 
@@
 
-% The Discussion should be succinct and must not contain subheadings.
-
-% \subsubsection*{Impact Now}
+\subsubsection*{Impact Now}
 
@@
 
-% \subsubsection*{Future Development}
+\subsubsection*{Future Development}
+
+% Dtypes
+
+The meaning of each element within a NumPy array is described by its datatype
+(dtype). NumPy arrays can hold all common numerical datatypes, string, datetimes
+and generic Python objects, each of these is identified by a dtype.
+We are busy overhauling the datatype system to make it consistent (aligned with
+expectations), to simplify the creation of custom dtypes.
+While numerical dtypes make the foundation for most users, there are
+a multitude of use cases which cannot prosper due to current limitations.
+These use cases include datatypes with physical units\cite{astropy,Goldbaum2018,pint},
+% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
+geometrical objects\cite{pygeos}, automatic differentiation\cite{pyadolc}, and many more.
+Projects addressing these exist, each struggles with current
+limitations.  % TODO: I may need citations, we could link github issues.
+
+While it is currently possible to define custom dtypes within NumPy, many of the
+use cases above cannot be solved using this system.
+This is evidenced also by code within NumPy which would require changes
+in many places to add such dtypes, even when adding them within NumPy itself.
+The new implementation will be simpler and more cleanly implemented in
+this regard.
+Custom dtypes and the basic dtypes included within NumPy will be implemented
+and behave identically as much as possible.
+At the same time redesigning the API will make NumPy dtypes more flexible to grow
+to future needs should they arise.
+
+As a further step, to make dtypes more accessible and allow a faster adoption of
+custom dtypes within the community, we plan to create a Python API to
+define such custom dtypes.
+
+A \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}
+is under development in consultation with the community.
 
 \section*{Data Availability}
-% The statement should be provided as a separate section (titled 'Data Availability') at the end of the main text, before the 'References' section.
 
@@
 
-%To include, in this order: \textbf{Accession codes} (where applicable);
-%\textbf{Competing financial interests} (mandatory statement).
-
-%The corresponding author is responsible for submitting a
-%\href{http://www.nature.com/srep/policies/index.html#competing}{competing
-%financial interests statement} on behalf of all authors of the paper. This
-%statement must be included in the submitted article file.
-
 \section*{Consortium}

### 2019-12-17 paper.tex
STY: (array_function) Use tightlist for array-like enumeration
Start funding section

--- 2019-12-13 paper.tex
@@
 
-% BIDS -- UCB, CZI -- NumFOCUS/QuanSight, Tidelift, ...
+Historically, NumPy was developed primarily by voluteers in their free time.
+% some funding at Berkeley, the documentation project, Travis funding ...
+Over the last few years, there has an increase in paid full-time developers due
+to new sources of funding.
+
+% how much do we say about who these developers are and
+% lessons learned?
+
+% BIDS -- UCB
+% https://www.moore.org/grant-detail?grantId=GBMF5447
+% $645,020 in 2016
+% https://sloan.org/grant-detail/8222
+% $659,359 in 2017
+% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
+
+The Berkeley Institute for Data Science (BIDS) received two grants to support
+the development of NumPy.
+At the end of 2016, the Gordon and Betty Moore Foundation awarded them
+\$645,020 for improving NumPy for better data science.
+In 2017, the Alfred P. Sloan Foundation complemented the funding with an
+additional \$659,359.
+Stéfan van der Walt, a senior research data scientist at BIDS and a member of
+the NumPy steering committee, is the PI of both grants and manages a team of
+four programmers working on the project.
+
+% technical focus 
+% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
+% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
+
+% CZI -- NumFOCUS/QuanSight
+% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
+% NumPy and OpenBLAS received $195,000 in 2019
+% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
+
+NumFOCUS received a grant for \$195,000 in 2019 for the development of NumPy
+and OpenBLAS from the Chan Zuckerberg Initiative. 
+Ralf Gommers, the Director of Quansight Labs and a member of
+the NumPy steering committee, is the PI.
+
+% focus on better serve NumPy's large number of beginner to intermediate level users
+% focusing all our time on growing the team and better serving new users and contributors
+
+% Tidelift, ...
+% https://tidelift.com/subscription/pkg/pypi-numpy
+% $10,000 over 24 months according to 
+% https://mail.python.org/pipermail/numpy-discussion/2018-September/078736.html
 

### 2019-12-17 paper.tex
Move funding to intro

@@
 sponsored by Travis Oliphant at Continuum Analytics, mostly the package had rapid growth
-and adoption without external investment.  In 2014, NumPy received its
-first large grants from the Moore and Sloan foundations, totaling 1.3M
-dollars.  This allowed for hiring several full time programmers and,
-with the support of community members, mobilized a project that had
-become stagnant for the several years prior.
+and adoption without external investment.
+
+% In 2014, NumPy received its
+% first large grants from the Moore and Sloan foundations, totaling 1.3M
+% dollars.  This allowed for hiring several full time programmers and,
+% with the support of community members, mobilized a project that had
+% become stagnant for the several years prior.
+
+% BIDS -- UCB
+% https://www.moore.org/grant-detail?grantId=GBMF5447
+% $645,020 in 2016
+% https://sloan.org/grant-detail/8222
+% $659,359 in 2017
+% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
+
+The Berkeley Institute for Data Science (BIDS) received two grants to support
+the development of NumPy.
+At the end of 2016, the Gordon and Betty Moore Foundation awarded them
+\$645,020 for improving NumPy for better data science.
+In 2017, the Alfred P. Sloan Foundation complemented the funding with an
+additional \$659,359.
+Stéfan van der Walt, a senior research data scientist at BIDS and a member of
+the NumPy steering committee, is the PI of both grants and manages a team of
+four programmers working on the project.
+
+% how much do we say about who these developers are and
+% lessons learned?
+
+% technical focus 
+% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
+% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
+
+% CZI -- NumFOCUS/QuanSight
+% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
+% NumPy and OpenBLAS received $195,000 in 2019
+% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
+
+NumFOCUS received a grant for \$195,000 in 2019 for the development of NumPy
+and OpenBLAS from the Chan Zuckerberg Initiative. 
+Ralf Gommers, the Director of Quansight Labs and a member of
+the NumPy steering committee, is the PI.
+
+% focus on better serve NumPy's large number of beginner to intermediate level users
+% focusing all our time on growing the team and better serving new users and contributors
+
+% Tidelift, ...
+% https://tidelift.com/subscription/pkg/pypi-numpy
+% $10,000 over 24 months according to 
+% https://mail.python.org/pipermail/numpy-discussion/2018-September/078736.html
+
+% Intel, QuanSight, ...
+
+\subsection*{Current status}
 
@@
 multiple active grants (but still almost no industry support).
+% check almost no industry
 
-For a more detailed history of the scientific Python ecosystem, please
-refer to ``SciPy 1.0---Fundamental Algorithms for Scientific Computing
-in Python'' \cite{virtanen2019scipy}.
+% (maturity, users)
+
+\section*{Architecture and implementation choices}
 
@@
 
-\subsection*{Current status (maturity, users)}
-
-\subsection*{Enabled ecosystem }
-
-% NEP 29 (which releases are supported in rolling window)
-
-% NumPyDoc standard, recent improvements
-
-\section*{Architecture and implementation choices}
+% https://numpy.org/neps/scope.html
 
@@
 
+% NEP 29 (which releases are supported in rolling window)
+
+% NumPyDoc standard, recent improvements
+
 \subsection*{Language choices}
@@
 We highlight a few of the implemented NEPs below.
+
+% https://numpy.org/neps/roadmap.html
 
@@
 
-\subsection*{Maintainers and contributors}
-
-\subsection*{Funding}
-
-Historically, NumPy was developed primarily by voluteers in their free time.
-% some funding at Berkeley, the documentation project, Travis funding ...
-Over the last few years, there has an increase in paid full-time developers due
-to new sources of funding.
-
-% how much do we say about who these developers are and
-% lessons learned?
-
-% BIDS -- UCB
-% https://www.moore.org/grant-detail?grantId=GBMF5447
-% $645,020 in 2016
-% https://sloan.org/grant-detail/8222
-% $659,359 in 2017
-% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
-
-The Berkeley Institute for Data Science (BIDS) received two grants to support
-the development of NumPy.
-At the end of 2016, the Gordon and Betty Moore Foundation awarded them
-\$645,020 for improving NumPy for better data science.
-In 2017, the Alfred P. Sloan Foundation complemented the funding with an
-additional \$659,359.
-Stéfan van der Walt, a senior research data scientist at BIDS and a member of
-the NumPy steering committee, is the PI of both grants and manages a team of
-four programmers working on the project.
-
-% technical focus 
-% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
-% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
-
-% CZI -- NumFOCUS/QuanSight
-% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
-% NumPy and OpenBLAS received $195,000 in 2019
-% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
-
-NumFOCUS received a grant for \$195,000 in 2019 for the development of NumPy
-and OpenBLAS from the Chan Zuckerberg Initiative. 
-Ralf Gommers, the Director of Quansight Labs and a member of
-the NumPy steering committee, is the PI.
-
-% focus on better serve NumPy's large number of beginner to intermediate level users
-% focusing all our time on growing the team and better serving new users and contributors
-
-% Tidelift, ...
-% https://tidelift.com/subscription/pkg/pypi-numpy
-% $10,000 over 24 months according to 
-% https://mail.python.org/pipermail/numpy-discussion/2018-September/078736.html
-
 \subsection*{GSoD}
 
-\subsection*{Maintainer overview}
+\subsection*{Maintainers and contributors}
 
@@
 
-\subsection*{Industry collaboration}
-
-% Intel, QuanSight, ...
+% may be useful:
+% - https://github.com/docathon/watchtower
+% - https://github.com/machine-shop/oss_devkit/blob/master/plot_pr.py % (PR over time plot, there may be other better versions out there)
 

### 2019-12-17 paper.tex
Update outline
Add NEP 29 paragraph

@@
 and OpenBLAS from the Chan Zuckerberg Initiative. 
+NumPy depends on OpenBLAS to provide its underlying accelerated linear algebra implementation.
 Ralf Gommers, the Director of Quansight Labs and a member of
@@
 
-\section*{Architecture and implementation choices}
-
-\subsection*{Project goals / scope of library}
+\subsubsection*{Project goals / scope of library}
 
 % https://numpy.org/neps/scope.html
+
+\subsubsection*{NEPS revived and roadmap}
+
+Following Python's practice of using Python Enhancement Proposals (PEPs) for
+``proposing major new features, for collecting community input on an
+issue, and for documenting the design decisions that have gone into Python,''
+we began using NumPy Enhancement Proposals (NEPs) in 2007.
+This process was informal and only occasionally used until 2017 when
+we adopted a more formal process \footnote{\url{https://numpy.org/neps/nep-0000.html}}.
+Since then there have been 19 proposed NEPS---6 have been implemented,
+4 have been accepted and are being implemented, 4 are under consideration,
+3 have deferred or superseded, and 2 have been rejected or withdrawn.
+We highlight a few of the implemented NEPs below.
+
+% https://numpy.org/neps/roadmap.html
+
+\section*{Architecture and implementation choices}
 
@@
 
+% https://numpy.org/neps/nep-0029-deprecation_policy.html
+In NEP 29, we recommend, along with leaders from various other
+projects, that all projects across the Scientific Python ecosystem
+adopt a common ``time window-based'' policy for support of Python and
+NumPy versions. This standard will simplify downstream project and
+release planning.
+
 % NumPyDoc standard, recent improvements
@@
 \section*{Recent technical improvements}
-
-\subsection*{NEPS revived and roadmap}
-
-Following Python's practice of using Python Enhancement Proposals (PEPs) for
-``proposing major new features, for collecting community input on an
-issue, and for documenting the design decisions that have gone into Python,''
-we began using NumPy Enhancement Proposals (NEPs) in 2007.
-This process was informal and only occasionally used until 2017 when
-we adopted a more formal process \footnote{\url{https://numpy.org/neps/nep-0000.html}}.
-Since then there have been 19 proposed NEPS---6 have been implemented,
-4 have been accepted and are being implemented, 4 are under consideration,
-3 have deferred or superseded, and 2 have been rejected or withdrawn.
-We highlight a few of the implemented NEPs below.
-
-% https://numpy.org/neps/roadmap.html
 
@@
 
-\subsection*{GSoD}
+% incl. CoC
 
@@
 
+\subsection*{GSoD}
+
 \subsection*{Website}
@@
 
-% \subsubsection*{Future Development}
+\subsubsection*{Future Development}
+
+% Dtypes
+
+We are busy overhauling the datatype system to make it consistent (aligned with
+expectations), to simplify the creation of custom dtypes, and to enable much
+requested features such as units.  The new implementation should be simpler and
+more cleanly implemented.
+
+A \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}
+is under development in consultation with the community.
 
 \section*{Data Availability}
-% The statement should be provided as a separate section (titled 'Data Availability') at the end of the main text, before the 'References' section.
 

### 2019-12-17 paper.tex
Update outline
Add a few paragraphs about wheel building.

@@
 
-% Wheels build system
+\subsubsection*{Wheels build system}
 
-% NEP 29 (which releases are supported in rolling window)
+A Python \emph{wheel}\cite{PEP427} is a standard file format for
+distributing Python libraries.  In addition to Python code, a
+wheel may include compiled C extensions and other binary data.
+This is important, because many libraries, including NumPy,
+require a C compiler and other build tools to build the software
+from the source code, making it difficult for many users to install
+the software on their own.  The introduction of wheels to the Python
+packaging system has made it much easier for users to install
+precompiled libraries.
+
+A GitHub repository containing scripts to build NumPy wheels has
+been configured so that a simple commit to the repository triggers
+an automated build system that creates NumPy wheels for several
+computer platforms, including Windows, Mac OSX and Linux.  The wheels
+are uploaded to a public server and made available for anyone to use.
+This system makes it easy for users to install precompiled versions
+of NumPy on these platforms.
+
+The technology that is used to build the wheels evolves continually.
+At the time this paper is being written, a key component is the
+\code{multibuild} suite of tools developed by Matthew Brett and
+other developers\cite{multibuild}.  Currently, scripts using
+\code{multibuild} are written for the continuous integration
+platforms Travis-CI (for Linux and Mac OSX) and Appveyor
+(for Windows).
+
+\subsubsection*{NEP 29}
 
@@
 
-% NumPyDoc standard, recent improvements
+\subsubsection*{NumPyDoc standard}
+
+% recent improvements
 

### 2019-12-17 paper.tex
NEW: Start on a new dtypes section

@@
 
+The meaning of each element within a NumPy array is described by its datatype
+(dtype). NumPy arrays can hold all common numerical datatypes, string, datetimes
+and generic Python objects, each of these is identified by a dtype.
 We are busy overhauling the datatype system to make it consistent (aligned with
-expectations), to simplify the creation of custom dtypes, and to enable much
-requested features such as units.  The new implementation should be simpler and
-more cleanly implemented.
+expectations), to simplify the creation of custom dtypes.
+While numerical dtypes make the foundation for most users, there are
+a multitude of use cases which cannot prosper due to current limitations.
+These use cases include datatypes with physical units\cite{astropy,Goldbaum2018,pint},
+% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
+geometrical objects\cite{pygeos}, automatic differentiation\cite{pyadolc}, and many more.
+Projects addressing these exist, each struggles with current
+limitations.  % TODO: I may need citations, we could link github issues.
+
+While it is currently possible to define custom dtypes within NumPy, many of the
+use cases above cannot be solved using this system.
+This is evidenced also by code within NumPy which would require changes
+in many places to add such dtypes, even when adding them within NumPy itself.
+The new implementation will be simpler and more cleanly implemented in
+this regard.
+Custom dtypes and the basic dtypes included within NumPy will be implemented
+and behave identically as much as possible.
+At the same time redesigning the API will make NumPy dtypes more flexible to grow
+to future needs should they arise.
+
+As a further step, to make dtypes more accessible and allow a faster adoption of
+custom dtypes within the community, we plan to create a Python API to
+define such custom dtypes.
 

### 2019-12-17 paper.tex
NEW: Start on contributions section and create plot/script for that
Note that the pickle data is not included here, because it contains
semi-private data (although it probably does not matter), plus it
is a bit large at 1.5MB

@@
 \newcommand{\inlinecite}[1]{\footnotesize\citen{#1}}
+\newcommand{\tightlist}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 
-\subsection*{\_\_array\_func\_\_}
+\subsection*{The \_\_array\_function\_\_ Protocol}
+
+The central position which NumPy takes within the python ecosystem with
+respect to scientific computing and numerical works,
+means that a vast number of projects are build on it.
+These projects are consumers of the NumPy API.
+
+On the other side, projects are providers of a
+\emph{NumPy-like API} and array objects targeting audiences with specialized needs beyond
+NumPy's capabilities. Some of these projects provide:
+
+\begin{description}
+    \tightlist
+    \item[SciPy's \code{scipy.sparse} and \code{pydata.sparse}:] Sparse arrays
+    \item[CuPy:] Computation on the GPU
+    \item[Dask:] Parallel, distributed computing
+    \item[PyTorch and TensorFlow:] Deep learning frameworks.
+\end{description}
+
+The multitude of specialized projects creates the difficulty that consumers
+of these NumPy-like APIs write code specific to a single project and do not support
+all of the above array providers.
+This is a burden for users relying on the specialized Array-like, since
+a tool they need may not work for them.
+Additionally, a user who has the need of transitioning from NumPy to a more
+specialized array, is faced with a challenge.
+These issue can lead into a fragmentation of the user bases into consumers
+of different NumPy-like APIs.
+
+To address these issues NumPy has the goal of providing the fundamental
+API for \emph{interoperability} between the various NumPy-like APIs.
+An earlier step in this direction was the implementation of the
+\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
+for most mathematical functions.\cite{NEP13}
+In 2019 this was expanded more generally with the inclusion of the
+\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
+These two protocols allows providers of Array-like APIs to be interoperable
+with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
+For the users relying on specialized array projects it means that even though
+much code is written specifically for NumPy arrays and uses the NumPy API as
+\code{import numpy as np}, it can nevertheless work for them.
+A brief example of such a function which is written using NumPy
+but can now be used more generally is:
+\begin{lstlisting}
+def f(x):
+    y = np.tensordot(x, x.T)
+    return np.mean(np.exp(y))
+\end{lstlisting}
+For example, a user relying on this function and normally using it with a
+NumPy array, may now use a distributed Dask array when challenged
+by a very large dataset, without any modification to \code{f(x)}.
+
+While, as of writing this, the interface is still considered experimental
+and may be modified, it has been implemented by multiple projects already
+and proven useful.
+% TODO: Should likely have citation/example if I claim it was useful (based
+%       mainly on Matt's statement).
 
@@
 %  nr of core devs / commit rights) & contributors
+
+NumPy is currently maintained a group of 23 contributors with commit rights
+to the NumPy code base. Out of these 17 maintainers were active in 2019.
+Additionally, there are a few long term developers who contributed and maintain
+specific parts of NumPy, but are not officially maintainers.
+At a release cycle of about every half year, the five recent releases in the years
+2018 and 2019 have averaged about 450~pull requests~(PR) each.\footnote{
+    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
+    All data provided here is based on the development which happened through github
+    pull requests. In some cases contributions by maintainers may not be categorized as such.}
+% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
+With each release attracting more than a hundred new contributors.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
+    \caption{Number of pull requests merged into the NumPy master branch for each
+        quarter since 2012. The total number of PR is indicated with the
+        lower blue area showing the portion contributed by current or previous
+        maintainers.}\label{fig:prs-over-time}
+\end{figure}
+
+The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
+PR merged into the NumPy master branch.
+In this plot PR opened by current and old maintainers are indicated.
+Although the number of PRs being merged fluctuates,
+the plot indicates an increased number of contributions over the past
+years.
+Over the course of its history, NumPy has attracted PR by 823 contributors.
+However, its development still relies heavily on the relative small number
+of stable maintainers, who share more than half of the contributions among
+themselves.
 

### 2019-12-17 scripts/fetch-pr-data.py

@@ -1,0 +1,80 @@
+#!/usr/bin/env python3
+
+import sys
+
+import github
+import pickle
+
+
+token = sys.argv[1]
+
+# Using a token/login is required to download these statistics:
+g = github.Github(token)
+numpy = g.get_repo('numpy/numpy')
+
+# Closed will ignore PRs that are not yet merged...
+merged_pulls = numpy.get_pulls(state="closed")
+
+NUMPY_ORG_MEMBER_IDS = {u.id for u in numpy.get_collaborators()}
+
+try:
+    with open("pr_data.pkl", "rb") as pickle_file:
+        pr_list = pickle.load(pickle_file)
+
+    if "number" not in pr_list[0]:
+        for pr in pr_list:
+            pr["number"] = int(pr["url"].replace("https://github.com/numpy/numpy/pull/", ""))
+
+except Exception as e:
+    print("no data loaded:", e)
+    pr_list = []
+
+known = {pr["number"] for pr in pr_list}
+
+
+for pr in merged_pulls:
+    if pr.number in known:
+        print("Already knew PR number:", pr.number)
+        continue
+
+    if not pr.base.label == "numpy:master":
+        # skip non-master (usually not interesting w.r.t. review)
+        continue
+
+    if not pr.merged:
+        continue
+
+    user = pr.user
+    pr_author_id = user.id
+    pr_author = user.login
+
+    merger = pr.merged_by
+    try:
+
+        merged_by = merger.login
+        merged_by_id = merger.id
+    except:
+        merged_by = merger
+        merged_by_id = None
+
+    info = {
+        "number": pr.number,
+        "title": pr.title,
+        # There is probably a better way to get the URL...
+        "url": pr.url.replace("//api.", "//").replace("/pulls/", "/pull/").replace("/repos/", "/"),
+        "created_at": pr.created_at,
+        "merged_at": pr.merged_at,
+        "merged_by": merged_by,
+        "merged_by_id": merged_by_id,
+        "author": pr_author,
+        "author_id": pr_author_id,
+        "is_collaborator": pr_author_id in NUMPY_ORG_MEMBER_IDS,
+        "merge_commit_sha": pr.merge_commit_sha,
+        "labels": [l.name for l in pr.labels],
+        }
+    print(info)
+    pr_list.append(info)
+
+
+with open("pr_data.pkl", "wb") as pickle_file:
+    pickle.dump(pr_list, pickle_file)

### 2019-12-17 scripts/plot_merged_prs.py

@@ -1,0 +1,177 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import pickle
+import datetime
+import pandas
+
+from matplotlib import pyplot as plt
+
+from pandas.plotting import register_matplotlib_converters
+register_matplotlib_converters()
+
+# Choose either "CURRENT_MAINTAINERS" (list of those with commit rights
+# and users who had commit rights based on having done a merge),
+# or "HAS_MERGED_A_PR" which defines a "Maintainer" as anyone who has
+# merged a PR _before_ their PR was merged.
+DEFINE_MAINTAINERS_BY = "CURRENT_MAINTAINERS"  #  "HAS_MERGED_A_PR"
+
+
+# List of maintainers (NumPy users with commit rights)
+CURRENT_MAINTAINERS = """
+    ahaldane
+    certik
+    charris
+    cournape
+    eric-wieser
+    FrancescAlted
+    jaimefrio
+    jjhelmus
+    juliantaylor
+    mhvk
+    mwiebe
+    njsmith
+    pv
+    rgommers
+    rkern
+    seberg
+    shoyer
+    stefanv
+    teoliphant
+    matthew-brett
+    mattip
+    tylerjereddy
+    WarrenWeckesser
+    """.split()
+assert(len(CURRENT_MAINTAINERS) == 23)
+
+# Users who have merged a PR and/or served on the Steering Council
+OLDER_MAINTAINERS = """
+    argriffing
+    timcera
+    pletnes
+    jennystone
+""".split()
+CURRENT_MAINTAINERS.extend(OLDER_MAINTAINERS)
+
+
+def is_current_maintainer(user_login):
+    return user_login in CURRENT_MAINTAINERS
+
+# Filled using the date they first merged a PR.
+MAINTAINERS_SINCE = {}
+
+with open("pr_data.pkl", "rb") as pickle_file:
+    pr_list = pickle.load(pickle_file)
+
+
+def maintainer_by_first_merge(pr_list):
+    maintainer_since = MAINTAINERS_SINCE
+    names = {}
+    for pr in pr_list:
+        merger = pr["merged_by_id"]
+        if merger is None:
+            print("PR merged by unknown user:", pr["number"])
+        if pr["merged_by"] == "homu":
+            continue
+
+        if merger not in maintainer_since:
+            maintainer_since[merger] = pr["merged_at"]
+            names[merger] = pr["merged_by"]
+            continue
+
+
[…]

### 2019-12-18 authors.tex
Add abstract
ENH: Add initial draft of random
Add random section draft

@@
-\author[1]{Jane Doe}
-\affil[1]{Berkeley Institute for Data Science, University of California, Berkeley, CA 94720, USA}
+\author[1]{Charles R Harris}
+\author[2]{Travis E. Oliphant}
+\author[3,*]{St\'efan J. van der Walt}
+\author[3]{Warren Weckesser}
+\author[3,4,*]{K. Jarrod Millman}
+\author[5]{Kevin Sheppard}
+\affil[1]{Independent Researcher}
+\affil[2]{Quansight LLC, Austin, TX, USA}
+\affil[3]{Berkeley Institute for Data Science, University of California, Berkeley, CA 94720, USA}
+\affil[4]{Division of Biostatistics, University of California, Berkeley, CA 94720, USA}
+\affil[*]{numpy.articles@gmail.com}
+\affil[5]{kevin.sheppard@economics.ox.ac.uk, Department of Economics, University of Oxford, Oxford, OX1 3UQ, UK}

### 2019-12-18 paper.tex
ENH: Add initial draft of random
Add random section draft

@@
 
+The NumPy random module provides pseudorandom numbers from a wide range of
+distributions. In legacy versions of NumPy, simulated random values are produced
+by a RandomState object that: handles seeding and state initialization; wraps
+the core pseudorandom number generator based on a 32-bit implementation of
+MT19937; interfaces with the underlying code that transforms random bits into
+deviates from other distributions; and supplies a singleton instance exposed in
+the root of the random module.
+
+The RandomState object makes a compatibility guarantee so that a fixed seed and
+sequence of function calls produce the same set of values. This guarantee has
+slowed progress since improving the underlying code requires extending the API
+with additional keyword arguments. This guarantee continues to apply to
+RandomState.
+
+NumPy 1.17 introduced a new API for generating random numbers that use a more
+flexible structure that can be extended by libraries or end-users. The new API
+is build using components separate the tasks needed when generating random
+variates. Pseudorandom bits are generated by a bit generator. These bits are
+then transformed into variates from complex distributions by a generator.
+Finally, seeding is handled by an object that produces sequences of high-quality
+initial values.
+
+Bit generators are simple classes that manage the state of an underlying
+pseudorandom number generator. NumPy ships with four bit generators. The default
+bit generator is a 64-bit implementation of the Permuted Congruential Generator
+\cite{pcg64} (\code{PCG64}). The three other bit generators are a 64-bit version
+of the Philox generator\cite{random123} (\code{Philox}), Chris Doty-Humphrey's
+Small Fast Chaotic generator\cite{practrand} (\code{SFC64}), and the 32-bit
+Mersenne Twister\cite{mt19937} (\code{MT19937}) which has been used in older
+versions of NumPy.\footnote{The randomgen project supplies a wide range of
+alternative bit generators such as a cryptographic counter-based generators
+(\code{AESCtr}) and generators that expose hardware random number generators
+(\code{RDRAND})\cite{randomgen}.} Bit generators export a capsule containing
+pointers to functions that produce 64 bits, 32 bits, or a random double. They
+also expose CFFI and ctypes interfaces to the same three functions.
+
+The Generator consumes one of the bit generators and produces variates from
+complicated distributions. Many improved methods for generating random variates
+from common distributions were implemented, including the Ziggurat method for
+normal, exponential and gamma variates\cite{ziggurat}, and Lemire's method for
+bounded random integer generation\cite{lemire}. The Generator is more similar to
+the legacy RandomState, and its API is substantially the same. The key
+differences all relate to state management, which has been delegated to the bit
+generator. The Generator does not make the same stream guarantee as the
+RandomState object, and so variates may differ across versions as improved
+generation algorithms are introduced.\footnote{Despite the removal of the
+compatibility guarantee, simple reproducibility across versions is encouraged,
+and minor changes that do not produce meaningful performance gains or fix
+underlying bug are not generally adopted.}
+
+Finally, a SeedSequence is used to initialize a bit generator. The seed sequence
+can be initialized with no arguments, in which case it reads entropy from a
+system-dependent provider, or with a user-provided seed. The seed sequence then
+transforms the initial set of entropy into a sequence of high-quality
+pseudorandom integers, which can be used to initialize multiple bit generators
+deterministically. A key design goal of a seed sequence was to be splittable in
+the sense that a seed sequence can be used to produce child sequences that are
+distinct from their parents or other ancestors. This capability allows a seed
+sequence to be used in large distributed applications where the number of
+workers required is not known. The sequences generated from the same initial
+entropy and the same splits are fully deterministic to ensure reproducibility.
+
+The three components are combined to construct a complete random number
+generator.
+
+\begin{lstlisting}
+from numpy.random import Generator, PCG64, SeedSequence
+seq = SeedSequence(10304245474441179933310169597522947680)
+pcg = PCG64(seq)
+gen = Generator(pcg)
+\end{lstlisting}
+
+\noindent This approach retains access to the seed sequence which can then be used to
+spawn additional generators.
+
+\begin{lstlisting}
+children = seq.spawn(2)
+gen_0 = Generator(PCG64(children[0]))
+gen_1 = Generator(PCG64(children[1]))
+\end{lstlisting}
+
+\noindent While this approach retains complete flexibility, the method
+\code{np.random.default\_rng} can be used to instantize a Generator when
+reproducibility is not needed.
+
+The final goal of the new API is to improve extensibility. RandomState is a
+monolithic object that obscures all of the underlying state and functions. The
+component architecture is one part of the extensibility improvements. The
+underlying functions (written in C) which transform the output of a bit
+generator to other distributions are available for use in CFFI. This allows the
+same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
+numba. Both the bit generators and the low-level functions can also be used in C
+or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
+NumPy source. Alternative approaches that avoid this extra step are being
+explored.}
+
 \subsection*{The \_\_array\_function\_\_ Protocol}

### 2019-12-18 paper.tex
BUG: Fix grammar
Fix error
ENH: Wrap classes in code
Wrap classes in code decoration

@@
 distributions. In legacy versions of NumPy, simulated random values are produced
-by a RandomState object that: handles seeding and state initialization; wraps
+by a \code{RandomState} object that: handles seeding and state initialization; wraps
 the core pseudorandom number generator based on a 32-bit implementation of
@@
 
-The RandomState object makes a compatibility guarantee so that a fixed seed and
+The \code{RandomState} object makes a compatibility guarantee so that a fixed seed and
 sequence of function calls produce the same set of values. This guarantee has
@@
 with additional keyword arguments. This guarantee continues to apply to
-RandomState.
+\code{RandomState}.
 
@@
 flexible structure that can be extended by libraries or end-users. The new API
-is build using components separate the tasks needed when generating random
+is built using components that separate the steps required to generate random
 variates. Pseudorandom bits are generated by a bit generator. These bits are
@@
 
-The Generator consumes one of the bit generators and produces variates from
+The \code{Generator} consumes one of the bit generators and produces variates from
 complicated distributions. Many improved methods for generating random variates
@@
 normal, exponential and gamma variates\cite{ziggurat}, and Lemire's method for
-bounded random integer generation\cite{lemire}. The Generator is more similar to
-the legacy RandomState, and its API is substantially the same. The key
+bounded random integer generation\cite{lemire}. The \code{Generator} is more similar to
+the legacy \code{RandomState}, and its API is substantially the same. The key
 differences all relate to state management, which has been delegated to the bit
-generator. The Generator does not make the same stream guarantee as the
-RandomState object, and so variates may differ across versions as improved
+generator. The \code{Generator} does not make the same stream guarantee as the
+\code{RandomState} object, and so variates may differ across versions as improved
 generation algorithms are introduced.\footnote{Despite the removal of the
@@
 
-Finally, a SeedSequence is used to initialize a bit generator. The seed sequence
+Finally, a \code{SeedSequence} is used to initialize a bit generator. The seed sequence
 can be initialized with no arguments, in which case it reads entropy from a
@@
 \noindent While this approach retains complete flexibility, the method
-\code{np.random.default\_rng} can be used to instantize a Generator when
+\code{np.random.default\_rng} can be used to instantize a \code{Generator} when
 reproducibility is not needed.
 
-The final goal of the new API is to improve extensibility. RandomState is a
+The final goal of the new API is to improve extensibility. \code{RandomState} is a
 monolithic object that obscures all of the underlying state and functions. The

### 2019-12-18 paper.tex
ENH: Add URL for randomgen
Add URL and reformat

@@
 distributions. In legacy versions of NumPy, simulated random values are produced
-by a \code{RandomState} object that: handles seeding and state initialization; wraps
-the core pseudorandom number generator based on a 32-bit implementation of
+by a \code{RandomState} object that: handles seeding and state initialization;
+wraps the core pseudorandom number generator based on a 32-bit implementation of
 MT19937; interfaces with the underlying code that transforms random bits into
@@
 
-The \code{RandomState} object makes a compatibility guarantee so that a fixed seed and
-sequence of function calls produce the same set of values. This guarantee has
-slowed progress since improving the underlying code requires extending the API
-with additional keyword arguments. This guarantee continues to apply to
-\code{RandomState}.
+The \code{RandomState} object makes a compatibility guarantee so that a fixed
+seed and sequence of function calls produce the same set of values. This
+guarantee has slowed progress since improving the underlying code requires
+extending the API with additional keyword arguments. This guarantee continues to
+apply to \code{RandomState}.
 
@@
 Mersenne Twister\cite{mt19937} (\code{MT19937}) which has been used in older
-versions of NumPy.\footnote{The randomgen project supplies a wide range of
-alternative bit generators such as a cryptographic counter-based generators
-(\code{AESCtr}) and generators that expose hardware random number generators
-(\code{RDRAND})\cite{randomgen}.} Bit generators export a capsule containing
-pointers to functions that produce 64 bits, 32 bits, or a random double. They
-also expose CFFI and ctypes interfaces to the same three functions.
+versions of NumPy.\footnote{The
+\href{https://github.com/bashtage/randomgen}{randomgen project} supplies a wide
+range of alternative bit generators such as a cryptographic counter-based
+generators (\code{AESCtr}) and generators that expose hardware random number
+generators (\code{RDRAND})\cite{randomgen}.} Bit generators export a capsule
+containing pointers to functions that produce 64 bits, 32 bits, or a random
+double. They also expose CFFI and ctypes interfaces to the same three functions.
 
-The \code{Generator} consumes one of the bit generators and produces variates from
-complicated distributions. Many improved methods for generating random variates
-from common distributions were implemented, including the Ziggurat method for
-normal, exponential and gamma variates\cite{ziggurat}, and Lemire's method for
-bounded random integer generation\cite{lemire}. The \code{Generator} is more similar to
-the legacy \code{RandomState}, and its API is substantially the same. The key
-differences all relate to state management, which has been delegated to the bit
-generator. The \code{Generator} does not make the same stream guarantee as the
-\code{RandomState} object, and so variates may differ across versions as improved
-generation algorithms are introduced.\footnote{Despite the removal of the
-compatibility guarantee, simple reproducibility across versions is encouraged,
-and minor changes that do not produce meaningful performance gains or fix
-underlying bug are not generally adopted.}
+The \code{Generator} consumes one of the bit generators and produces variates
+from complicated distributions. Many improved methods for generating random
+variates from common distributions were implemented, including the Ziggurat
+method for normal, exponential and gamma variates\cite{ziggurat}, and Lemire's
+method for bounded random integer generation\cite{lemire}. The \code{Generator}
+is more similar to the legacy \code{RandomState}, and its API is substantially
+the same. The key differences all relate to state management, which has been
+delegated to the bit generator. The \code{Generator} does not make the same
+stream guarantee as the \code{RandomState} object, and so variates may differ
+across versions as improved generation algorithms are
+introduced.\footnote{Despite the removal of the compatibility guarantee, simple
+reproducibility across versions is encouraged, and minor changes that do not
+produce meaningful performance gains or fix underlying bug are not generally
+adopted.}
 
-Finally, a \code{SeedSequence} is used to initialize a bit generator. The seed sequence
-can be initialized with no arguments, in which case it reads entropy from a
-system-dependent provider, or with a user-provided seed. The seed sequence then
-transforms the initial set of entropy into a sequence of high-quality
-pseudorandom integers, which can be used to initialize multiple bit generators
-deterministically. A key design goal of a seed sequence was to be splittable in
-the sense that a seed sequence can be used to produce child sequences that are
-distinct from their parents or other ancestors. This capability allows a seed
-sequence to be used in large distributed applications where the number of
-workers required is not known. The sequences generated from the same initial
-entropy and the same splits are fully deterministic to ensure reproducibility.
+Finally, a \code{SeedSequence} is used to initialize a bit generator. The seed
+sequence can be initialized with no arguments, in which case it reads entropy
+from a system-dependent provider, or with a user-provided seed. The seed
+sequence then transforms the initial set of entropy into a sequence of
+high-quality pseudorandom integers, which can be used to initialize multiple bit
+generators deterministically. A key design goal of a seed sequence was to be
+splittable in the sense that a seed sequence can be used to produce child
+sequences that are distinct from their parents or other ancestors. This
+capability allows a seed sequence to be used in large distributed applications
+where the number of workers required is not known. The sequences generated from
+the same initial entropy and the same splits are fully deterministic to ensure
+reproducibility.
 
@@
 
-\noindent This approach retains access to the seed sequence which can then be used to
-spawn additional generators.
+\noindent This approach retains access to the seed sequence which can then be
+used to spawn additional generators.
 
@@
 
-The final goal of the new API is to improve extensibility. \code{RandomState} is a
-monolithic object that obscures all of the underlying state and functions. The
+The final goal of the new API is to improve extensibility. \code{RandomState} is
+a monolithic object that obscures all of the underlying state and functions. The
 component architecture is one part of the extensibility improvements. The

### 2019-12-19 paper.tex
Add linenumbers and wordcounts
Update outline and minor rearrangement

@@
 \usepackage{csquotes}
+\usepackage{moreverb} % for verbatim ouput (for wordcount)
 \usepackage{listings, textcomp}
@@
 
+\usepackage{lineno}
+\linenumbers
+
 \title{NumPy---Array Computation for Python}
@@
 \thispagestyle{empty}
+
+% Count of words
+\verbatiminput{wordcount.tex}
 
@@
 
-\subsubsection*{Project goals / scope of library}
+% \subsubsection*{Project goals / scope of library}
 
@@
 
-\subsubsection*{NEPS revived and roadmap}
+% \subsubsection*{NEPS revived and roadmap}
 
@@
 
-\section*{Architecture and implementation choices}
+NumPy, while having been wildly successful as a library and array exchange
+project, has now also become a standard language for expressing array and
+tensor computations.
 
-\subsection*{Library organization}
+\section*{Library organization}
+
 Broadly, the NumPy library consists of the following parts:
@@
 
-\paragraph{\emph{Core.}}  The \code{ndarray} data structure and the
-universal functions make up the core of NumPy.
+\subsection*{Core.}
+
+The \code{ndarray} data structure and the universal functions make up the core
+of NumPy.
 
@@
 
-\paragraph{\emph{Computing libraries.}}
+\subsection*{Computing libraries}
+
 NumPy provides a large library of functions for array manipulation
@@
 
-\paragraph{\emph{Infrastructure libraries.}} NumPy provides utilities
-for writing unit tests and for building Python packages.
+\subsection*{Infrastructure libraries}
+
+NumPy provides utilities for writing unit tests and for building Python
+packages.
 
@@
 
-\paragraph{\emph{F2PY.}}  The program \code{f2py} is a tool for
+The program \code{f2py} is a tool for
 building NumPy-aware Python wrappers of Fortran functions.
@@
 
-\subsection*{Common infrastructure}
+\section*{Common infrastructure}
 
-\subsubsection*{Wheels build system}
+In addition to the testing and building packages described above,
+as the fundamental package for numerical and scientific computing
+in Python, NumPy also plays a central role in building and
+standardizing much of the community infrastructure.
+The docstring standard mentioned in the history section is now
+widely adopted.
+
+% https://numpy.org/neps/nep-0029-deprecation_policy.html
+In NEP 29, we recommend, along with leaders from various other
+projects, that all projects across the Scientific Python ecosystem
+adopt a common ``time window-based'' policy for support of Python and
+NumPy versions. This standard will simplify downstream project and
+release planning.
+
+\subsection*{Wheels build system}
 
@@
 (for Windows).
-
-\subsubsection*{NEP 29}
-
-% https://numpy.org/neps/nep-0029-deprecation_policy.html
-In NEP 29, we recommend, along with leaders from various other
-projects, that all projects across the Scientific Python ecosystem
-adopt a common ``time window-based'' policy for support of Python and
-NumPy versions. This standard will simplify downstream project and
-release planning.
-
-\subsubsection*{NumPyDoc standard}
-
-% recent improvements
-
-\subsection*{Language choices}
 
@@
 
-\subsection*{The \_\_array\_function\_\_ Protocol}
+\subsection*{The Array Function Protocol}
 
-The central position which NumPy takes within the python ecosystem with
-respect to scientific computing and numerical works,
-means that a vast number of projects are build on it.
+The central position which NumPy takes within the numerical and scientific
+Python ecosystem  means that a vast number of projects are built on it.
 These projects are consumers of the NumPy API.
 
-On the other side, projects are providers of a
-\emph{NumPy-like API} and array objects targeting audiences with specialized needs beyond
-NumPy's capabilities. Some of these projects provide:
-
-\begin{description}
-    \tightlist
-    \item[SciPy's \code{scipy.sparse} and \code{pydata.sparse}:] Sparse arrays
-    \item[CuPy:] Computation on the GPU
-    \item[Dask:] Parallel, distributed computing
-    \item[PyTorch and TensorFlow:] Deep learning frameworks.
-\end{description}
+On the other side, projects are providers of a \emph{NumPy-like API} and array
+objects targeting audiences with specialized needs beyond NumPy's capabilities.
+For example, the NumPy API is implemented by
+several popular tensor computation libraries including
+CuPy\footnote{\url{https://cupy.chainer.org/}},
+Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
+Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
+and to a lesser extent
+TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.
+It is also implemented in packages that support sparse arrays
+such as \code{scipy.sparse} and \code{pydata.sparse}.
+Another notable example is Dask, a library for parallel computing in
+Python.  Dask adopts the NumPy API and therefore presents a familiar
+interface to existing NumPy users, while adding powerful abilities to
+parallelize and distribute tasks.
 
@@
 
-While, as of writing this, the interface is still considered experimental
-and may be modified, it has been implemented by multiple projects already
-and proven useful.
-% TODO: Should likely have citation/example if I claim it was useful (based
-%       mainly on Matt's statement).
+%In order to aid better interaction between NumPy and these libraries,
+%an ``array function'' protocol was added to NumPy in 2018 (see Section
+%X). The new protocol allows for external array objects (sometimes
+%called ``duck arrays'') to pass through NumPy:
+%
+%\begin{lstlisting}
+%  import numpy as np
+%  import cupy as cp
+%
+%  x_gpu = cp.array([1, 2, 3])
+%  y = np.sum(x_gpu)  # This works, and returns a GPU array
+%\end{lstlisting}
 
@@
 
-\subsection*{GSoD}
-
-\subsection*{Website}
+% \subsection*{GSoD}
 
@@
 
-\subsubsection*{Impact Now}
+Most of the technical work for the Moore and Sloan grant is now complete.
+The main remaining item is an overhaul of the data type system, which is discussed
+below.  Now that years of technical debt have been addressed, the project
+is in good shape for growing the community and better addressing the needs
+of new users.  That is the main focus of the CZI grant, which we discuss
+below.
 
-NumPy, while having been wildly successful as a library and array
-exchange project, has now also become a standard language for expressing array
-and tensor computations.  For example, the NumPy API is implemented by
-several popular tensor computation libraries including
-CuPy\footnote{\url{https://cupy.chainer.org/}},
-Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
-Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
-and to a lesser extent
-TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.
-
-Another notable example is Dask, a library for parallel computing in
-Python.  Dask adopts the NumPy API and therefore presents a familiar
-interface to existing NumPy users, while adding powerful abilities to
-parallelize and distribute tasks.
-
-In order to aid better interaction between NumPy and these libraries,
-an ``array function'' protocol was added to NumPy in 2018 (see Section
-X). The new protocol allows for external array objects (sometimes
-called ``duck arrays'') to pass through NumPy:
-
-\begin{lstlisting}
-  import numpy as np
-  import cupy as cp
-
-  x_gpu = cp.array([1, 2, 3])
-  y = np.sum(x_gpu)  # This works, and returns a GPU array
-\end{lstlisting}
-
-\subsubsection*{Future Development}
+% \subsubsection*{Future Development}
 
@@
 
+% outline plan for the CZI grant and an invitation to new contributors
+
+%\subsection*{Website}
+
 \section*{Data Availability}

### 2019-12-19 paper.tex
Fix typo
Add Governance section

@@
 It underpins almost every library that does numerical
-computation, including SciPy.  NumPy consists of the `ndarray`-object
-along with utility functions to operate on such arrays.  Because of
+computation, including SciPy.  NumPy consists of the \code{ndarray}-object
+along with utility functions to operate on it.  Because of
 its inherent simplicity—being a pointer to memory with some
 associated meta-data about shape, data-type, and so forth—the NumPy
-array has become the {\it de facto} exchange format for array data in
-Python.  That said, NumPy has now seen such widespread adoption that
-not only the array object but also its {\it Application Programming
-  Interface} (API) has become ubiquitous, and is imitated by, e.g.,
-popular tensor computation libraries.
-
-Over the past few years, with external financial support, NumPy has
-seen renewed developer interest.  It is actively developed to meet the
-needs of its community, including the addition of some fairly significant
-changes as outlined in the Recent technical improvements section below.
+array is the {\it de facto} exchange format for array data in
+Python.
+NumPy, while having been wildly successful as a library and array exchange
+project, has now also become a standard language for expressing array and
+tensor computations.  
+It has such widespread adoption that not only the array object but also its
+{\it Application Programming Interface} (API) has become ubiquitous, and is
+implemented by, for example, popular tensor computation libraries.
 
@@
 Stéfan van der Walt, a senior research data scientist at BIDS and a member of
-the NumPy steering committee, is the PI of both grants and manages a team of
+the NumPy steering council, is the PI of both grants and manages a team of
 four programmers working on the project.
@@
 Ralf Gommers, the Director of Quansight Labs and a member of
-the NumPy steering committee, is the PI.
+the NumPy steering council, is the PI.
 
@@
 
-NumPy is now actively developed, with weekly online developer
-meetings, a formalized NumPy enhancement proposal process, and
-multiple active grants (but still almost no industry support).
-% check almost no industry
-
 % (maturity, users)
 
-% \subsubsection*{Project goals / scope of library}
-
-% https://numpy.org/neps/scope.html
-
-% \subsubsection*{NEPS revived and roadmap}
+Over the past few years, with external financial support, NumPy has seen
+renewed developer interest.
+It is actively developed, with weekly online developer
+meetings and a formalized NumPy enhancement proposal process.
 
@@
 
+% https://numpy.org/neps/scope.html
 % https://numpy.org/neps/roadmap.html
-
-NumPy, while having been wildly successful as a library and array exchange
-project, has now also become a standard language for expressing array and
-tensor computations.
 
@@
 
-Broadly, the NumPy library consists of the following parts:
-the NumPy array data structure; the so-called \emph{universal functions};
-a set of library functions for manipulating arrays and doing scientific
-computation; infrastructure libraries for unit tests and Python package
-building; and the program \code{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
-The \code{ndarray} and the universal functions are generally considered
-the core of the library.
-In the following, we give a brief summary of these components of the
-library.
+The NumPy array data structure and universal functions, which we describe
+below, make up the core of NumPy.
+The NumPy library also includes
+a set of library functions for manipulating arrays and doing scientific computation,
+libraries for unit tests and Python package building,
+and the program \code{f2py}\cite{peterson2009f2py} for wrapping Fortran code in Python.
 
-\subsection*{Core.}
+\subsection*{Core}
 
-The \code{ndarray} data structure and the universal functions make up the core
-of NumPy.
-
-The \code{ndarray} is the data structure at the heart of NumPy.
-The data structure stores regularly strided homogeneous data types
-in a contiguous block memory, allowing for the efficient representation
+The \code{ndarray} data structure stores regularly strided homogeneous data
+types in a contiguous block memory, allowing for the efficient representation
 of $n$-dimensional data.
+A \emph{masked array} class, built on top of the \code{ndarray}, is
+provided as well. This class allows elements of an array to be masked,
+indicating that the data is missing or unknown.
 More details about the data structure are given in ``The NumPy array:
@@
 
-The \emph{universal functions}, or more concisely, \emph{ufuncs},
+The \emph{universal functions} (or \emph{ufuncs})
 are functions written in C that implement efficient looping over
@@
 NumPy provides a large library of functions for array manipulation
-and scientific computing, including functions for: creating, reshaping,
-concatenating, and padding arrays; searching, sorting and counting data
+including functions for: creating, reshaping, concatenating, and padding arrays;
+searching, sorting and counting data
 in arrays; computing elementary statistics, such as the mean, median,
 variance, and standard deviation; file I/O; and more.
+For historical reasons, NumPy also includes basic functionality for
+linear algebra,
+fast Fourier transforms and windowing,
+pseudorandom number generators,
+and polynomial fitting.
+We will keep supporting these, but we will not expand beyond them.
 
-A suite of functions for computing the \emph{fast Fourier transform (FFT)}
-and its inverse is provided.
+%A suite of functions for computing the \emph{fast Fourier transform (FFT)}
+%and its inverse is provided.
 
-NumPy's linear algebra library includes functions for: solving linear
-systems of equations; computing various functions of a matrix, including
-the determinant, the norm, the inverse, and the pseudo-inverse;
-computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
-and more.
+%NumPy's linear algebra library includes functions for: solving linear
+%systems of equations; computing various functions of a matrix, including
+%the determinant, the norm, the inverse, and the pseudo-inverse;
+%computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
+%and more.
+%
+%The random number generator library in NumPy provides several alternative
+%\emph{bit stream generators} that provide the core function of generating
+%random integers.
+%A higher-level generator class that implements an assortment of
+%probability distributions is provided. It includes the beta, gamma
+%and Weibull distributions, the univariate and multivariate normal
+%distributions, and more.
 
-The random number generator library in NumPy provides several alternative
-\emph{bit stream generators} that provide the core function of generating
-random integers.
-A higher-level generator class that implements an assortment of
-probability distributions is provided. It includes the beta, gamma
-and Weibull distributions, the univariate and multivariate normal
-distributions, and more.
-
-A \emph{masked array} class, built on top of the \code{ndarray}, is
-provided.  This class allows elements of an array to be masked,
-indicating that the data is missing or unknown.
-
-There is also a \code{matrix} class the implements 2-d arrays with
-specialized behavior.  We mention it here for completeness, but the
-use of \code{matrix} is discouraged.  The quirks of the \code{matrix}
-class have proven to be troublesome, and with the addition of the
-infix \code{@} operator to Python for matrix multiplication, there was no
-longer a need for a separate \code{matrix} class, since one can
-use a 2-d array instead.
+%For historical reasons, we provide a \code{matrix} class the implements 2-d
+%arrays with specialized behavior.
+%But we strongly recommend you use a 2-d array and use the infix \code{@} operator
+%for matrix multiplication.
+%
+%There is also a \code{matrix} class the implements 2-d arrays with
+%specialized behavior.  We mention it here for completeness, but the
+%use of \code{matrix} is discouraged.  The quirks of the \code{matrix}
+%class have proven to be troublesome, and with the addition of the
+%infix \code{@} operator to Python for matrix multiplication, there was no
+%longer a need for a separate \code{matrix} class, since one can
+%use a 2-d array instead.
 
@@
 
-NumPy provides utilities for writing unit tests and for building Python
-packages.
-
-The \code{testing} subpackage provides functions such as
+Being at the base of the scientific Python stack, NumPy provides some basic
+infrastructure for other packages in the scientific Python ecosystem.
+The \code{numpy.testing} subpackage provides functions such as
 \code{assert\_allclose(actual, desired)} that may be used in unit
 test suites for code that uses NumPy arrays.
-
-NumPy provides a subpackage includes functions and classes
-that can be used by libraries that have NumPy as dependency,
-to be used in the library's configuration code that installs the
-library and that builds a package that can be used, for example,
-on the PyPI web site.
-
+The \code{numpy.distutils} subpackage provides build support for C++, Fortran,
+BLAS/LAPACK, and other relevant libraries for scientific computing.
 The program \code{f2py} is a tool for
@@
 widely adopted.
-
+We are also using the NEP system as a way to help coordinate
+the larger scientific Python community.
 % https://numpy.org/neps/nep-0029-deprecation_policy.html
-In NEP 29, we recommend, along with leaders from various other
-projects, that all projects across the Scientific Python ecosystem
-adopt a common ``time window-based'' policy for support of Python and
-NumPy versions. This standard will simplify downstream project and
-release planning.
+For example, in NEP 29, we recommend, along with leaders from various other
+projects, that all projects across the Scientific Python ecosystem adopt a
+common ``time window-based'' policy for support of Python and NumPy versions.
+This standard will simplify downstream project and release planning.
 
@@
 \section*{Recent technical improvements}
+
+\subsection*{The Array Function Protocol}
+
+The central position which NumPy takes within the numerical and scientific
+Python ecosystem  means that a vast number of projects are built on it.
+These projects are consumers of the NumPy API.
+
+On the other side, projects are providers of a \emph{NumPy-like API} and array
+objects targeting audiences with specialized needs beyond NumPy's capabilities.
+For example, the NumPy API is implemented by
+several popular tensor computation libraries including
+CuPy\footnote{\url{https://cupy.chainer.org/}},
+Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
+Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
+and to a lesser extent
+TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.
+It is also implemented in packages that support sparse arrays
+such as \code{scipy.sparse} and \code{pydata.sparse}.
+Another notable example is Dask, a library for parallel computing in
+Python.  Dask adopts the NumPy API and therefore presents a familiar
+interface to existing NumPy users, while adding powerful abilities to
+parallelize and distribute tasks.
+
+The multitude of specialized projects creates the difficulty that consumers
+of these NumPy-like APIs write code specific to a single project and do not support
+all of the above array providers.
+This is a burden for users relying on the specialized Array-like, since
+a tool they need may not work for them.
+Additionally, a user who has the need of transitioning from NumPy to a more
+specialized array, is faced with a challenge.
+These issue can lead into a fragmentation of the user bases into consumers
+of different NumPy-like APIs.
+
+To address these issues NumPy has the goal of providing the fundamental
+API for \emph{interoperability} between the various NumPy-like APIs.
+An earlier step in this direction was the implementation of the
+\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
+for most mathematical functions.\cite{NEP13}
+In 2019 this was expanded more generally with the inclusion of the
+\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
+These two protocols allows providers of Array-like APIs to be interoperable
+with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
+For the users relying on specialized array projects it means that even though
+much code is written specifically for NumPy arrays and uses the NumPy API as
+\code{import numpy as np}, it can nevertheless work for them.
+A brief example of such a function which is written using NumPy
+but can now be used more generally is:
+\begin{lstlisting}
+def f(x):
+    y = np.tensordot(x, x.T)
+    return np.mean(np.exp(y))
+\end{lstlisting}
+For example, a user relying on this function and normally using it with a
+NumPy array, may now use a distributed Dask array when challenged
+by a very large dataset, without any modification to \code{f(x)}.
+
+%In order to aid better interaction between NumPy and these libraries,
+%an ``array function'' protocol was added to NumPy in 2018 (see Section
+%X). The new protocol allows for external array objects (sometimes
+%called ``duck arrays'') to pass through NumPy:
+%
+%\begin{lstlisting}
+%  import numpy as np
+%  import cupy as cp
+%
+%  x_gpu = cp.array([1, 2, 3])
+%  y = np.sum(x_gpu)  # This works, and returns a GPU array
+%\end{lstlisting}
 
@@
 \noindent While this approach retains complete flexibility, the method
-\code{np.random.default\_rng} can be used to instantize a \code{Generator} when
+\code{np.random.default\_rng} can be used to instantiate a \code{Generator} when
 reproducibility is not needed.
@@
 explored.}
-
-\subsection*{The Array Function Protocol}
-
-The central position which NumPy takes within the numerical and scientific
-Python ecosystem  means that a vast number of projects are built on it.
-These projects are consumers of the NumPy API.
-
-On the other side, projects are providers of a \emph{NumPy-like API} and array
-objects targeting audiences with specialized needs beyond NumPy's capabilities.
-For example, the NumPy API is implemented by
-several popular tensor computation libraries including
-CuPy\footnote{\url{https://cupy.chainer.org/}},
-Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
-Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
-and to a lesser extent
-TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.
-It is also implemented in packages that support sparse arrays
-such as \code{scipy.sparse} and \code{pydata.sparse}.
-Another notable example is Dask, a library for parallel computing in
-Python.  Dask adopts the NumPy API and therefore presents a familiar
-interface to existing NumPy users, while adding powerful abilities to
-parallelize and distribute tasks.
-
-The multitude of specialized projects creates the difficulty that consumers
-of these NumPy-like APIs write code specific to a single project and do not support
-all of the above array providers.
-This is a burden for users relying on the specialized Array-like, since
-a tool they need may not work for them.
-Additionally, a user who has the need of transitioning from NumPy to a more
-specialized array, is faced with a challenge.
-These issue can lead into a fragmentation of the user bases into consumers
-of different NumPy-like APIs.
-
-To address these issues NumPy has the goal of providing the fundamental
-API for \emph{interoperability} between the various NumPy-like APIs.
-An earlier step in this direction was the implementation of the
-\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
-for most mathematical functions.\cite{NEP13}
-In 2019 this was expanded more generally with the inclusion of the
-\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
-These two protocols allows providers of Array-like APIs to be interoperable
-with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
-For the users relying on specialized array projects it means that even though
-much code is written specifically for NumPy arrays and uses the NumPy API as
-\code{import numpy as np}, it can nevertheless work for them.
-A brief example of such a function which is written using NumPy
-but can now be used more generally is:
-\begin{lstlisting}
-def f(x):
-    y = np.tensordot(x, x.T)
-    return np.mean(np.exp(y))
-\end{lstlisting}
-For example, a user relying on this function and normally using it with a
-NumPy array, may now use a distributed Dask array when challenged
-by a very large dataset, without any modification to \code{f(x)}.
-
-%In order to aid better interaction between NumPy and these libraries,
-%an ``array function'' protocol was added to NumPy in 2018 (see Section
-%X). The new protocol allows for external array objects (sometimes
-%called ``duck arrays'') to pass through NumPy:
-%
-%\begin{lstlisting}
-%  import numpy as np
-%  import cupy as cp
-%
-%  x_gpu = cp.array([1, 2, 3])
-%  y = np.sum(x_gpu)  # This works, and returns a GPU array
-%\end{lstlisting}
 
@@
 
-\subsection*{Governance}
-
-% incl. CoC
-
-\subsection*{Maintainers and contributors}
-
-%  nr of core devs / commit rights) & contributors
-
-NumPy is currently maintained a group of 23 contributors with commit rights
+NumPy is currently maintained by a group of 23 contributors with commit rights
 to the NumPy code base. Out of these 17 maintainers were active in 2019.
@@
 Over the course of its history, NumPy has attracted PR by 823 contributors.
-However, its development still relies heavily on the relative small number
-of stable maintainers, who share more than half of the contributions among
+However, its development still relies heavily on a relatively small number
+of active maintainers, who share more than half of the contributions among
 themselves.
 
-% may be useful:
-% - https://github.com/docathon/watchtower
-% - https://github.com/machine-shop/oss_devkit/blob/master/plot_pr.py % (PR over time plot, there may be other better versions out there)
+% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
+% https://github.com/numpy/numpy/pull/6352
+NumPy adopted an official Governance Document on October~5,
+2015\cite{NumPyProjectGovernance}.
+Project decisions are usually made by consensus of interested contributors.
+This means for many thing everyone is entrusted with veto power.
+A Steering Council, currently composed of 12~members, facilitates this
+process and oversees daily development of the project by contributing code
+and reviewing contributions from the community.
+
+% https://mail.python.org/pipermail/numpy-discussion/2018-July/078476.html
+% https://github.com/numpy/numpy/pull/11865
+NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
+In brief, we strive to:
+\emph{be open};
+\emph{be empathetic, welcoming, friendly, and patient};
+\emph{be collaborative};
+\emph{be inquisitive}; and
+\emph{be careful in the words that we choose}.
+The Code of Conduct also specifies how breaches can be reported and outlines
+the process for responding to such reports.
 

### 2019-12-19 paper.tex
Edit first four sections

@@
 
-NumPy is currently maintained by a group of 23 contributors with commit rights
+\subsection*{Governance}
+
+% incl. CoC
+
+\subsection*{Maintainers and contributors}
+
+%  nr of core devs / commit rights) & contributors
+
+NumPy is currently maintained a group of 23 contributors with commit rights
 to the NumPy code base. Out of these 17 maintainers were active in 2019.
@@
 Over the course of its history, NumPy has attracted PR by 823 contributors.
-However, its development still relies heavily on a relatively small number
-of active maintainers, who share more than half of the contributions among
+However, its development still relies heavily on the relative small number
+of stable maintainers, who share more than half of the contributions among
 themselves.
 
-% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
-% https://github.com/numpy/numpy/pull/6352
-NumPy adopted an official Governance Document on October~5,
-2015\cite{NumPyProjectGovernance}.
-Project decisions are usually made by consensus of interested contributors.
-This means for many thing everyone is entrusted with veto power.
-A Steering Council, currently composed of 12~members, facilitates this
-process and oversees daily development of the project by contributing code
-and reviewing contributions from the community.
-
-% https://mail.python.org/pipermail/numpy-discussion/2018-July/078476.html
-% https://github.com/numpy/numpy/pull/11865
-NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
-In brief, we strive to:
-\emph{be open};
-\emph{be empathetic, welcoming, friendly, and patient};
-\emph{be collaborative};
-\emph{be inquisitive}; and
-\emph{be careful in the words that we choose}.
-The Code of Conduct also specifies how breaches can be reported and outlines
-the process for responding to such reports.
+% may be useful:
+% - https://github.com/docathon/watchtower
+% - https://github.com/machine-shop/oss_devkit/blob/master/plot_pr.py % (PR over time plot, there may be other better versions out there)
 

### 2019-12-19 paper.tex
Move project organization section

--- 2019-12-19 paper.tex
@@
 
+NumPy plays a central role in building and standardizing much of the scientific
+Python community infrastructure.
+The docstring standard mentioned in the history section is now widely adopted.
+We are also using the NEP system as a way to help coordinate the larger
+scientific Python community.
+% https://numpy.org/neps/nep-0029-deprecation_policy.html
+For example, in NEP 29, we recommend, along with leaders from various other
+projects, that all projects across the Scientific Python ecosystem adopt a
+common ``time window-based'' policy for support of Python and NumPy versions.
+This standard will simplify downstream project and release planning.
+
 % https://numpy.org/neps/scope.html
 % https://numpy.org/neps/roadmap.html
+
+\section*{Project organization and community}
+
+NumPy is currently maintained by a group of 23 contributors with commit rights
+to the NumPy code base. Out of these 17 maintainers were active in 2019.
+Additionally, there are a few long term developers who contributed and maintain
+specific parts of NumPy, but are not officially maintainers.
+At a release cycle of about every half year, the five recent releases in the years
+2018 and 2019 have averaged about 450~pull requests~(PR) each.\footnote{
+    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
+    All data provided here is based on the development which happened through github
+    pull requests. In some cases contributions by maintainers may not be categorized as such.}
+% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
+With each release attracting more than a hundred new contributors.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
+    \caption{Number of pull requests merged into the NumPy master branch for each
+        quarter since 2012. The total number of PR is indicated with the
+        lower blue area showing the portion contributed by current or previous
+        maintainers.}\label{fig:prs-over-time}
+\end{figure}
+
+The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
+PR merged into the NumPy master branch.
+In this plot PR opened by current and old maintainers are indicated.
+Although the number of PRs being merged fluctuates,
+the plot indicates an increased number of contributions over the past
+years.
+Over the course of its history, NumPy has attracted PR by 823 contributors.
+However, its development still relies heavily on a relatively small number
+of active maintainers, who share more than half of the contributions among
+themselves.
+
+% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
+% https://github.com/numpy/numpy/pull/6352
+NumPy adopted an official Governance Document on October~5,
+2015\cite{NumPyProjectGovernance}.
+Project decisions are usually made by consensus of interested contributors.
+This means for many thing everyone is entrusted with veto power.
+A Steering Council, currently composed of 12~members, facilitates this
+process and oversees daily development of the project by contributing code
+and reviewing contributions from the community.
+
+% https://mail.python.org/pipermail/numpy-discussion/2018-July/078476.html
+% https://github.com/numpy/numpy/pull/11865
+NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
+In brief, we strive to:
+\emph{be open};
+\emph{be empathetic, welcoming, friendly, and patient};
+\emph{be collaborative};
+\emph{be inquisitive}; and
+\emph{be careful in the words that we choose}.
+The Code of Conduct also specifies how breaches can be reported and outlines
+the process for responding to such reports.
+
+% \subsection*{GSoD}
 
@@
 
-\section*{Common infrastructure}
+% \section*{Common infrastructure}
 
-In addition to the testing and building packages described above,
-as the fundamental package for numerical and scientific computing
-in Python, NumPy also plays a central role in building and
-standardizing much of the community infrastructure.
-The docstring standard mentioned in the history section is now
-widely adopted.
-We are also using the NEP system as a way to help coordinate
-the larger scientific Python community.
-% https://numpy.org/neps/nep-0029-deprecation_policy.html
-For example, in NEP 29, we recommend, along with leaders from various other
-projects, that all projects across the Scientific Python ecosystem adopt a
-common ``time window-based'' policy for support of Python and NumPy versions.
-This standard will simplify downstream project and release planning.
-
-\subsection*{Wheels build system}
-
-A Python \emph{wheel}\cite{PEP427} is a standard file format for
-distributing Python libraries.  In addition to Python code, a
-wheel may include compiled C extensions and other binary data.
-This is important, because many libraries, including NumPy,
-require a C compiler and other build tools to build the software
-from the source code, making it difficult for many users to install
-the software on their own.  The introduction of wheels to the Python
-packaging system has made it much easier for users to install
-precompiled libraries.
-
-A GitHub repository containing scripts to build NumPy wheels has
-been configured so that a simple commit to the repository triggers
-an automated build system that creates NumPy wheels for several
-computer platforms, including Windows, Mac OSX and Linux.  The wheels
-are uploaded to a public server and made available for anyone to use.
-This system makes it easy for users to install precompiled versions
-of NumPy on these platforms.
-
-The technology that is used to build the wheels evolves continually.
-At the time this paper is being written, a key component is the
-\code{multibuild} suite of tools developed by Matthew Brett and
-other developers\cite{multibuild}.  Currently, scripts using
-\code{multibuild} are written for the continuous integration
-platforms Travis-CI (for Linux and Mac OSX) and Appveyor
-(for Windows).
+%\subsection*{Wheels build system}
+%
+%A Python \emph{wheel}\cite{PEP427} is a standard file format for
+%distributing Python libraries.  In addition to Python code, a
+%wheel may include compiled C extensions and other binary data.
+%This is important, because many libraries, including NumPy,
+%require a C compiler and other build tools to build the software
+%from the source code, making it difficult for many users to install
+%the software on their own.  The introduction of wheels to the Python
+%packaging system has made it much easier for users to install
+%precompiled libraries.
+%
+%A GitHub repository containing scripts to build NumPy wheels has
+%been configured so that a simple commit to the repository triggers
+%an automated build system that creates NumPy wheels for several
+%computer platforms, including Windows, Mac OSX and Linux.  The wheels
+%are uploaded to a public server and made available for anyone to use.
+%This system makes it easy for users to install precompiled versions
+%of NumPy on these platforms.
+%
+%The technology that is used to build the wheels evolves continually.
+%At the time this paper is being written, a key component is the
+%\code{multibuild} suite of tools developed by Matthew Brett and
+%other developers\cite{multibuild}.  Currently, scripts using
+%\code{multibuild} are written for the continuous integration
+%platforms Travis-CI (for Linux and Mac OSX) and Appveyor
+%(for Windows).
 
@@
 
-\section*{Project organization and community}
-
-NumPy is currently maintained by a group of 23 contributors with commit rights
-to the NumPy code base. Out of these 17 maintainers were active in 2019.
-Additionally, there are a few long term developers who contributed and maintain
-specific parts of NumPy, but are not officially maintainers.
-At a release cycle of about every half year, the five recent releases in the years
-2018 and 2019 have averaged about 450~pull requests~(PR) each.\footnote{
-    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
-    All data provided here is based on the development which happened through github
-    pull requests. In some cases contributions by maintainers may not be categorized as such.}
-% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
-With each release attracting more than a hundred new contributors.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
-    \caption{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012. The total number of PR is indicated with the
-        lower blue area showing the portion contributed by current or previous
-        maintainers.}\label{fig:prs-over-time}
-\end{figure}
-
-The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
-PR merged into the NumPy master branch.
-In this plot PR opened by current and old maintainers are indicated.
-Although the number of PRs being merged fluctuates,
-the plot indicates an increased number of contributions over the past
-years.
-Over the course of its history, NumPy has attracted PR by 823 contributors.
-However, its development still relies heavily on a relatively small number
-of active maintainers, who share more than half of the contributions among
-themselves.
-
-% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
-% https://github.com/numpy/numpy/pull/6352
-NumPy adopted an official Governance Document on October~5,
-2015\cite{NumPyProjectGovernance}.
-Project decisions are usually made by consensus of interested contributors.
-This means for many thing everyone is entrusted with veto power.
-A Steering Council, currently composed of 12~members, facilitates this
-process and oversees daily development of the project by contributing code
-and reviewing contributions from the community.
-
-% https://mail.python.org/pipermail/numpy-discussion/2018-July/078476.html
-% https://github.com/numpy/numpy/pull/11865
-NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
-In brief, we strive to:
-\emph{be open};
-\emph{be empathetic, welcoming, friendly, and patient};
-\emph{be collaborative};
-\emph{be inquisitive}; and
-\emph{be careful in the words that we choose}.
-The Code of Conduct also specifies how breaches can be reported and outlines
-the process for responding to such reports.
-
-% \subsection*{GSoD}
-
 \section*{Discussion}

### 2019-12-19 paper.tex
Add notes
Rework history, funding, and current status sections (w/ Jarrod Millman)

@@
 
+The arrival of a stable version of NumPy (1.0) in October 2006 unified
+the scientific Python community, and heralded the era of popular
+scientific computing in Python that has led to formidable successes
+such as [cite some impacts here].
+
+[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
+  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
+  IDL is now below Python, Pangeo <-> xarray, Earthcube,
+  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
+  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
+  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]
+
 \subsection*{History}
@@
 objects had different strengths and weaknesses, and therefore tended
-to divide the already small user community.  In 2005, Travis Oliphant
-started NumPy, a ``best of both worlds'' unification of the two
-projects, based on Numeric.  The arrival of stable version 1.0 in October
-2006 unified the scientific Python
-community, and heralded the era of popular scientific computing in
-Python that has led to formidable successes such as [cite some impacts here].
+to divide the already small user community.  Furthermore, the divide
+posed a challenge for SciPy, which built on Numeric and provided
+fundamental scientific routines \cite{virtanen2019scipy}.
+In response, the maintainer of Numeric, Travis Oliphant, started a
+major rewrite that aimed to be a ``best of both worlds'' unification
+of the two projects.  This became known as NumPy.
 
 While, by 2006, NumPy was already eminently usable, it lacked good
-documentation. In order to use NumPy in his classes, Joe Harrington
+documentation. In order to use NumPy and SciPy in his classes, Joe Harrington
 from the University of South Florida sponsored several rounds of the
@@
 During these intensive
-documentation improvement sessions, a Wikipedia-like system was
-developed to allow community members to contribute so-called
+documentation improvement sessions, we developed a Wikipedia-like
+system for community members to contribute so-called
 ``docstrings''—text describing each NumPy function. A community
@@
 
-For a more detailed history of the scientific Python ecosystem, please
-refer to ``SciPy 1.0---Fundamental Algorithms for Scientific Computing
-in Python'' \cite{virtanen2019scipy}.
+% developed best practices, version control, unit testing, code review,
+% runable documentation, \cite{millman2014developing}
+% Travis steps back, no focused community effort until grant
+
+To coordinate work as a global community, to ensure accuracy of
+computed results, and to allow modification without breakage, the
+NumPy community pioneered best software engineering practices in the
+scientific Python community.  For example, NumPy early on adopted
+distributed revision control, unit testing, continuous integration,
+code review, standardized executable documentation, ...
+
+However, after his initial effort to build NumPy, Travis Oliphant
+stepped back from development to found a company and several projects,
+including Numba---an accelerated Python compiler that could vastly
+accelerate NumPy expressions.  By 2009, the project had entered a
+period of uncoordinated, opportunistic development, and was
+increasingly weighed down by technical debt that had accrued since the
+days of Numeric.
 
 \subsection*{Funding}
+
+% built up technical debt, from a legacy of code from two projects
+% and a lack of fulltime attention left ...
 
@@
 contributions were made, notably those by Mark Wiebe when he was
-sponsored by Travis Oliphant at Continuum Analytics, mostly the package had rapid growth
-and adoption without external investment.
-
-% In 2014, NumPy received its
-% first large grants from the Moore and Sloan foundations, totaling 1.3M
-% dollars.  This allowed for hiring several full time programmers and,
-% with the support of community members, mobilized a project that had
-% become stagnant for the several years prior.
+sponsored by Travis Oliphant at Continuum Analytics, mostly the
+package had rapid adoption and became central to scientific computing
+without external investment.
 
@@
 
-The Berkeley Institute for Data Science (BIDS) received two grants to support
-the development of NumPy.
-At the end of 2016, the Gordon and Betty Moore Foundation awarded them
-\$645,020 for improving NumPy for better data science.
-In 2017, the Alfred P. Sloan Foundation complemented the funding with an
-additional \$659,359.
-Stéfan van der Walt, a senior research data scientist at BIDS and a member of
-the NumPy steering council, is the PI of both grants and manages a team of
-four programmers working on the project.
+% From grant text/summary on Sloan website:
+%
+% to remove the technical debt and set in place standards and
+% architecture to encourage more sustainable development
+
+% In 2014, NumPy received its
+% first large grants from the Moore and Sloan foundations, totaling 1.3M
+% dollars.  This allowed for hiring several full time programmers and,
+% with the support of community members, mobilized a project that had
+% become stagnant for the several years prior.
+
+In 2017, the Berkeley Institute for Data Science
+(BIDS) received grants totaling 1.3M USD from the Gordon \& Betty
+Moore and the Alfred P. Sloan foundations to support the development
+of NumPy for better data science.
+Specifically, the aim was to address technical debt and set in place
+standards and architecture to encourage more sustainable development.
+Stéfan van der Walt, a senior research data scientist at BIDS and a
+member of the NumPy steering council, is the PI and manages four
+programmers working on the project.
 
@@
 % lessons learned?
-
-% technical focus 
-% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
-% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
 
@@
 
-NumFOCUS received a grant for \$195,000 in 2019 for the development of NumPy
-and OpenBLAS from the Chan Zuckerberg Initiative. 
-NumPy depends on OpenBLAS to provide its underlying accelerated linear algebra implementation.
-Ralf Gommers, the Director of Quansight Labs and a member of
-the NumPy steering council, is the PI.
+In 2019, NumFOCUS received a grant for \$195,000 in 2019 for the
+development of NumPy and OpenBLAS (on which NumPy depends for
+accelerated linear algebra) from the Chan Zuckerberg Initiative.
+This grant, building on the technical work supported by the previous
+grant, will focus on better serving NumPy's large number of beginner
+to intermediate level users and on growing the community of NumPy
+contributors.
+Ralf Gommers, the Director of Quansight Labs and a member of the NumPy
+steering council, is the PI.
 
-% focus on better serve NumPy's large number of beginner to intermediate level users
-% focusing all our time on growing the team and better serving new users and contributors
+The project receives around 10K USD per year from Tidelift, which is
+used to fund documentation and website developers.
 
@@
 
-% Intel, QuanSight, ...
-
 \subsection*{Current status}
@@
 
-Over the past few years, with external financial support, NumPy has seen
-renewed developer interest.
-It is actively developed, with weekly online developer
-meetings and a formalized NumPy enhancement proposal process.
+In 2007, the NumPy release manager, Jarrod Millman, adopted the notion
+of enhancement proposals---used infrequently until 2017 when, to
+facilitate work expected from funded development, he formalized the
+NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
+Python Enhancement Proposals (PEPs) for ``proposing major new
+features, for collecting community input on an issue, and for
+documenting the design decisions that have gone into
+Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
+Since then there have been 19 proposed NEPS---6 have been implemented,
+4 have been accepted and are being implemented, 4 are under
+consideration, 3 have been deferred or superseded, and 2 have been rejected
+or withdrawn.
 
-Following Python's practice of using Python Enhancement Proposals (PEPs) for
-``proposing major new features, for collecting community input on an
-issue, and for documenting the design decisions that have gone into Python,''
-we began using NumPy Enhancement Proposals (NEPs) in 2007.
-This process was informal and only occasionally used until 2017 when
-we adopted a more formal process \footnote{\url{https://numpy.org/neps/nep-0000.html}}.
-Since then there have been 19 proposed NEPS---6 have been implemented,
-4 have been accepted and are being implemented, 4 are under consideration,
-3 have deferred or superseded, and 2 have been rejected or withdrawn.
-We highlight a few of the implemented NEPs below.
+The first activities organized for the new grant were a NEP development
+sprint, a weekly community call, and a meeting to write a draft roadmap.
+
+Since the roadmap was meant to represent community consensus on future
+development, it was presented for comment at the annual SciPy
+conference during a so-called ``Birds of a Feather'' session, attended
+by more than 100 people.  One additional meeting was held to discuss
+specific enhancements, such as the development of a new data-type
+system.  The final roadmap was vetted by the community via online
+discussion before being accepted.
+
+% technical focus
+% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
+% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
+
+% Intel, QuanSight, ...
+
+The weekly community calls alternate weekly between triage meetings and
+higher level discussion.  The calls not only involve developers from
+the community, but provide a venue for vendors and other external
+groups to provide input.  For example, after Intel produced a forked
+version of NumPy, one of their developers joined a call to discuss
+community concerns.
 
@@
 The docstring standard mentioned in the history section is now widely adopted.
-We are also using the NEP system as a way to help coordinate the larger
+We are also now using the NEP system as a way to help coordinate the larger
 scientific Python community.
@@
 % https://numpy.org/neps/roadmap.html
+
+We showed that NumPy is actively developed according to a community
+vetted plan that integrates formal enhancement proposals.  Next, we
+give a more in-depth overview of the team developing NumPy, their
+activities, as well as the governance and community structures that
+guide the collaboration.
 
@@
 fast Fourier transforms and windowing,
-pseudorandom number generators,
 and polynomial fitting.
 We will keep supporting these, but we will not expand beyond them.
+
+% use scipy for linear algebra etc...
 
@@
 %
-%The random number generator library in NumPy provides several alternative
-%\emph{bit stream generators} that provide the core function of generating
-%random integers.
-%A higher-level generator class that implements an assortment of
-%probability distributions is provided. It includes the beta, gamma
-%and Weibull distributions, the univariate and multivariate normal
-%distributions, and more.
 
@@
 
+%The random number generator library in NumPy provides several alternative
+%\emph{bit stream generators} that provide the core function of generating
+%random integers.
+%A higher-level generator class that implements an assortment of
+%probability distributions is provided. It includes the beta, gamma
+%and Weibull distributions, the univariate and multivariate normal
+%distributions, and more.
+
 The NumPy random module provides pseudorandom numbers from a wide range of
@@
 apply to \code{RandomState}.
+
+% Mention explicitly the NEPs involved
 
@@
 
+% started at as focused small matrices, class examples
+
 At the time of writing the two fastest supercomputers in the
@@
 \section*{Discussion}
+
+% 1st grant done, technical cleanup finished except dtypes
+% 2nd grant starting, community growth and new user focus
 

### 2019-12-20 paper.tex
Minor changes

@@
 
-% developed best practices, version control, unit testing, code review,
-% runable documentation, \cite{millman2014developing}
-% Travis steps back, no focused community effort until grant
-
-To coordinate work as a global community, to ensure accuracy of
-computed results, and to allow modification without breakage, the
-NumPy community pioneered best software engineering practices in the
-scientific Python community.  For example, NumPy early on adopted
-distributed revision control, unit testing, continuous integration,
-code review, standardized executable documentation, ...
+To coordinate work as a global community, to ensure accuracy of computed
+results, and to allow modification without breakage, the NumPy community
+pioneered best software engineering practices in the scientific Python
+community  \cite{millman2014developing}.
+For example, NumPy early on adopted distributed revision control, unit testing,
+continuous integration, code review, and standardized executable documentation.
 
@@
 \subsection*{Funding}
-
-% built up technical debt, from a legacy of code from two projects
-% and a lack of fulltime attention left ...
 
@@
 
-% From grant text/summary on Sloan website:
-%
-% to remove the technical debt and set in place standards and
-% architecture to encourage more sustainable development
-
-% In 2014, NumPy received its
-% first large grants from the Moore and Sloan foundations, totaling 1.3M
-% dollars.  This allowed for hiring several full time programmers and,
-% with the support of community members, mobilized a project that had
-% become stagnant for the several years prior.
-
 In 2017, the Berkeley Institute for Data Science
@@
 programmers working on the project.
-
-% how much do we say about who these developers are and
-% lessons learned?
 
@@
 used to fund documentation and website developers.
-
-% Tidelift, ...
-% https://tidelift.com/subscription/pkg/pypi-numpy
-% $10,000 over 24 months according to
-% https://mail.python.org/pipermail/numpy-discussion/2018-September/078736.html
 
@@
 discussion before being accepted.
-
-% technical focus
-% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
-% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
-
-% Intel, QuanSight, ...
 
@@
 \section*{Recent technical improvements}
+
+% technical focus
+% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
+% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
+
+% Intel, QuanSight, ...
 

### 2019-12-20 paper.tex
Update recent technical improvements and discussion sections

@@
 
-% technical focus
-% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
-% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
-
-% Intel, QuanSight, ...
+NumPy has, throughout its lifetime, seen continuous development.
+E.g., bugs fixes, minor improvements, cleanups, and releases have
+happened regularly.  However, with the NEP process in place we now
+have the flexibility to plan and implement changes of larger scope.
+We highlight two of those below, as well as changes made to our
+testing infrastructure to support hardware platforms used in large
+scale computing.
 
@@
 
-The central position which NumPy takes within the numerical and scientific
+The central position of NumPy takes within the numerical and scientific
 Python ecosystem means that a vast number of projects are built on it.
@@
 much code is written specifically for NumPy arrays and uses the NumPy API as
-\code{import numpy as np}, it can nevertheless work for them.
-A brief example of such a function which is written using NumPy
-but can now be used more generally is:
+\code{import numpy as np}, it can nevertheless work for them.  For
+example, here is how a CuPy GPU array can be passed through NumPy for
+processing, with all operations being dispatched back to CuPy:
+
 \begin{lstlisting}
+  import numpy as np
+  import cupy as cp
+
+  x_gpu = cp.array([1, 2, 3])
+  y = np.sum(x_gpu)  # This works, and returns a GPU array
+\end{lstlisting}
+
+Similarly, user defined functions composed using NumPy can now be
+applied to, e.g., multi-node distributed Dask arrays:
+
+\begin{lstlisting}
+  import numpy as np
+  import dask.array as da
+
   def f(x):
+      """Function using NumPy API calls"""
       y = np.tensordot(x, x.T)
-    return np.mean(np.exp(y))
+      return np.mean(np.log(y + 1))
+
+  x_local = np.random.random([10000, 10000])  # random local array
+  x_distr = da.random.random([10000, 10000])  # random distributed array
+
+  f(x_local)  # returns a NumPy array
+  f(x_distr)  # works, returns a Dask array
 \end{lstlisting}
-For example, a user relying on this function and normally using it with a
-NumPy array, may now use a distributed Dask array when challenged
-by a very large dataset, without any modification to \code{f(x)}.
-
-%In order to aid better interaction between NumPy and these libraries,
-%an ``array function'' protocol was added to NumPy in 2018 (see Section
-%X). The new protocol allows for external array objects (sometimes
-%called ``duck arrays'') to pass through NumPy:
-%
-%\begin{lstlisting}
-%  import numpy as np
-%  import cupy as cp
-%
-%  x_gpu = cp.array([1, 2, 3])
-%  y = np.sum(x_gpu)  # This works, and returns a GPU array
-%\end{lstlisting}
 
@@
 Most of the technical work for the Moore and Sloan grant is now complete.
-The main remaining item is an overhaul of the data type system, which is discussed
-below.  Now that years of technical debt have been addressed, the project
-is in good shape for growing the community and better addressing the needs
-of new users.  That is the main focus of the CZI grant, which we discuss
-below.
+The main remaining item is an overhaul of the data type system.
+
+The meaning of each element within a NumPy array is described by its
+datatype (dtype). NumPy arrays can hold all common numerical
+datatypes, strings, datetimes, and generic Python objects, each of
+which is identified by a dtype.  We are busy overhauling the datatype
+system to make it behave consistently and to simplify the creation of
+custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
+of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
+While numerical dtypes make the foundation for most users,
+there are a multitude of use cases which cannot prosper due to current
+limitations, such as physical units\cite{astropy,Goldbaum2018,pint},
+% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
+geometrical objects\cite{pygeos}, and automatic
+differentiation\cite{pyadolc}.
+%Projects addressing these exist, each struggles with current
+%limitations.  % TODO: I may need citations, we could link github issues.
+
+%% While it is currently possible to define custom dtypes within NumPy, many of the
+%% use cases above cannot be solved using this system.
+%% This is evidenced also by code within NumPy which would require changes
+%% in many places to add such dtypes, even when adding them within NumPy itself.
+%% The new implementation will be simpler and more cleanly implemented in
+%% this regard.
+%% Custom dtypes and the basic dtypes included within NumPy will be implemented
+%% and behave identically as much as possible.
+%% At the same time redesigning the API will make NumPy dtypes more flexible to grow
+%% to future needs should they arise.
+
+%% As a further step, to make dtypes more accessible and allow a faster adoption of
+%% custom dtypes within the community, we plan to create a Python API to
+%% define such custom dtypes.
+
+Now that we have made good progress on addressing technical debt, the project
+is ready to grow its community of contributors and to better address the needs
+of an ever-growing number of new users.  This is the main focus of the
+forthcoming CZI grant.
+
+[... CZI grant ...]
+
+[... welcome new users ...]
+
+% technical focus
+% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
+% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
 
@@
 % Dtypes
-
-The meaning of each element within a NumPy array is described by its datatype
-(dtype). NumPy arrays can hold all common numerical datatypes, string, datetimes
-and generic Python objects, each of these is identified by a dtype.
-We are busy overhauling the datatype system to make it consistent (aligned with
-expectations), to simplify the creation of custom dtypes.
-While numerical dtypes make the foundation for most users, there are
-a multitude of use cases which cannot prosper due to current limitations.
-These use cases include datatypes with physical units\cite{astropy,Goldbaum2018,pint},
-% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
-geometrical objects\cite{pygeos}, automatic differentiation\cite{pyadolc}, and many more.
-Projects addressing these exist, each struggles with current
-limitations.  % TODO: I may need citations, we could link github issues.
-
-While it is currently possible to define custom dtypes within NumPy, many of the
-use cases above cannot be solved using this system.
-This is evidenced also by code within NumPy which would require changes
-in many places to add such dtypes, even when adding them within NumPy itself.
-The new implementation will be simpler and more cleanly implemented in
-this regard.
-Custom dtypes and the basic dtypes included within NumPy will be implemented
-and behave identically as much as possible.
-At the same time redesigning the API will make NumPy dtypes more flexible to grow
-to future needs should they arise.
-
-As a further step, to make dtypes more accessible and allow a faster adoption of
-custom dtypes within the community, we plan to create a Python API to
-define such custom dtypes.
-
-A \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}
-is under development in consultation with the community.
 

### 2019-12-20 body.tex
Split body from header

--- 2019-12-20 paper.tex
@@
-\documentclass[fleqn,10pt]{wlscirep}
-
-% Packages
-\usepackage{pifont}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-\usepackage{float}
-% \usepackage[multiple]{footmisc}
-\usepackage{booktabs}
-\usepackage{csquotes}
-\usepackage{moreverb} % for verbatim ouput (for wordcount)
-\usepackage{listings, textcomp}
-\lstset{language=Python}
-
-\lstset{ %
-  basicstyle=\ttfamily\footnotesize,  % size of fonts used for the code
-  breaklines=true,   % automatic line breaking only at whitespace
-  captionpos=b,   % sets the caption-position to bottom
-  commentstyle=\color{gray},  % comment style
-  keywordstyle=\color{blue},  % keyword style
-  stringstyle=\color{red},  % string literal style
-  upquote=true  %straight single quotes (requires textcomp)
-}
-
-% New Commands
-\newcommand{\cmark}{\ding{51}}%
-\newcommand{\xmark}{\ding{55}}%
-\newcommand{\code}[1]{\texttt{#1}}
-\newcommand{\fixme}[1]{\textcolor{red}{{#1}}}
-\newcommand{\inlinecite}[1]{\footnotesize\citen{#1}}
-\newcommand{\tightlist}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd{\@maketitle}
- {\noindent
-{\parbox{\dimexpr\linewidth-2\fboxsep\relax}{\color{color1}\large\sffamily\textbf{ABSTRACT}}}}
- {\noindent
-{\parbox{\dimexpr\linewidth-2\fboxsep\relax}{\color{color1}\large\sffamily\textbf{ABSTRACT}}}}
- {}{}
-\makeatother
-
-\usepackage{lineno}
-\linenumbers
-
-\title{NumPy---Array Computation for Python}
-
-\input{authors}
-
-\keywords{Scientific computing, Python, Mathematics}
-
-\begin{abstract}
-  NumPy is the fundamental array computing library for Python.
-  It has been actively developed for over two decades and has
-  more than 850 contributors, nearly 300,000 dependent repositories
-  on GitHub, and millions of downloads per year.
-  It is extensively used in research and teaching.
-  According to a 2018 survey, NumPy is the second most commonly used machine
-  learning tool among enterprises \cite{451report2018}.
-  In this article, we provide an overview of the capabilities and development
-  practices of the NumPy library and highlight some recent technical
-  developments.
-\end{abstract}
-
-\begin{document}
-
-\flushbottom
-\maketitle
-\thispagestyle{empty}
-
-% Count of words
-\verbatiminput{wordcount.tex}
-
 \section*{Introduction}
@@
 %\subsection*{Website}
-
-\section*{Data Availability}
-
-All NumPy source code and most data generated for the current study are
-available in the NumPy GitHub repository, \url{https://github.com/numpy}. Some
-supporting code and data have also been stored in other public repositories
-cited by this manuscript.
-
-\bibliography{references}
-
-\section*{Acknowledgments}
-
-\section*{Author Contributions Statement}
-
-%Must include all authors, identified by initials, for example: A.A.
-%conceived the experiment(s),  A.A. and B.A. conducted the experiment(s), C.A.
-%and D.A. analysed the results.  All authors reviewed the manuscript.
-
-\section*{Competing Interests}
-
-The authors declare no competing interests.
-
-\section*{Consortium}
-\subsection*{NumPy Contributors}
-\input{consortium}
-
-\end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:

### 2019-12-20 paper.tex
Split body from header

--- 2019-12-20 paper.tex
@@
 
-\section*{Introduction}
-
-NumPy is the fundamental array computation library for the Python
-ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-It underpins almost every library that does numerical
-computation, including SciPy.  NumPy consists of the \code{ndarray}-object
-along with utility functions to operate on it.  Because of
-its inherent simplicity—being a pointer to memory with some
-associated meta-data about shape, data-type, and so forth—the NumPy
-array is the {\it de facto} exchange format for array data in
-Python.
-NumPy, while having been wildly successful as a library and array exchange
-project, has now also become a standard language for expressing array and
-tensor computations.  
-It has such widespread adoption that not only the array object but also its
-{\it Application Programming Interface} (API) has become ubiquitous, and is
-implemented by, for example, popular tensor computation libraries.
-
-The arrival of a stable version of NumPy (1.0) in October 2006 unified
-the scientific Python community, and heralded the era of popular
-scientific computing in Python that has led to formidable successes
-such as [cite some impacts here].
-
-[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
-  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
-  IDL is now below Python, Pangeo <-> xarray, Earthcube,
-  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
-  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
-  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]
-
-\subsection*{History}
-
-NumPy was, at its inception, the unifier of the Python array
-processing world.  Before NumPy, two packages—Numeric and
-numarray—existed.  The first was developed by Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
-Hinsen, Travis Oliphant, and others, and was aimed at fast manipulation of many
-small arrays.  In contrast, to handle large astronomical images,
-NumArray was developed at the Space Telescope Science Institute (STScI).
-In other words, these array
-objects had different strengths and weaknesses, and therefore tended
-to divide the already small user community.  Furthermore, the divide
-posed a challenge for SciPy, which built on Numeric and provided
-fundamental scientific routines \cite{virtanen2019scipy}.
-In response, the maintainer of Numeric, Travis Oliphant, started a
-major rewrite that aimed to be a ``best of both worlds'' unification
-of the two projects.  This became known as NumPy.
-
-While, by 2006, NumPy was already eminently usable, it lacked good
-documentation. In order to use NumPy and SciPy in his classes, Joe Harrington
-from the University of South Florida sponsored several rounds of the
-NumPy Documentation Marathon \cite{harrington2008scipy,vanderwalt2008scipy}.
-During these intensive
-documentation improvement sessions, we developed a Wikipedia-like
-system for community members to contribute so-called
-``docstrings''—text describing each NumPy function. A community
-documentation standard was published so that these pieces of
-documentation would all provide similar information, and conform to a
-basic standard of quality. Experienced community members reviewed and
-integrated contributed docstrings, and soon NumPy had thorough
-documentation coverage.
-
-To coordinate work as a global community, to ensure accuracy of computed
-results, and to allow modification without breakage, the NumPy community
-pioneered best software engineering practices in the scientific Python
-community  \cite{millman2014developing}.
-For example, NumPy early on adopted distributed revision control, unit testing,
-continuous integration, code review, and standardized executable documentation.
-
-However, after his initial effort to build NumPy, Travis Oliphant
-stepped back from development to found a company and several projects,
-including Numba---an accelerated Python compiler that could vastly
-accelerate NumPy expressions.  By 2009, the project had entered a
-period of uncoordinated, opportunistic development, and was
-increasingly weighed down by technical debt that had accrued since the
-days of Numeric.
-
-\subsection*{Funding}
-
-For many years, NumPy operated without external funding, development
-time mostly being contributed freely by students and researchers in
-their spare time.  In-person meetings were sponsored off of
-individual's research grants, and while industry sponsored
-contributions were made, notably those by Mark Wiebe when he was
-sponsored by Travis Oliphant at Continuum Analytics, mostly the
-package had rapid adoption and became central to scientific computing
-without external investment.
-
-% BIDS -- UCB
-% https://www.moore.org/grant-detail?grantId=GBMF5447
-% $645,020 in 2016
-% https://sloan.org/grant-detail/8222
-% $659,359 in 2017
-% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
-
-In 2017, the Berkeley Institute for Data Science
-(BIDS) received grants totaling 1.3M USD from the Gordon \& Betty
-Moore and the Alfred P. Sloan foundations to support the development
-of NumPy for better data science.
-Specifically, the aim was to address technical debt and set in place
-standards and architecture to encourage more sustainable development.
-Stéfan van der Walt, a senior research data scientist at BIDS and a
-member of the NumPy steering council, is the PI and manages four
-programmers working on the project.
-
-% CZI -- NumFOCUS/QuanSight
-% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
-% NumPy and OpenBLAS received $195,000 in 2019
-% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
-
-In 2019, NumFOCUS received a grant for \$195,000 in 2019 for the
-development of NumPy and OpenBLAS (on which NumPy depends for
-accelerated linear algebra) from the Chan Zuckerberg Initiative.
-This grant, building on the technical work supported by the previous
-grant, will focus on better serving NumPy's large number of beginner
-to intermediate level users and on growing the community of NumPy
-contributors.
-Ralf Gommers, the Director of Quansight Labs and a member of the NumPy
-steering council, is the PI.
-
-The project receives around 10K USD per year from Tidelift, which is
-used to fund documentation and website developers.
-
-\subsection*{Current status}
-
-% (maturity, users)
-
-In 2007, the NumPy release manager, Jarrod Millman, adopted the notion
-of enhancement proposals---used infrequently until 2017 when, to
-facilitate work expected from funded development, he formalized the
-NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
-Python Enhancement Proposals (PEPs) for ``proposing major new
-features, for collecting community input on an issue, and for
-documenting the design decisions that have gone into
-Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
-Since then there have been 19 proposed NEPS---6 have been implemented,
-4 have been accepted and are being implemented, 4 are under
-consideration, 3 have been deferred or superseded, and 2 have been rejected
-or withdrawn.
-
-The first activities organized for the new grant were a NEP development
-sprint, a weekly community call, and a meeting to write a draft roadmap.
-
-Since the roadmap was meant to represent community consensus on future
-development, it was presented for comment at the annual SciPy
-conference during a so-called ``Birds of a Feather'' session, attended
-by more than 100 people.  One additional meeting was held to discuss
-specific enhancements, such as the development of a new data-type
-system.  The final roadmap was vetted by the community via online
-discussion before being accepted.
-
-The weekly community calls alternate weekly between triage meetings and
-higher level discussion.  The calls not only involve developers from
-the community, but provide a venue for vendors and other external
-groups to provide input.  For example, after Intel produced a forked
-version of NumPy, one of their developers joined a call to discuss
-community concerns.
-
-NumPy plays a central role in building and standardizing much of the scientific
-Python community infrastructure.
-The docstring standard mentioned in the history section is now widely adopted.
-We are also now using the NEP system as a way to help coordinate the larger
-scientific Python community.
-% https://numpy.org/neps/nep-0029-deprecation_policy.html
-For example, in NEP 29, we recommend, along with leaders from various other
-projects, that all projects across the Scientific Python ecosystem adopt a
-common ``time window-based'' policy for support of Python and NumPy versions.
-This standard will simplify downstream project and release planning.
-
-% https://numpy.org/neps/scope.html
-% https://numpy.org/neps/roadmap.html
-
-We showed that NumPy is actively developed according to a community
-vetted plan that integrates formal enhancement proposals.  Next, we
-give a more in-depth overview of the team developing NumPy, their
-activities, as well as the governance and community structures that
-guide the collaboration.
-
-\section*{Project organization and community}
-
-NumPy is currently maintained by a group of 23 contributors with commit rights
-to the NumPy code base. Out of these 17 maintainers were active in 2019.
-Additionally, there are a few long term developers who contributed and maintain
-specific parts of NumPy, but are not officially maintainers.
-At a release cycle of about every half year, the five recent releases in the years
-2018 and 2019 have averaged about 450~pull requests~(PR) each.\footnote{
-    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
-    All data provided here is based on the development which happened through github
-    pull requests. In some cases contributions by maintainers may not be categorized as such.}
-% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
-With each release attracting more than a hundred new contributors.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
-    \caption{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012. The total number of PR is indicated with the
-        lower blue area showing the portion contributed by current or previous
-        maintainers.}\label{fig:prs-over-time}
-\end{figure}
-
-The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
-PR merged into the NumPy master branch.
-In this plot PR opened by current and old maintainers are indicated.
-Although the number of PRs being merged fluctuates,
-the plot indicates an increased number of contributions over the past
-years.
-Over the course of its history, NumPy has attracted PR by 823 contributors.
-However, its development still relies heavily on a relatively small number
-of active maintainers, who share more than half of the contributions among
-themselves.
-
-% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
-% https://github.com/numpy/numpy/pull/6352
-NumPy adopted an official Governance Document on October~5,
-2015\cite{NumPyProjectGovernance}.
-Project decisions are usually made by consensus of interested contributors.
-This means for many thing everyone is entrusted with veto power.
-A Steering Council, currently composed of 12~members, facilitates this
-process and oversees daily development of the project by contributing code
-and reviewing contributions from the community.
-
-% https://mail.python.org/pipermail/numpy-discussion/2018-July/078476.html
-% https://github.com/numpy/numpy/pull/11865
-NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
-In brief, we strive to:
-\emph{be open};
-\emph{be empathetic, welcoming, friendly, and patient};
-\emph{be collaborative};
-\emph{be inquisitive}; and
-\emph{be careful in the words that we choose}.
-The Code of Conduct also specifies how breaches can be reported and outlines
-the process for responding to such reports.
-
-% \subsection*{GSoD}
-
-\section*{Library organization}
-
-The NumPy array data structure and universal functions, which we describe
-below, make up the core of NumPy.
-The NumPy library also includes
-a set of library functions for manipulating arrays and doing scientific computation,
-libraries for unit tests and Python package building,
-and the program \code{f2py}\cite{peterson2009f2py} for wrapping Fortran code in Python.
-
-\subsection*{Core}
-
-The \code{ndarray} data structure stores regularly strided homogeneous data
-types in a contiguous block memory, allowing for the efficient representation
-of $n$-dimensional data.
-A \emph{masked array} class, built on top of the \code{ndarray}, is
-provided as well. This class allows elements of an array to be masked,
-indicating that the data is missing or unknown.
-More details about the data structure are given in ``The NumPy array:
-a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.
-
-The \emph{universal functions} (or \emph{ufuncs})
-are functions written in C that implement efficient looping over
-NumPy arrays. An important feature of ufuncs is the built-in
-implementation of \emph{broadcasting}.  For example, the function
-\code{arctan2(x, y)} is a ufunc that accepts two values and computes
-$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
-the ufunc will take care of looping over the dimensions of the inputs
-in such a way that if, say, \code{x} is a 1-d array with length 3, and
-\code{y} is a 2-d array with shape $2 \times 1$, the output will be
-an array with shape $2 \times 3$.  The ufunc machinery takes care
-of calling the function with all the appropriate combinations of
-input array elements to complete the output array.
-The elementary arithmetic operations of addition, multiplication, etc.,
-are implemented as ufuncs, so broadcasting applies even in expressions
-such as \code{x + y * z}.
-
-\subsection*{Computing libraries}
-
-NumPy provides a large library of functions for array manipulation
-including functions for: creating, reshaping, concatenating, and padding arrays;
-searching, sorting and counting data
-in arrays; computing elementary statistics, such as the mean, median,
-variance, and standard deviation; file I/O; and more.
-For historical reasons, NumPy also includes basic functionality for
-linear algebra,
-fast Fourier transforms and windowing,
-and polynomial fitting.
-We will keep supporting these, but we will not expand beyond them.
-
-% use scipy for linear algebra etc...
-
-%A suite of functions for computing the \emph{fast Fourier transform (FFT)}
-%and its inverse is provided.
-
-%NumPy's linear algebra library includes functions for: solving linear
-%systems of equations; computing various functions of a matrix, including
-%the determinant, the norm, the inverse, and the pseudo-inverse;
-%computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
-%and more.
-%
-
-%For historical reasons, we provide a \code{matrix} class the implements 2-d
-%arrays with specialized behavior.
-%But we strongly recommend you use a 2-d array and use the infix \code{@} operator
-%for matrix multiplication.
-%
-%There is also a \code{matrix} class the implements 2-d arrays with
-%specialized behavior.  We mention it here for completeness, but the
-%use of \code{matrix} is discouraged.  The quirks of the \code{matrix}
-%class have proven to be troublesome, and with the addition of the
-%infix \code{@} operator to Python for matrix multiplication, there was no
-%longer a need for a separate \code{matrix} class, since one can
-%use a 2-d array instead.
-
-\subsection*{Infrastructure libraries}
-
-Being at the base of the scientific Python stack, NumPy provides some basic
-infrastructure for other packages in the scientific Python ecosystem.
-The \code{numpy.testing} subpackage provides functions such as
-\code{assert\_allclose(actual, desired)} that may be used in unit
-test suites for code that uses NumPy arrays.
-The \code{numpy.distutils} subpackage provides build support for C++, Fortran,
-BLAS/LAPACK, and other relevant libraries for scientific computing.
-The program \code{f2py} is a tool for
-building NumPy-aware Python wrappers of Fortran functions.
-NumPy itself does not use any Fortran code;  F2PY is part of NumPy
-for historical reasons.
-
-% \section*{Common infrastructure}
-
-%\subsection*{Wheels build system}
-%
-%A Python \emph{wheel}\cite{PEP427} is a standard file format for
-%distributing Python libraries.  In addition to Python code, a
-%wheel may include compiled C extensions and other binary data.
-%This is important, because many libraries, including NumPy,
-%require a C compiler and other build tools to build the software
-%from the source code, making it difficult for many users to install
-%the software on their own.  The introduction of wheels to the Python
-%packaging system has made it much easier for users to install
-%precompiled libraries.
-%
-%A GitHub repository containing scripts to build NumPy wheels has
-%been configured so that a simple commit to the repository triggers
-%an automated build system that creates NumPy wheels for several
-%computer platforms, including Windows, Mac OSX and Linux.  The wheels
-%are uploaded to a public server and made available for anyone to use.
-%This system makes it easy for users to install precompiled versions
-%of NumPy on these platforms.
-%
-%The technology that is used to build the wheels evolves continually.
-%At the time this paper is being written, a key component is the
-%\code{multibuild} suite of tools developed by Matthew Brett and
-%other developers\cite{multibuild}.  Currently, scripts using
-%\code{multibuild} are written for the continuous integration
-%platforms Travis-CI (for Linux and Mac OSX) and Appveyor
-%(for Windows).
-
-\section*{Recent technical improvements}
-
-% technical focus
-% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
-% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
-
-% Intel, QuanSight, ...
-
-\subsection*{The Array Function Protocol}
-
-The central position which NumPy takes within the numerical and scientific
-Python ecosystem  means that a vast number of projects are built on it.
-These projects are consumers of the NumPy API.
-
-On the other side, projects are providers of a \emph{NumPy-like API} and array
-objects targeting audiences with specialized needs beyond NumPy's capabilities.
-For example, the NumPy API is implemented by
-several popular tensor computation libraries including
-CuPy\footnote{\url{https://cupy.chainer.org/}},
-Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
-Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
-and to a lesser extent
-TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.
-It is also implemented in packages that support sparse arrays
-such as \code{scipy.sparse} and \code{pydata.sparse}.
-Another notable example is Dask, a library for parallel computing in
-Python.  Dask adopts the NumPy API and therefore presents a familiar
-interface to existing NumPy users, while adding powerful abilities to
-parallelize and distribute tasks.
-
-The multitude of specialized projects creates the difficulty that consumers
-of these NumPy-like APIs write code specific to a single project and do not support
-all of the above array providers.
-This is a burden for users relying on the specialized Array-like, since
-a tool they need may not work for them.
-Additionally, a user who has the need of transitioning from NumPy to a more
-specialized array, is faced with a challenge.
-These issue can lead into a fragmentation of the user bases into consumers
-of different NumPy-like APIs.
-
-To address these issues NumPy has the goal of providing the fundamental
-API for \emph{interoperability} between the various NumPy-like APIs.
-An earlier step in this direction was the implementation of the
-\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
-for most mathematical functions.\cite{NEP13}
-In 2019 this was expanded more generally with the inclusion of the
-\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
-These two protocols allows providers of Array-like APIs to be interoperable
-with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
-For the users relying on specialized array projects it means that even though
-much code is written specifically for NumPy arrays and uses the NumPy API as
-\code{import numpy as np}, it can nevertheless work for them.
-A brief example of such a function which is written using NumPy
-but can now be used more generally is:
-\begin{lstlisting}
-def f(x):
-    y = np.tensordot(x, x.T)
-    return np.mean(np.exp(y))
-\end{lstlisting}
-For example, a user relying on this function and normally using it with a
-NumPy array, may now use a distributed Dask array when challenged
-by a very large dataset, without any modification to \code{f(x)}.
-
-%In order to aid better interaction between NumPy and these libraries,
-%an ``array function'' protocol was added to NumPy in 2018 (see Section
-%X). The new protocol allows for external array objects (sometimes
-%called ``duck arrays'') to pass through NumPy:
-%
-%\begin{lstlisting}
-%  import numpy as np
-%  import cupy as cp
-%
-%  x_gpu = cp.array([1, 2, 3])
-%  y = np.sum(x_gpu)  # This works, and returns a GPU array
-%\end{lstlisting}
-
-\subsection*{Random}
-
-%The random number generator library in NumPy provides several alternative
-%\emph{bit stream generators} that provide the core function of generating
-%random integers.
-%A higher-level generator class that implements an assortment of
-%probability distributions is provided. It includes the beta, gamma
-%and Weibull distributions, the univariate and multivariate normal
-%distributions, and more.
-
-The NumPy random module provides pseudorandom numbers from a wide range of
-distributions. In legacy versions of NumPy, simulated random values are produced
-by a \code{RandomState} object that: handles seeding and state initialization;
-wraps the core pseudorandom number generator based on a 32-bit implementation of
-MT19937; interfaces with the underlying code that transforms random bits into
-deviates from other distributions; and supplies a singleton instance exposed in
-the root of the random module.
-
-The \code{RandomState} object makes a compatibility guarantee so that a fixed
-seed and sequence of function calls produce the same set of values. This
-guarantee has slowed progress since improving the underlying code requires
-extending the API with additional keyword arguments. This guarantee continues to
-apply to \code{RandomState}.
-
-% Mention explicitly the NEPs involved
-
-NumPy 1.17 introduced a new API for generating random numbers that use a more
-flexible structure that can be extended by libraries or end-users. The new API
-is built using components that separate the steps required to generate random
-variates. Pseudorandom bits are generated by a bit generator. These bits are
-then transformed into variates from complex distributions by a generator.
-Finally, seeding is handled by an object that produces sequences of high-quality
-initial values.
-
-Bit generators are simple classes that manage the state of an underlying
-pseudorandom number generator. NumPy ships with four bit generators. The default
-bit generator is a 64-bit implementation of the Permuted Congruential Generator
-\cite{pcg64} (\code{PCG64}). The three other bit generators are a 64-bit version
-of the Philox generator\cite{random123} (\code{Philox}), Chris Doty-Humphrey's
-Small Fast Chaotic generator\cite{practrand} (\code{SFC64}), and the 32-bit
-Mersenne Twister\cite{mt19937} (\code{MT19937}) which has been used in older
-versions of NumPy.\footnote{The
-\href{https://github.com/bashtage/randomgen}{randomgen project} supplies a wide
-range of alternative bit generators such as a cryptographic counter-based
-generators (\code{AESCtr}) and generators that expose hardware random number
-generators (\code{RDRAND})\cite{randomgen}.} Bit generators export a capsule
-containing pointers to functions that produce 64 bits, 32 bits, or a random
-double. They also expose CFFI and ctypes interfaces to the same three functions.
-
-The \code{Generator} consumes one of the bit generators and produces variates
-from complicated distributions. Many improved methods for generating random
-variates from common distributions were implemented, including the Ziggurat
-method for normal, exponential and gamma variates\cite{ziggurat}, and Lemire's
-method for bounded random integer generation\cite{lemire}. The \code{Generator}
-is more similar to the legacy \code{RandomState}, and its API is substantially
-the same. The key differences all relate to state management, which has been
-delegated to the bit generator. The \code{Generator} does not make the same
-stream guarantee as the \code{RandomState} object, and so variates may differ
-across versions as improved generation algorithms are
-introduced.\footnote{Despite the removal of the compatibility guarantee, simple
-reproducibility across versions is encouraged, and minor changes that do not
-produce meaningful performance gains or fix underlying bug are not generally
-adopted.}
-
-Finally, a \code{SeedSequence} is used to initialize a bit generator. The seed
-sequence can be initialized with no arguments, in which case it reads entropy
-from a system-dependent provider, or with a user-provided seed. The seed
-sequence then transforms the initial set of entropy into a sequence of
-high-quality pseudorandom integers, which can be used to initialize multiple bit
-generators deterministically. A key design goal of a seed sequence was to be
-splittable in the sense that a seed sequence can be used to produce child
-sequences that are distinct from their parents or other ancestors. This
-capability allows a seed sequence to be used in large distributed applications
-where the number of workers required is not known. The sequences generated from
-the same initial entropy and the same splits are fully deterministic to ensure
-reproducibility.
-
-The three components are combined to construct a complete random number
-generator.
-
-\begin{lstlisting}
-from numpy.random import Generator, PCG64, SeedSequence
-seq = SeedSequence(10304245474441179933310169597522947680)
-pcg = PCG64(seq)
-gen = Generator(pcg)
-\end{lstlisting}
-
-\noindent This approach retains access to the seed sequence which can then be
-used to spawn additional generators.
-
-\begin{lstlisting}
-children = seq.spawn(2)
-gen_0 = Generator(PCG64(children[0]))
-gen_1 = Generator(PCG64(children[1]))
-\end{lstlisting}
-
-\noindent While this approach retains complete flexibility, the method
-\code{np.random.default\_rng} can be used to instantiate a \code{Generator} when
-reproducibility is not needed.
-
-The final goal of the new API is to improve extensibility. \code{RandomState} is
-a monolithic object that obscures all of the underlying state and functions. The
-component architecture is one part of the extensibility improvements. The
-underlying functions (written in C) which transform the output of a bit
-generator to other distributions are available for use in CFFI. This allows the
-same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
-numba. Both the bit generators and the low-level functions can also be used in C
-or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
-NumPy source. Alternative approaches that avoid this extra step are being
-explored.}
-
-\subsection*{Testing on multiple architectures}
-
-% started at as focused small matrices, class examples
-
-At the time of writing the two fastest supercomputers in the
-world, Summit and Sierra, both have IBM POWER9 architectures
-\cite{top500nov2019}. In late 2018, Astra, the first ARM-based
-supercomputer to enter the TOP500 list, went into production\cite{
-astra-wiki}. Furthermore, over 100 billion ARM processors have been
-produced as of 2017\cite{arm-architecture}, making it the most 
-widely used instruction set architecture in the world.
-
-Clearly there are motivations for a large scientific computing
-software library to support POWER and ARM architectures. We've extended
-our continuous integration (CI) testing to include \texttt{ppc64le}
-(POWER8 on Travis CI) and ARMv8 (on Shippable service). We also test
-with the s390x architecture (IBM Z CPUs on Travis CI) so that we
-can probe the behavior of our library on a big-endian machine.
-This satisfies one of the major components of
-improved CI testing laid out in a version of our roadmap
-\cite{numpy-roadmap}---specifically, ``CI for more exotic
-platforms."
-
-PEP 599\cite{PEP599} lays out a plan for new Python binary wheel
-distribution support, \texttt{manylinux2014}, that adds
-support for a number of architectures supported by the CentOS
-Alternative Architecture Special Interest Group, including
-ARMv8, ppc64le, as well as s390x. We are thus well-positioned
-for a future where provision of binaries on these architectures
-will be expected for a library at the base of the ecosystem.
-
-\section*{Discussion}
-
-% 1st grant done, technical cleanup finished except dtypes
-% 2nd grant starting, community growth and new user focus
-
-Most of the technical work for the Moore and Sloan grant is now complete.
-The main remaining item is an overhaul of the data type system, which is discussed
-below.  Now that years of technical debt have been addressed, the project
-is in good shape for growing the community and better addressing the needs
-of new users.  That is the main focus of the CZI grant, which we discuss
-below.
-
-% \subsubsection*{Future Development}
-
-% Dtypes
-
-The meaning of each element within a NumPy array is described by its datatype
-(dtype). NumPy arrays can hold all common numerical datatypes, string, datetimes
-and generic Python objects, each of these is identified by a dtype.
-We are busy overhauling the datatype system to make it consistent (aligned with
-expectations), to simplify the creation of custom dtypes.
-While numerical dtypes make the foundation for most users, there are
-a multitude of use cases which cannot prosper due to current limitations.
-These use cases include datatypes with physical units\cite{astropy,Goldbaum2018,pint},
-% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
-geometrical objects\cite{pygeos}, automatic differentiation\cite{pyadolc}, and many more.
-Projects addressing these exist, each struggles with current
-limitations.  % TODO: I may need citations, we could link github issues.
-
-While it is currently possible to define custom dtypes within NumPy, many of the
-use cases above cannot be solved using this system.
-This is evidenced also by code within NumPy which would require changes
-in many places to add such dtypes, even when adding them within NumPy itself.
-The new implementation will be simpler and more cleanly implemented in
-this regard.
-Custom dtypes and the basic dtypes included within NumPy will be implemented
-and behave identically as much as possible.
-At the same time redesigning the API will make NumPy dtypes more flexible to grow
-to future needs should they arise.
-
-As a further step, to make dtypes more accessible and allow a faster adoption of
-custom dtypes within the community, we plan to create a Python API to
-define such custom dtypes.
-
-A \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}
-is under development in consultation with the community.
-
-% outline plan for the CZI grant and an invitation to new contributors
-
-%\subsection*{Website}
+\input{body}
 
@@
 \end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:

### 2019-12-20 paper.tex
Add acknowledgments and contributions

@@
 
+S.J.v.d.W and K.J.M. were funded in part by the Gordon and Betty Moore
+Foundation through Grant GBMF3834 and by the Alfred P. Sloan Foundation through
+Grant 2013-10-27 to the University of California, Berkeley.
+S.B., T.J.R., and W.W. were funded in part by the Gordon and Betty Moore 
+Foundation through Grant GBMF5447 and by the Alfred P. Sloan Foundation through
+Grant G-2017-9960 to the University of California, Berkeley.
+
 \section*{Author Contributions Statement}
 
-%Must include all authors, identified by initials, for example: A.A.
-%conceived the experiment(s),  A.A. and B.A. conducted the experiment(s), C.A.
-%and D.A. analysed the results.  All authors reviewed the manuscript.
+%C.R.H and T.E.O.
+S.J.v.d.W and K.J.M. composed the manuscript with input from others.
+All authors have contributed significant code, documentation, and/or expertise
+to the NumPy project.
+All authors reviewed the manuscript.
 

### 2019-12-20 body.tex
Remove cruft
Revise intro

@@
 ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-It underpins almost every library that does numerical
-computation, including SciPy.  NumPy consists of the \code{ndarray}-object
-along with utility functions to operate on it.  Because of
-its inherent simplicity—being a pointer to memory with some
-associated meta-data about shape, data-type, and so forth—the NumPy
-array is the {\it de facto} exchange format for array data in
-Python.
+Since its first stable release (1.0) in October 2006 it has unified
+the scientific Python community and now underpins almost every library
+that does numerical computation, including SciPy\cite{virtanen2019scipy},
+pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
+and scikit-image\cite{vanderwalt2014scikit}.
+Not surprisingly, many important scientific research projects and
+results rely on NumPy.
+%[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
+%  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
+%  IDL is now below Python, Pangeo <-> xarray, Earthcube,
+%  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
+%  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
+%  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]
+
+NumPy consists of the \code{ndarray}-object along with utility functions to
+operate on it.
+Because of its inherent simplicity—being a pointer to memory with some
+associated meta-data about shape, data-type, and so forth—the NumPy array is
+the {\it de facto} exchange format for array data in Python.
 NumPy, while having been wildly successful as a library and array exchange
@@
 
-The arrival of a stable version of NumPy (1.0) in October 2006 unified
-the scientific Python community, and heralded the era of popular
-scientific computing in Python that has led to formidable successes
-such as [cite some impacts here].
-
-[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
-  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
-  IDL is now below Python, Pangeo <-> xarray, Earthcube,
-  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
-  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
-  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]
-
 \subsection*{History}
 
-NumPy was, at its inception, the unifier of the Python array
-processing world.  Before NumPy, two packages—Numeric and
+Before NumPy, two different Python array packages—Numeric and
 numarray—existed.  The first was developed by Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
@@
 NumArray was developed at the Space Telescope Science Institute (STScI).
-In other words, these array
-objects had different strengths and weaknesses, and therefore tended
-to divide the already small user community.  Furthermore, the divide
-posed a challenge for SciPy, which built on Numeric and provided
-fundamental scientific routines \cite{virtanen2019scipy}.
+These array objects had different strengths and weaknesses, and therefore
+divided the already small community.  Furthermore, the divide
+posed a challenge for SciPy, which was built on Numeric and provided
+fundamental scientific routines.
 In response, the maintainer of Numeric, Travis Oliphant, started a
@@
 
-While, by 2006, NumPy was already eminently usable, it lacked good
-documentation. In order to use NumPy and SciPy in his classes, Joe Harrington
-from the University of South Florida sponsored several rounds of the
-NumPy Documentation Marathon \cite{harrington2008scipy,vanderwalt2008scipy}.
-During these intensive
-documentation improvement sessions, we developed a Wikipedia-like
-system for community members to contribute so-called
-``docstrings''—text describing each NumPy function. A community
-documentation standard was published so that these pieces of
-documentation would all provide similar information, and conform to a
-basic standard of quality. Experienced community members reviewed and
-integrated contributed docstrings, and soon NumPy had thorough
-documentation coverage.
-
+This spurred renewed development in both SciPy and NumPy and 
+heralded the current era of wide-spread use of Python for scientific computing.
 To coordinate work as a global community, to ensure accuracy of computed
-results, and to allow modification without breakage, the NumPy community
+results, and to allow modification without breakage, the NumPy (and SciPy) community
 pioneered best software engineering practices in the scientific Python
@@
 continuous integration, code review, and standardized executable documentation.
+It also developed a standard for ``docstrings''—text describing each NumPy function.
+This quickly became the standard for the larger scientific Python community.
 
-However, after his initial effort to build NumPy, Travis Oliphant
-stepped back from development to found a company and several projects,
-including Numba---an accelerated Python compiler that could vastly
-accelerate NumPy expressions.  By 2009, the project had entered a
-period of uncoordinated, opportunistic development, and was
-increasingly weighed down by technical debt that had accrued since the
-days of Numeric.
+As the scientific Python community grew and new challenges arose, Travis Oliphant
+eventually stepped back from leading NumPy to found a company and start
+new projects including Numba---an accelerated Python compiler that could vastly
+accelerate NumPy expressions.
+Development of NumPy continued by a core group that informally operated by consensus.
+
+%By 2009, the project had entered a
+%period of uncoordinated, opportunistic development, and was
+%increasingly weighed down by technical debt that had accrued since the
+%days of Numeric.
 
@@
 
-For many years, NumPy operated without external funding, development
-time mostly being contributed freely by students and researchers in
+For many years, NumPy had no dedicated funding, development
+time was mostly contributed freely by students and researchers in
 their spare time.  In-person meetings were sponsored off of
@@
 % https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
-
-In 2017, the Berkeley Institute for Data Science
-(BIDS) received grants totaling 1.3M USD from the Gordon \& Betty
-Moore and the Alfred P. Sloan foundations to support the development
-of NumPy for better data science.
-Specifically, the aim was to address technical debt and set in place
-standards and architecture to encourage more sustainable development.
-Stéfan van der Walt, a senior research data scientist at BIDS and a
-member of the NumPy steering council, is the PI and manages four
-programmers working on the project.
+In 2017, NumPy received its first large grants totaling 1.3M USD from the
+Gordon \& Betty Moore and the Alfred P. Sloan foundations.
+Stéfan van der Walt is the PI and manages four programmers working on the project.
+These two grants focus on addressing the technical debt accrued over the years and
+on setting in place standards and architecture to encourage more sustainable development.
 
@@
 % https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
-
-In 2019, NumFOCUS received a grant for \$195,000 in 2019 for the
-development of NumPy and OpenBLAS (on which NumPy depends for
-accelerated linear algebra) from the Chan Zuckerberg Initiative.
+NumPy received another grant for \$195,000 from the Chan Zuckerberg
+Initiative at the end of 2019.
+Ralf Gommers is the PI.
 This grant, building on the technical work supported by the previous
@@
 contributors.
-Ralf Gommers, the Director of Quansight Labs and a member of the NumPy
-steering council, is the PI.
+It also will provide some support for OpenBLAS (on which NumPy depends for
+accelerated linear algebra).
 
-The project receives around 10K USD per year from Tidelift, which is
+The project also receives around 10K USD per year from Tidelift, which is
 used to fund documentation and website developers.
@@
 
-In 2007, the NumPy release manager, Jarrod Millman, adopted the notion
-of enhancement proposals---used infrequently until 2017 when, to
-facilitate work expected from funded development, he formalized the
+As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
+of enhancement proposals---used infrequently and informally until 2017 when,
+to facilitate work expected from funded development, he formalized the
 NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
@@
 
-% \subsection*{GSoD}
-
 \section*{Library organization}
@@
 
-% use scipy for linear algebra etc...
-
-%A suite of functions for computing the \emph{fast Fourier transform (FFT)}
-%and its inverse is provided.
-
-%NumPy's linear algebra library includes functions for: solving linear
-%systems of equations; computing various functions of a matrix, including
-%the determinant, the norm, the inverse, and the pseudo-inverse;
-%computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
-%and more.
-%
-
-%For historical reasons, we provide a \code{matrix} class the implements 2-d
-%arrays with specialized behavior.
-%But we strongly recommend you use a 2-d array and use the infix \code{@} operator
-%for matrix multiplication.
-%
-%There is also a \code{matrix} class the implements 2-d arrays with
-%specialized behavior.  We mention it here for completeness, but the
-%use of \code{matrix} is discouraged.  The quirks of the \code{matrix}
-%class have proven to be troublesome, and with the addition of the
-%infix \code{@} operator to Python for matrix multiplication, there was no
-%longer a need for a separate \code{matrix} class, since one can
-%use a 2-d array instead.
-
 \subsection*{Infrastructure libraries}
@@
 for historical reasons.
-
-% \section*{Common infrastructure}
-
-%\subsection*{Wheels build system}
-%
-%A Python \emph{wheel}\cite{PEP427} is a standard file format for
-%distributing Python libraries.  In addition to Python code, a
-%wheel may include compiled C extensions and other binary data.
-%This is important, because many libraries, including NumPy,
-%require a C compiler and other build tools to build the software
-%from the source code, making it difficult for many users to install
-%the software on their own.  The introduction of wheels to the Python
-%packaging system has made it much easier for users to install
-%precompiled libraries.
-%
-%A GitHub repository containing scripts to build NumPy wheels has
-%been configured so that a simple commit to the repository triggers
-%an automated build system that creates NumPy wheels for several
-%computer platforms, including Windows, Mac OSX and Linux.  The wheels
-%are uploaded to a public server and made available for anyone to use.
-%This system makes it easy for users to install precompiled versions
-%of NumPy on these platforms.
-%
-%The technology that is used to build the wheels evolves continually.
-%At the time this paper is being written, a key component is the
-%\code{multibuild} suite of tools developed by Matthew Brett and
-%other developers\cite{multibuild}.  Currently, scripts using
-%\code{multibuild} are written for the continuous integration
-%platforms Travis-CI (for Linux and Mac OSX) and Appveyor
-%(for Windows).
 
@@
 
-% started at as focused small matrices, class examples
-
 At the time of writing the two fastest supercomputers in the

### 2019-12-20 body.tex
More work on intro

@@
 
-Before NumPy, two different Python array packages—Numeric and
-numarray—existed.  The first was developed by Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
-Hinsen, Travis Oliphant, and others, and was aimed at fast manipulation of many
-small arrays.  In contrast, to handle large astronomical images,
-NumArray was developed at the Space Telescope Science Institute (STScI).
-These array objects had different strengths and weaknesses, and therefore
-divided the already small community.  Furthermore, the divide
-posed a challenge for SciPy, which was built on Numeric and provided
-fundamental scientific routines.
-In response, the maintainer of Numeric, Travis Oliphant, started a
-major rewrite that aimed to be a ``best of both worlds'' unification
-of the two projects.  This became known as NumPy.
+Before NumPy, two different Python array packages—Numeric and NumArray—existed.
+The first began in the mid-90s and was developed by Jim Fulton, Jim Hugunin,
+Paul Dubois, Konrad Hinsen, Travis Oliphant, and others, and was aimed at fast
+manipulation of many small arrays.
+By the beginning of the millennium, several libraries built on top of Numeric 
+the provided fundamental scientific routines `were combined to form SciPy.
+Due to the poor performance of Numeric on large data sets, NumArray was
+developed at the Space Telescope Science Institute (STScI) to handle
+large astronomical images such as those produced by the Hubble Space Telescope.
+It was initially considered the successor to Numeric.
+However, it performed poorly on small arrays.
+This divide posed a challenge for SciPy as it wasn't clear how to support
+both libraries.
+Working with the NumArray and SciPy community, the maintainer of Numeric,
+Travis Oliphant, started a major rewrite that aimed to be a ``best of both
+worlds'' unification of the two projects.  This became known as NumPy.
 
-This spurred renewed development in both SciPy and NumPy and 
-heralded the current era of wide-spread use of Python for scientific computing.
+The creation of NumPy spurred renewed development in both SciPy and the larger
+scientific Python ecosystem and heralded the current era of wide-spread use of
+Python for scientific computing.
 To coordinate work as a global community, to ensure accuracy of computed

### 2019-12-20 body.tex
Minor edits
Updates to abstract and intro

@@
 results rely on NumPy.
+It plays an essential role in many research analysis pipelines in
+fields as diverse as physics, biology, astronomy, neuroscience,
+material science, engineering, and chemistry.
+% w/ citations
+In particular, NumPy played a fundamental role in discovering
+gravitational waves\cite{pycbc}, imaging black
+holes\cite{eht-imaging}, and exploring the solar system (... Kepler? https://ui.adsabs.harvard.edu/abs/2012ascl.soft08004S/abstract).
+
+...
 %[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
@@
 
-NumPy consists of the \code{ndarray}-object along with utility functions to
+NumPy consists of the \code{ndarray}-object along with utility functions that
 operate on it.
@@
 the {\it de facto} exchange format for array data in Python.
-NumPy, while having been wildly successful as a library and array exchange
-project, has now also become a standard language for expressing array and
+NumPy, having been wildly successful as a library and array exchange
+project, now also has become a standard language for expressing array and
 tensor computations.
@@
 {\it Application Programming Interface} (API) has become ubiquitous, and is
-implemented by, for example, popular tensor computation libraries.
+implemented by, among others, popular tensor computation libraries\cite{pytorch}.
 
@@
 By the beginning of the millennium, several libraries built on top of Numeric 
-the provided fundamental scientific routines `were combined to form SciPy.
+the provided fundamental scientific routines were combined to form SciPy.
 Due to the poor performance of Numeric on large data sets, NumArray was
@@
 Travis Oliphant, started a major rewrite that aimed to be a ``best of both
-worlds'' unification of the two projects.  This became known as NumPy.
+worlds'' unification of the two projects.  %This became known as NumPy.
+This merger of Numeric and NumArray was called NumPy.
 
@@
 on setting in place standards and architecture to encourage more sustainable development.
+% http://doi.org/10.5281/zenodo.3585761
+% http://doi.org/10.5281/zenodo.3585767
 

### 2019-12-20 paper.tex
Revise intro
Add SJvdW to NumPy grant funding
Updates to abstract and intro

@@
   learning tool among enterprises \cite{451report2018}.
+  NumPy continues to evolve to meet the needs of its community, to
+  remain relevant in an ever-changing computing landscape, and to be a
+  fundamental pillar of support for the next decade of data science in
+  Python.
   In this article, we provide an overview of the capabilities and development
-  practices of the NumPy library and highlight some recent technical
+  practices of the library and highlight some recent technical
   developments.
@@
 Grant 2013-10-27 to the University of California, Berkeley.
-S.B., T.J.R., and W.W. were funded in part by the Gordon and Betty Moore 
-Foundation through Grant GBMF5447 and by the Alfred P. Sloan Foundation through
-Grant G-2017-9960 to the University of California, Berkeley.
+S.J.v.d.W, S.B., T.J.R., and W.W. were funded in part by the Gordon
+and Betty Moore Foundation through Grant GBMF5447 and by the Alfred
+P. Sloan Foundation through Grant G-2017-9960 to the University of
+California, Berkeley.
 
@@
 S.J.v.d.W and K.J.M. composed the manuscript with input from others.
+S.B., K.S., T.J.R., and W.W. drafted sections.
 All authors have contributed significant code, documentation, and/or expertise

### 2019-12-20 body.tex
Proofread manuscript

@@
 
-NumPy is the fundamental array computation library for the Python
-ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-Since its first stable release (1.0) in October 2006 it has unified
+%NumPy is the fundamental array computation library for the Python
+%ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
+Since its first stable release in October 2006, NumPy has unified
 the scientific Python community and now underpins almost every library
@@
 Not surprisingly, many important scientific research projects and
-results rely on NumPy.
+results crucially rely on NumPy.
 It plays an essential role in many research analysis pipelines in
@@
 
-...
 %[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
@@
 
-We showed that NumPy is actively developed according to a community
-vetted plan that integrates formal enhancement proposals.  Next, we
-give a more in-depth overview of the team developing NumPy, their
-activities, as well as the governance and community structures that
-guide the collaboration.
+%We showed that NumPy is actively developed according to a community
+%vetted plan that integrates formal enhancement proposals.  Next, we
+%give a more in-depth overview of the team developing NumPy, their
+%activities, as well as the governance and community structures that
+%guide the collaboration.
 
@@
 NumPy is currently maintained by a group of 23 contributors with commit rights
-to the NumPy code base. Out of these 17 maintainers were active in 2019.
+to the NumPy code base. Out of these 17 maintainers were active in 2019 and 4
+of those were paid to work on the project full-time.
 Additionally, there are a few long term developers who contributed and maintain
@@
 At a release cycle of about every half year, the five recent releases in the years
-2018 and 2019 have averaged about 450~pull requests~(PR) each.\footnote{
+2018 and 2019 have averaged about 450~pull requests~(PRs) each.\footnote{
     Note that before mid 2011, NumPy development did not happen on \url{github.com}.
@@
     \caption{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012. The total number of PR is indicated with the
+        quarter since 2012. The total number of PRs is indicated with the
         lower blue area showing the portion contributed by current or previous
@@
 The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
-PR merged into the NumPy master branch.
-In this plot PR opened by current and old maintainers are indicated.
+PRs merged into the NumPy master branch.
+In this plot PRs opened by current and old maintainers are indicated.
 Although the number of PRs being merged fluctuates,
@@
 years.
-Over the course of its history, NumPy has attracted PR by 823 contributors.
-However, its development still relies heavily on a relatively small number
+Over the course of its history, NumPy has attracted PRs by 823 contributors.
+However, its development relies heavily on a relatively small number
 of active maintainers, who share more than half of the contributions among
@@
 % https://github.com/numpy/numpy/pull/6352
+To formalize its unwritten development policy,
 NumPy adopted an official Governance Document on October~5,
@@
 
-The NumPy array data structure and universal functions, which we describe
-below, make up the core of NumPy.
-The NumPy library also includes
-a set of library functions for manipulating arrays and doing scientific computation,
-libraries for unit tests and Python package building,
-and the program \code{f2py}\cite{peterson2009f2py} for wrapping Fortran code in Python.
+The NumPy library consists of core functionality (e.g., the NumPy array data structure),
+functions for manipulating arrays and doing scientific computation, and infrastructure
+libraries.
 
@@
 
-The \code{ndarray} data structure stores regularly strided homogeneous data
-types in a contiguous block memory, allowing for the efficient representation
-of $n$-dimensional data.
-A \emph{masked array} class, built on top of the \code{ndarray}, is
-provided as well. This class allows elements of an array to be masked,
-indicating that the data is missing or unknown.
+The \code{ndarray} data structure stores regularly spaced elements of a single
+data type in a contiguous block of memory; along with meta-data such as shape,
+it allows for the efficient representation of $n$-dimensional arrays.
 More details about the data structure are given in ``The NumPy array:
@@
 implementation of \emph{broadcasting}.  For example, the function
-\code{arctan2(x, y)} is a ufunc that accepts two values and computes
-$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
-the ufunc will take care of looping over the dimensions of the inputs
+\code{arctan2(x, y)} is a ufunc that accepts two arguments and computes
+$\tan^{-1}(y/x)$.  The ufunc loops over the dimensions of the inputs
 in such a way that if, say, \code{x} is a 1-d array with length 3, and
@@
 of calling the function with all the appropriate combinations of
-input array elements to complete the output array.
+input array elements to compute the output array.
 The elementary arithmetic operations of addition, multiplication, etc.,
@@
 
-NumPy provides a large library of functions for array manipulation
+NumPy provides a large collection of functions for array manipulation
 including functions for: creating, reshaping, concatenating, and padding arrays;
@@
 variance, and standard deviation; file I/O; and more.
+It also provides extensive support for generating pseudorandom numbers
+as well as an assortment of probability distributions.
 For historical reasons, NumPy also includes basic functionality for
-linear algebra,
-fast Fourier transforms and windowing,
+linear algebra, fast Fourier transforms and windowing,
 and polynomial fitting.
-We will keep supporting these, but we will not expand beyond them.
+We will continue supporting these, but we will not expand beyond them.
 
@@
 
-Being at the base of the scientific Python stack, NumPy provides some basic
-infrastructure for other packages in the scientific Python ecosystem.
+NumPy also provides basic infrastructure for other packages in the scientific
+Python ecosystem.
 The \code{numpy.testing} subpackage provides functions such as
@@
 
-NumPy has, throughout its lifetime, seen continuous development.
-E.g., bugs fixes, minor improvements, cleanups, and releases have
-happened regularly.  However, with the NEP process in place we now
-have the flexibility to plan and implement changes of larger scope.
-We highlight two of those below, as well as changes made to our
-testing infrastructure to support hardware platforms used in large
-scale computing.
+The NumPy codebase has been actively developed for over two decades (first as
+Numeric and NumArray).
+Given the complexity of the codebase and the massive number of projects depending
+on it, large changes require careful planning and substantial work.
+Moreover, with the now prevalent need for distributed computing and the
+emergence of specialized hardware such as GPUs, many new usecases have
+arisen, which NumPy was not originally designed for.
+With the recent infusion of funding and the newly formalized NEP process in
+place we have been able to tackle a number of important large scale changes.
+We highlight two of those below, as well as changes made to our testing
+infrastructure to support hardware platforms used in large scale computing.
 
@@
 
-The central position of NumPy takes within the numerical and scientific
-Python ecosystem means that a vast number of projects are built on it.
-These projects are consumers of the NumPy API.
-
-On the other side, projects are providers of a \emph{NumPy-like API} and array
-objects targeting audiences with specialized needs beyond NumPy's capabilities.
-For example, the NumPy API is implemented by
-several popular tensor computation libraries including
-CuPy\footnote{\url{https://cupy.chainer.org/}},
+A vast number of projects are built on NumPy;
+these projects are consumers of the NumPy API.
+%The central position of NumPy takes within the numerical and scientific
+%Python ecosystem means that a vast number of projects are built on it.
+%These projects are consumers of the NumPy API.
+Over the last several years, a growing number of projects are providers of
+a \emph{NumPy-like API} and array objects targeting audiences with specialized
+needs beyond NumPy's capabilities.
+For example, the NumPy API is implemented by several popular tensor computation
+libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
 Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
 PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
-Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
-and to a lesser extent
-TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.
+and Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}}.
 It is also implemented in packages that support sparse arrays
@@
 all of the above array providers.
-This is a burden for users relying on the specialized Array-like, since
+This is a burden for users relying on the specialized array-like, since
 a tool they need may not work for them.
-Additionally, a user who has the need of transitioning from NumPy to a more
-specialized array, is faced with a challenge.
-These issue can lead into a fragmentation of the user bases into consumers
-of different NumPy-like APIs.
+It also creates new challenges for end-users who need to transition
+from NumPy to a more specialized array.
+The growing multitude of specialized projects with NumPy-like APIs threatened
+to again fracture the scientific Python community.
 
@@
 \code{\_\_array\_function\_\_} protocol into NumPy~1.17.
-These two protocols allows providers of Array-like APIs to be interoperable
+These two protocols allows providers of array-like APIs to be interoperable
 with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
@@
 from numpy.random import Generator, PCG64, SeedSequence
+
 seq = SeedSequence(10304245474441179933310169597522947680)
@@
 same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
-numba. Both the bit generators and the low-level functions can also be used in C
+Numba. Both the bit generators and the low-level functions can also be used in C
 or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
@@
 Most of the technical work for the Moore and Sloan grant is now complete.
-The main remaining item is an overhaul of the data type system.
-
+The main remaining item is an overhaul of the data type system, which
+we are currently working on \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
 The meaning of each element within a NumPy array is described by its
@@
 datatypes, strings, datetimes, and generic Python objects, each of
-which is identified by a dtype.  We are busy overhauling the datatype
-system to make it behave consistently and to simplify the creation of
-custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
-of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
+which is identified by a dtype.
+%We are busy overhauling the datatype
+%system to make it behave consistently and to simplify the creation of
+%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
+%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
 While numerical dtypes make the foundation for most users,
@@
 %limitations.  % TODO: I may need citations, we could link github issues.
+The proposed overhaul of the datatype system will make it behave consistently and
+to simplify the creation of custom dtypes, both in C and in Python
 
@@
 
-Now that we have made good progress on addressing technical debt, the project
+Having made progress on addressing technical debt accrued over two decades of
+development and modernizing the code to handle new use cases, the project
 is ready to grow its community of contributors and to better address the needs

### 2019-12-20 paper.tex
Proofread manuscript

@@
   basicstyle=\ttfamily\footnotesize,  % size of fonts used for the code
+  xleftmargin=2\parindent,
   breaklines=true,   % automatic line breaking only at whitespace
@@
 \begin{abstract}
-  NumPy is the fundamental array computing library for Python.
+
+  NumPy is the fundamental array computation library for the Python ecosystem
+  \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
   It has been actively developed for over two decades and has
@@
   Python.
-  In this article, we provide an overview of the capabilities and development
-  practices of the library and highlight some recent technical
-  developments.
+  In this article, we provide an overview of the project and library
+  organization,  highlight some recent technical developments, and
+  outline are future plans.
 \end{abstract}

### 2019-12-20 body.tex
Add LSST reference

@@
 gravitational waves\cite{pycbc}, imaging black
-holes\cite{eht-imaging}, and exploring the solar system (... Kepler? https://ui.adsabs.harvard.edu/abs/2012ascl.soft08004S/abstract).
+holes\cite{eht-imaging}, and exploring the solar system \cite{jenness2018lsst}.
 

### 2019-12-20 body.tex
Add figure to illustrate broadcasting

@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=0.25\textwidth]{static/broadcasting}
+  \caption{
+    Broadcasting a $1 \times 3$ array against a $2 \times 1$ array
+    yields a $2 \times 3$ result.
+  }
+  \label{fig:broadcasting}
+\end{figure}
+
 The \emph{universal functions} (or \emph{ufuncs})
@@
 \code{y} is a 2-d array with shape $2 \times 1$, the output will be
-an array with shape $2 \times 3$.  The ufunc machinery takes care
+an array with shape $2 \times 3$ (see Figure~\ref{fig:broadcasting}).  The ufunc machinery takes care
 of calling the function with all the appropriate combinations of

### 2019-12-21 body.tex
Polishing intro sections

@@
 %ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-Since its first stable release in October 2006, NumPy has unified
+With its first stable release in October 2006, NumPy unified
 the scientific Python community and now underpins almost every library
@@
 In particular, NumPy played a fundamental role in discovering
-gravitational waves\cite{pycbc}, imaging black
-holes\cite{eht-imaging}, and exploring the solar system \cite{jenness2018lsst}.
+gravitational waves\cite{pycbc} and the first imaging of a black
+hole\cite{eht-imaging}, and lets us explore our
+universe\cite{jenness2018lsst}.
 
@@
 the {\it de facto} exchange format for array data in Python.
-NumPy, having been wildly successful as a library and array exchange
-project, now also has become a standard language for expressing array and
-tensor computations.
-It has such widespread adoption that not only the array object but also its
-{\it Application Programming Interface} (API) has become ubiquitous, and is
-implemented by, among others, popular tensor computation libraries\cite{pytorch}.
+The library has such widespread adoption that not only the array object but also its
+{\it Application Programming Interface} (API) has become ubiquitous as
+a language for tensor computation---witnessed by its use in popular
+deep learning libraries such as PyTorch\cite{pytorch}.
 
@@
 By the beginning of the millennium, several libraries built on top of Numeric 
-the provided fundamental scientific routines were combined to form SciPy.
+to provide fundamental scientific routines were combined into SciPy.
 Due to the poor performance of Numeric on large data sets, NumArray was
@@
 large astronomical images such as those produced by the Hubble Space Telescope.
-It was initially considered the successor to Numeric.
-However, it performed poorly on small arrays.
+However, while initially considered the successor to Numeric, NumArray was no longer
+performant on multiple small arrays.
 This divide posed a challenge for SciPy as it wasn't clear how to support
 both libraries.
-Working with the NumArray and SciPy community, the maintainer of Numeric,
-Travis Oliphant, started a major rewrite that aimed to be a ``best of both
-worlds'' unification of the two projects.  %This became known as NumPy.
-This merger of Numeric and NumArray was called NumPy.
+Working with the NumArray and SciPy communities, the then maintainer of Numeric,
+Travis Oliphant, embarked upon a major rewrite that aimed to be a ``best of both
+worlds'' unification of Numeric and NumArray. This merger resulted in NumPy.
 
@@
 To coordinate work as a global community, to ensure accuracy of computed
-results, and to allow modification without breakage, the NumPy (and SciPy) community
+results, and to allow modification without breakage, the NumPy (and SciPy) teams
 pioneered best software engineering practices in the scientific Python
-community  \cite{millman2014developing}.
+community\cite{millman2014developing}.
 For example, NumPy early on adopted distributed revision control, unit testing,
-continuous integration, code review, and standardized executable documentation.
+continuous integration, code review, and executable documentation.
 It also developed a standard for ``docstrings''—text describing each NumPy function.
-This quickly became the standard for the larger scientific Python community.
+This quickly became a documentation standard for the larger scientific Python community.
 
-As the scientific Python community grew and new challenges arose, Travis Oliphant
-eventually stepped back from leading NumPy to found a company and start
-new projects including Numba---an accelerated Python compiler that could vastly
-accelerate NumPy expressions.
-Development of NumPy continued by a core group that informally operated by consensus.
+Travis Oliphant eventually stepped back from leading NumPy to found a
+company and to start new projects including Numba---an accelerated Python
+compiler that could vastly accelerate NumPy expressions.  Development
+of NumPy continued by a core group that informally operated by
+consensus.
+
+% While the scientific Python community grew and new challenges arose,
 
@@
 
-For many years, NumPy had no dedicated funding, development
-time was mostly contributed freely by students and researchers in
-their spare time.  In-person meetings were sponsored off of
-individual's research grants, and while industry sponsored
-contributions were made, notably those by Mark Wiebe when he was
-sponsored by Travis Oliphant at Continuum Analytics, mostly the
-package had rapid adoption and became central to scientific computing
-without external investment.
+For many years, NumPy had no dedicated funding, and development time
+was mostly contributed freely by students and researchers in their
+spare time.  In-person meetings were sponsored off of individual's
+research grants, and while industry sponsored contributions were made,
+notably those by Mark Wiebe when he was sponsored by Travis Oliphant
+at Continuum Analytics, the package's rapid adoption in scientific
+computing took place without external investment.
 
@@
 NumPy received another grant for \$195,000 from the Chan Zuckerberg
-Initiative at the end of 2019.
-Ralf Gommers is the PI.
+Initiative at the end of 2019 with Ralf Gommers as the PI.
 This grant, building on the technical work supported by the previous
-grant, will focus on better serving NumPy's large number of beginner
+grant, will focus on better serving NumPy's large number of beginning
 to intermediate level users and on growing the community of NumPy
 contributors.
-It also will provide some support for OpenBLAS (on which NumPy depends for
-accelerated linear algebra).
+It will also provide support to OpenBLAS, on which NumPy depends for
+accelerated linear algebra.
 
-The project also receives around 10K USD per year from Tidelift, which is
-used to fund documentation and website developers.
+Finally, the project receives around 10K USD annually from Tidelift, which is
+typically used to fund documentation writers and website developers.
 
@@
 
-The weekly community calls alternate weekly between triage meetings and
+Weekly community calls alternate between triage and
 higher level discussion.  The calls not only involve developers from
@@
 NumPy is currently maintained by a group of 23 contributors with commit rights
-to the NumPy code base. Out of these 17 maintainers were active in 2019 and 4
-of those were paid to work on the project full-time.
+to the NumPy code base. Out of these, 17 maintainers were active in
+2019, 4 of whom were paid to work on the project full-time.
 Additionally, there are a few long term developers who contributed and maintain
@@
 At a release cycle of about every half year, the five recent releases in the years
-2018 and 2019 have averaged about 450~pull requests~(PRs) each.\footnote{
+2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
     Note that before mid 2011, NumPy development did not happen on \url{github.com}.
-    All data provided here is based on the development which happened through github
+    All data provided here is based on the development which happened through GitHub
     pull requests. In some cases contributions by maintainers may not be categorized as such.}
 % Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
-With each release attracting more than a hundred new contributors.
+with each release attracting more than a hundred new contributors.
 
@@
 \end{figure}
+
+% -- The paragraphs above and below can be tightened up a bit still --
 

### 2019-12-21 paper.tex
Fix typo

@@
   organization,  highlight some recent technical developments, and
-  outline are future plans.
+  outline our future plans.
 \end{abstract}

### 2019-12-24 body.tex
Use section numbers
Restructure based on feedback from Ralf
This is just a start.  We still need to cleanup
the prose.

@@
-\section*{Introduction}
+\section{Introduction}
 
@@
 
-\subsection*{History}
+\subsection{History}
 
@@
 
-\subsection*{Funding}
+\subsection{Funding}
 
@@
 
-\subsection*{Current status}
+\section{Organization and infrastructure}
 
-% (maturity, users)
-
-As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
-of enhancement proposals---used infrequently and informally until 2017 when,
-to facilitate work expected from funded development, he formalized the
-NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
-Python Enhancement Proposals (PEPs) for ``proposing major new
-features, for collecting community input on an issue, and for
-documenting the design decisions that have gone into
-Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
-Since then there have been 19 proposed NEPS---6 have been implemented,
-4 have been accepted and are being implemented, 4 are under
-consideration, 3 have been deferred or superseded, and 2 have been rejected
-or withdrawn.
-
-The first activities organized for the new grant were a NEP development
-sprint, a weekly community call, and a meeting to write a draft roadmap.
-
-Since the roadmap was meant to represent community consensus on future
-development, it was presented for comment at the annual SciPy
-conference during a so-called ``Birds of a Feather'' session, attended
-by more than 100 people.  One additional meeting was held to discuss
-specific enhancements, such as the development of a new data-type
-system.  The final roadmap was vetted by the community via online
-discussion before being accepted.
-
-Weekly community calls alternate between triage and
-higher level discussion.  The calls not only involve developers from
-the community, but provide a venue for vendors and other external
-groups to provide input.  For example, after Intel produced a forked
-version of NumPy, one of their developers joined a call to discuss
-community concerns.
-
-NumPy plays a central role in building and standardizing much of the scientific
-Python community infrastructure.
-The docstring standard mentioned in the history section is now widely adopted.
-We are also now using the NEP system as a way to help coordinate the larger
-scientific Python community.
-% https://numpy.org/neps/nep-0029-deprecation_policy.html
-For example, in NEP 29, we recommend, along with leaders from various other
-projects, that all projects across the Scientific Python ecosystem adopt a
-common ``time window-based'' policy for support of Python and NumPy versions.
-This standard will simplify downstream project and release planning.
-
-% https://numpy.org/neps/scope.html
-% https://numpy.org/neps/roadmap.html
-
-%We showed that NumPy is actively developed according to a community
-%vetted plan that integrates formal enhancement proposals.  Next, we
-%give a more in-depth overview of the team developing NumPy, their
-%activities, as well as the governance and community structures that
-%guide the collaboration.
-
-\section*{Project organization and community}
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
+    \caption{Number of pull requests merged into the NumPy master branch for each
+        quarter since 2012. The total number of PRs is indicated with the
+        lower blue area showing the portion contributed by current or previous
+        maintainers.}\label{fig:prs-over-time}
+\end{figure}
 
@@
 with each release attracting more than a hundred new contributors.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
-    \caption{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012. The total number of PRs is indicated with the
-        lower blue area showing the portion contributed by current or previous
-        maintainers.}\label{fig:prs-over-time}
-\end{figure}
-
-% -- The paragraphs above and below can be tightened up a bit still --
-
-The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
-PRs merged into the NumPy master branch.
-In this plot PRs opened by current and old maintainers are indicated.
+Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
+master branch.
 Although the number of PRs being merged fluctuates,
@@
 years.
-Over the course of its history, NumPy has attracted PRs by 823 contributors.
-However, its development relies heavily on a relatively small number
-of active maintainers, who share more than half of the contributions among
-themselves.
-
-% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
-% https://github.com/numpy/numpy/pull/6352
-To formalize its unwritten development policy,
-NumPy adopted an official Governance Document on October~5,
-2015\cite{NumPyProjectGovernance}.
-Project decisions are usually made by consensus of interested contributors.
-This means for many thing everyone is entrusted with veto power.
-A Steering Council, currently composed of 12~members, facilitates this
-process and oversees daily development of the project by contributing code
-and reviewing contributions from the community.
-
-% https://mail.python.org/pipermail/numpy-discussion/2018-July/078476.html
-% https://github.com/numpy/numpy/pull/11865
-NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
-In brief, we strive to:
-\emph{be open};
-\emph{be empathetic, welcoming, friendly, and patient};
-\emph{be collaborative};
-\emph{be inquisitive}; and
-\emph{be careful in the words that we choose}.
-The Code of Conduct also specifies how breaches can be reported and outlines
-the process for responding to such reports.
-
-\section*{Library organization}
 
@@
 
-\subsection*{Core}
+\subsection{Core functionality}
 
@@
 
-\subsection*{Computing libraries}
+\subsection{Computing libraries}
 
@@
 
-\subsection*{Infrastructure libraries}
+\subsection{Infrastructure libraries}
 
@@
 
-\section*{Recent technical improvements}
+\section{Recent technical improvements}
 
@@
 on it, large changes require careful planning and substantial work.
-Moreover, with the now prevalent need for distributed computing and the
-emergence of specialized hardware such as GPUs, many new usecases have
-arisen, which NumPy was not originally designed for.
+As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
+of enhancement proposals---used infrequently and informally until 2017 when,
+to facilitate work expected from funded development, he formalized the
+NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
+Python Enhancement Proposals (PEPs) for ``proposing major new
+features, for collecting community input on an issue, and for
+documenting the design decisions that have gone into
+Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
+%Since then there have been 19 proposed NEPS---6 have been implemented,
+%4 have been accepted and are being implemented, 4 are under
+%consideration, 3 have been deferred or superseded, and 2 have been rejected
+%or withdrawn.
+
+The first activities organized for the new grant were a NEP development
+sprint, a weekly community call, and a meeting to write a draft roadmap.
+Since the roadmap was meant to represent community consensus on future
+development, it was presented for comment at the annual SciPy
+conference during a so-called ``Birds of a Feather'' session, attended
+by more than 100 people.  One additional meeting was held to discuss
+specific enhancements, such as the development of a new data-type
+system.  The final roadmap was vetted by the community via online
+discussion before being accepted.
+
+%Weekly community calls alternate between triage and
+%higher level discussion.  The calls not only involve developers from
+%the community, but provide a venue for vendors and other external
+%groups to provide input.  For example, after Intel produced a forked
+%version of NumPy, one of their developers joined a call to discuss
+%community concerns.
+
+%NumPy plays a central role in building and standardizing much of the scientific
+%Python community infrastructure.
+%The docstring standard mentioned in the history section is now widely adopted.
+%We are also now using the NEP system as a way to help coordinate the larger
+%scientific Python community.
+%% https://numpy.org/neps/nep-0029-deprecation_policy.html
+%For example, in NEP 29, we recommend, along with leaders from various other
+%projects, that all projects across the Scientific Python ecosystem adopt a
+%common ``time window-based'' policy for support of Python and NumPy versions.
+%This standard will simplify downstream project and release planning.
+
+% https://numpy.org/neps/scope.html
+% https://numpy.org/neps/roadmap.html
+
+%We showed that NumPy is actively developed according to a community
+%vetted plan that integrates formal enhancement proposals.  Next, we
+%give a more in-depth overview of the team developing NumPy, their
+%activities, as well as the governance and community structures that
+%guide the collaboration.
+
+%Moreover, with the now prevalent need for distributed computing and the
+%emergence of specialized hardware such as GPUs, many new usecases have
+%arisen, which NumPy was not originally designed for.
 With the recent infusion of funding and the newly formalized NEP process in
@@
 
-\subsection*{The Array Function Protocol}
+\subsection{The Array Function Protocol}
 
@@
 
-\subsection*{Random}
+\subsection{Random}
 
@@
 
-\subsection*{Testing on multiple architectures}
+\subsection{Testing on multiple architectures}
 
@@
 
-\section*{Discussion}
+\section{Discussion}
 

### 2020-01-03 authors.tex
Rework history, funding, and current status sections (w/ Jarrod Millman)
Restructure second section

@@
 \author[1]{Charles R Harris}
-\author[2]{Travis E. Oliphant}
-\author[3,*]{St\'efan J. van der Walt}
-\author[3]{Warren Weckesser}
-\author[3,4,*]{K. Jarrod Millman}
+\author[2,*]{St\'efan J. van der Walt}
+\author[3,*]{Ralf Gommers}
+\author[2,4,*]{K. Jarrod Millman}
+\author[2]{Sebastian Berg}
 \author[5]{Kevin Sheppard}
+\author[6]{Tyler J. Reddy}
+\author[2]{Warren Weckesser}
+\author[3]{Travis E. Oliphant}
 \affil[1]{Independent Researcher}
-\affil[2]{Quansight LLC, Austin, TX, USA}
-\affil[3]{Berkeley Institute for Data Science, University of California, Berkeley, CA 94720, USA}
+\affil[2]{Berkeley Institute for Data Science, University of California, Berkeley, CA 94720, USA}
+\affil[3]{Quansight LLC, Austin, TX, USA}
 \affil[4]{Division of Biostatistics, University of California, Berkeley, CA 94720, USA}
+\affil[5]{Department of Economics, University of Oxford, Oxford, OX1 3UQ, UK}
+\affil[6]{Los Alamos National Laboratory, Los Alamos, NM 87544, USA}
 \affil[*]{numpy.articles@gmail.com}
-\affil[5]{kevin.sheppard@economics.ox.ac.uk, Department of Economics, University of Oxford, Oxford, OX1 3UQ, UK}

### 2020-01-03 body.tex
Restructure second section

@@
 
-\section{Organization and infrastructure}
+\section{Community and library}
 
@@
 
-The NumPy library consists of core functionality (e.g., the NumPy array data structure),
-functions for manipulating arrays and doing scientific computation, and infrastructure
-libraries.
-
-\subsection{Core functionality}
+\subsection{Basic functionality}
 
@@
 it allows for the efficient representation of $n$-dimensional arrays.
+\fixme{discuss Figure~\ref{fig:basic-functionality} .. vectors / matrices / and then more ...}
 More details about the data structure are given in ``The NumPy array:
@@
 \centering
-  \includegraphics[width=0.25\textwidth]{static/broadcasting}
-  \caption{
-    Broadcasting a $1 \times 3$ array against a $2 \times 1$ array
-    yields a $2 \times 3$ result.
-  }
-  \label{fig:broadcasting}
+\begin{subfigure}{.5\textwidth}
+\begin{lstlisting}
+>>> import numpy as np
+>>> x = np.array([0, 1, 2, 3])
+>>> x.dtype
+dtype('int64')
+>>> x.shape
+(4,)
+
+>>> y = 2 * x
+>>> y
+array([0, 2, 4, 6])
+
+>>> x * y
+array([ 0,  2,  8, 18])
+
+>>> x @ y
+28
+\end{lstlisting}
+  \caption{Vectors (1-d arrays)}
+  \label{fig:sub1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+\begin{lstlisting}
+>>> import numpy as np
+>>> x = np.array([1, 2])
+>>> A = np.array([[0.1, 0.2], [0.3, 0.2]])
+>>> A.dtype 
+dtype('float64')
+>>> A.shape
+(2, 2)
+
+>>> A * x
+array([[0.1, 0.4],
+       [0.3, 0.4]])
+
+>>> A @ x
+array([0.5, 0.7])
+>>> x @ A
+array([0.7, 0.6])
+\end{lstlisting}
+  \caption{Vectors (1-d arrays) and Matrices (2-d arrays)}
+  \label{fig:sub2}
+\end{subfigure}
+\caption{Basic functionality}
+\label{fig:basic-functionality}
 \end{figure}
+
+NumPy provides a large collection of functions for array manipulation
+including functions for: creating, reshaping, concatenating, and padding arrays;
+searching, sorting and counting data
+in arrays; computing elementary statistics, such as the mean, median,
+variance, and standard deviation; file I/O; and more.
+It also provides extensive support for generating pseudorandom numbers
+as well as an assortment of probability distributions.
+For historical reasons, NumPy also includes basic functionality for
+linear algebra, fast Fourier transforms and windowing,
+and polynomial fitting.
+We will continue supporting these, but we will not expand beyond them.
+
+\subsection{Advanced functionality}
 
@@
 
-\subsection{Computing libraries}
-
-NumPy provides a large collection of functions for array manipulation
-including functions for: creating, reshaping, concatenating, and padding arrays;
-searching, sorting and counting data
-in arrays; computing elementary statistics, such as the mean, median,
-variance, and standard deviation; file I/O; and more.
-It also provides extensive support for generating pseudorandom numbers
-as well as an assortment of probability distributions.
-For historical reasons, NumPy also includes basic functionality for
-linear algebra, fast Fourier transforms and windowing,
-and polynomial fitting.
-We will continue supporting these, but we will not expand beyond them.
-
-\subsection{Infrastructure libraries}
+\begin{figure}
+  \centering
+  \includegraphics[width=0.25\textwidth]{static/broadcasting}
+  \caption{
+    Broadcasting a $1 \times 3$ array against a $2 \times 1$ array
+    yields a $2 \times 3$ result.
+  }
+  \label{fig:broadcasting}
+\end{figure}
 

### 2020-01-03 body.tex
Remove testing section

@@
 place we have been able to tackle a number of important large scale changes.
-We highlight two of those below, as well as changes made to our testing
-infrastructure to support hardware platforms used in large scale computing.
+We highlight two of those below.
 
@@
 
-\subsection{Testing on multiple architectures}
-
-At the time of writing the two fastest supercomputers in the
-world, Summit and Sierra, both have IBM POWER9 architectures
-\cite{top500nov2019}. In late 2018, Astra, the first ARM-based
-supercomputer to enter the TOP500 list, went into production\cite{
-astra-wiki}. Furthermore, over 100 billion ARM processors have been
-produced as of 2017\cite{arm-architecture}, making it the most 
-widely used instruction set architecture in the world.
-
-Clearly there are motivations for a large scientific computing
-software library to support POWER and ARM architectures. We've extended
-our continuous integration (CI) testing to include \texttt{ppc64le}
-(POWER8 on Travis CI) and ARMv8 (on Shippable service). We also test
-with the s390x architecture (IBM Z CPUs on Travis CI) so that we
-can probe the behavior of our library on a big-endian machine.
-This satisfies one of the major components of
-improved CI testing laid out in a version of our roadmap
-\cite{numpy-roadmap}---specifically, ``CI for more exotic
-platforms."
-
-PEP 599\cite{PEP599} lays out a plan for new Python binary wheel
-distribution support, \texttt{manylinux2014}, that adds
-support for a number of architectures supported by the CentOS
-Alternative Architecture Special Interest Group, including
-ARMv8, ppc64le, as well as s390x. We are thus well-positioned
-for a future where provision of binaries on these architectures
-will be expected for a library at the base of the ecosystem.
-
 \section{Discussion}

### 2020-01-04 paper.tex
Remove template stuff from SciPy paper
Restructure second section
Add abstract word count

@@
 }
+
+\usepackage{subcaption}
 
@@
 \begin{abstract}
-
-  NumPy is the fundamental array computation library for the Python ecosystem
-  \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-  It has been actively developed for over two decades and has
-  more than 850 contributors, nearly 300,000 dependent repositories
-  on GitHub, and millions of downloads per year.
-  It is extensively used in research and teaching.
-  According to a 2018 survey, NumPy is the second most commonly used machine
-  learning tool among enterprises \cite{451report2018}.
-  NumPy continues to evolve to meet the needs of its community, to
-  remain relevant in an ever-changing computing landscape, and to be a
-  fundamental pillar of support for the next decade of data science in
-  Python.
-  In this article, we provide an overview of the project and library
-  organization,  highlight some recent technical developments, and
-  outline our future plans.
+\input{summary} 
 \end{abstract}
@@
 
-% Count of words
+% Count of words in summary
+\verbatiminput{summarycount.tex}
+
+% Count of words in text
 \verbatiminput{wordcount.tex}
@@
 
-\section*{Consortium}
-\subsection*{NumPy Contributors}
-\input{consortium}
-
 \end{document}

### 2020-01-04 summary.tex
Add abstract word count

% https://www.nature.com/nature/for-authors/formatting-guide

% Articles start with a fully referenced summary paragraph, ideally of no more
% than 200 words, which is separate from the main text and avoids numbers,
% abbreviations, acronyms or measurements unless essential. It is aimed at
% readers outside the discipline.

% This summary paragraph should be structured as follows:
% - 2-3 sentences of basic-level introduction to the field;
% - a brief account of the background and rationale of the work;
% - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
% - 2-3 sentences putting the main findings into general context so it is
%   clear how the results described in the paper have moved the field forwards.

% Please refer to our annotated example  to see how the summary paragraph should
% be constructed. https://www.nature.com/documents/nature-summary-paragraph.pdf

NumPy is the fundamental array computation library for the Python ecosystem
\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
It has been actively developed for over two decades and has
more than 850 contributors, nearly 300,000 dependent repositories
on GitHub, and millions of downloads per year.
It is extensively used in research and teaching.
According to a 2018 survey, NumPy is the second most commonly used machine
learning tool among enterprises \cite{451report2018}.
NumPy continues to evolve to meet the needs of its community, to
remain relevant in an ever-changing computing landscape, and to be a
fundamental pillar of support for the next decade of data science in
Python.
In this article, we provide an overview of the project and library
organization,  highlight some recent technical developments, and
outline our future plans.

### 2020-01-04 body.tex
Rework intro / summary

@@
 %ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-With its first stable release in October 2006, NumPy unified
-the scientific Python community and now underpins almost every library
-that does numerical computation, including SciPy\cite{virtanen2019scipy},
-pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
-and scikit-image\cite{vanderwalt2014scikit}.
-Not surprisingly, many important scientific research projects and
-results crucially rely on NumPy.
-It plays an essential role in many research analysis pipelines in
-fields as diverse as physics, biology, astronomy, neuroscience,
-material science, engineering, and chemistry.
-% w/ citations
-In particular, NumPy played a fundamental role in discovering
-gravitational waves\cite{pycbc} and the first imaging of a black
-hole\cite{eht-imaging}, and lets us explore our
-universe\cite{jenness2018lsst}.
+%With its first stable release in October 2006, NumPy unified
+%the scientific Python community and now underpins almost every library
+%that does numerical computation, including SciPy\cite{virtanen2019scipy},
+%pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
+%and scikit-image\cite{vanderwalt2014scikit}.
+%Not surprisingly, many important scientific research projects and
+%results crucially rely on NumPy.
+%It plays an essential role in many research analysis pipelines in
+%fields as diverse as physics, biology, astronomy, neuroscience,
+%material science, engineering, and chemistry.
+%% w/ citations
+%In particular, NumPy played a fundamental role in discovering
+%gravitational waves\cite{pycbc} and the first imaging of a black
+%hole\cite{eht-imaging}, and lets us explore our
+%universe\cite{jenness2018lsst}.
 
@@
 
+With its first stable release in October 2006, NumPy unified
+the scientific Python community and now underpins almost every library
+that does numerical computation, including SciPy\cite{virtanen2019scipy},
+pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
+and scikit-image\cite{vanderwalt2014scikit}.
 NumPy consists of the \code{ndarray}-object along with utility functions that
@@
 it allows for the efficient representation of $n$-dimensional arrays.
-\fixme{discuss Figure~\ref{fig:basic-functionality} .. vectors / matrices / and then more ...}
+
+\fixme{discuss Figure~\ref{fig:basic-functionality} .. vectors / matrices / and then more ...
+indexing and slicing}
+
 More details about the data structure are given in ``The NumPy array:

### 2020-01-04 summary.tex
Rework intro / summary

@@
 
+% - 2-3 sentences of basic-level introduction to the field;
+Software libraries for array computation provide powerful notation
+for accessing, manipulating, and operating on aggregate data \cite{iverson1980notation}.
+% - a brief account of the background and rationale of the work;
 NumPy is the fundamental array computation library for the Python ecosystem
 \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-It has been actively developed for over two decades and has
-more than 850 contributors, nearly 300,000 dependent repositories
-on GitHub, and millions of downloads per year.
-It is extensively used in research and teaching.
-According to a 2018 survey, NumPy is the second most commonly used machine
-learning tool among enterprises \cite{451report2018}.
-NumPy continues to evolve to meet the needs of its community, to
-remain relevant in an ever-changing computing landscape, and to be a
-fundamental pillar of support for the next decade of data science in
-Python.
-In this article, we provide an overview of the project and library
-organization,  highlight some recent technical developments, and
-outline our future plans.
+Not surprisingly, many important scientific research projects and results
+crucially rely on NumPy.
+It plays an essential role in many research analysis pipelines in fields as
+diverse as physics, biology, astronomy, neuroscience, material science,
+engineering, and chemistry.
+In particular, NumPy played a fundamental role in discovering gravitational
+waves\cite{pycbc} and the first imaging of a black hole\cite{eht-imaging}, and
+lets us explore our universe\cite{jenness2018lsst}.
+It is one of the most commonly used machine learning tools among enterprises
+\cite{451report2018}.
+% - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
+Here we provide an overview of the project and library organization,  highlight
+some recent technical developments, and outline our future plans.
+% - 2-3 sentences putting the main findings into general context so it is
+%   clear how the results described in the paper have moved the field forwards.
+NumPy continues to evolve to meet the needs of its community, to remain
+relevant in an ever-changing computing landscape, and to be a fundamental
+pillar of support for the next decade of data science in Python.

### 2020-01-04 summary.tex
Minor edits

@@
 Software libraries for array computation provide powerful notation
-for accessing, manipulating, and operating on aggregate data \cite{iverson1980notation}.
+for accessing, manipulating, and operating on aggregate data in vectors,
+matrices, or higher-dimensional arrays \cite{iverson1980notation}.
 % - a brief account of the background and rationale of the work;
@@
 lets us explore our universe\cite{jenness2018lsst}.
-It is one of the most commonly used machine learning tools among enterprises
-\cite{451report2018}.
+It is also one of the most commonly used machine learning tools among
+enterprises \cite{451report2018}.
 % - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);

### 2020-01-04 body.tex
Minor edits
First pass - to give an idea of direction.

@@
 
-\subsection{History}
+\subsection{Background}
 
-Before NumPy, two different Python array packages—Numeric and NumArray—existed.
-The first began in the mid-90s and was developed by Jim Fulton, Jim Hugunin,
-Paul Dubois, Konrad Hinsen, Travis Oliphant, and others, and was aimed at fast
-manipulation of many small arrays.
-By the beginning of the millennium, several libraries built on top of Numeric 
-to provide fundamental scientific routines were combined into SciPy.
-Due to the poor performance of Numeric on large data sets, NumArray was
-developed at the Space Telescope Science Institute (STScI) to handle
-large astronomical images such as those produced by the Hubble Space Telescope.
-However, while initially considered the successor to Numeric, NumArray was no longer
-performant on multiple small arrays.
-This divide posed a challenge for SciPy as it wasn't clear how to support
-both libraries.
-Working with the NumArray and SciPy communities, the then maintainer of Numeric,
-Travis Oliphant, embarked upon a major rewrite that aimed to be a ``best of both
-worlds'' unification of Numeric and NumArray. This merger resulted in NumPy.
+% https://books.google.com/books?id=QQJODwAAQBAJ&pg=PA101&lpg=PA101&dq=%E2%80%9Cthe+secret+to+successful+hiring+is+this:+look+for+the+people+who+want+to+change+the+world.%E2%80%9D&source=bl&ots=D6l8JXfB20&sig=ACfU3U0x1X9Gh8k6MlNV0BqhBa_73Q0yfQ&hl=en&sa=X&ved=2ahUKEwj48pqg_oHgAhVohuAKHYlJAsc4ChDoATAAegQIChAB#v=onepage&q=%E2%80%9Cthe%20secret%20to%20successful%20hiring%20is%20this%3A%20look%20for%20the%20people%20who%20want%20to%20change%20the%20world.%E2%80%9D&f=false
+% "The secret to successful hiring is this: look for the people who want to change the world.".  Marc Benioff.
 
-The creation of NumPy spurred renewed development in both SciPy and the larger
-scientific Python ecosystem and heralded the current era of wide-spread use of
-Python for scientific computing.
-To coordinate work as a global community, to ensure accuracy of computed
-results, and to allow modification without breakage, the NumPy (and SciPy) teams
-pioneered best software engineering practices in the scientific Python
-community\cite{millman2014developing}.
-For example, NumPy early on adopted distributed revision control, unit testing,
-continuous integration, code review, and executable documentation.
-It also developed a standard for ``docstrings''—text describing each NumPy function.
-This quickly became a documentation standard for the larger scientific Python community.
+NumPy is both a cause and an effect of a striking shift in the nature of
+scientific computing.  The traditional model of scientific software has been of
+the user as {\it consumer}.  Writing software is a task for specialists and
+professionals, supported by income from selling the software to scientists to
+use.  The new model is of the user as {\it owner}.  The Python language, with
+the NumPy library, provides a framework that scientists can consume, but has
+also been successful in encouraging scientists to develop as programmers, and
+to write code for the library, and the system of packages around it.  This has
+fostered a new culture of contribution and collaboration in scientific
+computing that resulted in a virtuous cycle of increasing use, and increasing
+development.
 
-Travis Oliphant eventually stepped back from leading NumPy to found a
-company and to start new projects including Numba---an accelerated Python
-compiler that could vastly accelerate NumPy expressions.  Development
-of NumPy continued by a core group that informally operated by
-consensus.
+This cycle is one of several explanations for the astonishing growth of NumPy
+in scientific computing.
+
+The first incarnation of Numpy was a package called "Numeric", written in 1995
+by Jim Hugunin, then completing his master's thesis at MIT on
+superconductor-semiconductor junctions.  Huginin based his package on previous
+work by Jim Fulton, then working at the US Geological Survey, and acknowledges
+many others in his initial report of the package \cite{Hugunin-whitepaper}.
+Numeric provided an array object in Python, written in C, and linking to
+standard fast implementations of linear algebra.
+
+As Hugunin points out, at the time he wrote Numeric, commercial languages for
+interactive computing with numerical arrays were already well-established,
+including Matlab and IDL.  At the time, it must have seemed a safe bet that
+Numeric would not succeed in this market.  The commercial languages were
+developed and maintained by large numbers of full-time professional programmers
+and supported by formal quality assurance, release managers and writers of
+scientific documentation.  In contrast, Numeric, and later, Numpy, would
+continue to be developed almost entirely by scientists, including graduate
+students and faculty, without specific funding, and often to the detriment of
+the scientists' careers.
+
+Table X1 about here: backgrounds of initial designers (mainly math, physics).
+
+Table X2 about here: backgrounds of major committers (mainly math, physics).
+
+Around 1998, the Space Telescope Science Institute (STScI) software group began
+to use Python heavily, and, around 2000, wrote a re-implementation of much of
+Numeric, called Numarray, to support their need to work on large memory-mapped
+arrays, and arrays of mixed data type records \cite{STScI-slither}.  This
+briefly caused the Numeric and Numarray communities to diverge, until 2005,
+when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
+both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
+This merger resulted in NumPy.  At the time, Oliphant was an assistant
+professor of Electrical and Computer Engineering at Brigham Young University.
+
+The creation of NumPy spurred renewed development in the larger scientific
+Python ecosystem and heralded the current era of wide-spread use of Python for
+scientific computing.
+
+Over the next 10 years, Numpy attracted a relatively large pool of scientist
+developers (see Figure ~\ref{fig:prs-over-time}).  To benefit from these
+user-developers, the project developed an increasingly strong culture of using
+software-engineering practice to improve collaboration and reduce error
+\cite{millman2014developing}. The NumPy team was early in adopting distributed
+revision control and code review to improved collaboration on code, and
+continuous testing so we run a large battery automated tests for every proposed
+change in Numpy. At first, Numpy had rather sparse documentation, but the project
+developed a new standard for ``docstrings'' —- text describing each NumPy
+function -- and used this standard to write comprehensive high-quality
+documention, integrated with the source code. The Numpy documentation format
+quickly became a standard for the larger Python community.
+
+As we implied above, there were several factors that allowed this rapid growth
+and successful development.
+
+Hugunin identified the first factor in his initial description of the Numeric
+package \cite{Hugunin-whitepaper}.  Numerical computing is usually one
+component of a larger programming task.  Therefore:
+
+\begin{quote}
+    Rather than trying to retrofit an existing numerical language to support
+    the wealth of features found in a powerful, modern, general-purpose
+    programming language, it makes much more sense to attack the problem from
+    the other direction and add the features of a powerful numerical
+    programming language to Python.
+\end{quote}
+
+Python is already a good choice for many standard programming tasks such as
+cleaning data, interacting with web resources and parsing text.  It has a wide
+range of libraries for many different tasks. Adding fast array operations and
+linear algebra allows the scientist to do all their work with within the same
+language - and one that has the advantage of being famously easy to learn and
+teach.
+
+The second factor flows from Python's nature as an open-source language,
+embedded within the open-source community.  Python attracts developers who like
+to build and contribute.  As a result, Numpy has always been a library that was
+developed by scientists in order to do their work.  Commercial specialist
+numerical languages can encourage consumer-users, who pick up a tool and use
+it, but do not expect to contribute to development.  In contrast, Numpy often
+attracts user-developers, who want to work with Numpy, and are prepared to fix
+and extend the library as they work. User-developers have been the fundamental
+force in developing Numpy, and shaping it into a library that has been refined
+in the furnace of practical scientific work.
+
+The third factor is much like the second: a culture in Python that concentrates
+on high quality of code and distribution. Because Python is open-source, it has
+a strong culture of contribution, and therefore, of using tools and process to
+improve collaborative software development, such as distributed version
+control, code review, and automated testing.  These have allowed the project to
+scale safely as it attracted more users and developers, even though these
+developers rarely came to the project with training in software engineering.
+
+Lastly, we believe Numpy has benefitted greatly from a well-chosen application
+programming interface (API).  Scientific and numerical programming needs to be
+fast, and scale to very large datasets.  The Numpy API defines a simple wrapper
+for data in memory so it can be represented as a one- or multi-dimensional
+array.  The simplicity of this wrapper has made it very successful as a
+standard way of representing arrays in memory, and has made it relatively easy
+for other libraries to develop fast and memory-efficient compiled code, usually
+in C or Fortran, that can manipulate these arrays and pass them back to Python.
+This has been a large factor in the growth of numerical computing libaries
+around Numpy and Python.
 
@@
 \fixme{discuss Figure~\ref{fig:basic-functionality} .. vectors / matrices / and then more ...
-indexing and slicing}
+indexing and slicing, one-liners that would be tens or hundreds of lines of code in
+languages such as C.}
 

### 2020-01-04 paper.tex
Use two columns

@@
-\documentclass[fleqn,10pt]{wlscirep}
+\documentclass[fleqn,10pt,twocolumn]{wlscirep}
 
@@
   basicstyle=\ttfamily\footnotesize,  % size of fonts used for the code
-  xleftmargin=2\parindent,
+  xleftmargin=\parindent,
   breaklines=true,   % automatic line breaking only at whitespace
@@
 
-\usepackage{lineno}
+\usepackage[switch]{lineno}
 \linenumbers

### 2020-01-04 body.tex
Use two columns
Reorganize text

@@
 
-With its first stable release in October 2006, NumPy unified
-the scientific Python community and now underpins almost every library
-that does numerical computation, including SciPy\cite{virtanen2019scipy},
-pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
-and scikit-image\cite{vanderwalt2014scikit}.
-NumPy consists of the \code{ndarray}-object along with utility functions that
-operate on it.
-Because of its inherent simplicity—being a pointer to memory with some
-associated meta-data about shape, data-type, and so forth—the NumPy array is
-the {\it de facto} exchange format for array data in Python.
-The library has such widespread adoption that not only the array object but also its
-{\it Application Programming Interface} (API) has become ubiquitous as
-a language for tensor computation---witnessed by its use in popular
-deep learning libraries such as PyTorch\cite{pytorch}.
-
-\subsection{Background}
-
 % https://books.google.com/books?id=QQJODwAAQBAJ&pg=PA101&lpg=PA101&dq=%E2%80%9Cthe+secret+to+successful+hiring+is+this:+look+for+the+people+who+want+to+change+the+world.%E2%80%9D&source=bl&ots=D6l8JXfB20&sig=ACfU3U0x1X9Gh8k6MlNV0BqhBa_73Q0yfQ&hl=en&sa=X&ved=2ahUKEwj48pqg_oHgAhVohuAKHYlJAsc4ChDoATAAegQIChAB#v=onepage&q=%E2%80%9Cthe%20secret%20to%20successful%20hiring%20is%20this%3A%20look%20for%20the%20people%20who%20want%20to%20change%20the%20world.%E2%80%9D&f=false
@@
 
-The first incarnation of Numpy was a package called "Numeric", written in 1995
+\subsection{History}
+
+The first incarnation of NumPy was a package called "Numeric", written in 1995
 by Jim Hugunin, then completing his master's thesis at MIT on
-superconductor-semiconductor junctions.  Huginin based his package on previous
+superconductor-semiconductor junctions.  Hugunin based his package on previous
 work by Jim Fulton, then working at the US Geological Survey, and acknowledges
@@
 and supported by formal quality assurance, release managers and writers of
-scientific documentation.  In contrast, Numeric, and later, Numpy, would
+scientific documentation.  In contrast, Numeric, and later, NumPy, would
 continue to be developed almost entirely by scientists, including graduate
@@
 to use Python heavily, and, around 2000, wrote a re-implementation of much of
-Numeric, called Numarray, to support their need to work on large memory-mapped
+Numeric, called NumArray, to support their need to work on large memory-mapped
 arrays, and arrays of mixed data type records \cite{STScI-slither}.  This
-briefly caused the Numeric and Numarray communities to diverge, until 2005,
+briefly caused the Numeric and NumArray communities to diverge, until 2005,
 when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
@@
 
-The creation of NumPy spurred renewed development in the larger scientific
-Python ecosystem and heralded the current era of wide-spread use of Python for
-scientific computing.
-
-Over the next 10 years, Numpy attracted a relatively large pool of scientist
-developers (see Figure ~\ref{fig:prs-over-time}).  To benefit from these
+Over the next 10 years, NumPy attracted a relatively large pool of scientist
+developers.  To benefit from these
 user-developers, the project developed an increasingly strong culture of using
@@
 continuous testing so we run a large battery automated tests for every proposed
-change in Numpy. At first, Numpy had rather sparse documentation, but the project
-developed a new standard for ``docstrings'' —- text describing each NumPy
-function -- and used this standard to write comprehensive high-quality
-documention, integrated with the source code. The Numpy documentation format
+change in NumPy. At first, NumPy had rather sparse documentation, but the project
+developed a new standard for ``docstrings''---text describing each NumPy
+function---and used this standard to write comprehensive high-quality
+documentation, integrated with the source code. The NumPy documentation format
 quickly became a standard for the larger Python community.
 
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
+    \caption{Number of pull requests merged into the NumPy master branch for each
+        quarter since 2012. The total number of PRs is indicated with the
+        lower blue area showing the portion contributed by current or previous
+        maintainers.}\label{fig:prs-over-time}
+\end{figure}
+
+NumPy is currently maintained by a group of 23 contributors with commit rights
+to the NumPy code base. Out of these, 17 maintainers were active in
+2019, 4 of whom were paid to work on the project full-time.
+Additionally, there are a few long term developers who contributed and maintain
+specific parts of NumPy, but are not officially maintainers.
+At a release cycle of about every half year, the five recent releases in the years
+2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
+    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
+    All data provided here is based on the development which happened through GitHub
+    pull requests. In some cases contributions by maintainers may not be categorized as such.}
+% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
+with each release attracting more than a hundred new contributors.
+Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
+master branch.
+Although the number of PRs being merged fluctuates,
+the plot indicates an increased number of contributions over the past
+years.
+
+\subsection{Enabling factors}
+
+The creation of NumPy spurred renewed development in the larger scientific
+Python ecosystem and heralded the current era of wide-spread use of Python for
+scientific computing.
 As we implied above, there were several factors that allowed this rapid growth
@@
 linear algebra allows the scientist to do all their work with within the same
-language - and one that has the advantage of being famously easy to learn and
+language---and one that has the advantage of being famously easy to learn and
 teach.
@@
 embedded within the open-source community.  Python attracts developers who like
-to build and contribute.  As a result, Numpy has always been a library that was
+to build and contribute.  As a result, NumPy has always been a library that was
 developed by scientists in order to do their work.  Commercial specialist
 numerical languages can encourage consumer-users, who pick up a tool and use
-it, but do not expect to contribute to development.  In contrast, Numpy often
-attracts user-developers, who want to work with Numpy, and are prepared to fix
+it, but do not expect to contribute to development.  In contrast, NumPy often
+attracts user-developers, who want to work with NumPy, and are prepared to fix
 and extend the library as they work. User-developers have been the fundamental
-force in developing Numpy, and shaping it into a library that has been refined
+force in developing NumPy, and shaping it into a library that has been refined
 in the furnace of practical scientific work.
@@
 
-Lastly, we believe Numpy has benefitted greatly from a well-chosen application
+Lastly, we believe NumPy has benefited greatly from a well-chosen application
 programming interface (API).  Scientific and numerical programming needs to be
-fast, and scale to very large datasets.  The Numpy API defines a simple wrapper
+fast, and scale to very large datasets.  The NumPy API defines a simple wrapper
 for data in memory so it can be represented as a one- or multi-dimensional
@@
 in C or Fortran, that can manipulate these arrays and pass them back to Python.
-This has been a large factor in the growth of numerical computing libaries
-around Numpy and Python.
+This has been a large factor in the growth of numerical computing libraries
+around NumPy and Python.
 
@@
 
-\subsection{Funding}
+\subsection{Hybrid development}
 
@@
 
-\section{Community and library}
+\section{The \code{ndarray} data structure}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
-    \caption{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012. The total number of PRs is indicated with the
-        lower blue area showing the portion contributed by current or previous
-        maintainers.}\label{fig:prs-over-time}
-\end{figure}
-
-NumPy is currently maintained by a group of 23 contributors with commit rights
-to the NumPy code base. Out of these, 17 maintainers were active in
-2019, 4 of whom were paid to work on the project full-time.
-Additionally, there are a few long term developers who contributed and maintain
-specific parts of NumPy, but are not officially maintainers.
-At a release cycle of about every half year, the five recent releases in the years
-2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
-    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
-    All data provided here is based on the development which happened through GitHub
-    pull requests. In some cases contributions by maintainers may not be categorized as such.}
-% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
-with each release attracting more than a hundred new contributors.
-Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
-master branch.
-Although the number of PRs being merged fluctuates,
-the plot indicates an increased number of contributions over the past
-years.
-
-\subsection{Basic functionality}
+With its first stable release in October 2006, NumPy unified
+the scientific Python community and now underpins almost every library
+that does numerical computation, including SciPy\cite{virtanen2019scipy},
+pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
+and scikit-image\cite{vanderwalt2014scikit}.
+NumPy consists of the \code{ndarray}-object along with utility functions that
+operate on it.
+Because of its inherent simplicity---being a pointer to memory with some
+associated meta-data about shape, data-type, and so forth---the NumPy array is
+the {\it de facto} exchange format for array data in Python.
+The library has such widespread adoption that not only the array object but also its
+{\it Application Programming Interface} (API) has become ubiquitous as
+a language for tensor computation---witnessed by its use in popular
+deep learning libraries such as PyTorch\cite{pytorch}.
 
@@
 
-\fixme{discuss Figure~\ref{fig:basic-functionality} .. vectors / matrices / and then more ...
+\fixme{discuss .. vectors / matrices / and then more ...
 indexing and slicing, one-liners that would be tens or hundreds of lines of code in
@@
 
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
-  \caption{Vectors (1-d arrays)}
-  \label{fig:sub1}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
+
 \begin{lstlisting}
@@
 \end{lstlisting}
-  \caption{Vectors (1-d arrays) and Matrices (2-d arrays)}
-  \label{fig:sub2}
-\end{subfigure}
-\caption{Basic functionality}
-\label{fig:basic-functionality}
-\end{figure}
 
@@
 
-\subsection{Advanced functionality}
-
-The \emph{universal functions} (or \emph{ufuncs})
-are functions written in C that implement efficient looping over
-NumPy arrays. An important feature of ufuncs is the built-in
-implementation of \emph{broadcasting}.  For example, the function
-\code{arctan2(x, y)} is a ufunc that accepts two arguments and computes
-$\tan^{-1}(y/x)$.  The ufunc loops over the dimensions of the inputs
-in such a way that if, say, \code{x} is a 1-d array with length 3, and
-\code{y} is a 2-d array with shape $2 \times 1$, the output will be
-an array with shape $2 \times 3$ (see Figure~\ref{fig:broadcasting}).  The ufunc machinery takes care
-of calling the function with all the appropriate combinations of
-input array elements to compute the output array.
-The elementary arithmetic operations of addition, multiplication, etc.,
-are implemented as ufuncs, so broadcasting applies even in expressions
-such as \code{x + y * z}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.25\textwidth]{static/broadcasting}
-  \caption{
-    Broadcasting a $1 \times 3$ array against a $2 \times 1$ array
-    yields a $2 \times 3$ result.
-  }
-  \label{fig:broadcasting}
-\end{figure}
-
-NumPy also provides basic infrastructure for other packages in the scientific
-Python ecosystem.
-The \code{numpy.testing} subpackage provides functions such as
-\code{assert\_allclose(actual, desired)} that may be used in unit
-test suites for code that uses NumPy arrays.
-The \code{numpy.distutils} subpackage provides build support for C++, Fortran,
-BLAS/LAPACK, and other relevant libraries for scientific computing.
-The program \code{f2py} is a tool for
-building NumPy-aware Python wrappers of Fortran functions.
-NumPy itself does not use any Fortran code;  F2PY is part of NumPy
-for historical reasons.
+%\subsection{Advanced functionality}
+%
+%The \emph{universal functions} (or \emph{ufuncs})
+%are functions written in C that implement efficient looping over
+%NumPy arrays. An important feature of ufuncs is the built-in
+%implementation of \emph{broadcasting}.  For example, the function
+%\code{arctan2(x, y)} is a ufunc that accepts two arguments and computes
+%$\tan^{-1}(y/x)$.  The ufunc loops over the dimensions of the inputs
+%in such a way that if, say, \code{x} is a 1-d array with length 3, and
+%\code{y} is a 2-d array with shape $2 \times 1$, the output will be
+%an array with shape $2 \times 3$ (see Figure~\ref{fig:broadcasting}).  The ufunc machinery takes care
+%of calling the function with all the appropriate combinations of
+%input array elements to compute the output array.
+%The elementary arithmetic operations of addition, multiplication, etc.,
+%are implemented as ufuncs, so broadcasting applies even in expressions
+%such as \code{x + y * z}.
+%
+%\begin{figure}
+%  \centering
+%  \includegraphics[width=0.5\linewidth]{static/broadcasting}
+%  \caption{
+%    Broadcasting a $1 \times 3$ array against a $2 \times 1$ array
+%    yields a $2 \times 3$ result.
+%  }
+%  \label{fig:broadcasting}
+%\end{figure}
+%
+%NumPy also provides basic infrastructure for other packages in the scientific
+%Python ecosystem.
+%The \code{numpy.testing} subpackage provides functions such as
+%\code{assert\_allclose(actual, desired)} that may be used in unit
+%test suites for code that uses NumPy arrays.
+%The \code{numpy.distutils} subpackage provides build support for C++, Fortran,
+%BLAS/LAPACK, and other relevant libraries for scientific computing.
+%The program \code{f2py} is a tool for
+%building NumPy-aware Python wrappers of Fortran functions.
+%NumPy itself does not use any Fortran code;  F2PY is part of NumPy
+%for historical reasons.
 
@@
 
-The NumPy codebase has been actively developed for over two decades (first as
-Numeric and NumArray).
 Given the complexity of the codebase and the massive number of projects depending
@@
 x_gpu = cp.array([1, 2, 3])
-y = np.sum(x_gpu)  # This works, and returns a GPU array
+y = np.sum(x_gpu)  # Returns a GPU array
 \end{lstlisting}
 
-Similarly, user defined functions composed using NumPy can now be
-applied to, e.g., multi-node distributed Dask arrays:
-
-\begin{lstlisting}
-import numpy as np
-import dask.array as da
-
-def f(x):
-    """Function using NumPy API calls"""
-    y = np.tensordot(x, x.T)
-    return np.mean(np.log(y + 1))
-
-x_local = np.random.random([10000, 10000])  # random local array
-x_distr = da.random.random([10000, 10000])  # random distributed array
-
-f(x_local)  # returns a NumPy array
-f(x_distr)  # works, returns a Dask array
-\end{lstlisting}
+%
+%Similarly, user defined functions composed using NumPy can now be
+%applied to, e.g., multi-node distributed Dask arrays:
+%
+%\begin{lstlisting}
+%import numpy as np
+%import dask.array as da
+%
+%
+%def f(x):
+%    """Function using NumPy API calls"""
+%    y = np.tensordot(x, x.T)
+%    return np.mean(np.log(y + 1))
+%
+%
+%x_local = np.random.random([10000, 10000])  # random local array
+%x_distr = da.random.random([10000, 10000])  # random distributed array
+%
+%f(x_local)  # returns a NumPy array
+%f(x_distr)  # works, returns a Dask array
+%\end{lstlisting}
 
@@
 \begin{lstlisting}
-from numpy.random import Generator, PCG64, SeedSequence
+from numpy.random import (
+    Generator,
+    PCG64,
+    SeedSequence,
+)
 
-seq = SeedSequence(10304245474441179933310169597522947680)
+seq = SeedSequence(1030424547444117993331016959)
 pcg = PCG64(seq)

### 2020-01-05 body.tex
More work on outline

@@
 % https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
+% http://doi.org/10.5281/zenodo.3585761
+% http://doi.org/10.5281/zenodo.3585767
 In 2017, NumPy received its first large grants totaling 1.3M USD from the
 Gordon \& Betty Moore and the Alfred P. Sloan foundations.
-Stéfan van der Walt is the PI and manages four programmers working on the project.
 These two grants focus on addressing the technical debt accrued over the years and
 on setting in place standards and architecture to encourage more sustainable development.
-% http://doi.org/10.5281/zenodo.3585761
-% http://doi.org/10.5281/zenodo.3585767
-
 % CZI -- NumFOCUS/QuanSight
@@
 % https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
-NumPy received another grant for \$195,000 from the Chan Zuckerberg
-Initiative at the end of 2019 with Ralf Gommers as the PI.
-This grant, building on the technical work supported by the previous
-grant, will focus on better serving NumPy's large number of beginning
+NumPy received a third grant for 200K USD from the Chan Zuckerberg
+Initiative at the end of 2019.
+This grant focuses on better serving NumPy's large number of beginning
 to intermediate level users and on growing the community of NumPy
 contributors.
-It will also provide support to OpenBLAS, on which NumPy depends for
-accelerated linear algebra.
-
 Finally, the project receives around 10K USD annually from Tidelift, which is
 typically used to fund documentation writers and website developers.
+
+The first activities organized for the new grants were a development planning
+sprint, a weekly community call, and a meeting to write a draft roadmap.
+Since the roadmap was meant to represent community consensus on future
+development, it was presented for comment at the annual SciPy
+conference during a so-called ``Birds of a Feather'' session, attended
+by more than 100 people.  One additional meeting was held to discuss
+specific enhancements, such as the development of a new data-type
+system.  The final roadmap was vetted by the community via online
+discussion before being accepted.
 
@@
 
-Given the complexity of the codebase and the massive number of projects depending
-on it, large changes require careful planning and substantial work.
-As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
-of enhancement proposals---used infrequently and informally until 2017 when,
-to facilitate work expected from funded development, he formalized the
-NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
-Python Enhancement Proposals (PEPs) for ``proposing major new
-features, for collecting community input on an issue, and for
-documenting the design decisions that have gone into
-Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
-%Since then there have been 19 proposed NEPS---6 have been implemented,
-%4 have been accepted and are being implemented, 4 are under
-%consideration, 3 have been deferred or superseded, and 2 have been rejected
-%or withdrawn.
-
-The first activities organized for the new grant were a NEP development
-sprint, a weekly community call, and a meeting to write a draft roadmap.
-Since the roadmap was meant to represent community consensus on future
-development, it was presented for comment at the annual SciPy
-conference during a so-called ``Birds of a Feather'' session, attended
-by more than 100 people.  One additional meeting was held to discuss
-specific enhancements, such as the development of a new data-type
-system.  The final roadmap was vetted by the community via online
-discussion before being accepted.
-
 %Weekly community calls alternate between triage and
@@
 %arisen, which NumPy was not originally designed for.
-With the recent infusion of funding and the newly formalized NEP process in
-place we have been able to tackle a number of important large scale changes.
+
+With the recent infusion of funding and a clear process for coordinating with
+the developer community, we have been able to tackle a number of important
+large scale changes.
 We highlight two of those below.

### 2020-01-05 body.tex
Remove old text

@@
 \section{Introduction}
-
-%NumPy is the fundamental array computation library for the Python
-%ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-%With its first stable release in October 2006, NumPy unified
-%the scientific Python community and now underpins almost every library
-%that does numerical computation, including SciPy\cite{virtanen2019scipy},
-%pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
-%and scikit-image\cite{vanderwalt2014scikit}.
-%Not surprisingly, many important scientific research projects and
-%results crucially rely on NumPy.
-%It plays an essential role in many research analysis pipelines in
-%fields as diverse as physics, biology, astronomy, neuroscience,
-%material science, engineering, and chemistry.
-%% w/ citations
-%In particular, NumPy played a fundamental role in discovering
-%gravitational waves\cite{pycbc} and the first imaging of a black
-%hole\cite{eht-imaging}, and lets us explore our
-%universe\cite{jenness2018lsst}.
-
-%[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
-%  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
-%  IDL is now below Python, Pangeo <-> xarray, Earthcube,
-%  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
-%  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
-%  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]
 
@@
 
-The first incarnation of NumPy was a package called "Numeric", written in 1995
+The first incarnation of NumPy was a package called Numeric, written in 1995
 by Jim Hugunin, then completing his master's thesis at MIT on
@@
 
-% While the scientific Python community grew and new challenges arose,
-
-%By 2009, the project had entered a
-%period of uncoordinated, opportunistic development, and was
-%increasingly weighed down by technical debt that had accrued since the
-%days of Numeric.
-
 \subsection{Hybrid development}
@@
 
-%\subsection{Advanced functionality}
-%
-%The \emph{universal functions} (or \emph{ufuncs})
-%are functions written in C that implement efficient looping over
-%NumPy arrays. An important feature of ufuncs is the built-in
-%implementation of \emph{broadcasting}.  For example, the function
-%\code{arctan2(x, y)} is a ufunc that accepts two arguments and computes
-%$\tan^{-1}(y/x)$.  The ufunc loops over the dimensions of the inputs
-%in such a way that if, say, \code{x} is a 1-d array with length 3, and
-%\code{y} is a 2-d array with shape $2 \times 1$, the output will be
-%an array with shape $2 \times 3$ (see Figure~\ref{fig:broadcasting}).  The ufunc machinery takes care
-%of calling the function with all the appropriate combinations of
-%input array elements to compute the output array.
-%The elementary arithmetic operations of addition, multiplication, etc.,
-%are implemented as ufuncs, so broadcasting applies even in expressions
-%such as \code{x + y * z}.
-%
-%\begin{figure}
-%  \centering
-%  \includegraphics[width=0.5\linewidth]{static/broadcasting}
-%  \caption{
-%    Broadcasting a $1 \times 3$ array against a $2 \times 1$ array
-%    yields a $2 \times 3$ result.
-%  }
-%  \label{fig:broadcasting}
-%\end{figure}
-%
-%NumPy also provides basic infrastructure for other packages in the scientific
-%Python ecosystem.
-%The \code{numpy.testing} subpackage provides functions such as
-%\code{assert\_allclose(actual, desired)} that may be used in unit
-%test suites for code that uses NumPy arrays.
-%The \code{numpy.distutils} subpackage provides build support for C++, Fortran,
-%BLAS/LAPACK, and other relevant libraries for scientific computing.
-%The program \code{f2py} is a tool for
-%building NumPy-aware Python wrappers of Fortran functions.
-%NumPy itself does not use any Fortran code;  F2PY is part of NumPy
-%for historical reasons.
-
 \section{Recent technical improvements}
-
-%Weekly community calls alternate between triage and
-%higher level discussion.  The calls not only involve developers from
-%the community, but provide a venue for vendors and other external
-%groups to provide input.  For example, after Intel produced a forked
-%version of NumPy, one of their developers joined a call to discuss
-%community concerns.
-
-%NumPy plays a central role in building and standardizing much of the scientific
-%Python community infrastructure.
-%The docstring standard mentioned in the history section is now widely adopted.
-%We are also now using the NEP system as a way to help coordinate the larger
-%scientific Python community.
-%% https://numpy.org/neps/nep-0029-deprecation_policy.html
-%For example, in NEP 29, we recommend, along with leaders from various other
-%projects, that all projects across the Scientific Python ecosystem adopt a
-%common ``time window-based'' policy for support of Python and NumPy versions.
-%This standard will simplify downstream project and release planning.
-
-% https://numpy.org/neps/scope.html
-% https://numpy.org/neps/roadmap.html
-
-%We showed that NumPy is actively developed according to a community
-%vetted plan that integrates formal enhancement proposals.  Next, we
-%give a more in-depth overview of the team developing NumPy, their
-%activities, as well as the governance and community structures that
-%guide the collaboration.
-
-%Moreover, with the now prevalent need for distributed computing and the
-%emergence of specialized hardware such as GPUs, many new usecases have
-%arisen, which NumPy was not originally designed for.
 
@@
 these projects are consumers of the NumPy API.
-%The central position of NumPy takes within the numerical and scientific
-%Python ecosystem means that a vast number of projects are built on it.
-%These projects are consumers of the NumPy API.
 Over the last several years, a growing number of projects are providers of
@@
 
-%
-%Similarly, user defined functions composed using NumPy can now be
-%applied to, e.g., multi-node distributed Dask arrays:
-%
-%\begin{lstlisting}
-%import numpy as np
-%import dask.array as da
-%
-%
-%def f(x):
-%    """Function using NumPy API calls"""
-%    y = np.tensordot(x, x.T)
-%    return np.mean(np.log(y + 1))
-%
-%
-%x_local = np.random.random([10000, 10000])  # random local array
-%x_distr = da.random.random([10000, 10000])  # random distributed array
-%
-%f(x_local)  # returns a NumPy array
-%f(x_distr)  # works, returns a Dask array
-%\end{lstlisting}
-
 \subsection{Random}
-
-%The random number generator library in NumPy provides several alternative
-%\emph{bit stream generators} that provide the core function of generating
-%random integers.
-%A higher-level generator class that implements an assortment of
-%probability distributions is provided. It includes the beta, gamma
-%and Weibull distributions, the univariate and multivariate normal
-%distributions, and more.
 
@@
 apply to \code{RandomState}.
-
-% Mention explicitly the NEPs involved
 
@@
 The main remaining item is an overhaul of the data type system, which
-we are currently working on \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
+we are currently working on.
 The meaning of each element within a NumPy array is described by its

### 2020-01-05 body.tex
Draft data structure section

@@
 it allows for the efficient representation of $n$-dimensional arrays.
+More details about the \code{ndarray} data structure are given in ``The NumPy array:
+a structure for efficient numerical computation''\cite{vanderwalt2011numpy}
+and our online documentation.
 
-\fixme{discuss .. vectors / matrices / and then more ...
-indexing and slicing, one-liners that would be tens or hundreds of lines of code in
-languages such as C.}
+The need for such a data structure arises often in science.
+For instance, they may store measurements from an experiment or simulation
+at discrete time points.
+Data analysis frequently involves operations $(+, -, \times, /, \log, \arctan, \dots)$ on
+the individual measurements.
+In programming languages such as C, this would involve looping over every element
+to perform these operations.
+Array computations generalize these element-wise operations on scalars to
+vectors (1-d arrays), matrices (2-d arrays), and higher dimensional arrays (n-d
+arrays).
+For instance, if \code{x} is an \code{ndarray}, then \code{2 * x} produces
+an \code{ndarray} whose values are twice those in \code{x}.
+Similarly, if \code{x} and \code{y} have the same shape, then
+\code{x * y} produces an \code{ndarray} whose values are the products of
+the corresponding entries of \code{x} and \code{y}.
+These \emph{vectorized} operations often produce one-liners that would tens
+or hundreds of lines of code in languages such as C.
 
-More details about the data structure are given in ``The NumPy array:
-a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.
+Element-wise operations are not always desired.
+Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is 1-d array.
+Then \code{A @ x} computes the matrix-vector product $Ax$ and
+\code{x @ A} computes the vector-matrix product $x^\top A$.
+This matrix multiplication operation extends to higher-dimensions by
+treating the array as a stack of matrices.
 
-\begin{lstlisting}
->>> import numpy as np
->>> x = np.array([0, 1, 2, 3])
->>> x.dtype
-dtype('int64')
->>> x.shape
-(4,)
-
->>> y = 2 * x
->>> y
-array([0, 2, 4, 6])
-
->>> x * y
-array([ 0,  2,  8, 18])
-
->>> x @ y
-28
-\end{lstlisting}
-
-\begin{lstlisting}
->>> import numpy as np
->>> x = np.array([1, 2])
->>> A = np.array([[0.1, 0.2], [0.3, 0.2]])
->>> A.dtype 
-dtype('float64')
->>> A.shape
-(2, 2)
-
->>> A * x
-array([[0.1, 0.4],
-       [0.3, 0.4]])
-
->>> A @ x
-array([0.5, 0.7])
->>> x @ A
-array([0.7, 0.6])
-\end{lstlisting}
-
-NumPy provides a large collection of functions for array manipulation
+NumPy also provides a large collection of functions for array manipulation
 including functions for: creating, reshaping, concatenating, and padding arrays;

### 2020-01-05 body.tex
Minor fix
Update section title
Update outline

@@
-\section{Introduction}
-
 % https://books.google.com/books?id=QQJODwAAQBAJ&pg=PA101&lpg=PA101&dq=%E2%80%9Cthe+secret+to+successful+hiring+is+this:+look+for+the+people+who+want+to+change+the+world.%E2%80%9D&source=bl&ots=D6l8JXfB20&sig=ACfU3U0x1X9Gh8k6MlNV0BqhBa_73Q0yfQ&hl=en&sa=X&ved=2ahUKEwj48pqg_oHgAhVohuAKHYlJAsc4ChDoATAAegQIChAB#v=onepage&q=%E2%80%9Cthe%20secret%20to%20successful%20hiring%20is%20this%3A%20look%20for%20the%20people%20who%20want%20to%20change%20the%20world.%E2%80%9D&f=false
@@
 
-\subsection{History}
+\section*{History}
 
@@
 
-\subsection{Enabling factors}
+\section*{Enabling factors}
 
@@
 
-\subsection{Hybrid development}
+\section*{Hybrid development}
 
@@
 % https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
-NumPy received a third grant for 200K USD from the Chan Zuckerberg
+NumPy received a third grant for 195K USD from the Chan Zuckerberg
 Initiative at the end of 2019.
@@
 
-\section{The \code{ndarray} data structure}
+\section*{NumPy arrays}
 
@@
 data type in a contiguous block of memory; along with meta-data such as shape,
-it allows for the efficient representation of $n$-dimensional arrays.
-More details about the \code{ndarray} data structure are given in ``The NumPy array:
-a structure for efficient numerical computation''\cite{vanderwalt2011numpy}
-and our online documentation.
-
+it allows for the efficient representation of $n$-dimensional arrays \cite{vanderwalt2011numpy}.
 The need for such a data structure arises often in science.
@@
 at discrete time points.
-Data analysis frequently involves operations $(+, -, \times, /, \log, \arctan, \dots)$ on
+
+Data analysis frequently involves operations $(+, -, \times, /, \log, \sin, \dots)$ on
 the individual measurements.
-In programming languages such as C, this would involve looping over every element
-to perform these operations.
+In programming languages such as C or Java, this would involve looping over
+every element to perform these operations.
 Array computations generalize these element-wise operations on scalars to
@@
 the corresponding entries of \code{x} and \code{y}.
-These \emph{vectorized} operations often produce one-liners that would tens
+These \emph{vectorized} operations often produce one-liners that would be tens
 or hundreds of lines of code in languages such as C.
@@
 
-\section{Recent technical improvements}
-
 With the recent infusion of funding and a clear process for coordinating with
@@
 
-\subsection{The Array Function Protocol}
+\section*{Array function protocol}
 
@@
 
-\subsection{Random}
+\section*{Random}
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 

### 2020-01-06 body.tex
Move hybrid development
Move figure

@@
 
-Table X1 about here: backgrounds of initial designers (mainly math, physics).
+%Table X1 about here: backgrounds of initial designers (mainly math, physics).
 
-Table X2 about here: backgrounds of major committers (mainly math, physics).
+%Table X2 about here: backgrounds of major committers (mainly math, physics).
 
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
-    \caption{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012. The total number of PRs is indicated with the
-        lower blue area showing the portion contributed by current or previous
-        maintainers.}\label{fig:prs-over-time}
-\end{figure}
-
-NumPy is currently maintained by a group of 23 contributors with commit rights
-to the NumPy code base. Out of these, 17 maintainers were active in
-2019, 4 of whom were paid to work on the project full-time.
-Additionally, there are a few long term developers who contributed and maintain
-specific parts of NumPy, but are not officially maintainers.
-At a release cycle of about every half year, the five recent releases in the years
-2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
-    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
-    All data provided here is based on the development which happened through GitHub
-    pull requests. In some cases contributions by maintainers may not be categorized as such.}
-% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
-with each release attracting more than a hundred new contributors.
-Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
-master branch.
-Although the number of PRs being merged fluctuates,
-the plot indicates an increased number of contributions over the past
-years.
+NumPy now underpins almost every Python library that does scientific or
+numerical computation, including SciPy\cite{virtanen2019scipy},
+matplotlib\cite{hunter2007matplotlib}, pandas\cite{mckinney-proc-scipy-2010},
+scikit-learn\cite{pedregosa2011scikit}, and
+scikit-image\cite{vanderwalt2014scikit}.
+NumPy consists of the $n$-dimensional array object along with utility functions
+that operate on it.
+Because of its inherent simplicity---being a pointer to memory with some
+associated meta-data about shape, data-type, and so forth---the NumPy array is
+the {\it de facto} exchange format for array data in Python.
+The library has such widespread adoption that not only the array object but also its
+{\it Application Programming Interface} (API) has become ubiquitous as
+a language for tensor computation---witnessed by its use in popular
+deep learning libraries such as PyTorch\cite{pytorch}.
 
@@
 
+\section*{NumPy arrays}
+
+The NumPy array data structure stores regularly spaced elements of a single
+data type in a contiguous block of memory; along with meta-data such as shape,
+it allows for the efficient representation of $n$-dimensional arrays \cite{vanderwalt2011numpy}.
+The need for such a data structure arises often in science.
+For instance, they may store measurements from an experiment or simulation
+at discrete time points.
+
+Data analysis frequently involves operations $(+, -, \times, /, \log, \sin, \dots)$ on
+the individual measurements.
+In programming languages such as C or Java, this would involve looping over
+every element to perform these operations.
+Array computations generalize these element-wise operations on scalars to
+vectors (1-d arrays), matrices (2-d arrays), and higher dimensional arrays (n-d
+arrays).
+For instance, if \code{x} is an array, then \code{2 * x} produces
+an array whose values are twice those in \code{x}.
+Similarly, if \code{x} and \code{y} have the same shape, then
+\code{x * y} produces an array whose values are the products of
+the corresponding entries of \code{x} and \code{y}.
+When \code{x} and \code{y} have different shapes, the smaller array is \emph{broadcast}
+across the larger array to a common shape when sensible.
+These vectorized operations often produce one-liners that would be many tens
+of lines of code in languages such as C.
+
+Element-wise operations are not always desired.
+Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is 1-d array.
+Then \code{A @ x} computes the matrix-vector product $Ax$ and
+\code{x @ A} computes the vector-matrix product $x^\top A$.
+This matrix multiplication operation extends to higher-dimensions by
+treating the array as a stack of matrices.
+
+NumPy also provides a large collection of functions for array manipulation
+including functions for:
+creating, reshaping, concatenating, and padding arrays;
+searching, sorting and counting data in arrays;
+computing elementary statistics, such as the mean, median, variance, and standard deviation;
+reading and writing files;
+and more.
+It also provides extensive support for generating pseudorandom numbers
+as well as an assortment of probability distributions.
+For historical reasons, NumPy also includes basic functionality for
+linear algebra, fast Fourier transforms and windowing,
+and polynomial fitting.
+We will continue supporting these, but we will not expand beyond them.
+
 \section*{Hybrid development}
@@
 
-\section*{NumPy arrays}
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
+    \caption{Number of pull requests merged into the NumPy master branch for each
+        quarter since 2012. The total number of PRs is indicated with the
+        lower blue area showing the portion contributed by current or previous
+        maintainers.}\label{fig:prs-over-time}
+\end{figure}
 
-With its first stable release in October 2006, NumPy unified
-the scientific Python community and now underpins almost every library
-that does numerical computation, including SciPy\cite{virtanen2019scipy},
-pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
-and scikit-image\cite{vanderwalt2014scikit}.
-NumPy consists of the \code{ndarray}-object along with utility functions that
-operate on it.
-Because of its inherent simplicity---being a pointer to memory with some
-associated meta-data about shape, data-type, and so forth---the NumPy array is
-the {\it de facto} exchange format for array data in Python.
-The library has such widespread adoption that not only the array object but also its
-{\it Application Programming Interface} (API) has become ubiquitous as
-a language for tensor computation---witnessed by its use in popular
-deep learning libraries such as PyTorch\cite{pytorch}.
+NumPy is currently maintained by a group of 23 contributors with commit rights
+to the NumPy code base. Out of these, 17 maintainers were active in
+2019, 4 of whom were paid to work on the project full-time.
+Additionally, there are a few long term developers who contributed and maintain
+specific parts of NumPy, but are not officially maintainers.
+At a release cycle of about every half year, the five recent releases in the years
+2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
+    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
+    All data provided here is based on the development which happened through GitHub
+    pull requests. In some cases contributions by maintainers may not be categorized as such.}
+% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
+with each release attracting more than a hundred new contributors.
+Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
+master branch.
+Although the number of PRs being merged fluctuates,
+the plot indicates an increased number of contributions over the past
+years.
 
-The \code{ndarray} data structure stores regularly spaced elements of a single
-data type in a contiguous block of memory; along with meta-data such as shape,
-it allows for the efficient representation of $n$-dimensional arrays \cite{vanderwalt2011numpy}.
-The need for such a data structure arises often in science.
-For instance, they may store measurements from an experiment or simulation
-at discrete time points.
-
-Data analysis frequently involves operations $(+, -, \times, /, \log, \sin, \dots)$ on
-the individual measurements.
-In programming languages such as C or Java, this would involve looping over
-every element to perform these operations.
-Array computations generalize these element-wise operations on scalars to
-vectors (1-d arrays), matrices (2-d arrays), and higher dimensional arrays (n-d
-arrays).
-For instance, if \code{x} is an \code{ndarray}, then \code{2 * x} produces
-an \code{ndarray} whose values are twice those in \code{x}.
-Similarly, if \code{x} and \code{y} have the same shape, then
-\code{x * y} produces an \code{ndarray} whose values are the products of
-the corresponding entries of \code{x} and \code{y}.
-These \emph{vectorized} operations often produce one-liners that would be tens
-or hundreds of lines of code in languages such as C.
-
-Element-wise operations are not always desired.
-Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is 1-d array.
-Then \code{A @ x} computes the matrix-vector product $Ax$ and
-\code{x @ A} computes the vector-matrix product $x^\top A$.
-This matrix multiplication operation extends to higher-dimensions by
-treating the array as a stack of matrices.
-
-NumPy also provides a large collection of functions for array manipulation
-including functions for: creating, reshaping, concatenating, and padding arrays;
-searching, sorting and counting data
-in arrays; computing elementary statistics, such as the mean, median,
-variance, and standard deviation; file I/O; and more.
-It also provides extensive support for generating pseudorandom numbers
-as well as an assortment of probability distributions.
-For historical reasons, NumPy also includes basic functionality for
-linear algebra, fast Fourier transforms and windowing,
-and polynomial fitting.
-We will continue supporting these, but we will not expand beyond them.
-
-With the recent infusion of funding and a clear process for coordinating with
-the developer community, we have been able to tackle a number of important
-large scale changes.
-We highlight two of those below.
+We briefly highlight two of those below.
 
@@
 much code is written specifically for NumPy arrays and uses the NumPy API as
-\code{import numpy as np}, it can nevertheless work for them.  For
-example, here is how a CuPy GPU array can be passed through NumPy for
-processing, with all operations being dispatched back to CuPy:
-
-\begin{lstlisting}
-import numpy as np
-import cupy as cp
-
-x_gpu = cp.array([1, 2, 3])
-y = np.sum(x_gpu)  # Returns a GPU array
-\end{lstlisting}
+\code{import numpy as np}, it can nevertheless work for them.
+%For example, here is how a CuPy GPU array can be passed through NumPy for
+%processing, with all operations being dispatched back to CuPy:
+%
+%\begin{lstlisting}
+%import numpy as np
+%import cupy as cp
+%
+%x_gpu = cp.array([1, 2, 3])
+%y = np.sum(x_gpu)  # Returns a GPU array
+%\end{lstlisting}
 
@@
 generator.
-
-\begin{lstlisting}
-from numpy.random import (
-    Generator,
-    PCG64,
-    SeedSequence,
-)
-
-seq = SeedSequence(1030424547444117993331016959)
-pcg = PCG64(seq)
-gen = Generator(pcg)
-\end{lstlisting}
-
-\noindent This approach retains access to the seed sequence which can then be
+%
+%\begin{lstlisting}
+%from numpy.random import (
+%    Generator,
+%    PCG64,
+%    SeedSequence,
+%)
+%
+%seq = SeedSequence(1030424547444117993331016959)
+%pcg = PCG64(seq)
+%gen = Generator(pcg)
+%\end{lstlisting}
+%
+This approach retains access to the seed sequence which can then be
 used to spawn additional generators.
-
-\begin{lstlisting}
-children = seq.spawn(2)
-gen_0 = Generator(PCG64(children[0]))
-gen_1 = Generator(PCG64(children[1]))
-\end{lstlisting}
-
-\noindent While this approach retains complete flexibility, the method
+%
+%\begin{lstlisting}
+%children = seq.spawn(2)
+%gen_0 = Generator(PCG64(children[0]))
+%gen_1 = Generator(PCG64(children[1]))
+%\end{lstlisting}
+%
+While this approach retains complete flexibility, the method
 \code{np.random.default\_rng} can be used to instantiate a \code{Generator} when

### 2020-01-06 body.tex
Switch figures

@@
 across the larger array to a common shape when sensible.
+For example, if \code{x} is a 1-d array with length 3, and \code{y} is a 2-d
+array with shape $2 \times 1$, the output will be an array with shape $2 \times 3$
+(see Figure~\ref{fig:broadcasting}).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{static/broadcasting}
+  \caption{
+    Broadcasting a $1 \times 3$ array against a $2 \times 1$ array
+    yields a $2 \times 3$ result.
+  }
+  \label{fig:broadcasting}
+\end{figure}
+
 These vectorized operations often produce one-liners that would be many tens
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
-    \caption{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012. The total number of PRs is indicated with the
-        lower blue area showing the portion contributed by current or previous
-        maintainers.}\label{fig:prs-over-time}
-\end{figure}
-
-NumPy is currently maintained by a group of 23 contributors with commit rights
-to the NumPy code base. Out of these, 17 maintainers were active in
-2019, 4 of whom were paid to work on the project full-time.
-Additionally, there are a few long term developers who contributed and maintain
-specific parts of NumPy, but are not officially maintainers.
-At a release cycle of about every half year, the five recent releases in the years
-2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
-    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
-    All data provided here is based on the development which happened through GitHub
-    pull requests. In some cases contributions by maintainers may not be categorized as such.}
-% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
-with each release attracting more than a hundred new contributors.
-Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
-master branch.
-Although the number of PRs being merged fluctuates,
-the plot indicates an increased number of contributions over the past
-years.
+%\begin{figure}
+%    \centering
+%    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
+%    \caption{Number of pull requests merged into the NumPy master branch for each
+%        quarter since 2012. The total number of PRs is indicated with the
+%        lower blue area showing the portion contributed by current or previous
+%        maintainers.}\label{fig:prs-over-time}
+%\end{figure}
+%
+%NumPy is currently maintained by a group of 23 contributors with commit rights
+%to the NumPy code base. Out of these, 17 maintainers were active in
+%2019, 4 of whom were paid to work on the project full-time.
+%Additionally, there are a few long term developers who contributed and maintain
+%specific parts of NumPy, but are not officially maintainers.
+%At a release cycle of about every half year, the five recent releases in the years
+%2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
+%    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
+%    All data provided here is based on the development which happened through GitHub
+%    pull requests. In some cases contributions by maintainers may not be categorized as such.}
+%% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
+%with each release attracting more than a hundred new contributors.
+%Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
+%master branch.
+%Although the number of PRs being merged fluctuates,
+%the plot indicates an increased number of contributions over the past
+%years.
 

### 2020-01-06 body.tex
Some correctness and readability fixes

@@
 notably those by Mark Wiebe when he was sponsored by Travis Oliphant
-at Continuum Analytics, the package's rapid adoption in scientific
+at Enthought, the package's rapid adoption in scientific
 computing took place without external investment.
@@
 contributors.
-Finally, the project receives around 10K USD annually from Tidelift, which is
-typically used to fund documentation writers and website developers.
+Finally, since May 2019 the project receives around 10K USD annually from
+Tidelift, which is typically used to fund documentation and website
+improvements.
 
@@
 libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
-Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
-and Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}}.
+JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
+and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}. 
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
+and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
+provide tensor APIs with NumPy-inspired semantics.
 It is also implemented in packages that support sparse arrays
@@
 
-\section*{Random}
+\section*{Random number generation}
 
-The NumPy random module provides pseudorandom numbers from a wide range of
+The NumPy \code{random} module provides pseudorandom numbers from a wide range of
 distributions. In legacy versions of NumPy, simulated random values are produced
 by a \code{RandomState} object that: handles seeding and state initialization;
-wraps the core pseudorandom number generator based on a 32-bit implementation of
-MT19937; interfaces with the underlying code that transforms random bits into
+wraps the core pseudorandom number generator based on a Mersenne Twister
+implementation\footnote{to be precise, the standard 32-bit version of MT19937};
+interfaces with the underlying code that transforms random bits into
 deviates from other distributions; and supplies a singleton instance exposed in
@@
 generators (\code{RDRAND})\cite{randomgen}.} Bit generators export a capsule
-containing pointers to functions that produce 64 bits, 32 bits, or a random
-double. They also expose CFFI and ctypes interfaces to the same three functions.
+containing pointers to functions that produce
+FIXME this doesn't make sense (it's the uniform stream of numbers that matters
+rather that "capsule of pointers", also what's "a random double"?):
+64 bits, 32 bits, or a random double.
+They also expose CFFI and ctypes interfaces to the same three functions.
 

### 2020-01-06 body.tex
Adjust random section minimally to address Ralf's comment

@@
 generators (\code{AESCtr}) and generators that expose hardware random number
-generators (\code{RDRAND})\cite{randomgen}.} Bit generators export a capsule
-containing pointers to functions that produce
-FIXME this doesn't make sense (it's the uniform stream of numbers that matters
-rather that "capsule of pointers", also what's "a random double"?):
-64 bits, 32 bits, or a random double.
-They also expose CFFI and ctypes interfaces to the same three functions.
+generators (\code{RDRAND})\cite{randomgen}.} Bit generators provide
+functions, exposed both in Python and C, for generating random integer
+and floating point numbers.
 

### 2020-01-06 body.tex
Remove additional sections

@@
 
-% BIDS -- UCB
-% https://www.moore.org/grant-detail?grantId=GBMF5447
-% $645,020 in 2016
-% https://sloan.org/grant-detail/8222
-% $659,359 in 2017
-% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
-% http://doi.org/10.5281/zenodo.3585761
-% http://doi.org/10.5281/zenodo.3585767
-In 2017, NumPy received its first large grants totaling 1.3M USD from the
-Gordon \& Betty Moore and the Alfred P. Sloan foundations.
-These two grants focus on addressing the technical debt accrued over the years and
-on setting in place standards and architecture to encourage more sustainable development.
-% CZI -- NumFOCUS/QuanSight
-% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
-% NumPy and OpenBLAS received $195,000 in 2019
-% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
-NumPy received a third grant for 195K USD from the Chan Zuckerberg
-Initiative at the end of 2019.
-This grant focuses on better serving NumPy's large number of beginning
-to intermediate level users and on growing the community of NumPy
-contributors.
-Finally, since May 2019 the project receives around 10K USD annually from
-Tidelift, which is typically used to fund documentation and website
-improvements.
+In 2018, we hired our first full-time developers ...
 
@@
 
-%\begin{figure}
-%    \centering
-%    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
-%    \caption{Number of pull requests merged into the NumPy master branch for each
-%        quarter since 2012. The total number of PRs is indicated with the
-%        lower blue area showing the portion contributed by current or previous
-%        maintainers.}\label{fig:prs-over-time}
-%\end{figure}
-%
-%NumPy is currently maintained by a group of 23 contributors with commit rights
-%to the NumPy code base. Out of these, 17 maintainers were active in
-%2019, 4 of whom were paid to work on the project full-time.
-%Additionally, there are a few long term developers who contributed and maintain
-%specific parts of NumPy, but are not officially maintainers.
-%At a release cycle of about every half year, the five recent releases in the years
-%2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
-%    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
-%    All data provided here is based on the development which happened through GitHub
-%    pull requests. In some cases contributions by maintainers may not be categorized as such.}
-%% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
-%with each release attracting more than a hundred new contributors.
-%Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
-%master branch.
-%Although the number of PRs being merged fluctuates,
-%the plot indicates an increased number of contributions over the past
-%years.
-
 We briefly highlight two of those below.
 
-\section*{Array function protocol}
+- array function protocol
 
-A vast number of projects are built on NumPy;
-these projects are consumers of the NumPy API.
-Over the last several years, a growing number of projects are providers of
-a \emph{NumPy-like API} and array objects targeting audiences with specialized
-needs beyond NumPy's capabilities.
-For example, the NumPy API is implemented by several popular tensor computation
-libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
-JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
-and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}. 
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
-and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
-provide tensor APIs with NumPy-inspired semantics.
-It is also implemented in packages that support sparse arrays
-such as \code{scipy.sparse} and \code{pydata.sparse}.
-Another notable example is Dask, a library for parallel computing in
-Python.  Dask adopts the NumPy API and therefore presents a familiar
-interface to existing NumPy users, while adding powerful abilities to
-parallelize and distribute tasks.
+- random number generation
 
-The multitude of specialized projects creates the difficulty that consumers
-of these NumPy-like APIs write code specific to a single project and do not support
-all of the above array providers.
-This is a burden for users relying on the specialized array-like, since
-a tool they need may not work for them.
-It also creates new challenges for end-users who need to transition
-from NumPy to a more specialized array.
-The growing multitude of specialized projects with NumPy-like APIs threatened
-to again fracture the scientific Python community.
+Ongoing work
 
-To address these issues NumPy has the goal of providing the fundamental
-API for \emph{interoperability} between the various NumPy-like APIs.
-An earlier step in this direction was the implementation of the
-\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
-for most mathematical functions.\cite{NEP13}
-In 2019 this was expanded more generally with the inclusion of the
-\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
-These two protocols allows providers of array-like APIs to be interoperable
-with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
-For the users relying on specialized array projects it means that even though
-much code is written specifically for NumPy arrays and uses the NumPy API as
-\code{import numpy as np}, it can nevertheless work for them.
-%For example, here is how a CuPy GPU array can be passed through NumPy for
-%processing, with all operations being dispatched back to CuPy:
-%
-%\begin{lstlisting}
-%import numpy as np
-%import cupy as cp
-%
-%x_gpu = cp.array([1, 2, 3])
-%y = np.sum(x_gpu)  # Returns a GPU array
-%\end{lstlisting}
-
-\section*{Random number generation}
-
-The NumPy \code{random} module provides pseudorandom numbers from a wide range of
-distributions. In legacy versions of NumPy, simulated random values are produced
-by a \code{RandomState} object that: handles seeding and state initialization;
-wraps the core pseudorandom number generator based on a Mersenne Twister
-implementation\footnote{to be precise, the standard 32-bit version of MT19937};
-interfaces with the underlying code that transforms random bits into
-deviates from other distributions; and supplies a singleton instance exposed in
-the root of the random module.
-
-The \code{RandomState} object makes a compatibility guarantee so that a fixed
-seed and sequence of function calls produce the same set of values. This
-guarantee has slowed progress since improving the underlying code requires
-extending the API with additional keyword arguments. This guarantee continues to
-apply to \code{RandomState}.
-
-NumPy 1.17 introduced a new API for generating random numbers that use a more
-flexible structure that can be extended by libraries or end-users. The new API
-is built using components that separate the steps required to generate random
-variates. Pseudorandom bits are generated by a bit generator. These bits are
-then transformed into variates from complex distributions by a generator.
-Finally, seeding is handled by an object that produces sequences of high-quality
-initial values.
-
-Bit generators are simple classes that manage the state of an underlying
-pseudorandom number generator. NumPy ships with four bit generators. The default
-bit generator is a 64-bit implementation of the Permuted Congruential Generator
-\cite{pcg64} (\code{PCG64}). The three other bit generators are a 64-bit version
-of the Philox generator\cite{random123} (\code{Philox}), Chris Doty-Humphrey's
-Small Fast Chaotic generator\cite{practrand} (\code{SFC64}), and the 32-bit
-Mersenne Twister\cite{mt19937} (\code{MT19937}) which has been used in older
-versions of NumPy.\footnote{The
-\href{https://github.com/bashtage/randomgen}{randomgen project} supplies a wide
-range of alternative bit generators such as a cryptographic counter-based
-generators (\code{AESCtr}) and generators that expose hardware random number
-generators (\code{RDRAND})\cite{randomgen}.} Bit generators provide
-functions, exposed both in Python and C, for generating random integer
-and floating point numbers.
-
-The \code{Generator} consumes one of the bit generators and produces variates
-from complicated distributions. Many improved methods for generating random
-variates from common distributions were implemented, including the Ziggurat
-method for normal, exponential and gamma variates\cite{ziggurat}, and Lemire's
-method for bounded random integer generation\cite{lemire}. The \code{Generator}
-is more similar to the legacy \code{RandomState}, and its API is substantially
-the same. The key differences all relate to state management, which has been
-delegated to the bit generator. The \code{Generator} does not make the same
-stream guarantee as the \code{RandomState} object, and so variates may differ
-across versions as improved generation algorithms are
-introduced.\footnote{Despite the removal of the compatibility guarantee, simple
-reproducibility across versions is encouraged, and minor changes that do not
-produce meaningful performance gains or fix underlying bug are not generally
-adopted.}
-
-Finally, a \code{SeedSequence} is used to initialize a bit generator. The seed
-sequence can be initialized with no arguments, in which case it reads entropy
-from a system-dependent provider, or with a user-provided seed. The seed
-sequence then transforms the initial set of entropy into a sequence of
-high-quality pseudorandom integers, which can be used to initialize multiple bit
-generators deterministically. A key design goal of a seed sequence was to be
-splittable in the sense that a seed sequence can be used to produce child
-sequences that are distinct from their parents or other ancestors. This
-capability allows a seed sequence to be used in large distributed applications
-where the number of workers required is not known. The sequences generated from
-the same initial entropy and the same splits are fully deterministic to ensure
-reproducibility.
-
-The three components are combined to construct a complete random number
-generator.
-%
-%\begin{lstlisting}
-%from numpy.random import (
-%    Generator,
-%    PCG64,
-%    SeedSequence,
-%)
-%
-%seq = SeedSequence(1030424547444117993331016959)
-%pcg = PCG64(seq)
-%gen = Generator(pcg)
-%\end{lstlisting}
-%
-This approach retains access to the seed sequence which can then be
-used to spawn additional generators.
-%
-%\begin{lstlisting}
-%children = seq.spawn(2)
-%gen_0 = Generator(PCG64(children[0]))
-%gen_1 = Generator(PCG64(children[1]))
-%\end{lstlisting}
-%
-While this approach retains complete flexibility, the method
-\code{np.random.default\_rng} can be used to instantiate a \code{Generator} when
-reproducibility is not needed.
-
-The final goal of the new API is to improve extensibility. \code{RandomState} is
-a monolithic object that obscures all of the underlying state and functions. The
-component architecture is one part of the extensibility improvements. The
-underlying functions (written in C) which transform the output of a bit
-generator to other distributions are available for use in CFFI. This allows the
-same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
-Numba. Both the bit generators and the low-level functions can also be used in C
-or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
-NumPy source. Alternative approaches that avoid this extra step are being
-explored.}
+- dtype
 
 \section*{Discussion}
-
-% 1st grant done, technical cleanup finished except dtypes
-% 2nd grant starting, community growth and new user focus
-
-Most of the technical work for the Moore and Sloan grant is now complete.
-The main remaining item is an overhaul of the data type system, which
-we are currently working on.
-The meaning of each element within a NumPy array is described by its
-datatype (dtype). NumPy arrays can hold all common numerical
-datatypes, strings, datetimes, and generic Python objects, each of
-which is identified by a dtype.
-%We are busy overhauling the datatype
-%system to make it behave consistently and to simplify the creation of
-%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
-%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
-While numerical dtypes make the foundation for most users,
-there are a multitude of use cases which cannot prosper due to current
-limitations, such as physical units\cite{astropy,Goldbaum2018,pint},
-% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
-geometrical objects\cite{pygeos}, and automatic
-differentiation\cite{pyadolc}.
-%Projects addressing these exist, each struggles with current
-%limitations.  % TODO: I may need citations, we could link github issues.
-The proposed overhaul of the datatype system will make it behave consistently and
-to simplify the creation of custom dtypes, both in C and in Python
-
-%% While it is currently possible to define custom dtypes within NumPy, many of the
-%% use cases above cannot be solved using this system.
-%% This is evidenced also by code within NumPy which would require changes
-%% in many places to add such dtypes, even when adding them within NumPy itself.
-%% The new implementation will be simpler and more cleanly implemented in
-%% this regard.
-%% Custom dtypes and the basic dtypes included within NumPy will be implemented
-%% and behave identically as much as possible.
-%% At the same time redesigning the API will make NumPy dtypes more flexible to grow
-%% to future needs should they arise.
-
-%% As a further step, to make dtypes more accessible and allow a faster adoption of
-%% custom dtypes within the community, we plan to create a Python API to
-%% define such custom dtypes.
-
-Having made progress on addressing technical debt accrued over two decades of
-development and modernizing the code to handle new use cases, the project
-is ready to grow its community of contributors and to better address the needs
-of an ever-growing number of new users.  This is the main focus of the
-forthcoming CZI grant.
-
-[... CZI grant ...]
-
-[... welcome new users ...]
-
-% technical focus
-% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays).
-% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.
-
-% \subsubsection*{Future Development}
-
-% Dtypes
-
-% outline plan for the CZI grant and an invitation to new contributors
-
-%\subsection*{Website}

### 2020-01-06 body.tex
Add methods section

@@
 
+The random number generator library in NumPy provides several alternative
+\emph{bit stream generators} that provide the core function of generating
+random integers.
+A higher-level generator class that implements an assortment of
+probability distributions is provided. It includes the beta, gamma
+and Weibull distributions, the univariate and multivariate normal
+distributions, and more.
+
 Ongoing work

### 2020-01-06 methods.tex
Add methods section

\section*{Methods}

Over the course of its history, NumPy has attracted PRs by 823 contributors.
However, its development relies heavily on a relatively small number
of active maintainers, who share more than half of the contributions among
themselves.

% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
% https://github.com/numpy/numpy/pull/6352
To formalize its unwritten development policy,
NumPy adopted an official Governance Document on October~5,
2015\cite{NumPyProjectGovernance}.
Project decisions are usually made by consensus of interested contributors.
This means for many thing everyone is entrusted with veto power.
A Steering Council, currently composed of 12~members, facilitates this
process and oversees daily development of the project by contributing code
and reviewing contributions from the community.

% https://mail.python.org/pipermail/numpy-discussion/2018-July/078476.html
% https://github.com/numpy/numpy/pull/11865
NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
In brief, we strive to:
\emph{be open};
\emph{be empathetic, welcoming, friendly, and patient};
\emph{be collaborative};
\emph{be inquisitive}; and
\emph{be careful in the words that we choose}.
The Code of Conduct also specifies how breaches can be reported and outlines
the process for responding to such reports.

NumPy is currently maintained by a group of 23 contributors with commit rights
to the NumPy code base. Out of these, 17 maintainers were active in
2019, 4 of whom were paid to work on the project full-time.
Additionally, there are a few long term developers who contributed and maintain
specific parts of NumPy, but are not officially maintainers.

At a release cycle of about every half year, the five recent releases in the years
2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
    All data provided here is based on the development which happened through GitHub
    pull requests. In some cases contributions by maintainers may not be categorized as such.}
% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
with each release attracting more than a hundred new contributors.
Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
master branch.
Although the number of PRs being merged fluctuates,
the plot indicates an increased number of contributions over the past
years.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
    \caption{Number of pull requests merged into the NumPy master branch for each
        quarter since 2012. The total number of PRs is indicated with the
        lower blue area showing the portion contributed by current or previous
        maintainers.}\label{fig:prs-over-time}
\end{figure}

\subsection*{Funding}

% BIDS -- UCB
% https://www.moore.org/grant-detail?grantId=GBMF5447
% $645,020 in 2016
% https://sloan.org/grant-detail/8222
% $659,359 in 2017
% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
% http://doi.org/10.5281/zenodo.3585761
% http://doi.org/10.5281/zenodo.3585767
In 2017, NumPy received its first large grants totaling 1.3M USD from the
Gordon \& Betty Moore and the Alfred P. Sloan foundations.
Stéfan van der Walt is the PI and manages four programmers working on the project.
These two grants focus on addressing the technical debt accrued over the years and
on setting in place standards and architecture to encourage more sustainable development.

% CZI -- NumFOCUS/QuanSight
% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
% NumPy and OpenBLAS received $195,000 in 2019
% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
NumPy received a third grant for 195K USD from the Chan Zuckerberg
Initiative at the end of 2019 with Ralf Gommers as the PI.
This grant focuses on better serving NumPy's large number of beginning
to intermediate level users and on growing the community of NumPy
contributors.
It will also provide support to OpenBLAS, on which NumPy depends for
accelerated linear algebra.

Finally, since May 2019 the project receives around 10K USD annually from
Tidelift, which is typically used to fund documentation and website
improvements.

The first activities organized for the new grant were a NEP development
sprint, a weekly community call, and a meeting to write a draft roadmap.

Weekly community calls alternate between triage and
higher level discussion.  The calls not only involve developers from
the community, but provide a venue for vendors and other external
groups to provide input.  For example, after Intel produced a forked
version of NumPy, one of their developers joined a call to discuss
community concerns.

NumPy plays a central role in building and standardizing much of the scientific
Python community infrastructure.
The docstring standard mentioned in the history section is now widely adopted.
We are also now using the NEP system as a way to help coordinate the larger
scientific Python community.
% https://numpy.org/neps/nep-0029-deprecation_policy.html
For example, in NEP 29, we recommend, along with leaders from various other
projects, that all projects across the Scientific Python ecosystem adopt a
common ``time window-based'' policy for support of Python and NumPy versions.
This standard will simplify downstream project and release planning.

\subsection*{Wheels build system}

A Python \emph{wheel}\cite{PEP427} is a standard file format for
distributing Python libraries.  In addition to Python code, a
wheel may include compiled C extensions and other binary data.
This is important, because many libraries, including NumPy,
require a C compiler and other build tools to build the software
from the source code, making it difficult for many users to install
the software on their own.  The introduction of wheels to the Python
packaging system has made it much easier for users to install
precompiled libraries.

A GitHub repository containing scripts to build NumPy wheels has
been configured so that a simple commit to the repository triggers
an automated build system that creates NumPy wheels for several
computer platforms, including Windows, Mac OSX and Linux.  The wheels
are uploaded to a public server and made available for anyone to use.
This system makes it easy for users to install precompiled versions
of NumPy on these platforms.

The technology that is used to build the wheels evolves continually.
At the time this paper is being written, a key component is the
\code{multibuild} suite of tools developed by Matthew Brett and
other developers\cite{multibuild}.  Currently, scripts using
\code{multibuild} are written for the continuous integration
platforms Travis-CI (for Linux and Mac OSX) and Appveyor
(for Windows).

\subsection*{Recent technical improvements}

Given the complexity of the codebase and the massive number of projects depending
on it, large changes require careful planning and substantial work.
As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
of enhancement proposals---used infrequently and informally until 2017 when,
to facilitate work expected from funded development, he formalized the
NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
Python Enhancement Proposals (PEPs) for ``proposing major new
features, for collecting community input on an issue, and for
documenting the design decisions that have gone into
Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
Since then there have been 19 proposed NEPS---6 have been implemented,
4 have been accepted and are being implemented, 4 are under
consideration, 3 have been deferred or superseded, and 2 have been rejected
or withdrawn.

With the recent infusion of funding and a clear process for coordinating with
the developer community, we have been able to tackle a number of important
large scale changes.
We highlight two of those below, as well as changes made to our testing
infrastructure to support hardware platforms used in large scale computing.

\subsection*{Array function protocol}

A vast number of projects are built on NumPy;
these projects are consumers of the NumPy API.
Over the last several years, a growing number of projects are providers of
a \emph{NumPy-like API} and array objects targeting audiences with specialized
needs beyond NumPy's capabilities.
For example, the NumPy API is implemented by several popular tensor computation
libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
provide tensor APIs with NumPy-inspired semantics.
It is also implemented in packages that support sparse arrays
such as \code{scipy.sparse} and \code{pydata.sparse}.
Another notable example is Dask, a library for parallel computing in
Python.  Dask adopts the NumPy API and therefore presents a familiar
interface to existing NumPy users, while adding powerful abilities to
parallelize and distribute tasks.

The multitude of specialized projects creates the difficulty that consumers
of these NumPy-like APIs write code specific to a single project and do not support
all of the above array providers.
This is a burden for users relying on the specialized array-like, since
a tool they need may not work for them.
It also creates new challenges for end-users who need to transition
from NumPy to a more specialized array.
The growing multitude of specialized projects with NumPy-like APIs threatened
to again fracture the scientific Python community.

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
An earlier step in this direction was the implementation of the
\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
for most mathematical functions.\cite{NEP13}
In 2019 this was expanded more generally with the inclusion of the
\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
These two protocols allows providers of array-like APIs to be interoperable
with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
For the users relying on specialized array projects it means that even though
much code is written specifically for NumPy arrays and uses the NumPy API as
\code{import numpy as np}, it can nevertheless work for them.
For example, here is how a CuPy GPU array can be passed through NumPy for
processing, with all operations being dispatched back to CuPy:

\begin{lstlisting}
import numpy as np
import cupy as cp

x_gpu = cp.array([1, 2, 3])
y = np.sum(x_gpu)  # Returns a GPU array
\end{lstlisting}

Similarly, user defined functions composed using NumPy can now be
applied to, e.g., multi-node distributed Dask arrays:

\begin{lstlisting}
import numpy as np
import dask.array as da

def f(x):
    """Function using NumPy API calls"""
    y = np.tensordot(x, x.T)
    return np.mean(np.log(y + 1))

x_local = np.random.random([10000, 10000])  # random local array
x_distr = da.random.random([10000, 10000])  # random distributed array

f(x_local)  # returns a NumPy array
f(x_distr)  # works, returns a Dask array
\end{lstlisting}

\subsection*{Random number generation}

The NumPy \code{random} module provides pseudorandom numbers from a wide range of
distributions. In legacy versions of NumPy, simulated random values are produced
by a \code{RandomState} object that: handles seeding and state initialization;
wraps the core pseudorandom number generator based on a Mersenne Twister
implementation\footnote{to be precise, the standard 32-bit version of MT19937};
interfaces with the underlying code that transforms random bits into
deviates from other distributions; and supplies a singleton instance exposed in
the root of the random module.

The \code{RandomState} object makes a compatibility guarantee so that a fixed
seed and sequence of function calls produce the same set of values. This
guarantee has slowed progress since improving the underlying code requires
extending the API with additional keyword arguments. This guarantee continues to
apply to \code{RandomState}.

NumPy 1.17 introduced a new API for generating random numbers that use a more
flexible structure that can be extended by libraries or end-users. The new API
is built using components that separate the steps required to generate random
variates. Pseudorandom bits are generated by a bit generator. These bits are
then transformed into variates from complex distributions by a generator.
Finally, seeding is handled by an object that produces sequences of high-quality
initial values.

Bit generators are simple classes that manage the state of an underlying
pseudorandom number generator. NumPy ships with four bit generators. The default
bit generator is a 64-bit implementation of the Permuted Congruential Generator
\cite{pcg64} (\code{PCG64}). The three other bit generators are a 64-bit version
of the Philox generator\cite{random123} (\code{Philox}), Chris Doty-Humphrey's
Small Fast Chaotic generator\cite{practrand} (\code{SFC64}), and the 32-bit
Mersenne Twister\cite{mt19937} (\code{MT19937}) which has been used in older
versions of NumPy.\footnote{The
\href{https://github.com/bashtage/randomgen}{randomgen project} supplies a wide
range of alternative bit generators such as a cryptographic counter-based
generators (\code{AESCtr}) and generators that expose hardware random number
generators (\code{RDRAND})\cite{randomgen}.} Bit generators provide
functions, exposed both in Python and C, for generating random integer
and floating point numbers.

The \code{Generator} consumes one of the bit generators and produces variates
from complicated distributions. Many improved methods for generating random
variates from common distributions were implemented, including the Ziggurat
method for normal, exponential and gamma variates\cite{ziggurat}, and Lemire's
method for bounded random integer generation\cite{lemire}. The \code{Generator}
is more similar to the legacy \code{RandomState}, and its API is substantially
the same. The key differences all relate to state management, which has been
delegated to the bit generator. The \code{Generator} does not make the same
stream guarantee as the \code{RandomState} object, and so variates may differ
across versions as improved generation algorithms are
introduced.\footnote{Despite the removal of the compatibility guarantee, simple
reproducibility across versions is encouraged, and minor changes that do not
produce meaningful performance gains or fix underlying bug are not generally
adopted.}

Finally, a \code{SeedSequence} is used to initialize a bit generator. The seed
sequence can be initialized with no arguments, in which case it reads entropy
from a system-dependent provider, or with a user-provided seed. The seed
sequence then transforms the initial set of entropy into a sequence of
high-quality pseudorandom integers, which can be used to initialize multiple bit
generators deterministically. A key design goal of a seed sequence was to be
splittable in the sense that a seed sequence can be used to produce child
sequences that are distinct from their parents or other ancestors. This
capability allows a seed sequence to be used in large distributed applications
where the number of workers required is not known. The sequences generated from
the same initial entropy and the same splits are fully deterministic to ensure
reproducibility.

The three components are combined to construct a complete random number
generator.

\begin{lstlisting}
from numpy.random import (
    Generator,
    PCG64,
    SeedSequence,
)

seq = SeedSequence(1030424547444117993331016959)
pcg = PCG64(seq)
gen = Generator(pcg)
\end{lstlisting}

This approach retains access to the seed sequence which can then be
used to spawn additional generators.

\begin{lstlisting}
children = seq.spawn(2)
gen_0 = Generator(PCG64(children[0]))
gen_1 = Generator(PCG64(children[1]))
\end{lstlisting}

While this approach retains complete flexibility, the method
\code{np.random.default\_rng} can be used to instantiate a \code{Generator} when
reproducibility is not needed.

The final goal of the new API is to improve extensibility. \code{RandomState} is
a monolithic object that obscures all of the underlying state and functions. The
component architecture is one part of the extensibility improvements. The
underlying functions (written in C) which transform the output of a bit
generator to other distributions are available for use in CFFI. This allows the
same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
Numba. Both the bit generators and the low-level functions can also be used in C
or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
NumPy source. Alternative approaches that avoid this extra step are being
explored.}

\subsection*{Testing on multiple architectures}

At the time of writing the two fastest supercomputers in the
world, Summit and Sierra, both have IBM POWER9 architectures
\cite{top500nov2019}. In late 2018, Astra, the first ARM-based
supercomputer to enter the TOP500 list, went into production\cite{
astra-wiki}. Furthermore, over 100 billion ARM processors have been
produced as of 2017\cite{arm-architecture}, making it the most 
widely used instruction set architecture in the world.

Clearly there are motivations for a large scientific computing
software library to support POWER and ARM architectures. We've extended
our continuous integration (CI) testing to include \texttt{ppc64le}
(POWER8 on Travis CI) and ARMv8 (on Shippable service). We also test
with the s390x architecture (IBM Z CPUs on Travis CI) so that we
can probe the behavior of our library on a big-endian machine.
This satisfies one of the major components of
improved CI testing laid out in a version of our roadmap
\cite{numpy-roadmap}---specifically, ``CI for more exotic
platforms."

PEP 599\cite{PEP599} lays out a plan for new Python binary wheel
distribution support, \texttt{manylinux2014}, that adds
support for a number of architectures supported by the CentOS
Alternative Architecture Special Interest Group, including
ARMv8, ppc64le, as well as s390x. We are thus well-positioned
for a future where provision of binaries on these architectures
will be expected for a library at the base of the ecosystem.

\subsection*{Dtype}

The meaning of each element within a NumPy array is described by its
datatype (dtype). NumPy arrays can hold all common numerical
datatypes, strings, datetimes, and generic Python objects, each of
which is identified by a dtype.
%We are busy overhauling the datatype
%system to make it behave consistently and to simplify the creation of
%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
While numerical dtypes make the foundation for most users,
there are a multitude of use cases which cannot prosper due to current
limitations, such as physical units\cite{astropy,Goldbaum2018,pint},
% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
geometrical objects\cite{pygeos}, and automatic
differentiation\cite{pyadolc}.
%Projects addressing these exist, each struggles with current
%limitations.  % TODO: I may need citations, we could link github issues.
The proposed overhaul of the datatype system will make it behave consistently and
to simplify the creation of custom dtypes, both in C and in Python

### 2020-01-06 body.tex
Added notes and small edits

@@
 NumPy is both a cause and an effect of a striking shift in the nature of
-scientific computing.  The traditional model of scientific software has been of
-the user as {\it consumer}.  Writing software is a task for specialists and
-professionals, supported by income from selling the software to scientists to
-use.  The new model is of the user as {\it owner}.  The Python language, with
-the NumPy library, provides a framework that scientists can consume, but has
-also been successful in encouraging scientists to develop as programmers, and
-to write code for the library, and the system of packages around it.  This has
+scientific computing.  In the traditional scientific software model, the user is a {\it consumer}.  Writing software is a task for specialists and
+professionals, supported by income from selling the software to
+scientists.  In the new model, the user is an {\it owner}.  The Python language, with
+the NumPy library, provides a framework that scientists can consume,
+but also encourages scientists to participate as programmers---to
+write code for the library and the ecosystem of packages around it.  This has
 fostered a new culture of contribution and collaboration in scientific
@@
 standard fast implementations of linear algebra.
+% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
+% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
+% May want to mention Yorick here (origin of broadcasting idea)
 
@@
 interactive computing with numerical arrays were already well-established,
-including Matlab and IDL.  At the time, it must have seemed a safe bet that
+including Matlab and IDL.  Back then, it must have seemed a safe bet that
 Numeric would not succeed in this market.  The commercial languages were
-developed and maintained by large numbers of full-time professional programmers
-and supported by formal quality assurance, release managers and writers of
-scientific documentation.  In contrast, Numeric, and later, NumPy, would
+developed and maintained by large groups of full-time professional programmers,
+supported by formal quality assurance, release managers, and writers of
+scientific documentation.  In contrast, Numeric, and later NumPy, would
 continue to be developed almost entirely by scientists, including graduate
-students and faculty, without specific funding, and often to the detriment of
-the scientists' careers.
+students and faculty, without specific funding and often to the detriment of their careers.
+% S: The "Back then" sentence sets up an expectation.
+Later in this section, we then explain that Numeric/NumPy *was* successful, but we never explicitly close the loop: "despite the odds, great success".
 
@@
 
+% S TODO: the first sentence below could flow a bit better
 Around 1998, the Space Telescope Science Institute (STScI) software group began
-to use Python heavily, and, around 2000, wrote a re-implementation of much of
+to use Python heavily and, around 2000, wrote a re-implementation of much of
 Numeric, called NumArray, to support their need to work on large memory-mapped
-arrays, and arrays of mixed data type records \cite{STScI-slither}.  This
+arrays and arrays of mixed data type records \cite{STScI-slither}.  This
 briefly caused the Numeric and NumArray communities to diverge, until 2005,
@@
 \cite{millman2014developing}. The NumPy team was early in adopting distributed
-revision control and code review to improved collaboration on code, and
-continuous testing so we run a large battery automated tests for every proposed
-change in NumPy. At first, NumPy had rather sparse documentation, but the project
+revision control and code review to improve collaboration on code, and
+continuous testing that runs a large battery of automated tests for every proposed
+change to NumPy. At first, NumPy had rather sparse documentation, but the project
 developed a new standard for ``docstrings''---text describing each NumPy
-function---and used this standard to write comprehensive high-quality
-documentation, integrated with the source code. The NumPy documentation format
+function---and used this standard to write comprehensive, high-quality
+documentation, integrated with the source code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}. The NumPy documentation format
 quickly became a standard for the larger Python community.
@@
 package \cite{Hugunin-whitepaper}.  Numerical computing is usually one
-component of a larger programming task.  Therefore:
+component of a larger programming task, therefore:
 
@@
 numerical languages can encourage consumer-users, who pick up a tool and use
-it, but do not expect to contribute to development.  In contrast, NumPy often
+it, but do not expect to contribute to development%<-sentence a bit tough to parse.
+In contrast, NumPy often
 attracts user-developers, who want to work with NumPy, and are prepared to fix
@@
 around NumPy and Python.
+
+% S: I think two issues are conflated here: the simple underlying memory model, and the API.
+The API has developed organically, while the underlying memory model is a design decision that shapes much of how the rest of the library was built.
+
+% S:
+%
+% Thinking about enabling factors, I'd characterize them into three categories:
+%
+% - Practical
+% - Philosophical
+% - Social
+%
+% E.g., practical: reason why we use Python (learn one language for everything); students don't have money, so want to avoid impracticalities of license dongles/servers, etc.
+%
+% Philosophical: science should be open, transparent; our software should be controlled by scientists, not designers that we don't have access to
+%
+% Social: joy of building these things together, friendly welcome into the community for many of us---appreciation of our work and hours
+%
 

### 2020-01-06 summary.tex
Added notes and small edits

@@
 Software libraries for array computation provide powerful notation
-for accessing, manipulating, and operating on aggregate data in vectors,
-matrices, or higher-dimensional arrays \cite{iverson1980notation}.
+for accessing, manipulating, and operating on data in vectors,
+matrices, and higher-dimensional arrays \cite{iverson1980notation}.
+% Stéfan: why "aggregate" data?
 % - a brief account of the background and rationale of the work;
@@
 Not surprisingly, many important scientific research projects and results
-crucially rely on NumPy.
-It plays an essential role in many research analysis pipelines in fields as
+depend on NumPy.
+It plays an essential role in research analysis pipelines in fields as
 diverse as physics, biology, astronomy, neuroscience, material science,
 engineering, and chemistry.
-In particular, NumPy played a fundamental role in discovering gravitational
-waves\cite{pycbc} and the first imaging of a black hole\cite{eht-imaging}, and
-lets us explore our universe\cite{jenness2018lsst}.
+For example, in astronomy, NumPy was fundamental to discovering gravitational
+waves\cite{pycbc}, the first imaging of a black
+hole\cite{eht-imaging}, and continues underpinning the future exploration of our universe\cite{jenness2018lsst}.
 It is also one of the most commonly used machine learning tools among

### 2020-01-07 body.tex
Organize methods section

@@
 
+The meaning of each element within a NumPy array is described by its
+datatype (dtype). NumPy arrays can hold all common numerical
+datatypes, strings, datetimes, and generic Python objects, each of
+which is identified by a dtype.
+%We are busy overhauling the datatype
+%system to make it behave consistently and to simplify the creation of
+%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
+%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
+While numerical dtypes make the foundation for most users,
+there are a multitude of use cases which cannot prosper due to current
+limitations, such as physical units\cite{astropy,Goldbaum2018,pint},
+% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
+geometrical objects\cite{pygeos}, and automatic
+differentiation\cite{pyadolc}.
+%Projects addressing these exist, each struggles with current
+%limitations.  % TODO: I may need citations, we could link github issues.
+The proposed overhaul of the datatype system will make it behave consistently and
+to simplify the creation of custom dtypes, both in C and in Python.
+
 \section*{Discussion}

### 2020-01-07 methods.tex
Organize methods section

@@
 
-Over the course of its history, NumPy has attracted PRs by 823 contributors.
-However, its development relies heavily on a relatively small number
-of active maintainers, who share more than half of the contributions among
-themselves.
+We use Git for version control and GitHub as the public hosting service for our
+official \emph{upstream} repository (\url{https://github.com/numpy/numpy}).
+We each work in our own copy (or fork) of the project and use the
+upstream repository as our integration point.
+To get new code into the upstream repository, we use GitHub's
+pull request (PR) mechanism.
+This allows us to review code before integrating it as well as to run a
+large number of tests on the modificatied code to ensure that it does not
+break things.
+
+We also use GitHub's issue tracking system to collect and triage bugs.
+
+\subsection*{Governance}
 
@@
 % https://github.com/numpy/numpy/pull/6352
-To formalize its unwritten development policy,
 NumPy adopted an official Governance Document on October~5,
@@
 the process for responding to such reports.
-
-NumPy is currently maintained by a group of 23 contributors with commit rights
-to the NumPy code base. Out of these, 17 maintainers were active in
-2019, 4 of whom were paid to work on the project full-time.
-Additionally, there are a few long term developers who contributed and maintain
-specific parts of NumPy, but are not officially maintainers.
-
-At a release cycle of about every half year, the five recent releases in the years
-2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
-    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
-    All data provided here is based on the development which happened through GitHub
-    pull requests. In some cases contributions by maintainers may not be categorized as such.}
-% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
-with each release attracting more than a hundred new contributors.
-Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
-master branch.
-Although the number of PRs being merged fluctuates,
-the plot indicates an increased number of contributions over the past
-years.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
-    \caption{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012. The total number of PRs is indicated with the
-        lower blue area showing the portion contributed by current or previous
-        maintainers.}\label{fig:prs-over-time}
-\end{figure}
 
@@
 
-The first activities organized for the new grant were a NEP development
-sprint, a weekly community call, and a meeting to write a draft roadmap.
+\subsection*{Developers}
+
+NumPy is currently maintained by a group of 23 contributors with commit rights
+to the NumPy code base. Out of these, 17 maintainers were active in
+2019, 4 of whom were paid to work on the project full-time.
+Additionally, there are a few long term developers who contributed and maintain
+specific parts of NumPy, but are not officially maintainers.
+
+Over the course of its history, NumPy has attracted PRs by 823 contributors.
+However, its development relies heavily on a relatively small number
+of active maintainers, who share more than half of the contributions among
+themselves.
+
+At a release cycle of about every half year, the five recent releases in the years
+2018 and 2019 have averaged about 450~PRs each,\footnote{
+    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
+    All data provided here is based on the development which happened through GitHub
+    PRs. In some cases contributions by maintainers may not be categorized as such.}
+% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
+with each release attracting more than a hundred new contributors.
+Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
+master branch.
+Although the number of PRs being merged fluctuates,
+the plot indicates an increased number of contributions over the past
+years.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
+    \caption{Number of pull requests merged into the NumPy master branch for each
+        quarter since 2012. The total number of PRs is indicated with the
+        lower blue area showing the portion contributed by current or previous
+        maintainers.}\label{fig:prs-over-time}
+\end{figure}
+
+\subsection*{Community calls}
 
@@
 
-NumPy plays a central role in building and standardizing much of the scientific
-Python community infrastructure.
-The docstring standard mentioned in the history section is now widely adopted.
-We are also now using the NEP system as a way to help coordinate the larger
-scientific Python community.
-% https://numpy.org/neps/nep-0029-deprecation_policy.html
-For example, in NEP 29, we recommend, along with leaders from various other
-projects, that all projects across the Scientific Python ecosystem adopt a
-common ``time window-based'' policy for support of Python and NumPy versions.
-This standard will simplify downstream project and release planning.
+\subsection*{NumPy enhancement proposals}
+
+Given the complexity of the codebase and the massive number of projects depending
+on it, large changes require careful planning and substantial work.
+As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
+of enhancement proposals---used infrequently and informally until 2017 when,
+to facilitate work expected from funded development, he formalized the
+NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
+Python Enhancement Proposals (PEPs) for ``proposing major new
+features, for collecting community input on an issue, and for
+documenting the design decisions that have gone into
+Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
+Since then there have been 19 proposed NEPS---6 have been implemented,
+4 have been accepted and are being implemented, 4 are under
+consideration, 3 have been deferred or superseded, and 2 have been rejected
+or withdrawn.
+
+%NumPy plays a central role in building and standardizing much of the scientific
+%Python community infrastructure.
+%The docstring standard mentioned in the history section is now widely adopted.
+%We are also now using the NEP system as a way to help coordinate the larger
+%scientific Python community.
+%% https://numpy.org/neps/nep-0029-deprecation_policy.html
+%For example, in NEP 29, we recommend, along with leaders from various other
+%projects, that all projects across the Scientific Python ecosystem adopt a
+%common ``time window-based'' policy for support of Python and NumPy versions.
+%This standard will simplify downstream project and release planning.
 
@@
 \subsection*{Recent technical improvements}
-
-Given the complexity of the codebase and the massive number of projects depending
-on it, large changes require careful planning and substantial work.
-As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
-of enhancement proposals---used infrequently and informally until 2017 when,
-to facilitate work expected from funded development, he formalized the
-NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
-Python Enhancement Proposals (PEPs) for ``proposing major new
-features, for collecting community input on an issue, and for
-documenting the design decisions that have gone into
-Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
-Since then there have been 19 proposed NEPS---6 have been implemented,
-4 have been accepted and are being implemented, 4 are under
-consideration, 3 have been deferred or superseded, and 2 have been rejected
-or withdrawn.
 
@@
 will be expected for a library at the base of the ecosystem.
-
-\subsection*{Dtype}
-
-The meaning of each element within a NumPy array is described by its
-datatype (dtype). NumPy arrays can hold all common numerical
-datatypes, strings, datetimes, and generic Python objects, each of
-which is identified by a dtype.
-%We are busy overhauling the datatype
-%system to make it behave consistently and to simplify the creation of
-%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
-%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
-While numerical dtypes make the foundation for most users,
-there are a multitude of use cases which cannot prosper due to current
-limitations, such as physical units\cite{astropy,Goldbaum2018,pint},
-% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
-geometrical objects\cite{pygeos}, and automatic
-differentiation\cite{pyadolc}.
-%Projects addressing these exist, each struggles with current
-%limitations.  % TODO: I may need citations, we could link github issues.
-The proposed overhaul of the datatype system will make it behave consistently and
-to simplify the creation of custom dtypes, both in C and in Python

### 2020-01-07 body.tex
Rewrite ndarray section

@@
 
-The NumPy array data structure stores regularly spaced elements of a single
-data type in a contiguous block of memory; along with meta-data such as shape,
-it allows for the efficient representation of $n$-dimensional arrays \cite{vanderwalt2011numpy}.
+The NumPy array is a structure that efficiently stores and accesses $n$-dimensional array data\cite{vanderwalt2011numpy}.
 The need for such a data structure arises often in science.
-For instance, they may store measurements from an experiment or simulation
-at discrete time points.
+For instance, it may be used to store experimental measurements made at discrete time intervals (1-D arrays), gray-level photographs (2-D arrays), spectral measurement on a regular grid (3-D arrays), or color videos (4-D arrays).
 
-Data analysis frequently involves operations $(+, -, \times, /, \log, \sin, \dots)$ on
-the individual measurements.
-In programming languages such as C or Java, this would involve looping over
-every element to perform these operations.
-Array computations generalize these element-wise operations on scalars to
-vectors (1-d arrays), matrices (2-d arrays), and higher dimensional arrays (n-d
-arrays).
-For instance, if \code{x} is an array, then \code{2 * x} produces
-an array whose values are twice those in \code{x}.
-Similarly, if \code{x} and \code{y} have the same shape, then
-\code{x * y} produces an array whose values are the products of
-the corresponding entries of \code{x} and \code{y}.
-When \code{x} and \code{y} have different shapes, the smaller array is \emph{broadcast}
-across the larger array to a common shape when sensible.
-For example, if \code{x} is a 1-d array with length 3, and \code{y} is a 2-d
-array with shape $2 \times 1$, the output will be an array with shape $2 \times 3$
-(see Figure~\ref{fig:broadcasting}).
+The NumPy array consists of a pointer to memory, along with meta-data used to interpret the data stored there.
+For example, while computer memory stores elements linearly and has no notion of higher-dimensions, we can interpret it as $n$-dimensional arrays by also providing its shape, strides (the number of bytes to advance in memory to skip along rows, columns, etc.), and data-type.
+
+Once data is viewed as an $n$-dimensional array, more abstract operations such as reductions (e.g., summation across columns), products, and so forth, are defined.
+Users predominantly interact with NumPy arrays using these operators and associated utility functions through an easily readable high-level API (Application Programming Interface).
+
+Sometimes, it is desirable to combine arrays of different dimensions.
+One may, for example, want to add a scalar to a vector, or scale each column of an array by a specific scalar value.
+This is done through {\em broadcasting}, where one of the arrays is virtually duplicated (i.e., without copying any data in memory), so that the shapes of the operands match.
+Broadcasting is illustrated in \ref{fig:broadcasting}.
 
@@
 
-These vectorized operations often produce one-liners that would be many tens
-of lines of code in languages such as C.
+NumPy include numerous functions that perform element-wise and reduction operations on arrays, including arithmetic, statistics, and trigonometry.
+They aim to loop over array elements near-optimally (taking into consideration, e.g., strides, in order to best utilize the computer's fast cache memory), and implement broadcasting.
+Vectorized operations that would take many tens of lines to express in languages such as C can often be implemented as a single, clear Python expression.
 
-Element-wise operations are not always desired.
-Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is 1-d array.
+NumPy also provides support for accelerated linear algebra, and supports several established backends such as OpenBLAS and Intel MKL.
+Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is a 1-d array.
 Then \code{A @ x} computes the matrix-vector product $Ax$ and
@@
 
-NumPy also provides a large collection of functions for array manipulation
-including functions for:
-creating, reshaping, concatenating, and padding arrays;
-searching, sorting and counting data in arrays;
-computing elementary statistics, such as the mean, median, variance, and standard deviation;
-reading and writing files;
-and more.
-It also provides extensive support for generating pseudorandom numbers
-as well as an assortment of probability distributions.
-For historical reasons, NumPy also includes basic functionality for
-linear algebra, fast Fourier transforms and windowing,
-and polynomial fitting.
+In addition, NumPy provides a large collection of utility functions for creating, reshaping, concatenating, and padding arrays; for searching, sorting and counting data in arrays; and for reading and writing files.
+It provides extensive support for generating pseudorandom numbers and includes an assortment of probability distributions.
+For historical reasons, NumPy also includes basic functionality for linear algebra, fast Fourier transforms and windowing, and polynomial fitting.
 We will continue supporting these, but we will not expand beyond them.

### 2020-01-07 body.tex
Add notes
Start rewriting Recent Development section

@@
 
-\section*{Hybrid development}
+\section*{Recent development}
 
@@
 
-In 2018, we hired our first full-time developers ...
+The project entered a new phase of development in 2018 with the hiring of two full-time developers.
+With this new resource, we were able to address long-outstanding technical debt, review more volunteer work, and implement several larger features.
 
-The first activities organized for the new grants were a development planning
-sprint, a weekly community call, and a meeting to write a draft roadmap.
-Since the roadmap was meant to represent community consensus on future
-development, it was presented for comment at the annual SciPy
-conference during a so-called ``Birds of a Feather'' session, attended
-by more than 100 people.  One additional meeting was held to discuss
-specific enhancements, such as the development of a new data-type
-system.  The final roadmap was vetted by the community via online
-discussion before being accepted.
+In an open source project that has relied on volunteer effort for so many years, the introduction of paid work can be disruptive.
+In such communities, decision making power is often tied to contributed effort, and when some individuals are paid it skews the equation.
+To guard against this, the project set up an explicit governance structure, formalized the feature proposal process, and constructed a development roadmap with extensive feedback from the community.
+Individuals were hired that were already active in NumPy and its surrounding communities, who already shared the project's norms and values.
+Funded developers were also expected to review code as a part of their job, thereby supporting the work done by volunteers.
+
+Still, funded development allowed for the implementation of several larger features, which require a sizeable amount of time, conversations with multiple stakeholders, and longer term focus; and were therefore more difficult to accomplish by volunteer effort alone.
+
+% --- I'm here now ---
 

### 2020-01-08 body.tex
First draft of rewrite of Recent Developments

@@
 deep learning libraries such as PyTorch\cite{pytorch}.
+
+% S: Maybe we want to emphasize Jax, which has rapidly become very popular
 
@@
 
-The project entered a new phase of development in 2018 with the hiring of two full-time developers.
-With this new resource, we were able to address long-outstanding technical debt, review more volunteer work, and implement several larger features.
+The project entered a new phase of development in 2018 with the hiring of two (and later four) full-time developers.
+With this new resource, NumPy was able to address long-outstanding technical debt, review more volunteer work, and implement several larger features.
 
 In an open source project that has relied on volunteer effort for so many years, the introduction of paid work can be disruptive.
-In such communities, decision making power is often tied to contributed effort, and when some individuals are paid it skews the equation.
+In such communities, decision making power is often tied to contributed effort, and when some individuals are paid it skews the balance.
 To guard against this, the project set up an explicit governance structure, formalized the feature proposal process, and constructed a development roadmap with extensive feedback from the community.
-Individuals were hired that were already active in NumPy and its surrounding communities, who already shared the project's norms and values.
+In addition, individuals were hired that were already active in NumPy and its surrounding communities: they already shared the project's norms and values.
 Funded developers were also expected to review code as a part of their job, thereby supporting the work done by volunteers.
 
-Still, funded development allowed for the implementation of several larger features, which require a sizeable amount of time, conversations with multiple stakeholders, and longer term focus; and were therefore more difficult to accomplish by volunteer effort alone.
+Still, funded development allowed for the implementation of several bigger features, which required a sizeable amount of time, conversations with multiple stakeholders, as well as longer term focus, and were therefore difficult to accomplish by volunteer effort alone.
 
-% --- I'm here now ---
+New work divides roughly into two categories: improving the library's interaction with external packages, and, with the benefit of several years' worth of experience, redesigning core parts of the library.
 
-We briefly highlight two of those below.
+For example, a new array protocol was implemented that allows NumPy to consume array-like objects produced by libraries outside of NumPy.
+When asking NumPy to, e.g., sum such an array, instead of attempting that operation itself, NumPy requests the original array to complete the computation.
+This means that the NumPy API---the syntax for interacting with arrays---can still be used, but now with a library that supports GPUs, does distributed computation processing, or stores data in the cloud.
 
-- array function protocol
+A second example is the rewrite of NumPy's random number generation facilities.
+These have been maintained but unchanged for many years to preserve bit-by-bit compatibility of random number streams.
+A rewrite that allowed for the underlying pseudorandom algorithms to be exchanged and extended was completed by a volunteer contributor and integrated by a paid developer.
 
-- random number generation
+An example of ongoing work is a redesign of the data type system.
+Data types (dtypes for short) tell us more about the elements stored in a NumPy array.
+Arrays can hold, e.g., numerical data (of lower and higher precision), strings, datetimes, or Python objects, each identified by its own dtype.
 
-The random number generator library in NumPy provides several alternative
-\emph{bit stream generators} that provide the core function of generating
-random integers.
-A higher-level generator class that implements an assortment of
-probability distributions is provided. It includes the beta, gamma
-and Weibull distributions, the univariate and multivariate normal
-distributions, and more.
-
-Ongoing work
-
-- dtype
-
-The meaning of each element within a NumPy array is described by its
-datatype (dtype). NumPy arrays can hold all common numerical
-datatypes, strings, datetimes, and generic Python objects, each of
-which is identified by a dtype.
 %We are busy overhauling the datatype
@@
 %of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
-While numerical dtypes make the foundation for most users,
-there are a multitude of use cases which cannot prosper due to current
-limitations, such as physical units\cite{astropy,Goldbaum2018,pint},
+While the majority of users rely only on numerical dtypes,
+there are a multitude of use cases for which the current implementation is unsuited, such as physical units\cite{astropy,Goldbaum2018,pint},
 % pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
@@
 %limitations.  % TODO: I may need citations, we could link github issues.
-The proposed overhaul of the datatype system will make it behave consistently and
-to simplify the creation of custom dtypes, both in C and in Python.
+The proposed overhaul of the datatype system will make it behave consistently and will simplify the creation of custom dtypes, both in C and in Python.
 

### 2020-01-08 body.tex
Revisions to recent developments

@@
   \includegraphics[width=0.5\linewidth]{static/broadcasting}
-  \caption{
-    Broadcasting a $1 \times 3$ array against a $2 \times 1$ array
-    yields a $2 \times 3$ result.
+  \caption{\fixme{PLACEHOLDER: Stefan is working on some new figures to illustrate arrays, broadcasting, etc.}
    }
@@
 
-For many years, NumPy had no dedicated funding, and development time
-was mostly contributed freely by students and researchers in their
-spare time.  In-person meetings were sponsored off of individual's
-research grants, and while industry sponsored contributions were made,
-notably those by Mark Wiebe when he was sponsored by Travis Oliphant
-at Enthought, the package's rapid adoption in scientific
-computing took place without external investment.
+%For many years, NumPy had no dedicated funding, and development time
+%was mostly contributed freely by students and researchers in their
+%spare time.  In-person meetings were sponsored off of individual's
+%research grants, and while industry sponsored contributions were made,
+%notably those by Mark Wiebe when he was sponsored by Travis Oliphant
+%at Enthought, the package's rapid adoption in scientific
+%computing took place without external investment.
 
-The project entered a new phase of development in 2018 with the hiring of two (and later four) full-time developers.
-With this new resource, NumPy was able to address long-outstanding technical debt, review more volunteer work, and implement several larger features.
+NumPy was initially developed by students, faculty, and researchers in their
+spare time to provide a modern array computation library for Python.
+They were influenced by their experiences with powerful interactive programming
+languages for scientific computing like Yurik \cite{munro1995using} as well
+as commercial languages like IDL and Matlab.
+Often they developed code to solve their own or their colleagues problems.
 
-In an open source project that has relied on volunteer effort for so many years, the introduction of paid work can be disruptive.
-In such communities, decision making power is often tied to contributed effort, and when some individuals are paid it skews the balance.
-To guard against this, the project set up an explicit governance structure, formalized the feature proposal process, and constructed a development roadmap with extensive feedback from the community.
-In addition, individuals were hired that were already active in NumPy and its surrounding communities: they already shared the project's norms and values.
-Funded developers were also expected to review code as a part of their job, thereby supporting the work done by volunteers.
+As Python's popularity for scientific computing, data science, and machine
+learning increased, many new use cases arose.
+At the same time, the massive number of scientific Python packages that
+build on NumPy mean that it has an unusually high need for stability.
 
-Still, funded development allowed for the implementation of several bigger features, which required a sizeable amount of time, conversations with multiple stakeholders, as well as longer term focus, and were therefore difficult to accomplish by volunteer effort alone.
+In 2018, we hired two (and later four) full-time developers.
+This enabled us to implement several bigger features,
+which required a sizeable amount of time, conversations with multiple
+stakeholders, as well as longer term focus, and were therefore difficult to
+accomplish by volunteer effort alone.
+To ensure that this paid effort did not disrupt our community owned character
+we formalized the feature proposal process, and constructed a development
+roadmap with extensive input and feedback from the community.   In addition,
+we hired people who were already active in NumPy and its surrounding communities:
+they already shared the project's norms and values.
+Funded developers also regularly reviewed other people's code as a part of
+their job, thereby supporting the work done by volunteers.
 
-New work divides roughly into two categories: improving the library's interaction with external packages, and, with the benefit of several years' worth of experience, redesigning core parts of the library.
+Much of the recent work falls into two categories: improving the library's
+interaction with external packages, and, with the benefit of several years'
+worth of experience, redesigning core parts of the library.
 
-For example, a new array protocol was implemented that allows NumPy to consume array-like objects produced by libraries outside of NumPy.
-When asking NumPy to, e.g., sum such an array, instead of attempting that operation itself, NumPy requests the original array to complete the computation.
-This means that the NumPy API---the syntax for interacting with arrays---can still be used, but now with a library that supports GPUs, does distributed computation processing, or stores data in the cloud.
+For example, a new array protocol was implemented that allows NumPy to consume
+array-like objects produced by libraries outside of NumPy.  When asking NumPy
+to, e.g., sum such an array, instead of attempting that operation itself, NumPy
+requests the original array to complete the computation.  This means that the
+NumPy API---the syntax for interacting with arrays---can still be used, but now
+with a library that supports GPUs, does distributed computation processing, or
+stores data in the cloud.
 
 A second example is the rewrite of NumPy's random number generation facilities.
-These have been maintained but unchanged for many years to preserve bit-by-bit compatibility of random number streams.
-A rewrite that allowed for the underlying pseudorandom algorithms to be exchanged and extended was completed by a volunteer contributor and integrated by a paid developer.
+These have been maintained but unchanged for many years to preserve bit-by-bit
+compatibility of random number streams.  A rewrite that allowed for the
+underlying pseudorandom algorithms to be exchanged and extended was completed
+by a volunteer contributor and integrated by a paid developer.
 
-An example of ongoing work is a redesign of the data type system.
-Data types (dtypes for short) tell us more about the elements stored in a NumPy array.
-Arrays can hold, e.g., numerical data (of lower and higher precision), strings, datetimes, or Python objects, each identified by its own dtype.
+An example of ongoing work is a redesign of the data type system.  Data types
+(dtypes for short) tell us more about the elements stored in a NumPy array.
+Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
+datetimes, or Python objects, each identified by its own dtype.
 
@@
 \section*{Discussion}
+
+\fixme{To be written.}

### 2020-01-08 summary.tex
Update abstract
Revisions to recent developments

@@
 % - 2-3 sentences of basic-level introduction to the field;
-Software libraries for array computation provide powerful notation
-for accessing, manipulating, and operating on data in vectors,
-matrices, and higher-dimensional arrays \cite{iverson1980notation}.
-% Stéfan: why "aggregate" data?
+Array computation libraries provide powerful notation for accessing,
+manipulating, and operating on data in vectors, matrices, and
+higher-dimensional arrays \cite{iverson1980notation}.
 % - a brief account of the background and rationale of the work;
@@
 \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-Not surprisingly, many important scientific research projects and results
-depend on NumPy.
+It has been actively developed for over two decades and has more than 850
+contributors, nearly 300,000 dependent repositories on GitHub, and millions of
+downloads per year.
 It plays an essential role in research analysis pipelines in fields as
@@
 For example, in astronomy, NumPy was fundamental to discovering gravitational
-waves\cite{pycbc}, the first imaging of a black
-hole\cite{eht-imaging}, and continues underpinning the future exploration of our universe\cite{jenness2018lsst}.
-It is also one of the most commonly used machine learning tools among
+waves\cite{pycbc}, the first imaging of a black hole\cite{eht-imaging}, and
+continues underpinning the future exploration of our
+universe\cite{jenness2018lsst}.
+It is one of the most commonly used machine learning tools among
 enterprises \cite{451report2018}.
 % - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
-Here we provide an overview of the project and library organization, highlight
-some recent technical developments, and outline our future plans.
+Here we show how a small group of students and scientists created this
+popular tool, explain the core ideas that helped make it it successful,
+and highlight recent development to better serve new use cases
+and future improvements.
 % - 2-3 sentences putting the main findings into general context so it is
@@
 relevant in an ever-changing computing landscape, and to be a fundamental
-pillar of support for the next decade of data science in Python.
+pillar of support for the next decade of scientific computing, data science,
+and machine learning in Python.

### 2020-01-08 body.tex
Fix typos

@@
 They were influenced by their experiences with powerful interactive programming
-languages for scientific computing like Yurik \cite{munro1995using} as well
+languages for scientific computing like Yorick \cite{munro1995using} as well
 as commercial languages like IDL and Matlab.
-Often they developed code to solve their own or their colleagues problems.
+Often they developed code to solve their own or their colleagues' problems.
 
@@
 At the same time, the massive number of scientific Python packages that
-build on NumPy mean that it has an unusually high need for stability.
+built on NumPy meant that it had an unusually high need for stability.
 

### 2020-01-08 body.tex
Edit intro/history

--- 2020-01-08 body.tex
@@
 % "The secret to successful hiring is this: look for the people who want to change the world.".  Marc Benioff.
-
-NumPy is both a cause and an effect of a striking shift in the nature of
-scientific computing.  In the traditional scientific software model, the user is a {\it consumer}.  Writing software is a task for specialists and
-professionals, supported by income from selling the software to
-scientists.  In the new model, the user is an {\it owner}.  The Python language, with
-the NumPy library, provides a framework that scientists can consume,
-but also encourages scientists to participate as programmers---to
-write code for the library and the ecosystem of packages around it.  This has
-fostered a new culture of contribution and collaboration in scientific
-computing that resulted in a virtuous cycle of increasing use, and increasing
-development.
-
-This cycle is one of several explanations for the astonishing growth of NumPy
-in scientific computing.
-
-\section*{History}
 
@@
 % https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
-% May want to mention Yorick here (origin of broadcasting idea)
+% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)
 
-As Hugunin points out, at the time he wrote Numeric, commercial languages for
-interactive computing with numerical arrays were already well-established,
-including Matlab and IDL.  Back then, it must have seemed a safe bet that
-Numeric would not succeed in this market.  The commercial languages were
-developed and maintained by large groups of full-time professional programmers,
-supported by formal quality assurance, release managers, and writers of
-scientific documentation.  In contrast, Numeric, and later NumPy, would
-continue to be developed almost entirely by scientists, including graduate
-students and faculty, without specific funding and often to the detriment of their careers.
-% S: The "Back then" sentence sets up an expectation.
+%At the time, commercial languages for interactive computing with numerical
+%arrays were already well-established, including Matlab and IDL.
+%They were developed and maintained by large groups of
+%full-time professional programmers, supported by formal quality assurance,
+%release managers, and writers of scientific documentation.
+%In contrast, Numeric, and later NumPy, would continue to be developed almost
+%entirely by scientists, including graduate students and faculty.
+%%without specific funding and often to the detriment of their careers.
+%% S: The "Back then" sentence sets up an expectation.
 Later in this section, we then explain that Numeric/NumPy *was* successful, but we never explicitly close the loop: "despite the odds, great success".
+%NumPy had no dedicated funding, and development time
+%was mostly contributed freely by students and researchers in their
+%spare time.
 
@@
 both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
-This merger resulted in NumPy.  At the time, Oliphant was an assistant
-professor of Electrical and Computer Engineering at Brigham Young University.
+This merger resulted in NumPy.
 
-Over the next 10 years, NumPy attracted a relatively large pool of scientist
-developers.  To benefit from these
-user-developers, the project developed an increasingly strong culture of using
+Over the next 10 years, NumPy attracted more scientist developers.
+The project developed an increasingly strong culture of using
 software-engineering practice to improve collaboration and reduce error
@@
 scientific computing.
-As we implied above, there were several factors that allowed this rapid growth
+There were several factors that allowed this rapid growth
 and successful development.
@@
 It provides extensive support for generating pseudorandom numbers and includes an assortment of probability distributions.
-For historical reasons, NumPy also includes basic functionality for linear algebra, fast Fourier transforms and windowing, and polynomial fitting.
-We will continue supporting these, but we will not expand beyond them.
 

### 2020-01-08 body.tex
More work outline

@@
 The first incarnation of NumPy was a package called Numeric, written in 1995
-by Jim Hugunin, then completing his master's thesis at MIT on
-superconductor-semiconductor junctions.  Hugunin based his package on previous
-work by Jim Fulton, then working at the US Geological Survey, and acknowledges
+by Jim Hugunin, then completing his master's thesis at MIT.
+Hugunin based his package on previous work by Jim Fulton,
+then working at the US Geological Survey, and acknowledges
 many others in his initial report of the package \cite{Hugunin-whitepaper}.
@@
 % S TODO: the first sentence below could flow a bit better
-Around 1998, the Space Telescope Science Institute (STScI) software group began
-to use Python heavily and, around 2000, wrote a re-implementation of much of
-Numeric, called NumArray, to support their need to work on large memory-mapped
-arrays and arrays of mixed data type records \cite{STScI-slither}.  This
-briefly caused the Numeric and NumArray communities to diverge, until 2005,
+Around 200, the Space Telescope Science Institute (STScI) software group wrote
+a re-implementation of much of Numeric, called NumArray, to support their
+need to work on large memory-mapped arrays and arrays of mixed data type
+records \cite{STScI-slither}.
+This briefly caused the Numeric and NumArray communities to diverge, until 2005,
 when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
@@
 continuous testing that runs a large battery of automated tests for every proposed
-change to NumPy. At first, NumPy had rather sparse documentation, but the project
-developed a new standard for ``docstrings''---text describing each NumPy
-function---and used this standard to write comprehensive, high-quality
-documentation, integrated with the source code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}. The NumPy documentation format
-quickly became a standard for the larger Python community.
+change to NumPy.
+The project has comprehensive, high-quality documentation,
+integrated with the source
+code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
 
@@
 
-\section*{Enabling factors}
+\section*{NumPy arrays}
+
+The NumPy array is a structure that efficiently stores and accesses $n$-dimensional array data\cite{vanderwalt2011numpy}.
+The need for such a data structure arises often in science.
+For instance, it may be used to store experimental measurements made at discrete time intervals (1-D arrays), gray-level photographs (2-D arrays), spectral measurement on a regular grid (3-D arrays), or color videos (4-D arrays).
+
+The NumPy array consists of a pointer to memory, along with meta-data used to interpret the data stored there.
+For example, while computer memory stores elements linearly and has no notion of higher-dimensions, we can interpret it as $n$-dimensional arrays by also providing its shape, strides (the number of bytes to advance in memory to skip along rows, columns, etc.), and data-type.
+
+Once data is viewed as an $n$-dimensional array, more abstract operations such as reductions (e.g., summation across columns), products, and so forth, are defined.
+Users predominantly interact with NumPy arrays using these operators and associated utility functions through an easily readable high-level API (Application Programming Interface).
+
+Sometimes, it is desirable to combine arrays of different dimensions.
+One may, for example, want to add a scalar to a vector, or scale each column of an array by a specific scalar value.
+This is done through {\em broadcasting}, where one of the arrays is virtually duplicated (i.e., without copying any data in memory), so that the shapes of the operands match.
+Broadcasting is illustrated in \ref{fig:broadcasting}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{static/broadcasting}
+  \caption{\fixme{PLACEHOLDER: Stefan is working on some new figures to illustrate arrays, broadcasting, etc.}
+   }
+  \label{fig:broadcasting}
+\end{figure}
+
+NumPy include numerous functions that perform element-wise and reduction operations on arrays, including arithmetic, statistics, and trigonometry.
+They aim to loop over array elements near-optimally (taking into consideration, e.g., strides, in order to best utilize the computer's fast cache memory), and implement broadcasting.
+Vectorized operations that would take many tens of lines to express in languages such as C can often be implemented as a single, clear Python expression.
+
+NumPy also provides support for accelerated linear algebra, and supports several established backends such as OpenBLAS and Intel MKL.
+Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is a 1-d array.
+Then \code{A @ x} computes the matrix-vector product $Ax$ and
+\code{x @ A} computes the vector-matrix product $x^\top A$.
+This matrix multiplication operation extends to higher-dimensions by
+treating the array as a stack of matrices.
+
+In addition, NumPy provides a large collection of utility functions for creating, reshaping, concatenating, and padding arrays; for searching, sorting and counting data in arrays; and for reading and writing files.
+It provides extensive support for generating pseudorandom numbers and includes an assortment of probability distributions.
+
+\section*{Recent development}
+
+%For many years, NumPy had no dedicated funding, and development time
+%was mostly contributed freely by students and researchers in their
+%spare time.  In-person meetings were sponsored off of individual's
+%research grants, and while industry sponsored contributions were made,
+%notably those by Mark Wiebe when he was sponsored by Travis Oliphant
+%at Enthought, the package's rapid adoption in scientific
+%computing took place without external investment.
+
+NumPy was initially developed by students, faculty, and researchers in their
+spare time to provide a modern array computation library for Python.
+They were influenced by their experiences with powerful interactive programming
+languages for scientific computing like Yorick \cite{munro1995using} as well
+as commercial languages like IDL and Matlab.
+Often they developed code to solve their own or their colleagues' problems.
+
+As Python's popularity for scientific computing, data science, and machine
+learning increased, many new use cases arose.
+At the same time, the massive number of scientific Python packages that
+built on NumPy meant that it had an unusually high need for stability.
+So to guide our development we formalized the feature proposal process, and
+constructed a development roadmap with extensive input and feedback from the
+community.
+
+In 2018, we hired two (and later four) full-time developers.
+This enabled us, with the benefit of several years' worth of experience,
+to redesign core parts of the library.
+
+One example of redesigning core parts of the library is the rewrite of NumPy's
+random number generation facilities.
+These have been maintained but unchanged for many years to preserve bit-by-bit
+compatibility of random number streams.  A rewrite that allowed for the
+underlying pseudorandom algorithms to be exchanged and extended was completed
+by a volunteer contributor and integrated by a paid developer.
+
+Another ongoing example is a redesign of the data type system.  Data types
+(dtypes for short) tell us more about the elements stored in a NumPy array.
+Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
+datetimes, or Python objects, each identified by its own dtype.
+
+%We are busy overhauling the datatype
+%system to make it behave consistently and to simplify the creation of
+%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
+%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
+While the majority of users rely only on numerical dtypes,
+there are a multitude of use cases for which the current implementation is unsuited, such as physical units\cite{astropy,Goldbaum2018,pint},
+% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
+geometrical objects\cite{pygeos}, and automatic
+differentiation\cite{pyadolc}.
+%Projects addressing these exist, each struggles with current
+%limitations.  % TODO: I may need citations, we could link github issues.
+The proposed overhaul of the datatype system will make it behave consistently and will simplify the creation of custom dtypes, both in C and in Python.
+
+\section*{Array protocol ??}
+
+We are increasingly focused on improving the library's interaction with external
+packages.
+
+For example, a new array protocol was implemented that allows NumPy to consume
+array-like objects produced by libraries outside of NumPy.  When asking NumPy
+to, e.g., sum such an array, instead of attempting that operation itself, NumPy
+requests the original array to complete the computation.  This means that the
+NumPy API---the syntax for interacting with arrays---can still be used, but now
+with a library that supports GPUs, does distributed computation processing, or
+stores data in the cloud.
+
+%"""
+%NumPy’s high level ndarray API has been implemented several times outside of
+%NumPy itself for different architectures, such as for GPU arrays (CuPy), Sparse
+%arrays (scipy.sparse, pydata/sparse) and parallel arrays (Dask array) as well
+%as various NumPy-like implementations in the deep learning frameworks, like
+%TensorFlow and PyTorch.
+%
+%Similarly there are many projects that build on top of the NumPy API for
+%labeled and indexed arrays (XArray), automatic differentiation (Autograd,
+%Tangent), masked arrays (numpy.ma), physical units (astropy.units, pint, unyt),
+%etc. that add additional functionality on top of the NumPy API. Most of these
+%project also implement a close variation of NumPy’s level high API.
+%
+%We would like to be able to use these libraries together, for example we would
+%like to be able to place a CuPy array within XArray, or perform automatic
+%differentiation on Dask array code. This would be easier to accomplish if code
+%written for NumPy ndarrays could also be used by other NumPy-like projects.
+%"""
+
+\section*{Discussion}
 
@@
 %
-
-\section*{NumPy arrays}
-
-The NumPy array is a structure that efficiently stores and accesses $n$-dimensional array data\cite{vanderwalt2011numpy}.
-The need for such a data structure arises often in science.
-For instance, it may be used to store experimental measurements made at discrete time intervals (1-D arrays), gray-level photographs (2-D arrays), spectral measurement on a regular grid (3-D arrays), or color videos (4-D arrays).
-
-The NumPy array consists of a pointer to memory, along with meta-data used to interpret the data stored there.
-For example, while computer memory stores elements linearly and has no notion of higher-dimensions, we can interpret it as $n$-dimensional arrays by also providing its shape, strides (the number of bytes to advance in memory to skip along rows, columns, etc.), and data-type.
-
-Once data is viewed as an $n$-dimensional array, more abstract operations such as reductions (e.g., summation across columns), products, and so forth, are defined.
-Users predominantly interact with NumPy arrays using these operators and associated utility functions through an easily readable high-level API (Application Programming Interface).
-
-Sometimes, it is desirable to combine arrays of different dimensions.
-One may, for example, want to add a scalar to a vector, or scale each column of an array by a specific scalar value.
-This is done through {\em broadcasting}, where one of the arrays is virtually duplicated (i.e., without copying any data in memory), so that the shapes of the operands match.
-Broadcasting is illustrated in \ref{fig:broadcasting}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.5\linewidth]{static/broadcasting}
-  \caption{\fixme{PLACEHOLDER: Stefan is working on some new figures to illustrate arrays, broadcasting, etc.}
-   }
-  \label{fig:broadcasting}
-\end{figure}
-
-NumPy include numerous functions that perform element-wise and reduction operations on arrays, including arithmetic, statistics, and trigonometry.
-They aim to loop over array elements near-optimally (taking into consideration, e.g., strides, in order to best utilize the computer's fast cache memory), and implement broadcasting.
-Vectorized operations that would take many tens of lines to express in languages such as C can often be implemented as a single, clear Python expression.
-
-NumPy also provides support for accelerated linear algebra, and supports several established backends such as OpenBLAS and Intel MKL.
-Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is a 1-d array.
-Then \code{A @ x} computes the matrix-vector product $Ax$ and
-\code{x @ A} computes the vector-matrix product $x^\top A$.
-This matrix multiplication operation extends to higher-dimensions by
-treating the array as a stack of matrices.
-
-In addition, NumPy provides a large collection of utility functions for creating, reshaping, concatenating, and padding arrays; for searching, sorting and counting data in arrays; and for reading and writing files.
-It provides extensive support for generating pseudorandom numbers and includes an assortment of probability distributions.
-
-\section*{Recent development}
-
-%For many years, NumPy had no dedicated funding, and development time
-%was mostly contributed freely by students and researchers in their
-%spare time.  In-person meetings were sponsored off of individual's
-%research grants, and while industry sponsored contributions were made,
-%notably those by Mark Wiebe when he was sponsored by Travis Oliphant
-%at Enthought, the package's rapid adoption in scientific
-%computing took place without external investment.
-
-NumPy was initially developed by students, faculty, and researchers in their
-spare time to provide a modern array computation library for Python.
-They were influenced by their experiences with powerful interactive programming
-languages for scientific computing like Yorick \cite{munro1995using} as well
-as commercial languages like IDL and Matlab.
-Often they developed code to solve their own or their colleagues' problems.
-
-As Python's popularity for scientific computing, data science, and machine
-learning increased, many new use cases arose.
-At the same time, the massive number of scientific Python packages that
-built on NumPy meant that it had an unusually high need for stability.
-
-In 2018, we hired two (and later four) full-time developers.
-This enabled us to implement several bigger features,
-which required a sizeable amount of time, conversations with multiple
-stakeholders, as well as longer term focus, and were therefore difficult to
-accomplish by volunteer effort alone.
-To ensure that this paid effort did not disrupt our community owned character
-we formalized the feature proposal process, and constructed a development
-roadmap with extensive input and feedback from the community.   In addition,
-we hired people who were already active in NumPy and its surrounding communities:
-they already shared the project's norms and values.
-Funded developers also regularly reviewed other people's code as a part of
-their job, thereby supporting the work done by volunteers.
-
-Much of the recent work falls into two categories: improving the library's
-interaction with external packages, and, with the benefit of several years'
-worth of experience, redesigning core parts of the library.
-
-For example, a new array protocol was implemented that allows NumPy to consume
-array-like objects produced by libraries outside of NumPy.  When asking NumPy
-to, e.g., sum such an array, instead of attempting that operation itself, NumPy
-requests the original array to complete the computation.  This means that the
-NumPy API---the syntax for interacting with arrays---can still be used, but now
-with a library that supports GPUs, does distributed computation processing, or
-stores data in the cloud.
-
-A second example is the rewrite of NumPy's random number generation facilities.
-These have been maintained but unchanged for many years to preserve bit-by-bit
-compatibility of random number streams.  A rewrite that allowed for the
-underlying pseudorandom algorithms to be exchanged and extended was completed
-by a volunteer contributor and integrated by a paid developer.
-
-An example of ongoing work is a redesign of the data type system.  Data types
-(dtypes for short) tell us more about the elements stored in a NumPy array.
-Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
-datetimes, or Python objects, each identified by its own dtype.
-
-%We are busy overhauling the datatype
-%system to make it behave consistently and to simplify the creation of
-%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
-%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
-While the majority of users rely only on numerical dtypes,
-there are a multitude of use cases for which the current implementation is unsuited, such as physical units\cite{astropy,Goldbaum2018,pint},
-% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
-geometrical objects\cite{pygeos}, and automatic
-differentiation\cite{pyadolc}.
-%Projects addressing these exist, each struggles with current
-%limitations.  % TODO: I may need citations, we could link github issues.
-The proposed overhaul of the datatype system will make it behave consistently and will simplify the creation of custom dtypes, both in C and in Python.
-
-\section*{Discussion}
-
-\fixme{To be written.}

### 2020-01-08 body.tex
Reorg based on discussion w/ Ralf and Stefan

@@
-% https://books.google.com/books?id=QQJODwAAQBAJ&pg=PA101&lpg=PA101&dq=%E2%80%9Cthe+secret+to+successful+hiring+is+this:+look+for+the+people+who+want+to+change+the+world.%E2%80%9D&source=bl&ots=D6l8JXfB20&sig=ACfU3U0x1X9Gh8k6MlNV0BqhBa_73Q0yfQ&hl=en&sa=X&ved=2ahUKEwj48pqg_oHgAhVohuAKHYlJAsc4ChDoATAAegQIChAB#v=onepage&q=%E2%80%9Cthe%20secret%20to%20successful%20hiring%20is%20this%3A%20look%20for%20the%20people%20who%20want%20to%20change%20the%20world.%E2%80%9D&f=false
-% "The secret to successful hiring is this: look for the people who want to change the world.".  Marc Benioff.
-
 The first incarnation of NumPy was a package called Numeric, written in 1995
@@
 
-%At the time, commercial languages for interactive computing with numerical
-%arrays were already well-established, including Matlab and IDL.
-%They were developed and maintained by large groups of
-%full-time professional programmers, supported by formal quality assurance,
-%release managers, and writers of scientific documentation.
-%In contrast, Numeric, and later NumPy, would continue to be developed almost
-%entirely by scientists, including graduate students and faculty.
-%%without specific funding and often to the detriment of their careers.
-%% S: The "Back then" sentence sets up an expectation.
-Later in this section, we then explain that Numeric/NumPy *was* successful, but we never explicitly close the loop: "despite the odds, great success".
-%NumPy had no dedicated funding, and development time
-%was mostly contributed freely by students and researchers in their
-%spare time.
-
-%Table X1 about here: backgrounds of initial designers (mainly math, physics).
-
-%Table X2 about here: backgrounds of major committers (mainly math, physics).
-
 % S TODO: the first sentence below could flow a bit better
@@
 This merger resulted in NumPy.
-
-Over the next 10 years, NumPy attracted more scientist developers.
-The project developed an increasingly strong culture of using
-software-engineering practice to improve collaboration and reduce error
-\cite{millman2014developing}. The NumPy team was early in adopting distributed
-revision control and code review to improve collaboration on code, and
-continuous testing that runs a large battery of automated tests for every proposed
-change to NumPy.
-The project has comprehensive, high-quality documentation,
-integrated with the source
-code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
 
@@
 
-The NumPy array is a structure that efficiently stores and accesses $n$-dimensional array data\cite{vanderwalt2011numpy}.
+The NumPy array is a structure that efficiently stores and accesses
+$n$-dimensional array data\cite{vanderwalt2011numpy}.
 The need for such a data structure arises often in science.
-For instance, it may be used to store experimental measurements made at discrete time intervals (1-D arrays), gray-level photographs (2-D arrays), spectral measurement on a regular grid (3-D arrays), or color videos (4-D arrays).
+For instance, it may be used to store experimental measurements made at
+discrete time intervals (1-D arrays), gray-level photographs (2-D arrays),
+spectral measurement on a regular grid (3-D arrays), or color videos (4-D
+arrays).
 
-The NumPy array consists of a pointer to memory, along with meta-data used to interpret the data stored there.
-For example, while computer memory stores elements linearly and has no notion of higher-dimensions, we can interpret it as $n$-dimensional arrays by also providing its shape, strides (the number of bytes to advance in memory to skip along rows, columns, etc.), and data-type.
+The NumPy array consists of a pointer to memory, along with meta-data used to
+interpret the data stored there.
+For example, while computer memory stores elements linearly and has no notion
+of higher-dimensions, we can interpret it as $n$-dimensional arrays by also
+providing its shape, strides (the number of bytes to advance in memory to skip
+along rows, columns, etc.), and data-type.
 
-Once data is viewed as an $n$-dimensional array, more abstract operations such as reductions (e.g., summation across columns), products, and so forth, are defined.
-Users predominantly interact with NumPy arrays using these operators and associated utility functions through an easily readable high-level API (Application Programming Interface).
+Data types (dtypes for short) tell us more about the elements stored in a NumPy
+array.
+Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
+datetimes, or Python objects, each identified by its own dtype.
+
+Once data is viewed as an $n$-dimensional array, more abstract operations such as
+reductions (e.g., summation across columns), products, and so forth, are defined.
+Users predominantly interact with NumPy arrays using these operators and
+associated utility functions through an easily readable high-level API
+(Application Programming Interface).
 
 Sometimes, it is desirable to combine arrays of different dimensions.
-One may, for example, want to add a scalar to a vector, or scale each column of an array by a specific scalar value.
-This is done through {\em broadcasting}, where one of the arrays is virtually duplicated (i.e., without copying any data in memory), so that the shapes of the operands match.
-Broadcasting is illustrated in \ref{fig:broadcasting}.
+One may, for example, want to add a scalar to a vector, or scale each column of
+an array by a specific scalar value.
+This is done through {\em broadcasting}, where one of the arrays is virtually
+duplicated (i.e., without copying any data in memory), so that the shapes of
+the operands match.
+Broadcasting is illustrated in Figure~\ref{fig:broadcasting}.
 
@@
 
-NumPy include numerous functions that perform element-wise and reduction operations on arrays, including arithmetic, statistics, and trigonometry.
-They aim to loop over array elements near-optimally (taking into consideration, e.g., strides, in order to best utilize the computer's fast cache memory), and implement broadcasting.
-Vectorized operations that would take many tens of lines to express in languages such as C can often be implemented as a single, clear Python expression.
+NumPy include numerous functions that perform element-wise and reduction
+operations on arrays, including arithmetic, statistics, and trigonometry.
+They aim to loop over array elements near-optimally (taking into consideration,
+e.g., strides, in order to best utilize the computer's fast cache memory), and
+implement broadcasting.
+Vectorized operations that would take many tens of lines to express in
+languages such as C can often be implemented as a single, clear Python
+expression.
 
-NumPy also provides support for accelerated linear algebra, and supports several established backends such as OpenBLAS and Intel MKL.
+NumPy also provides support for accelerated linear algebra, and supports
+several established backends such as OpenBLAS and Intel MKL.
 Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is a 1-d array.
-Then \code{A @ x} computes the matrix-vector product $Ax$ and
-\code{x @ A} computes the vector-matrix product $x^\top A$.
-This matrix multiplication operation extends to higher-dimensions by
-treating the array as a stack of matrices.
+Then \code{A @ x} computes the matrix-vector product $Ax$ and \code{x @ A} computes
+the vector-matrix product $x^\top A$.
+This matrix multiplication operation extends to higher-dimensions by treating
+the array as a stack of matrices.
 
-In addition, NumPy provides a large collection of utility functions for creating, reshaping, concatenating, and padding arrays; for searching, sorting and counting data in arrays; and for reading and writing files.
-It provides extensive support for generating pseudorandom numbers and includes an assortment of probability distributions.
+In addition, NumPy provides a large collection of utility functions for
+creating, reshaping, concatenating, and padding arrays; for searching, sorting
+and counting data in arrays; and for reading and writing files.
+It provides extensive support for generating pseudorandom numbers and includes
+an assortment of probability distributions.
 
-\section*{Recent development}
+\section*{Changing computational landscape}
 
-%For many years, NumPy had no dedicated funding, and development time
-%was mostly contributed freely by students and researchers in their
-%spare time.  In-person meetings were sponsored off of individual's
-%research grants, and while industry sponsored contributions were made,
-%notably those by Mark Wiebe when he was sponsored by Travis Oliphant
-%at Enthought, the package's rapid adoption in scientific
-%computing took place without external investment.
+\fixme{Ralf will take a first pass on the section on 1/12.}
 
@@
 learning increased, many new use cases arose.
-At the same time, the massive number of scientific Python packages that
-built on NumPy meant that it had an unusually high need for stability.
-So to guide our development we formalized the feature proposal process, and
-constructed a development roadmap with extensive input and feedback from the
-community.
 
-In 2018, we hired two (and later four) full-time developers.
-This enabled us, with the benefit of several years' worth of experience,
-to redesign core parts of the library.
+- GPUs
 
-One example of redesigning core parts of the library is the rewrite of NumPy's
-random number generation facilities.
-These have been maintained but unchanged for many years to preserve bit-by-bit
-compatibility of random number streams.  A rewrite that allowed for the
-underlying pseudorandom algorithms to be exchanged and extended was completed
-by a volunteer contributor and integrated by a paid developer.
+- distributed computing
 
-Another ongoing example is a redesign of the data type system.  Data types
-(dtypes for short) tell us more about the elements stored in a NumPy array.
-Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
-datetimes, or Python objects, each identified by its own dtype.
+- other languages w/ arrays
 
-%We are busy overhauling the datatype
-%system to make it behave consistently and to simplify the creation of
-%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
-%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
-While the majority of users rely only on numerical dtypes,
-there are a multitude of use cases for which the current implementation is unsuited, such as physical units\cite{astropy,Goldbaum2018,pint},
-% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
-geometrical objects\cite{pygeos}, and automatic
-differentiation\cite{pyadolc}.
-%Projects addressing these exist, each struggles with current
-%limitations.  % TODO: I may need citations, we could link github issues.
-The proposed overhaul of the datatype system will make it behave consistently and will simplify the creation of custom dtypes, both in C and in Python.
+- massive explosion in data science, machine learning, and artificial intelligence
 
-\section*{Array protocol ??}
+% NumPy also went from a small community of mostly expert users to
+% one of the most widely used foundations of the scientific Python
+% ecosystem.
+% Now the user base is very different....
 
-We are increasingly focused on improving the library's interaction with external
-packages.
+Another ongoing example is a redesign of the data type system.
+While the majority of users rely only on numerical dtypes, there are a
+multitude of use cases for which the current implementation is unsuited, such
+as physical units\cite{astropy,Goldbaum2018,pint}, geometrical
+objects\cite{pygeos}, and automatic differentiation\cite{pyadolc}.
+The proposed overhaul of the datatype system will make it behave consistently
+and will simplify the creation of custom dtypes, both in C and in Python.
+
+\section*{Array protocols}
+
+\fixme{NumPy’s high level ndarray API has been implemented several times outside of
+NumPy itself for different architectures, such as for GPU arrays (CuPy), Sparse
+arrays (scipy.sparse, pydata/sparse) and parallel arrays (Dask array) as well
+as various NumPy-like implementations in the deep learning frameworks, like
+TensorFlow and PyTorch.}
+
+\fixme{Similarly there are many projects that build on top of the NumPy API for
+labeled and indexed arrays (XArray), automatic differentiation (Autograd,
+Tangent), masked arrays (numpy.ma), physical units (astropy.units, pint, unyt),
+etc. that add additional functionality on top of the NumPy API. Most of these
+project also implement a close variation of NumPy’s level high API.}
+
+\fixme{We would like to be able to use these libraries together, for example we would
+like to be able to place a CuPy array within XArray, or perform automatic
+differentiation on Dask array code. This would be easier to accomplish if code
+written for NumPy ndarrays could also be used by other NumPy-like projects.}
 
@@
 
-%"""
-%NumPy’s high level ndarray API has been implemented several times outside of
-%NumPy itself for different architectures, such as for GPU arrays (CuPy), Sparse
-%arrays (scipy.sparse, pydata/sparse) and parallel arrays (Dask array) as well
-%as various NumPy-like implementations in the deep learning frameworks, like
-%TensorFlow and PyTorch.
-%
-%Similarly there are many projects that build on top of the NumPy API for
-%labeled and indexed arrays (XArray), automatic differentiation (Autograd,
-%Tangent), masked arrays (numpy.ma), physical units (astropy.units, pint, unyt),
-%etc. that add additional functionality on top of the NumPy API. Most of these
-%project also implement a close variation of NumPy’s level high API.
-%
-%We would like to be able to use these libraries together, for example we would
-%like to be able to place a CuPy array within XArray, or perform automatic
-%differentiation on Dask array code. This would be easier to accomplish if code
-%written for NumPy ndarrays could also be used by other NumPy-like projects.
-%"""
+\section*{Discussion}
 
-\section*{Discussion}
+% aka why is this so successful
+% some sense of ongoing work / future directions
+
+Over the last decade, NumPy has attracted more scientist developers.
+The project has strong culture of using software-engineering practice to
+improve collaboration and reduce error \cite{millman2014developing}.
+The NumPy team was early in adopting distributed
+revision control and code review to improve collaboration on code, and
+continuous testing that runs a large battery of automated tests for every proposed
+change to NumPy.
+The project has comprehensive, high-quality documentation,
+integrated with the source
+code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
 

### 2020-01-09 body.tex
Add some notes

@@
 
+\fixme{The idea here is something like:  classic numpy (as described above)
+was built to solve previous problems; new problems have arisen as
+the computational landscape has changed; new hardware, new languages,
+new problems; as a result np's role has changed; it is a foundational
+tool widely-used not a specialist tool used by scientist developers;
+many array-like libraries for various uses and np is serving as
+a standard)}
+
 NumPy was initially developed by students, faculty, and researchers in their
@@
 
+% https://numpy.org/neps/nep-0016-abstract-array.html
+% https://numpy.org/neps/nep-0018-array-function-protocol.html
+% https://numpy.org/neps/nep-0022-ndarray-duck-typing-overview.html
+% https://numpy.org/neps/nep-0030-duck-array-protocol.html
+% https://numpy.org/neps/nep-0037-array-module.html
+
 \fixme{NumPy’s high level ndarray API has been implemented several times outside of
@@
 
-Over the last decade, NumPy has attracted more scientist developers.
-The project has strong culture of using software-engineering practice to
-improve collaboration and reduce error \cite{millman2014developing}.
-The NumPy team was early in adopting distributed
-revision control and code review to improve collaboration on code, and
-continuous testing that runs a large battery of automated tests for every proposed
-change to NumPy.
-The project has comprehensive, high-quality documentation,
-integrated with the source
-code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
-
 The creation of NumPy spurred renewed development in the larger scientific
@@
 component of a larger programming task, therefore:
-
 \begin{quote}
@@
 \end{quote}
-
 Python is already a good choice for many standard programming tasks such as
@@
 
-The third factor is much like the second: a culture in Python that concentrates
-on high quality of code and distribution. Because Python is open-source, it has
-a strong culture of contribution, and therefore, of using tools and process to
-improve collaborative software development, such as distributed version
-control, code review, and automated testing.  These have allowed the project to
-scale safely as it attracted more users and developers, even though these
-developers rarely came to the project with training in software engineering.
+The third factor is much like the second: 
+%a culture in Python that concentrates
+%on high quality of code and distribution.  Because Python is open-source, it has
+%a strong culture of contribution, and therefore, of using tools and process to
+%improve collaborative software development, such as distributed version
+%control, code review, and automated testing.  These have allowed the project to
+%scale safely as it attracted more users and developers, even though these
+%developers rarely came to the project with training in software engineering.
+the project has strong culture of using software-engineering practice to
+improve collaboration and reduce error \cite{millman2014developing}.
+The NumPy team was early in adopting distributed
+revision control and code review to improve collaboration on code, and
+continuous testing that runs a large battery of automated tests for every proposed
+change to NumPy.
+The project has comprehensive, high-quality documentation,
+integrated with the source
+code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
 

### 2020-01-09 body.tex
Typo

@@
 % S TODO: the first sentence below could flow a bit better
-Around 200, the Space Telescope Science Institute (STScI) software group wrote
+Around 2000, the Space Telescope Science Institute (STScI) software group wrote
 a re-implementation of much of Numeric, called NumArray, to support their

### 2020-01-09 body.tex
Add back some missing content

@@
 many others in his initial report of the package \cite{Hugunin-whitepaper}.
+Later, Paul Dubois at Lawrence Livermore National Laboratory toke over the project
+and funded the writing of the first
+manual\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}
 Numeric provided an array object in Python, written in C, and linking to
@@
 % May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)
+% https://aip.scitation.org/doi/pdf/10.1063/1.4823451 
+% https://www.dyalog.com/uploads/documents/Papers/Arrays%20of%20Objects.pdf
 
@@
 
-\fixme{NumPy’s high level ndarray API has been implemented several times outside of
-NumPy itself for different architectures, such as for GPU arrays (CuPy), Sparse
-arrays (scipy.sparse, pydata/sparse) and parallel arrays (Dask array) as well
-as various NumPy-like implementations in the deep learning frameworks, like
-TensorFlow and PyTorch.}
+A vast number of projects are built on NumPy;
+these projects are consumers of the NumPy API.
+Over the last several years, a growing number of projects are providers of
+a \emph{NumPy-like API} and array objects targeting audiences with specialized
+needs beyond NumPy's capabilities.
+For example, the NumPy API is implemented by several popular tensor computation
+libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
+JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
+and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
+and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
+provide tensor APIs with NumPy-inspired semantics.
+It is also implemented in packages that support sparse arrays
+such as \code{scipy.sparse} and \code{pydata.sparse}.
+Another notable example is Dask, a library for parallel computing in
+Python.  Dask adopts the NumPy API and therefore presents a familiar
+interface to existing NumPy users, while adding powerful abilities to
+parallelize and distribute tasks.
 
@@
 
+The multitude of specialized projects creates the difficulty that consumers
+of these NumPy-like APIs write code specific to a single project and do not support
+all of the above array providers.
+This is a burden for users relying on the specialized array-like, since
+a tool they need may not work for them.
+It also creates new challenges for end-users who need to transition
+from NumPy to a more specialized array.
+The growing multitude of specialized projects with NumPy-like APIs threatened
+to again fracture the scientific Python community.
+
 \fixme{We would like to be able to use these libraries together, for example we would
@@
 written for NumPy ndarrays could also be used by other NumPy-like projects.}
+
+To address these issues NumPy has the goal of providing the fundamental
+API for \emph{interoperability} between the various NumPy-like APIs.
 

### 2020-01-09 methods.tex
Reorg based on discussion w/ Ralf and Stefan
Add back some missing content

@@
 We also use GitHub's issue tracking system to collect and triage bugs.
+
+\subsection*{Library organization}
+
+Broadly, the NumPy library consists of the following parts:
+the NumPy array data structure; the so-called \emph{universal functions};
+a set of library functions for manipulating arrays and doing scientific
+computation; infrastructure libraries for unit tests and Python package
+building; and the program \code{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
+The \code{ndarray} and the universal functions are generally considered
+the core of the library.
+In the following, we give a brief summary of these components of the
+library.
+
+\paragraph{\emph{Core.}}  The \code{ndarray} data structure and the
+universal functions make up the core of NumPy.
+
+The \code{ndarray} is the data structure at the heart of NumPy.
+The data structure stores regularly strided homogeneous data types
+in a contiguous block memory, allowing for the efficient representation
+of $n$-dimensional data.
+More details about the data structure are given in ``The NumPy array:
+a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.
+
+The \emph{universal functions}, or more concisely, \emph{ufuncs},
+are functions written in C that implement efficient looping over
+NumPy arrays. An important feature of ufuncs is the built-in
+implementation of \emph{broadcasting}.  For example, the function
+\code{arctan2(x, y)} is a ufunc that accepts two values and computes
+$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
+the ufunc will take care of looping over the dimensions of the inputs
+in such a way that if, say, \code{x} is a 1-d array with length 3, and
+\code{y} is a 2-d array with shape $2 \times 1$, the output will be
+an array with shape $2 \times 3$.  The ufunc machinery takes care
+of calling the function with all the appropriate combinations of
+input array elements to complete the output array.
+The elementary arithmetic operations of addition, multiplication, etc.,
+are implemented as ufuncs, so broadcasting applies even in expressions
+such as \code{x + y * z}.
+
+\paragraph{\emph{Computing libraries.}}
+NumPy provides a large library of functions for array manipulation
+and scientific computing, including functions for: creating, reshaping,
+concatenating, and padding arrays; searching, sorting and counting data
+in arrays; computing elementary statistics, such as the mean, median,
+variance, and standard deviation; file I/O; and more.
+
+A suite of functions for computing the \emph{fast Fourier transform (FFT)}
+and its inverse is provided.
+
+NumPy's linear algebra library includes functions for: solving linear
+systems of equations; computing various functions of a matrix, including
+the determinant, the norm, the inverse, and the pseudo-inverse;
+computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
+and more.
+
+The random number generator library in NumPy provides several alternative
+\emph{bit stream generators} that provide the core function of generating
+random integers.
+A higher-level generator class that implements an assortment of
+probability distributions is provided. It includes the beta, gamma
+and Weibull distributions, the univariate and multivariate normal
+distributions, and more.
+
+\paragraph{\emph{Infrastructure libraries.}} NumPy provides utilities
+for writing unit tests and for building Python packages.
+
+The \code{testing} subpackage provides functions such as
+\code{assert\_allclose(actual, desired)} that may be used in unit
+test suites for code that uses NumPy arrays.
+
+NumPy provides a subpackage includes functions and classes
+that can be used by libraries that have NumPy as dependency,
+to be used in the library's configuration code that installs the
+library and that builds a package that can be used, for example,
+on the PyPI web site.
+
+\paragraph{\emph{F2PY.}}  The program \code{f2py} is a tool for
+building NumPy-aware Python wrappers of Fortran functions.
+NumPy itself does not use any Fortran code;  F2PY is part of NumPy
+for historical reasons.
 
@@
 
+The massive number of scientific Python packages that
+built on NumPy meant that it had an unusually high need for stability.
+So to guide our development we formalized the feature proposal process, and
+constructed a development roadmap with extensive input and feedback from the
+community.
+
 Weekly community calls alternate between triage and
@@
 on it, large changes require careful planning and substantial work.
-As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
-of enhancement proposals---used infrequently and informally until 2017 when,
-to facilitate work expected from funded development, he formalized the
-NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
+NumPy Enhancement Proposals (NEPs) are modeled after
 Python Enhancement Proposals (PEPs) for ``proposing major new
@@
 
-%NumPy plays a central role in building and standardizing much of the scientific
-%Python community infrastructure.
-%The docstring standard mentioned in the history section is now widely adopted.
-%We are also now using the NEP system as a way to help coordinate the larger
-%scientific Python community.
-%% https://numpy.org/neps/nep-0029-deprecation_policy.html
-%For example, in NEP 29, we recommend, along with leaders from various other
-%projects, that all projects across the Scientific Python ecosystem adopt a
-%common ``time window-based'' policy for support of Python and NumPy versions.
-%This standard will simplify downstream project and release planning.
+\subsection*{Central role}
+
+NumPy plays a central role in building and standardizing much of the scientific
+Python community infrastructure.
+The docstring standard mentioned in the history section is now widely adopted.
+We are also now using the NEP system as a way to help coordinate the larger
+scientific Python community.
+% https://numpy.org/neps/nep-0029-deprecation_policy.html
+For example, in NEP 29, we recommend, along with leaders from various other
+projects, that all projects across the Scientific Python ecosystem adopt a
+common ``time window-based'' policy for support of Python and NumPy versions.
+This standard will simplify downstream project and release planning.
 
@@
 
-A vast number of projects are built on NumPy;
-these projects are consumers of the NumPy API.
-Over the last several years, a growing number of projects are providers of
-a \emph{NumPy-like API} and array objects targeting audiences with specialized
-needs beyond NumPy's capabilities.
-For example, the NumPy API is implemented by several popular tensor computation
-libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
-JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
-and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
-and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
-provide tensor APIs with NumPy-inspired semantics.
-It is also implemented in packages that support sparse arrays
-such as \code{scipy.sparse} and \code{pydata.sparse}.
-Another notable example is Dask, a library for parallel computing in
-Python.  Dask adopts the NumPy API and therefore presents a familiar
-interface to existing NumPy users, while adding powerful abilities to
-parallelize and distribute tasks.
-
-The multitude of specialized projects creates the difficulty that consumers
-of these NumPy-like APIs write code specific to a single project and do not support
-all of the above array providers.
-This is a burden for users relying on the specialized array-like, since
-a tool they need may not work for them.
-It also creates new challenges for end-users who need to transition
-from NumPy to a more specialized array.
-The growing multitude of specialized projects with NumPy-like APIs threatened
-to again fracture the scientific Python community.
-
-To address these issues NumPy has the goal of providing the fundamental
-API for \emph{interoperability} between the various NumPy-like APIs.
 An earlier step in this direction was the implementation of the

### 2020-01-09 body.tex
Typo

@@
 and funded the writing of the first
-manual\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}
+manual\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
 Numeric provided an array object in Python, written in C, and linking to

### 2020-01-09 body.tex
Add draft diagram sketches

@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=\textwidth]{static/sketches/array concepts}
+  \caption{\fixme{PLACEHOLDER: This figure needs some text to describe
+  it}
+   }
+  \label{fig:arrayconcepts}
+\end{figure*}
+
 The NumPy array is a structure that efficiently stores and accesses
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{static/broadcasting}
-  \caption{\fixme{PLACEHOLDER: Stefan is working on some new figures to illustrate arrays, broadcasting, etc.}
+  \includegraphics[width=\linewidth]{static/sketches/fancy indexing}
+  \caption{\fixme{Indexing using arrays, or ``fancy indexing''.}
    }
@@
 expression.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{static/sketches/reductions}
+  \caption{\fixme{Reducing an array along one of several axes.}
+   }
+  \label{fig:reductions}
+\end{figure}
 
@@
 \section*{Array protocols}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=0.8\textwidth]{static/sketches/duck arrays}
+  \caption{\fixme{Examples of so-called ``duck arrays'': arrays
+  implemented by another library that behave like NumPy arrays, and
+  support NumPy operations.  Through protocols, it is possible to have
+  a call such as $\cos(\cdot)$ succeed on each of these.}}
+  \label{fig:arrayconcepts}
+\end{figure*}
 

### 2020-01-09 body.tex
Fine-tuning outline

@@
 % https://aip.scitation.org/doi/pdf/10.1063/1.4823451 
-% https://www.dyalog.com/uploads/documents/Papers/Arrays%20of%20Objects.pdf
+% https://aip.scitation.org/doi/pdf/10.1063/1.4822400
 
@@
   \centering
-  \includegraphics[width=\textwidth]{static/sketches/array concepts}
+  \includegraphics[width=\textwidth]{static/sketches/array-concepts}
   \caption{\fixme{PLACEHOLDER: This figure needs some text to describe
@@
 
-Once data is viewed as an $n$-dimensional array, more abstract operations such as
-reductions (e.g., summation across columns), products, and so forth, are defined.
+Once data is viewed as an $n$-dimensional array, more abstract operations such
+as reductions (e.g., summation across columns), products, and so forth, are
+defined (Figure~\ref{fig:reductions}).
 Users predominantly interact with NumPy arrays using these operators and
@@
   \centering
-  \includegraphics[width=\linewidth]{static/sketches/fancy indexing}
+  \includegraphics[width=\linewidth]{static/sketches/fancy-indexing}
   \caption{\fixme{Indexing using arrays, or ``fancy indexing''.}
@@
 
-% NumPy also went from a small community of mostly expert users to
-% one of the most widely used foundations of the scientific Python
-% ecosystem.
-% Now the user base is very different....
+NumPy also went from a small community of mostly expert users to
+one of the most widely used foundations of the scientific Python
+ecosystem.
+Now the user base is very different....
+% maybe worth a sentence or two about how we are working on web and
+% user docs to make it easier for new users to get started...
 
-Another ongoing example is a redesign of the data type system.
-While the majority of users rely only on numerical dtypes, there are a
-multitude of use cases for which the current implementation is unsuited, such
-as physical units\cite{astropy,Goldbaum2018,pint}, geometrical
-objects\cite{pygeos}, and automatic differentiation\cite{pyadolc}.
-The proposed overhaul of the datatype system will make it behave consistently
-and will simplify the creation of custom dtypes, both in C and in Python.
-
-\section*{Array protocols}
+\section*{Interoperability and extensibility}
 
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{static/sketches/duck arrays}
+  \includegraphics[width=0.8\textwidth]{static/sketches/duck-arrays}
   \caption{\fixme{Examples of so-called ``duck arrays'': arrays
@@
   a call such as $\cos(\cdot)$ succeed on each of these.}}
-  \label{fig:arrayconcepts}
+  \label{fig:duck-arrays}
 \end{figure*}
@@
 a \emph{NumPy-like API} and array objects targeting audiences with specialized
-needs beyond NumPy's capabilities.
+needs beyond NumPy's capabilities (Figure~\ref{fig:duck-arrays}).
 For example, the NumPy API is implemented by several popular tensor computation
@@
 stores data in the cloud.
+
+Another ongoing example is a redesign of the data type system.
+While the majority of users rely only on numerical dtypes, there are a
+multitude of use cases for which the current implementation is unsuited, such
+as physical units\cite{astropy,Goldbaum2018,pint}, geometrical
+objects\cite{pygeos}, and automatic differentiation\cite{pyadolc}.
+The proposed overhaul of the datatype system will make it behave consistently
+and will simplify the creation of custom dtypes, both in C and in Python.
 

### 2020-01-10 body.tex
Add missing link
Another pass through array structure

@@
    }
-  \label{fig:arrayconcepts}
+  \label{fig:array-concepts}
 \end{figure*}
@@
 interpret the data stored there.
-For example, while computer memory stores elements linearly and has no notion
-of higher-dimensions, we can interpret it as $n$-dimensional arrays by also
-providing its shape, strides (the number of bytes to advance in memory to skip
-along rows, columns, etc.), and data-type.
+Computer memory stores elements linearly and has no notion
+of higher-dimensions, but we can interpret it as an $n$-dimensional array by also
+providing the array's shape, strides (the number of bytes to advance in memory to skip
+along rows, columns, etc.), and data-type (see Figure~\ref{fig:array-concepts}).
 
@@
 Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
-datetimes, or Python objects, each identified by its own dtype.
+timestamps, or Python objects, each identified by its own dtype.
 
 Once data is viewed as an $n$-dimensional array, more abstract operations such
-as reductions (e.g., summation across columns), products, and so forth, are
-defined (Figure~\ref{fig:reductions}).
-Users predominantly interact with NumPy arrays using these operators and
-associated utility functions through an easily readable high-level API
-(Application Programming Interface).
+as reductions (e.g., summation across columns, see Figure~\ref{fig:reductions}) products, and so forth, are
+defined.
 
-Sometimes, it is desirable to combine arrays of different dimensions.
-One may, for example, want to add a scalar to a vector, or scale each column of
-an array by a specific scalar value.
-This is done through {\em broadcasting}, where one of the arrays is virtually
-duplicated (i.e., without copying any data in memory), so that the shapes of
-the operands match.
-Broadcasting is illustrated in Figure~\ref{fig:broadcasting}.
+Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and $@$, for matrix multiplication), together with {\em utility functions}; together, these provide an easily readable, expressive, high-level API (Application Programming Interface) for array manipulation.
+
+NumPy includes numerous functions that perform element-wise and reduction operations on arrays, including arithmetic, statistics, and trigonometry.
+They aim to loop over array elements near-optimally, taking into consideration, e.g., strides, in order to best utilize the computer's fast cache memory.
+Vectorized operations that would take many tens of lines to express in languages such as C can often be implemented as a single, clear Python expression.
+
+Other utility functions include creating, reshaping, concatenating, and padding arrays; searching, sorting and counting data; and reading and writing files.
+NumPy provides extensive support for generating pseudorandom numbers and includes an assortment of probability distributions.
+It also performs accelerated linear algebra, utilising one of several backends such as OpenBLAS and Intel MKL optimized for the CPUs at hand.
+
+%Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is a 1-d array.
+Then \code{A @ x} computes %the matrix-vector product $Ax$ and \code{x @ A} computes the vector-matrix product $x^\top A$.
+This matrix %multiplication operation extends to higher-dimensions by treating the array as a stack of matrices.
+
+Finally, a powerful array feature that affects the way NumPy indexes and combines arrays, is {\em broadcasting}.
+When performing an operation (such as addition) on two arrays, one intuitively expect that these arrays should be required to have the same shape.
+However, through broadcasting, NumPy allows the dimensions to differ, while still producing results that appeal to intuition.
+A trivial example is the addition of a scalar value to an array, but it also generalizes to more complex examples such as scaling each column of an array, or generating a grid of coordinates.
+In broadcasting, one or both arrays are virtually duplicated (that is, without copying any data in memory), so that the shapes of the operands match (see Figure~\ref{fig:array-concepts}).
+Broadcasting is also applied whenever an array is indexed using arrays of indices (see Figure~\ref{fig:broadcasting}).
 
@@
 
-NumPy include numerous functions that perform element-wise and reduction
-operations on arrays, including arithmetic, statistics, and trigonometry.
-They aim to loop over array elements near-optimally (taking into consideration,
-e.g., strides, in order to best utilize the computer's fast cache memory), and
-implement broadcasting.
-Vectorized operations that would take many tens of lines to express in
-languages such as C can often be implemented as a single, clear Python
-expression.
-
 \begin{figure}
@@
   \includegraphics[width=\linewidth]{static/sketches/reductions}
-  \caption{\fixme{Reducing an array along one of several axes.}
+  \caption{\fixme{Reducing an array along one of several axes. Note:
+  may want to fix to have right-hand axis system.}
    }
@@
 \end{figure}
-
-NumPy also provides support for accelerated linear algebra, and supports
-several established backends such as OpenBLAS and Intel MKL.
-Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is a 1-d array.
-Then \code{A @ x} computes the matrix-vector product $Ax$ and \code{x @ A} computes
-the vector-matrix product $x^\top A$.
-This matrix multiplication operation extends to higher-dimensions by treating
-the array as a stack of matrices.
-
-In addition, NumPy provides a large collection of utility functions for
-creating, reshaping, concatenating, and padding arrays; for searching, sorting
-and counting data in arrays; and for reading and writing files.
-It provides extensive support for generating pseudorandom numbers and includes
-an assortment of probability distributions.
 
@@
 % https://numpy.org/neps/nep-0030-duck-array-protocol.html
+% https://github.com/numpy/numpy/blob/a111b551ae940d7d5f8523fef1cf3589c6ba00a0/doc/neps/nep-0033-extensible-dtypes.rst
 % https://numpy.org/neps/nep-0037-array-module.html

### 2020-01-10 body.tex
Revise intro

@@
-The first incarnation of NumPy was a package called Numeric, written in 1995
-by Jim Hugunin, then completing his master's thesis at MIT.
-Hugunin based his package on previous work by Jim Fulton,
-then working at the US Geological Survey, and acknowledges
-many others in his initial report of the package \cite{Hugunin-whitepaper}.
-Later, Paul Dubois at Lawrence Livermore National Laboratory toke over the project
-and funded the writing of the first
-manual\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
-Numeric provided an array object in Python, written in C, and linking to
-standard fast implementations of linear algebra.
-% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
-% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
-% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)
-% https://aip.scitation.org/doi/pdf/10.1063/1.4823451 
-% https://aip.scitation.org/doi/pdf/10.1063/1.4822400
-
-% S TODO: the first sentence below could flow a bit better
+Two different Python array packages existed before NumPy.
+The Numeric package began in the mid-90s and provided an array object in
+Python, written in C, and linking to standard fast implementations of linear
+algebra.
 Around 2000, the Space Telescope Science Institute (STScI) software group wrote
@@
 records \cite{STScI-slither}.
-This briefly caused the Numeric and NumArray communities to diverge, until 2005,
-when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
-both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
-This merger resulted in NumPy.
+This briefly caused the two communities to diverge, until
+2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
+and NumArray \cite{oliphant2006guide}.
 
-NumPy now underpins almost every Python library that does scientific or
-numerical computation, including SciPy\cite{virtanen2019scipy},
+%By the beginning of the millennium, several libraries built on top of Numeric 
+%to provide fundamental scientific routines were combined into SciPy.
+%Due to the poor performance of Numeric on large data sets, NumArray was
+%developed at the Space Telescope Science Institute (STScI) to handle
+%large astronomical images such as those produced by the Hubble Space Telescope.
+%However, while initially considered the successor to Numeric, NumArray was no longer
+%performant on multiple small arrays.
+%This divide posed a challenge for SciPy as it wasn't clear how to support
+%both libraries.
+%Working with the NumArray and SciPy communities, the then maintainer of Numeric,
+%Travis Oliphant, embarked upon a major rewrite that aimed to be a ``best of both
+%worlds'' unification of Numeric and NumArray. This merger resulted in NumPy.
+%
+%The first incarnation of NumPy was a package called Numeric, written in 1995
+%by Jim Hugunin, then completing his master's thesis at MIT.
+%Hugunin based his package on previous work by Jim Fulton,
+%then working at the US Geological Survey, and acknowledges
+%many others in his initial report of the package \cite{Hugunin-whitepaper}.
+%Later, Paul Dubois at Lawrence Livermore National Laboratory toke over the project
+%and funded the writing of the first
+%manual\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
+%Numeric provided an array object in Python, written in C, and linking to
+%standard fast implementations of linear algebra.
+%% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
+%% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
+%% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)
+%% https://aip.scitation.org/doi/pdf/10.1063/1.4823451 
+%% https://aip.scitation.org/doi/pdf/10.1063/1.4822400
+%
+%% S TODO: the first sentence below could flow a bit better
+%Around 2000, the Space Telescope Science Institute (STScI) software group wrote
+%a re-implementation of much of Numeric, called NumArray, to support their
+%need to work on large memory-mapped arrays and arrays of mixed data type
+%records \cite{STScI-slither}.
+%This briefly caused the Numeric and NumArray communities to diverge, until 2005,
+%when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
+%both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
+%This merger resulted in NumPy.
+
+Today, NumPy underpins almost every Python library that does scientific or
+numerical computation including SciPy\cite{virtanen2019scipy},
 matplotlib\cite{hunter2007matplotlib}, pandas\cite{mckinney-proc-scipy-2010},
@@
 scikit-image\cite{vanderwalt2014scikit}.
-NumPy consists of the $n$-dimensional array object along with utility functions
+It consists of an $n$-dimensional array object along with utility functions
 that operate on it.
 Because of its inherent simplicity---being a pointer to memory with some
-associated meta-data about shape, data-type, and so forth---the NumPy array is
+associated meta-data---the NumPy array is
 the {\it de facto} exchange format for array data in Python.
@@
 {\it Application Programming Interface} (API) has become ubiquitous as
-a language for tensor computation---witnessed by its use in popular
-deep learning libraries such as PyTorch\cite{pytorch}.
+a language for array computation.
 
@@
 discrete time intervals (1-D arrays), gray-level photographs (2-D arrays),
-spectral measurement on a regular grid (3-D arrays), or color videos (4-D
+spectral measurements on a regular grid (3-D arrays), or color videos (4-D
 arrays).
@@
 providing the array's shape, strides (the number of bytes to advance in memory to skip
-along rows, columns, etc.), and data-type (see Figure~\ref{fig:array-concepts}).
+along rows, columns, etc.), and data type (see Figure~\ref{fig:array-concepts}).
 
@@
 array.
+% can hold ...?, e.g.,
 Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
@@
 As Python's popularity for scientific computing, data science, and machine
-learning increased, many new use cases arose.
+learning increased, many new challenges and use cases arose.
 

### 2020-01-10 body.tex
Simplify figure

@@
 In broadcasting, one or both arrays are virtually duplicated (that is, without copying any data in memory), so that the shapes of the operands match (see Figure~\ref{fig:array-concepts}).
-Broadcasting is also applied whenever an array is indexed using arrays of indices (see Figure~\ref{fig:broadcasting}).
+Broadcasting is also applied whenever an array is indexed using arrays of indices.% (see Figure~\ref{fig:broadcasting}).
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{static/sketches/fancy-indexing}
-  \caption{\fixme{Indexing using arrays, or ``fancy indexing''.}
-   }
-  \label{fig:broadcasting}
-\end{figure}
+%\begin{figure}
+%  \centering
+%  \includegraphics[width=\linewidth]{static/sketches/fancy-indexing}
+%  \caption{\fixme{Indexing using arrays, or ``fancy indexing''.}
+%   }
+%  \label{fig:broadcasting}
+%\end{figure}
 

### 2020-01-10 body.tex
Work on array outline
Another pass on NumPy array

@@
 The NumPy array is a structure that efficiently stores and accesses
-$n$-dimensional array data\cite{vanderwalt2011numpy}.
-The need for such a data structure arises often in science.
-For instance, it may be used to store experimental measurements made at
-discrete time intervals (1-D arrays), gray-level photographs (2-D arrays),
-spectral measurements on a regular grid (3-D arrays), or color videos (4-D
-arrays).
+$n$-dimensional arrays, also known as tensors\cite{vanderwalt2011numpy}.
+The need for such a data structure arises often in science, as it can be used to represent everything from time-series measurements to time-varying spectral measurements on a grid.
 
 The NumPy array consists of a pointer to memory, along with meta-data used to
-interpret the data stored there.
-Computer memory stores elements linearly and has no notion
-of higher-dimensions, but we can interpret it as an $n$-dimensional array by also
-providing the array's shape, strides (the number of bytes to advance in memory to skip
-along rows, columns, etc.), and data type (see Figure~\ref{fig:array-concepts}).
+interpret the data stored there, notably {\em data type}, {\em shape}, and {\em strides} (see Figure~\ref{fig:array-concepts}).
 
-Data types (dtypes for short) tell us more about the elements stored in a NumPy
-array.
-% can hold ...?, e.g.,
-Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
-timestamps, or Python objects, each identified by its own dtype.
+A {\em data type} describes the nature of elements stored in an array.
+An array has a single data type, and each array element occupies the same number of bytes in memory.
+Examples of data types include real and complex numbers (of lower and higher precision), strings, timestamps, and pointers to Python objects.
+
+The {\em shape} of an array determines the number of axes, or dimensionality, of the array.
+For example, a vector of numbers can be stored as a 1-D array of shape $N$, while color videos are 4-D arrays of shape $T \times M \times N \times 3$.
+
+{\em Strides} are necessary to interpret computer memory, which stores elements linearly, as $n$-dimensional arrays.
+It describes the number of bytes to move forward in memory to jump from row to row, column to column, and so forth.
+Consider, for example, a 2-D array of floating point numbers with shape $4 \times 3$, where each element occupies 8 bytes in memory.
+To move between consecutive columns we need to jump forward 8 bytes in memory, and to access the next row $3 \times 8 = 24$ bytes.
+The strides of that array are therefore $(24, 8)$.
+
+Portions of an array can be accessed through an {\em indexing} operation.
+Arrays can be indexed in several ways to return single elements, sub-arrays, or elements that satisfy a specific condition.
+Arrays can even be indexed using other arrays (see Figure~\ref{fig:array-concepts}).
+Wherever possible, indexing that retrieves a subarray returns a {\em view} on the original array, such that data is shared between the two arrays.
+This provides a powerful way to operate on subsets of array data, while limiting memory usage.
+
+Finally, a powerful array feature that affects the way NumPy indexes and combines arrays, is {\em broadcasting}.
+When performing an operation (such as addition) on two arrays, one intuitively expect that these arrays should be required to have the same shape.
+However, through broadcasting, NumPy allows the dimensions to differ, while still producing results that appeal to intuition.
+A trivial example is the addition of a scalar value to an array, but it also generalizes to more complex examples such as scaling each column of an array, or generating a grid of coordinates.
+In broadcasting, one or both arrays are virtually duplicated (that is, without copying any data in memory), so that the shapes of the operands match (see Figure~\ref{fig:array-concepts}).
+Broadcasting is also applied whenever an array is indexed using arrays of indices.% (see Figure~\ref{fig:broadcasting}).
 
@@
 
-Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and $@$, for matrix multiplication), together with {\em utility functions}; together, these provide an easily readable, expressive, high-level API (Application Programming Interface) for array manipulation.
+Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and, for matrix multiplication, $@$), together with {\em utility functions}; together, these provide an easily readable, expressive, high-level API (Application Programming Interface) for array manipulation.
 
@@
 This matrix %multiplication operation extends to higher-dimensions by treating the array as a stack of matrices.
-
-Finally, a powerful array feature that affects the way NumPy indexes and combines arrays, is {\em broadcasting}.
-When performing an operation (such as addition) on two arrays, one intuitively expect that these arrays should be required to have the same shape.
-However, through broadcasting, NumPy allows the dimensions to differ, while still producing results that appeal to intuition.
-A trivial example is the addition of a scalar value to an array, but it also generalizes to more complex examples such as scaling each column of an array, or generating a grid of coordinates.
-In broadcasting, one or both arrays are virtually duplicated (that is, without copying any data in memory), so that the shapes of the operands match (see Figure~\ref{fig:array-concepts}).
-Broadcasting is also applied whenever an array is indexed using arrays of indices.% (see Figure~\ref{fig:broadcasting}).
 

### 2020-01-10 body.tex
Minor changes
Add caption titles

@@
   \includegraphics[width=\textwidth]{static/sketches/array-concepts}
-  \caption{\fixme{PLACEHOLDER: This figure needs some text to describe
-  it}
+  \caption{\textbf{Fundamental Array Concepts.}
+    \textbf{a,} The NumPy array data structure ...
+    \textbf{b,} Indexing ...
+    \textbf{c,} Broadcasting ...
+    \textbf{d,} Indexing with arrays ... combined with broadcasting ...
    }
@@
 
+% shape > num. of axes and number of elements in each axis
+
 {\em Strides} are necessary to interpret computer memory, which stores elements linearly, as $n$-dimensional arrays.
@@
 The strides of that array are therefore $(24, 8)$.
+
+% on top of this structure we add a powerful, expressive syntax for indexing and element-wise
+% calculations ....
+% this high-level syntax allows us to be efficient (broadcasting) and implement in C for
+% optimizing
 
@@
 
-Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and, for matrix multiplication, $@$), together with {\em utility functions}; together, these provide an easily readable, expressive, high-level API (Application Programming Interface) for array manipulation.
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{static/sketches/reductions}
+  \caption{\textbf{Reductions.} \fixme{Reducing an array along one of several axes. Note:
+  may want to fix to have right-hand axis system.}
+   }
+  \label{fig:reductions}
+\end{figure}
 
@@
 
-%Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is a 1-d array.
-Then \code{A @ x} computes %the matrix-vector product $Ax$ and \code{x @ A} computes the vector-matrix product $x^\top A$.
-This matrix %multiplication operation extends to higher-dimensions by treating the array as a stack of matrices.
+Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and, for matrix multiplication, $@$), together with {\em utility functions}; together, these provide an easily readable, expressive, high-level API (Application Programming Interface) for array manipulation.
 
-%\begin{figure}
-%  \centering
-%  \includegraphics[width=\linewidth]{static/sketches/fancy-indexing}
-%  \caption{\fixme{Indexing using arrays, or ``fancy indexing''.}
-%   }
-%  \label{fig:broadcasting}
-%\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{static/sketches/reductions}
-  \caption{\fixme{Reducing an array along one of several axes. Note:
-  may want to fix to have right-hand axis system.}
-   }
-  \label{fig:reductions}
-\end{figure}
+% + interactivity
+% users tend to "live in their data",
+% interactivity and powerful, expressive syntax ... inspect, feel their way forward, experiment
+% collect these into imperative or functional programs ....
+% not that they never plan, often do ... but the ability to stopping "programming"
+% and experiment is a key factor in why people use numpy ....
 
 \section*{Changing computational landscape}
-
-\fixme{Ralf will take a first pass on the section on 1/12.}
-
-\fixme{The idea here is something like:  classic numpy (as described above)
-was built to solve previous problems; new problems have arisen as
-the computational landscape has changed; new hardware, new languages,
-new problems; as a result np's role has changed; it is a foundational
-tool widely-used not a specialist tool used by scientist developers;
-many array-like libraries for various uses and np is serving as
-a standard)}
 
@@
 learning increased, many new challenges and use cases arose.
+
+\fixme{The idea here is something like:  classic numpy (as described above)
+was built to solve previous problems; new problems have arisen as
+the computational landscape has changed; new hardware, new languages,
+new problems; as a result np's role has changed; it is a foundational
+tool widely-used not a specialist tool used by scientist developers;
+many array-like libraries for various uses and np is serving as
+a standard)}
 
@@
   \includegraphics[width=0.8\textwidth]{static/sketches/duck-arrays}
-  \caption{\fixme{Examples of so-called ``duck arrays'': arrays
+  \caption{\textbf{Duck Arrays.} \fixme{Examples of so-called ``duck arrays'': arrays
   implemented by another library that behave like NumPy arrays, and

### 2020-01-10 body.tex
Use array programming

@@
 {\it Application Programming Interface} (API) has become ubiquitous as
-a language for array computation.
+a language for array programming.
 
@@
 NumPy was initially developed by students, faculty, and researchers in their
-spare time to provide a modern array computation library for Python.
+spare time to provide a modern array programming library for Python.
 They were influenced by their experiences with powerful interactive programming

### 2020-01-10 summary.tex
Use array programming

@@
 % - 2-3 sentences of basic-level introduction to the field;
-Array computation libraries provide powerful notation for accessing,
+Array programming provides powerful notation for accessing,
 manipulating, and operating on data in vectors, matrices, and
@@
 % - a brief account of the background and rationale of the work;
-NumPy is the fundamental array computation library for the Python ecosystem
+NumPy is the fundamental array programming library for the Python ecosystem
 \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.

### 2020-01-10 authors.tex
More work on outline
Add authors
This is the tentative list that Ralf, Stefan, and I agreed on.
We went back to ancient history (`git shortlog -sn v0.9.2..HEAD`), and
tried to weigh merge commits (indicative of review effort) and other
contributions like release management, docs, mentoring, and so on.

It is open to discussion, but we think it is fairly reasonable.

--- 2019-12-06 authors.tex
@@
-\author[1]{Jane Doe}
-\affil[1]{Berkeley Institute for Data Science, University of California, Berkeley, CA 94720, USA}
+\author[1]{Charles R Harris}
+\author[2,3,*]{K. Jarrod Millman}
+\author[2,*]{St\'efan J. van der Walt}
+\author[4,*]{Ralf Gommers}
+\author[8]{Pauli Virtanen}
+\author[8]{David Cournapeau}
+\author[8]{Eric Wieser}
+\author[8]{Julian Taylor}
+\author[2]{Sebastian Berg}
+\author[8]{Nathaniel J. Smith}
+\author[8]{Robert Kern}
+\author[8]{Matti Picus}
+\author[8]{Stephan Hoyer}
+\author[8]{Marten van Kerkwijk}
+\author[7]{Matthew Brett}
+\author[8]{Allan Haldane}
+\author[8]{Jaime Fernandez del Rio}
+\author[8]{Mark Wiebe}
+\author[8]{Pearu Peterson}
+\author[8]{David M. Cooke}
+\author[8]{Pierre Gerard-Marchant}
+\author[5]{Kevin Sheppard}
+\author[6]{Tyler J. Reddy}
+\author[2]{Warren Weckesser}
+\author[8]{Hameer Abbasi}
+\author[8]{Christoph Gohlke}
+\author[4]{Travis E. Oliphant}
+\affil[1]{Independent Researcher}
+\affil[2]{Berkeley Institute for Data Science, University of California, Berkeley, CA, USA}
+\affil[3]{Division of Biostatistics, University of California, Berkeley, CA, USA}
+\affil[4]{Quansight LLC, Austin, TX, USA}
+\affil[5]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[6]{Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[7]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
+\affil[8]{These authors still need to review paper!}
+\affil[*]{numpy.articles@gmail.com}

### 2020-01-10 body.tex
Minor edit
Another round of edits on the array section

@@
 Today, NumPy underpins almost every Python library that does scientific or
-numerical computation including SciPy\cite{virtanen2019scipy},
-matplotlib\cite{hunter2007matplotlib}, pandas\cite{mckinney-proc-scipy-2010},
-scikit-learn\cite{pedregosa2011scikit}, and
-scikit-image\cite{vanderwalt2014scikit}.
+numerical computation including SciPy \cite{virtanen2019scipy},
+matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
+scikit-learn \cite{pedregosa2011scikit}, and
+scikit-image \cite{vanderwalt2014scikit}.
 It consists of an $n$-dimensional array object along with utility functions
@@
 
-The NumPy array is a structure that efficiently stores and accesses
-$n$-dimensional arrays, also known as tensors\cite{vanderwalt2011numpy}.
-The need for such a data structure arises often in science, as it can be used to represent everything from time-series measurements to time-varying spectral measurements on a grid.
-
-The NumPy array consists of a pointer to memory, along with meta-data used to
-interpret the data stored there, notably {\em data type}, {\em shape}, and {\em strides} (see Figure~\ref{fig:array-concepts}).
+The NumPy array is a structure that efficiently stores and accesses $n$-dimensional arrays, also known as tensors \cite{vanderwalt2011numpy}, that enables a wide variety of scientific computation.
+It consists of a pointer to memory, along with meta-data used to interpret the data stored there, notably a {\em data type}, {\em shape}, and {\em strides} (see Figure~\ref{fig:array-concepts}).
 
@@
 
-The {\em shape} of an array determines the number of axes, or dimensionality, of the array.
-For example, a vector of numbers can be stored as a 1-D array of shape $N$, while color videos are 4-D arrays of shape $T \times M \times N \times 3$.
-
-% shape > num. of axes and number of elements in each axis
+The {\em shape} of an array determines the number of elements along each axis, and the number of axes is the array's dimensionality.
+For example, a vector of numbers can be stored as a one dimensional array of shape $N$, while color videos are four dimensional arrays of shape $T \times M \times N \times 3$.
 
@@
 
-% on top of this structure we add a powerful, expressive syntax for indexing and element-wise
-% calculations ....
-% this high-level syntax allows us to be efficient (broadcasting) and implement in C for
-% optimizing
+To manipulate the data structure in an intuitive manner, we provide users with a simple yet powerful array expression syntax.
+It supports operations such as indexing (selecting subarrays) and element-wise calculations.
+A high level syntax allows users to express themselves succinctly, while NumPy deals with the underlying mechanics of making operations fast.
 
-Portions of an array can be accessed through an {\em indexing} operation.
-Arrays can be indexed in several ways to return single elements, sub-arrays, or elements that satisfy a specific condition.
+Indexing an array returns single elements, sub-arrays, or elements that satisfy a specific condition.
 Arrays can even be indexed using other arrays (see Figure~\ref{fig:array-concepts}).
@@
 
-Finally, a powerful array feature that affects the way NumPy indexes and combines arrays, is {\em broadcasting}.
+A powerful array feature that affects the way NumPy indexes and combines arrays, is {\em broadcasting}.
 When performing an operation (such as addition) on two arrays, one intuitively expect that these arrays should be required to have the same shape.
@@
 Broadcasting is also applied whenever an array is indexed using arrays of indices.% (see Figure~\ref{fig:broadcasting}).
-
-Once data is viewed as an $n$-dimensional array, more abstract operations such
-as reductions (e.g., summation across columns, see Figure~\ref{fig:reductions}) products, and so forth, are
-defined.
 
@@
 
-NumPy includes numerous functions that perform element-wise and reduction operations on arrays, including arithmetic, statistics, and trigonometry.
+%Once data is viewed as an $n$-dimensional array, more abstract operations such
+%as reductions can be defined.
+
+%(e.g., summation across columns, see Figure~\ref{fig:reductions}) products, and %so forth, are
+%defined.
+
+To complement the array syntax, NumPy includes several functions that perform element-wise calculations on arrays, including arithmetic, statistics, and trigonometry.
 They aim to loop over array elements near-optimally, taking into consideration, e.g., strides, in order to best utilize the computer's fast cache memory.
 Vectorized operations that would take many tens of lines to express in languages such as C can often be implemented as a single, clear Python expression.
+
+Many of these functions, such as summation, support reductions: aggregating results across over or more dimension of the array.
+For example, summing a $D$-dimensional array over $p$ axes results in a $D-p$-dimensional array, as shown in Figure~\ref{fig:reductions}.
 
@@
 
-Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and, for matrix multiplication, $@$), together with {\em utility functions}; together, these provide an easily readable, expressive, high-level API (Application Programming Interface) for array manipulation.
+Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and, for matrix multiplication, $@$), as well as {\em utility functions}; together, these provide an easily readable, expressive, high-level API for array programming.
+Exposing these array programming primitives in Python---an interpreted language---creates an ideal environment for exploratory data analysis \cite{perez2007ipython} where users may inspect, manipulate, and visualize their data.
+
+\fixme{This end of paragraph needs fixing up}Instead of focusing on programming, users can bring their attention to understanding and interacting with their data.
 
@@
 API for \emph{interoperability} between the various NumPy-like APIs.
-
 For example, a new array protocol was implemented that allows NumPy to consume
@@
 
-Another ongoing example is a redesign of the data type system.
+NumPy also aims to improve our \emph{extensibility}.
+For example, we are currently redesigning the data type system.
 While the majority of users rely only on numerical dtypes, there are a
 multitude of use cases for which the current implementation is unsuited, such
-as physical units\cite{astropy,Goldbaum2018,pint}, geometrical
-objects\cite{pygeos}, and automatic differentiation\cite{pyadolc}.
+as physical units \cite{astropy,Goldbaum2018,pint}, geometrical
+objects \cite{pygeos}, and automatic differentiation \cite{pyadolc}.
 The proposed overhaul of the datatype system will make it behave consistently
@@
 integrated with the source
-code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
+code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
 

### 2020-01-11 body.tex
Minor edits

@@
 Around 2000, the Space Telescope Science Institute (STScI) software group wrote
-a re-implementation of much of Numeric, called NumArray, to support their
-need to work on large memory-mapped arrays and arrays of mixed data type
+a re-implementation of much of Numeric, called NumArray, to support their work
+on large memory-mapped arrays and arrays of mixed data type
 records \cite{STScI-slither}.
@@
 and NumArray \cite{oliphant2006guide}.
-
-%By the beginning of the millennium, several libraries built on top of Numeric 
-%to provide fundamental scientific routines were combined into SciPy.
-%Due to the poor performance of Numeric on large data sets, NumArray was
-%developed at the Space Telescope Science Institute (STScI) to handle
-%large astronomical images such as those produced by the Hubble Space Telescope.
-%However, while initially considered the successor to Numeric, NumArray was no longer
-%performant on multiple small arrays.
-%This divide posed a challenge for SciPy as it wasn't clear how to support
-%both libraries.
-%Working with the NumArray and SciPy communities, the then maintainer of Numeric,
-%Travis Oliphant, embarked upon a major rewrite that aimed to be a ``best of both
-%worlds'' unification of Numeric and NumArray. This merger resulted in NumPy.
-%
-%The first incarnation of NumPy was a package called Numeric, written in 1995
-%by Jim Hugunin, then completing his master's thesis at MIT.
-%Hugunin based his package on previous work by Jim Fulton,
-%then working at the US Geological Survey, and acknowledges
-%many others in his initial report of the package \cite{Hugunin-whitepaper}.
-%Later, Paul Dubois at Lawrence Livermore National Laboratory toke over the project
-%and funded the writing of the first
-%manual\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
-%Numeric provided an array object in Python, written in C, and linking to
-%standard fast implementations of linear algebra.
-%% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
-%% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
-%% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)
-%% https://aip.scitation.org/doi/pdf/10.1063/1.4823451 
-%% https://aip.scitation.org/doi/pdf/10.1063/1.4822400
-%
-%% S TODO: the first sentence below could flow a bit better
-%Around 2000, the Space Telescope Science Institute (STScI) software group wrote
-%a re-implementation of much of Numeric, called NumArray, to support their
-%need to work on large memory-mapped arrays and arrays of mixed data type
-%records \cite{STScI-slither}.
-%This briefly caused the Numeric and NumArray communities to diverge, until 2005,
-%when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
-%both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
-%This merger resulted in NumPy.
 
@@
 
-% S: Maybe we want to emphasize Jax, which has rapidly become very popular
-
 \section*{NumPy arrays}
@@
 
-The NumPy array is a structure that efficiently stores and accesses $n$-dimensional arrays, also known as tensors \cite{vanderwalt2011numpy}, that enables a wide variety of scientific computation.
-It consists of a pointer to memory, along with meta-data used to interpret the data stored there, notably a {\em data type}, {\em shape}, and {\em strides} (see Figure~\ref{fig:array-concepts}).
+The NumPy array is a data structure that efficiently stores and accesses $n$-dimensional arrays, also known as tensors \cite{vanderwalt2011numpy}, that enables a wide variety of scientific computation.
+It consists of a pointer to memory, along with meta-data used to interpret the data stored there, notably {\em data type}, {\em shape}, and {\em strides} (see Figure~\ref{fig:array-concepts}).
 
-A {\em data type} describes the nature of elements stored in an array.
+The {\em data type} describes the nature of elements stored in an array.
 An array has a single data type, and each array element occupies the same number of bytes in memory.
@@
 
-\fixme{Similarly there are many projects that build on top of the NumPy API for
-labeled and indexed arrays (XArray), automatic differentiation (Autograd,
-Tangent), masked arrays (numpy.ma), physical units (astropy.units, pint, unyt),
-etc. that add additional functionality on top of the NumPy API. Most of these
-project also implement a close variation of NumPy’s level high API.}
-
 The multitude of specialized projects creates the difficulty that consumers
@@
 stores data in the cloud.
+
+\fixme{Similarly there are many projects that build on top of the NumPy API for
+labeled and indexed arrays (XArray), automatic differentiation (Autograd,
+Tangent), masked arrays (numpy.ma), physical units (astropy.units, pint, unyt),
+etc. that add additional functionality on top of the NumPy API. Most of these
+project also implement a close variation of NumPy’s level high API.}
 

### 2020-01-11 body.tex
Add text from Sebastian

@@
 
+While NumPy remains the fundamental tool of choice for a large portion
+of applications the growth of the computational landscape and methods
+cannot be supported within a single library.
+However, the NumPy API is a preferred language for array computation
+and often one of the first tools learned.
+
 A vast number of projects are built on NumPy;
@@
 API for \emph{interoperability} between the various NumPy-like APIs.
+We believe that improving the interoperability and moving NumPy forward
+to serve new arising needs where possible is the best way to balance
+the need for stability on the one hand while supporting the growth of
+specialized solutions and new tools on the other.
+
 For example, a new array protocol was implemented that allows NumPy to consume
@@
 numerical languages can encourage consumer-users, who pick up a tool and use
-it, but do not expect to contribute to development%<-sentence a bit tough to parse.
+it, but do not expect to contribute to development. %<-sentence a bit tough to parse.
 In contrast, NumPy often

### 2020-01-11 body.tex
Re-draft interoperability section

--- 2020-01-11 body.tex
@@
 
-\section*{Interoperability and extensibility}
+\section*{Extensibility and Interoperability}
 
@@
 
+NumPy's API has become ubiquitous in scientific computation, but NumPy itself can never hope to satisfy every single specialized need of the community.
+The advent of new technologies, such as GPUs, create opportunities that should be exploited, but are out of scope for NumPy development.
+Still, the widespread adoption of the NumPy API is valuable, as it lowers the barrier to entry and provides stability to the community.
+What can NumPy do to support the use of its API without blocking innovation?
+We believe that improving interoperability by having NumPy be a central coordinator among array objects is a good way to balance stability while supporting growth of a larger surrounding ecosystem of specialized solutions and new tools.
+
+Primarily, there exist two types of Python array objects: subclasses of NumPy arrays, and arrays that mimic NumPy arrays but are fundamentally different.
+
+Subclasses of NumPy often exist due to difficulty in constructing richer data types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint}, geometrical objects \cite{pygeos}, and missing numbers.
+To improve {\em extensibility}, we are currently overhauling the data type system to make it easier to implement data types both in Python and C, and to support these and other applications.
+
+There has been a steady increase in the number of external libraries that mimic NumPy, i.e.
+provide arrays and a NumPy-like API for manipulating them.
+This includes popular tensor computation libraries such as CuPy\footnote{\url{https://cupy.chainer.org/}}, JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}, and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}} and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}} provide tensor APIs with NumPy-inspired semantics.
+Ideally, operating on these arrays using NumPy would simply work, so that end users could write code once, and would then benefit from switching between NumPy arrays, GPU arrays, distributed arrays, and so forth, as appropriate.
+
+To facilitate {\em interoperability}, the NumPy team is defining so-called ``protocols'' (or contracts of operation), that allow for these arrays to be passed to NumPy functions.
+NumPy, in turn, dispatches operations to the originating library, as required.
+
+\url{https://numpy.org/neps/nep-0037-array-module.html}
+
 % https://numpy.org/neps/nep-0016-abstract-array.html
@@
 % https://numpy.org/neps/nep-0037-array-module.html
-
-A vast number of projects are built on NumPy;
-these projects are consumers of the NumPy API.
-Over the last several years, a growing number of projects are providers of
-a \emph{NumPy-like API} and array objects targeting audiences with specialized
-needs beyond NumPy's capabilities (Figure~\ref{fig:duck-arrays}).
-For example, the NumPy API is implemented by several popular tensor computation
-libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
-JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
-and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
-and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
-provide tensor APIs with NumPy-inspired semantics.
-It is also implemented in packages that support sparse arrays
-such as \code{scipy.sparse} and \code{pydata.sparse}.
-Another notable example is Dask, a library for parallel computing in
-Python.  Dask adopts the NumPy API and therefore presents a familiar
-interface to existing NumPy users, while adding powerful abilities to
-parallelize and distribute tasks.
-
-The multitude of specialized projects creates the difficulty that consumers
-of these NumPy-like APIs write code specific to a single project and do not support
-all of the above array providers.
-This is a burden for users relying on the specialized array-like, since
-a tool they need may not work for them.
-It also creates new challenges for end-users who need to transition
-from NumPy to a more specialized array.
-The growing multitude of specialized projects with NumPy-like APIs threatened
-to again fracture the scientific Python community.
-
-\fixme{We would like to be able to use these libraries together, for example we would
-like to be able to place a CuPy array within XArray, or perform automatic
-differentiation on Dask array code. This would be easier to accomplish if code
-written for NumPy ndarrays could also be used by other NumPy-like projects.}
-
-To address these issues NumPy has the goal of providing the fundamental
-API for \emph{interoperability} between the various NumPy-like APIs.
-For example, a new array protocol was implemented that allows NumPy to consume
-array-like objects produced by libraries outside of NumPy.  When asking NumPy
-to, e.g., sum such an array, instead of attempting that operation itself, NumPy
-requests the original array to complete the computation.  This means that the
-NumPy API---the syntax for interacting with arrays---can still be used, but now
-with a library that supports GPUs, does distributed computation processing, or
-stores data in the cloud.
-
-\fixme{Similarly there are many projects that build on top of the NumPy API for
-labeled and indexed arrays (XArray), automatic differentiation (Autograd,
-Tangent), masked arrays (numpy.ma), physical units (astropy.units, pint, unyt),
-etc. that add additional functionality on top of the NumPy API. Most of these
-project also implement a close variation of NumPy’s level high API.}
-
-NumPy also aims to improve our \emph{extensibility}.
-For example, we are currently redesigning the data type system.
-While the majority of users rely only on numerical dtypes, there are a
-multitude of use cases for which the current implementation is unsuited, such
-as physical units \cite{astropy,Goldbaum2018,pint}, geometrical
-objects \cite{pygeos}, and automatic differentiation \cite{pyadolc}.
-The proposed overhaul of the datatype system will make it behave consistently
-and will simplify the creation of custom dtypes, both in C and in Python.
 

### 2020-01-11 body.tex
Minor revisions

@@
-Two different Python array packages existed before NumPy.
+Two Python array packages existed before NumPy.
 The Numeric package began in the mid-90s and provided an array object in
@@
 
-The NumPy array is a data structure that efficiently stores and accesses $n$-dimensional arrays, also known as tensors \cite{vanderwalt2011numpy}, that enables a wide variety of scientific computation.
-It consists of a pointer to memory, along with meta-data used to interpret the data stored there, notably {\em data type}, {\em shape}, and {\em strides} (see Figure~\ref{fig:array-concepts}).
+The NumPy array is a data structure that efficiently stores and accesses
+$n$-dimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, that
+enables a wide variety of scientific computation.
+It consists of a pointer to memory, along with meta-data used to interpret the
+data stored there, notably {\em data type}, {\em shape}, and {\em strides}
+(Fig.~\ref{fig:array-concepts}a).
 
 The {\em data type} describes the nature of elements stored in an array.
-An array has a single data type, and each array element occupies the same number of bytes in memory.
-Examples of data types include real and complex numbers (of lower and higher precision), strings, timestamps, and pointers to Python objects.
+An array has a single data type, and each array element occupies the same
+number of bytes in memory.
+Examples of data types include real and complex numbers (of lower and higher
+precision), strings, timestamps, and pointers to Python objects.
 
-The {\em shape} of an array determines the number of elements along each axis, and the number of axes is the array's dimensionality.
-For example, a vector of numbers can be stored as a one dimensional array of shape $N$, while color videos are four dimensional arrays of shape $T \times M \times N \times 3$.
+The {\em shape} of an array determines the number of elements along each axis,
+and the number of axes is the array's dimensionality.
+For example, a vector of numbers can be stored as a one dimensional array of
+shape $N$, while color videos are four dimensional arrays of shape
+$T \times M \times N \times 3$.
 
-{\em Strides} are necessary to interpret computer memory, which stores elements linearly, as $n$-dimensional arrays.
-It describes the number of bytes to move forward in memory to jump from row to row, column to column, and so forth.
-Consider, for example, a 2-D array of floating point numbers with shape $4 \times 3$, where each element occupies 8 bytes in memory.
-To move between consecutive columns we need to jump forward 8 bytes in memory, and to access the next row $3 \times 8 = 24$ bytes.
+{\em Strides} are necessary to interpret computer memory, which stores elements
+linearly, as $n$-dimensional arrays.
+It describes the number of bytes to move forward in memory to jump from row to
+row, column to column, and so forth.
+Consider, for example, a 2-D array of floating point numbers with shape
+$4 \times 3$, where each element occupies 8 bytes in memory.
+To move between consecutive columns we need to jump forward 8 bytes in memory,
+and to access the next row $3 \times 8 = 24$ bytes.
 The strides of that array are therefore $(24, 8)$.
 
-To manipulate the data structure in an intuitive manner, we provide users with a simple yet powerful array expression syntax.
-It supports operations such as indexing (selecting subarrays) and element-wise calculations.
-A high level syntax allows users to express themselves succinctly, while NumPy deals with the underlying mechanics of making operations fast.
+To manipulate the data structure in an intuitive manner, we provide users with
+a simple yet powerful array expression syntax.
+It supports operations such as indexing (selecting subarrays) and element-wise
+calculations.
+A high level syntax allows users to express themselves succinctly, while NumPy
+deals with the underlying mechanics of making operations fast.
 
-Indexing an array returns single elements, sub-arrays, or elements that satisfy a specific condition.
-Arrays can even be indexed using other arrays (see Figure~\ref{fig:array-concepts}).
-Wherever possible, indexing that retrieves a subarray returns a {\em view} on the original array, such that data is shared between the two arrays.
-This provides a powerful way to operate on subsets of array data, while limiting memory usage.
+Indexing an array returns single elements, sub-arrays, or elements that satisfy
+a specific condition (Fig.~\ref{fig:array-concepts}b).
+Arrays can even be indexed using other arrays (Fig.~\ref{fig:array-concepts}d).
+Wherever possible, indexing that retrieves a subarray returns a {\em view} on
+the original array, such that data is shared between the two arrays.
+This provides a powerful way to operate on subsets of array data, while
+limiting memory usage.
 
-A powerful array feature that affects the way NumPy indexes and combines arrays, is {\em broadcasting}.
-When performing an operation (such as addition) on two arrays, one intuitively expect that these arrays should be required to have the same shape.
-However, through broadcasting, NumPy allows the dimensions to differ, while still producing results that appeal to intuition.
-A trivial example is the addition of a scalar value to an array, but it also generalizes to more complex examples such as scaling each column of an array, or generating a grid of coordinates.
-In broadcasting, one or both arrays are virtually duplicated (that is, without copying any data in memory), so that the shapes of the operands match (see Figure~\ref{fig:array-concepts}).
-Broadcasting is also applied whenever an array is indexed using arrays of indices.% (see Figure~\ref{fig:broadcasting}).
+A powerful array feature that affects the way NumPy indexes and combines
+arrays, is {\em broadcasting}.
+When performing an operation (such as addition) on two arrays, one intuitively
+expect that these arrays should be required to have the same shape.
+However, through broadcasting, NumPy allows the dimensions to differ, while
+still producing results that appeal to intuition.
+A trivial example is the addition of a scalar value to an array, but it also
+generalizes to more complex examples such as scaling each column of an array,
+or generating a grid of coordinates.
+In broadcasting, one or both arrays are virtually duplicated (that is, without
+copying any data in memory), so that the shapes of the operands match
+(Fig.~\ref{fig:array-concepts}c).
+Broadcasting is also applied whenever an array is indexed using arrays of
+indices (Fig.~\ref{fig:array-concepts}d).
 
@@
 
-%(e.g., summation across columns, see Figure~\ref{fig:reductions}) products, and %so forth, are
+%(e.g., summation across columns, see Fig.~\ref{fig:reductions}) products, and %so forth, are
 %defined.
 
-To complement the array syntax, NumPy includes several functions that perform element-wise calculations on arrays, including arithmetic, statistics, and trigonometry.
-They aim to loop over array elements near-optimally, taking into consideration, e.g., strides, in order to best utilize the computer's fast cache memory.
-Vectorized operations that would take many tens of lines to express in languages such as C can often be implemented as a single, clear Python expression.
+To complement the array syntax, NumPy includes several functions that perform
+element-wise calculations on arrays, including arithmetic, statistics, and
+trigonometry.
+They aim to loop over array elements near-optimally, taking into consideration,
+e.g., strides, in order to best utilize the computer's fast cache memory.
+Vectorized operations that would take many tens of lines to express in
+languages such as C can often be implemented as a single, clear Python
+expression.
 
-Many of these functions, such as summation, support reductions: aggregating results across over or more dimension of the array.
-For example, summing a $D$-dimensional array over $p$ axes results in a $D-p$-dimensional array, as shown in Figure~\ref{fig:reductions}.
+Many of these functions, such as summation, support reductions: aggregating
+results across over or more dimension of the array.
+For example, summing a $D$-dimensional array over $p$ axes results in a
+$D-p$-dimensional array (Fig.~\ref{fig:reductions}).
 
-Other utility functions include creating, reshaping, concatenating, and padding arrays; searching, sorting and counting data; and reading and writing files.
-NumPy provides extensive support for generating pseudorandom numbers and includes an assortment of probability distributions.
-It also performs accelerated linear algebra, utilising one of several backends such as OpenBLAS and Intel MKL optimized for the CPUs at hand.
+Other utility functions include creating, reshaping, concatenating, and padding
+arrays; searching, sorting and counting data; and reading and writing files.
+NumPy provides extensive support for generating pseudorandom numbers and
+includes an assortment of probability distributions.
+It also performs accelerated linear algebra, utilising one of several backends
+such as OpenBLAS and Intel MKL optimized for the CPUs at hand.
 
-Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and, for matrix multiplication, $@$), as well as {\em utility functions}; together, these provide an easily readable, expressive, high-level API for array programming.
-Exposing these array programming primitives in Python---an interpreted language---creates an ideal environment for exploratory data analysis \cite{perez2007ipython} where users may inspect, manipulate, and visualize their data.
+Users predominantly interact with NumPy arrays using {\em indexing} (to access
+sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
+and, for matrix multiplication, $@$), as well as {\em utility functions};
+together, these provide an easily readable, expressive, high-level API for
+array programming.
+Exposing these array programming primitives in Python---an interpreted
+language---creates an ideal environment for exploratory data analysis where
+users may inspect, manipulate, and visualize their data
+\cite{perez2007ipython}.
 
-\fixme{This end of paragraph needs fixing up}Instead of focusing on programming, users can bring their attention to understanding and interacting with their data.
-
-% + interactivity
-% users tend to "live in their data",
-% interactivity and powerful, expressive syntax ... inspect, feel their way forward, experiment
-% collect these into imperative or functional programs ....
-% not that they never plan, often do ... but the ability to stopping "programming"
-% and experiment is a key factor in why people use numpy ....
-
-\section*{Changing computational landscape}
-
-NumPy was initially developed by students, faculty, and researchers in their
-spare time to provide a modern array programming library for Python.
-They were influenced by their experiences with powerful interactive programming
-languages for scientific computing like Yorick \cite{munro1995using} as well
-as commercial languages like IDL and Matlab.
-Often they developed code to solve their own or their colleagues' problems.
-
-As Python's popularity for scientific computing, data science, and machine
-learning increased, many new challenges and use cases arose.
-
-\fixme{The idea here is something like:  classic numpy (as described above)
-was built to solve previous problems; new problems have arisen as
-the computational landscape has changed; new hardware, new languages,
-new problems; as a result np's role has changed; it is a foundational
-tool widely-used not a specialist tool used by scientist developers;
-many array-like libraries for various uses and np is serving as
-a standard)}
-
-- GPUs
-
-- distributed computing
-
-- other languages w/ arrays
-
-- massive explosion in data science, machine learning, and artificial intelligence
-
-NumPy also went from a small community of mostly expert users to
-one of the most widely used foundations of the scientific Python
-ecosystem.
-Now the user base is very different....
-% maybe worth a sentence or two about how we are working on web and
-% user docs to make it easier for new users to get started...
-
-\section*{Extensibility and Interoperability}
+\section*{The proliferation of arrays}
 
@@
   \includegraphics[width=0.8\textwidth]{static/sketches/duck-arrays}
-  \caption{\textbf{Duck Arrays.} \fixme{Examples of so-called ``duck arrays'': arrays
-  implemented by another library that behave like NumPy arrays, and
-  support NumPy operations.  Through protocols, it is possible to have
-  a call such as $\cos(\cdot)$ succeed on each of these.}}
-  \label{fig:duck-arrays}
+  \caption{\textbf{Proliferation of Arrays.} Many ...}
+  \label{fig:array-proliferation}
 \end{figure*}
 
-NumPy's API has become ubiquitous in scientific computation, but NumPy itself can never hope to satisfy every single specialized need of the community.
-The advent of new technologies, such as GPUs, create opportunities that should be exploited, but are out of scope for NumPy development.
-Still, the widespread adoption of the NumPy API is valuable, as it lowers the barrier to entry and provides stability to the community.
+(Fig.~\ref{fig:array-proliferation})
+
+- distributed arrays
+
+- GPUs
+
+- computation graph
+
+- sparse arrays
+
+- other languages w/ arrays
+
+- massive explosion in data science, machine learning, and artificial intelligence (tensors)
+
+\section*{Extensibility and Interoperability}
+
+NumPy's API has become ubiquitous in scientific computation, but NumPy itself
+can never hope to satisfy every single specialized need of the community.
+The advent of new technologies, such as GPUs, create opportunities that should be
+exploited, but are out of scope for NumPy development.
+Still, the widespread adoption of the NumPy API is valuable, as it lowers the
+barrier to entry and provides stability to the community.
 What can NumPy do to support the use of its API without blocking innovation?
-We believe that improving interoperability by having NumPy be a central coordinator among array objects is a good way to balance stability while supporting growth of a larger surrounding ecosystem of specialized solutions and new tools.
+We believe that improving interoperability by having NumPy be a central
+coordinator among array objects is a good way to balance stability while
+supporting growth of a larger surrounding ecosystem of specialized solutions
+and new tools.
+Primarily, there exist two types of Python array objects: subclasses of NumPy
+arrays, and arrays that mimic NumPy arrays but are fundamentally different.
 
-Primarily, there exist two types of Python array objects: subclasses of NumPy arrays, and arrays that mimic NumPy arrays but are fundamentally different.
+Subclasses of NumPy often exist due to difficulty in constructing richer data
+types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
+geometrical objects \cite{pygeos}, and missing numbers.
+To improve \emph{extensibility}, we are currently overhauling the data type
+system to make it easier to implement data types both in Python and C, and to
+support these and other applications.
 
-Subclasses of NumPy often exist due to difficulty in constructing richer data types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint}, geometrical objects \cite{pygeos}, and missing numbers.
-To improve {\em extensibility}, we are currently overhauling the data type system to make it easier to implement data types both in Python and C, and to support these and other applications.
+There has been a steady increase in the number of external libraries that mimic
+NumPy, i.e. provide arrays and a NumPy-like API for manipulating them.
+This includes popular tensor computation libraries such as
+CuPy\footnote{\url{https://cupy.chainer.org/}},
+JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}, and
+Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
+and
+Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
+provide tensor APIs with NumPy-inspired semantics.
+Ideally, operating on these arrays using NumPy would simply work, so that end
+users could write code once, and would then benefit from switching between
+NumPy arrays, GPU arrays, distributed arrays, and so forth, as appropriate.
 
-There has been a steady increase in the number of external libraries that mimic NumPy, i.e.
-provide arrays and a NumPy-like API for manipulating them.
-This includes popular tensor computation libraries such as CuPy\footnote{\url{https://cupy.chainer.org/}}, JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}, and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}} and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}} provide tensor APIs with NumPy-inspired semantics.
-Ideally, operating on these arrays using NumPy would simply work, so that end users could write code once, and would then benefit from switching between NumPy arrays, GPU arrays, distributed arrays, and so forth, as appropriate.
-
-To facilitate {\em interoperability}, the NumPy team is defining so-called ``protocols'' (or contracts of operation), that allow for these arrays to be passed to NumPy functions.
+To facilitate \emph{interoperability}, the NumPy team is defining so-called
+``protocols'' (or contracts of operation), that allow for these arrays to be
+passed to NumPy functions.
 NumPy, in turn, dispatches operations to the originating library, as required.
 
-\url{https://numpy.org/neps/nep-0037-array-module.html}
+%\url{https://numpy.org/neps/nep-0037-array-module.html}
 
@@
 
-The creation of NumPy spurred renewed development in the larger scientific
-Python ecosystem and heralded the current era of wide-spread use of Python for
-scientific computing.
-There were several factors that allowed this rapid growth
-and successful development.
+NumPy was initially developed by students, faculty, and researchers in their
+spare time to provide a modern array programming library for Python.
+They were influenced by their experiences with powerful interactive programming
+languages for scientific computing like Yorick \cite{munro1995using} as well
+as commercial languages like IDL and Matlab.
+Often they developed code to solve their own or their colleagues' problems.
 
-Hugunin identified the first factor in his initial description of the Numeric
-package \cite{Hugunin-whitepaper}.  Numerical computing is usually one
-component of a larger programming task, therefore:
-\begin{quote}
-    Rather than trying to retrofit an existing numerical language to support
-    the wealth of features found in a powerful, modern, general-purpose
-    programming language, it makes much more sense to attack the problem from
-    the other direction and add the features of a powerful numerical
-    programming language to Python.
-\end{quote}
+NumPy spurred the development of the larger scientific Python ecosystem and
+heralded the current era of wide-spread use of Python for scientific computing.
+There were several factors that allowed this rapid growth and successful
+development.
+
 Python is already a good choice for many standard programming tasks such as
-cleaning data, interacting with web resources and parsing text.  It has a wide
-range of libraries for many different tasks. Adding fast array operations and
-linear algebra allows the scientist to do all their work with within the same
-language---and one that has the advantage of being famously easy to learn and
-teach.
+cleaning data, interacting with web resources and parsing text.
+It has a wide range of libraries for many different tasks.
+Adding fast array operations and linear algebra allows the scientist to do all
+their work with within the same language---and one that has the advantage of
+being famously easy to learn and teach.
 
-The second factor flows from Python's nature as an open-source language,
-embedded within the open-source community.  Python attracts developers who like
-to build and contribute.  As a result, NumPy has always been a library that was
-developed by scientists in order to do their work.  Commercial specialist
-numerical languages can encourage consumer-users, who pick up a tool and use
-it, but do not expect to contribute to development. %<-sentence a bit tough to parse.
-In contrast, NumPy often
-attracts user-developers, who want to work with NumPy, and are prepared to fix
-and extend the library as they work. User-developers have been the fundamental
-force in developing NumPy, and shaping it into a library that has been refined
-in the furnace of practical scientific work.
+NumPy has strong culture of using software-engineering practice to
+improve collaboration and reduce error \cite{millman2014developing}.
+The NumPy team was early in adopting distributed revision control and code
+review to improve collaboration on code, and continuous testing that runs a
+large battery of automated tests for every proposed change to NumPy.
+The project has comprehensive, high-quality documentation, integrated with the
+source code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
 
-The third factor is much like the second: 
-%a culture in Python that concentrates
-%on high quality of code and distribution.  Because Python is open-source, it has
-%a strong culture of contribution, and therefore, of using tools and process to
-%improve collaborative software development, such as distributed version
-%control, code review, and automated testing.  These have allowed the project to
-%scale safely as it attracted more users and developers, even though these
-%developers rarely came to the project with training in software engineering.
-the project has strong culture of using software-engineering practice to
-improve collaboration and reduce error \cite{millman2014developing}.
-The NumPy team was early in adopting distributed
-revision control and code review to improve collaboration on code, and
-continuous testing that runs a large battery of automated tests for every proposed
-change to NumPy.
-The project has comprehensive, high-quality documentation,
-integrated with the source
-code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
-
-Lastly, we believe NumPy has benefited greatly from a well-chosen application
-programming interface (API).  Scientific and numerical programming needs to be
-fast, and scale to very large datasets.  The NumPy API defines a simple wrapper
-for data in memory so it can be represented as a one- or multi-dimensional
-array.  The simplicity of this wrapper has made it very successful as a
-standard way of representing arrays in memory, and has made it relatively easy
-for other libraries to develop fast and memory-efficient compiled code, usually
-in C or Fortran, that can manipulate these arrays and pass them back to Python.
+Lastly, NumPy has benefited greatly from a well-chosen API.
+Scientific and numerical programming needs to be fast, and scale to very large
+datasets.
+The NumPy API defines a simple wrapper for data in memory so it can be
+represented as a one- or multi-dimensional array.
+The simplicity of this wrapper has made it very successful as a standard way of
+representing arrays in memory, and has made it relatively easy for other
+libraries to develop fast and memory-efficient compiled code, usually in C or
+Fortran, that can manipulate these arrays and pass them back to Python.
 This has been a large factor in the growth of numerical computing libraries

### 2020-01-11 summary.tex
Minor revisions

@@
 % - a brief account of the background and rationale of the work;
-NumPy is the fundamental array programming library for the Python ecosystem
+NumPy is the fundamental array programming library for the Python programming language
 \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
@@
 % - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
-Here we show how a small group of students and scientists created this
-popular tool, explain the core ideas that helped make it it successful,
-and highlight recent development to better serve new use cases
-and future improvements.
+Here we show how a few core ideas lead to a simple and powerful array syntax.
 % - 2-3 sentences putting the main findings into general context so it is
 %   clear how the results described in the paper have moved the field forwards.
-NumPy continues to evolve to meet the needs of its community, to remain
-relevant in an ever-changing computing landscape, and to be a fundamental
-pillar of support for the next decade of scientific computing, data science,
-and machine learning in Python.
+NumPy arrays are so popular that several project have developed their own
+NumPy-like interface and array objects targeting audiences with specialized
+needs.
+Because of this, NumPy plays a unique role in enabling disparate technologies
+to interact and exchange.
+%NumPy continues to evolve to meet the needs of its community, to remain
+%relevant in an ever-changing computing landscape, and to be a fundamental
+%pillar of support for the next decade of scientific computing, data science,
+%and machine learning in Python.

### 2020-01-11 body.tex
Minor edits

@@
 includes an assortment of probability distributions.
-It also performs accelerated linear algebra, utilising one of several backends
+It also performs accelerated linear algebra, utilizing one of several backends
 such as OpenBLAS and Intel MKL optimized for the CPUs at hand.
@@
 
-- massive explosion in data science, machine learning, and artificial intelligence (tensors)
+- massive explosion in data science, machine learning, and artificial intelligence
 

### 2020-01-11 summary.tex
Minor edits

@@
 % - a brief account of the background and rationale of the work;
-NumPy is the fundamental array programming library for the Python programming language
+NumPy is the fundamental array programming library for the Python language
 \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
@@
 % - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
-Here we show how a few core ideas lead to a simple and powerful array syntax.
+Here we show how a few fundamental concepts lead to a simple and powerful array syntax.
 % - 2-3 sentences putting the main findings into general context so it is
 %   clear how the results described in the paper have moved the field forwards.
-NumPy arrays are so popular that several project have developed their own
+NumPy is so popular that several project have developed their own
 NumPy-like interface and array objects targeting audiences with specialized
@@
 Because of this, NumPy plays a unique role in enabling disparate technologies
-to interact and exchange.
+to interact and exchange data.
 %NumPy continues to evolve to meet the needs of its community, to remain

### 2020-01-11 body.tex
Minor edits to interop section

@@
 Still, the widespread adoption of the NumPy API is valuable, as it lowers the
-barrier to entry and provides stability to the community.
+barrier to entry and provides the wider community with a stable array programming interface.
 What can NumPy do to support the use of its API without blocking innovation?
-We believe that improving interoperability by having NumPy be a central
-coordinator among array objects is a good way to balance stability while
+We believe that having NumPy as a central point of coordination between array objects can improve interoperability while
 supporting growth of a larger surrounding ecosystem of specialized solutions
@@
 passed to NumPy functions.
-NumPy, in turn, dispatches operations to the originating library, as required.
+NumPy, in turn, dispatches operations to the originating library, as required.  Early versions of these protocols have been successfully deployed.
 

### 2020-01-11 body.tex
Add caption for Fig 3: Proliferation of Arrays

@@
   \includegraphics[width=0.8\textwidth]{static/sketches/duck-arrays}
-  \caption{\textbf{Proliferation of Arrays.} Many ...}
+  \caption{\textbf{Proliferation of Arrays.} The NumPy array is a classic and simple in-memory representation of array data.
+Since its inception, numerous other array representations have been developed, including distributed arrays---which are stored on multiple computers, GPU arrays---which utilize graphics processing units for fast computation, computation graphs---which postpone and combine calculations before executing them efficiently, and sparse arrays---which typically contain few non-zero values, and store only those in memory.}
   \label{fig:array-proliferation}

### 2020-01-11 body.tex
Add caption for Fig. 2: Reductions

@@
   \includegraphics[width=\linewidth]{static/sketches/reductions}
-  \caption{\textbf{Reductions.} \fixme{Reducing an array along one of several axes. Note:
-  may want to fix to have right-hand axis system.}
+  \caption{\textbf{Reductions.} Reduction operations act along one or more axes.
+In this example, a three-dimensional array is shown to be summed along various single axes to produce two-dimensional results, or along two axes consecutively to produce a one-dimensional result.
+\fixme{May want to fix to have right-hand axis system.}
    }

### 2020-01-11 body.tex
Flesh out caption for Fig. 1: Fundamental Array Concepts

@@
   \caption{\textbf{Fundamental Array Concepts.}
-    \textbf{a,} The NumPy array data structure ...
-    \textbf{b,} Indexing ...
-    \textbf{c,} Broadcasting ...
-    \textbf{d,} Indexing with arrays ... combined with broadcasting ...
+    \textbf{a,} The NumPy array data structure and its associated meta-data fields.
+    \textbf{b,} Indexing an array with various types of arguments.
+    \textbf{c,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
+    \textbf{d,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
    }

### 2020-01-11 body.tex
Add some notes

@@
 
-(Fig.~\ref{fig:array-proliferation})
+- massive explosion in data science, machine learning, and artificial intelligence
+
+\fixme{NumPy is now omnipresent.  Many packages depend on it.}
+
+- new hardware platforms; large scales of data; computational barriers; ...
+
+\fixme{arrays are a nice idea in principle, but not all problems are that easily
+solved; it breaks down when you have very large data sets, when your data
+doesn't live on a single machine, when you have to use specialized hardware, or
+when several resources need to be combined to solve a single array computation}
+
+This has lead to a proliferation of arrays (Fig.~\ref{fig:array-proliferation}).
 
@@
 - other languages w/ arrays
-
-- massive explosion in data science, machine learning, and artificial intelligence
 

### 2020-01-11 body.tex
Light editing on discussion

@@
 Python is already a good choice for many standard programming tasks such as
-cleaning data, interacting with web resources and parsing text.
+cleaning data, interacting with web resources, and parsing text.
 It has a wide range of libraries for many different tasks.
@@
 
-Lastly, NumPy has benefited greatly from a well-chosen API.
-Scientific and numerical programming needs to be fast, and scale to very large
-datasets.
-The NumPy API defines a simple wrapper for data in memory so it can be
-represented as a one- or multi-dimensional array.
-The simplicity of this wrapper has made it very successful as a standard way of
-representing arrays in memory, and has made it relatively easy for other
+Lastly, NumPy benefited greatly from the simplicity of its underlying data structure, which made it a natural standard for exchanging array data between libraries.
+It also made it relatively easy for other
 libraries to develop fast and memory-efficient compiled code, usually in C or
-Fortran, that can manipulate these arrays and pass them back to Python.
-This has been a large factor in the growth of numerical computing libraries
-around NumPy and Python.
+Fortran, that could manipulate these arrays and pass them back to Python.
+
+Over time, and with the arrival of different technologies and array types, the role of NumPy has changed from merely being an array provider, to being a widely adopted array programming language.
 

### 2020-01-11 body.tex
Minor edits

@@
 shape $N$, while color videos are four dimensional arrays of shape
-$T \times M \times N \times 3$.
+$(T, M, N, 3)$.
 
@@
 Consider, for example, a 2-D array of floating point numbers with shape
-$4 \times 3$, where each element occupies 8 bytes in memory.
+$(4, 3)$, where each element occupies 8 bytes in memory.
 To move between consecutive columns we need to jump forward 8 bytes in memory,
@@
 exploited, but are out of scope for NumPy development.
+
 Still, the widespread adoption of the NumPy API is valuable, as it lowers the
 barrier to entry and provides the wider community with a stable array programming interface.
+
 What can NumPy do to support the use of its API without blocking innovation?
@@
 
-There has been a steady increase in the number of external libraries that mimic
-NumPy, i.e. provide arrays and a NumPy-like API for manipulating them.
+There has been a steady increase in the number of external libraries that
+provide arrays and a NumPy-like API for manipulating them.
 This includes popular tensor computation libraries such as
@@
 passed to NumPy functions.
-NumPy, in turn, dispatches operations to the originating library, as required.  Early versions of these protocols have been successfully deployed.
-
+NumPy, in turn, dispatches operations to the originating library, as required.
+% FIXME: expand to 3 or more sentences discussing current efforts and
+% issues with those, how did adoption go, what is being done next
 %\url{https://numpy.org/neps/nep-0037-array-module.html}
+Early versions of these protocols have been successfully deployed.
 

### 2020-01-11 paper.tex
More work on outline
Update outline
Add methods section
Use array programming
Use Fig. instead of Figure in figure captions
Minor edits

@@
 
+\usepackage{caption}
+\captionsetup{%
+   figurename=Fig.,
+   tablename=Tab.
+}
 \usepackage{subcaption}
@@
 
-\title{NumPy---Array Computation for Python}
+\title{NumPy---Array Programming in Python}
 
@@
 
-\keywords{Scientific computing, Python, Mathematics}
+\keywords{Scientific computing, Python, Mathematics, Array Programming}
 
@@
 
-% Count of words in summary
-\verbatiminput{summarycount.tex}
-
-% Count of words in text
+% Count of words
 \verbatiminput{wordcount.tex}
@@
 
-\section*{Data Availability}
+\bibliography{references}
 
-All NumPy source code and most data generated for the current study are
-available in the NumPy GitHub repository, \url{https://github.com/numpy}. Some
-supporting code and data have also been stored in other public repositories
-cited by this manuscript.
+\newpage
 
-\bibliography{references}
+\input{methods}
 
@@
 S.J.v.d.W and K.J.M. composed the manuscript with input from others.
-S.B., K.S., T.J.R., and W.W. drafted sections.
+S.B., K.S., T.J.R., W.W., and M.B. drafted sections.
 All authors have contributed significant code, documentation, and/or expertise

### 2020-01-13 methods.tex
Add new section title

@@
 
+A vast number of projects are built on NumPy;
+these projects are consumers of the NumPy API.
+Over the last several years, a growing number of projects are providers of
+a \emph{NumPy-like API} and array objects targeting audiences with specialized
+needs beyond NumPy's capabilities.
+For example, the NumPy API is implemented by several popular tensor computation
+libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
+JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
+and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
+and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
+provide tensor APIs with NumPy-inspired semantics.
+It is also implemented in packages that support sparse arrays
+such as \code{scipy.sparse} and \code{pydata.sparse}.
+Another notable example is Dask, a library for parallel computing in
+Python.  Dask adopts the NumPy API and therefore presents a familiar
+interface to existing NumPy users, while adding powerful abilities to
+parallelize and distribute tasks.
+
+The multitude of specialized projects creates the difficulty that consumers
+of these NumPy-like APIs write code specific to a single project and do not support
+all of the above array providers.
+This is a burden for users relying on the specialized array-like, since
+a tool they need may not work for them.
+It also creates new challenges for end-users who need to transition
+from NumPy to a more specialized array.
+The growing multitude of specialized projects with NumPy-like APIs threatened
+to again fracture the scientific Python community.
+
+To address these issues NumPy has the goal of providing the fundamental
+API for \emph{interoperability} between the various NumPy-like APIs.
 An earlier step in this direction was the implementation of the

### 2020-01-13 body.tex
Add new section title
Start expanding Supporting the Ecosystem

@@
 and to access the next row $3 \times 8 = 24$ bytes.
-The strides of that array are therefore $(24, 8)$.
+The strides of that array are therefore $(24, 8)$.  NumPy is able to
+store arrays in both so-called ``C'' and ``Fortran''-order, iterating
+first over either rows or columns.  This allows external libraries
+written in those languages to directly access NumPy array data in memory.
 
@@
 
-%Once data is viewed as an $n$-dimensional array, more abstract operations such
-%as reductions can be defined.
-
-%(e.g., summation across columns, see Fig.~\ref{fig:reductions}) products, and %so forth, are
-%defined.
-
 To complement the array syntax, NumPy includes several functions that perform
@@
 
+\section*{Supporting the ecosystem}
+
 Users predominantly interact with NumPy arrays using {\em indexing} (to access
@@
 array programming.
-Exposing these array programming primitives in Python---an interpreted
-language---creates an ideal environment for exploratory data analysis where
-users may inspect, manipulate, and visualize their data
-\cite{perez2007ipython}.
+
+Various libraries build on and extend this functionality to further enrich the ecosystem of available tools.
+For example, SciPy adds various mathematical, scientific, and engineering routines; Pandas provides indexed arrays, known as dataframes; xarray implements arrays with labeled axes; and Matplotlib generates publication-ready figures and visualizations.
+Employing these low-level utilities are several widely used technique-specific libraries, including scikit-learn (machine learning), scikit-image (image processing), statmodels (statistics),  networkx (graphs and complex networks), and Napari (volumetric image visualization) \fixme{citations for all these}.
+
+Finally, domain-specific libraries and applications are built on this diverse foundation of tooling; these include astropy (astronomy) \fixme{citation}, NiPy (neuroimaging) \fixme{citation}, and pangeo (earth science) \fixme{citation}.
+
+% ecosystem
+% core: scipy, pandas, matplolib ;
+% scikits: scikit-learn, scikit-image, statmodels, networkx
+% domain specific: astropy (2 citations?), neuroscience (3 citations?), pangeo, napari, etc.
+% Ipython / Jupyter
+
+\fixme{Maybe some wording around the actual *support* element?}
+
+Exposing array programming primitives, as well as the surrounding ecosystem of tools, in Python---an interpreted language---creates an ideal environment for interactive, exploratory data analysis where users may iteratively inspect, manipulate, and visualize their data \cite{perez2007ipython}.
 
@@
 
+% Fortran / C
+% IDL / FITS / ROOT / Parquet / Arrow ?
+
 NumPy spurred the development of the larger scientific Python ecosystem and
@@
 source code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
+
+% teaching (neuroimaging paper / etc.?)
 

### 2020-01-13 body.tex
Edit Proliferation and Discussion

@@
 
-- massive explosion in data science, machine learning, and artificial intelligence
+Recent years have seen a massive explosion in data science, machine learning, and artificial intelligence.
+NumPy and its API, underlying almost every tool in the computational toolchain, has become ubiquitous.
+Still, NumPy itself could never hope to satisfy every single specialized need of the community; it is limited in its ability to work with very large datasets, datasets split across multiple systems, and computations that require specialized hardware.
 
-\fixme{NumPy is now omnipresent.  Many packages depend on it.}
-
-- new hardware platforms; large scales of data; computational barriers; ...
-
-\fixme{arrays are a nice idea in principle, but not all problems are that easily
-solved; it breaks down when you have very large data sets, when your data
-doesn't live on a single machine, when you have to use specialized hardware, or
-when several resources need to be combined to solve a single array computation}
-
-This has lead to a proliferation of arrays (Fig.~\ref{fig:array-proliferation}).
-
-- distributed arrays
-
-- GPUs
-
-- computation graph
-
-- sparse arrays
-
-- other languages w/ arrays
+Furthermore, pursuing solutions to these challenges are deemed out of scope for NumPy development.
+The community's efforts to fill the resultant gaps therefore led to a proliferation of arrays (Fig.~\ref{fig:array-proliferation}): distributed arrays, GPU arrays, delayed execution arrays, sparse arrays, foreign language arrays, and more.
 
@@
 
-NumPy's API has become ubiquitous in scientific computation, but NumPy itself
-can never hope to satisfy every single specialized need of the community.
-The advent of new technologies, such as GPUs, create opportunities that should be
-exploited, but are out of scope for NumPy development.
-
-Still, the widespread adoption of the NumPy API is valuable, as it lowers the
-barrier to entry and provides the wider community with a stable array programming interface.
+Given a rapid proliferation of array implementations, a widespread adoption of the NumPy API remains valuable: it lowers the barrier to entry for newcomers, and provides the wider community with a stable array programming interface.
 
@@
 
-NumPy was initially developed by students, faculty, and researchers in their
-spare time to provide a modern array programming library for Python.
-They were influenced by their experiences with powerful interactive programming
-languages for scientific computing like Yorick \cite{munro1995using} as well
-as commercial languages like IDL and Matlab.
-Often they developed code to solve their own or their colleagues' problems.
+NumPy was initially developed by students, faculty, and researchers in their spare time to provide a modern array programming library for Python.
+They were influenced by their experiences with powerful interactive programming languages for scientific computing like Yorick \cite{munro1995using} as well as commercial languages and environments like IDL and Matlab.
+They also frequently developed code from scratch to solve their own or their colleagues' problems, often in low-level languages that preceeded Python like as Fortran \fixme{Cite https://www.netlib.org/} and C.
+It is also hard to imagine that scientific Python was not in part influenced by its contemporaries, such as ROOT.
 
@@
 
-% teaching (neuroimaging paper / etc.?)
+% https://ras.ac.uk/sites/default/files/2020-01/Group%20Award%20-%20Astropy.pdf
+% https://ras.ac.uk/news-and-press/news/leading-astronomers-and-geophysicists-honoured-ras-bicentenary-year-0
+
+This culture of using best practices for producing reliable scientific software has been eagerly adopted by the ecosystem of libraries that build on NumPy.
+For example, in a recent award given by the Royal Astronomical Society to AstroPy, they state:
+\begin{quotation}
+\noindent\emph{The Astropy Project has provided hundreds of junior scientists with experience in professional-standard software development practices including use of version control, unit testing, code review and issue tracking procedures.
+This is a vital skillset for modern researchers that is often missing from formal university education in physics or astronomy.}
+\end{quotation}
+Community members explicitly work to address this lack of formal education through formal courses and workshops\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}\fixme{cite https://scipy-school.org/}.
 

### 2020-01-14 body.tex
Note to add references to Parquet, Arrow, Zarr, and Ruby's Numo
Combine figures

@@
   \centering
-  \includegraphics[width=\textwidth]{static/sketches/array-concepts}
+  \includegraphics[width=.9\textwidth]{static/sketches/array-concepts}   
   \caption{\textbf{Fundamental Array Concepts.}
@@
     \textbf{b,} Indexing an array with various types of arguments.
-    \textbf{c,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
-    \textbf{d,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
+    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
+    \textbf{d,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
+    \textbf{e,} Reduction operations act along one or more axes. In this
+    example, a three-dimensional array is shown to be summed along various single
+    axes to produce two-dimensional results, or along two axes consecutively to
+    produce a one-dimensional result.
    }
@@
 a specific condition (Fig.~\ref{fig:array-concepts}b).
-Arrays can even be indexed using other arrays (Fig.~\ref{fig:array-concepts}d).
+Arrays can even be indexed using other arrays (Fig.~\ref{fig:array-concepts}c).
 Wherever possible, indexing that retrieves a subarray returns a {\em view} on
@@
 copying any data in memory), so that the shapes of the operands match
-(Fig.~\ref{fig:array-concepts}c).
+(Fig.~\ref{fig:array-concepts}d).
 Broadcasting is also applied whenever an array is indexed using arrays of
-indices (Fig.~\ref{fig:array-concepts}d).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{static/sketches/reductions}
-  \caption{\textbf{Reductions.} Reduction operations act along one or more axes.
-In this example, a three-dimensional array is shown to be summed along various single axes to produce two-dimensional results, or along two axes consecutively to produce a one-dimensional result.
-\fixme{May want to fix to have right-hand axis system.}
-   }
-  \label{fig:reductions}
-\end{figure}
+indices (Fig.~\ref{fig:array-concepts}c).
 
@@
 For example, summing a $D$-dimensional array over $p$ axes results in a
-$D-p$-dimensional array (Fig.~\ref{fig:reductions}).
+$D-p$-dimensional array (Fig.~\ref{fig:array-concepts}e).
 
@@
 \section*{The proliferation of arrays}
+
+\fixme{Add references somewhere to Zarr, https://arrow.apache.org/, https://parquet.apache.org/, https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
 

### 2020-01-16 body.tex
Minor edits

@@
 It also performs accelerated linear algebra, utilizing one of several backends
-such as OpenBLAS and Intel MKL optimized for the CPUs at hand.
+such as OpenBLAS \cite{wang2013augem,xianyi2012model} and Intel MKL optimized for the CPUs at hand.
 
 \section*{Supporting the ecosystem}
+
+\fixme{Maybe some wording around the actual *support* element?}
 
@@
 
-Various libraries build on and extend this functionality to further enrich the ecosystem of available tools.
-For example, SciPy adds various mathematical, scientific, and engineering routines; Pandas provides indexed arrays, known as dataframes; xarray implements arrays with labeled axes; and Matplotlib generates publication-ready figures and visualizations.
-Employing these low-level utilities are several widely used technique-specific libraries, including scikit-learn (machine learning), scikit-image (image processing), statmodels (statistics),  networkx (graphs and complex networks), and Napari (volumetric image visualization) \fixme{citations for all these}.
+Various libraries build on and extend this functionality to further enrich the
+ecosystem of available tools.  For example, SciPy adds various mathematical,
+scientific, and engineering routines; Pandas provides indexed arrays, known as
+dataframes; xarray implements arrays with labeled axes; and Matplotlib
+generates publication-ready figures and visualizations.  Employing these
+low-level utilities are several widely used technique-specific libraries,
+including scikit-learn (machine learning), scikit-image (image processing),
+statmodels (statistics),  networkx (graphs and complex networks) \cite{SciPyProceedings_11}, and Napari
+(volumetric image visualization).
 
-Finally, domain-specific libraries and applications are built on this diverse foundation of tooling; these include astropy (astronomy) \fixme{citation}, NiPy (neuroimaging) \fixme{citation}, and pangeo (earth science) \fixme{citation}.
+Finally, domain-specific libraries and applications are built on this diverse
+foundation of tooling; these include Astropy (astronomy) \cite{astropy:2013,
+astropy:2018}, Biopython (computational biology and bioinformatics) \cite{cock2009biopython},
+NIPY (neuroimaging) \cite{millman2007analysis}, and Pangeo (earth science) \cite{2018EGUGA..2012146H}.
 
-% ecosystem
-% core: scipy, pandas, matplolib ;
-% scikits: scikit-learn, scikit-image, statmodels, networkx
-% domain specific: astropy (2 citations?), neuroscience (3 citations?), pangeo, napari, etc.
-% Ipython / Jupyter
-
-\fixme{Maybe some wording around the actual *support* element?}
-
-Exposing array programming primitives, as well as the surrounding ecosystem of tools, in Python---an interpreted language---creates an ideal environment for interactive, exploratory data analysis where users may iteratively inspect, manipulate, and visualize their data \cite{perez2007ipython}.
+Exposing array programming primitives, as well as the surrounding ecosystem of
+tools, in Python---an interpreted language---creates an ideal environment for
+interactive, exploratory data analysis where users may iteratively inspect,
+manipulate, and visualize their data \cite{perez2007ipython}.
 
@@
 
-\fixme{Add references somewhere to Zarr, https://arrow.apache.org/, https://parquet.apache.org/, https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
+\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
+https://parquet.apache.org/,
+https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
 
@@
 They were influenced by their experiences with powerful interactive programming languages for scientific computing like Yorick \cite{munro1995using} as well as commercial languages and environments like IDL and Matlab.
-They also frequently developed code from scratch to solve their own or their colleagues' problems, often in low-level languages that preceeded Python like as Fortran \fixme{Cite https://www.netlib.org/} and C.
+They also frequently developed code from scratch to solve their own or their colleagues' problems, often in low-level languages that preceded Python like as Fortran \cite{dongarra2008netlib} and C.
 It is also hard to imagine that scientific Python was not in part influenced by its contemporaries, such as ROOT.
@@
 % Fortran / C
-% IDL / FITS / ROOT / Parquet / Arrow ?
 
@@
 \end{quotation}
-Community members explicitly work to address this lack of formal education through formal courses and workshops\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}\fixme{cite https://scipy-school.org/}.
+Community members explicitly work to address this lack of formal education
+through formal courses and workshops
+\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
+%\fixme{cite https://scipy-school.org/}.
 

### 2020-01-16 summary.tex
Fix typo
Make end of abstract a bit more punchy

@@
 downloads per year.
+%It is one of the most commonly used machine learning tools among
+%enterprises \cite{451report2018}.
 It plays an essential role in research analysis pipelines in fields as
@@
 universe\cite{jenness2018lsst}.
-It is one of the most commonly used machine learning tools among
-enterprises \cite{451report2018}.
 % - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
-Here we show how a few fundamental concepts lead to a simple and powerful array syntax.
+Here we show how a few fundamental concepts lead to a simple and
+powerful programming paradigm for organizing, exploring, and analyzing
+scientific data.
 % - 2-3 sentences putting the main findings into general context so it is
 %   clear how the results described in the paper have moved the field forwards.
-NumPy is so popular that several project have developed their own
-NumPy-like interface and array objects targeting audiences with specialized
-needs.
-Because of this, NumPy plays a unique role in enabling disparate technologies
-to interact and exchange data.
+NumPy is the foundation upon which the entire scientific Python
+universe is constructed. It is so pervasive that several projects,
+targeting audiences with specialized needs, have developed their own
+NumPy-like interfaces and array objects.  Because of its central position in the
+ecosystem, NumPy increasingly is used to bridge disparate
+technologies, opening new possibilities in previously unexplored computational frontiers.
+
+%Because of this, NumPy plays a unique role in enabling disparate technologies
+%to interact and exchange data.
 %NumPy continues to evolve to meet the needs of its community, to remain

### 2020-01-16 body.tex
Minor fixes

@@
 numerical computation including SciPy \cite{virtanen2019scipy},
-matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
+Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
 scikit-learn \cite{pedregosa2011scikit}, and
@@
 
+% - review supporting the ecosystem (actual support element)
+%   - docstrings / testing / packaging
+%   - coordinating mechanism
+%     - deprecation policy
+
 Users predominantly interact with NumPy arrays using {\em indexing} (to access
@@
 including scikit-learn (machine learning), scikit-image (image processing),
-statmodels (statistics),  networkx (graphs and complex networks) \cite{SciPyProceedings_11}, and Napari
+statsmodels (statistics),  networkx (graphs and complex networks) \cite{SciPyProceedings_11}, and Napari
 (volumetric image visualization).
@@
 This culture of using best practices for producing reliable scientific software has been eagerly adopted by the ecosystem of libraries that build on NumPy.
-For example, in a recent award given by the Royal Astronomical Society to AstroPy, they state:
+For example, in a recent award given by the Royal Astronomical Society to Astropy, they state:
 \begin{quotation}

### 2020-01-16 body.tex
Replace utility w/ array-aware

@@
 scikit-image \cite{vanderwalt2014scikit}.
-It consists of an $n$-dimensional array object along with utility functions
+It consists of an $n$-dimensional array object along with array-aware functions
 that operate on it.
@@
 
-Other utility functions include creating, reshaping, concatenating, and padding
+Other array-aware functions include creating, reshaping, concatenating, and padding
 arrays; searching, sorting and counting data; and reading and writing files.
@@
 sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
-and, for matrix multiplication, $@$), as well as {\em utility functions};
+and, for matrix multiplication, $@$), as well as {\em array-aware functions};
 together, these provide an easily readable, expressive, high-level API for
@@
 generates publication-ready figures and visualizations.  Employing these
-low-level utilities are several widely used technique-specific libraries,
+low-level array-aware functions are several widely used technique-specific libraries,
 including scikit-learn (machine learning), scikit-image (image processing),

### 2020-01-16 body.tex
Minor edits / comments from Ralf

@@
 
-To facilitate \emph{interoperability}, the NumPy team is defining so-called
+To facilitate \emph{interoperability}, the NumPy team has designed
 ``protocols'' (or contracts of operation), that allow for these arrays to be
@@
 %\url{https://numpy.org/neps/nep-0037-array-module.html}
-Early versions of these protocols have been successfully deployed.
+Versions of these protocols have been successfully deployed.
+% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
+% possible to create distributed GPU arrays, enabling .... [find a reference to
+% work by Peter Entschev]".
 

### 2020-01-16 methods.tex
Fix typo

@@
 This allows us to review code before integrating it as well as to run a
-large number of tests on the modificatied code to ensure that it does not
+large number of tests on the modified code to ensure that it does not
 break things.

### 2020-01-16 body.tex
Avoid repetition
Minor edit

@@
 that operate on it.
-Because of its inherent simplicity---being a pointer to memory with some
-associated meta-data---the NumPy array is
+Because of its inherent simplicity, the NumPy array is
 the {\it de facto} exchange format for array data in Python.
@@
 The strides of that array are therefore $(24, 8)$.  NumPy is able to
-store arrays in both so-called ``C'' and ``Fortran''-order, iterating
+store arrays in either C or Fortran memory order, iterating
 first over either rows or columns.  This allows external libraries

### 2020-01-16 body.tex
More minor edits

@@
 generates publication-ready figures and visualizations.  Employing these
-low-level array-aware functions are several widely used technique-specific libraries,
+low-level array-aware functions are widely used technique-specific libraries,
 including scikit-learn (machine learning), scikit-image (image processing),
-statsmodels (statistics),  networkx (graphs and complex networks) \cite{SciPyProceedings_11}, and Napari
-(volumetric image visualization).
+statsmodels (statistics),  networkx (graphs and complex networks)
+\cite{SciPyProceedings_11}, and Napari (volumetric image visualization).
 
@@
 
-Furthermore, pursuing solutions to these challenges are deemed out of scope for NumPy development.
-The community's efforts to fill the resultant gaps therefore led to a proliferation of arrays (Fig.~\ref{fig:array-proliferation}): distributed arrays, GPU arrays, delayed execution arrays, sparse arrays, foreign language arrays, and more.
+Furthermore, pursuing solutions to these challenges is out of scope for
+NumPy development. The community's efforts to fill the resultant gaps therefore
+led to a proliferation of arrays (Fig.~\ref{fig:array-proliferation}):
+distributed arrays, GPU arrays, delayed execution arrays, sparse arrays,
+foreign language arrays, and more.
 
@@
 
-Python is already a good choice for many standard programming tasks such as
+Python is already a good choice for standard programming tasks such as
 cleaning data, interacting with web resources, and parsing text.
-It has a wide range of libraries for many different tasks.
+It has a wide range of libraries for different tasks.
 Adding fast array operations and linear algebra allows the scientist to do all
@@
 
-Lastly, NumPy benefited greatly from the simplicity of its underlying data structure, which made it a natural standard for exchanging array data between libraries.
-It also made it relatively easy for other
+Lastly, NumPy benefited greatly from the simplicity of its underlying data structure, which made it a natural standard for exchanging array data between libraries.  It also made it easy for other
 libraries to develop fast and memory-efficient compiled code, usually in C or

### 2020-01-16 methods.tex
More minor edits

@@
 
-The random number generator library in NumPy provides several alternative
+The random number generator library in NumPy provides alternative
 \emph{bit stream generators} that provide the core function of generating
@@
 Over the course of its history, NumPy has attracted PRs by 823 contributors.
-However, its development relies heavily on a relatively small number
+However, its development relies heavily on a small number
 of active maintainers, who share more than half of the contributions among

### 2020-01-16 body.tex
Fix typo

@@
 Many of these functions, such as summation, support reductions: aggregating
-results across over or more dimension of the array.
-For example, summing a $D$-dimensional array over $p$ axes results in a
-$D-p$-dimensional array (Fig.~\ref{fig:array-concepts}e).
+results across one or more dimensions of the array.
+For example, summing a $d$-dimensional array over $c$ axes results in a
+$(d-c)$-dimensional array (Fig.~\ref{fig:array-concepts}e).
 

### 2020-01-20 body.tex
Minor edits

@@
 scikit-image \cite{vanderwalt2014scikit}.
-It consists of an $n$-dimensional array object along with array-aware functions
+It consists of an multidimensional array object along with array-aware functions
 that operate on it.
@@
   \centering
-  \includegraphics[width=.9\textwidth]{static/sketches/array-concepts}   
+  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
   \caption{\textbf{Fundamental Array Concepts.}
-    \textbf{a,} The NumPy array data structure and its associated meta-data fields.
+    \textbf{a,} The NumPy array data structure and its associated metadata fields.
     \textbf{b,} Indexing an array with various types of arguments.
@@
 The NumPy array is a data structure that efficiently stores and accesses
-$n$-dimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, that
+multidimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, that
 enables a wide variety of scientific computation.
-It consists of a pointer to memory, along with meta-data used to interpret the
+It consists of a pointer to memory, along with metadata used to interpret the
 data stored there, notably {\em data type}, {\em shape}, and {\em strides}
@@
 The {\em data type} describes the nature of elements stored in an array.
-An array has a single data type, and each array element occupies the same
+n array has a single data type, and each array element occupies the same
 number of bytes in memory.
@@
 {\em Strides} are necessary to interpret computer memory, which stores elements
-linearly, as $n$-dimensional arrays.
+linearly, as multidimensional arrays.
 It describes the number of bytes to move forward in memory to jump from row to
@@
 a simple yet powerful array expression syntax.
-It supports operations such as indexing (selecting subarrays) and element-wise
+It supports operations such as indexing (selecting subarrays) and elementwise
 calculations.
@@
 
-To complement the array syntax, NumPy includes several functions that perform
-element-wise calculations on arrays, including arithmetic, statistics, and
+To complement the array syntax, NumPy includes functions that perform
+elementwise calculations on arrays, including arithmetic, statistics, and
 trigonometry.
@@
 results across one or more dimensions of the array.
-For example, summing a $d$-dimensional array over $c$ axes results in a
-$(d-c)$-dimensional array (Fig.~\ref{fig:array-concepts}e).
+For example, summing a $n$-dimensional array over $d$ axes results in a
+$(n-d)$-dimensional array (Fig.~\ref{fig:array-concepts}e).
 
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{static/sketches/duck-arrays}
+  \includegraphics[width=0.6\textwidth]{static/sketches/duck-arrays}
   \caption{\textbf{Proliferation of Arrays.} The NumPy array is a classic and simple in-memory representation of array data.
@@
 
-Given a rapid proliferation of array implementations, a widespread adoption of the NumPy API remains valuable: it lowers the barrier to entry for newcomers, and provides the wider community with a stable array programming interface.
+Given the proliferation of array implementations, a widespread adoption of the NumPy API remains valuable: it lowers the barrier to entry for newcomers, and provides the wider community with a stable array programming interface.
 

### 2020-01-20 methods.tex
Minor edits

@@
 the ufunc will take care of looping over the dimensions of the inputs
-in such a way that if, say, \code{x} is a 1-d array with length 3, and
-\code{y} is a 2-d array with shape $2 \times 1$, the output will be
+in such a way that if, say, \code{x} is a 1-D array with length 3, and
+\code{y} is a 2-D array with shape $2 \times 1$, the output will be
 an array with shape $2 \times 3$.  The ufunc machinery takes care

### 2020-01-20 body.tex
Rethinking Fig. 2 (duck array / interoperability)

@@
 
-\begin{figure*}
-  \centering
-  \includegraphics[width=0.6\textwidth]{static/sketches/duck-arrays}
-  \caption{\textbf{Proliferation of Arrays.} The NumPy array is a classic and simple in-memory representation of array data.
-Since its inception, numerous other array representations have been developed, including distributed arrays---which are stored on multiple computers, GPU arrays---which utilize graphics processing units for fast computation, computation graphs---which postpone and combine calculations before executing them efficiently, and sparse arrays---which typically contain few non-zero values, and store only those in memory.}
-  \label{fig:array-proliferation}
-\end{figure*}
+Recent years have seen a massive explosion in data science, machine learning,
+and artificial intelligence.  NumPy and its API, underlying almost every tool
+in the computational toolchain, has become ubiquitous.  Still, NumPy itself
+could never hope to satisfy every single specialized need of the community; it
+is limited in its ability to work with very large datasets, datasets split
+across multiple systems, and computations that require specialized hardware.
+Furthermore, pursuing solutions to these challenges is out of scope for NumPy
+development.
 
-Recent years have seen a massive explosion in data science, machine learning, and artificial intelligence.
-NumPy and its API, underlying almost every tool in the computational toolchain, has become ubiquitous.
-Still, NumPy itself could never hope to satisfy every single specialized need of the community; it is limited in its ability to work with very large datasets, datasets split across multiple systems, and computations that require specialized hardware.
-
-Furthermore, pursuing solutions to these challenges is out of scope for
-NumPy development. The community's efforts to fill the resultant gaps therefore
-led to a proliferation of arrays (Fig.~\ref{fig:array-proliferation}):
-distributed arrays, GPU arrays, delayed execution arrays, sparse arrays,
-foreign language arrays, and more.
+The community's efforts to fill the resultant gaps therefore led to a
+proliferation of arrays including distributed arrays---which are stored on
+multiple computers, GPU arrays---which utilize graphics processing units for
+fast computation, computation graphs---which postpone and combine calculations
+before executing them efficiently, and sparse arrays---which typically contain
+few non-zero values, and store only those in memory.
+%distributed arrays, GPU arrays, delayed execution arrays, sparse arrays,
+%foreign language arrays, and more.
 
@@
 
-Given the proliferation of array implementations, a widespread adoption of the NumPy API remains valuable: it lowers the barrier to entry for newcomers, and provides the wider community with a stable array programming interface.
+Given the proliferation of array implementations, a widespread adoption of the
+NumPy API remains valuable: it lowers the barrier to entry for newcomers, and
+provides the wider community with a stable array programming interface.
 
 What can NumPy do to support the use of its API without blocking innovation?
-We believe that having NumPy as a central point of coordination between array objects can improve interoperability while
-supporting growth of a larger surrounding ecosystem of specialized solutions
-and new tools.
-Primarily, there exist two types of Python array objects: subclasses of NumPy
-arrays, and arrays that mimic NumPy arrays but are fundamentally different.
+We believe that having NumPy as a central point of coordination between array
+objects can improve interoperability while supporting growth of a larger
+surrounding ecosystem of specialized solutions and new tools.  Primarily, there
+exist two types of Python array objects: subclasses of NumPy arrays, and arrays
+that mimic NumPy arrays but are fundamentally different.
 
@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=\textwidth]{static/sketches/duck-arrays}
+  \caption{\textbf{Interoperability.} \fixme{This figure needs work.}}  \label{fig:duck-arrays}
+\end{figure*}
+
 To facilitate \emph{interoperability}, the NumPy team has designed
 ``protocols'' (or contracts of operation), that allow for these arrays to be
-passed to NumPy functions.
+passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
 NumPy, in turn, dispatches operations to the originating library, as required.
@@
 
-NumPy was initially developed by students, faculty, and researchers in their spare time to provide a modern array programming library for Python.
-They were influenced by their experiences with powerful interactive programming languages for scientific computing like Yorick \cite{munro1995using} as well as commercial languages and environments like IDL and Matlab.
-They also frequently developed code from scratch to solve their own or their colleagues' problems, often in low-level languages that preceded Python like as Fortran \cite{dongarra2008netlib} and C.
-It is also hard to imagine that scientific Python was not in part influenced by its contemporaries, such as ROOT.
+NumPy was initially developed by students, faculty, and researchers in their
+spare time to provide a modern array programming library for Python.  They were
+influenced by their experiences with powerful interactive programming languages
+for scientific computing like Yorick \cite{munro1995using} as well as
+commercial languages and environments like IDL and Matlab.  They also
+frequently developed code from scratch to solve their own or their colleagues'
+problems, often in low-level languages that preceded Python like as Fortran
+\cite{dongarra2008netlib} and C.  It is also hard to imagine that scientific
+Python was not in part influenced by its contemporaries, such as ROOT.
 
@@
 
-This culture of using best practices for producing reliable scientific software has been eagerly adopted by the ecosystem of libraries that build on NumPy.
-For example, in a recent award given by the Royal Astronomical Society to Astropy, they state:
+This culture of using best practices for producing reliable scientific software
+has been eagerly adopted by the ecosystem of libraries that build on NumPy.
+For example, in a recent award given by the Royal Astronomical Society to
+Astropy, they state:
 \begin{quotation}
-\noindent\emph{The Astropy Project has provided hundreds of junior scientists with experience in professional-standard software development practices including use of version control, unit testing, code review and issue tracking procedures.
-This is a vital skillset for modern researchers that is often missing from formal university education in physics or astronomy.}
+\noindent\emph{The Astropy Project has provided hundreds of junior scientists
+with experience in professional-standard software development practices
+including use of version control, unit testing, code review and issue tracking
+procedures. This is a vital skillset for modern researchers that is often
+missing from formal university education in physics or astronomy.}
 \end{quotation}
@@
 
-Lastly, NumPy benefited greatly from the simplicity of its underlying data structure, which made it a natural standard for exchanging array data between libraries.  It also made it easy for other
-libraries to develop fast and memory-efficient compiled code, usually in C or
-Fortran, that could manipulate these arrays and pass them back to Python.
+Lastly, NumPy benefited greatly from the simplicity of its underlying data
+structure, which made it a natural standard for exchanging array data between
+libraries.  It also made it easy for other libraries to develop fast and
+memory-efficient compiled code, usually in C or Fortran, that could manipulate
+these arrays and pass them back to Python.
 
-Over time, and with the arrival of different technologies and array types, the role of NumPy has changed from merely being an array provider, to being a widely adopted array programming language.
-
-% S: I think two issues are conflated here: the simple underlying memory model, and the API.
-The API has developed organically, while the underlying memory model is a design decision that shapes much of how the rest of the library was built.
+Over time, and with the arrival of different technologies and array types, the
+role of NumPy has changed from merely being an array provider, to being a
+widely adopted array programming language.
 
@@
 %
-% E.g., practical: reason why we use Python (learn one language for everything); students don't have money, so want to avoid impracticalities of license dongles/servers, etc.
+% E.g., practical: reason why we use Python (learn one language for
+% everything); students don't have money, so want to avoid impracticalities of
+% license dongles/servers, etc.
 %
-% Philosophical: science should be open, transparent; our software should be controlled by scientists, not designers that we don't have access to
+% Philosophical: science should be open, transparent; our software should be
+% controlled by scientists, not designers that we don't have access to
 %
-% Social: joy of building these things together, friendly welcome into the community for many of us---appreciation of our work and hours
-%
+% Social: joy of building these things together, friendly welcome into the
+% community for many of us---appreciation of our work and hours

### 2020-01-20 body.tex
More work on Fig. 2
Mention we are community developed

@@
 scikit-image \cite{vanderwalt2014scikit}.
-It consists of an multidimensional array object along with array-aware functions
+It is a community developed, open source library, which provides a
+multidimensional Python array object along with array-aware functions
 that operate on it.
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
-  \includegraphics[width=\textwidth]{static/sketches/duck-arrays}
-  \caption{\textbf{Interoperability.} \fixme{This figure needs work.}}  \label{fig:duck-arrays}
-\end{figure*}
+  \includegraphics[width=.45\textwidth]{static/sketches/duck-arrays}
+  \caption{\textbf{Interoperability.} \fixme{This figure needs work.}}\label{fig:duck-arrays}
+\end{figure}
 

### 2020-01-22 body.tex
Add some notes

@@
 frequently developed code from scratch to solve their own or their colleagues'
-problems, often in low-level languages that preceded Python like as Fortran
-\cite{dongarra2008netlib} and C.  It is also hard to imagine that scientific
-Python was not in part influenced by its contemporaries, such as ROOT.
+problems, often in low-level languages that preceded Python like Fortran
+\cite{dongarra2008netlib} and C.
 
 % Fortran / C
+% A few sentences on the history of array programming, referencing APL and Fortran, are missing.
 
@@
 % - Social
+% - Technical
 %

### 2020-01-22 body.tex
Remove unneeded hypens

@@
 Around 2000, the Space Telescope Science Institute (STScI) software group wrote
-a re-implementation of much of Numeric, called NumArray, to support their work
+a reimplementation of much of Numeric, called NumArray, to support their work
 on large memory-mapped arrays and arrays of mixed data type
@@
 
-Indexing an array returns single elements, sub-arrays, or elements that satisfy
+Indexing an array returns single elements, subarrays, or elements that satisfy
 a specific condition (Fig.~\ref{fig:array-concepts}b).
@@
 Users predominantly interact with NumPy arrays using {\em indexing} (to access
-sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
+subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
 and, for matrix multiplication, $@$), as well as {\em array-aware functions};
@@
 
-NumPy has strong culture of using software-engineering practice to
+NumPy has strong culture of using software engineering practice to
 improve collaboration and reduce error \cite{millman2014developing}.

### 2020-01-22 body.tex
Switch interoperability/extensibility order

@@
 
-\section*{Extensibility and Interoperability}
+\section*{Interoperability and extensibility}
 
@@
 surrounding ecosystem of specialized solutions and new tools.  Primarily, there
-exist two types of Python array objects: subclasses of NumPy arrays, and arrays
-that mimic NumPy arrays but are fundamentally different.
-
-Subclasses of NumPy often exist due to difficulty in constructing richer data
-types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
-geometrical objects \cite{pygeos}, and missing numbers.
-To improve \emph{extensibility}, we are currently overhauling the data type
-system to make it easier to implement data types both in Python and C, and to
-support these and other applications.
+exist two types of Python array objects: (1) arrays that mimic NumPy arrays but are
+fundamentally different and (2) subclasses of NumPy arrays.
 
@@
 % https://numpy.org/neps/nep-0037-array-module.html
+
+Subclasses of NumPy often exist due to difficulty in constructing richer data
+types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
+geometrical objects \cite{pygeos}, and missing numbers.
+To improve \emph{extensibility}, we are currently overhauling the data type
+system to make it easier to implement data types both in Python and C, and to
+support these and other applications.
 

### 2020-01-24 body.tex
Fix typos

@@
 
-The {\em data type} describes the nature of elements stored in an array.
-n array has a single data type, and each array element occupies the same
+The \emph{data type} describes the nature of elements stored in an array.
+An array has a single data type, and each array element occupies the same
 number of bytes in memory.
@@
 
-The {\em shape} of an array determines the number of elements along each axis,
+The \emph{shape} of an array determines the number of elements along each axis,
 and the number of axes is the array's dimensionality.
@@
 
-{\em Strides} are necessary to interpret computer memory, which stores elements
+\emph{Strides} are necessary to interpret computer memory, which stores elements
 linearly, as multidimensional arrays.
@@
 
-Indexing an array returns single elements, subarrays, or elements that satisfy
+\emph{Indexing} an array returns single elements, subarrays, or elements that satisfy
 a specific condition (Fig.~\ref{fig:array-concepts}b).
@@
 arrays, is {\em broadcasting}.
-When performing an operation (such as addition) on two arrays, one intuitively
-expect that these arrays should be required to have the same shape.
+When performing an elementwise operation (such as addition) on two arrays, one
+expects that these arrays should have the same shape.
 However, through broadcasting, NumPy allows the dimensions to differ, while
@@
 
-Many of these functions, such as summation, support reductions: aggregating
+Many of these functions, such as summation, support \emph{reductions}: aggregating
 results across one or more dimensions of the array.

### 2020-01-24 body.tex
Add quote about APL users

@@
 
+This description of how APL users work easily applies to scientific Python
+users \cite{kromberg2007apl}:
+\begin{quote}
+Users of APL tend to ``live in their data''. Interactive interpreters allow them
+to inspect, almost feel their  way forward, discovering successful snippets
+of code through experimentation and collecting them
+into imperative or functional programs according to taste.
+\end{quote}
+
 NumPy spurred the development of the larger scientific Python ecosystem and

### 2020-01-24 body.tex
Rewrite first paragraph of interoperability

@@
 
-Given the proliferation of array implementations, a widespread adoption of the
-NumPy API remains valuable: it lowers the barrier to entry for newcomers, and
-provides the wider community with a stable array programming interface.
+NumPy supports array operations between external array objects by
+acting as a central coordination mechanism with a well-specified API.
+With the proliferation of array implementations discussed above, a
+widespread adoption of the NumPy API is valuable: it lowers the
+barrier to entry for newcomers and provides the wider community with a
+stable array programming interface. This, in turn, prevents disruptive
+schisms like the divergence of Numeric and NumArray, by facilitating
+the development of specialized solutions and new tools that operate in
+concert.
 
-What can NumPy do to support the use of its API without blocking innovation?
-We believe that having NumPy as a central point of coordination between array
-objects can improve interoperability while supporting growth of a larger
-surrounding ecosystem of specialized solutions and new tools.  Primarily, there
+Primarily, there
 exist two types of Python array objects: (1) arrays that mimic NumPy arrays but are

### 2020-01-25 body.tex
Think about expanding interop section

@@
 
-To manipulate the data structure in an intuitive manner, we provide users with
-a simple yet powerful array expression syntax.
-It supports operations such as indexing (selecting subarrays) and elementwise
-calculations.
-A high level syntax allows users to express themselves succinctly, while NumPy
+Users interact with NumPy arrays using {\em indexing} (to access
+subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$
+for elementwise operations and $@$ for matrix multiplication), as well as {\em array-aware functions};
+together, these provide an easily readable, expressive, high-level API for
+array programming, while NumPy
 deals with the underlying mechanics of making operations fast.
@@
 
-A powerful array feature that affects the way NumPy indexes and combines
-arrays, is {\em broadcasting}.
+To complement the array syntax, NumPy includes functions that perform
+elementwise calculations on arrays, including arithmetic, statistics, and
+trigonometry.
+They aim to loop over array elements near-optimally, taking into consideration,
+e.g., strides, in order to best utilize the computer's fast cache memory.
+Vectorized operations that would take many tens of lines to express in
+languages such as C can often be implemented as a single, clear Python
+expression.
+
 When performing an elementwise operation (such as addition) on two arrays, one
 expects that these arrays should have the same shape.
-However, through broadcasting, NumPy allows the dimensions to differ, while
+However, through \emph{broadcasting}, NumPy allows the dimensions to differ, while
 still producing results that appeal to intuition.
@@
 
-To complement the array syntax, NumPy includes functions that perform
-elementwise calculations on arrays, including arithmetic, statistics, and
-trigonometry.
-They aim to loop over array elements near-optimally, taking into consideration,
-e.g., strides, in order to best utilize the computer's fast cache memory.
-Vectorized operations that would take many tens of lines to express in
-languages such as C can often be implemented as a single, clear Python
-expression.
-
-Many of these functions, such as summation, support \emph{reductions}: aggregating
+Many of the array-aware functions, such as summation, support \emph{reductions}: aggregating
 results across one or more dimensions of the array.
@@
 
-\section*{Supporting the ecosystem}
-
-\fixme{Maybe some wording around the actual *support* element?}
+\section*{Scientific Python ecosystem}
 
@@
 
-Users predominantly interact with NumPy arrays using {\em indexing} (to access
-subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
-and, for matrix multiplication, $@$), as well as {\em array-aware functions};
-together, these provide an easily readable, expressive, high-level API for
-array programming.
+The brevity and expressive power of NumPy's array programming syntax combined with 
+an advanced interactive environment like IPython \cite{perez2007ipython}
+or Jupyter \cite{Kluyver:2016aa} makes interactive data exploration fluid and simple.
+It is common for NumPy users to work interactively with their data,
+discovering through experimentation bits of code that they stitch together into
+imperative or functional programs.
+The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
+this solid foundation with core (e.g., SciPy and Matplotlib) libraries,
+widely used technique-specific
+libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
+and several domain-specific projects and libraries
+\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
+Exposing these simple array programming primitives, as well as the surrounding ecosystem of
+tools, in Python---an interpreted, general purpose language---creates an ideal environment for
+interactive, exploratory data analysis where users may iteratively inspect,
+manipulate, and visualize their data.
+This rich environment has made the scientific Python ecosystem popular for scientific
+research.
 
-Various libraries build on and extend this functionality to further enrich the
-ecosystem of available tools.  For example, SciPy adds various mathematical,
-scientific, and engineering routines; Pandas provides indexed arrays, known as
-dataframes; xarray implements arrays with labeled axes; and Matplotlib
-generates publication-ready figures and visualizations.  Employing these
-low-level array-aware functions are widely used technique-specific libraries,
-including scikit-learn (machine learning), scikit-image (image processing),
-statsmodels (statistics),  networkx (graphs and complex networks)
-\cite{SciPyProceedings_11}, and Napari (volumetric image visualization).
+\begin{figure}
+  \centering
+  \includegraphics[width=.48\textwidth]{static/ecosystem}
+  \caption{\textbf{Scientific Python Ecosystem.}
+   This figure highlights some of the essential tools and libraries in the vast
+   scientific Python ecosystem.  Most of the ecosystem builds on NumPy to
+   provide a rich and powerful collection of array-aware functionality.
+  }
+  \label{fig:ecosystem}
+\end{figure}
 
-Finally, domain-specific libraries and applications are built on this diverse
-foundation of tooling; these include Astropy (astronomy) \cite{astropy:2013,
-astropy:2018}, Biopython (computational biology and bioinformatics) \cite{cock2009biopython},
-NIPY (neuroimaging) \cite{millman2007analysis}, and Pangeo (earth science) \cite{2018EGUGA..2012146H}.
+For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
+collaboration relies on many components of the scientific Python ecosystem.
+NumPy arrays are used to store and manipulate numerical data at every step
+in the processing chain: from raw data through calibration and image 
+reconstruction.
+Scipy is used to provide tools used in general image processing such as 
+filtering and image alignment, while scikit-image provides higher-level
+functionality such as edge filters and Hough transforms.
+SciPy provides fundamental algorithms for scientific computing on arrays
+including many mathematical, scientific, and engineering routines.
+SciPy and NumPy are closely related and were developed together often
+by the same people.
+scikit-image is a popular technique-specific library,
+which vastly extends the basic image processing tools provided by SciPy.
+Other components of Scipy are used throughout the library, such as the
+\code{scipy.optimize} package for handling general optimization tasks.
+NetworkX \cite{SciPyProceedings_11} (the standard package for complex
+network analysis) is used for analysis of consistency among image
+comparisons.
+Astropy \cite{astropy:2013, astropy:2018} (a community-developed astronomy
+library) is used throughout the library,
+primarily for I/O from standard astronomical file formats and time/coordinate
+transformations.
+Matplotlib (the core library for generating publication-ready figures
+and visualizations) is used for visualizing data throughout the analysis pipeline,
+including the generation of the final image of the black hole.
+% N.B. - most of this comes from introspecting the Image class defined in
+% ehtim/image.py and grepping the dependencies in requirements.txt
 
-Exposing array programming primitives, as well as the surrounding ecosystem of
-tools, in Python---an interpreted language---creates an ideal environment for
-interactive, exploratory data analysis where users may iteratively inspect,
-manipulate, and visualize their data \cite{perez2007ipython}.
+Tools for analyzing data from gravitational wave observatories such as LIGO
+and Virgo are provided by the \code{pycbc} package, which was used in the first
+detection of gravitational waves \cite{abbott2016observation} and in 
+on-going analysis of data from LIGO and Virgo.
+\code{pycbc} make extensive use of the scientific Python ecosystem.
+The time-series data from the interferometers are stored in NumPy arrays,
+while the \code{scipy.signal} is used to construct and apply filters to the
+data.
+Matplotlib is used to visualize data at all points along the analysis chain
+including raw data from the instrumentation and the time-frequency 
+visualization of the ``chirps'' that result from mergers of compact binaries
+such as black holes\cite{abbott2016observation} and
+neutron stars\cite{abbott2017multi}.
 
-\section*{The proliferation of arrays}
+The rich ecosystem of tools building on NumPy and the larger community
+of scientists and programmers using those tools mean that NumPy has a
+responsibility to maintain considerable stability.
+This need creates interesting challenges for NumPy development as the
+computational environment scientists are analyzing their data on continues
+to evolve and change.
 
-\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
-https://parquet.apache.org/,
-https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
+\section*{Proliferation of arrays}
+
+%\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
+%https://parquet.apache.org/,
+%https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
+
+% pandas, xarray, ...
 
@@
 
+This includes popular tensor computation libraries such as
+CuPy, JAX, and Apache MXNet.
+PyTorch and Tensorflow provide tensor APIs with NumPy-inspired semantics.
+% \footnote{\url{https://cupy.chainer.org/}}
+% \footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}
+% \footnote{\url{https://numpy.mxnet.io/}}
+% \footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
+% \footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
+
 \section*{Interoperability and extensibility}
@@
 provide arrays and a NumPy-like API for manipulating them.
-This includes popular tensor computation libraries such as
-CuPy\footnote{\url{https://cupy.chainer.org/}},
-JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}, and
-Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
-and
-Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
-provide tensor APIs with NumPy-inspired semantics.
 Ideally, operating on these arrays using NumPy would simply work, so that end
@@
 % work by Peter Entschev]".
+% Below an attempt to start on this:
+Thanks to these developments users of Dask and NumPy arrays are now capable
+to scale their computation to the GPU using Dasks CuPy integration with
+only minimal code changes.
+However, general adoption has not yet been achieved mainly for compatibility
+concerns.
+Thus, we are actively exploring to address these by revising and extending
+these protocols with the aim of achieving a situation where this
+interoperability spans the whole scientific Python ecosystem up to and including domain libraries.
 
@@
 types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
-geometrical objects \cite{pygeos}, and missing numbers.
+geometrical objects \cite{pygeos}, and missing numbers. % pandas: categoricals, IP addresses?
 To improve \emph{extensibility}, we are currently overhauling the data type
@@
 support these and other applications.
+The widespread adoption of rich, user defined, data types can allow more
+concise, feature rich scientific computing.
+By making these user implemented data types a fundamental design of NumPy
+these tools will enabled and interact naturally with different libraries.
 
@@
 
-% aka why is this so successful
-% some sense of ongoing work / future directions
+NumPy was initially developed by students, faculty, and researchers to
+provide an advanced, open source array programming library for Python,
+which was free to use and unencumbered by license servers, dongles, and the like.
+There was a sense of building something
+impactful together, for the benefit of many others.  Participating in
+such an endeavor, within a welcoming community of like-minded
+indivuals, held a powerful attraction for many early contributors.
 
-NumPy was initially developed by students, faculty, and researchers in their
-spare time to provide a modern array programming library for Python.  They were
-influenced by their experiences with powerful interactive programming languages
-for scientific computing like Yorick \cite{munro1995using} as well as
-commercial languages and environments like IDL and Matlab.  They also
-frequently developed code from scratch to solve their own or their colleagues'
-problems, often in low-level languages that preceded Python like Fortran
-\cite{dongarra2008netlib} and C.
+These user-developers frequently had to write code from scratch to solve
+their or their colleagues' problems---often in low-level languages
+that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
+To them, the advantages of an interactive, high-level array library
+were evident. The design of this new tool was informed by their
+experiences with powerful interactive programming languages for
+scientific computing such as APL \cite{iverson1962programming} and
+Yorick \cite{munro1995using}, as well as commercial languages and
+environments like IDL and Matlab.
 
-% Fortran / C
-% A few sentences on the history of array programming, referencing APL and Fortran, are missing.
+NumPy has subsequently seen prolific adoption, with millions of users,
+first from research and later also industry.
+It spurred the development of the larger scientific Python ecosystem and
+heralded the current era of wide-spread use of Python for scientific computing.
+The reasons for its success are varied and complex, but include the fact that
+NumPy (a) is built on top of Python,
+(b) has a strong culture of collaboration and good software development practices, and
+(c) provides a cutting-edge array programming environment for scientific data analysis.
 
-This description of how APL users work easily applies to scientific Python
-users \cite{kromberg2007apl}:
-\begin{quote}
-Users of APL tend to ``live in their data''. Interactive interpreters allow them
-to inspect, almost feel their  way forward, discovering successful snippets
-of code through experimentation and collecting them
-into imperative or functional programs according to taste.
-\end{quote}
+\emph{Python} is an open source, general-purpose, interpreted programming language
+well-suited to standard programming tasks such as cleaning data,
+interacting with web resources, and parsing text.
+Adding fast array operations and linear algebra allows scientist to do all
+their work within a single language---and one that has the advantage of
+being famously easy to learn and teach.
+In fact, Python's adoption as a primary learning language in universities
+(e.g., many lower division UC Berkeley courses use Python including
+Foundations of Data Science,
+The Beauty and Joy of Computing,
+Introduction to Computational Thinking with Data,
+and The Structure and Interpretation of Computer Programs) has
+further popularized NumPy as a tool of choice for modern data science.
 
-NumPy spurred the development of the larger scientific Python ecosystem and
-heralded the current era of wide-spread use of Python for scientific computing.
-There were several factors that allowed this rapid growth and successful
-development.
+Even though NumPy is not part of Python's standard library,
+we benefit from a good relationship with the Python developers.
+Over the years, the Python language has added new features and
+special syntax so that NumPy would have more succinct and 
+easy to read array notation.
+Since we are not part of the standard library, we are able to
+dictate our own release policies and development patterns.
+%\fixme{we've gotten a number of features: slicing syntax, implicit tuples,
+%complex numbers, buffer protocol, matrix multiplication operator}
+%\fixme{but we benefit from separate governance}
 
-Python is already a good choice for standard programming tasks such as
-cleaning data, interacting with web resources, and parsing text.
-It has a wide range of libraries for different tasks.
-Adding fast array operations and linear algebra allows the scientist to do all
-their work with within the same language---and one that has the advantage of
-being famously easy to learn and teach.
-
-NumPy has strong culture of using software engineering practice to
-improve collaboration and reduce error \cite{millman2014developing}.
-The NumPy team was early in adopting distributed revision control and code
-review to improve collaboration on code, and continuous testing that runs a
-large battery of automated tests for every proposed change to NumPy.
-The project has comprehensive, high-quality documentation, integrated with the
-source code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
+NumPy is a community of practice with a strong \emph{culture} of
+employing software engineering practice to improve collaboration and
+reduce error \cite{millman2014developing}.  This culture is not only
+adopted by leaders in the project, but also enthusiastically taught to
+newcomers. The NumPy team was early in adopting distributed revision
+control and code review to improve collaboration on code, and
+continuous testing that runs a large battery of automated tests for
+every proposed change to NumPy.  The project has comprehensive,
+high-quality documentation, integrated with the source
+code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
 
@@
 including use of version control, unit testing, code review and issue tracking
-procedures. This is a vital skillset for modern researchers that is often
+procedures. This is a vital skill set for modern researchers that is often
 missing from formal university education in physics or astronomy.}
@@
 \cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
-%\fixme{cite https://scipy-school.org/}.
 
-Lastly, NumPy benefited greatly from the simplicity of its underlying data
-structure, which made it a natural standard for exchanging array data between
-libraries.  It also made it easy for other libraries to develop fast and
+NumPy combines the expressive power of \emph{array programming}, 
+the performance of C, and the Python programming language in a mature,
+well-tested, well-documented, and community developed library.
+It builds on decades of work refining the array programming
+syntax from APL through Yorick and others.
+It also made it easy for other libraries to develop fast and
 memory-efficient compiled code, usually in C or Fortran, that could manipulate
 these arrays and pass them back to Python.
+Popular array-aware libraries increased the user community and that
+in turn led to more libraries adding new functionality
+to the scientific Python ecosystem.
 
-Over time, and with the arrival of different technologies and array types, the
-role of NumPy has changed from merely being an array provider, to being a
-widely adopted array programming language.
-
-% S:
-%
-% Thinking about enabling factors, I'd characterize them into three categories:
-%
-% - Practical
-% - Philosophical
-% - Social
-% - Technical
-%
-% E.g., practical: reason why we use Python (learn one language for
-% everything); students don't have money, so want to avoid impracticalities of
-% license dongles/servers, etc.
-%
-% Philosophical: science should be open, transparent; our software should be
-% controlled by scientists, not designers that we don't have access to
-%
-% Social: joy of building these things together, friendly welcome into the
-% community for many of us---appreciation of our work and hours
+Over time the role of the NumPy has evolved. It is no longer just the
+foundational array library underlying the scientific Python ecosystem,
+but it is also a standard API for tensor computation and a
+central coordinating mechanism between array types and technologies.

### 2020-01-25 body.tex
Restructure discussion

--- 2020-01-24 body.tex
@@
 
-NumPy was initially developed by students, faculty, and researchers in their
-spare time to provide a modern array programming library for Python.  They were
-influenced by their experiences with powerful interactive programming languages
-for scientific computing like Yorick \cite{munro1995using} as well as
-commercial languages and environments like IDL and Matlab.  They also
-frequently developed code from scratch to solve their own or their colleagues'
-problems, often in low-level languages that preceded Python like Fortran
-\cite{dongarra2008netlib} and C.
+NumPy was initially developed by students, faculty, and researchers to
+provide a modern array programming library for Python.  These
+user-developers frequently had to write code from scratch to solve
+their or their colleagues' problems---often in low-level languages
+that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
+To them, the advantages of an interactive, high-level array library
+were evident. Their design of this new tool was informed by their
+experiences with powerful interactive programming languages for
+scientific computing such as APL \cite{iverson1962programming} and
+Yorick \cite{munro1995using}, as well as commercial languages and
+environments like IDL and Matlab.
+
+NumPy has subsequently seen prolific adoption, with millions of users,
+first from research and later also industry.  It spurred the
+development of the larger scientific Python ecosystem and heralded the
+current era of wide-spread use of Python for scientific computing. The
+reasons for its success are certainly complex, but we argue that they
+are broadly of \emph{practical}, \emph{philosophical},
+\emph{social}, and \emph{technical} origins.
 
@@
 
-This description of how APL users work easily applies to scientific Python
-users \cite{kromberg2007apl}:
+The \emph{practical} appeal of NumPy and its ecosystem is apparent
+when you consider the following description of how APL users
+work---one which
+easily applies to scientific Python users too \cite{kromberg2007apl}:
 \begin{quote}
@@
 
-NumPy spurred the development of the larger scientific Python ecosystem and
-heralded the current era of wide-spread use of Python for scientific computing.
-There were several factors that allowed this rapid growth and successful
-development.
+This modality of interactive scientific exploration would develop
+further and eventually culminate in the Jupyter ecosystem of tools \fixme{citation needed}.
 
-Python is already a good choice for standard programming tasks such as
+NumPy also benefited from being written in Python, a
+general-purpose language well-suited to standard programming tasks such as
 cleaning data, interacting with web resources, and parsing text.
-It has a wide range of libraries for different tasks.
-Adding fast array operations and linear algebra allows the scientist to do all
-their work with within the same language---and one that has the advantage of
-being famously easy to learn and teach.
+Adding fast array operations and linear algebra allows scientist to do all
+their work with within a single language---and one that has the advantage of
+being famously easy to learn and teach.  In fact, Python's adoption as
+a primary learning language in many universities \fixme{citation
+needed} may well have further popularize NumPy as a tool of choice for
+modern data science.
 
-NumPy has strong culture of using software engineering practice to
-improve collaboration and reduce error \cite{millman2014developing}.
-The NumPy team was early in adopting distributed revision control and code
-review to improve collaboration on code, and continuous testing that runs a
-large battery of automated tests for every proposed change to NumPy.
-The project has comprehensive, high-quality documentation, integrated with the
-source code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
+As an open source project, NumPy also did not cost money, and was
+unencumbered by license servers, dongles, and the likes.
+
+In terms of \emph{philosophical} appeal, NumPy is open source
+software.  So, of course, are many other packages that are no longer
+relevant, and by itself it does not motivate NumPy's sucess. However,
+many developers would never have participated in building a closed
+project.  Open source software also appealed the many proponents of
+reproducible science, who wanted a transparent, introspectable stack
+of software underlying their results.
+
+\emph{Socially}, there was a sense of building something
+impactful together, for the benefit of many others.  Participating in
+such an endeavor, within a welcoming community of like-minded
+indivuals, held a powerful attraction for many early contributors.
+
+Throughout its development, NumPy has upheld strong \emph{technical}
+principles.  It is a community of practice that enshrines a culture of
+employing software engineering practice to improve collaboration and
+reduce error \cite{millman2014developing}.  This culture is not only
+adopted by leaders in the project, but also enthusiastically taught to
+newcomers. The NumPy team was early in adopting distributed revision
+control and code review to improve collaboration on code, and
+continuous testing that runs a large battery of automated tests for
+every proposed change to NumPy.  The project has comprehensive,
+high-quality documentation, integrated with the source
+code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
 
@@
 
-Over time, and with the arrival of different technologies and array types, the
-role of NumPy has changed from merely being an array provider, to being a
-widely adopted array programming language.
+Over time the role of the NumPy has changed. It is no longer merely an
+array library, but also a standard API for tensor computation and a
+central coordinating mechanism between array types and technologies.
+NumPy has become a fully-fledged array programming paradigm, that
+continues to drive numerical exploration and computation in scientific
+Python.
 

### 2020-01-25 methods.tex
Some small tweaks/suggestions for the method section

@@
 This allows us to review code before integrating it as well as to run a
-large number of tests on the modified code to ensure that it does not
-break things.
+large number of tests on the modified code to ensure a bug free and stable
+experience for our users.
 
@@
 Broadly, the NumPy library consists of the following parts:
-the NumPy array data structure; the so-called \emph{universal functions};
+the NumPy array data structure \code{ndarray}; the so-called \emph{universal functions};
 a set of library functions for manipulating arrays and doing scientific
@@
 The data structure stores regularly strided homogeneous data types
-in a contiguous block memory, allowing for the efficient representation
+inside a contiguous block memory, allowing for the efficient representation
 of $n$-dimensional data.
@@
 \code{y} is a 2-D array with shape $2 \times 1$, the output will be
-an array with shape $2 \times 3$.  The ufunc machinery takes care
+an array with shape $2 \times 3$ (Fig.~\ref{fig:array-concepts}c).
+The ufunc machinery takes care
 of calling the function with all the appropriate combinations of
@@
 The elementary arithmetic operations of addition, multiplication, etc.,
-are implemented as ufuncs, so broadcasting applies even in expressions
+are implemented as ufuncs, so broadcasting applies also to expressions
 such as \code{x + y * z}.
@@
 \paragraph{\emph{Infrastructure libraries.}} NumPy provides utilities
-for writing unit tests and for building Python packages.
+for writing tests and for building Python packages.
 
 The \code{testing} subpackage provides functions such as
-\code{assert\_allclose(actual, desired)} that may be used in unit
+\code{assert\_allclose(actual, desired)} that may be used in
 test suites for code that uses NumPy arrays.
 
-NumPy provides a subpackage includes functions and classes
-that can be used by libraries that have NumPy as dependency,
-to be used in the library's configuration code that installs the
-library and that builds a package that can be used, for example,
-on the PyPI web site.
+NumPy provides the subpackage \code{distutils} which includes functions and classes
+to facilitate configuration, installation, and packaging of libraries depending on NumPy.
+These can be used, for example, to as part of publishing to the PyPI website.
 
@@
 Project decisions are usually made by consensus of interested contributors.
-This means for many thing everyone is entrusted with veto power.
+This means for most decisions everyone is entrusted with veto power.
 A Steering Council, currently composed of 12~members, facilitates this
@@
 Finally, since May 2019 the project receives around 10K USD annually from
+% typically is correct right now, but do we want to write it so specifically?
 Tidelift, which is typically used to fund documentation and website
@@
 a tool they need may not work for them.
-It also creates new challenges for end-users who need to transition
+It also creates challenges for end-users who need to transition
 from NumPy to a more specialized array.
@@
 \code{\_\_array\_function\_\_} protocol into NumPy~1.17.
-These two protocols allows providers of array-like APIs to be interoperable
+These two protocols allow providers of array objects to be interoperable
 with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
@@
 interfaces with the underlying code that transforms random bits into
-deviates from other distributions; and supplies a singleton instance exposed in
+variates from other distributions; and supplies a singleton instance exposed in
 the root of the random module.
@@
 high-quality pseudorandom integers, which can be used to initialize multiple bit
-generators deterministically. A key design goal of a seed sequence was to be
-splittable in the sense that a seed sequence can be used to produce child
-sequences that are distinct from their parents or other ancestors. This
-capability allows a seed sequence to be used in large distributed applications
+generators deterministically. A key design element of a seed sequence is that
+it can be used to spawn child \code{SeedSequence}s which are distinct from
+their parent and all ancestors.
+This capability allows a seed sequence to facilitate large distributed applications
 where the number of workers required is not known. The sequences generated from
-the same initial entropy and the same splits are fully deterministic to ensure
+the same initial entropy and spawns are fully deterministic to ensure
 reproducibility.

### 2020-01-25 body.tex
Minor edits

@@
 
+This includes popular tensor computation libraries such as
+CuPy\footnote{\url{https://cupy.chainer.org/}},
+JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}, and
+Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
+and
+Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
+provide tensor APIs with NumPy-inspired semantics.
+
 \section*{Interoperability and extensibility}
@@
 \end{quote}
-
 This modality of interactive scientific exploration would develop
-further and eventually culminate in the Jupyter ecosystem of tools \fixme{citation needed}.
+further and eventually culminate in the Jupyter ecosystem of tools \cite{Kluyver:2016aa}.
 
@@
 being famously easy to learn and teach.  In fact, Python's adoption as
-a primary learning language in many universities \fixme{citation
-needed} may well have further popularize NumPy as a tool of choice for
-modern data science.
+a primary learning language in many universities (e.g., at UC Berkeley
+many lower division courses use Python including Foundations of Data Science,
+Introduction to Computational Thinking with Data,
+and The Structure and Interpretation of Computer Programs) have
+further popularized NumPy as a tool of choice for modern data science.
 

### 2020-01-25 paper.tex
Acknowledge larger community

@@
 
+We would also like to thank the many members of the community who have provided
+feedback, submitted bug reports, made minor improvements to the documentation,
+code, or website, promoted its use in their scientific fields, and built
+the vast ecosystem of tools and libraries around NumPy.
+
 \section*{Author Contributions Statement}
@@
 %C.R.H and T.E.O.
-S.J.v.d.W and K.J.M. composed the manuscript with input from others.
+K.J.M. and S.J.v.d.W composed the manuscript with input from others.
 S.B., K.S., T.J.R., W.W., and M.B. drafted sections.

### 2020-01-25 body.tex
Revise discussion

--- 2020-01-25 body.tex
@@
 interactive, exploratory data analysis where users may iteratively inspect,
-manipulate, and visualize their data \cite{perez2007ipython}.
+manipulate, and visualize their data.
+
+IPython \cite{perez2007ipython}
+
+Jupyter \cite{Kluyver:2016aa}
 
@@
 
-% aka why is this so successful
-% some sense of ongoing work / future directions
-
 NumPy was initially developed by students, faculty, and researchers to
@@
 NumPy has subsequently seen prolific adoption, with millions of users,
-first from research and later also industry.  It spurred the
-development of the larger scientific Python ecosystem and heralded the
-current era of wide-spread use of Python for scientific computing. The
-reasons for its success are certainly complex, but we argue that they
-are broadly of \emph{practical}, \emph{philosophical},
-\emph{social}, and \emph{technical} origins.
+first from research and later also industry.
+It spurred the development of the larger scientific Python ecosystem and
+heralded the current era of wide-spread use of Python for scientific computing.
+The reasons for its success are complex, but include the fact NumPy
+(a) is built on top of Python, (b) has a strong culture of collaboration
+and good software development practices, and (c) provides one of the
+best array programming environments for scientific data analysis.
 
-% Fortran / C
-% A few sentences on the history of array programming, referencing APL and Fortran, are missing.
+\emph{Python} is an open source, general-purpose, interpreted programming language
+well-suited to standard programming tasks such as cleaning data,
+interacting with web resources, and parsing text.
+Adding fast array operations and linear algebra allows scientist to do all
+their work within a single language---and one that has the advantage of
+being famously easy to learn and teach.
+In fact, Python's adoption as
+a primary learning language in many universities (e.g., many lower division
+UC Berkeley courses use Python including Foundations of Data Science,
+The Beauty and Joy of Computing,
+Introduction to Computational Thinking with Data,
+and The Structure and Interpretation of Computer Programs) has
+further popularized NumPy as a tool of choice for modern data science.
 
-The \emph{practical} appeal of NumPy and its ecosystem is apparent
-when you consider the following description of how APL users
-work---one which
-easily applies to scientific Python users too \cite{kromberg2007apl}:
-\begin{quote}
-Users of APL tend to ``live in their data''. Interactive interpreters allow them
-to inspect, almost feel their  way forward, discovering successful snippets
-of code through experimentation and collecting them
-into imperative or functional programs according to taste.
-\end{quote}
+Even though NumPy is not part of Python's standard library,
+we have a health relationship to the core development team.
+\fixme{we've gotten a number of features: slicing syntax, implicit tuples,
+complex numbers, buffer protocol, matrix multiplication operator}
+\fixme{but we benefit from separate governance}
 
-This modality of interactive scientific exploration would develop
-further and eventually culminate in the Jupyter ecosystem of tools \fixme{citation needed}.
-
-NumPy also benefited from being written in Python, a
-general-purpose language well-suited to standard programming tasks such as
-cleaning data, interacting with web resources, and parsing text.
-Adding fast array operations and linear algebra allows scientist to do all
-their work with within a single language---and one that has the advantage of
-being famously easy to learn and teach.  In fact, Python's adoption as
-a primary learning language in many universities \fixme{citation
-needed} may well have further popularize NumPy as a tool of choice for
-modern data science.
-
-As an open source project, NumPy also did not cost money, and was
-unencumbered by license servers, dongles, and the likes.
-
-In terms of \emph{philosophical} appeal, NumPy is open source
-software.  So, of course, are many other packages that are no longer
-relevant, and by itself it does not motivate NumPy's sucess. However,
-many developers would never have participated in building a closed
-project.  Open source software also appealed the many proponents of
-reproducible science, who wanted a transparent, introspectable stack
-of software underlying their results.
-
-\emph{Socially}, there was a sense of building something
-impactful together, for the benefit of many others.  Participating in
-such an endeavor, within a welcoming community of like-minded
-indivuals, held a powerful attraction for many early contributors.
-
-Throughout its development, NumPy has upheld strong \emph{technical}
-principles.  It is a community of practice that enshrines a culture of
+NumPy is a community of practice with a strong \emph{culture} of
 employing software engineering practice to improve collaboration and
@@
 \cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
-%\fixme{cite https://scipy-school.org/}.
 
-Lastly, NumPy benefited greatly from the simplicity of its underlying data
-structure, which made it a natural standard for exchanging array data between
-libraries.  It also made it easy for other libraries to develop fast and
+The \emph{array programming} paradigm originated with APL in the 50s and 60s.
+It is characterized by working interactively with array data using
+vectorized operations.
+Its draw for scientist is captured by the following description of
+how APL users work \cite{kromberg2007apl}:
+\begin{quote}
+Users of APL tend to ``live in their data''. Interactive interpreters allow them
+to inspect, almost feel their  way forward, discovering successful snippets
+of code through experimentation and collecting them
+into imperative or functional programs according to taste.
+\end{quote}
+Scientist exploring data at the IPython command prompt with the
+whole scientific Python ecosystem available to them often work in
+this way.  This type of interactive scientific exploration is
+increasingly taking place in Jupyter notebooks.
+
+NumPy combines the performance of C, the expressive power of array
+programming, and the Python programming language in a mature
+well-tested and well-documented community developed library.
+It builds on decades of work refining the array programming
+syntax from APL through Yorick and others.
+It also made it easy for other libraries to develop fast and
 memory-efficient compiled code, usually in C or Fortran, that could manipulate
 these arrays and pass them back to Python.
+Popular libraries increased the user community and that
+in turn led to new libraries adding additional functionality
+to the scientific Python ecosystem in a virtuous cycle.
 
-Over time the role of the NumPy has changed. It is no longer merely an
-array library, but also a standard API for tensor computation and a
+Over time the role of the NumPy has evolved. It is no longer just the
+foundational array library underlying the scientific Python ecosystem,
+but it is also a standard API for tensor computation and a
 central coordinating mechanism between array types and technologies.
-NumPy has become a fully-fledged array programming paradigm, that
-continues to drive numerical exploration and computation in scientific
-Python.
-
-% S:
-%
-% Thinking about enabling factors, I'd characterize them into three categories:
-%
-% - Practical
-% - Philosophical
-% - Social
-% - Technical
-%
-% E.g., practical: reason why we use Python (learn one language for
-% everything); students don't have money, so want to avoid impracticalities of
-% license dongles/servers, etc.
-%
-% Philosophical: science should be open, transparent; our software should be
-% controlled by scientists, not designers that we don't have access to
-%
-% Social: joy of building these things together, friendly welcome into the
-% community for many of us---appreciation of our work and hours

### 2020-01-25 methods.tex
Method fixups based on review comments

@@
 The elementary arithmetic operations of addition, multiplication, etc.,
-are implemented as ufuncs, so broadcasting applies also to expressions
+are implemented as ufuncs, so that broadcasting also applies to expressions
 such as \code{x + y * z}.
@@
 to facilitate configuration, installation, and packaging of libraries depending on NumPy.
-These can be used, for example, to as part of publishing to the PyPI website.
+% Remove the PyPI reference entirely?
+These can be used, for example, when publishing to the PyPI website.
 
@@
 Project decisions are usually made by consensus of interested contributors.
-This means for most decisions everyone is entrusted with veto power.
+This means that, for most decisions, everyone is entrusted with veto power.
 A Steering Council, currently composed of 12~members, facilitates this
@@
 Finally, since May 2019 the project receives around 10K USD annually from
-% typically is correct right now, but do we want to write it so specifically?
-Tidelift, which is typically used to fund documentation and website
+Tidelift, a part of which is used to fund documentation and website
 improvements.
@@
 high-quality pseudorandom integers, which can be used to initialize multiple bit
-generators deterministically. A key design element of a seed sequence is that
-it can be used to spawn child \code{SeedSequence}s which are distinct from
-their parent and all ancestors.
+generators deterministically. The key feature of a seed sequence is that
+it can be used to spawn child \code{SeedSequence}s which allow to initialize
+multiple distinct bit generators.
+% Of course there is a diminishing chance of collisions...
+% Only found http://www.pcg-random.org/posts/developing-a-seed_seq-alternative.html
+% as a reference for seed sequence, it would be nice to cite something.
 This capability allows a seed sequence to facilitate large distributed applications

### 2020-01-25 body.tex
Adopt review comments from Stefan in interop sectio

@@
 
-Users predominantly interact with NumPy arrays using {\em indexing} (to access
-subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
-and, for matrix multiplication, $@$), as well as {\em array-aware functions};
-together, these provide an easily readable, expressive, high-level API for
-array programming.
-
-Various libraries build on and extend this functionality to further enrich the
-ecosystem of available tools.  For example, SciPy adds various mathematical,
-scientific, and engineering routines; Pandas provides indexed arrays, known as
-dataframes; xarray implements arrays with labeled axes; and Matplotlib
-generates publication-ready figures and visualizations.  Employing these
-low-level array-aware functions are widely used technique-specific libraries,
-including scikit-learn (machine learning), scikit-image (image processing),
-statsmodels (statistics),  networkx (graphs and complex networks)
-\cite{SciPyProceedings_11}, and Napari (volumetric image visualization).
-
-Finally, domain-specific libraries and applications are built on this diverse
-foundation of tooling; these include Astropy (astronomy) \cite{astropy:2013,
-astropy:2018}, Biopython (computational biology and bioinformatics) \cite{cock2009biopython},
-NIPY (neuroimaging) \cite{millman2007analysis}, and Pangeo (earth science) \cite{2018EGUGA..2012146H}.
-
-Exposing array programming primitives, as well as the surrounding ecosystem of
-tools, in Python---an interpreted language---creates an ideal environment for
+The brevity and expressive power of NumPy's array programming syntax combined with 
+an advanced interactive environment like IPython \cite{perez2007ipython}
+or Jupyter \cite{Kluyver:2016aa} makes interactive data exploration fluid and simple.
+It is common for NumPy users to work interactively with their data,
+discovering through experimentation bits of code that they stitch together into
+imperative or functional programs.
+The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
+this solid foundation with core (e.g., SciPy and Matplotlib) libraries,
+widely used technique-specific
+libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
+and several domain-specific projects and libraries
+\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
+Exposing these simple array programming primitives, as well as the surrounding ecosystem of
+tools, in Python---an interpreted, general purpose language---creates an ideal environment for
 interactive, exploratory data analysis where users may iteratively inspect,
 manipulate, and visualize their data.
+This rich environment has made the scientific Python ecosystem popular for scientific
+research.
 
-IPython \cite{perez2007ipython}
+\begin{figure}
+  \centering
+  \includegraphics[width=.48\textwidth]{static/ecosystem}
+  \caption{\textbf{Scientific Python Ecosystem.}
+   This figure highlights some of the essential tools and libraries in the vast
+   scientific Python ecosystem.  Most of the ecosystem builds on NumPy to
+   provide a rich and powerful collection of array-aware functionality.
+  }
+  \label{fig:ecosystem}
+\end{figure}
 
-Jupyter \cite{Kluyver:2016aa}
+For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
+collaboration relies on many components of the scientific Python ecosystem.
+NumPy arrays are used to store and manipulate numerical data at every step
+in the processing chain: from raw data through calibration and image 
+reconstruction.
+Scipy is used to provide tools used in general image processing such as 
+filtering and image alignment, while scikit-image provides higher-level
+functionality such as edge filters and Hough transforms.
+SciPy provides fundamental algorithms for scientific computing on arrays
+including many mathematical, scientific, and engineering routines.
+SciPy and NumPy are closely related and were developed together often
+by the same people.
+scikit-image is a popular technique-specific library,
+which vastly extends the basic image processing tools provided by SciPy.
+Other components of Scipy are used throughout the library, such as the
+\code{scipy.optimize} package for handling general optimization tasks.
+NetworkX \cite{SciPyProceedings_11} (the standard package for complex
+network analysis) is used for analysis of consistency among image
+comparisons.
+Astropy \cite{astropy:2013, astropy:2018} (a community-developed astronomy
+library) is used throughout the library,
+primarily for I/O from standard astronomical file formats and time/coordinate
+transformations.
+Matplotlib (the core library for generating publication-ready figures
+and visualizations) is used for visualizing data throughout the analysis pipeline,
+including the generation of the final image of the black hole.
+% N.B. - most of this comes from introspecting the Image class defined in
+% ehtim/image.py and grepping the dependencies in requirements.txt
 
-\section*{The proliferation of arrays}
+Tools for analyzing data from gravitational wave observatories such as LIGO
+and Virgo are provided by the \code{pycbc} package, which was used in the first
+detection of gravitational waves \cite{abbott2016observation} and in 
+on-going analysis of data from LIGO and Virgo.
+\code{pycbc} make extensive use of the scientific Python ecosystem.
+The time-series data from the interferometers are stored in NumPy arrays,
+while the \code{scipy.signal} is used to construct and apply filters to the
+data.
+Matplotlib is used to visualize data at all points along the analysis chain
+including raw data from the instrumentation and the time-frequency 
+visualization of the ``chirps'' that result from mergers of compact binaries
+such as black holes\cite{abbott2016observation} and
+neutron stars\cite{abbott2017multi}.
 
-\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
-https://parquet.apache.org/,
-https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
+The rich ecosystem of tools building on NumPy and the larger community
+of scientists and programmers using those tools mean that NumPy has a
+responsibility to maintain considerable stability.
+This need creates interesting challenges for NumPy development as the
+computational environment scientists are analyzing their data on continues
+to evolve and change.
+
+\section*{Proliferation of arrays}
+
+%\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
+%https://parquet.apache.org/,
+%https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
+
+% pandas, xarray, ...
 
@@
 This includes popular tensor computation libraries such as
-CuPy\footnote{\url{https://cupy.chainer.org/}},
-JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}, and
-Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
-and
-Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
-provide tensor APIs with NumPy-inspired semantics.
+CuPy, JAX, and Apache MXNet.
+PyTorch and Tensorflow provide tensor APIs with NumPy-inspired semantics.
+% \footnote{\url{https://cupy.chainer.org/}}
+% \footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}
+% \footnote{\url{https://numpy.mxnet.io/}}
+% \footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
+% \footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
 
@@
 % work by Peter Entschev]".
+% Below an attempt to start on this:
+Thanks to these developments users of Dask and NumPy arrays are now capable
+to scale their computation to the GPU using Dasks CuPy integration with
+only minimal code changes.
+However, general adoption has not yet been achieved mainly for compatibility
+concerns.
+Thus, we are actively addressing these concerns by revising and extending
+these protocols with the aim of wide interoperability that spans the whole
+scientific Python ecosystem up to and including domain libraries.
 
@@
 NumPy was initially developed by students, faculty, and researchers to
-provide a modern array programming library for Python.  These
-user-developers frequently had to write code from scratch to solve
+provide an advanced, open source array programming library for Python,
+which was free to use and unencumbered by license servers, dongles, and the like.
+There was a sense of building something
+impactful together, for the benefit of many others.  Participating in
+such an endeavor, within a welcoming community of like-minded
+indivuals, held a powerful attraction for many early contributors.
+
+These user-developers frequently had to write code from scratch to solve
 their or their colleagues' problems---often in low-level languages
@@
 To them, the advantages of an interactive, high-level array library
-were evident. Their design of this new tool was informed by their
+were evident. The design of this new tool was informed by their
 experiences with powerful interactive programming languages for
@@
 heralded the current era of wide-spread use of Python for scientific computing.
-The reasons for its success are complex, but include the fact NumPy
-(a) is built on top of Python, (b) has a strong culture of collaboration
-and good software development practices, and (c) provides one of the
-best array programming environments for scientific data analysis.
+The reasons for its success are varied and complex, but include the fact that
+NumPy (a) is built on top of Python,
+(b) has a strong culture of collaboration and good software development practices, and
+(c) provides a cutting-edge array programming environment for scientific data analysis.
 
@@
 being famously easy to learn and teach.
-In fact, Python's adoption as
-a primary learning language in many universities (e.g., many lower division
-UC Berkeley courses use Python including Foundations of Data Science,
+In fact, Python's adoption as a primary learning language in universities
+(e.g., many lower division UC Berkeley courses use Python including
+Foundations of Data Science,
 The Beauty and Joy of Computing,
@@
 Even though NumPy is not part of Python's standard library,
-we have a health relationship to the core development team.
-\fixme{we've gotten a number of features: slicing syntax, implicit tuples,
-complex numbers, buffer protocol, matrix multiplication operator}
-\fixme{but we benefit from separate governance}
+we benefit from a good relationship with the Python developers.
+Over the years, the Python language has added new features and
+special syntax so that NumPy would have more succinct and 
+easy to read array notation.
+Since we are not part of the standard library, we are able to
+dictate our own release policies and development patterns.
+%\fixme{we've gotten a number of features: slicing syntax, implicit tuples,
+%complex numbers, buffer protocol, matrix multiplication operator}
+%\fixme{but we benefit from separate governance}
 
@@
 
-The \emph{array programming} paradigm originated with APL in the 50s and 60s.
-It is characterized by working interactively with array data using
-vectorized operations.
-Its draw for scientist is captured by the following description of
-how APL users work \cite{kromberg2007apl}:
-\begin{quote}
-Users of APL tend to ``live in their data''. Interactive interpreters allow them
-to inspect, almost feel their  way forward, discovering successful snippets
-of code through experimentation and collecting them
-into imperative or functional programs according to taste.
-\end{quote}
-Scientist exploring data at the IPython command prompt with the
-whole scientific Python ecosystem available to them often work in
-this way.  This type of interactive scientific exploration is
-increasingly taking place in Jupyter notebooks.
-
-NumPy combines the performance of C, the expressive power of array
-programming, and the Python programming language in a mature
-well-tested and well-documented community developed library.
+NumPy combines the expressive power of \emph{array programming}, 
+the performance of C, and the Python programming language in a mature,
+well-tested, well-documented, and community developed library.
 It builds on decades of work refining the array programming
@@
 these arrays and pass them back to Python.
-Popular libraries increased the user community and that
-in turn led to new libraries adding additional functionality
-to the scientific Python ecosystem in a virtuous cycle.
+Popular array-aware libraries increased the user community and that
+in turn led to more libraries adding new functionality
+to the scientific Python ecosystem.
 

### 2020-01-25 body.tex
Minor edits

--- 2020-01-25 body.tex
@@
 heralded the current era of wide-spread use of Python for scientific computing.
-The reasons for its success are complex, but include the fact NumPy
-(a) is built on top of Python, (b) has a strong culture of collaboration
-and good software development practices, and (c) provides one of the
-best array programming environments for scientific data analysis.
+The reasons for its success are complex, but include the fact that
+NumPy (a) is built on top of Python,
+(b) has a strong culture of collaboration and good software development practices, and
+(c) provides a cutting-edge array programming environment for scientific data analysis.
 

### 2020-01-25 methods.tex
Minor edits

@@
 
-Finally, since May 2019 the project receives around 10K USD annually from
-Tidelift, a part of which is used to fund documentation and website
+Finally, since May 2019 the project receives a small amount annually from
+Tidelift, which is used to fund things like documentation and website
 improvements.
@@
 generators deterministically. The key feature of a seed sequence is that
-it can be used to spawn child \code{SeedSequence}s which allow to initialize
+it can be used to spawn child \code{SeedSequence}s to initialize
 multiple distinct bit generators.

### 2020-01-26 body.tex
Work on outline and ecosystem section

@@
 
-To manipulate the data structure in an intuitive manner, we provide users with
-a simple yet powerful array expression syntax.
-It supports operations such as indexing (selecting subarrays) and elementwise
-calculations.
-A high level syntax allows users to express themselves succinctly, while NumPy
+Users predominantly interact with NumPy arrays using {\em indexing} (to access
+subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
+and, for matrix multiplication, $@$), as well as {\em array-aware functions};
+together, these provide an easily readable, expressive, high-level API for
+array programming, while NumPy
 deals with the underlying mechanics of making operations fast.
@@
 
-\section*{Supporting the ecosystem}
+The expressive power and brevity of NumPy's array programming syntax combined with 
+an advanced interactive environment like IPython \cite{perez2007ipython}
+or Jupyter \cite{Kluyver:2016aa} make interactive data exploration fluid and simple.
 
-\fixme{Maybe some wording around the actual *support* element?}
+\section*{Scientific Python ecosystem}
 
@@
 
-Users predominantly interact with NumPy arrays using {\em indexing} (to access
-subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
-and, for matrix multiplication, $@$), as well as {\em array-aware functions};
-together, these provide an easily readable, expressive, high-level API for
-array programming.
+While NumPy provides a solid foundation for array programming, the scientific
+Python ecosystem is composed of many different libraries, which build on and
+extend this functionality.
+These libraries can roughly be organized into three layers:
+core libraries, technique-specific libraries, and domain-specific libraries.
 
-Various libraries build on and extend this functionality to further enrich the
-ecosystem of available tools.  For example, SciPy adds various mathematical,
-scientific, and engineering routines; Pandas provides indexed arrays, known as
-dataframes; xarray implements arrays with labeled axes; and Matplotlib
-generates publication-ready figures and visualizations.  Employing these
-low-level array-aware functions are widely used technique-specific libraries,
-including scikit-learn (machine learning), scikit-image (image processing),
-statsmodels (statistics),  networkx (graphs and complex networks)
-\cite{SciPyProceedings_11}, and Napari (volumetric image visualization).
+The core libraries include SciPy, Matplotlib, Pandas, and xarray.
+SciPy adds various mathematical, scientific, and engineering routines;
+Matplotlib generates publication-ready figures and visualizations,
+Pandas provides indexed arrays, known as dataframes; and
+xarray implements arrays with labeled axes.
+
+Employing these low-level array-aware functions are widely used
+technique-specific libraries, including scikit-learn (machine learning),
+scikit-image (image processing), statsmodels (statistics),  networkx (graphs
+and complex networks) \cite{SciPyProceedings_11}, and Napari (volumetric image
+visualization).
 
@@
 
+% we should have an example of how these tools may all be used in an
+% example analysis.  maybe one of the ones mentioned in the abstract?
+
 Exposing array programming primitives, as well as the surrounding ecosystem of
@@
 manipulate, and visualize their data.
+The rich ecosystem of tools building on NumPy and the larger community
+of scientists and programmers using those tools mean that NumPy has a
+responsibility to maintain considerable stability.
+This need creates interesting challenges for NumPy development as the
+computational environment scientists are analyzing their data on continues
+to evolve and change.
 
-IPython \cite{perez2007ipython}
+\section*{Proliferation of arrays}
 
-Jupyter \cite{Kluyver:2016aa}
-
-\section*{The proliferation of arrays}
-
-\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
-https://parquet.apache.org/,
-https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
+%\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
+%https://parquet.apache.org/,
+%https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
 

### 2020-01-26 body.tex
Revise discussion

@@
 or Jupyter \cite{Kluyver:2016aa} make interactive data exploration fluid and simple.
+It is common for NumPy users to work interactively with their data,
+discovering through experimentation bits of code that they stich together into
+imperative or functional programs.
 
@@
 
-The \emph{array programming} paradigm originated with APL in the 50s and 60s.
-It is characterized by working interactively with array data using
-vectorized operations.
-Its draw for scientist is captured by the following description of
-how APL users work \cite{kromberg2007apl}:
-\begin{quote}
-Users of APL tend to ``live in their data''. Interactive interpreters allow them
-to inspect, almost feel their  way forward, discovering successful snippets
-of code through experimentation and collecting them
-into imperative or functional programs according to taste.
-\end{quote}
-Scientist exploring data at the IPython command prompt with the
-whole scientific Python ecosystem available to them often work in
-this way.  This type of interactive scientific exploration is
-increasingly taking place in Jupyter notebooks.
-
-NumPy combines the performance of C, the expressive power of array
-programming, and the Python programming language in a mature
+NumPy combines the expressive power of \emph{array programming}, 
+the performance of C, and the Python programming language in a mature
 well-tested and well-documented community developed library.

### 2020-01-26 body.tex
Typo

@@
 NumPy combines the expressive power of \emph{array programming}, 
-the performance of C, and the Python programming language in a mature
-well-tested and well-documented community developed library.
+the performance of C, and the Python programming language in a mature,
+well-tested, well-documented, and community developed library.
 It builds on decades of work refining the array programming

### 2020-01-26 body.tex
Brief accounting of eht-imaging and pycbc libraries.
Added a few sentences on how components of the scientific
python ecosystem are used by pycbc and eht-imaging for the
analysis of gravitational wave data and black hole imaging,
respectively.

Focused mostly on numpy, scipy, and matplotlib.

@@
 % example analysis.  maybe one of the ones mentioned in the abstract?
+
+The \tt{eht-imaging} library developed by the Event Horizon Telescope 
+collabortion relies on many components of the scientific Python ecosystem.
+NumPy arrays were used to store and manipulate numerical data at every step
+in the processing chain: from raw data through calibration and image 
+reconstruction.
+Scipy was used to provide tools used in general image processing such as 
+filtering and image alignment, while scikit\_image provided higher-level
+functionality such as edge filters and Hough transforms.
+Other components of Scipy are used throughout the library, such as the
+\tt{scipy.optimize} package for handling general optimization tasks.
+Matplotlib is used for visualizing data throughout the analysis pipeline,
+including the generation of the final image of the black hole.
+% N.B. - most of this comes from introspecting the Image class defined in
+% ehtim/image.py.
+
+Tools for analyzing data from gravitational wave observatories such as LIGO
+and Virgo are provided by the \tt{pycbc} package, which was used in the first
+detection of gravitationaly waves \cite{abbott2016observation} and in 
+on-going analysis of data from LIGO and Virgo.
+\tt{pycbc} make extensive use of the scientific Python ecosystem.
+The time-series data from the interferometers are stored in NumPy arrays,
+while the \tt{scipy.signal} is used to construct and apply filters to the
+data.
+Matplotlib is used to visualize data at all points along the analysis chain
+including raw data from the instrumentation and the time-frequency 
+visualization of the ``chirp'' from the binary black hole merger.
 

### 2020-01-26 body.tex
Rearrange ecosystem section

@@
 Python ecosystem is composed of many different libraries, which build on and
-extend this functionality.
-These libraries can roughly be organized into three layers:
-core libraries, technique-specific libraries, and domain-specific libraries.
+extend this functionality \fixme{(Fig.)}.
+SciPy provides fundamental algorithms for scientific computing on arrays
+including many mathematical, scientific, and engineering routines.
+SciPy and NumPy are closely related and were developed together often
+by the same people.
+Matplotlib generates publication-ready figures and visualizations.
+Employing these core array-aware libraries are widely used
+technique-specific libraries\cite{SciPyProceedings_11}.
+Finally, several domain-specific
+libraries \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}
+are built on this diverse foundation of tooling.
 
-The core libraries include SciPy, Matplotlib, Pandas, and xarray.
-SciPy adds various mathematical, scientific, and engineering routines;
-Matplotlib generates publication-ready figures and visualizations,
-Pandas provides indexed arrays, known as dataframes; and
-xarray implements arrays with labeled axes.
+Exposing array programming primitives, as well as the surrounding ecosystem of
+tools, in Python---an interpreted language---creates an ideal environment for
+interactive, exploratory data analysis where users may iteratively inspect,
+manipulate, and visualize their data.
+This has made the scientific Python ecosystem popular for scientific
+research.
 
-Employing these low-level array-aware functions are widely used
-technique-specific libraries, including scikit-learn (machine learning),
-scikit-image (image processing), statsmodels (statistics),  networkx (graphs
-and complex networks) \cite{SciPyProceedings_11}, and Napari (volumetric image
-visualization).
+%These libraries can roughly be organized into three layers:
+%core libraries, technique-specific libraries, and domain-specific libraries.
 
-Finally, domain-specific libraries and applications are built on this diverse
-foundation of tooling; these include Astropy (astronomy) \cite{astropy:2013,
-astropy:2018}, Biopython (computational biology and bioinformatics) \cite{cock2009biopython},
-NIPY (neuroimaging) \cite{millman2007analysis}, and Pangeo (earth science) \cite{2018EGUGA..2012146H}.
+%SciPy adds various mathematical, scientific, and engineering routines;
+%Matplotlib generates publication-ready figures and visualizations,
+%Pandas provides indexed arrays, known as dataframes; and
+%xarray implements arrays with labeled axes.
 
@@
 
-The \tt{eht-imaging} library developed by the Event Horizon Telescope 
-collabortion relies on many components of the scientific Python ecosystem.
+For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
+collaboration relies on many components of the scientific Python ecosystem.
 NumPy arrays were used to store and manipulate numerical data at every step
@@
 Other components of Scipy are used throughout the library, such as the
-\tt{scipy.optimize} package for handling general optimization tasks.
+\code{scipy.optimize} package for handling general optimization tasks.
+NetworkX \cite{SciPyProceedings_11} was used for image comparisons.
+Astropy \cite{astropy:2013, astropy:2018} is used in quite a few places,
+primarily for I/O and time/coordinate transformations.
 Matplotlib is used for visualizing data throughout the analysis pipeline,
@@
 Tools for analyzing data from gravitational wave observatories such as LIGO
-and Virgo are provided by the \tt{pycbc} package, which was used in the first
+and Virgo are provided by the \code{pycbc} package, which was used in the first
 detection of gravitationaly waves \cite{abbott2016observation} and in 
 on-going analysis of data from LIGO and Virgo.
-\tt{pycbc} make extensive use of the scientific Python ecosystem.
+\code{pycbc} make extensive use of the scientific Python ecosystem.
 The time-series data from the interferometers are stored in NumPy arrays,
-while the \tt{scipy.signal} is used to construct and apply filters to the
+while the \code{scipy.signal} is used to construct and apply filters to the
 data.
@@
 
-Exposing array programming primitives, as well as the surrounding ecosystem of
-tools, in Python---an interpreted language---creates an ideal environment for
-interactive, exploratory data analysis where users may iteratively inspect,
-manipulate, and visualize their data.
 The rich ecosystem of tools building on NumPy and the larger community

### 2020-01-26 body.tex
Minor edits

@@
 
-Users predominantly interact with NumPy arrays using {\em indexing} (to access
-subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
-and, for matrix multiplication, $@$), as well as {\em array-aware functions};
+Users interact with NumPy arrays using {\em indexing} (to access
+subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and $/$
+for elementwise operations and $@$ for matrix multiplication), as well as {\em array-aware functions};
 together, these provide an easily readable, expressive, high-level API for
@@
 
-The expressive power and brevity of NumPy's array programming syntax combined with 
-an advanced interactive environment like IPython \cite{perez2007ipython}
-or Jupyter \cite{Kluyver:2016aa} make interactive data exploration fluid and simple.
-It is common for NumPy users to work interactively with their data,
-discovering through experimentation bits of code that they stich together into
-imperative or functional programs.
-
 \section*{Scientific Python ecosystem}
@@
 
-While NumPy provides a solid foundation for array programming, the scientific
-Python ecosystem is composed of many different libraries, which build on and
-extend this functionality \fixme{(Fig.)}.
+The expressive power and brevity of NumPy's array programming syntax combined with 
+an advanced interactive environment like IPython \cite{perez2007ipython}
+or Jupyter \cite{Kluyver:2016aa} makes interactive data exploration fluid and simple.
+It is common for NumPy users to work interactively with their data,
+discovering through experimentation bits of code that they stich together into
+imperative or functional programs.
+Exposing array programming primitives, as well as the surrounding ecosystem of
+tools, in Python---an interpreted language---creates an ideal environment for
+interactive, exploratory data analysis where users may iteratively inspect,
+manipulate, and visualize their data.
+
+The scientific Python ecosystem \fixme{(Fig.)} is composed of many different libraries.
 SciPy provides fundamental algorithms for scientific computing on arrays
@@
 are built on this diverse foundation of tooling.
-
-Exposing array programming primitives, as well as the surrounding ecosystem of
-tools, in Python---an interpreted language---creates an ideal environment for
-interactive, exploratory data analysis where users may iteratively inspect,
-manipulate, and visualize their data.
 This has made the scientific Python ecosystem popular for scientific
@@
 collaboration relies on many components of the scientific Python ecosystem.
-NumPy arrays were used to store and manipulate numerical data at every step
+NumPy arrays are used to store and manipulate numerical data at every step
 in the processing chain: from raw data through calibration and image 
 reconstruction.
-Scipy was used to provide tools used in general image processing such as 
-filtering and image alignment, while scikit\_image provided higher-level
+Scipy is used to provide tools used in general image processing such as 
+filtering and image alignment, while scikit-image provided higher-level
 functionality such as edge filters and Hough transforms.
@@
 \code{scipy.optimize} package for handling general optimization tasks.
-NetworkX \cite{SciPyProceedings_11} was used for image comparisons.
+NetworkX \cite{SciPyProceedings_11} is used for image comparisons.
 Astropy \cite{astropy:2013, astropy:2018} is used in quite a few places,
@@
 heralded the current era of wide-spread use of Python for scientific computing.
-The reasons for its success are complex, but include the fact that
+The reasons for its success are varied and complex, but include the fact that
 NumPy (a) is built on top of Python,

### 2020-01-26 body.tex
Minor edits

@@
 Users interact with NumPy arrays using {\em indexing} (to access
-subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and $/$
+subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$
 for elementwise operations and $@$ for matrix multiplication), as well as {\em array-aware functions};
@@
 Matplotlib generates publication-ready figures and visualizations.
-Employing these core array-aware libraries are widely used
-technique-specific libraries\cite{SciPyProceedings_11}.
-Finally, several domain-specific
-libraries \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}
-are built on this diverse foundation of tooling.
+Employing these array-aware libraries are widely used
+technique-specific libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11}
+and several domain-specific
+libraries \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
 This has made the scientific Python ecosystem popular for scientific

### 2020-01-26 body.tex
Add ecosystem figure

@@
 
-The scientific Python ecosystem \fixme{(Fig.)} is composed of many different libraries.
+\begin{figure}
+  \centering
+  \includegraphics[width=.48\textwidth]{static/ecosystem}
+  \caption{\textbf{Scientific Python Ecosystem.}
+   }
+  \label{fig:ecosystem}
+\end{figure}
+
+The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) is composed of many different libraries.
 SciPy provides fundamental algorithms for scientific computing on arrays
@@
 technique-specific libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11}
-and several domain-specific
+and several domain-specific projects and
 libraries \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.

### 2020-01-26 body.tex
More work on ecosystem section

@@
 
-The expressive power and brevity of NumPy's array programming syntax combined with 
+The brevity and expressive power of NumPy's array programming syntax combined with 
 an advanced interactive environment like IPython \cite{perez2007ipython}
@@
 imperative or functional programs.
-Exposing array programming primitives, as well as the surrounding ecosystem of
-tools, in Python---an interpreted language---creates an ideal environment for
+The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
+this solid foundation with core (e.g., SciPy and Matplotlib) libraries,
+widely used technique-specific
+libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
+and several domain-specific projects and libraries
+\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
+Exposing these simple array programming primitives, as well as the surrounding ecosystem of
+tools, in Python---an interpreted, general purpose language---creates an ideal environment for
 interactive, exploratory data analysis where users may iteratively inspect,
 manipulate, and visualize their data.
+This rich environment has made the scientific Python ecosystem popular for scientific
+research.
 
@@
   \caption{\textbf{Scientific Python Ecosystem.}
+   This figure highlights some of the essential tools and libraries in the vast
+   scientific Python ecosystem.  Most of the ecosystem builds on NumPy to
+   provide a rich and powerful collection of array-aware functionality.
   }
@@
 \end{figure}
-
-The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) is composed of many different libraries.
-SciPy provides fundamental algorithms for scientific computing on arrays
-including many mathematical, scientific, and engineering routines.
-SciPy and NumPy are closely related and were developed together often
-by the same people.
-Matplotlib generates publication-ready figures and visualizations.
-Employing these array-aware libraries are widely used
-technique-specific libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11}
-and several domain-specific projects and
-libraries \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
-This has made the scientific Python ecosystem popular for scientific
-research.
-
-%These libraries can roughly be organized into three layers:
-%core libraries, technique-specific libraries, and domain-specific libraries.
-
-%SciPy adds various mathematical, scientific, and engineering routines;
-%Matplotlib generates publication-ready figures and visualizations,
-%Pandas provides indexed arrays, known as dataframes; and
-%xarray implements arrays with labeled axes.
-
-% we should have an example of how these tools may all be used in an
-% example analysis.  maybe one of the ones mentioned in the abstract?
 
@@
 functionality such as edge filters and Hough transforms.
+SciPy provides fundamental algorithms for scientific computing on arrays
+including many mathematical, scientific, and engineering routines.
+SciPy and NumPy are closely related and were developed together often
+by the same people.
+scikit-image is a popular technique-specific library,
+which vastly extends the basic image processing tools provided by SciPy.
 Other components of Scipy are used throughout the library, such as the
 \code{scipy.optimize} package for handling general optimization tasks.
-NetworkX \cite{SciPyProceedings_11} is used for image comparisons.
-Astropy \cite{astropy:2013, astropy:2018} is used in quite a few places,
+NetworkX \cite{SciPyProceedings_11} (the standard package for complex
+network analysis) is used for image comparisons.
+Astropy \cite{astropy:2013, astropy:2018} (a community-developed astronomy
+libarary) is used in quite a few places,
 primarily for I/O and time/coordinate transformations.
-Matplotlib is used for visualizing data throughout the analysis pipeline,
+Matplotlib (the core library for generating publication-ready figures
+and visualizations) is used for visualizing data throughout the analysis pipeline,
 including the generation of the final image of the black hole.
@@
 %https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
+
+% pandas, xarray, ...
 

### 2020-01-26 body.tex
Fix typo

@@
 Scipy is used to provide tools used in general image processing such as 
-filtering and image alignment, while scikit-image provided higher-level
+filtering and image alignment, while scikit-image provides higher-level
 functionality such as edge filters and Hough transforms.

### 2020-01-26 body.tex
Fix typo

@@
 It is common for NumPy users to work interactively with their data,
-discovering through experimentation bits of code that they stich together into
+discovering through experimentation bits of code that they stitch together into
 imperative or functional programs.

### 2020-01-26 body.tex
Modified pycbc and eht-imaging text - more explicit.
I've modified the text related to a more detailed look at the
pycbc and eht-imaging libraries to be a bit more explicit.

I also added the neutron star merger example to the end of the
pycbc discussion as it is another incredible scientific
discovery supported by GWOs and pycbc.

@@
 NetworkX \cite{SciPyProceedings_11} (the standard package for complex
-network analysis) is used for image comparisons.
+network analysis) is used for analysis of consistency among image
+comparisons.
 Astropy \cite{astropy:2013, astropy:2018} (a community-developed astronomy
-libarary) is used in quite a few places,
-primarily for I/O and time/coordinate transformations.
+library) is used throughout the library,
+primarily for I/O from standard astronomical file formats and time/coordinate
+transformations.
 Matplotlib (the core library for generating publication-ready figures
@@
 % N.B. - most of this comes from introspecting the Image class defined in
-% ehtim/image.py.
+% ehtim/image.py and grepping the dependencies in requirements.txt
 
@@
 and Virgo are provided by the \code{pycbc} package, which was used in the first
-detection of gravitationaly waves \cite{abbott2016observation} and in 
+detection of gravitational waves \cite{abbott2016observation} and in 
 on-going analysis of data from LIGO and Virgo.
@@
 including raw data from the instrumentation and the time-frequency 
-visualization of the ``chirp'' from the binary black hole merger.
+visualization of the ``chirps'' that result from mergers of compact binaries
+such as black holes\cite{abbott2016observation} and
+neutron stars\cite{abbott2017multi}.
 

### 2020-01-26 summary.tex
Fix bibtex

@@
 It plays an essential role in research analysis pipelines in fields as
-diverse as physics, biology, astronomy, neuroscience, material science,
-engineering, and chemistry.
+diverse as physics, chemistry, biology, astronomy, geoscience, psychology,
+material science, and engineering.
 For example, in astronomy, NumPy was fundamental to discovering gravitational

### 2020-01-26 body.tex
Add Stefan's comments (from PR #48)

@@
 % work by Peter Entschev]".
+% Below an attempt to start on this:
+Thanks to these developments users of Dask and NumPy arrays are now capable
+of scaling their computation to the GPU using Dask's CuPy integration with
+only minimal code changes.
+% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
+% The video cited embedded there could be a citation:i
+% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
+However, general adoption has not yet been achieved mainly for compatibility
+concerns.
+% explain a bit what "compatibility concerns" means ...
+Thus, we are actively addressing these concerns by revising and extending
+these protocols with the aim of wide interoperability that spans the whole
+scientific Python ecosystem up to and including domain libraries.
 
@@
 support these and other applications.
+The widespread adoption of rich, user defined, data types can allow more
+concise, feature rich scientific computing.
+By making these user implemented data types a fundamental design of NumPy
+these tools will enabled and interact naturally with different libraries.
+% need to add a bit of clarity on what "enable and interact naturally with different libraries" means.
+%
+%  1. I expect multiple user dtypes can interact (naturally); 2. libraries such
+%  as sklearn (or dask) might "just work" with e.g. physical Units (and
+%  Categoricals, which they probably really would like to use e.g. for labels);
+%  3. Things being inside a dtype rather than subclass is a better user
+%  experience.
 

### 2020-01-26 body.tex
Fix typos and spelling

@@
 reconstruction.
-Scipy is used to provide tools used in general image processing such as 
+SciPy is used to provide tools used in general image processing such as 
 filtering and image alignment, while scikit-image provides higher-level
@@
 which vastly extends the basic image processing tools provided by SciPy.
-Other components of Scipy are used throughout the library, such as the
+Other components of SciPy are used throughout the library, such as the
 \code{scipy.optimize} package for handling general optimization tasks.
@@
 transformations.
-Matplotlib (the core library for generating publication-ready figures
-and visualizations) is used for visualizing data throughout the analysis pipeline,
+Matplotlib (the core library for generating publication-ready figures) is
+used for visualizing data throughout the analysis pipeline,
 including the generation of the final image of the black hole.
@@
 such an endeavor, within a welcoming community of like-minded
-indivuals, held a powerful attraction for many early contributors.
+individuals, held a powerful attraction for many early contributors.
 

### 2020-01-26 summary.tex
Fix typos and spelling

--- 2020-01-16 summary.tex
@@
 It plays an essential role in research analysis pipelines in fields as
-diverse as physics, biology, astronomy, neuroscience, material science,
-engineering, and chemistry.
+diverse as physics, chemistry, astronomy, geoscience, biology, psychology,
+material science, and engineering.
 For example, in astronomy, NumPy was fundamental to discovering gravitational

### 2020-01-26 body.tex
Improve ecosystem figure

@@
 imperative or functional programs.
+SciPy and Matplotlib are tightly integrated with NumPy to expand the suite of
+standard algorithms and provide basic plotting capabilities.
+SciPy provides fundamental algorithms for scientific computing on arrays
+including many mathematical, scientific, and engineering routines.
+Matplotlib generates publication-ready figures and visualizations.
 The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
-this solid foundation with core (e.g., SciPy and Matplotlib) libraries,
-widely used technique-specific
-libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
+this solid foundation providing many, widely used technique-specific
+libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11}
 and several domain-specific projects and libraries
@@
 functionality such as edge filters and Hough transforms.
-SciPy provides fundamental algorithms for scientific computing on arrays
-including many mathematical, scientific, and engineering routines.
-SciPy and NumPy are closely related and were developed together often
-by the same people.
 scikit-image is a popular technique-specific library,
@@
 Other components of SciPy are used throughout the library, such as the
-\code{scipy.optimize} package for handling general optimization tasks.
+\code{scipy.optimize} subpackage for handling general optimization tasks.
 NetworkX \cite{SciPyProceedings_11} (the standard package for complex
@@
 library) is used throughout the library,
-primarily for I/O from standard astronomical file formats and time/coordinate
+primarily to handle standard astronomical file formats and time/coordinate
 transformations.
-Matplotlib (the core library for generating publication-ready figures) is
+Matplotlib is
 used for visualizing data throughout the analysis pipeline,
@@
 The time-series data from the interferometers are stored in NumPy arrays,
-while the \code{scipy.signal} is used to construct and apply filters to the
+while the \code{scipy.signal} subpackage is used to construct and apply filters to the
 data.

### 2020-01-27 body.tex
Use vectorize and minor edits

@@
 subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$
-for elementwise operations and $@$ for matrix multiplication), as well as {\em array-aware functions};
+for vectorized operations and $@$ for matrix multiplication), as well as {\em array-aware functions};
 together, these provide an easily readable, expressive, high-level API for
@@
 
+\fixme{Up to here we were talking about arrays, now we introduce array programming.
+Need to better explain vectorization in array programming.
+This is perhaps the essential thing about it so we should make it slightly
+more explicit.  It is just shorter to type, but it is easier to think
+about. }
 To complement the array syntax, NumPy includes functions that perform
-elementwise calculations on arrays, including arithmetic, statistics, and
+vectorized calculations on arrays, including arithmetic, statistics, and
 trigonometry.
@@
 
-When performing an elementwise operation (such as addition) on two arrays, one
+When performing a vectorized operation (such as addition) on two arrays, one
 expects that these arrays should have the same shape.
+% might expect if you are coming from other languages
+% wouldn't expect if you were doing math
+% scalar vector multiplication, scalar matrix multiplication, etc.
 However, through \emph{broadcasting}, NumPy allows the dimensions to differ, while
@@
 
-% - review supporting the ecosystem (actual support element)
-%   - docstrings / testing / packaging
-%   - coordinating mechanism
-%     - deprecation policy
-
 The brevity and expressive power of NumPy's array programming syntax combined with 
@@
 or Jupyter \cite{Kluyver:2016aa} makes interactive data exploration fluid and simple.
-It is common for NumPy users to work interactively with their data,
-discovering through experimentation bits of code that they stitch together into
-imperative or functional programs.
 SciPy and Matplotlib are tightly integrated with NumPy to expand the suite of
 standard algorithms and provide basic plotting capabilities.
-SciPy provides fundamental algorithms for scientific computing on arrays
-including many mathematical, scientific, and engineering routines.
+SciPy provides many fundamental algorithms for scientific computing on arrays
+including mathematical, scientific, and engineering routines.
 Matplotlib generates publication-ready figures and visualizations.
+
+% say something about this already being a good foundation comparable to
+% other scientific computing languages.
+
 The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
-this solid foundation providing many, widely used technique-specific
+this foundation providing many, widely used technique-specific
 libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11}
@@
 manipulate, and visualize their data.
-This rich environment has made the scientific Python ecosystem popular for scientific
+Working interactively with their data, users discover through experimentation
+bits of code that they stitch together into imperative or functional programs.
+This rich and productive environment has made the scientific Python ecosystem popular for scientific
 research.
@@
   \centering
-  \includegraphics[width=.48\textwidth]{static/ecosystem}
+  \includegraphics[width=.45\textwidth]{static/ecosystem}
   \caption{\textbf{Scientific Python Ecosystem.}
@@
 SciPy is used to provide tools used in general image processing such as 
-filtering and image alignment, while scikit-image provides higher-level
-functionality such as edge filters and Hough transforms.
-scikit-image is a popular technique-specific library,
-which vastly extends the basic image processing tools provided by SciPy.
+filtering and image alignment, while scikit-image---a popular image processing
+library that extends SciPy---provides higher-level functionality such as
+edge filters and Hough transforms.
 Other components of SciPy are used throughout the library, such as the
 \code{scipy.optimize} subpackage for handling general optimization tasks.
-NetworkX \cite{SciPyProceedings_11} (the standard package for complex
-network analysis) is used for analysis of consistency among image
+NetworkX \cite{SciPyProceedings_11}---a popular package for complex
+network analysis---is used for analysis of consistency among image
 comparisons.
-Astropy \cite{astropy:2013, astropy:2018} (a community-developed astronomy
-library) is used throughout the library,
+The community developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
+is used throughout \code{eht-imaging},
 primarily to handle standard astronomical file formats and time/coordinate
@@
 
-Tools for analyzing data from gravitational wave observatories such as LIGO
-and Virgo are provided by the \code{pycbc} package, which was used in the first
+Tools for analyzing data from gravitational wave observatories are
+provided by the \code{pycbc} package, which was used in the first
 detection of gravitational waves \cite{abbott2016observation} and in 
 on-going analysis of data from LIGO and Virgo.
-\code{pycbc} make extensive use of the scientific Python ecosystem.
+\code{pycbc} also make extensive use of the scientific Python ecosystem.
 The time-series data from the interferometers are stored in NumPy arrays,
@@
 
+% need better transition
+% similar tools exist in physics, chemistry, biology, earth sciences,
+% engineering
+% this means there is an extraordinary need for stability and reliability
+% this conflicts with competing need to innovate
+% resolving this tension requires balancing trade-offs
+% this lead to a proliferation of arrays
+
 The rich ecosystem of tools building on NumPy and the larger community
@@
 to evolve and change.
+
+% - review supporting the ecosystem (actual support element)
+%   - docstrings / testing / packaging
+%   - coordinating mechanism
+%     - deprecation policy
 
@@
 
-To facilitate \emph{interoperability}, the NumPy team has designed
+To facilitate \emph{interoperability}, the NumPy provides
 ``protocols'' (or contracts of operation), that allow for these arrays to be
@@
 heralded the current era of wide-spread use of Python for scientific computing.
-The reasons for its success are varied and complex, but include the fact that
+The reasons for its popularity are varied and complex, but include the fact that
 NumPy (a) is built on top of Python,

### 2020-01-27 summary.tex
Use vectorize and minor edits

@@
 higher-dimensional arrays \cite{iverson1980notation}.
+Vectorization---operating on whole arrays rather than their individual
+elements---is essential to array programming.
 % - a brief account of the background and rationale of the work;
@@
 \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-It has been actively developed for over two decades and has more than 850
-contributors, nearly 300,000 dependent repositories on GitHub, and millions of
+Developed for over two decades by more than 850 contributors, 
+it has nearly 300,000 dependent repositories on GitHub and millions of
 downloads per year.
@@
 waves\cite{pycbc}, the first imaging of a black hole\cite{eht-imaging}, and
-continues underpinning the future exploration of our
+continues underpinning the exploration of our
 universe\cite{jenness2018lsst}.

### 2020-01-27 body.tex
Revise ecosystem

@@
 
-\fixme{Up to here we were talking about arrays, now we introduce array programming.
-Need to better explain vectorization in array programming.
-This is perhaps the essential thing about it so we should make it slightly
-more explicit.  It is just shorter to type, but it is easier to think
-about. }
 To complement the array syntax, NumPy includes functions that perform
-vectorized calculations on arrays, including arithmetic, statistics, and
+\emph{vectorized} calculations on arrays, including arithmetic, statistics, and
 trigonometry.
-They aim to loop over array elements near-optimally, taking into consideration,
-e.g., strides, in order to best utilize the computer's fast cache memory.
-Vectorized operations that would take many tens of lines to express in
+Vectorization---operating on whole arrays rather than their individual
+elements---is essential to array programming.
+This means that operations that would take many tens of lines to express in
 languages such as C can often be implemented as a single, clear Python
 expression.
+This results in concise code and frees users to focus on the details of
+their analysis, while NumPy handles looping over array elements near-optimally,
+taking into consideration, for example, strides, in order to best utilize the
+computer's fast cache memory.
 
-When performing a vectorized operation (such as addition) on two arrays, one
-expects that these arrays should have the same shape.
-% might expect if you are coming from other languages
-% wouldn't expect if you were doing math
-% scalar vector multiplication, scalar matrix multiplication, etc.
-However, through \emph{broadcasting}, NumPy allows the dimensions to differ, while
+When performing a vectorized operation (such as addition) on two arrays with
+the same shape, it is clear what should happen.
+Through \emph{broadcasting}, NumPy allows the dimensions to differ, while
 still producing results that appeal to intuition.
@@
 (Fig.~\ref{fig:array-concepts}d).
-Broadcasting is also applied whenever an array is indexed using arrays of
+Broadcasting is also applied when an array is indexed using arrays of
 indices (Fig.~\ref{fig:array-concepts}c).
@@
 
-The brevity and expressive power of NumPy's array programming syntax combined with 
-an advanced interactive environment like IPython \cite{perez2007ipython}
-or Jupyter \cite{Kluyver:2016aa} makes interactive data exploration fluid and simple.
-SciPy and Matplotlib are tightly integrated with NumPy to expand the suite of
-standard algorithms and provide basic plotting capabilities.
-SciPy provides many fundamental algorithms for scientific computing on arrays
+SciPy and Matplotlib are tightly integrated with NumPy.
+% to expand the suite of standard algorithms and
+% provide basic plotting capabilities.
+SciPy provides fundamental algorithms for scientific computing
 including mathematical, scientific, and engineering routines.
 Matplotlib generates publication-ready figures and visualizations.
-
-% say something about this already being a good foundation comparable to
-% other scientific computing languages.
-
+The combination of NumPy, SciPy, and Matplotlib with 
+an advanced interactive environment like IPython \cite{perez2007ipython}
+or Jupyter \cite{Kluyver:2016aa}
+provides a complete foundation for array-based scientific computing
+and makes interactive data exploration fluid and simple.
 The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
@@
 \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
-Exposing these simple array programming primitives, as well as the surrounding ecosystem of
-tools, in Python---an interpreted, general purpose language---creates an ideal environment for
-interactive, exploratory data analysis where users may iteratively inspect,
+NumPy plays a coordinating role for the ecosystem of array-aware
+libraries including setting documentation standards,
+providing array testing infrastracture,
+and coordinating minimum version depedencies.
+
+Exposing this foundation for array programming, as well as the surrounding ecosystem of
+tools, in IPython or Jupyter, creates an ideal environment for
+interactive, exploratory data analysis where users iteratively inspect,
 manipulate, and visualize their data.
 Working interactively with their data, users discover through experimentation
-bits of code that they stitch together into imperative or functional programs.
+bits of code that they stitch together into imperative or functional programs
+or narrative computational notebooks.
 This rich and productive environment has made the scientific Python ecosystem popular for scientific
 research.
+As a result, important scientific software packages build on the ecosystem
+further exanding it for specific research projects.
 
@@
 
-% need better transition
-% similar tools exist in physics, chemistry, biology, earth sciences,
-% engineering
-% this means there is an extraordinary need for stability and reliability
-% this conflicts with competing need to innovate
-% resolving this tension requires balancing trade-offs
-% this lead to a proliferation of arrays
-
+Similar examples can be found in physics, earth sciences, biology,
+and more.
+These researchers depend on NumPy being correct, reliable,
+and repeatable.
 The rich ecosystem of tools building on NumPy and the larger community
@@
 responsibility to maintain considerable stability.
+This conflicts with the need to innovate;
+resolving it requires balancing trade-offs.
+% our focus is ...
 This need creates interesting challenges for NumPy development as the
@@
 to evolve and change.
-
-% - review supporting the ecosystem (actual support element)
-%   - docstrings / testing / packaging
-%   - coordinating mechanism
-%     - deprecation policy
+% this leads to a proliferation of arrays
 
@@
 
-%\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
-%https://parquet.apache.org/,
-%https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}
-
 % pandas, xarray, ...
+% distributed arrays, delayed execution arrays
+% GPU arrays, tensor computation libraries
+% sparse arrays
+% foreign language arrays
 
@@
 few non-zero values, and store only those in memory.
-%distributed arrays, GPU arrays, delayed execution arrays, sparse arrays,
-%foreign language arrays, and more.
 
@@
 Primarily, there
-exist two types of Python array objects: (1) arrays that mimic NumPy arrays but are
-fundamentally different and (2) subclasses of NumPy arrays.
+exist two types of Python array objects: (a) arrays that mimic NumPy arrays but are
+fundamentally different and (b) subclasses of NumPy arrays.
 
@@
 This culture of using best practices for producing reliable scientific software
-has been eagerly adopted by the ecosystem of libraries that build on NumPy.
+has been adopted by the ecosystem of libraries that build on NumPy.
 For example, in a recent award given by the Royal Astronomical Society to

### 2020-01-27 body.tex
Edit beginning of ecosystem section

@@
 
-SciPy and Matplotlib are tightly integrated with NumPy.
+SciPy and Matplotlib are tightly coupled with NumPy---in terms of
+history, development, and usage.
 % to expand the suite of standard algorithms and
@@
 Matplotlib generates publication-ready figures and visualizations.
-The combination of NumPy, SciPy, and Matplotlib with 
+The combination of NumPy, SciPy, and Matplotlib together with
 an advanced interactive environment like IPython \cite{perez2007ipython}
 or Jupyter \cite{Kluyver:2016aa}
-provides a complete foundation for array-based scientific computing
-and makes interactive data exploration fluid and simple.
+provides a complete foundation for array-based scientific computing.
 The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
-this foundation providing many, widely used technique-specific
-libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11}
-and several domain-specific projects and libraries
+this foundation to provide several, widely used \emph{technique-specific}
+libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
+that in turn underlay numerous \emph{domain-specific} projects
 \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
-NumPy plays a coordinating role for the ecosystem of array-aware
-libraries including setting documentation standards,
-providing array testing infrastracture,
-and coordinating minimum version depedencies.
+NumPy, as the focal point for the ecosystem of array-aware libraries,
+sets documentation standards, provides array testing infrastracture,
+and defines minimum version depedencies.
 
-Exposing this foundation for array programming, as well as the surrounding ecosystem of
-tools, in IPython or Jupyter, creates an ideal environment for
-interactive, exploratory data analysis where users iteratively inspect,
-manipulate, and visualize their data.
-Working interactively with their data, users discover through experimentation
-bits of code that they stitch together into imperative or functional programs
-or narrative computational notebooks.
-This rich and productive environment has made the scientific Python ecosystem popular for scientific
-research.
-As a result, important scientific software packages build on the ecosystem
-further exanding it for specific research projects.
+The interactive environment created by this array programming
+foundation---along with the surrounding ecosystem of tools---inside of
+IPython or Jupyter, is ideally suited to exploratory data analysis.
+Users fluidly inspect, manipulate, and visualize their data, and
+rapidly iterate to refine programming statements. These statements are
+then stitched together into imperative or functional programs, or
+narrative computational notebooks.  This rich and productive
+environment has made the scientific Python ecosystem hugely popular
+for scientific research.
+
+Despite its tight integration, the ecosystem was conceived to have
+clear functional separation, with well defined interfaces.  This makes
+possible the design of large, complex scientific libraries, that
+add specialized functionality for solving specific problems.
+
+%As a result, important scientific software
+%packages build on the ecosystem, and further expand it for specific
+%research projects.
 
@@
 and more.
-These researchers depend on NumPy being correct, reliable,
+Researchers in these fields depend on NumPy being correct, reliable,
 and repeatable.
@@
 responsibility to maintain considerable stability.
-This conflicts with the need to innovate;
+At times, this conflicts with the need to innovate;
 resolving it requires balancing trade-offs.
 % our focus is ...
+% NumPy purpose engages with this problem by ...
+% [this also defines the scope of NumPy]
 This need creates interesting challenges for NumPy development as the
@@
 % pandas, xarray, ...
+% numba?
 % distributed arrays, delayed execution arrays

### 2020-01-27 body.tex
Small fix

@@
 NumPy combines the expressive power of \emph{array programming}, 
-the performance of C, and the Python programming language in a mature,
+the performance of C, and
+the readability, usability, and versatility of Python in a mature,
 well-tested, well-documented, and community developed library.

### 2020-01-27 body.tex
Comment out 2nd example

@@
 
-Tools for analyzing data from gravitational wave observatories are
-provided by the \code{pycbc} package, which was used in the first
-detection of gravitational waves \cite{abbott2016observation} and in 
-on-going analysis of data from LIGO and Virgo.
-\code{pycbc} also make extensive use of the scientific Python ecosystem.
-The time-series data from the interferometers are stored in NumPy arrays,
-while the \code{scipy.signal} subpackage is used to construct and apply filters to the
-data.
-Matplotlib is used to visualize data at all points along the analysis chain
-including raw data from the instrumentation and the time-frequency 
-visualization of the ``chirps'' that result from mergers of compact binaries
-such as black holes\cite{abbott2016observation} and
-neutron stars\cite{abbott2017multi}.
+%Tools for analyzing data from gravitational wave observatories are
+%provided by the \code{pycbc} package, which was used in the first
+%detection of gravitational waves \cite{abbott2016observation} and in 
+%on-going analysis of data from LIGO and Virgo.
+%\code{pycbc} also make extensive use of the scientific Python ecosystem.
+%The time-series data from the interferometers are stored in NumPy arrays,
+%while the \code{scipy.signal} subpackage is used to construct and apply filters to the
+%data.
+%Matplotlib is used to visualize data at all points along the analysis chain
+%including raw data from the instrumentation and the time-frequency 
+%visualization of the ``chirps'' that result from mergers of compact binaries
+%such as black holes\cite{abbott2016observation} and
+%neutron stars\cite{abbott2017multi}.
 

### 2020-01-28 body.tex
Minor edits

@@
 SciPy and Matplotlib are tightly coupled with NumPy---in terms of
-history, development, and usage.
+history, development, and use.
 % to expand the suite of standard algorithms and
@@
 that in turn underlay numerous \emph{domain-specific} projects
-\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
+\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
 NumPy, as the focal point for the ecosystem of array-aware libraries,
-sets documentation standards, provides array testing infrastracture,
-and defines minimum version depedencies.
+sets documentation standards, provides array testing infrastructure,
+and defines minimum version dependencies.
 
@@
 
-Similar examples can be found in physics, earth sciences, biology,
+Similar examples can be found in physics, earth science, biology,
 and more.

### 2020-01-28 body.tex
Minor update to Scientific Python ecosystem.
Most important change is to talk about numpy.distutils and f2py
rather than minimum versions; the latter really is a small detail.

@@
 NumPy, as the focal point for the ecosystem of array-aware libraries,
-sets documentation standards, provides array testing infrastructure,
-and defines minimum version dependencies.
+sets documentation standards, and provides array testing infrastructure
+and build support for integrating with Fortran code and using specialized
+compilers and computing platforms.
 
@@
 Other components of SciPy are used throughout the library, such as the
-\code{scipy.optimize} subpackage for handling general optimization tasks.
+\code{scipy.optimize} module for handling general optimization tasks.
 NetworkX \cite{SciPyProceedings_11}---a popular package for complex
@@
 comparisons.
-The community developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
-is used throughout \code{eht-imaging},
-primarily to handle standard astronomical file formats and time/coordinate
-transformations.
-Matplotlib is
-used for visualizing data throughout the analysis pipeline,
+The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
+is used throughout \code{eht-imaging}, primarily to handle standard
+astronomical file formats and time/coordinate transformations.
+Matplotlib is used for visualizing data throughout the analysis pipeline,
 including the generation of the final image of the black hole.
@@
 This need creates interesting challenges for NumPy development as the
-computational environment scientists are analyzing their data on continues
+computational environment scientists are analyzing their data in continues
 to evolve and change.

### 2020-01-28 body.tex
Update of proliferation section (still needs extending).

@@
 
-% pandas, xarray, ...
-% numba?
+% pandas, xarray, ... - leave out Pandas since it's not an array?
 % distributed arrays, delayed execution arrays
@@
 in the computational toolchain, has become ubiquitous.  Still, NumPy itself
-could never hope to satisfy every single specialized need of the community; it
-is limited in its ability to work with very large datasets, datasets split
-across multiple systems, and computations that require specialized hardware.
-Furthermore, pursuing solutions to these challenges is out of scope for NumPy
-development.
+could never hope to satisfy every single specialized need of the community;
+due to its in-memory data model, it is limited in its ability to work with very
+large datasets, datasets split across multiple systems, and computations that
+require graphics processing units (GPUs), tensor processing units (TPUs) or
+other specialized hardware for performance reasons.  Pursuing solutions to
+these challenges is out of scope for NumPy development.
 
 The community's efforts to fill the resultant gaps therefore led to a
-proliferation of arrays including distributed arrays---which are stored on
-multiple computers, GPU arrays---which utilize graphics processing units for
-fast computation, computation graphs---which postpone and combine calculations
-before executing them efficiently, and sparse arrays---which typically contain
-few non-zero values, and store only those in memory.
+proliferation of array implementations. Each deep learning framework created
+its own arrays; PyTorch, Tensorflow, Apache MXNet and JAX arrays all have the
+capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
+execution mode to allow for performance optimizations.  SciPy and Sparse both
+provide sparse arrays---which typically contain few non-zero values and store
+only those in memory for efficiency.
+In addition there are projects that build on top of NumPy arrays as a data
+container and \textit{extend} its capabilities.  Distributed arrays are
+made possible that way by Dask, and labeled arrays---referring to dimensions of
+an array by name rather than by index for clarity, compare \code{x[:, 1]} vs.
+\code{x.loc[:, 'time']}---by Xarray.
 
-This includes popular tensor computation libraries such as
-CuPy, JAX, and Apache MXNet.
-PyTorch and Tensorflow provide tensor APIs with NumPy-inspired semantics.
-% \footnote{\url{https://cupy.chainer.org/}}
+Importantly, some projects chose to follow the NumPy API exactly---with Dask,
+CuPy, Sparse and JAX the most prominent examples---while others have
+NumPy-inspired semantics but chose to deviate in places.
+
+% \footnote{\url{https://cupy.chainer.org}}
 % \footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}
-% \footnote{\url{https://numpy.mxnet.io/}}
+% \footnote{\url{https://mxnet.incubator.apache.org}}
+% \footnote{\url{https://dask.org}}
+% JAX has a citation, see https://github.com/google/jax#citing-jax
+% Xarray has a citation, see http://xarray.pydata.org/en/stable/faq.html#how-should-i-cite-xarray 
 % \footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
 % \footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
+
+% proliferation of arrays including distributed arrays---which are stored on
+% multiple computers, GPU arrays---which utilize graphics processing units for
+% fast computation, computation graphs---which postpone and combine calculations
+% before executing them efficiently, and sparse arrays---which typically contain
+% few non-zero values, and store only those in memory.
+
+%This includes popular tensor computation libraries such as
+%CuPy, JAX, and Apache MXNet.
+%PyTorch and Tensorflow provide tensor APIs with NumPy-inspired semantics.
 
@@
 
+% Ralf: FIXME I don't think this is actually important - should work exactly the same
+% for users whether a NumPy-compatible API is a subclass or a duck array
 Primarily, there
@@
 provide arrays and a NumPy-like API for manipulating them.
-Ideally, operating on these arrays using NumPy would simply work, so that end
-users could write code once, and would then benefit from switching between
-NumPy arrays, GPU arrays, distributed arrays, and so forth, as appropriate.
+Ideally, operating on these arrays using NumPy functions or semantics would
+simply work, so that end users could write code once, and would then benefit
+from switching between NumPy arrays, GPU arrays, distributed arrays, and so
+forth, as appropriate.
 
@@
 % work by Peter Entschev]".
+% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
 % Below an attempt to start on this:

### 2020-01-29 body.tex
Minor edits

@@
 including the generation of the final image of the black hole.
-% N.B. - most of this comes from introspecting the Image class defined in
-% ehtim/image.py and grepping the dependencies in requirements.txt
-
-%Tools for analyzing data from gravitational wave observatories are
-%provided by the \code{pycbc} package, which was used in the first
-%detection of gravitational waves \cite{abbott2016observation} and in 
-%on-going analysis of data from LIGO and Virgo.
-%\code{pycbc} also make extensive use of the scientific Python ecosystem.
-%The time-series data from the interferometers are stored in NumPy arrays,
-%while the \code{scipy.signal} subpackage is used to construct and apply filters to the
-%data.
-%Matplotlib is used to visualize data at all points along the analysis chain
-%including raw data from the instrumentation and the time-frequency 
-%visualization of the ``chirps'' that result from mergers of compact binaries
-%such as black holes\cite{abbott2016observation} and
-%neutron stars\cite{abbott2017multi}.
 
@@
 resolving it requires balancing trade-offs.
-% our focus is ...
-% NumPy purpose engages with this problem by ...
-% [this also defines the scope of NumPy]
-This need creates interesting challenges for NumPy development as the
-computational environment scientists are analyzing their data in continues
-to evolve and change.
-% this leads to a proliferation of arrays
+
+% FIXME: this should be expanded
+Over the last few years there has been a massive increase in the use
+of scientific Python due in part to the rapid growth of data science,
+machine learning, and artificial intelligence.  
+As a result, NumPy and its API has become ubiquitous.
 
@@
 
-Recent years have seen a massive explosion in data science, machine learning,
-and artificial intelligence.  NumPy and its API, underlying almost every tool
-in the computational toolchain, has become ubiquitous.  Still, NumPy itself
-could never hope to satisfy every single specialized need of the community;
+% Ralf: we should describe our role at the foundation of the stack
+% i was thinking something along these lines ():
+NumPy provides in-memory, multidimensional, homogeneously typed
+(single pointer + strided) arrays on CPUs.
+It supports many types of CPUs (e.g., POWER and ARM) with similar
+performance as native C code.
+While NumPy suits most user's needs, it cannot support every specialized need of the community;
 due to its in-memory data model, it is limited in its ability to work with very
@@
 require graphics processing units (GPUs), tensor processing units (TPUs) or
-other specialized hardware for performance reasons.  Pursuing solutions to
-these challenges is out of scope for NumPy development.
+other specialized hardware for performance reasons.
 
-The community's efforts to fill the resultant gaps therefore led to a
+% Ralf: should mention that there are typically utilities to convert to and from numpy arrays
+% and why (presumably to get access to ecosystem)
+The community's efforts to fill the resultant gaps led to a
 proliferation of array implementations. Each deep learning framework created
-its own arrays; PyTorch, Tensorflow, Apache MXNet and JAX arrays all have the
+its own arrays; PyTorch \cite{NEURIPS2019_9015},
+Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
+and JAX \cite{jax2018github} arrays all have the
 capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
@@
 made possible that way by Dask, and labeled arrays---referring to dimensions of
-an array by name rather than by index for clarity, compare \code{x[:, 1]} vs.
-\code{x.loc[:, 'time']}---by Xarray.
+an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
+\code{x.loc[:,~'time']}---by Xarray \cite{hoyer2017xarray}.
 
@@
 NumPy-inspired semantics but chose to deviate in places.
+% Ralf: may want to add something like:
+% when they deviate, they sometimes fix it, e.g.,
+% https://github.com/pytorch/pytorch/pull/1983
+% https://github.com/pytorch/pytorch/issues/2228
+% many have for numpy users
+% https://github.com/torch/torch7/wiki/Torch-for-Numpy-users
 
 % \footnote{\url{https://cupy.chainer.org}}
-% \footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}
-% \footnote{\url{https://mxnet.incubator.apache.org}}
-% \footnote{\url{https://dask.org}}
-% JAX has a citation, see https://github.com/google/jax#citing-jax
-% Xarray has a citation, see http://xarray.pydata.org/en/stable/faq.html#how-should-i-cite-xarray 
-% \footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
-% \footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
 
-% proliferation of arrays including distributed arrays---which are stored on
-% multiple computers, GPU arrays---which utilize graphics processing units for
-% fast computation, computation graphs---which postpone and combine calculations
-% before executing them efficiently, and sparse arrays---which typically contain
-% few non-zero values, and store only those in memory.
+% computation graphs---which postpone and combine calculations
+% before executing them efficiently,
 
-%This includes popular tensor computation libraries such as
-%CuPy, JAX, and Apache MXNet.
-%PyTorch and Tensorflow provide tensor APIs with NumPy-inspired semantics.
+% Ralf: I am not sure where this goes, but it would be good to have something
+% about dtype limitation in this section and i also think there should
+% be some discussion of missing values somewhere
+% (it is mentioned below, but not discussed)
+While it supports a wide range of datatypes, it does not have good support
+for missing values. % may be worth saying a few sentences about missing values
+% if we want to discuss Pandas, perhaps we could do so here (i.e, missing values)
 
@@
 concert.
-
-% Ralf: FIXME I don't think this is actually important - should work exactly the same
-% for users whether a NumPy-compatible API is a subclass or a duck array
-Primarily, there
-exist two types of Python array objects: (a) arrays that mimic NumPy arrays but are
-fundamentally different and (b) subclasses of NumPy arrays.
 
@@
 
+% should we hint at some of the community interaction part
+% they have helped us develop the neps and e.g.,
+% https://github.com/pytorch/pytorch/issues/22402
 To facilitate \emph{interoperability}, the NumPy provides

### 2020-01-29 body.tex
A few more changes

@@
 and more.
-Researchers in these fields depend on NumPy being correct, reliable,
-and repeatable.
-The rich ecosystem of tools building on NumPy and the larger community
-of scientists and programmers using those tools mean that NumPy has a
-responsibility to maintain considerable stability.
-At times, this conflicts with the need to innovate;
-resolving it requires balancing trade-offs.
-
 % FIXME: this should be expanded 
@@
 NumPy provides in-memory, multidimensional, homogeneously typed
-(single pointer + strided) arrays on CPUs.
+(i.e., single pointer and strided) arrays on CPUs.
 It supports many types of CPUs (e.g., POWER and ARM) with similar
 performance as native C code.
-While NumPy suits most user's needs, it cannot support every specialized need of the community;
-due to its in-memory data model, it is limited in its ability to work with very
-large datasets, datasets split across multiple systems, and computations that
-require graphics processing units (GPUs), tensor processing units (TPUs) or
-other specialized hardware for performance reasons.
+% a brief word about accelerators perhaps?
+% end with how this fits the bill for a huge number of users
+
+While NumPy suits most user's needs, it cannot support every specialized need
+of the community; due to its in-memory data model, it is limited in its ability
+to work with very large datasets, datasets split across multiple systems, and
+computations that require graphics processing units (GPUs), tensor processing
+units (TPUs) or other specialized hardware for performance reasons.
 
@@
 While it supports a wide range of datatypes, it does not have good support
-for missing values. % may be worth saying a few sentences about missing values
+for missing values or categorical data. % may be worth saying a few sentences about missing values
 % if we want to discuss Pandas, perhaps we could do so here (i.e, missing values)
+% maybe we need to rename the section, but this could anticipate
+% the dtypes discussion below
 
@@
 % https://github.com/pytorch/pytorch/issues/22402
-To facilitate \emph{interoperability}, the NumPy provides
+To facilitate \emph{interoperability}, NumPy provides
 ``protocols'' (or contracts of operation), that allow for these arrays to be
@@
 types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
-geometrical objects \cite{pygeos}, and missing numbers. % pandas: categoricals, IP addresses?
+geometrical objects \cite{pygeos}, and missing values. % pandas: categoricals, IP addresses?
 To improve \emph{extensibility}, we are currently overhauling the data type
@@
 
-Over time the role of the NumPy has evolved. It is no longer just the
-foundational array library underlying the scientific Python ecosystem,
-but it is also a standard API for tensor computation and a
-central coordinating mechanism between array types and technologies.
+Over time the role of the NumPy has evolved.
+It began as an attempt to add an array object to Python
+and became the foundation of a rich ecosystem of tools.
+Now a large amount of scientific work depends on NumPy
+being correct, fast, and stable.
+Yet, it is no longer just the foundational array library underlying the
+scientific Python ecosystem; it is also a standard API for tensor
+computation and a central coordinating mechanism between array types and
+technologies.

### 2020-01-29 body.tex
Clean up
Remove future dtype discussion from main text

@@
 
-% pandas, xarray, ... - leave out Pandas since it's not an array?
-% distributed arrays, delayed execution arrays
-% GPU arrays, tensor computation libraries
-% sparse arrays
-% foreign language arrays
-
-% Ralf: we should describe our role at the foundation of the stack
-% i was thinking something along these lines ():
 NumPy provides in-memory, multidimensional, homogeneously typed
@@
 
-% Ralf: should mention that there are typically utilities to convert to and from numpy arrays
+% maybe mention that there are typically utilities
+% to convert to and from numpy arrays
 % and why (presumably to get access to ecosystem)
@@
 NumPy-inspired semantics but chose to deviate in places.
-% Ralf: may want to add something like:
 % when they deviate, they sometimes fix it, e.g.,
@@
 
-% \footnote{\url{https://cupy.chainer.org}}
-
 % computation graphs---which postpone and combine calculations
@@
 
-% Ralf: I am not sure where this goes, but it would be good to have something
-% about dtype limitation in this section and i also think there should
-% be some discussion of missing values somewhere
-% (it is mentioned below, but not discussed)
-While it supports a wide range of datatypes, it does not have good support
-for missing values or categorical data. % may be worth saying a few sentences about missing values
-% if we want to discuss Pandas, perhaps we could do so here (i.e, missing values)
-% maybe we need to rename the section, but this could anticipate
-% the dtypes discussion below
-
-\section*{Interoperability and extensibility}
-
-NumPy supports array operations between external array objects by
-acting as a central coordination mechanism with a well-specified API.
-With the proliferation of array implementations discussed above, a
-widespread adoption of the NumPy API is valuable: it lowers the
+Ideally, operating on these arrays using NumPy functions or semantics would
+simply work, so that end users could write code once, and would then benefit
+from switching between NumPy arrays, GPU arrays, distributed arrays, and so
+forth, as appropriate.
+To support array operations between external array objects, NumPy
+acts as a central coordination mechanism with a well-specified API.
+The widespread adoption of the NumPy API is valuable: it lowers the
 barrier to entry for newcomers and provides the wider community with a
@@
 concert.
-
-There has been a steady increase in the number of external libraries that
-provide arrays and a NumPy-like API for manipulating them.
-Ideally, operating on these arrays using NumPy functions or semantics would
-simply work, so that end users could write code once, and would then benefit
-from switching between NumPy arrays, GPU arrays, distributed arrays, and so
-forth, as appropriate.
 
@@
 % https://numpy.org/neps/nep-0037-array-module.html
-
-Subclasses of NumPy often exist due to difficulty in constructing richer data
-types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
-geometrical objects \cite{pygeos}, and missing values. % pandas: categoricals, IP addresses?
-To improve \emph{extensibility}, we are currently overhauling the data type
-system to make it easier to implement data types both in Python and C, and to
-support these and other applications.
-The widespread adoption of rich, user defined, data types can allow more
-concise, feature rich scientific computing.
-By making these user implemented data types a fundamental design of NumPy
-these tools will enabled and interact naturally with different libraries.
-% need to add a bit of clarity on what "enable and interact naturally with different libraries" means.
-%
-%  1. I expect multiple user dtypes can interact (naturally); 2. libraries such
-%  as sklearn (or dask) might "just work" with e.g. physical Units (and
-%  Categoricals, which they probably really would like to use e.g. for labels);
-%  3. Things being inside a dtype rather than subclass is a better user
-%  experience.
 

### 2020-01-30 body.tex
Revise section title
Fix name
Edit ecosystem

@@
 The library has such widespread adoption that not only the array object but also its
-{\it Application Programming Interface} (API) has become ubiquitous as
+{\it Application Programming Interface} (API) has become widespread as
 a language for array programming.
@@
 narrative computational notebooks.  This rich and productive
-environment has made the scientific Python ecosystem hugely popular
+environment has made Python popular
 for scientific research.
-
-Despite its tight integration, the ecosystem was conceived to have
-clear functional separation, with well defined interfaces.  This makes
-possible the design of large, complex scientific libraries, that
-add specialized functionality for solving specific problems.
 
@@
   \caption{\textbf{Scientific Python Ecosystem.}
-   This figure highlights some of the essential tools and libraries in the vast
-   scientific Python ecosystem.  Most of the ecosystem builds on NumPy to
+   This figure highlights just a few of the essential tools and libraries in the vast
+   scientific Python ecosystem.  Most of the ecosystem builds on NumPy arrays to
    provide a rich and powerful collection of array-aware functionality.
@@
 
+Many research groups have designed large,
+complex scientific libraries, that add specialized functionality for
+solving problems in specific domains.
 For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
-collaboration relies on many components of the scientific Python ecosystem.
+collaboration for radio interferometry imaging, analysis, and simulation
+relies on many components of the scientific Python ecosystem.
 NumPy arrays are used to store and manipulate numerical data at every step
@@
 SciPy is used to provide tools used in general image processing such as 
-filtering and image alignment, while scikit-image---a popular image processing
-library that extends SciPy---provides higher-level functionality such as
+filtering and image alignment, while scikit-image, a popular image processing
+library that extends SciPy, provides higher-level functionality such as
 edge filters and Hough transforms.
@@
 \code{scipy.optimize} module for handling general optimization tasks.
-NetworkX \cite{SciPyProceedings_11}---a popular package for complex
-network analysis---is used for analysis of consistency among image
+NetworkX \cite{SciPyProceedings_11}, a popular package for complex
+network analysis, is used for analysis of consistency among image
 comparisons.
@@
 and more.
-% FIXME: this should be expanded 
-Over the last few years there has been a massive increase in the use
-of scientific Python due in part to the rapid growth of data science,
-machine learning, and artificial intelligence.  
-As a result, NumPy and its API has become ubiquitous.
+In many fields, the scientific Python ecosystem is the primary
+software environment.
+The recent rapid growth of data science, machine learning, and
+artificial intelligence has dramatically increased scientific
+Python use.
+NumPy and its ecosystem is now taught in
+numerous university courses, bootcamps, and summer schools.
+There are multiple community conferences and workshops.
+NumPy and its API has become ubiquitous.
 
-\section*{Proliferation of arrays}
+\section*{Array proliferation and interoperability}
 

### 2020-01-30 body.tex
Minor edits

@@
 
-Other array-aware functions include creating, reshaping, concatenating, and padding
-arrays; searching, sorting and counting data; and reading and writing files.
-NumPy provides extensive support for generating pseudorandom numbers and
-includes an assortment of probability distributions.
-It also performs accelerated linear algebra, utilizing one of several backends
-such as OpenBLAS \cite{wang2013augem,xianyi2012model} and Intel MKL optimized for the CPUs at hand.
+NumPy also includes array-aware functions for creating, reshaping, concatenating, and padding
+arrays; searching, sorting, and counting data; and reading and writing files.
+It provides extensive support for generating pseudorandom numbers,
+includes an assortment of probability distributions, and
+performs accelerated linear algebra, utilizing one of several backends
+such as OpenBLAS \cite{wang2013augem,xianyi2012model} or Intel MKL optimized
+for the CPUs at hand.
 
@@
 history, development, and use.
-% to expand the suite of standard algorithms and
-% provide basic plotting capabilities.
 SciPy provides fundamental algorithms for scientific computing
@@
 
-%As a result, important scientific software
-%packages build on the ecosystem, and further expand it for specific
-%research projects.
-
 \begin{figure}
@@
 Many research groups have designed large,
-complex scientific libraries, that add specialized functionality for
+complex scientific libraries, which add specialized functionality for
 solving problems in specific domains.
@@
 dictate our own release policies and development patterns.
-%\fixme{we've gotten a number of features: slicing syntax, implicit tuples,
-%complex numbers, buffer protocol, matrix multiplication operator}
-%\fixme{but we benefit from separate governance}
 
@@
 Yet, it is no longer just the foundational array library underlying the
-scientific Python ecosystem; it is also a standard API for tensor
+scientific Python ecosystem; it is also the standard API for tensor
 computation and a central coordinating mechanism between array types and
-technologies.
+technologies in Python.
+NumPy continues to evolve.
+For instance, the NumPy data type system is currently being redesigned to make it
+easier to implement data types both in Python and C, and to better handle
+richer data types including physical units \cite{astropy,Goldbaum2018,pint},
+geometrical objects \cite{pygeos}, and missing values.

### 2020-01-30 body.tex
Thoughts about expanding the prolif section based on Jarrods ideas

@@
 (i.e., single pointer and strided) arrays on CPUs.
-It supports many types of CPUs (e.g., POWER and ARM) with similar
-performance as native C code.
+It runs on machines ranging from embedded devices up to the worlds
+largest supercomputers [citation needed] and provides a performance
+which often approaches compiled languages.
+When algorithms do not fit the array programming approach,
+or require a faster implementation,
+a number of tools (e.g. Cython, Numba, Pythran, ...)
+are available to transparently accelerate crucial parts.
+These tools allow for a high level array orientated code base without sacrificing
+the speed expected from a compiled language.
+In many cases fast algorithms are implemented by libraries in the scientific
+python ecosystem so that most user programs are sufficiently fast being
+written in Python and NumPy.
+% smooth experience/intengration from high level <-> low level?
+
 % a brief word about accelerators perhaps?
@@
 
-While NumPy suits most user's needs, it cannot support every specialized need
+NumPy suits most user's needs, but it cannot support every specialized need
 of the community; due to its in-memory data model, it is limited in its ability
@@
 units (TPUs) or other specialized hardware for performance reasons.
+However, both distributed data and the parallel execution of GPUs and TPUs
+map well to the paradigm of array orientated programming.
+This leads to a gap between available modern hardware architecture and the
+necessary tools to leverage their computational power.
 
@@
 The community's efforts to fill the resultant gaps led to a
-proliferation of array implementations. Each deep learning framework created
+proliferation of new array implementations. Each deep learning framework created
 its own arrays; PyTorch \cite{NEURIPS2019_9015},
@@
 geometrical objects \cite{pygeos}, and missing values.
+% Maybe add a sentence in the direction of:
+% We hope/expect, that this will allow the proliferation of rich datatypes
+% which in turn can propel the whole ecosystem forward? (too much :)?
+% As a part of other changes/developements...)
+% Discussed with Jarrod, if it fits to get back to protocols, it is fine
+% to call DTypes a (rich) protocol.

### 2020-01-31 body.tex
Update Fig. 3
Edit NumPy array section

@@
 
-Many of the array-aware functions, such as summation, support \emph{reductions}: aggregating
-results across one or more dimensions of the array.
-For example, summing a $n$-dimensional array over $d$ axes results in a
+Other array-aware functions, such as \code{maximum}, perform
+element-by-element \emph{reductions}, aggregating results across one,
+multiple, or all axes of a single array.
+For example, summing an $n$-dimensional array over $d$ axes results in a
 $(n-d)$-dimensional array (Fig.~\ref{fig:array-concepts}e).
@@
 for the CPUs at hand.
+
+Altogether, the combination of a simple in-memory array
+representation, a syntax that closely mimics mathematics, and a
+variety of array-aware utility functions forms a productive and
+powerfully expressive array programming language.
 
@@
 
-\begin{figure}
+\begin{figure*}
   \centering
-  \includegraphics[width=.45\textwidth]{static/sketches/duck-arrays}
+  \includegraphics[width=.9\textwidth]{static/sketches/array_protocol.jpg}
   \caption{\textbf{Interoperability.} \fixme{This figure needs work.}}\label{fig:duck-arrays}
-\end{figure}
+\end{figure*}
 

### 2020-01-31 body.tex
Rework last paragraph of ecosystem section

@@
 or Jupyter \cite{Kluyver:2016aa}
-provides a complete foundation for array-based scientific computing.
+provides a solid foundation for array-based scientific computing.
 The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
@@
 SciPy is used to provide tools used in general image processing such as 
-filtering and image alignment, while scikit-image, a popular image processing
+filtering and image alignment, while scikit-image, an image processing
 library that extends SciPy, provides higher-level functionality such as
@@
 \code{scipy.optimize} module for handling general optimization tasks.
-NetworkX \cite{SciPyProceedings_11}, a popular package for complex
+NetworkX \cite{SciPyProceedings_11}, a package for complex
 network analysis, is used for analysis of consistency among image
@@
 Matplotlib is used for visualizing data throughout the analysis pipeline,
-including the generation of the final image of the black hole.
+including for the generation of the final image of the black hole.
 
-Similar examples can be found in physics, earth science, biology,
-and more.
-In many fields, the scientific Python ecosystem is the primary
-software environment.
-The recent rapid growth of data science, machine learning, and
-artificial intelligence has dramatically increased scientific
-Python use.
-NumPy and its ecosystem is now taught in
-numerous university courses, bootcamps, and summer schools.
-There are multiple community conferences and workshops.
-NumPy and its API has become ubiquitous.
+The rapid recent growth of data science, machine learning, and
+artificial intelligence has further and dramatically boosted usage of
+scientific Python.  Examples of its impactful application, such as the
+above, now exist in almost every discipline in the natural and social
+sciences, and in many fields these tools have become \emph{the
+primary} software environment.  NumPy and its ecosystem are commonly
+taught in university courses, bootcamps, and summer schools, and are
+at the focus of community conferences and workshops worldwide.
+
+NumPy and its API have become truly ubiquitous.
 

### 2020-01-31 body.tex
Edit additions to proliferation section

@@
     \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
-    \textbf{d,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
+    \textbf{d,} Vectorization efficiently applies operations to groups of elements
+    \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
     \textbf{e,} Reduction operations act along one or more axes. In this
@@
 (i.e., single pointer and strided) arrays on CPUs.
-It runs on machines ranging from embedded devices up to the worlds
-largest supercomputers [citation needed] and provides a performance
-which often approaches compiled languages.
-When algorithms do not fit the array programming approach,
-or require a faster implementation,
-a number of tools (e.g. Cython, Numba, Pythran, ...)
-are available to transparently accelerate crucial parts.
-These tools allow for a high level array orientated code base without sacrificing
-the speed expected from a compiled language.
-In many cases fast algorithms are implemented by libraries in the scientific
-python ecosystem so that most user programs are sufficiently fast being
-written in Python and NumPy.
+It runs on machines ranging from embedded devices to the world's
+largest supercomputers \fixme{citation needed} and yields performance
+approaching that of compiled languages.
+For algorithms that do not fit the array programming approach,
+or require even faster implementations, a variety of tools exist (Cython, Numba, or Pythran, ...) that transparently accelerates bottlenecks.
+These tools allow programmers to express concepts using high level array programming without sacrificing efficiency.
+In many cases, fast algorithms already are implemented by libraries in the scientific Python ecosystem; using these primitives, efficient codes can be written in pure Python.
 % smooth experience/intengration from high level <-> low level?

### 2020-01-31 authors.tex
Add author affiliations

--- 2020-01-03 authors.tex
@@
 \author[1]{Charles R Harris}
+\author[2,3,*]{K. Jarrod Millman}
 \author[2,*]{St\'efan J. van der Walt}
-\author[3,*]{Ralf Gommers}
-\author[2,4,*]{K. Jarrod Millman}
+\author[4,*]{Ralf Gommers}
+\author[5]{Pauli Virtanen}
+\author[6]{David Cournapeau}
+\author[7]{Eric Wieser}
+\author[80]{Julian Taylor}
 \author[2]{Sebastian Berg}
-\author[5]{Kevin Sheppard}
-\author[6]{Tyler J. Reddy}
+\author[80]{Nathaniel J. Smith}
+\author[10]{Robert Kern}
+\author[2]{Matti Picus}
+\author[11]{Stephan Hoyer}
+\author[12]{Marten van Kerkwijk}
+\author[13]{Matthew Brett}
+\author[14]{Allan Haldane}
+\author[15]{Jaime Fernandez del Rio}
+\author[16]{Mark Wiebe}
+\author[4]{Pearu Peterson}
+\author[80]{David M. Cooke}
+\author[18]{Pierre Gerard-Marchant}
+\author[19]{Kevin Sheppard}
+\author[20]{Tyler J. Reddy}
 \author[2]{Warren Weckesser}
-\author[3]{Travis E. Oliphant}
+\author[4]{Hameer Abbasi}
+\author[21]{Christoph Gohlke}
+\author[4]{Travis E. Oliphant}
 \affil[1]{Independent Researcher}
-\affil[2]{Berkeley Institute for Data Science, University of California, Berkeley, CA 94720, USA}
-\affil[3]{Quansight LLC, Austin, TX, USA}
-\affil[4]{Division of Biostatistics, University of California, Berkeley, CA 94720, USA}
-\affil[5]{Department of Economics, University of Oxford, Oxford, OX1 3UQ, UK}
-\affil[6]{Los Alamos National Laboratory, Los Alamos, NM 87544, USA}
+\affil[2]{Berkeley Institute for Data Science, University of California, Berkeley, CA, USA}
+\affil[3]{Division of Biostatistics, University of California, Berkeley, CA, USA}
+\affil[4]{Quansight LLC, Austin, TX, USA}
+\affil[5]{Department of Physics and Nanoscience Center, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
+\affil[6]{Mercai JP, Tokyo, Japan}
+\affil[7]{University of Cambridge, Cambridge, UK}
+\affil[10]{Enthought, Inc., Austin, TX, USA}
+\affil[11]{Google Research, Mountain View, CA, USA}
+\affil[12]{Department of Astronomy \& Astrophysics, University of Toronto, Toronto, ON, Canada}
+\affil[13]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
+\affil[14]{Department of Chemistry, Temple University, Philadelphia, PA, USA}
+\affil[15]{Google, Zurich, Switzerland}
+\affil[16]{AWS Thinkbox, Seattle, Washington, USA}
+\affil[18]{France-IX, France}
+\affil[19]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[20]{Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[21]{Biomedical Engineering Department, University of California, Irvine, USA}
+\affil[80]{Not sure}
 \affil[*]{numpy.articles@gmail.com}

### 2020-01-31 body.tex
Update captions

@@
     \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
-    \textbf{e,} Reduction operations act along one or more axes. In this
+    \textbf{f,} Reduction operations act along one or more axes. In this
     example, a three-dimensional array is shown to be summed along various single
@@
     produce a one-dimensional result.
+    \textbf{g,} Example NumPy code, illustrating some of these concepts.
    }
@@
   \includegraphics[width=.9\textwidth]{static/sketches/array_protocol.jpg}
-  \caption{\textbf{Interoperability.} \fixme{This figure needs work.}}\label{fig:duck-arrays}
+  \caption{\textbf{Interoperability.} Through its array protocol, the NumPy API can operate on various types of array objects, including those written outside of NumPy.
+In this example, NumPy's \code{sum} function is called on a NumPy and a dask array, respectively.
+The call gets succeeds by dispatching to the appropriate library implementation, and results in different output array types.}\label{fig:duck-arrays}
 \end{figure*}

### 2020-01-31 body.tex
Strengthen final paragraph; still needs editing

@@
 
-Over time the role of the NumPy has evolved.
-It began as an attempt to add an array object to Python
-and became the foundation of a rich ecosystem of tools.
-Now a large amount of scientific work depends on NumPy
-being correct, fast, and stable.
-Yet, it is no longer just the foundational array library underlying the
-scientific Python ecosystem; it is also the standard API for tensor
-computation and a central coordinating mechanism between array types and
-technologies in Python.
-NumPy continues to evolve.
-For instance, the NumPy data type system is currently being redesigned to make it
+NumPy began as an attempt to add an array object to Python and became
+the foundation of a rich ecosystem of tools.
+Now, a large amount of scientific work depends on NumPy being correct,
+fast, and stable.
+Work on NumPy continues, so that it may meet the needs of its users.  For instance, the NumPy data type system is currently being redesigned to make it
 easier to implement data types both in Python and C, and to better handle
-richer data types including physical units \cite{astropy,Goldbaum2018,pint},
+rich data types such as physical units \cite{astropy,Goldbaum2018,pint},
 geometrical objects \cite{pygeos}, and missing values.
@@
 % to call DTypes a (rich) protocol.
+
+NumPy's role in the ecosystem continues to evolve. It is no longer just the foundational array library underlying the
+scientific Python ecosystem, but has become the standard API for tensor
+computation and a central coordinating mechanism between array types and
+technologies in Python.

### 2020-01-31 body.tex
Move figure

@@
 and NumArray \cite{oliphant2006guide}.
+
+\begin{figure*}[h]
+  \centering
+  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
+  \caption{\textbf{Fundamental Array Concepts.}
+    \textbf{a,} The NumPy array data structure and its associated metadata fields.
+    \textbf{b,} Indexing an array with various types of arguments.
+    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
+    \textbf{d,} Vectorization efficiently applies operations to groups of elements
+    \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
+    \textbf{f,} Reduction operations act along one or more axes. In this
+    example, a three-dimensional array is shown to be summed along various single
+    axes to produce two-dimensional results, or along two axes consecutively to
+    produce a one-dimensional result.
+    \textbf{g,} Example NumPy code, illustrating some of these concepts.
+   }
+  \label{fig:array-concepts}
+\end{figure*}
 
@@
 \section*{NumPy arrays}
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
-  \caption{\textbf{Fundamental Array Concepts.}
-    \textbf{a,} The NumPy array data structure and its associated metadata fields.
-    \textbf{b,} Indexing an array with various types of arguments.
-    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
-    \textbf{d,} Vectorization efficiently applies operations to groups of elements
-    \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
-    \textbf{f,} Reduction operations act along one or more axes. In this
-    example, a three-dimensional array is shown to be summed along various single
-    axes to produce two-dimensional results, or along two axes consecutively to
-    produce a one-dimensional result.
-    \textbf{g,} Example NumPy code, illustrating some of these concepts.
-   }
-  \label{fig:array-concepts}
-\end{figure*}
 
@@
 It runs on machines ranging from embedded devices to the world's
-largest supercomputers \fixme{citation needed} and yields performance
+largest supercomputers and yields performance
 approaching that of compiled languages.

### 2020-01-31 body.tex
minor edits

@@
 
+% KJM: I moved this here to get the placement right 
 \begin{figure*}[h]
@@
 Because of its inherent simplicity, the NumPy array is
-the {\it de facto} exchange format for array data in Python.
+the \emph{de facto} exchange format for array data in Python.
 The library has such widespread adoption that not only the array object but also its
-{\it Application Programming Interface} (API) has become widespread as
+\emph{Application Programming Interface} (API) has become widespread as
 a language for array programming.
@@
 \emph{vectorized} calculations on arrays, including arithmetic, statistics, and
-trigonometry.
+trigonometry (Fig.~\ref{fig:array-concepts}d).
 Vectorization---operating on whole arrays rather than their individual
@@
 above, now exist in almost every discipline in the natural and social
-sciences, and in many fields these tools have become \emph{the
+sciences.  In many fields these tools have become \emph{the
 primary} software environment.  NumPy and its ecosystem are commonly
@@
 For algorithms that do not fit the array programming approach,
-or require even faster implementations, a variety of tools exist (Cython, Numba, or Pythran, ...) that transparently accelerates bottlenecks.
+or require even faster implementations, a variety of tools exist (e.g., Cython, Numba, Pythran)
+that transparently accelerate bottlenecks.
 These tools allow programmers to express concepts using high level array programming without sacrificing efficiency.
@@
   \centering
-  \includegraphics[width=.9\textwidth]{static/sketches/array_protocol.jpg}
-  \caption{\textbf{Interoperability.} Through its array protocol, the NumPy API can operate on various types of array objects, including those written outside of NumPy.
-In this example, NumPy's \code{sum} function is called on a NumPy and a dask array, respectively.
-The call gets succeeds by dispatching to the appropriate library implementation, and results in different output array types.}\label{fig:duck-arrays}
+  \includegraphics[width=\textwidth]{static/sketches/array_protocol.jpg}
+
+  \caption{\textbf{Interoperability.} Through its array protocol, the NumPy API
+     can operate on various types of array objects, including those written outside
+     of NumPy. In this example, NumPy's \code{sum} function is called on a NumPy and
+     a Dask array, respectively. The call succeeds by dispatching to the
+     appropriate library implementation, and results in different output array
+     types.
+  }\label{fig:duck-arrays}
+
 \end{figure*}
@@
 Even though NumPy is not part of Python's standard library,
-we benefit from a good relationship with the Python developers.
+it benefits from a good relationship with the Python developers.
 Over the years, the Python language has added new features and
@@
 easy to read array notation.
-Since we are not part of the standard library, we are able to
-dictate our own release policies and development patterns.
+Since it is not part of the standard library, NumPy is able to
+dictate its own release policies and development patterns.
 
@@
 these arrays and pass them back to Python.
-Popular array-aware libraries increased the user community and that
-in turn led to more libraries adding new functionality
-to the scientific Python ecosystem.
 

### 2020-01-31 body.tex
Trivial edits to interop

@@
 It runs on machines ranging from embedded devices to the world's
-largest supercomputers and yields performance
+largest supercomputers, with performance
 approaching that of compiled languages.
@@
 
-Importantly, some projects chose to follow the NumPy API exactly---with Dask,
+Importantly, some projects follow the NumPy API exactly---with Dask,
 CuPy, Sparse and JAX the most prominent examples---while others have
-NumPy-inspired semantics but chose to deviate in places.
+NumPy-inspired semantics but, often due to their development history, deviate in places.
 % when they deviate, they sometimes fix it, e.g.,

### 2020-01-31 body.tex
Update Fig. 1 and 3

@@
 NumPy-inspired semantics but, often due to their development history, deviate in places.
+Wherever there are such deviations, we often observe users gently requesting,
+and developers evolving the software step by step towards the NumPy API,
+provided this is possible and reasonable.
+
 % when they deviate, they sometimes fix it, e.g.,
@@
      of NumPy. In this example, NumPy's \code{sum} function is called on a NumPy and
-     a Dask array, respectively. The call succeeds by dispatching to the
-     appropriate library implementation, and results in different output array
+     a Dask array, respectively. The calls succeed by dispatching to the
+     appropriate library implementation, and result in different output array
      types.

### 2020-01-31 body.tex
Minor edits

@@
 Wherever there are such deviations, we often observe users gently requesting,
-and developers evolving the software step by step towards the NumPy API,
-provided this is possible and reasonable.
+and developers evolving the software step by step toward the NumPy API.
 

### 2020-01-31 body.tex
Cleanup

@@
 
-Other array-aware functions, such as \code{maximum}, perform
+Other array-aware functions, such as \code{sum}, \code{mean}, and \code{maximum}, perform
 element-by-element \emph{reductions}, aggregating results across one,
@@
 However, both distributed data and the parallel execution of GPUs and TPUs
-map well to the paradigm of array orientated programming.
+map well to the paradigm of array programming.
 This leads to a gap between available modern hardware architecture and the
@@
 In fact, Python's adoption as a primary learning language in universities
-(e.g., many lower division UC Berkeley courses use Python including
-Foundations of Data Science,
-The Beauty and Joy of Computing,
-Introduction to Computational Thinking with Data,
-and The Structure and Interpretation of Computer Programs) has
-further popularized NumPy as a tool of choice for modern data science.
+has further popularized NumPy as a tool of choice for modern data science.
 

### 2020-01-31 body.tex
Shape the story of the proliferation section, according to my and Jarrod's discussion

@@
 
-NumPy provides in-memory, multidimensional, homogeneously typed
-(i.e., single pointer and strided) arrays on CPUs.
-It runs on machines ranging from embedded devices to the world's
-largest supercomputers, with performance
-approaching that of compiled languages.
-For algorithms that do not fit the array programming approach,
-or require even faster implementations, a variety of tools exist (e.g., Cython, Numba, Pythran)
-that transparently accelerate bottlenecks.
-These tools allow programmers to express concepts using high level array programming without sacrificing efficiency.
-In many cases, fast algorithms already are implemented by libraries in the scientific Python ecosystem; using these primitives, efficient codes can be written in pure Python.
-% smooth experience/intengration from high level <-> low level?
+NumPy provides in-memory, multidimensional, homogeneously typed (i.e.,
+single pointer and strided) arrays on CPUs.  It runs on machines
+ranging from embedded devices to the world's largest supercomputers,
+with performance approaching that of compiled languages.  For the
+largest part of its existence, NumPy covered the vast majority of
+array computation use cases.
+
+%For algorithms that do not fit the array programming approach,
+%or require even faster implementations, a variety of tools exist (e.g., Cython, Numba, Pythran)
+%that transparently accelerate bottlenecks.
+%These tools allow programmers to express concepts using high level array programming without sacrificing efficiency.
+%In many cases, fast algorithms already are implemented by libraries in the scientific Python ecosystem; using these primitives, efficient codes can be written in pure %Python.
 
@@
 
-NumPy suits most user's needs, but it cannot support every specialized need
-of the community; due to its in-memory data model, it is limited in its ability
-to work with very large datasets, datasets split across multiple systems, and
-computations that require graphics processing units (GPUs), tensor processing
-units (TPUs) or other specialized hardware for performance reasons.
-However, both distributed data and the parallel execution of GPUs and TPUs
-map well to the paradigm of array programming.
-This leads to a gap between available modern hardware architecture and the
-necessary tools to leverage their computational power.
+In recentl years, there has been a rapid growth in the development of
+specialized array processing hardware, including graphics processing
+units (GPUs), tensor processing units (TPUs), and field-programmable
+gate arrays (FPGAs).  Routinely, data sets are now larger than a
+single machine's memory, and may be stored on multiple machines or in
+the cloud.  Due to its in-memory data model, NumPy is unable to
+utilize such specialized hardware and storage.  However, both
+distributed data and the parallel execution of GPUs and TPUs map well
+to the \emph{paradigm} of array programming: a gap therefore existed between
+available modern hardware architectures and the tools necessary to
+leverage their computational power.
 
@@
 % and why (presumably to get access to ecosystem)
-The community's efforts to fill the resultant gaps led to a
+The community's efforts to fill this gap led to a
 proliferation of new array implementations. Each deep learning framework created
@@
 
-Importantly, some projects follow the NumPy API exactly---with Dask,
-CuPy, Sparse and JAX the most prominent examples---while others have
-NumPy-inspired semantics but, often due to their development history, deviate in places.
-Wherever there are such deviations, we often observe users gently requesting,
-and developers evolving the software step by step toward the NumPy API.
+Such libraries often mimic the NumPy API, because it lowers the
+barrier to entry for newcomers and provides the wider community with a
+stable array programming interface. This, in turn, prevents disruptive
+schisms like the divergence of Numeric and NumArray.
+But exploring new ways of working with array is experimental by nature
+and, in fact, several promising libraries---such as Caffe, Theano, and
+biggus---have already ceased development. And each time that a user
+decides to try a new technology or backend library, they have to
+change import statements and ensure that the new library implements
+all the parts of the NumPy API they currently use.
+
+%CuPy, Sparse and JAX the most prominent examples---while others have
+%NumPy-inspired semantics but, often due to their development history, deviate in places.
+%Wherever there are such deviations, we often observe users gently requesting,
+%and developers evolving the software step by step toward the NumPy API.
 
@@
 
-Ideally, operating on these arrays using NumPy functions or semantics would
+Ideally, operating on specialized arrays using NumPy functions or semantics would
 simply work, so that end users could write code once, and would then benefit
@@
 To support array operations between external array objects, NumPy
-acts as a central coordination mechanism with a well-specified API.
-The widespread adoption of the NumPy API is valuable: it lowers the
-barrier to entry for newcomers and provides the wider community with a
-stable array programming interface. This, in turn, prevents disruptive
-schisms like the divergence of Numeric and NumArray, by facilitating
-the development of specialized solutions and new tools that operate in
-concert.
+therefore added the capability to act as a central coordination
+mechanism with a well-specified API.
 
@@
 % https://github.com/pytorch/pytorch/issues/22402
-To facilitate \emph{interoperability}, NumPy provides
+To facilitate this \emph{interoperability}, NumPy provides
 ``protocols'' (or contracts of operation), that allow for these arrays to be
@@
 only minimal code changes.
+As an example, consider the arrays shown in ...
+\fixme{this part to be fleshed out to match the code in the figure}
 % spelled out on this page: https://docs.dask.org/en/latest/gpu.html
@@
 % https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
-However, general adoption has not yet been achieved mainly for compatibility
-concerns.
+%However, general adoption has not yet been achieved mainly for compatibility
+%concerns.
 % explain a bit what "compatibility concerns" means ...
-Thus, we are actively addressing these concerns by revising and extending
-these protocols with the aim of wide interoperability that spans the whole
-scientific Python ecosystem up to and including domain libraries.
+%Thus, we are actively addressing these concerns by revising and extending
+%these protocols with the aim of wide interoperability that spans the whole
+%scientific Python ecosystem up to and including domain libraries.
 

### 2020-02-01 body.tex
Minor edits

@@
 
-In recentl years, there has been a rapid growth in the development of
-specialized array processing hardware, including graphics processing
+In recent years, there has been a rapid growth in the development of
+specialized hardware, including graphics processing
 units (GPUs), tensor processing units (TPUs), and field-programmable
@@
 single machine's memory, and may be stored on multiple machines or in
-the cloud.  Due to its in-memory data model, NumPy is unable to
-utilize such specialized hardware and storage.  However, both
-distributed data and the parallel execution of GPUs and TPUs map well
+the cloud.  Due to its in-memory data model, NumPy is currently unable to
+utilize such specialized hardware and storage directly.  However, both
+distributed data and the parallel execution of GPUs, TPUs, and FPGAs map well
 to the \emph{paradigm} of array programming: a gap therefore existed between
@@
 schisms like the divergence of Numeric and NumArray.
-But exploring new ways of working with array is experimental by nature
+But exploring new ways of working with arrays is experimental by nature
 and, in fact, several promising libraries---such as Caffe, Theano, and
@@
 
-  \caption{\textbf{Interoperability.} Through its array protocol, the NumPy API
+  \caption{\textbf{NumPy Array Protocol.} Through its array protocol, the NumPy API
      can operate on various types of array objects, including those written outside
@@
 
-% should we hint at some of the community interaction part
-% they have helped us develop the neps and e.g.,
-% https://github.com/pytorch/pytorch/issues/22402
 To facilitate this \emph{interoperability}, NumPy provides
@@
 NumPy, in turn, dispatches operations to the originating library, as required.
+Thanks to these developments users of Dask and NumPy arrays, for instance, are now capable
+of scaling their computation to the GPU using Dask's CuPy integration with
+only minimal code changes.
+
 % FIXME: expand to 3 or more sentences discussing current efforts and
@@
 %\url{https://numpy.org/neps/nep-0037-array-module.html}
-Versions of these protocols have been successfully deployed.
+%Versions of these protocols have been successfully deployed.
 % Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
@@
 % Below an attempt to start on this:
-Thanks to these developments users of Dask and NumPy arrays are now capable
-of scaling their computation to the GPU using Dask's CuPy integration with
-only minimal code changes.
-As an example, consider the arrays shown in ...
-\fixme{this part to be fleshed out to match the code in the figure}
+
+% should we hint at some of the community interaction part
+% they have helped us develop the neps and e.g.,
+% https://github.com/pytorch/pytorch/issues/22402
+
 % spelled out on this page: https://docs.dask.org/en/latest/gpu.html

### 2020-02-01 body.tex
One proliferation section change; some textual things and comments
We're getting to the stage where it's hard to find obvious issues;
a few things I wanted to change but I thought it'd be better to
add comments for discussion instead.

@@
 % KJM: I moved this here to get the placement right 
+% RG: minor comment on the example code: would be better to not copy IPython
+%     oddities but align output on brackets and In/Out on ":"
 \begin{figure*}[h]
@@
 The NumPy array is a data structure that efficiently stores and accesses
-multidimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, that
+multidimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, and
 enables a wide variety of scientific computation.
@@
 
+% RG: (to discuss) this figure plus associated text says that the layers about
+%     Foundation build on top of that whole foundation. In reality that's not
+%     true - everything builds on NumPy, almost everything on SciPy, and very
+%     little on Matplotlib/IPython/Jupyter. Perhaps "interactive environment"
+%     should get its own section instead.
+%     Also, I think Pandas can be added next to statsmodels as statistics. That
+%     avoids explaining dataframes; now it looks like a serious omission.
+%     I'd also be in favor of adding Spyder in the text; scientific computing
+%     beyond exploratory work is better done in an IDE.
 \begin{figure}
@@
 Other components of SciPy are used throughout the library, such as the
-\code{scipy.optimize} module for handling general optimization tasks.
+\code{scipy.optimize} module for handling mathematical optimization tasks.
 NetworkX \cite{SciPyProceedings_11}, a package for complex
@@
 
-NumPy provides in-memory, multidimensional, homogeneously typed (i.e.,
-single pointer and strided) arrays on CPUs.  It runs on machines
+NumPy provides in-memory, multidimensional, homogeneously typed
+arrays on CPUs.  It runs on machines
 ranging from embedded devices to the world's largest supercomputers,
@@
 But exploring new ways of working with arrays is experimental by nature
-and, in fact, several promising libraries---such as Caffe, Theano, and
-biggus---have already ceased development. And each time that a user
-decides to try a new technology or backend library, they have to
+and, in fact, several promising libraries---such as Theano and Caffe---have
+already ceased development. And each time that a user
+decides to try a new technology, they have to
 change import statements and ensure that the new library implements
@@
 NumPy, in turn, dispatches operations to the originating library, as required.
-Thanks to these developments users of Dask and NumPy arrays, for instance, are now capable
-of scaling their computation to the GPU using Dask's CuPy integration with
-only minimal code changes.
+
+Thanks to these developments users are now capable, for example, of scaling
+their computation to distributed systems with Dask. The protocols
+\textit{compose} well, so it is even possible to take NumPy code and scale it
+to multi-GPU systems with minimal code changes through Dask arrays containing
+multiple CuPy arrays \cite{entschev2019}.
 
@@
 
+% RG: this first sentence is duplicate with (b) three paragraphs up - on
+%     purpose or not?
 NumPy is a community of practice with a strong \emph{culture} of
@@
 Community members explicitly work to address this lack of formal education
-through formal courses and workshops
+through courses and workshops
 \cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.

### 2020-02-02 body.tex
Minor edits

@@
 %     oddities but align output on brackets and In/Out on ":"
+% KJM: agreed
+% FIXME we should figure out if we want to keep working on the figures in this format or not
 \begin{figure*}[h]
@@
 %     beyond exploratory work is better done in an IDE.
+% KJM: agreed
+% FIXME we should figure out if we want to keep working on the figures in this format or not
 \begin{figure}
@@
 NumPy (a) is built on top of Python,
-(b) has a strong culture of collaboration and good software development practices, and
+(b) has a culture of collaboration and good software development practices, and
 (c) provides a cutting-edge array programming environment for scientific data analysis.
@@
 %     purpose or not?
+% KJM: yes, sort of, (a) corresponds to the two para above
+% (b) corresponds to the next two para and
+% (c) corresponds to the last two (or three) para
 NumPy is a community of practice with a strong \emph{culture} of

### 2020-02-03 static/sketches/fig1.py
Update Fig 1 and 3

@@ -1,0 +1,41 @@
+DejaVu Sans Mono
+Size: 62
+
+# Fig. 1
+
+In [1]: import numpy as np
+
+In [2]: x = np.arange(12)
+
+In [3]: x = x.reshape(4, 3)
+
+In [4]: x
+Out[4]:
+array([[ 0,  1,  2],
+       [ 3,  4,  5],
+       [ 6,  7,  8],
+       [ 9, 10, 11]])
+
+In [5]: np.mean(x, axis=0)
+Out[5]: array([4.5, 5.5, 6.5])
+
+In [6]: x = x - np.mean(x, axis=0)
+
+In [7]: np.std(x, axis=0)
+Out[7]: array([3.35, 3.35, 3.35])
+
+In [8]: x = x / np.std(x, axis=0)
+
+In [9]: x
+Out[9]:
+array([[-1.34, -1.34, -1.34],
+       [-0.45, -0.45, -0.45],
+       [ 0.45,  0.45,  0.45],
+       [ 1.34,  1.34,  1.34]])
+
+--------
+
+array([[-1.34, -1.34164079, -1.34164079],
+       [-0.45 , -0.4472136 , -0.4472136 ],
+       [ 0.45 ,  0.4472136 ,  0.4472136 ],
+       [ 1.34,  1.34164079,  1.34164079]])

### 2020-02-03 static/sketches/fig3.py

@@ -1,0 +1,49 @@
+DejaVu Sans Mono
+Size: 62
+
+# Fig. 3
+
+In [1]: import numpy as np
+
+In [2]: import dask.array as da
+
+In [3]: x = da.arange(12)
+
+In [4]: x = np.reshape(x, (4, 3))
+
+In [5]: x
+Out[5]: dask.array<..., shape=(4, 3), ...>
+
+In [6]: np.mean(x, axis=0)
+Out[6]: dask.array<..., shape=(3,), ...>
+
+In [7]: x = x - np.mean(x, axis=0)
+
+In [8]: x = x / np.std(x, axis=0)
+
+In [9]: x
+Out[9]: dask.array<..., shape=(4, 3), ...>
+
+
+---------
+
+In [1]: import numpy as np
+
+In [2]: import dask.array as da
+
+In [3]: x = da.arange(12)
+
+In [4]: x = np.reshape(x, (4, 3))
+
+In [5]: x
+Out[5]: dask.array<reshape, shape=(4, 3), dtype=int64, chunksize=(4, 3), chunktype=numpy.ndarray>
+
+In [6]: np.mean(x, axis=0)
+Out[6]: dask.array<mean_agg-aggregate, shape=(3,), dtype=float64, chunksize=(3,), chunktype=numpy.ndarray>
+
+In [7]: x = x - np.mean(x, axis=0)
+
+In [8]: x = x / np.std(x, axis=0)
+
+In [9]: x
+Out[9]: dask.array<truediv, shape=(4, 3), dtype=float64, chunksize=(4, 3), chunktype=numpy.ndarray>

### 2020-02-03 body.tex
Update Fig 1 and 3
Minor edits

@@
 % KJM: I moved this here to get the placement right 
-% RG: minor comment on the example code: would be better to not copy IPython
-%     oddities but align output on brackets and In/Out on ":"
-% KJM: agreed
-% FIXME we should figure out if we want to keep working on the figures in this format or not
 \begin{figure*}[h]
@@
 The combination of NumPy, SciPy, and Matplotlib together with
-an advanced interactive environment like IPython \cite{perez2007ipython}
-or Jupyter \cite{Kluyver:2016aa}
-provides a solid foundation for array-based scientific computing.
+an advanced interactive environment like IPython \cite{perez2007ipython},
+or Jupyter \cite{Kluyver:2016aa} provides a solid foundation for array
+programming in Python.
 The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
@@
 then stitched together into imperative or functional programs, or
-narrative computational notebooks.  This rich and productive
-environment has made Python popular
+narrative computational notebooks.
+Scientific computing beyond exploratory work is often done in a text editor
+or an integrated development environment (IDEs) such as Spyder.
+This rich and productive environment has made Python popular
 for scientific research.
@@
   \centering
-  \includegraphics[width=.45\textwidth]{static/ecosystem}
+  \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
   \caption{\textbf{Scientific Python Ecosystem.}
@@
 
-  \caption{\textbf{NumPy Array Protocol.} Through its array protocol, the NumPy API
+  \caption{\textbf{NumPy API and Array Protocol.} Through its array protocol, the NumPy API
      can operate on various types of array objects, including those written outside
-     of NumPy. In this example, NumPy's \code{sum} function is called on a NumPy and
-     a Dask array, respectively. The calls succeed by dispatching to the
-     appropriate library implementation, and result in different output array
-     types.
+     of NumPy. In this example, NumPy's \code{mean} function is called on a a Dask
+     array.  The call succeeds by dispatching to the appropriate library implementation
+     (i.e., Dask in this case) and results in a new Dask array.  Compare this
+     code to the example code in Fig.~\ref{fig:array-concepts}g.
   }\label{fig:duck-arrays}
@@
 
-% RG: this first sentence is duplicate with (b) three paragraphs up - on
-%     purpose or not?
-% KJM: yes, sort of, (a) corresponds to the two para above
-% (b) corresponds to the next two para and
-% (c) corresponds to the last two (or three) para
 NumPy is a community of practice with a strong \emph{culture} of

### 2020-02-04 authors.tex
Update letters
Add NS affilitation
Update PP affil

@@
 \author[2]{Sebastian Berg}
-\author[80]{Nathaniel J. Smith}
+\author[9]{Nathaniel J. Smith}
 \author[10]{Robert Kern}
@@
 \author[16]{Mark Wiebe}
-\author[4]{Pearu Peterson}
+\author[4,17,18]{Pearu Peterson}
 \author[80]{David M. Cooke}
-\author[18]{Pierre Gerard-Marchant}
-\author[19]{Kevin Sheppard}
-\author[20]{Tyler J. Reddy}
+\author[20]{Pierre Gerard-Marchant}
+\author[21]{Kevin Sheppard}
+\author[22]{Tyler J. Reddy}
 \author[2]{Warren Weckesser}
 \author[4]{Hameer Abbasi}
-\author[21]{Christoph Gohlke}
+\author[23]{Christoph Gohlke}
 \author[4]{Travis E. Oliphant}
-\affil[1]{Independent Researcher}
+\affil[1]{Independent Researcher, Logan, Utah, USA}
 \affil[2]{Berkeley Institute for Data Science, University of California, Berkeley, CA, USA}
@@
 \affil[7]{University of Cambridge, Cambridge, UK}
+\affil[9]{Independent Researcher, Berkeley, CA, USA}
 \affil[10]{Enthought, Inc., Austin, TX, USA}
@@
 \affil[16]{AWS Thinkbox, Seattle, Washington, USA}
-\affil[18]{France-IX, France}
-\affil[19]{Department of Economics, University of Oxford, Oxford, UK}
-\affil[20]{Los Alamos National Laboratory, Los Alamos, NM, USA}
-\affil[21]{Biomedical Engineering Department, University of California, Irvine, USA}
+\affil[17]{Independent Researcher, Saue, Estonia}
+\affil[18]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
+\affil[20]{France-IX, France}
+\affil[21]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[22]{Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[24]{Biomedical Engineering Department, University of California, Irvine, USA}
 \affil[80]{Not sure}

### 2020-02-04 body.tex
First rough re-sketch of ecosystem in Inkscape
Remove old comments
Update captions

@@
 
-% RG: (to discuss) this figure plus associated text says that the layers about
-%     Foundation build on top of that whole foundation. In reality that's not
-%     true - everything builds on NumPy, almost everything on SciPy, and very
-%     little on Matplotlib/IPython/Jupyter. Perhaps "interactive environment"
-%     should get its own section instead.
-%     Also, I think Pandas can be added next to statsmodels as statistics. That
-%     avoids explaining dataframes; now it looks like a serious omission.
-%     I'd also be in favor of adding Spyder in the text; scientific computing
-%     beyond exploratory work is better done in an IDE.
-% KJM: agreed
-% FIXME we should figure out if we want to keep working on the figures in this format or not
 \begin{figure}
@@
    provide a rich and powerful collection of array-aware functionality.
+   This means that other array implementations are easily accessible to
+   the ecosystem if they use the NumPy API and array protocol (Fig.~\ref{fig:array-protocol}).
   }
@@
 therefore added the capability to act as a central coordination
-mechanism with a well-specified API.
+mechanism with a well-specified API (Fig.~\ref{fig:ecosystem}).
 
@@
      code to the example code in Fig.~\ref{fig:array-concepts}g.
-  }\label{fig:duck-arrays}
+  }\label{fig:array-protocol}
 
@@
 ``protocols'' (or contracts of operation), that allow for these arrays to be
-passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
+passed to NumPy functions (Fig.~\ref{fig:array-protocol}).
 NumPy, in turn, dispatches operations to the originating library, as required.

### 2020-02-04 body.tex
Update caption

@@
   \caption{\textbf{Scientific Python Ecosystem.}
-   This figure highlights just a few of the essential tools and libraries in the vast
-   scientific Python ecosystem.  Most of the ecosystem builds on NumPy arrays to
-   provide a rich and powerful collection of array-aware functionality.
-   This means that other array implementations are easily accessible to
-   the ecosystem if they use the NumPy API and array protocol (Fig.~\ref{fig:array-protocol}).
+   This figure highlights some of the essential tools and libraries, which depend
+   on NumPy.  Since most of the ecosystem
+   builds on NumPy arrays and its rich and powerful collection of array-aware
+   functionality, the simplest way for other array implementations to become
+   available to the vast scientific Python ecosystem is by using the NumPy API and array
+   protocols (Fig.~\ref{fig:array-protocol}).
   }

### 2020-02-04 body.tex
DOC: Address techincia grammar issues
Address technical grammar issues such as - and ,
Substitute some frequent works with synonyms

@@
 scikit-image \cite{vanderwalt2014scikit}.
-It is a community developed, open source library, which provides a
+It is a community-developed, open-source library, which provides a
 multidimensional Python array object along with array-aware functions
@@
 and the number of axes is the array's dimensionality.
-For example, a vector of numbers can be stored as a one dimensional array of
-shape $N$, while color videos are four dimensional arrays of shape
+For example, a vector of numbers can be stored as a one-dimensional array of
+shape $N$, while color videos are four-dimensional arrays of shape
 $(T, M, N, 3)$.
@@
 It describes the number of bytes to move forward in memory to jump from row to
-row, column to column, and so forth.
-Consider, for example, a 2-D array of floating point numbers with shape
+row or column to column.
+Consider, for example, a 2-D array of floating-point numbers with shape
 $(4, 3)$, where each element occupies 8 bytes in memory.
@@
 first over either rows or columns.  This allows external libraries
-written in those languages to directly access NumPy array data in memory.
+written in those languages to access NumPy array data in memory directly.
 
@@
 the original array, such that data is shared between the two arrays.
-This provides a powerful way to operate on subsets of array data, while
+This provides a powerful way to operate on subsets of array data while
 limiting memory usage.
@@
 A trivial example is the addition of a scalar value to an array, but it also
-generalizes to more complex examples such as scaling each column of an array,
+generalizes to more complex examples such as scaling each column of an array
 or generating a grid of coordinates.
@@
 history, development, and use.
-SciPy provides fundamental algorithms for scientific computing
+SciPy provides fundamental algorithms for scientific computing,
 including mathematical, scientific, and engineering routines.
 Matplotlib generates publication-ready figures and visualizations.
-The combination of NumPy, SciPy, and Matplotlib together with
+The combination of NumPy, SciPy, and Matplotlib, together with
 an advanced interactive environment like IPython \cite{perez2007ipython},
-or Jupyter \cite{Kluyver:2016aa} provides a solid foundation for array
+or Jupyter \cite{Kluyver:2016aa}, provides a solid foundation for array
 programming in Python.
@@
 rapidly iterate to refine programming statements. These statements are
-then stitched together into imperative or functional programs, or
-narrative computational notebooks.
+then stitched together into imperative or functional programs or
+notebooks containing both computation and narrative.
 Scientific computing beyond exploratory work is often done in a text editor
@@
 NetworkX \cite{SciPyProceedings_11}, a package for complex
-network analysis, is used for analysis of consistency among image
+network analysis, is used for the analysis of consistency among image
 comparisons.
@@
 The rapid recent growth of data science, machine learning, and
-artificial intelligence has further and dramatically boosted usage of
+artificial intelligence has further and dramatically boosted the usage of
 scientific Python.  Examples of its impactful application, such as the
 above, now exist in almost every discipline in the natural and social
-sciences.  In many fields these tools have become \emph{the
-primary} software environment.  NumPy and its ecosystem are commonly
-taught in university courses, bootcamps, and summer schools, and are
+sciences.  These tools have become \emph{the
+primary} software environment in many fields.  NumPy and its ecosystem are commonly
+taught in university courses, boot camps, and summer schools, and are
 at the focus of community conferences and workshops worldwide.
@@
 with performance approaching that of compiled languages.  For the
-largest part of its existence, NumPy covered the vast majority of
+most its existence, NumPy addressed the vast majority of
 array computation use cases.
@@
 %that transparently accelerate bottlenecks.
-%These tools allow programmers to express concepts using high level array programming without sacrificing efficiency.
-%In many cases, fast algorithms already are implemented by libraries in the scientific Python ecosystem; using these primitives, efficient codes can be written in pure %Python.
+%These tools allow programmers to express concepts using high-level array programming without sacrificing efficiency.
+%In many cases, fast algorithms are already implemented by libraries in the scientific Python ecosystem;
+%these primitives simplify writing code in pure Python.
 
@@
 units (GPUs), tensor processing units (TPUs), and field-programmable
-gate arrays (FPGAs).  Routinely, data sets are now larger than a
-single machine's memory, and may be stored on multiple machines or in
+gate arrays (FPGAs).  Data sets routinely exceed the memory capacity of a single machine and may be stored on multiple machines or in
 the cloud.  Due to its in-memory data model, NumPy is currently unable to
@@
 distributed data and the parallel execution of GPUs, TPUs, and FPGAs map well
-to the \emph{paradigm} of array programming: a gap therefore existed between
+to the \emph{paradigm} of array programming: a gap, therefore, existed between
 available modern hardware architectures and the tools necessary to
@@
 % to convert to and from numpy arrays
-% and why (presumably to get access to ecosystem)
+% and why (presumably to get access to the ecosystem)
 The community's efforts to fill this gap led to a
@@
 only those in memory for efficiency.
-In addition there are projects that build on top of NumPy arrays as a data
+In addition, there are projects that build on top of NumPy arrays as a data
 container and \textit{extend} its capabilities.  Distributed arrays are
@@
 an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
-\code{x.loc[:,~'time']}---by Xarray \cite{hoyer2017xarray}.
+\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.
 
@@
 already ceased development. And each time that a user
-decides to try a new technology, they have to
+decides to try new technology, they have to
 change import statements and ensure that the new library implements
@@
 
-%CuPy, Sparse and JAX the most prominent examples---while others have
+%CuPy, Sparse, and JAX the most prominent examples---while others have
 %NumPy-inspired semantics but, often due to their development history, deviate in places.
@@
 Ideally, operating on specialized arrays using NumPy functions or semantics would
-simply work, so that end users could write code once, and would then benefit
-from switching between NumPy arrays, GPU arrays, distributed arrays, and so
-forth, as appropriate.
-To support array operations between external array objects, NumPy
-therefore added the capability to act as a central coordination
+simply work, so that users could write code once, and would then benefit
+from switching between NumPy arrays, GPU arrays, or distributed arrays, as appropriate.
+To support array operations between external array objects, NumPy,
+therefore, added the capability to act as a central coordination
 mechanism with a well-specified API (Fig.~\ref{fig:ecosystem}).
@@
      can operate on various types of array objects, including those written outside
-     of NumPy. In this example, NumPy's \code{mean} function is called on a a Dask
+     of NumPy. In this example, NumPy's \code{mean} function is called on a Dask
      array.  The call succeeds by dispatching to the appropriate library implementation
@@
 
-Thanks to these developments users are now capable, for example, of scaling
+Thanks to these developments, users are now capable, for example, of scaling
 their computation to distributed systems with Dask. The protocols
@@
 % should we hint at some of the community interaction part
-% they have helped us develop the neps and e.g.,
+% they have helped us develop the neps and, e.g.
 % https://github.com/pytorch/pytorch/issues/22402
@@
 NumPy was initially developed by students, faculty, and researchers to
-provide an advanced, open source array programming library for Python,
+provide an advanced, open-source array programming library for Python,
 which was free to use and unencumbered by license servers, dongles, and the like.
@@
 first from research and later also industry.
-It spurred the development of the larger scientific Python ecosystem and
-heralded the current era of wide-spread use of Python for scientific computing.
-The reasons for its popularity are varied and complex, but include the fact that
+It spurred the development of the broader scientific Python ecosystem and
+heralded the current era of widespread use of Python for scientific computing.
+The reasons for its popularity are varied and complex but include the fact that
 NumPy (a) is built on top of Python,
@@
 
-\emph{Python} is an open source, general-purpose, interpreted programming language
+\emph{Python} is an open-source, general-purpose, interpreted programming language
 well-suited to standard programming tasks such as cleaning data,
@@
 reduce error \cite{millman2014developing}.  This culture is not only
-adopted by leaders in the project, but also enthusiastically taught to
+adopted by leaders in the project but also enthusiastically taught to
 newcomers. The NumPy team was early in adopting distributed revision
 control and code review to improve collaboration on code, and
-continuous testing that runs a large battery of automated tests for
+continuous testing that runs an extensive battery of automated tests for
 every proposed change to NumPy.  The project has comprehensive,
@@
 the readability, usability, and versatility of Python in a mature,
-well-tested, well-documented, and community developed library.
+well-tested, well-documented, and community-developed library.
 It builds on decades of work refining the array programming
@@
 NumPy began as an attempt to add an array object to Python and became
-the foundation of a rich ecosystem of tools.
+the foundation of a vibrant ecosystem of tools.
 Now, a large amount of scientific work depends on NumPy being correct,
@@
 % Maybe add a sentence in the direction of:
-% We hope/expect, that this will allow the proliferation of rich datatypes
+% We hope/expect, that this will allow the proliferation of rich data types
 % which in turn can propel the whole ecosystem forward? (too much :)?
-% As a part of other changes/developements...)
+% As a part of other changes/developments...)
 % Discussed with Jarrod, if it fits to get back to protocols, it is fine
@@
 NumPy's role in the ecosystem continues to evolve. It is no longer just the foundational array library underlying the
-scientific Python ecosystem, but has become the standard API for tensor
+scientific Python ecosystem but has become the standard API for tensor
 computation and a central coordinating mechanism between array types and

### 2020-02-04 summary.tex
Revise ecosystem
DOC: Update abstract
Small changes to abstract

@@
 % - 2-3 sentences of basic-level introduction to the field;
-Array programming provides powerful notation for accessing,
+Array programming provides a powerful, compact, expressive syntax for accessing,
 manipulating, and operating on data in vectors, matrices, and
 higher-dimensional arrays \cite{iverson1980notation}.
-Vectorization---operating on whole arrays rather than their individual
-elements---is essential to array programming.
 % - a brief account of the background and rationale of the work;
@@
 diverse as physics, chemistry, astronomy, geoscience, biology, psychology,
-material science, and engineering.
+material science, engineering, finance, and economics.
 For example, in astronomy, NumPy was fundamental to discovering gravitational

### 2020-02-04 body.tex
DOC: Add reference for Spyder
DOC: Additional changes
Small rewrites to simplify language

@@
 $(4, 3)$, where each element occupies 8 bytes in memory.
-To move between consecutive columns we need to jump forward 8 bytes in memory,
+To move between consecutive columns, we need to jump forward 8 bytes in memory,
 and to access the next row $3 \times 8 = 24$ bytes.
-The strides of that array are therefore $(24, 8)$.  NumPy is able to
+The strides of that array are therefore $(24, 8)$.  NumPy can
 store arrays in either C or Fortran memory order, iterating
@@
 their analysis, while NumPy handles looping over array elements near-optimally,
-taking into consideration, for example, strides, in order to best utilize the
+taking into consideration, for example, strides, to best utilize the
 computer's fast cache memory.
@@
 Scientific computing beyond exploratory work is often done in a text editor
-or an integrated development environment (IDEs) such as Spyder.
+or an integrated development environment (IDEs) such as Spyder\cite{spyder2020}.
 This rich and productive environment has made Python popular
@@
 and JAX \cite{jax2018github} arrays all have the
-capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
-execution mode to allow for performance optimizations.  SciPy and Sparse both
+capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
+to allow for additional performance optimizations.  SciPy and Sparse both
 provide sparse arrays---which typically contain few non-zero values and store
@@
 already ceased development. And each time that a user
-decides to try new technology, they have to
+decides to try new technology, they must
 change import statements and ensure that the new library implements
@@
 
-Thanks to these developments, users are now capable, for example, of scaling
-their computation to distributed systems with Dask. The protocols
-\textit{compose} well, so it is even possible to take NumPy code and scale it
-to multi-GPU systems with minimal code changes through Dask arrays containing
-multiple CuPy arrays \cite{entschev2019}.
+Thanks to these developments, users can now, for example, scale
+their computation from a single machine to distributed systems using Dask. The protocols
+\textit{compose} well, and so it is possible to redeploy NumPy code onto multi-GPU systems with minimal changes\cite{entschev2019}.
 
@@
 Yorick \cite{munro1995using}, as well as commercial languages and
-environments like IDL and Matlab.
+environments like IDL and {MATLAB}.
 
@@
 Over the years, the Python language has added new features and
-special syntax so that NumPy would have more succinct and 
-easy to read array notation.
+special syntax so that NumPy would have a more succinct and 
+easier to read array notation\cite{smith2014pep465}.
 Since it is not part of the standard library, NumPy is able to

### 2020-02-04 body.tex

@@
 It describes the number of bytes to move forward in memory to jump from row to
-row or column to column.
+row, column to column, and so forth.
 Consider, for example, a 2-D array of floating-point numbers with shape
@@
 rapidly iterate to refine programming statements. These statements are
-then stitched together into imperative or functional programs or
+then stitched together into imperative or functional programs, or
 notebooks containing both computation and narrative.
@@
 ranging from embedded devices to the world's largest supercomputers,
-with performance approaching that of compiled languages.  For the
+with performance approaching that of compiled languages.  For
 most its existence, NumPy addressed the vast majority of
@@
 units (GPUs), tensor processing units (TPUs), and field-programmable
-gate arrays (FPGAs).  Data sets routinely exceed the memory capacity of a single machine and may be stored on multiple machines or in
+gate arrays (FPGAs).  Data sets now routinely exceed the memory capacity of a single machine and may be stored on multiple machines or in
 the cloud.  Due to its in-memory data model, NumPy is currently unable to
@@
 already ceased development. And each time that a user
-decides to try new technology, they must
+decides to try a new technology, they must
 change import statements and ensure that the new library implements
@@
 simply work, so that users could write code once, and would then benefit
-from switching between NumPy arrays, GPU arrays, or distributed arrays, as appropriate.
+from switching between NumPy arrays, GPU arrays, distributed arrays,
+and so forth, as appropriate.
 To support array operations between external array objects, NumPy,
@@
 heralded the current era of widespread use of Python for scientific computing.
-The reasons for its popularity are varied and complex but include the fact that
+The reasons for its popularity are varied and complex but include the facts that
 NumPy (a) is built on top of Python,

### 2020-02-04 body.tex
Remove URL citations from main text

@@
 Scientific computing beyond exploratory work is often done in a text editor
-or an integrated development environment (IDEs) such as Spyder\cite{spyder2020}.
+or an integrated development environment (IDEs) such as Spyder.
 This rich and productive environment has made Python popular
@@
 Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
-and JAX \cite{jax2018github} arrays all have the
+and JAX arrays all have the
 capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
@@
 special syntax so that NumPy would have a more succinct and 
-easier to read array notation\cite{smith2014pep465}.
+easier to read array notation.
 Since it is not part of the standard library, NumPy is able to
@@
 easier to implement data types both in Python and C, and to better handle
-rich data types such as physical units \cite{astropy,Goldbaum2018,pint},
-geometrical objects \cite{pygeos}, and missing values.
+rich data types such as physical units \cite{astropy,Goldbaum2018},
+geometrical objects, and missing values.
 % Maybe add a sentence in the direction of:

### 2020-02-04 summary.tex
Remove URL citations from main text

@@
 For example, in astronomy, NumPy was fundamental to discovering gravitational
-waves\cite{pycbc}, the first imaging of a black hole\cite{eht-imaging}, and
+waves \cite{abbott2016observation}, the first imaging of a black hole \cite{eht-imaging}, and
 continues underpinning the exploration of our
-universe\cite{jenness2018lsst}.
+universe \cite{jenness2018lsst}.
 % - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);

### 2020-02-05 body.tex
Update references

@@
 Python, written in C, and linking to standard fast implementations of linear
-algebra.
+algebra \cite{Numericmanual}.
 Around 2000, the Space Telescope Science Institute (STScI) software group wrote
@@
 on large memory-mapped arrays and arrays of mixed data type
-records \cite{STScI-slither}.
+records \cite{greenfield2003numarray}.
 This briefly caused the two communities to diverge, until

### 2020-02-05 body.tex
Acknowledge Numeric/Numarray developers

@@
 Around 2000, the Space Telescope Science Institute (STScI) software group wrote
-a reimplementation of much of Numeric, called NumArray, to support their work
+a reimplementation of Numeric, called NumArray, to support their work
 on large memory-mapped arrays and arrays of mixed data type

### 2020-02-05 paper.tex
Move figure
Minor edits
Acknowledge Numeric/Numarray developers

--- 2020-01-11 paper.tex
@@
 
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd{\@maketitle}
- {\noindent
-{\parbox{\dimexpr\linewidth-2\fboxsep\relax}{\color{color1}\large\sffamily\textbf{ABSTRACT}}}}
- {\noindent
-{\parbox{\dimexpr\linewidth-2\fboxsep\relax}{\color{color1}\large\sffamily\textbf{ABSTRACT}}}}
- {}{}
-\makeatother
-
 \usepackage[switch]{lineno}
@@
 Grant 2013-10-27 to the University of California, Berkeley.
-S.J.v.d.W, S.B., T.J.R., and W.W. were funded in part by the Gordon
+S.J.v.d.W, S.B., M.P., and W.W. were funded in part by the Gordon
 and Betty Moore Foundation through Grant GBMF5447 and by the Alfred
@@
 
+Ross Barnowski and Michael Eickenberg suggested text and provided 
+helpful feedback.
+
+We would like to thank Jim Hugunin, Paul Dubois, Travis Oliphant, David Ascher,
+Tim Peters, Konrad Hinsen, and other Numeric developers.
+We would also like to thank Perry Greenfield, Jay Todd Miller, JT Hsu, Richard
+L White, Jochen Krupper, Phil Hodge, and other Numarray developers.
+
+Finally, we would like to thank the many members of the community who have provided
+feedback, submitted bug reports, made minor improvements to the documentation,
+code, or website, promoted its use in their scientific fields, and built
+the vast ecosystem of tools and libraries around NumPy.
+
 \section*{Author Contributions Statement}
 
-%C.R.H and T.E.O.
-S.J.v.d.W and K.J.M. composed the manuscript with input from others.
-S.B., K.S., T.J.R., W.W., and M.B. drafted sections.
+K.J.M. and S.J.v.d.W composed the manuscript with input from others.
+S.B., R.G., K.S., W.W., M.B., and T.J.R. contributed text.
 All authors have contributed significant code, documentation, and/or expertise

### 2020-02-05 authors.tex
Update Eric's institution
Update Marten's initial
Minor updates

@@
 \author[11]{Stephan Hoyer}
-\author[12]{Marten van Kerkwijk}
+\author[12]{Marten H. van Kerkwijk}
 \author[13]{Matthew Brett}
@@
 \affil[1]{Independent Researcher, Logan, Utah, USA}
-\affil[2]{Berkeley Institute for Data Science, University of California, Berkeley, CA, USA}
-\affil[3]{Division of Biostatistics, University of California, Berkeley, CA, USA}
+\affil[2]{Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA, USA}
+\affil[3]{Division of Biostatistics, University of California, Berkeley, Berkeley, CA, USA}
 \affil[4]{Quansight LLC, Austin, TX, USA}
@@
 \affil[6]{Mercai JP, Tokyo, Japan}
-\affil[7]{University of Cambridge, Cambridge, UK}
-\affil[9]{Independent Researcher, Berkeley, CA, USA}
+\affil[7]{Department of Engineering, University of Cambridge, Cambridge, UK}
+\affil[9]{Independent Researcher, Berkeley, Berkeley, CA, USA}
 \affil[10]{Enthought, Inc., Austin, TX, USA}
@@
 \affil[22]{Los Alamos National Laboratory, Los Alamos, NM, USA}
-\affil[24]{Biomedical Engineering Department, University of California, Irvine, USA}
+\affil[24]{Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
 \affil[80]{Not sure}

### 2020-02-05 body.tex
Rework intro

--- 2020-02-04 body.tex
@@
 Two Python array packages existed before NumPy.
-The Numeric package began in the mid-90s and provided an array object in
-Python, written in C, and linking to standard fast implementations of linear
-algebra.
-Around 2000, the Space Telescope Science Institute (STScI) software group wrote
-a reimplementation of much of Numeric, called NumArray, to support their work
-on large memory-mapped arrays and arrays of mixed data type
-records \cite{STScI-slither}.
-This briefly caused the two communities to diverge, until
+The Numeric package began in the mid-1990s and provided an array object and array-aware functions
+in Python, written in C, and linking to standard fast implementations of linear
+algebra \cite{Numericmanual}.
+To handle large astronomical images coming from the Hubble Space Telescope,
+a reimplementation of Numeric, called NumArray, added
+support for memory-mapped arrays and mixed data type records. 
+%Around 2000, the Space Telescope Science Institute (STScI) software group wrote
+%a reimplementation of Numeric, called NumArray, to support their work
+%on large memory-mapped arrays and arrays of mixed data type
+%records \cite{greenfield2003numarray}.
+While NumArray was highly compatible with Numeric, the two packages
+had enough differences that it divided the community, until
 2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
@@
 Today, NumPy underpins almost every Python library that does scientific or
-numerical computation including SciPy \cite{virtanen2019scipy},
+numerical computation including SciPy \cite{virtanen2020scipy},
 Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
@@
 a language for array programming.
+To exploit new specialized hardware and storage, there has been
+a recent proliferation of Python array packages.
+Instead of leading back to some kind of Numarray and Numeric divide that
+NumPy was created to bridge, NumPy has become ...
 

### 2020-02-05 paper.tex
Update acknowledgments

--- 2020-01-11 paper.tex
@@
 
+Jim Hugunin wrote Numeric in 1995, while a graduate student at MIT.
+Hugunin based his package on previous work by Jim Fulton, then working at the
+US Geological Survey, with input from many others.
+After he graduated, Paul Dubois at the Lawrence Livermore National Laboratory
+became the maintainer.
+Many people contributed to the project including Travis Oliphant, David Ascher,
+Tim Peters, and Konrad Hinsen.
+
+Around 1998, the Space Telescope Science Institute (STScI) software group began
+to use Python heavily, and, around 2000, wrote a reimplementation of
+Numeric, called Numarray.
+%https://lancesimms.com/programs/Python/pyraf/Docs/greenfield.pdf
+Perry Greenfield created the original framework.
+Many people contributed to it including Jay Todd Miller, JT Hsu, Richard
+L White, Jochen Krupper, and Phil Hodge.
+
+Eric Jones co-founded the SciPy community, gave early feedback on array
+implementations, and provided funding and travel support to several community
+members.
+Numerous people contributed to the creation and growth of the larger SciPy
+ecosystem, which gives NumPy much of its value.  Others have injected new
+energy and ideas by creating experimental array packages.
+
+Finally, we would like to thank the many members of the community who have provided
+feedback, submitted bug reports, made improvements to the documentation,
+code, or website, promoted its use in their scientific fields, and built
+the vast ecosystem of tools and libraries around NumPy.
+Ross Barnowski and Michael Eickenberg suggested text and provided
+helpful feedback on the manuscript.
+
 S.J.v.d.W and K.J.M. were funded in part by the Gordon and Betty Moore
@@
 
-%C.R.H and T.E.O.
-S.J.v.d.W and K.J.M. composed the manuscript with input from others.
-S.B., K.S., T.J.R., W.W., and M.B. drafted sections.
+K.J.M. and S.J.v.d.W composed the manuscript with input from others.
+S.B., R.G., K.S., W.W., M.B., and T.J.R. contributed text.
 All authors have contributed significant code, documentation, and/or expertise

### 2020-02-05 body.tex
Minor edits

--- 2020-02-05 body.tex
@@
 
-Today, NumPy underpins almost every Python library that does scientific or
+Fifteen years later, NumPy now underpins almost every Python library that does scientific or
 numerical computation including SciPy \cite{virtanen2020scipy},

### 2020-02-05 body.tex
Polish end of intro

--- 2020-02-05 body.tex
@@
 
-Today, NumPy underpins almost every Python library that does scientific or
+Now, fifteen years later, NumPy underpins almost every Python library that does scientific or
 numerical computation including SciPy \cite{virtanen2020scipy},
@@
 the \emph{de facto} exchange format for array data in Python.
-The library has such widespread adoption that not only the array object but also its
-\emph{Application Programming Interface} (API) has become widespread as
-a language for array programming.
-To exploit new specialized hardware and storage, there has been
-a recent proliferation of Python array packages.
-Instead of leading back to some kind of Numarray and Numeric divide that
-NumPy was created to bridge, NumPy has become ...
+
+NumPy operates on in-memory arrays using the CPU. To utilize modern,
+specialized storage and hardware, there has been a recent
+proliferation of Python array packages. Unlike with the Numarray and
+Numeric divide, it is now much harder for these new libraries to
+fracture the user community---given how much work already builds
+on-top of NumPy.  However, to provide the ecosystem with access to
+new and exploratory technologies, NumPy is transitioning into a
+central coordinating mechanism that specifies a well-defined array
+programming \emph{Application Programming Interface} (API) and
+dispatches it, as appropriate, to specialized array implementations.
 

### 2020-02-06 static/sketches/fig1.py
Redraw Fig. 1 in Inkscape

@@ -1,41 +1,32 @@
-DejaVu Sans Mono
-Size: 62
+import numpy as np
+x = np.arange(12)
+x = x.reshape(4, 3)
+x
+np.mean(x, axis=0)
+x = x - np.mean(x, axis=0)
+x
 
-# Fig. 1
+# In [1]: import numpy as np
 
-In [1]: import numpy as np
+# In [2]: x = np.arange(12)
 
-In [2]: x = np.arange(12)
+# In [3]: x = x.reshape(4, 3)
 
-In [3]: x = x.reshape(4, 3)
+# In [4]: x
+# Out[4]:
+# array([[ 0,  1,  2],
+#        [ 3,  4,  5],
+#        [ 6,  7,  8],
+#        [ 9, 10, 11]])
 
-In [4]: x
-Out[4]:
-array([[ 0,  1,  2],
-       [ 3,  4,  5],
-       [ 6,  7,  8],
-       [ 9, 10, 11]])
+# In [5]: np.mean(x, axis=0)
+# Out[5]: array([4.5, 5.5, 6.5])
 
-In [5]: np.mean(x, axis=0)
-Out[5]: array([4.5, 5.5, 6.5])
+# In [6]: x = x - np.mean(x, axis=0)
 
-In [6]: x = x - np.mean(x, axis=0)
-
-In [7]: np.std(x, axis=0)
-Out[7]: array([3.35, 3.35, 3.35])
-
-In [8]: x = x / np.std(x, axis=0)
-
-In [9]: x
-Out[9]:
-array([[-1.34, -1.34, -1.34],
-       [-0.45, -0.45, -0.45],
-       [ 0.45,  0.45,  0.45],
-       [ 1.34,  1.34,  1.34]])
-
---------
-
-array([[-1.34, -1.34164079, -1.34164079],
-       [-0.45 , -0.4472136 , -0.4472136 ],
-       [ 0.45 ,  0.4472136 ,  0.4472136 ],
-       [ 1.34,  1.34164079,  1.34164079]])
+# In [7]: x
+# Out[7]:
+# array([[-4.5, -4.5, -4.5],
+#        [-1.5, -1.5, -1.5],
+#        [ 1.5,  1.5,  1.5],
+#        [ 4.5,  4.5,  4.5]])

### 2020-02-06 body.tex
Redraw Fig. 1 in Inkscape
Refine conceps sketch; update captions

@@
 support for memory-mapped arrays and mixed data type records. 
-%Around 2000, the Space Telescope Science Institute (STScI) software group wrote
-%a reimplementation of Numeric, called NumArray, to support their work
-%on large memory-mapped arrays and arrays of mixed data type
-%records \cite{greenfield2003numarray}.
 While NumArray was highly compatible with Numeric, the two packages
@@
 % KJM: I moved this here to get the placement right
+% FIXME SJW: let's review caption (d); I don't think it's crystal clear what
+% we're trying to show there; double check caption (g)
 \begin{figure*}[h]
   \centering
-  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
+  \includegraphics[width=\textwidth]{static/sketches/concepts}
   \caption{\textbf{Fundamental Array Concepts.}
     \textbf{a,} The NumPy array data structure and its associated metadata fields.
-    \textbf{b,} Indexing an array with various types of arguments.
-    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
-    \textbf{d,} Vectorization efficiently applies operations to groups of elements
-    \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
+    \textbf{b,} Indexing an array with slices and steps. These
+  operations return a \emph{view} of the original data.
+    \textbf{c,} Indexing an array with masks, scalar coordinates, or
+  other arrays, so that it returns a copy of the original data. In the
+  bottom example, an array is indexed with other arrays; this broadcasts the indexing arguments before performing the lookup.
+    \textbf{d,} Vectorization efficiently applies operations to groups
+  of elements.
+    \textbf{e,} Broadcasting in the multiplication of two-dimensional arrays.
     \textbf{f,} Reduction operations act along one or more axes. In this
-    example, a three-dimensional array is shown to be summed along various single
-    axes to produce two-dimensional results, or along two axes consecutively to
-    produce a one-dimensional result.
+    example, an array is be summed along various axes to produce a vector, or along two axes consecutively to
+    produce a scalar.
     \textbf{g,} Example NumPy code, illustrating some of these concepts.

### 2020-02-06 body.tex
Redraw fig. 3: protocol. Still need to redraw icons.
Minor edits

@@
 specialized storage and hardware, there has been a recent
-proliferation of Python array packages. Unlike with the Numarray and
+proliferation of Python array packages. Unlike with the NumArray and
 Numeric divide, it is now much harder for these new libraries to
@@
 
+% FIXME SJW: Redraw the icons in vector format
 \begin{figure}
@@
 artificial intelligence has further and dramatically boosted the usage of
-scientific Python.  Examples of its impactful application, such as the
+scientific Python.  Examples of its significant application, such as the
 above, now exist in almost every discipline in the natural and social
-sciences.  These tools have become \emph{the
-primary} software environment in many fields.  NumPy and its ecosystem are commonly
+sciences.  These tools have become \emph{the primary}
+software environment in many fields.  NumPy and its ecosystem are commonly
 taught in university courses, boot camps, and summer schools, and are
@@
 capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
-to allow for additional performance optimizations.  SciPy and Sparse both
+to allow for additional performance optimizations.  SciPy and PyData/Sparse both
 provide sparse arrays---which typically contain few non-zero values and store
@@
   \centering
-  \includegraphics[width=\textwidth]{static/sketches/array_protocol.jpg}
+  \includegraphics[width=\textwidth]{static/sketches/protocol}
 
@@
 which was free to use and unencumbered by license servers, dongles, and the like.
-There was a sense of building something
-impactful together, for the benefit of many others.  Participating in
+There was a sense of building something consequential together,
+for the benefit of many others.  Participating in
 such an endeavor, within a welcoming community of like-minded
@@
 
-NumPy is a community of practice with a strong \emph{culture} of
+NumPy is a community of practice with a \emph{culture} of
 employing software engineering practice to improve collaboration and

### 2020-02-06 paper.tex
Minor edits

@@
 to use Python heavily, and, around 2000, wrote a reimplementation of
-Numeric, called Numarray.
+Numeric, called NumArray.
 %https://lancesimms.com/programs/Python/pyraf/Docs/greenfield.pdf

### 2020-02-07 authors.tex
Update author affiliation
MAINT: Tyler revisions
* adjust name & affiliation

* fix an incorrect reference to the reduction
operation in Figure 1, which should be "f" not
"e"

* fix a singular reference to "scientist" that should
be plural

@@
 \author[21]{Kevin Sheppard}
-\author[22]{Tyler J. Reddy}
+\author[22]{Tyler Reddy}
 \author[2]{Warren Weckesser}
@@
 \affil[21]{Department of Economics, University of Oxford, Oxford, UK}
-\affil[22]{Los Alamos National Laboratory, Los Alamos, NM, USA}
-\affil[24]{Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
+\affil[22]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[23]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
 \affil[80]{Not sure}

### 2020-02-07 body.tex
Fix spelling of PyData/Sparse.
MAINT: Tyler revisions
* adjust name & affiliation

* fix an incorrect reference to the reduction
operation in Figure 1, which should be "f" not
"e"

* fix a singular reference to "scientist" that should
be plural

@@
 For example, summing an $n$-dimensional array over $d$ axes results in a
-$(n-d)$-dimensional array (Fig.~\ref{fig:array-concepts}e).
+$(n-d)$-dimensional array (Fig.~\ref{fig:array-concepts}f).
 
@@
 
-%CuPy, Sparse, and JAX the most prominent examples---while others have
+%CuPy, PyData/Sparse, and JAX the most prominent examples---while others have
 %NumPy-inspired semantics but, often due to their development history, deviate in places.
@@
 interacting with web resources, and parsing text.
-Adding fast array operations and linear algebra allows scientist to do all
+Adding fast array operations and linear algebra allows scientists to do all
 their work within a single language---and one that has the advantage of

### 2020-02-07 body.tex
Minor edits / feedback from Tyler

@@
 solving problems in specific domains.
-For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
+For example, the \code{eht-imaging} library \cite{chael2019ehtim} developed by the Event Horizon Telescope 
 collaboration for radio interferometry imaging, analysis, and simulation

### 2020-02-07 methods.tex
Fix spelling of PyData/Sparse.
Minor edits / feedback from Tyler

@@
 Python community infrastructure.
-The docstring standard mentioned in the history section is now widely adopted.
+NumPy's docstring standard is now widely adopted.
 We are also now using the NEP system as a way to help coordinate the larger
@@
 It is also implemented in packages that support sparse arrays
-such as \code{scipy.sparse} and \code{pydata.sparse}.
+such as \code{scipy.sparse} and PyData/Sparse.
 Another notable example is Dask, a library for parallel computing in

### 2020-02-07 body.tex
Minor edits / feedback from Marten and Perry

@@
 To handle large astronomical images coming from the Hubble Space Telescope,
-a reimplementation of Numeric, called NumArray, added
-support for memory-mapped arrays and mixed data type records. 
-While NumArray was highly compatible with Numeric, the two packages
+a reimplementation of Numeric, called Numarray, added
+support for structured arrays, flexible indexing, memory mapping, byte-order variants,
+more efficient memory use, flexible IEEE error handling capabilities, and
+better type casting rules \cite{greenfield2003numarray}.
+While Numarray was highly compatible with Numeric, the two packages
 had enough differences that it divided the community, until
 2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
-and NumArray \cite{oliphant2006guide}.
+and Numarray \cite{oliphant2006guide}---combining Numarray's features with Numeric's
+performance on small arrays and its rich C \emph{Application Programming Interface} (API).
 
@@
 specialized storage and hardware, there has been a recent
-proliferation of Python array packages. Unlike with the NumArray and
+proliferation of Python array packages. Unlike with the Numarray and
 Numeric divide, it is now much harder for these new libraries to
@@
 central coordinating mechanism that specifies a well-defined array
-programming \emph{Application Programming Interface} (API) and
+programming API and
 dispatches it, as appropriate, to specialized array implementations.
@@
 NumPy provides in-memory, multidimensional, homogeneously typed
-arrays on CPUs.  It runs on machines
+(i.e., single pointer and strided) arrays on CPUs.  It runs on machines
 ranging from embedded devices to the world's largest supercomputers,
@@
 stable array programming interface. This, in turn, prevents disruptive
-schisms like the divergence of Numeric and NumArray.
+schisms like the divergence of Numeric and Numarray.
 But exploring new ways of working with arrays is experimental by nature

### 2020-02-07 paper.tex
Minor edits / feedback from Marten and Perry

--- 2020-02-05 paper.tex
@@
 
-S.J.v.d.W and K.J.M. were funded in part by the Gordon and Betty Moore
+We thank Ross Barnowski, Michael Eickenberg, and Perry Greenfield, who
+suggested text and provided helpful feedback on the manuscript.
+
+We also thank the many members of the community who provided
+feedback, submitted bug reports, made improvements to the documentation,
+code, or website, promoted NumPy's use in their scientific fields, and built
+the vast ecosystem of tools and libraries around NumPy.
+We also gratefully acknowledge the Numeric and Numarray developers
+on whose work we built.
+
+Jim Hugunin wrote Numeric in 1995, while a graduate student at MIT.
+Hugunin based his package on previous work by Jim Fulton, then working at the
+US Geological Survey, with input from many others.
+After he graduated, Paul Dubois at the Lawrence Livermore National Laboratory
+became the maintainer.
+Many people contributed to the project including Travis Oliphant (a co-author
+of this paper), David Ascher, Tim Peters, and Konrad Hinsen.
+
+In 1998 the Space Telescope Science Institute started using Python
+and in 2000 began developing a new array package called Numarray, written
+almost entirely by Jay Todd Miller, starting from a prototype developed by
+Perry Greenfield.  Other contributors included Richard L. White, J. C. Hsu,
+Jochen Krupper, and Phil Hodge.
+The Numeric/Numarray split divided the community, yet ultimately pushed
+progress much further and faster than would otherwise have been possible.
+
+Shortly after Numarray development started, T.E.O. took over maintenance of
+Numeric. In 2005, he led the effort and did most of the work to unify Numeric
+and Numarray, and produce the first version of NumPy.
+
+Eric Jones co-founded (along with T.E.O. and P.P.) the SciPy community, gave early feedback on array
+implementations, and provided funding and travel support to several
+community members.
+Numerous people contributed to the creation and
+growth of the larger SciPy ecosystem, which gives NumPy much of its
+value. Others injected new energy and ideas by creating experimental
+array packages.
+
+K.J.M. and S.J.v.d.W. were funded in part by the Gordon and Betty Moore
 Foundation through Grant GBMF3834 and by the Alfred P. Sloan Foundation through
 Grant 2013-10-27 to the University of California, Berkeley.
-S.J.v.d.W, S.B., M.P., and W.W. were funded in part by the Gordon
+S.J.v.d.W., S.B., M.P., and W.W. were funded in part by the Gordon
 and Betty Moore Foundation through Grant GBMF5447 and by the Alfred
@@
 
-Ross Barnowski and Michael Eickenberg suggested text and provided 
-helpful feedback.
-
-We would like to thank Jim Hugunin, Paul Dubois, Travis Oliphant, David Ascher,
-Tim Peters, Konrad Hinsen, and other Numeric developers.
-We would also like to thank Perry Greenfield, Jay Todd Miller, JT Hsu, Richard
-L White, Jochen Krupper, Phil Hodge, and other Numarray developers.
-
-Finally, we would like to thank the many members of the community who have provided
-feedback, submitted bug reports, made minor improvements to the documentation,
-code, or website, promoted its use in their scientific fields, and built
-the vast ecosystem of tools and libraries around NumPy.
-
 \section*{Author Contributions Statement}
 
-K.J.M. and S.J.v.d.W composed the manuscript with input from others.
+K.J.M. and S.J.v.d.W. composed the manuscript with input from others.
 S.B., R.G., K.S., W.W., M.B., and T.J.R. contributed text.

### 2020-02-08 body.tex
Move figure
Improve caption

@@
-Two Python array packages existed before NumPy.
-The Numeric package began in the mid-1990s and provided an array object and array-aware functions
-in Python, written in C, and linking to standard fast implementations of linear
-algebra \cite{Numericmanual}.
-To handle large astronomical images coming from the Hubble Space Telescope,
-a reimplementation of Numeric, called Numarray, added
-support for structured arrays, flexible indexing, memory mapping, byte-order variants,
-more efficient memory use, flexible IEEE error handling capabilities, and
-better type casting rules \cite{greenfield2003numarray}.
-While Numarray was highly compatible with Numeric, the two packages
-had enough differences that it divided the community, until
-2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
-and Numarray \cite{oliphant2006guide}---combining Numarray's features with Numeric's
-performance on small arrays and its rich C \emph{Application Programming Interface} (API).
-
 % KJM: I moved this here to get the placement right
@@
 \end{figure*}
+
+Two Python array packages existed before NumPy.
+The Numeric package began in the mid-1990s and provided an array object and array-aware functions
+in Python, written in C, and linking to standard fast implementations of linear
+algebra \cite{Numericmanual}.
+To handle large astronomical images coming from the Hubble Space Telescope,
+a reimplementation of Numeric, called Numarray, added
+support for structured arrays, flexible indexing, memory mapping, byte-order variants,
+more efficient memory use, flexible IEEE error handling capabilities, and
+better type casting rules \cite{greenfield2003numarray}.
+While Numarray was highly compatible with Numeric, the two packages
+had enough differences that it divided the community, until
+2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
+and Numarray \cite{oliphant2006guide}---combining Numarray's features with Numeric's
+performance on small arrays and its rich C \emph{Application Programming Interface} (API).
 
@@
   \caption{\textbf{Scientific Python Ecosystem.}
-   This figure highlights some of the essential tools and libraries, which depend
+   This figure highlights (in blue) some of the essential tools and libraries that depend
    on NumPy.  Since most of the ecosystem
    builds on NumPy arrays and its rich and powerful collection of array-aware
-   functionality, the simplest way for other array implementations to become
+   functionality, the simplest way for new array implementations to become
    available to the vast scientific Python ecosystem is by using the NumPy API and array

### 2020-02-08 scripts/plot_merged_prs.py
Update and tweak PR merge plot

@@ -107,8 +107,8 @@ if True:
 
 prs = prs.set_index("merged_at")
 
-plt.figure(figsize=(6, 4))
-plt.ylabel("number of merged PR per quarter")
+plt.figure(figsize=(6*0.75, 4*0.75))
+plt.ylabel("PRs merged each quarter")
 
 
 def find_category(row):
@@ -136,12 +136,13 @@ groups = {}
 for g, comm in prs.groupby("category"):
     resampled = comm["category"].resample(
             # Resample starting at each quarter and shift by about 1/2 quarter
-            "QS", loffset=datetime.timedelta(days=30*3/2)).count()
+            "QS", loffset=0 * datetime.timedelta(days=30*3/2)).count()
     groups[g] = resampled
 
 categories = ["Maintainer", "Other"]
 stacked = []
-for color, cat in zip(["C0", "C1"], categories):
+# other color:  "#0e50a8", "#077a0a" OR   "#cce0fb", "#71f775"
+for color, cat in zip(["#0e50a8aa", "#077a0aaa"], categories):
     next_stack = groups[cat]
 
     if len(stacked) > 0:
@@ -153,20 +154,22 @@ for color, cat in zip(["C0", "C1"], categories):
     stacked.append(next_stack)
 
     plt.fill_between(
-        next_stack.index, next_stack.values, prev_vals,
+        next_stack.index.repeat(2) + np.array([0, datetime.timedelta(days=30*3)] * len(next_stack.index), dtype="m8[D]"),
+        next_stack.values.repeat(2), prev_vals.repeat(2),
         label=cat, lw=0, zorder=4 if cat == "Community" else 3,
-        color=color, alpha=0.8)
-    plt.plot(
-        next_stack.index, next_stack.values, "o",
-        zorder=4 if cat == "Community" else 3, color=color,
-        mec="k")
+        color=color)
+    if False:
+        plt.plot(
+            next_stack.index + datetime.timedelta(days=30*3/2), next_stack.values, "o",
+            zorder=4 if cat == "Community" else 3, color="none",
+            mec="k", ms=4)
 
     print(f"Total number for {cat}:", (next_stack.values - prev_vals).sum())
 
-plt.xlim(resampled.index.min(), resampled.index[-2])
+plt.xlim(resampled.index.min(), resampled.index[-1])
 plt.ylim(0, None)
 
-plt.legend()
+plt.legend(loc="upper left")
 
 plt.tight_layout()
 

### 2020-02-10 body.tex
Feedback from Paul and Ross

@@
 in Python, written in C, and linking to standard fast implementations of linear
-algebra \cite{Numericmanual}.
+algebra \cite{dubois1996numerical,Numericmanual}.
 To handle large astronomical images coming from the Hubble Space Telescope,
@@
 
+% FIXME SJW: Redraw the icons in vector format
+\begin{figure}
+  \centering
+  \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
+  \caption{\textbf{Scientific Python Ecosystem.}
+   This figure highlights (in blue) some of the essential tools, projects, and libraries that depend
+   on NumPy.  Since most of the ecosystem
+   builds on NumPy arrays and its rich and powerful collection of array-aware
+   functionality, the simplest way for new array implementations to become
+   available to the vast scientific Python ecosystem is by using the NumPy API and array
+   protocols (Fig.~\ref{fig:array-protocol}).
+  }
+  \label{fig:ecosystem}
+\end{figure}
+
 \section*{Scientific Python ecosystem}
@@
 The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
-this foundation to provide several, widely used \emph{technique-specific}
+this foundation to provide several, widely used \emph{technique specific}
 libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
-that in turn underlay numerous \emph{domain-specific} projects
+that in turn underlay numerous \emph{domain specific} projects
 \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
@@
 
-The interactive environment created by this array programming
-foundation---along with the surrounding ecosystem of tools---inside of
-IPython or Jupyter, is ideally suited to exploratory data analysis.
-Users fluidly inspect, manipulate, and visualize their data, and
-rapidly iterate to refine programming statements. These statements are
-then stitched together into imperative or functional programs, or
-notebooks containing both computation and narrative.
-Scientific computing beyond exploratory work is often done in a text editor
-or an integrated development environment (IDEs) such as Spyder.
-This rich and productive environment has made Python popular
-for scientific research.
-
-% FIXME SJW: Redraw the icons in vector format
-\begin{figure}
-  \centering
-  \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
-  \caption{\textbf{Scientific Python Ecosystem.}
-   This figure highlights (in blue) some of the essential tools and libraries that depend
-   on NumPy.  Since most of the ecosystem
-   builds on NumPy arrays and its rich and powerful collection of array-aware
-   functionality, the simplest way for new array implementations to become
-   available to the vast scientific Python ecosystem is by using the NumPy API and array
-   protocols (Fig.~\ref{fig:array-protocol}).
-  }
-  \label{fig:ecosystem}
-\end{figure}
-
 Many research groups have designed large,
-complex scientific libraries, which add specialized functionality for
-solving problems in specific domains.
-For example, the \code{eht-imaging} library \cite{chael2019ehtim} developed by the Event Horizon Telescope 
-collaboration for radio interferometry imaging, analysis, and simulation
-relies on many components of the scientific Python ecosystem.
+complex scientific libraries, which add \emph{application specific} functionality
+to the ecosystem.
+For example, the \code{eht-imaging} library \cite{chael2019ehtim} developed by
+the Event Horizon Telescope collaboration for radio interferometry imaging,
+analysis, and simulation relies on many lower-level components of the scientific Python
+ecosystem.
 NumPy arrays are used to store and manipulate numerical data at every step
@@
 
+The interactive environment created by this array programming
+foundation---along with the surrounding ecosystem of tools---inside of
+IPython or Jupyter, is ideally suited to exploratory data analysis.
+Users fluidly inspect, manipulate, and visualize their data, and
+rapidly iterate to refine programming statements. These statements are
+then stitched together into imperative or functional programs, or
+notebooks containing both computation and narrative.
+Scientific computing beyond exploratory work is often done in a text editor
+or an integrated development environment (IDEs) such as Spyder.
+This rich and productive environment has made Python popular
+for scientific research.
+
 The rapid recent growth of data science, machine learning, and
@@
 scientific Python.  Examples of its significant application, such as the
-above, now exist in almost every discipline in the natural and social
+\code{eht-imaging} library, now exist in almost every discipline in the natural and social
 sciences.  These tools have become \emph{the primary}
@@
 To them, the advantages of an interactive, high-level array library
-were evident. The design of this new tool was informed by their
-experiences with powerful interactive programming languages for
-scientific computing such as APL \cite{iverson1962programming} and
-Yorick \cite{munro1995using}, as well as commercial languages and
-environments like IDL and {MATLAB}.
+were evident. The design of this new tool was informed by other
+powerful interactive programming languages for scientific computing
+such as Basis \cite{dubois1989basis}, Yorick \cite{munro1995using}, R \cite{ihaka1996r},
+and APL \cite{iverson1962programming},
+as well as commercial languages and environments like IDL and {MATLAB}.
 
@@
 It builds on decades of work refining the array programming
-syntax from APL through Yorick and others.
+syntax from APL through Basis, Yorick, and others.
 It also made it easy for other libraries to develop fast and

### 2020-02-10 paper.tex
Feedback from Paul and Ross
Update Fig. 2

@@
 
-We thank Ross Barnowski, Michael Eickenberg, and Perry Greenfield, who
+We thank Ross Barnowski, Paul Dubois, Michael Eickenberg, and Perry Greenfield, who
 suggested text and provided helpful feedback on the manuscript.
@@
 became the maintainer.
-Many people contributed to the project including Travis Oliphant (a co-author
+Many people contributed to the project including T.E.O. (a co-author
 of this paper), David Ascher, Tim Peters, and Konrad Hinsen.

### 2020-02-12 body.tex
Minor edits mostly based on PV feedback

@@
 
+Python is an open-source, general-purpose, interpreted programming language
+well-suited to standard programming tasks such as cleaning data,
+interacting with web resources, and parsing text.
+Adding fast array operations and linear algebra allows scientists to do all
+their work within a single language---and one that has the advantage of
+being famously easy to learn and teach.
+In fact, Python's adoption as a primary learning language in universities
+has further popularized NumPy as a tool of choice for modern data science.
+
+Even though NumPy is not part of Python's standard library,
+it benefits from a good relationship with the Python developers.
+Over the years, the Python language has added new features and
+special syntax so that NumPy would have a more succinct and 
+easier to read array notation.
+Since it is not part of the standard library, NumPy is able to
+dictate its own release policies and development patterns.
+
 SciPy and Matplotlib are tightly coupled with NumPy---in terms of
@@
 compilers and computing platforms.
+% FIXME: what are "specialized compilers and computing
+% platforms"? Does Fortran need to be mentioned here?
 
@@
 for scientific research.
+
+NumPy is a community of practice with a culture of
+employing software engineering practice to improve collaboration and
+reduce error \cite{millman2014developing}.  This culture is not only
+adopted by leaders in the project but also enthusiastically taught to
+newcomers. The NumPy team was early in adopting distributed revision
+control and code review to improve collaboration on code, and
+continuous testing that runs an extensive battery of automated tests for
+every proposed change to NumPy.  The project has comprehensive,
+high-quality documentation, integrated with the source
+code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
+
+% https://ras.ac.uk/sites/default/files/2020-01/Group%20Award%20-%20Astropy.pdf
+% https://ras.ac.uk/news-and-press/news/leading-astronomers-and-geophysicists-honoured-ras-bicentenary-year-0
+
+This culture of using best practices for producing reliable scientific software
+has been adopted by the ecosystem of libraries that build on NumPy.
+For example, in a recent award given by the Royal Astronomical Society to
+Astropy, they state:
+\begin{quotation}
+\noindent\emph{The Astropy Project has provided hundreds of junior scientists
+with experience in professional-standard software development practices
+including use of version control, unit testing, code review and issue tracking
+procedures. This is a vital skill set for modern researchers that is often
+missing from formal university education in physics or astronomy.}
+\end{quotation}
+Community members explicitly work to address this lack of formal education
+through courses and workshops
+\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
 
@@
 NumPy, in turn, dispatches operations to the originating library, as required.
+Currently around 50 or so functions, which are the most commonly
+used functions, support these protocols.
+This list is being refined as we gain more experience with how it works in
+practice.
 
 Thanks to these developments, users can now, for example, scale
-their computation from a single machine to distributed systems using Dask. The protocols
-\textit{compose} well, and so it is possible to redeploy NumPy code onto multi-GPU systems with minimal changes\cite{entschev2019}.
+their computation from a single machine to distributed systems using Dask.
+The protocols \textit{compose} well, and so it is possible to redeploy NumPy
+code onto multi-GPU systems with minimal changes\cite{entschev2019}.
 
@@
 
-NumPy was initially developed by students, faculty, and researchers to
+NumPy combines the expressive power of \emph{array programming}, 
+the performance of C, and
+the readability, usability, and versatility of Python in a mature,
+well-tested, well-documented, and community-developed library.
+%It builds on decades of work refining the array programming
+%syntax from APL through Basis, Yorick, and others.
+It also made it easy for other libraries to develop fast and
+memory-efficient compiled code, usually in C or Fortran, that could manipulate
+these arrays and pass them back to Python.
+
+It was initially developed by students, faculty, and researchers to
 provide an advanced, open-source array programming library for Python,
@@
 heralded the current era of widespread use of Python for scientific computing.
-The reasons for its popularity are varied and complex but include the facts that
-NumPy (a) is built on top of Python,
-(b) has a culture of collaboration and good software development practices, and
-(c) provides a cutting-edge array programming environment for scientific data analysis.
-
-\emph{Python} is an open-source, general-purpose, interpreted programming language
-well-suited to standard programming tasks such as cleaning data,
-interacting with web resources, and parsing text.
-Adding fast array operations and linear algebra allows scientists to do all
-their work within a single language---and one that has the advantage of
-being famously easy to learn and teach.
-In fact, Python's adoption as a primary learning language in universities
-has further popularized NumPy as a tool of choice for modern data science.
-
-Even though NumPy is not part of Python's standard library,
-it benefits from a good relationship with the Python developers.
-Over the years, the Python language has added new features and
-special syntax so that NumPy would have a more succinct and 
-easier to read array notation.
-Since it is not part of the standard library, NumPy is able to
-dictate its own release policies and development patterns.
-
-NumPy is a community of practice with a \emph{culture} of
-employing software engineering practice to improve collaboration and
-reduce error \cite{millman2014developing}.  This culture is not only
-adopted by leaders in the project but also enthusiastically taught to
-newcomers. The NumPy team was early in adopting distributed revision
-control and code review to improve collaboration on code, and
-continuous testing that runs an extensive battery of automated tests for
-every proposed change to NumPy.  The project has comprehensive,
-high-quality documentation, integrated with the source
-code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
-
-% https://ras.ac.uk/sites/default/files/2020-01/Group%20Award%20-%20Astropy.pdf
-% https://ras.ac.uk/news-and-press/news/leading-astronomers-and-geophysicists-honoured-ras-bicentenary-year-0
-
-This culture of using best practices for producing reliable scientific software
-has been adopted by the ecosystem of libraries that build on NumPy.
-For example, in a recent award given by the Royal Astronomical Society to
-Astropy, they state:
-\begin{quotation}
-\noindent\emph{The Astropy Project has provided hundreds of junior scientists
-with experience in professional-standard software development practices
-including use of version control, unit testing, code review and issue tracking
-procedures. This is a vital skill set for modern researchers that is often
-missing from formal university education in physics or astronomy.}
-\end{quotation}
-Community members explicitly work to address this lack of formal education
-through courses and workshops
-\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
-
-NumPy combines the expressive power of \emph{array programming}, 
-the performance of C, and
-the readability, usability, and versatility of Python in a mature,
-well-tested, well-documented, and community-developed library.
-It builds on decades of work refining the array programming
-syntax from APL through Basis, Yorick, and others.
-It also made it easy for other libraries to develop fast and
-memory-efficient compiled code, usually in C or Fortran, that could manipulate
-these arrays and pass them back to Python.
+%The reasons for its popularity are varied and complex but include the facts that
+%NumPy (a) is built on top of Python,
+%(b) has a culture of collaboration and good software development practices, and
+%(c) provides a cutting-edge array programming environment for scientific data analysis.
 

### 2020-02-12 summary.tex
Minor edits mostly based on PV feedback

--- 2020-02-04 summary.tex
@@
 material science, engineering, finance, and economics.
-For example, in astronomy, NumPy was fundamental to discovering gravitational
-waves\cite{pycbc}, the first imaging of a black hole\cite{eht-imaging}, and
-continues underpinning the exploration of our
-universe\cite{jenness2018lsst}.
+For example, in astronomy, NumPy is an important part of the data analysis
+pipeline used in the discovery of gravitational
+waves \cite{abbott2016observation}, the first imaging of a black hole \cite{eht-imaging}, and
+the continuing exploration of our universe \cite{jenness2018lsst}.
 % - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);

### 2020-02-13 authors.tex
Fix typo in David Cournapeau affiliation
Update AH affil
Update JT affil
Update PGM name and affil
Update MW affil

--- 2020-02-05 authors.tex
@@
 \author[7]{Eric Wieser}
-\author[80]{Julian Taylor}
+\author[8]{Julian Taylor}
 \author[2]{Sebastian Berg}
@@
 \author[16]{Mark Wiebe}
-\author[4,17,18]{Pearu Peterson}
+\author[4,18,19]{Pearu Peterson}
 \author[80]{David M. Cooke}
-\author[20]{Pierre Gerard-Marchant}
-\author[21]{Kevin Sheppard}
-\author[22]{Tyler J. Reddy}
+\author[21]{Pierre G\'erard-Marchant}
+\author[22]{Kevin Sheppard}
+\author[23]{Tyler Reddy}
 \author[2]{Warren Weckesser}
 \author[4]{Hameer Abbasi}
-\author[23]{Christoph Gohlke}
+\author[24]{Christoph Gohlke}
 \author[4]{Travis E. Oliphant}
@@
 \affil[5]{Department of Physics and Nanoscience Center, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
-\affil[6]{Mercai JP, Tokyo, Japan}
+\affil[6]{Mercari JP, Tokyo, Japan}
 \affil[7]{Department of Engineering, University of Cambridge, Cambridge, UK}
-\affil[9]{Independent Researcher, Berkeley, Berkeley, CA, USA}
+\affil[8]{Independent Researcher, Karlsruhe, Germany}
+\affil[9]{Independent Researcher, Berkeley, CA, USA}
 \affil[10]{Enthought, Inc., Austin, TX, USA}
@@
 \affil[13]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
-\affil[14]{Department of Chemistry, Temple University, Philadelphia, PA, USA}
+\affil[14]{Department of Physics, Temple University, Philadelphia, PA, USA}
 \affil[15]{Google, Zurich, Switzerland}
-\affil[16]{AWS Thinkbox, Seattle, Washington, USA}
-\affil[17]{Independent Researcher, Saue, Estonia}
-\affil[18]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
-\affil[20]{France-IX, France}
-\affil[21]{Department of Economics, University of Oxford, Oxford, UK}
-\affil[22]{Los Alamos National Laboratory, Los Alamos, NM, USA}
-\affil[24]{Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
+\affil[16]{Amazon, Seattle, Washington, USA}
+\affil[18]{Independent Researcher, Saue, Estonia}
+\affil[19]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
+\affil[21]{France-IX Services, Paris, France}
+\affil[22]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[23]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[24]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
 \affil[80]{Not sure}

### 2020-02-13 authors.tex
Add script to generate auth/affil file

@@
-\author[1]{Charles R Harris}
+\author[1]{Charles R. Harris}
 \author[2,3,*]{K. Jarrod Millman}
@@
 \author[16]{Mark Wiebe}
-\author[4,18,19]{Pearu Peterson}
-\author[80]{David M. Cooke}
-\author[21]{Pierre G\'erard-Marchant}
-\author[22]{Kevin Sheppard}
-\author[23]{Tyler Reddy}
+\author[4,17,18]{Pearu Peterson}
+\author[19]{Pierre G\'erard-Marchant}
+\author[20]{Kevin Sheppard}
+\author[21]{Tyler Reddy}
 \author[2]{Warren Weckesser}
 \author[4]{Hameer Abbasi}
-\author[24]{Christoph Gohlke}
+\author[22]{Christoph Gohlke}
 \author[4]{Travis E. Oliphant}
@@
 \affil[16]{Amazon, Seattle, Washington, USA}
-\affil[18]{Independent Researcher, Saue, Estonia}
-\affil[19]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
-\affil[21]{France-IX Services, Paris, France}
-\affil[22]{Department of Economics, University of Oxford, Oxford, UK}
-\affil[23]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
-\affil[24]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
-\affil[80]{Not sure}
+\affil[17]{Independent Researcher, Saue, Estonia}
+\affil[18]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
+\affil[19]{France-IX Services, Paris, France}
+\affil[20]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[21]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[22]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
 \affil[*]{numpy.articles@gmail.com}

### 2020-02-13 body.tex
Add script to generate auth/affil file

@@
 algebra \cite{dubois1996numerical,Numericmanual}.
+% FIXME steering physics codes at llnl
 To handle large astronomical images coming from the Hubble Space Telescope,
@@
 
+% this is the software engineering side of things
+% as opposed to the interactive, exploration described above
 NumPy is a community of practice with a culture of
@@
 ranging from embedded devices to the world's largest supercomputers,
-with performance approaching that of compiled languages.  For
-most its existence, NumPy addressed the vast majority of
+with performance approaching that of compiled languages.
+For most its existence, NumPy addressed the vast majority of
 array computation use cases.
-
-%For algorithms that do not fit the array programming approach,
-%or require even faster implementations, a variety of tools exist (e.g., Cython, Numba, Pythran)
-%that transparently accelerate bottlenecks.
-%These tools allow programmers to express concepts using high-level array programming without sacrificing efficiency.
-%In many cases, fast algorithms are already implemented by libraries in the scientific Python ecosystem;
-%these primitives simplify writing code in pure Python.
-
-% a brief word about accelerators perhaps?
-% end with how this fits the bill for a huge number of users
 
@@
 %syntax from APL through Basis, Yorick, and others.
-It also made it easy for other libraries to develop fast and
+It also makes it easy for other libraries to develop fast and
 memory-efficient compiled code, usually in C or Fortran, that could manipulate
 these arrays and pass them back to Python.
+For users that require faster implementations but do not want to write lower-level
+compiled code, a variety of tools exist
+(e.g., Cython, Numba, Pythran) that transparently accelerate bottlenecks.
+These tools allow programmers to express concepts using high-level array
+programming without sacrificing efficiency.
+In many cases, fast algorithms are already implemented by libraries in the
+scientific Python ecosystem; these primitives simplify writing code in pure
+Python.
 
-It was initially developed by students, faculty, and researchers to
+NumPy was initially developed by students, faculty, and researchers to
 provide an advanced, open-source array programming library for Python,

### 2020-02-13 mkauth.py

@@ -1,0 +1,30 @@
+import yaml
+
+with open("authors.yml") as f:
+    d = yaml.load(f, Loader=yaml.FullLoader)
+
+auths = d["auth"]
+
+star = len(d["inst"])
+affil_dict = {}
+affil_print = []
+for n, a in enumerate(d["inst"]):
+    for k, v in a.items():
+        num = n + 1
+        if num == star:
+            num = "*"
+        affil_dict.update({k: num})
+        affil_print.append(f"\\affil[{num}]{{{v}}}")
+
+with open("authors.tex", "w") as f:
+    for auth in auths:
+        for name, i in auth.items():
+            insts = ",".join(
+                [str(item) for item in [affil_dict[ins] for ins in i["institute"]]]
+            )
+            insts = f"[{insts}]"
+            f.write(f"\\author{insts}{{{name}}}")
+            f.write("\n")
+    for affil in affil_print:
+        f.write(affil)
+        f.write("\n")

### 2020-02-13 summary.tex
Tweak summary
- Clarify that NumPy was part of a pipeline for various scientific
discoveries
- Rephrase over-excited last sentence
- Remove mention of LSST and focus on past examples

--- 2020-02-04 summary.tex
@@
 material science, engineering, finance, and economics.
-For example, in astronomy, NumPy was fundamental to discovering gravitational
-waves\cite{pycbc}, the first imaging of a black hole\cite{eht-imaging}, and
-continues underpinning the exploration of our
-universe\cite{jenness2018lsst}.
+For example, in astronomy, NumPy was an important part of the software stack used
+in the discovery of gravitational waves \cite{abbott2016observation}
+and the first imaging of a black hole \cite{eht-imaging}.
 % - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
@@
 NumPy-like interfaces and array objects.  Because of its central position in the
-ecosystem, NumPy increasingly is used to bridge disparate
-technologies, opening new possibilities in previously unexplored computational frontiers.
-
-%Because of this, NumPy plays a unique role in enabling disparate technologies
-%to interact and exchange data.
-%NumPy continues to evolve to meet the needs of its community, to remain
-%relevant in an ever-changing computing landscape, and to be a fundamental
-%pillar of support for the next decade of scientific computing, data science,
-%and machine learning in Python.
+ecosystem, NumPy increasingly plays the role of an interoperability layer
+between these new array computation libraries.

### 2020-02-13 body.tex
Add sentence about computational steering
Refine protocols paragraph

@@
 algebra \cite{dubois1996numerical,Numericmanual}.
-% FIXME steering physics codes at llnl
+One of its earliest uses was to steer C++ applications for
+inertial confinement fusion research at
+Lawrence Livermore National Laboratory \cite{yang1997steering}.
 To handle large astronomical images coming from the Hubble Space Telescope,
@@
 and so forth, as appropriate.
-To support array operations between external array objects, NumPy,
-therefore, added the capability to act as a central coordination
+To support array operations between external array objects, NumPy
+therefore added the capability to act as a central coordination
 mechanism with a well-specified API (Fig.~\ref{fig:ecosystem}).
@@
 To facilitate this \emph{interoperability}, NumPy provides
-``protocols'' (or contracts of operation), that allow for these arrays to be
+``protocols'' (or contracts of operation), that allow for specialized arrays to be
 passed to NumPy functions (Fig.~\ref{fig:array-protocol}).
 NumPy, in turn, dispatches operations to the originating library, as required.
-Currently around 50 or so functions, which are the most commonly
-used functions, support these protocols.
-This list is being refined as we gain more experience with how it works in
-practice.
-
+Over three hundred of the most popular
+NumPy functions are supported.  The protocols are implemented by at
+least five libraries, including Dask, CuPy, xarray, and PyData/Sparse.
 Thanks to these developments, users can now, for example, scale
@@
 code onto multi-GPU systems with minimal changes\cite{entschev2019}.
+Working with the community, we have identified several issues that
+prevent further adoption, and are refining interfaces to reduce
+impedance mismatch.
 

### 2020-02-13 body.tex
Edit ecosystem section

@@
 their work within a single language---and one that has the advantage of
-being famously easy to learn and teach.
-In fact, Python's adoption as a primary learning language in universities
-has further popularized NumPy as a tool of choice for modern data science.
+being famously easy to learn and teach, as witnessed by its adoption
+as a primary learning language in many universities.
 
@@
 NumPy, as the focal point for the ecosystem of array-aware libraries,
-sets documentation standards, and provides array testing infrastructure
-and build support for integrating with Fortran code and using specialized
-compilers and computing platforms.
-% FIXME: what are "specialized compilers and computing
-% platforms"? Does Fortran need to be mentioned here?
+sets documentation standards, provides array testing infrastructure,
+and adds build support for Fortran and other compilers.
 
@@
 the Event Horizon Telescope collaboration for radio interferometry imaging,
-analysis, and simulation relies on many lower-level components of the scientific Python
+analysis, and simulation, relies on many lower-level components of the scientific Python
 ecosystem.
@@
 reconstruction.
-SciPy is used to provide tools used in general image processing such as
+SciPy supplies tools for general image processing tasks such as
 filtering and image alignment, while scikit-image, an image processing
@@
 edge filters and Hough transforms.
-Other components of SciPy are used throughout the library, such as the
-\code{scipy.optimize} module for handling mathematical optimization tasks.
+The \code{scipy.optimize} module performs mathematical optimization.
 NetworkX \cite{SciPyProceedings_11}, a package for complex
-network analysis, is used for the analysis of consistency among image
-comparisons.
-The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
-is used throughout \code{eht-imaging}, primarily to handle standard
-astronomical file formats and time/coordinate transformations.
-Matplotlib is used for visualizing data throughout the analysis pipeline,
-including for the generation of the final image of the black hole.
+network analysis, is used to verify image comparison consistency.
+Astropy \cite{astropy:2013, astropy:2018} handles standard
+astronomical file formats and computes time/coordinate transformations.
+Matplotlib is used to visualize data and to generate the final image of the black hole.
 
-The interactive environment created by this array programming
-foundation---along with the surrounding ecosystem of tools---inside of
-IPython or Jupyter, is ideally suited to exploratory data analysis.
+The interactive environment created by the array programming
+foundation along with the surrounding ecosystem of tools---inside of
+IPython or Jupyter---is ideally suited to exploratory data analysis.
 Users fluidly inspect, manipulate, and visualize their data, and
@@
 % as opposed to the interactive, exploration described above
-NumPy is a community of practice with a culture of
-employing software engineering practice to improve collaboration and
+
+To complement this facility for exploratory work and rapid
+prototyping, NumPy has developed a culture of
+employing time-tested software engineering practices to improve collaboration and
 reduce error \cite{millman2014developing}.  This culture is not only
@@
 continuous testing that runs an extensive battery of automated tests for
-every proposed change to NumPy.  The project has comprehensive,
+every proposed change to NumPy.  The project also has comprehensive,
 high-quality documentation, integrated with the source

### 2020-02-14 body.tex
Edit discussion

@@
 NumPy, in turn, dispatches operations to the originating library, as required.
-Over three hundred of the most popular
+Over four hundred of the most popular
 NumPy functions are supported.  The protocols are implemented by at
@@
 well-tested, well-documented, and community-developed library.
-%It builds on decades of work refining the array programming
-%syntax from APL through Basis, Yorick, and others.
-It also makes it easy for other libraries to develop fast and
-memory-efficient compiled code, usually in C or Fortran, that could manipulate
-these arrays and pass them back to Python.
-For users that require faster implementations but do not want to write lower-level
-compiled code, a variety of tools exist
-(e.g., Cython, Numba, Pythran) that transparently accelerate bottlenecks.
-These tools allow programmers to express concepts using high-level array
-programming without sacrificing efficiency.
-In many cases, fast algorithms are already implemented by libraries in the
-scientific Python ecosystem; these primitives simplify writing code in pure
+Many fast algorithms are already implemented by libraries in the
+scientific Python ecosystem.
+Where extreme optimization is warranted, compiled languages such as Cython, Numba, Pythran, that
+extend Python and transparently accelerate bottlenecks, can be
+used.
+Because of NumPy's simple memory model, it is easy to write low-level, hand-optimized code, usually in C
+or Fortran, to manipulate NumPy arrays and pass them back to
 Python.
+Furthermore, using array protocols, it is possible to utilize the full
+spectrum of specialized hardware acceleration with minimal changes to
+existing code.
 
@@
 their or their colleagues' problems---often in low-level languages
-that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
+that precede Python, like Fortran \cite{dongarra2008netlib} and C.
 To them, the advantages of an interactive, high-level array library
@@
 It spurred the development of the broader scientific Python ecosystem and
-heralded the current era of widespread use of Python for scientific computing.
-%The reasons for its popularity are varied and complex but include the facts that
-%NumPy (a) is built on top of Python,
-%(b) has a culture of collaboration and good software development practices, and
-%(c) provides a cutting-edge array programming environment for scientific data analysis.
+heralded the current era of widespread use of Python for scientific
+computing. % This and the "What began as" sentence below are more or less interchangeable.
 
-NumPy began as an attempt to add an array object to Python and became
-the foundation of a vibrant ecosystem of tools.
-Now, a large amount of scientific work depends on NumPy being correct,
-fast, and stable.
-Work on NumPy continues, so that it may meet the needs of its users.  For instance, the NumPy data type system is currently being redesigned to make it
-easier to implement data types both in Python and C, and to better handle
-rich data types such as physical units \cite{astropy,Goldbaum2018},
-geometrical objects, and missing values.
-% Maybe add a sentence in the direction of:
-% We hope/expect, that this will allow the proliferation of rich data types
-% which in turn can propel the whole ecosystem forward? (too much :)?
-% As a part of other changes/developments...)
-% Discussed with Jarrod, if it fits to get back to protocols, it is fine
-% to call DTypes a (rich) protocol.
+%What began as an attempt to add an array object to Python became
+%the foundation of a vibrant ecosystem of tools.
+%Now, a large amount of scientific work depends on NumPy being correct,
+%fast, and stable.
+%%
+%%
+%% v TO ADD v
+%%
+%% Over the next decade, the computational landscape will change
+%% drastically.
+%% - AI / data science mature
+%% - hardware / Moore's law, specialized devices
+%% - more users (diversity + numbers)
+%% - scale / type of data; scientific data, lightsheet microscopes, LSST
+%% - new ideas come from other languages like Julia, Rust, ...; JIT improvements
+%% NumPy is poised to tackle these challenges, and work continues.
+%% - Major refactoring of code
+%% - Governance
+%% - Process (NEP, roadmap, how we handle issues on GH, etc.)
+%% - Protocols: will be extending that work
+%% - Funding (some recent evidence that that is improving → grants, or
+%% Moore/Sloan/CZI) CZI EOSS
+%% - Redesigning website and documentation to be more friendly, easy to
+%% navigate
+%% - Next generation of developers (grad students) to sign on
+%%
 
-NumPy's role in the ecosystem continues to evolve. It is no longer just the foundational array library underlying the
-scientific Python ecosystem but has become the standard API for tensor
-computation and a central coordinating mechanism between array types and
-technologies in Python.
+NumPy's is no longer just the foundational array library underlying
+the scientific Python ecosystem, but has also become the standard API
+for tensor computation and a central coordinating mechanism between
+array types and technologies in Python.

### 2020-02-14 body.tex
Minor edits including feedback from Ross

@@
 better type casting rules \cite{greenfield2003numarray}.
-While Numarray was highly compatible with Numeric, the two packages
-had enough differences that it divided the community, until
-2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
-and Numarray \cite{oliphant2006guide}---combining Numarray's features with Numeric's
-performance on small arrays and its rich C \emph{Application Programming Interface} (API).
+While Numarray was highly compatible with Numeric, the two packages had enough
+differences that it divided the community, until 2005, when NumPy emerged as a
+``best of both worlds'' unification \cite{oliphant2006guide}---combining
+Numarray's features with Numeric's performance on small arrays and its rich C
+\emph{Application Programming Interface} (API).
 
@@
 still producing results that appeal to intuition.
-A trivial example is the addition of a scalar value to an array, but it also
+A trivial example is the addition of a scalar value to an array, but broadcasting also
 generalizes to more complex examples such as scaling each column of an array
@@
 The community's efforts to fill this gap led to a
-proliferation of new array implementations. Each deep learning framework created
+proliferation of new array implementations. For example, each deep learning framework created
 its own arrays; PyTorch \cite{NEURIPS2019_9015},
@@
 scientific Python ecosystem.
-Where extreme optimization is warranted, compiled languages such as Cython, Numba, Pythran, that
+Where extreme optimization is warranted, compiled languages such as
+Cython \cite{behnel2011cython}, Numba \cite{Lam:2015:NLP:2833157.2833162},
+and Pythran \cite{guelton2015pythran}, that
 extend Python and transparently accelerate bottlenecks, can be

### 2020-02-14 methods.tex
Minor edits including feedback from Ross

@@
 This allows us to review code before integrating it as well as to run a
-large number of tests on the modified code to ensure a bug free and stable
-experience for our users.
+large number of tests on the modified code to ensure that the changes
+do not break expected behavior.
 
-We also use GitHub's issue tracking system to collect and triage bugs.
+We also use GitHub's issue tracking system to collect and triage problems and
+proposed improvements.
 

### 2020-02-14 body.tex
Edit summary

@@
 fracture the user community---given how much work already builds
-on-top of NumPy.  However, to provide the ecosystem with access to
+on top of NumPy.  However, to provide the ecosystem with access to
 new and exploratory technologies, NumPy is transitioning into a

### 2020-02-14 authors.tex
Update affil policy

--- 2020-02-07 authors.tex
@@
-\author[1]{Charles R Harris}
-\author[2,3,*]{K. Jarrod Millman}
-\author[2,*]{St\'efan J. van der Walt}
-\author[4,*]{Ralf Gommers}
-\author[5]{Pauli Virtanen}
-\author[6]{David Cournapeau}
-\author[7]{Eric Wieser}
-\author[80]{Julian Taylor}
-\author[2]{Sebastian Berg}
-\author[9]{Nathaniel J. Smith}
-\author[10]{Robert Kern}
-\author[2]{Matti Picus}
-\author[11]{Stephan Hoyer}
-\author[12]{Marten H. van Kerkwijk}
-\author[13]{Matthew Brett}
-\author[14]{Allan Haldane}
-\author[15]{Jaime Fernandez del Rio}
-\author[16]{Mark Wiebe}
-\author[4,17,18]{Pearu Peterson}
-\author[80]{David M. Cooke}
-\author[20]{Pierre Gerard-Marchant}
+\author[1]{Charles R. Harris}
+\author[2,3,4,*]{K. Jarrod Millman}
+\author[2,4,*]{St\'efan J. van der Walt}
+\author[5,*]{Ralf Gommers}
+\author[6]{Pauli Virtanen}
+\author[7]{David Cournapeau}
+\author[8]{Eric Wieser}
+\author[9]{Julian Taylor}
+\author[4]{Sebastian Berg}
+\author[10]{Nathaniel J. Smith}
+\author[11]{Robert Kern}
+\author[4]{Matti Picus}
+\author[12]{Stephan Hoyer}
+\author[13]{Marten H. van Kerkwijk}
+\author[2,14]{Matthew Brett}
+\author[15]{Allan Haldane}
+\author[16]{Jaime Fernandez del Rio}
+\author[17]{Mark Wiebe}
+\author[5,18,19]{Pearu Peterson}
+\author[20]{Pierre G\'erard-Marchant}
 \author[21]{Kevin Sheppard}
 \author[22]{Tyler Reddy}
-\author[2]{Warren Weckesser}
-\author[4]{Hameer Abbasi}
+\author[4]{Warren Weckesser}
+\author[5]{Hameer Abbasi}
 \author[23]{Christoph Gohlke}
-\author[4]{Travis E. Oliphant}
+\author[5]{Travis E. Oliphant}
 \affil[1]{Independent Researcher, Logan, Utah, USA}
-\affil[2]{Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA, USA}
+\affil[2]{Helen Wills Neuroscience Institute, University of California, Berkeley, Berkeley, CA, USA}
 \affil[3]{Division of Biostatistics, University of California, Berkeley, Berkeley, CA, USA}
-\affil[4]{Quansight LLC, Austin, TX, USA}
-\affil[5]{Department of Physics and Nanoscience Center, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
-\affil[6]{Mercai JP, Tokyo, Japan}
-\affil[7]{Department of Engineering, University of Cambridge, Cambridge, UK}
-\affil[9]{Independent Researcher, Berkeley, Berkeley, CA, USA}
-\affil[10]{Enthought, Inc., Austin, TX, USA}
-\affil[11]{Google Research, Mountain View, CA, USA}
-\affil[12]{Department of Astronomy \& Astrophysics, University of Toronto, Toronto, ON, Canada}
-\affil[13]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
-\affil[14]{Department of Chemistry, Temple University, Philadelphia, PA, USA}
-\affil[15]{Google, Zurich, Switzerland}
-\affil[16]{AWS Thinkbox, Seattle, Washington, USA}
-\affil[17]{Independent Researcher, Saue, Estonia}
-\affil[18]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
-\affil[20]{France-IX, France}
+\affil[4]{Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA, USA}
+\affil[5]{Quansight LLC, Austin, TX, USA}
+\affil[6]{Department of Physics and Nanoscience Center, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
+\affil[7]{Mercari JP, Tokyo, Japan}
+\affil[8]{Department of Engineering, University of Cambridge, Cambridge, UK}
+\affil[9]{Independent Researcher, Karlsruhe, Germany}
+\affil[10]{Independent Researcher, Berkeley, CA, USA}
+\affil[11]{Enthought, Inc., Austin, TX, USA}
+\affil[12]{Google Research, Mountain View, CA, USA}
+\affil[13]{Department of Astronomy \& Astrophysics, University of Toronto, Toronto, ON, Canada}
+\affil[14]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
+\affil[15]{Department of Physics, Temple University, Philadelphia, PA, USA}
+\affil[16]{Google, Zurich, Switzerland}
+\affil[17]{Amazon, Seattle, Washington, USA}
+\affil[18]{Independent Researcher, Saue, Estonia}
+\affil[19]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
+\affil[20]{France-IX Services, Paris, France}
 \affil[21]{Department of Economics, University of Oxford, Oxford, UK}
@@
 \affil[23]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
-\affil[80]{Not sure}
 \affil[*]{numpy.articles@gmail.com}

### 2020-02-14 body.tex
Minor edits

@@
 
-The rapid recent growth of data science, machine learning, and
+The recent rapid growth of data science, machine learning, and
 artificial intelligence has further and dramatically boosted the usage of
@@
 
-In recent years, there has been a rapid growth in the development of
-specialized hardware, including graphics processing
-units (GPUs), tensor processing units (TPUs), and field-programmable
-gate arrays (FPGAs).  Data sets now routinely exceed the memory capacity of a single machine and may be stored on multiple machines or in
-the cloud.  Due to its in-memory data model, NumPy is currently unable to
-utilize such specialized hardware and storage directly.  However, both
+However, scientific data sets now routinely exceed the memory capacity of a single machine and may
+be stored on multiple machines or in the cloud.
+In addition, the recent need to accelerate deep learning and artificial intelligence applications
+has led to the emergence of specialized accelerator hardware,
+including graphics processing units (GPUs), tensor processing units (TPUs),
+and field-programmable gate arrays (FPGAs).
+Due to its in-memory data model, NumPy is currently unable to
+utilize such storage and specialized hardware directly.  However, both
 distributed data and the parallel execution of GPUs, TPUs, and FPGAs map well
@@
 
-% maybe mention that there are typically utilities
-% to convert to and from numpy arrays
-% and why (presumably to get access to the ecosystem)
 The community's efforts to fill this gap led to a
@@
 Over four hundred of the most popular
-NumPy functions are supported.  The protocols are implemented by at
-least five libraries, including Dask, CuPy, xarray, and PyData/Sparse.
+NumPy functions are supported.
+The protocols are implemented by widely used libraries such as Dask, CuPy,
+xarray, and PyData/Sparse.
 Thanks to these developments, users can now, for example, scale
 their computation from a single machine to distributed systems using Dask.
-The protocols \textit{compose} well, and so it is possible to redeploy NumPy
-code onto multi-GPU systems with minimal changes\cite{entschev2019}.
-Working with the community, we have identified several issues that
-prevent further adoption, and are refining interfaces to reduce
-impedance mismatch.
+The protocols also \textit{compose} well, allowing users to redeploy NumPy
+code at scale on distributed, multi-GPU systems via, for instance, CuPy arrays embedded in Dask
+arrays. Using NumPy's high-level API, users can leverage highly parallel code
+execution on multiple systems with millions of cores, all with minimal code
+changes \cite{entschev2019}.
+
+These array protocols are now a key feature of NumPy, and are expected to only
+increase in importance.  As with the rest of NumPy, we iteratively refine and
+add protocol designs to improve utility and simplify adoption.
 
@@
 
-NumPy has subsequently seen prolific adoption, with millions of users,
-first from research and later also industry.
-It spurred the development of the broader scientific Python ecosystem and
-heralded the current era of widespread use of Python for scientific
-computing. % This and the "What began as" sentence below are more or less interchangeable.
+%NumPy has subsequently seen prolific adoption, with millions of users,
+%first from research and later also industry.
+%It spurred the development of the broader scientific Python ecosystem and
+%heralded the current era of widespread use of Python for scientific
+%computing. % This and the "What began as" sentence below are more or less interchangeable.
 
-%What began as an attempt to add an array object to Python became
-%the foundation of a vibrant ecosystem of tools.
-%Now, a large amount of scientific work depends on NumPy being correct,
-%fast, and stable.
-%%
-%%
-%% v TO ADD v
-%%
-%% Over the next decade, the computational landscape will change
-%% drastically.
-%% - AI / data science mature
-%% - hardware / Moore's law, specialized devices
-%% - more users (diversity + numbers)
-%% - scale / type of data; scientific data, lightsheet microscopes, LSST
-%% - new ideas come from other languages like Julia, Rust, ...; JIT improvements
-%% NumPy is poised to tackle these challenges, and work continues.
-%% - Major refactoring of code
-%% - Governance
-%% - Process (NEP, roadmap, how we handle issues on GH, etc.)
-%% - Protocols: will be extending that work
-%% - Funding (some recent evidence that that is improving → grants, or
-%% Moore/Sloan/CZI) CZI EOSS
-%% - Redesigning website and documentation to be more friendly, easy to
-%% navigate
-%% - Next generation of developers (grad students) to sign on
-%%
+What began as an attempt to add an array object to Python became
+the foundation of a vibrant ecosystem of tools.
+Now, a large amount of scientific work depends on NumPy being correct,
+fast, and stable.
+It is no longer a small community project, but is core scientific
+infrastructure.
+The developer culture has changed as well.
+- Governance
+- Process (NEP, roadmap, how we handle issues on GH, etc.)
+- still primarily draw from graduate students and researchers
+- Funding (some recent evidence that that is improving  grants, or
+Moore/Sloan/CZI) CZI EOSS
+- Major refactoring of code
+- Redesigning website and documentation to be more friendly, easy to
+navigate
 
@@
 array types and technologies in Python.
+
+- Protocols: will be extending that work
+
+Over the next decade, we will face new challenges.
+- AI / data science mature
+- hardware / Moore's law, specialized devices
+- more users (diversity + numbers)
+- scale / type of data; scientific data, lightsheet microscopes, LSST
+- new ideas come from other languages like Julia, Rust, ...; JIT improvements
+
+NumPy is poised to tackle these challenges, and florish in the next decade.
+Need
+- Next generation of developers (grad students) to sign on
+- Funders and industry continue funding this project as well as increasing support for the ecosystem through new funding mechanisms.

### 2020-02-14 body.tex
Flesh out bullet points in discussion section

@@
 well-tested, well-documented, and community-developed library.
-Many fast algorithms are already implemented by libraries in the
+Many fast algorithms are implemented by libraries in the
 scientific Python ecosystem.
@@
 
-What began as an attempt to add an array object to Python became
-the foundation of a vibrant ecosystem of tools.
-Now, a large amount of scientific work depends on NumPy being correct,
-fast, and stable.
-It is no longer a small community project, but is core scientific
+What began as an attempt to add an array object to Python became the
+foundation of a vibrant ecosystem of tools.  Now, a large amount of
+scientific work depends on NumPy being correct, fast, and stable.  It
+is no longer a small community project, but is core scientific
 infrastructure.
-The developer culture has changed as well.
-- Governance
-- Process (NEP, roadmap, how we handle issues on GH, etc.)
-- still primarily draw from graduate students and researchers
-- Funding (some recent evidence that that is improving  grants, or
-Moore/Sloan/CZI) CZI EOSS
-- Major refactoring of code
-- Redesigning website and documentation to be more friendly, easy to
-navigate
 
-NumPy's is no longer just the foundational array library underlying
+The developer culture has matured: while initial development was
+highly informal, NumPy now has a roadmap and a process for proposing
+and discussing large changes.  The project has formal governance
+structures and is fiscally sponsored by NumFOCUS, a nonprofit that
+promotes open practices in research, data, and scientific computing.
+Over the past few years, the project attracted its first funded
+development, sponsored by the Moore and Sloan Foundations, and
+received an award as part of the Chan Zuckerberg Initiative's
+Essentials of Open Source Software program.  With this funding, the
+project was (and is) able to have sustained focus over multiple months
+to implement substantial new features and improvements.  That said, it
+still depends heavily on contributions made by graduate students and
+researchers.
+
+NumPy's is no longer \emph{just} the foundational array library underlying
 the scientific Python ecosystem, but has also become the standard API
 for tensor computation and a central coordinating mechanism between
-array types and technologies in Python.
+array types and technologies in Python. Work continues to expand on and
+improve these interoperability features.
 
-- Protocols: will be extending that work
+Over the next decade, we will face several new challenges.
+%Artificial
+%Intelligence and Data Science will mature ... [what do we say about
+%  this?].
+New devices will be developed, and existing specialized
+hardware will evolve, to meet diminishing returns on Moore's law.
+There will be more, and a wider variety of, data science
+practitioners, a significant proportion of whom will be using NumPy.
+The scale of scientific data gathering will continue to expand, with
+the adoption of devices and instruments such as light sheet
+microscopes and the LSST.  New generation languages, interpreters, and compilers,
+such as Rust, Julia, and LLVM, will invent and determine the viability
+of new concepts and data structures.
 
-Over the next decade, we will face new challenges.
-- AI / data science mature
-- hardware / Moore's law, specialized devices
-- more users (diversity + numbers)
-- scale / type of data; scientific data, lightsheet microscopes, LSST
-- new ideas come from other languages like Julia, Rust, ...; JIT improvements
-
-NumPy is poised to tackle these challenges, and florish in the next decade.
-Need
-- Next generation of developers (grad students) to sign on
-- Funders and industry continue funding this project as well as increasing support for the ecosystem through new funding mechanisms.
+Through various mechanisms described in this paper, NumPy is poised to
+embrace such a changing landscape, and to continue playing a leading
+role in interactive scientific computation.  To do so will require
+sustained funding from government, academia, and industry.  But,
+importantly, it will also need a new generation of graduate students
+and other developers to engage, to build a NumPy that meets the needs
+of the next decade of data science.

### 2020-02-15 authors.tex
Update MW affil

@@
 \author[15]{Allan Haldane}
-\author[16]{Jaime Fernandez del Rio}
-\author[17]{Mark Wiebe}
-\author[5,18,19]{Pearu Peterson}
-\author[20]{Pierre G\'erard-Marchant}
-\author[21]{Kevin Sheppard}
-\author[22]{Tyler Reddy}
+\author[16]{Jaime Fern\'andez del R\'io}
+\author[17,18]{Mark Wiebe}
+\author[5,19,20]{Pearu Peterson}
+\author[21]{Pierre G\'erard-Marchant}
+\author[22]{Kevin Sheppard}
+\author[23]{Tyler Reddy}
 \author[4]{Warren Weckesser}
 \author[5]{Hameer Abbasi}
-\author[23]{Christoph Gohlke}
+\author[24]{Christoph Gohlke}
 \author[5]{Travis E. Oliphant}
@@
 \affil[16]{Google, Zurich, Switzerland}
-\affil[17]{Amazon, Seattle, Washington, USA}
-\affil[18]{Independent Researcher, Saue, Estonia}
-\affil[19]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
-\affil[20]{France-IX Services, Paris, France}
-\affil[21]{Department of Economics, University of Oxford, Oxford, UK}
-\affil[22]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
-\affil[23]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
+\affil[17]{Department of Physics and Astronomy, The University of British Columbia, Vancouver, BC, Canada}
+\affil[18]{Amazon, Seattle, Washington, USA}
+\affil[19]{Independent Researcher, Saue, Estonia}
+\affil[20]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
+\affil[21]{France-IX Services, Paris, France}
+\affil[22]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[23]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[24]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
 \affil[*]{numpy.articles@gmail.com}

### 2020-02-15 body.tex
Minor change to Fig. 1
Add references to discussion

@@
 still depends heavily on contributions made by graduate students and
-researchers.
+researchers in their free time.
 
@@
 the adoption of devices and instruments such as light sheet
-microscopes and the LSST.  New generation languages, interpreters, and compilers,
-such as Rust, Julia, and LLVM, will invent and determine the viability
-of new concepts and data structures.
+microscopes and the LSST \cite{jenness2018lsst}.  New generation languages, interpreters, and compilers,
+such as Rust \cite{10.1145/2692956.2663188}, Julia \cite{Julia-2017}, and LLVM \cite{LLVM:CGO04}, will invent
+and determine the viability of new concepts and data structures.
 

### 2020-02-15 authors.tex
Add Stefan's older affiliations

--- 2020-02-14 authors.tex
@@
 \author[2,3,4,*]{K. Jarrod Millman}
-\author[2,4,*]{St\'efan J. van der Walt}
-\author[5,*]{Ralf Gommers}
-\author[6]{Pauli Virtanen}
-\author[7]{David Cournapeau}
-\author[8]{Eric Wieser}
-\author[9]{Julian Taylor}
+\author[5,2,4,*]{St\'efan J. van der Walt}
+\author[6,*]{Ralf Gommers}
+\author[7]{Pauli Virtanen}
+\author[8]{David Cournapeau}
+\author[9]{Eric Wieser}
+\author[10]{Julian Taylor}
 \author[4]{Sebastian Berg}
-\author[10]{Nathaniel J. Smith}
-\author[11]{Robert Kern}
+\author[11]{Nathaniel J. Smith}
+\author[12]{Robert Kern}
 \author[4]{Matti Picus}
-\author[12]{Stephan Hoyer}
-\author[13]{Marten H. van Kerkwijk}
-\author[2,14]{Matthew Brett}
-\author[15]{Allan Haldane}
-\author[16]{Jaime Fernandez del Rio}
-\author[17]{Mark Wiebe}
-\author[5,18,19]{Pearu Peterson}
-\author[20]{Pierre G\'erard-Marchant}
-\author[21]{Kevin Sheppard}
-\author[22]{Tyler Reddy}
+\author[13]{Stephan Hoyer}
+\author[14]{Marten H. van Kerkwijk}
+\author[2,15]{Matthew Brett}
+\author[16]{Allan Haldane}
+\author[17]{Jaime Fern\'andez del R\'io}
+\author[18,19]{Mark Wiebe}
+\author[6,20,21]{Pearu Peterson}
+\author[22]{Pierre G\'erard-Marchant}
+\author[23]{Kevin Sheppard}
+\author[24]{Tyler Reddy}
 \author[4]{Warren Weckesser}
-\author[5]{Hameer Abbasi}
-\author[23]{Christoph Gohlke}
-\author[5]{Travis E. Oliphant}
+\author[6]{Hameer Abbasi}
+\author[25]{Christoph Gohlke}
+\author[6]{Travis E. Oliphant}
 \affil[1]{Independent Researcher, Logan, Utah, USA}
@@
 \affil[4]{Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA, USA}
-\affil[5]{Quansight LLC, Austin, TX, USA}
-\affil[6]{Department of Physics and Nanoscience Center, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
-\affil[7]{Mercari JP, Tokyo, Japan}
-\affil[8]{Department of Engineering, University of Cambridge, Cambridge, UK}
-\affil[9]{Independent Researcher, Karlsruhe, Germany}
-\affil[10]{Independent Researcher, Berkeley, CA, USA}
-\affil[11]{Enthought, Inc., Austin, TX, USA}
-\affil[12]{Google Research, Mountain View, CA, USA}
-\affil[13]{Department of Astronomy \& Astrophysics, University of Toronto, Toronto, ON, Canada}
-\affil[14]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
-\affil[15]{Department of Physics, Temple University, Philadelphia, PA, USA}
-\affil[16]{Google, Zurich, Switzerland}
-\affil[17]{Amazon, Seattle, Washington, USA}
-\affil[18]{Independent Researcher, Saue, Estonia}
-\affil[19]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
-\affil[20]{France-IX Services, Paris, France}
-\affil[21]{Department of Economics, University of Oxford, Oxford, UK}
-\affil[22]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
-\affil[23]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
+\affil[5]{Applied Mathematics, Stellenbosch University, Stellenbosch, South Africa}
+\affil[6]{Quansight LLC, Austin, TX, USA}
+\affil[7]{Department of Physics and Nanoscience Center, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
+\affil[8]{Mercari JP, Tokyo, Japan}
+\affil[9]{Department of Engineering, University of Cambridge, Cambridge, UK}
+\affil[10]{Independent Researcher, Karlsruhe, Germany}
+\affil[11]{Independent Researcher, Berkeley, CA, USA}
+\affil[12]{Enthought, Inc., Austin, TX, USA}
+\affil[13]{Google Research, Mountain View, CA, USA}
+\affil[14]{Department of Astronomy \& Astrophysics, University of Toronto, Toronto, ON, Canada}
+\affil[15]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
+\affil[16]{Department of Physics, Temple University, Philadelphia, PA, USA}
+\affil[17]{Google, Zurich, Switzerland}
+\affil[18]{Department of Physics and Astronomy, The University of British Columbia, Vancouver, BC, Canada}
+\affil[19]{Amazon, Seattle, Washington, USA}
+\affil[20]{Independent Researcher, Saue, Estonia}
+\affil[21]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
+\affil[22]{France-IX Services, Paris, France}
+\affil[23]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[24]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[25]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
 \affil[*]{numpy.articles@gmail.com}

### 2020-02-15 body.tex
Two minor fixes

@@
 
-NumPy's is no longer \emph{just} the foundational array library underlying
+NumPy is no longer \emph{just} the foundational array library underlying
 the scientific Python ecosystem, but has also become the standard API
@@
 the adoption of devices and instruments such as light sheet
-microscopes and the LSST \cite{jenness2018lsst}.  New generation languages, interpreters, and compilers,
+microscopes and the Large Synoptic Survey Telescope (LSST) \cite{jenness2018lsst}.  New generation languages, interpreters, and compilers,
 such as Rust \cite{10.1145/2692956.2663188}, Julia \cite{Julia-2017}, and LLVM \cite{LLVM:CGO04}, will invent

### 2020-02-15 summary.tex
Edit summary

@@
 % - a brief account of the background and rationale of the work;
-NumPy is the fundamental array programming library for the Python language
+NumPy is the primary array programming library for the Python language
 \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-Developed for over two decades by more than 850 contributors, 
-it has nearly 300,000 dependent repositories on GitHub and millions of
-downloads per year.
+% https://pypistats.org/packages/numpy
+% millions of downloads per week
+% 
+%Developed for over two decades, it has nearly 300,000 dependent repositories on
+%GitHub and millions of downloads per year.
 %It is one of the most commonly used machine learning tools among
@@
 % - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
-Here we show how a few fundamental concepts lead to a simple and
+Here we show how a few fundamental array concepts lead to a simple and
 powerful programming paradigm for organizing, exploring, and analyzing
@@
 targeting audiences with specialized needs, have developed their own
-NumPy-like interfaces and array objects.  Because of its central position in the
+NumPy-like interfaces and array objects%
+%---many in response to the rise of deep learning and artificial
+%intelligence.
+. Because of its central position in the
 ecosystem, NumPy increasingly plays the role of an interoperability layer
 between these new array computation libraries.
+% protocols ?
+% It make it easier for new experimental array implementation a simple
+% way to access the scientific Python ecosystem.
+% This allows users to scale computations from single machines to distributed systems.

### 2020-02-15 scripts/plot_merged_prs.py
Adjust plot colors by eye; rotate legend 30°

@@ -142,7 +142,7 @@ for g, comm in prs.groupby("category"):
 categories = ["Maintainer", "Other"]
 stacked = []
 # other color:  "#0e50a8", "#077a0a" OR   "#cce0fb", "#71f775"
-for color, cat in zip(["#0e50a8aa", "#077a0aaa"], categories):
+for color, cat in zip(["#b0d0f5ff", "#077a0a77"], categories):
     next_stack = groups[cat]
 
     if len(stacked) > 0:
@@ -158,6 +158,7 @@ for color, cat in zip(["#0e50a8aa", "#077a0aaa"], categories):
         next_stack.values.repeat(2), prev_vals.repeat(2),
         label=cat, lw=0, zorder=4 if cat == "Community" else 3,
         color=color)
+    plt.xticks(rotation=30)
     if False:
         plt.plot(
             next_stack.index + datetime.timedelta(days=30*3/2), next_stack.values, "o",

### 2020-02-16 body.tex
Update Figs

@@
 
-  \caption{\textbf{NumPy API and Array Protocol.} Through its array protocol, the NumPy API
+  \caption{\textbf{NumPy API and Array Protocols.} Through its array protocols, the NumPy API
      can operate on various types of array objects, including those written outside

### 2020-02-16 authors.tex
Update PGM affil

@@
 \author[6,20,21]{Pearu Peterson}
-\author[22]{Pierre G\'erard-Marchant}
-\author[23]{Kevin Sheppard}
-\author[24]{Tyler Reddy}
+\author[22,23]{Pierre G\'erard-Marchant}
+\author[24]{Kevin Sheppard}
+\author[25]{Tyler Reddy}
 \author[4]{Warren Weckesser}
 \author[6]{Hameer Abbasi}
-\author[25]{Christoph Gohlke}
+\author[26]{Christoph Gohlke}
 \author[6]{Travis E. Oliphant}
@@
 \affil[21]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
-\affil[22]{France-IX Services, Paris, France}
-\affil[23]{Department of Economics, University of Oxford, Oxford, UK}
-\affil[24]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
-\affil[25]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
+\affil[22]{Department of Biological and Agricultural Engineering, University of Georgia, Athens, GA}
+\affil[23]{France-IX Services, Paris, France}
+\affil[24]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[25]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[26]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
 \affil[*]{numpy.articles@gmail.com}

### 2020-02-20 body.tex
Manually move fig. 3
Update caption for ecosystem sketch

@@
 
-% FIXME SJW: Redraw the icons in vector format
 \begin{figure}
@@
   \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
-  \caption{\textbf{Scientific Python Ecosystem.}
-   This figure highlights (in blue) some of the essential tools, projects, and libraries that depend
-   on NumPy.  Since most of the ecosystem
-   builds on NumPy arrays and its rich and powerful collection of array-aware
-   functionality, the simplest way for new array implementations to become
-   available to the vast scientific Python ecosystem is by using the NumPy API and array
-   protocols (Fig.~\ref{fig:array-protocol}).
+  \caption{\textbf{NumPy lies at the root of the scientific Python ecosystem.}
+   Essential libraries and projects that depend on NumPy's API gain access to new array
+   implementations that support NumPy's array protocols (Fig.~\ref{fig:array-protocol}).
   }
@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=\textwidth]{static/sketches/protocol}
+
+  \caption{\textbf{NumPy API and Array Protocols.} Through its array protocols, the NumPy API
+     can operate on various types of array objects, including those written outside
+     of NumPy. In this example, NumPy's \code{mean} function is called on a Dask
+     array.  The call succeeds by dispatching to the appropriate library implementation
+     (i.e., Dask in this case) and results in a new Dask array.  Compare this
+     code to the example code in Fig.~\ref{fig:array-concepts}g.
+  }\label{fig:array-protocol}
+
+\end{figure*}
+
 NumPy provides in-memory, multidimensional, homogeneously typed
@@
 mechanism with a well-specified API (Fig.~\ref{fig:ecosystem}).
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=\textwidth]{static/sketches/protocol}
-
-  \caption{\textbf{NumPy API and Array Protocols.} Through its array protocols, the NumPy API
-     can operate on various types of array objects, including those written outside
-     of NumPy. In this example, NumPy's \code{mean} function is called on a Dask
-     array.  The call succeeds by dispatching to the appropriate library implementation
-     (i.e., Dask in this case) and results in a new Dask array.  Compare this
-     code to the example code in Fig.~\ref{fig:array-concepts}g.
-  }\label{fig:array-protocol}
-
-\end{figure*}
 

### 2020-02-20 body.tex
Tweak body

@@
 \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
-NumPy, as the focal point for the ecosystem of array-aware libraries,
+NumPy, at the root of the ecosystem of array-aware libraries,
 sets documentation standards, provides array testing infrastructure,

### 2020-02-20 methods.tex
Make caption a sentence

@@
     \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
-    \caption{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012. The total number of PRs is indicated with the
+    \caption{\textbf{Number of pull requests merged into the NumPy master branch for each
+        quarter since 2012.} The total number of PRs is indicated with the
         lower blue area showing the portion contributed by current or previous

### 2020-02-20 body.tex
Make all captions sentences

@@
   \includegraphics[width=\textwidth]{static/sketches/concepts}
-  \caption{\textbf{Fundamental Array Concepts.}
+  \caption{\textbf{The NumPy array incorporates several fundamental array concepts.}
     \textbf{a,} The NumPy array data structure and its associated metadata fields.
@@
 
-  \caption{\textbf{NumPy API and Array Protocols.} Through its array protocols, the NumPy API
-     can operate on various types of array objects, including those written outside
-     of NumPy. In this example, NumPy's \code{mean} function is called on a Dask
+  \caption{\textbf{NumPy's API and array protocols expose new arrays to the ecosystem.}
+     In this example, NumPy's \code{mean} function is called on a Dask
      array.  The call succeeds by dispatching to the appropriate library implementation

### 2020-02-20 body.tex
Remove old text

@@
 % KJM: I moved this here to get the placement right
-% FIXME SJW: let's review caption (d); I don't think it's crystal clear what
-% we're trying to show there; double check caption (g)
 \begin{figure*}[h]
@@
 
-% this is the software engineering side of things
-% as opposed to the interactive, exploration described above
-
 To complement this facility for exploratory work and rapid
@@
 
-%CuPy, PyData/Sparse, and JAX the most prominent examples---while others have
-%NumPy-inspired semantics but, often due to their development history, deviate in places.
-%Wherever there are such deviations, we often observe users gently requesting,
-%and developers evolving the software step by step toward the NumPy API.
-
-% when they deviate, they sometimes fix it, e.g.,
-% https://github.com/pytorch/pytorch/pull/1983
-% https://github.com/pytorch/pytorch/issues/2228
-% many have for numpy users
-% https://github.com/torch/torch7/wiki/Torch-for-Numpy-users
-
-% computation graphs---which postpone and combine calculations
-% before executing them efficiently,
-
 Ideally, operating on specialized arrays using NumPy functions or semantics would
@@
 add protocol designs to improve utility and simplify adoption.
-
-% FIXME: expand to 3 or more sentences discussing current efforts and
-% issues with those, how did adoption go, what is being done next
-%\url{https://numpy.org/neps/nep-0037-array-module.html}
-%Versions of these protocols have been successfully deployed.
-% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
-% possible to create distributed GPU arrays, enabling .... [find a reference to
-% work by Peter Entschev]".
-% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
-% Below an attempt to start on this:
-
-% should we hint at some of the community interaction part
-% they have helped us develop the neps and, e.g.
-% https://github.com/pytorch/pytorch/issues/22402
-
-% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
-% The video cited embedded there could be a citation:i
-% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
-%However, general adoption has not yet been achieved mainly for compatibility
-%concerns.
-% explain a bit what "compatibility concerns" means ...
-%Thus, we are actively addressing these concerns by revising and extending
-%these protocols with the aim of wide interoperability that spans the whole
-%scientific Python ecosystem up to and including domain libraries.
-
-% https://numpy.org/neps/nep-0016-abstract-array.html
-% https://numpy.org/neps/nep-0018-array-function-protocol.html
-% https://numpy.org/neps/nep-0022-ndarray-duck-typing-overview.html
-% https://numpy.org/neps/nep-0030-duck-array-protocol.html
-% https://github.com/numpy/numpy/blob/a111b551ae940d7d5f8523fef1cf3589c6ba00a0/doc/neps/nep-0033-extensible-dtypes.rst
-% https://numpy.org/neps/nep-0037-array-module.html
 
@@
 
-%NumPy has subsequently seen prolific adoption, with millions of users,
-%first from research and later also industry.
-%It spurred the development of the broader scientific Python ecosystem and
-%heralded the current era of widespread use of Python for scientific
-%computing. % This and the "What began as" sentence below are more or less interchangeable.
-
 What began as an attempt to add an array object to Python became the
@@
 
-Over the next decade, we will face several new challenges.
-%Artificial
-%Intelligence and Data Science will mature ... [what do we say about
-%  this?].
-New devices will be developed, and existing specialized
-hardware will evolve, to meet diminishing returns on Moore's law.
-There will be more, and a wider variety of, data science
-practitioners, a significant proportion of whom will be using NumPy.
-The scale of scientific data gathering will continue to expand, with
-the adoption of devices and instruments such as light sheet
-microscopes and the Large Synoptic Survey Telescope (LSST) \cite{jenness2018lsst}.  New generation languages, interpreters, and compilers,
-such as Rust \cite{10.1145/2692956.2663188}, Julia \cite{Julia-2017}, and LLVM \cite{LLVM:CGO04}, will invent
-and determine the viability of new concepts and data structures.
+Over the next decade, we will face several new challenges.  New devices will be
+developed, and existing specialized hardware will evolve, to meet diminishing
+returns on Moore's law.  There will be more, and a wider variety of, data
+science practitioners, a significant proportion of whom will be using NumPy.
+The scale of scientific data gathering will continue to expand, with the
+adoption of devices and instruments such as light sheet microscopes and the
+Large Synoptic Survey Telescope (LSST) \cite{jenness2018lsst}.  New generation
+languages, interpreters, and compilers, such as Rust
+\cite{10.1145/2692956.2663188}, Julia \cite{Julia-2017}, and LLVM
+\cite{LLVM:CGO04}, will invent and determine the viability of new concepts and
+data structures.
 

### 2020-02-20 summary.tex
Remove old text

--- 2020-02-13 summary.tex
@@
 % - a brief account of the background and rationale of the work;
-NumPy is the fundamental array programming library for the Python language
+NumPy is the primary array programming library for the Python language
 \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
-Developed for over two decades by more than 850 contributors, 
-it has nearly 300,000 dependent repositories on GitHub and millions of
-downloads per year.
-%It is one of the most commonly used machine learning tools among
-%enterprises \cite{451report2018}.
 It plays an essential role in research analysis pipelines in fields as

### 2020-02-20 body.tex
Minor edits

@@
     \textbf{f,} Reduction operations act along one or more axes. In this
-    example, an array is be summed along various axes to produce a vector, or along two axes consecutively to
+    example, an array is summed along select axes to produce a vector, or along two axes consecutively to
     produce a scalar.
@@
 well-tested, well-documented, and community-developed library.
-Many fast algorithms are implemented by libraries in the
-scientific Python ecosystem.
+Libraries in the scientific Python ecosystem provide fast implementations of most important algorithms.
 Where extreme optimization is warranted, compiled languages such as
@@
 
-Over the next decade, we will face several new challenges.  New devices will be
+Over the next decade, we will face several challenges.  New devices will be
 developed, and existing specialized hardware will evolve, to meet diminishing

### 2020-02-20 body.tex
Tiny textual fix ("their own")

@@
 These user-developers frequently had to write code from scratch to solve
-their or their colleagues' problems---often in low-level languages
+their own or their colleagues' problems---often in low-level languages
 that precede Python, like Fortran \cite{dongarra2008netlib} and C.

### 2020-02-21 body.tex
Minor edits

@@
   \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
-  \caption{\textbf{NumPy lies at the root of the scientific Python ecosystem.}
+  \caption{\textbf{NumPy is the base of the scientific Python ecosystem.}
    Essential libraries and projects that depend on NumPy's API gain access to new array

### 2020-02-21 body.tex
Minor edits
Fix word

@@
   \centering
-  \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
+  \includegraphics[width=.5\textwidth]{static/sketches/ecosystem}
   \caption{\textbf{NumPy is the base of the scientific Python ecosystem.}
@@
 \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
-NumPy, at the root of the ecosystem of array-aware libraries,
+NumPy, at the base of the ecosystem of array-aware libraries,
 sets documentation standards, provides array testing infrastructure,

### 2020-05-23 methods.tex
Move methods to standalone PDF

@@
-\section*{Methods}
-
 We use Git for version control and GitHub as the public hosting service for our
@@
 
-\subsection*{Library organization}
+\section*{Library organization}
 
@@
 
-\paragraph{\emph{Core.}}  The \code{ndarray} data structure and the
+\subsection*{Core}
+
+The \code{ndarray} data structure and the
 universal functions make up the core of NumPy.
@@
 
-\paragraph{\emph{Computing libraries.}}
+\subsection*{Computing libraries.}
+
 NumPy provides a large library of functions for array manipulation
@@
 
-\paragraph{\emph{Infrastructure libraries.}} NumPy provides utilities
-for writing tests and for building Python packages.
+\subsection*{Infrastructure libraries}
+
+NumPy provides utilities for writing tests and for building Python packages.
 
@@
 
-\paragraph{\emph{F2PY.}}  The program \code{f2py} is a tool for
+\subsection*{F2PY}  The program \code{f2py} is a tool for
 building NumPy-aware Python wrappers of Fortran functions.
@@
 
-\subsection*{Governance}
+\section*{Governance}
 
@@
 
-\subsection*{Funding}
+\section*{Funding}
 
@@
 
-\subsection*{Developers}
+\section*{Developers}
 
@@
 
-\subsection*{Community calls}
+\section*{Community calls}
 
@@
 
-\subsection*{NumPy enhancement proposals}
+\section*{NumPy enhancement proposals}
 
@@
 
-\subsection*{Central role}
+\section*{Central role}
 
@@
 
-\subsection*{Wheels build system}
+\section*{Wheels build system}
 
@@
 
-\subsection*{Recent technical improvements}
+\section*{Recent technical improvements}
 
@@
 
-\subsection*{Array function protocol}
+\section*{Array function protocol}
 
@@
 
-\subsection*{Random number generation}
+\section*{Random number generation}
 
@@
 
-\subsection*{Testing on multiple architectures}
+\section*{Testing on multiple architectures}
 

### 2020-05-23 supplementary.tex
Move methods to standalone PDF

--- 2020-02-06 paper.tex
@@
 
-\title{NumPy---Array Programming in Python}
+\title{Supplementary Methods}
 
 \input{authors}
-
-\keywords{Scientific computing, Python, Mathematics, Array Programming}
 
@@
 % Count of words
-\verbatiminput{wordcount.tex}
+%\verbatiminput{wordcount.tex}
 
-\input{body}
+\input{methods}
 
@@
 
-\newpage
-
-\input{methods}
-
-\section*{Acknowledgments}
-
-Jim Hugunin wrote Numeric in 1995, while a graduate student at MIT.
-Hugunin based his package on previous work by Jim Fulton, then working at the
-US Geological Survey, with input from many others.
-After he graduated, Paul Dubois at the Lawrence Livermore National Laboratory
-became the maintainer.
-Many people contributed to the project including Travis Oliphant, David Ascher,
-Tim Peters, and Konrad Hinsen.
-
-Around 1998, the Space Telescope Science Institute (STScI) software group began
-to use Python heavily, and, around 2000, wrote a reimplementation of
-Numeric, called NumArray.
-%https://lancesimms.com/programs/Python/pyraf/Docs/greenfield.pdf
-Perry Greenfield created the original framework.
-Many people contributed to it including Jay Todd Miller, JT Hsu, Richard
-L White, Jochen Krupper, and Phil Hodge.
-
-Eric Jones co-founded the SciPy community, gave early feedback on array
-implementations, and provided funding and travel support to several community
-members.
-Numerous people contributed to the creation and growth of the larger SciPy
-ecosystem, which gives NumPy much of its value.  Others have injected new
-energy and ideas by creating experimental array packages.
-
-Finally, we would like to thank the many members of the community who have provided
-feedback, submitted bug reports, made improvements to the documentation,
-code, or website, promoted its use in their scientific fields, and built
-the vast ecosystem of tools and libraries around NumPy.
-Ross Barnowski and Michael Eickenberg suggested text and provided
-helpful feedback on the manuscript.
-
-S.J.v.d.W and K.J.M. were funded in part by the Gordon and Betty Moore
-Foundation through Grant GBMF3834 and by the Alfred P. Sloan Foundation through
-Grant 2013-10-27 to the University of California, Berkeley.
-S.J.v.d.W, S.B., M.P., and W.W. were funded in part by the Gordon
-and Betty Moore Foundation through Grant GBMF5447 and by the Alfred
-P. Sloan Foundation through Grant G-2017-9960 to the University of
-California, Berkeley.
-
-\section*{Author Contributions Statement}
-
-K.J.M. and S.J.v.d.W composed the manuscript with input from others.
-S.B., R.G., K.S., W.W., M.B., and T.J.R. contributed text.
-All authors have contributed significant code, documentation, and/or expertise
-to the NumPy project.
-All authors reviewed the manuscript.
-
-\section*{Competing Interests}
-
-The authors declare no competing interests.
-
 \end{document}

### 2020-05-23 body.tex
Revise abstract

@@
 Now, fifteen years later, NumPy underpins almost every Python library that does scientific or
-numerical computation including SciPy \cite{virtanen2020scipy},
+numerical computation \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}
+including SciPy \cite{virtanen2020scipy},
 Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},

### 2020-05-23 summary.tex
Revise abstract

@@
 manipulating, and operating on data in vectors, matrices, and
-higher-dimensional arrays \cite{iverson1980notation}.
+higher-dimensional arrays. % \cite{iverson1980notation}.
 % - a brief account of the background and rationale of the work;
-NumPy is the primary array programming library for the Python language
-\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
+NumPy is the primary array programming library for the Python language.
+%\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
 It plays an essential role in research analysis pipelines in fields as
@@
 For example, in astronomy, NumPy was an important part of the software stack used
-in the discovery of gravitational waves \cite{abbott2016observation}
-and the first imaging of a black hole \cite{eht-imaging}.
+in the discovery of gravitational waves% \cite{abbott2016observation}
+and the first imaging of a black hole. % \cite{eht-imaging}.
 % - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
-Here we show how a few fundamental array concepts lead to a simple and
+Here we review how a few fundamental array concepts lead to a simple and
 powerful programming paradigm for organizing, exploring, and analyzing
@@
 between these new array computation libraries.
+
+%Reviews start with a 200-word maximum abstract, which should set the stage and
+%end with a summary sentence.
+%This abstract cannot include references, and it's important that it makes
+%it clear that this article is a Review article.

### 2020-05-25 summary.tex
Add missing space

@@
 For example, in astronomy, NumPy was an important part of the software stack used
-in the discovery of gravitational waves% \cite{abbott2016observation}
+in the discovery of gravitational waves % \cite{abbott2016observation}
 and the first imaging of a black hole. % \cite{eht-imaging}.

### 2020-05-25 body.tex
Simplify tex for submission

@@
 
-Other array-aware functions, such as \code{sum}, \code{mean}, and \code{maximum}, perform
+Other array-aware functions, such as \texttt{sum}, \texttt{mean}, and \texttt{maximum}, perform
 element-by-element \emph{reductions}, aggregating results across one,
@@
 to the ecosystem.
-For example, the \code{eht-imaging} library \cite{chael2019ehtim} developed by
+For example, the \texttt{eht-imaging} library \cite{chael2019ehtim} developed by
 the Event Horizon Telescope collaboration for radio interferometry imaging,
@@
 edge filters and Hough transforms.
-The \code{scipy.optimize} module performs mathematical optimization.
+The \texttt{scipy.optimize} module performs mathematical optimization.
 NetworkX \cite{SciPyProceedings_11}, a package for complex
@@
 scientific Python.  Examples of its significant application, such as the
-\code{eht-imaging} library, now exist in almost every discipline in the natural and social
+\texttt{eht-imaging} library, now exist in almost every discipline in the natural and social
 sciences.  These tools have become \emph{the primary}
@@
   \caption{\textbf{NumPy's API and array protocols expose new arrays to the ecosystem.}
-     In this example, NumPy's \code{mean} function is called on a Dask
+     In this example, NumPy's \texttt{mean} function is called on a Dask
      array.  The call succeeds by dispatching to the appropriate library implementation
@@
 made possible that way by Dask, and labeled arrays---referring to dimensions of
-an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
-\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.
+an array by name rather than by index for clarity, compare \texttt{x[:,~1]} vs.
+\texttt{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.
 

### 2020-05-25 methods.tex
Simplify tex for submission

@@
 Broadly, the NumPy library consists of the following parts:
-the NumPy array data structure \code{ndarray}; the so-called \emph{universal functions};
+the NumPy array data structure \texttt{ndarray}; the so-called \emph{universal functions};
 a set of library functions for manipulating arrays and doing scientific
 computation; infrastructure libraries for unit tests and Python package
-building; and the program \code{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
-The \code{ndarray} and the universal functions are generally considered
+building; and the program \texttt{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
+The \texttt{ndarray} and the universal functions are generally considered
 the core of the library.
@@
 
-The \code{ndarray} data structure and the
+The \texttt{ndarray} data structure and the
 universal functions make up the core of NumPy.
 
-The \code{ndarray} is the data structure at the heart of NumPy.
+The \texttt{ndarray} is the data structure at the heart of NumPy.
 The data structure stores regularly strided homogeneous data types
@@
 implementation of \emph{broadcasting}.  For example, the function
-\code{arctan2(x, y)} is a ufunc that accepts two values and computes
+\texttt{arctan2(x, y)} is a ufunc that accepts two values and computes
 $\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
 the ufunc will take care of looping over the dimensions of the inputs
-in such a way that if, say, \code{x} is a 1-D array with length 3, and
-\code{y} is a 2-D array with shape $2 \times 1$, the output will be
+in such a way that if, say, \texttt{x} is a 1-D array with length 3, and
+\texttt{y} is a 2-D array with shape $2 \times 1$, the output will be
 an array with shape $2 \times 3$ (Fig.~\ref{fig:array-concepts}c).
@@
 are implemented as ufuncs, so that broadcasting also applies to expressions
-such as \code{x + y * z}.
+such as \texttt{x + y * z}.
 
@@
 
-The \code{testing} subpackage provides functions such as
-\code{assert\_allclose(actual, desired)} that may be used in
+The \texttt{testing} subpackage provides functions such as
+\texttt{assert\_allclose(actual, desired)} that may be used in
 test suites for code that uses NumPy arrays.
 
-NumPy provides the subpackage \code{distutils} which includes functions and classes
+NumPy provides the subpackage \texttt{distutils} which includes functions and classes
 to facilitate configuration, installation, and packaging of libraries depending on NumPy.
@@
 
-\subsection*{F2PY}  The program \code{f2py} is a tool for
+\subsection*{F2PY}  The program \texttt{f2py} is a tool for
 building NumPy-aware Python wrappers of Fortran functions.
@@
 At the time this paper is being written, a key component is the
-\code{multibuild} suite of tools developed by Matthew Brett and
+\texttt{multibuild} suite of tools developed by Matthew Brett and
 other developers\cite{multibuild}.  Currently, scripts using
-\code{multibuild} are written for the continuous integration
+\texttt{multibuild} are written for the continuous integration
 platforms Travis-CI (for Linux and Mac OSX) and Appveyor
@@
 It is also implemented in packages that support sparse arrays
-such as \code{scipy.sparse} and PyData/Sparse.
+such as \texttt{scipy.sparse} and PyData/Sparse.
 Another notable example is Dask, a library for parallel computing in
@@
 An earlier step in this direction was the implementation of the
-\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
+\texttt{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
 for most mathematical functions.\cite{NEP13}
 In 2019 this was expanded more generally with the inclusion of the
-\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
+\texttt{\_\_array\_function\_\_} protocol into NumPy~1.17.
 These two protocols allow providers of array objects to be interoperable
@@
 much code is written specifically for NumPy arrays and uses the NumPy API as
-\code{import numpy as np}, it can nevertheless work for them.
+\texttt{import numpy as np}, it can nevertheless work for them.
 For example, here is how a CuPy GPU array can be passed through NumPy for
@@
 
-The NumPy \code{random} module provides pseudorandom numbers from a wide range of
+The NumPy \texttt{random} module provides pseudorandom numbers from a wide range of
 distributions. In legacy versions of NumPy, simulated random values are produced
-by a \code{RandomState} object that: handles seeding and state initialization;
+by a \texttt{RandomState} object that: handles seeding and state initialization;
 wraps the core pseudorandom number generator based on a Mersenne Twister
@@
 
-The \code{RandomState} object makes a compatibility guarantee so that a fixed
+The \texttt{RandomState} object makes a compatibility guarantee so that a fixed
 seed and sequence of function calls produce the same set of values. This
@@
 extending the API with additional keyword arguments. This guarantee continues to
-apply to \code{RandomState}.
+apply to \texttt{RandomState}.
 
@@
 bit generator is a 64-bit implementation of the Permuted Congruential Generator
-\cite{pcg64} (\code{PCG64}). The three other bit generators are a 64-bit version
-of the Philox generator\cite{random123} (\code{Philox}), Chris Doty-Humphrey's
-Small Fast Chaotic generator\cite{practrand} (\code{SFC64}), and the 32-bit
-Mersenne Twister\cite{mt19937} (\code{MT19937}) which has been used in older
+\cite{pcg64} (\texttt{PCG64}). The three other bit generators are a 64-bit version
+of the Philox generator\cite{random123} (\texttt{Philox}), Chris Doty-Humphrey's
+Small Fast Chaotic generator\cite{practrand} (\texttt{SFC64}), and the 32-bit
+Mersenne Twister\cite{mt19937} (\texttt{MT19937}) which has been used in older
 versions of NumPy.\footnote{The
@@
 range of alternative bit generators such as a cryptographic counter-based
-generators (\code{AESCtr}) and generators that expose hardware random number
-generators (\code{RDRAND})\cite{randomgen}.} Bit generators provide
+generators (\texttt{AESCtr}) and generators that expose hardware random number
+generators (\texttt{RDRAND})\cite{randomgen}.} Bit generators provide
 functions, exposed both in Python and C, for generating random integer
@@
 
-The \code{Generator} consumes one of the bit generators and produces variates
+The \texttt{Generator} consumes one of the bit generators and produces variates
 from complicated distributions. Many improved methods for generating random
@@
 method for normal, exponential and gamma variates\cite{ziggurat}, and Lemire's
-method for bounded random integer generation\cite{lemire}. The \code{Generator}
-is more similar to the legacy \code{RandomState}, and its API is substantially
+method for bounded random integer generation\cite{lemire}. The \texttt{Generator}
+is more similar to the legacy \texttt{RandomState}, and its API is substantially
 the same. The key differences all relate to state management, which has been
-delegated to the bit generator. The \code{Generator} does not make the same
-stream guarantee as the \code{RandomState} object, and so variates may differ
+delegated to the bit generator. The \texttt{Generator} does not make the same
+stream guarantee as the \texttt{RandomState} object, and so variates may differ
 across versions as improved generation algorithms are
@@
 
-Finally, a \code{SeedSequence} is used to initialize a bit generator. The seed
+Finally, a \texttt{SeedSequence} is used to initialize a bit generator. The seed
 sequence can be initialized with no arguments, in which case it reads entropy
@@
 generators deterministically. The key feature of a seed sequence is that
-it can be used to spawn child \code{SeedSequence}s to initialize
+it can be used to spawn child \texttt{SeedSequence}s to initialize
 multiple distinct bit generators.
@@
 While this approach retains complete flexibility, the method
-\code{np.random.default\_rng} can be used to instantiate a \code{Generator} when
+\texttt{np.random.default\_rng} can be used to instantiate a \texttt{Generator} when
 reproducibility is not needed.
 
-The final goal of the new API is to improve extensibility. \code{RandomState} is
+The final goal of the new API is to improve extensibility. \texttt{RandomState} is
 a monolithic object that obscures all of the underlying state and functions. The

### 2020-05-25 paper.tex
Remove word count
Update title
Move methods to standalone PDF
Simplify tex for submission

@@
 % Packages
-\usepackage{pifont}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-\usepackage{float}
-% \usepackage[multiple]{footmisc}
-\usepackage{booktabs}
-\usepackage{csquotes}
 \usepackage{moreverb} % for verbatim ouput (for wordcount)
-\usepackage{listings, textcomp}
-\lstset{language=Python}
-
-\lstset{ %
-  basicstyle=\ttfamily\footnotesize,  % size of fonts used for the code
-  xleftmargin=\parindent,
-  breaklines=true,   % automatic line breaking only at whitespace
-  captionpos=b,   % sets the caption-position to bottom
-  commentstyle=\color{gray},  % comment style
-  keywordstyle=\color{blue},  % keyword style
-  stringstyle=\color{red},  % string literal style
-  upquote=true  %straight single quotes (requires textcomp)
-}
 
@@
 }
-\usepackage{subcaption}
-
-% New Commands
-\newcommand{\cmark}{\ding{51}}%
-\newcommand{\xmark}{\ding{55}}%
-\newcommand{\code}[1]{\texttt{#1}}
-\newcommand{\fixme}[1]{\textcolor{red}{{#1}}}
-\newcommand{\inlinecite}[1]{\footnotesize\citen{#1}}
-\newcommand{\tightlist}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 
-\title{NumPy---Array Programming in Python}
+\title{Array Programming with NumPy}
 
@@
 % Count of words
-\verbatiminput{wordcount.tex}
+%\verbatiminput{wordcount.tex}
 
@@
 \bibliography{references}
-
-\newpage
-
-\input{methods}
 

### 2020-05-25 body.tex
Simplify tex for submission

@@
 Now, fifteen years later, NumPy underpins almost every Python library that does scientific or
-numerical computation \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}
-including SciPy \cite{virtanen2020scipy},
+numerical computation including SciPy \cite{virtanen2020scipy},
 Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},

### 2020-05-25 methods.tex
Simplify tex for submission

@@
+\section*{Methods}
+
 We use Git for version control and GitHub as the public hosting service for our
@@
 
-\section*{Library organization}
+\subsection*{Library organization}
 
@@
 
-\subsection*{Core}
-
-The \texttt{ndarray} data structure and the
+\paragraph{\emph{Core.}}  The \texttt{ndarray} data structure and the
 universal functions make up the core of NumPy.
@@
 
-\subsection*{Computing libraries.}
-
+\paragraph{\emph{Computing libraries.}}
 NumPy provides a large library of functions for array manipulation
@@
 
-\subsection*{Infrastructure libraries}
-
-NumPy provides utilities for writing tests and for building Python packages.
+\paragraph{\emph{Infrastructure libraries.}} NumPy provides utilities
+for writing tests and for building Python packages.
 
@@
 
-\subsection*{F2PY}  The program \texttt{f2py} is a tool for
+\paragraph{\emph{F2PY.}}  The program \texttt{f2py} is a tool for
 building NumPy-aware Python wrappers of Fortran functions.
@@
 
-\section*{Governance}
+\subsection*{Governance}
 
@@
 
-\section*{Funding}
+\subsection*{Funding}
 
@@
 
-\section*{Developers}
+\subsection*{Developers}
 
@@
 
-\section*{Community calls}
+\subsection*{Community calls}
 
@@
 
-\section*{NumPy enhancement proposals}
+\subsection*{NumPy enhancement proposals}
 
@@
 
-\section*{Central role}
+\subsection*{Central role}
 
@@
 
-\section*{Wheels build system}
+\subsection*{Wheels build system}
 
@@
 
-\section*{Recent technical improvements}
+\subsection*{Recent technical improvements}
 
@@
 
-\section*{Array function protocol}
+\subsection*{Array function protocol}
 
@@
 
-\section*{Random number generation}
+\subsection*{Random number generation}
 
@@
 
-\section*{Testing on multiple architectures}
+\subsection*{Testing on multiple architectures}
 

### 2020-05-25 paper.tex
Simplify tex for submission

@@
 % Packages
-\usepackage{moreverb} % for verbatim ouput (for wordcount)
+\usepackage{listings, textcomp}
+\lstset{language=Python}
+
+\lstset{ %
+  basicstyle=\ttfamily\footnotesize,  % size of fonts used for the code
+  xleftmargin=\parindent,
+  breaklines=true,   % automatic line breaking only at whitespace
+  captionpos=b,   % sets the caption-position to bottom
+  commentstyle=\color{gray},  % comment style
+  keywordstyle=\color{blue},  % keyword style
+  stringstyle=\color{red},  % string literal style
+  upquote=true  %straight single quotes (requires textcomp)
+}
 
@@
 }
-
-\usepackage[switch]{lineno}
-\linenumbers
 
@@
 
-% Count of words
-%\verbatiminput{wordcount.tex}
-
 \input{body}
@@
 \bibliography{references}
+
+\newpage
+
+\input{methods}
 

### 2020-05-26 methods.tex
Remove URLs from references

--- 2020-05-25 methods.tex
@@
 NumPy adopted an official Governance Document on October~5,
-2015\cite{NumPyProjectGovernance}.
+2015 (\url{https://numpy.org/devdocs/dev/governance/governance.html}).
 Project decisions are usually made by consensus of interested contributors.
@@
 % https://github.com/numpy/numpy/pull/11865
-NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
+NumPy's official Code of Conduct was approved on September~1, 2018
+(\url{https://numpy.org/devdocs/dev/conduct/code_of_conduct.html}).
 In brief, we strive to:
@@
 
-A Python \emph{wheel}\cite{PEP427} is a standard file format for
-distributing Python libraries.  In addition to Python code, a
-wheel may include compiled C extensions and other binary data.
+A Python \emph{wheel} ({https://www.python.org/dev/peps/pep-0427/})
+is a standard file format for distributing Python libraries.
+In addition to Python code, a wheel may include compiled
+C extensions and other binary data.
 This is important, because many libraries, including NumPy,
@@
 \texttt{multibuild} suite of tools developed by Matthew Brett and
-other developers\cite{multibuild}.  Currently, scripts using
-\texttt{multibuild} are written for the continuous integration
-platforms Travis-CI (for Linux and Mac OSX) and Appveyor
-(for Windows).
+other developers (\url{https://github.com/matthew-brett/multibuild}).
+Currently, scripts using \texttt{multibuild} are written for
+the continuous integration platforms Travis-CI (for Linux and Mac OSX)
+and Appveyor (for Windows).
 
@@
 \texttt{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
-for most mathematical functions.\cite{NEP13}
+for most mathematical functions (\url{https://numpy.org/neps/nep-0013-ufunc-overrides.html}).
 In 2019 this was expanded more generally with the inclusion of the
@@
 These two protocols allow providers of array objects to be interoperable
-with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
+with the NumPy API: their arrays work correctly with almost all NumPy
+functions (\url{https://numpy.org/neps/nep-0018-array-function-protocol.html}).
 For the users relying on specialized array projects it means that even though
@@
 world, Summit and Sierra, both have IBM POWER9 architectures
-\cite{top500nov2019}. In late 2018, Astra, the first ARM-based
-supercomputer to enter the TOP500 list, went into production\cite{
-astra-wiki}. Furthermore, over 100 billion ARM processors have been
-produced as of 2017\cite{arm-architecture}, making it the most 
-widely used instruction set architecture in the world.
+(\url{https://www.top500.org/lists/2019/11/}).
+In late 2018, Astra, the first ARM-based supercomputer
+to enter the TOP500 list, went into production
+(\url{https://en.wikichip.org/wiki/supercomputers/astra}).
+Furthermore, over 100 billion ARM processors have been produced as
+of 2017 (\url{https://en.wikipedia.org/wiki/ARM_architecture}),
+making it the most widely used instruction set architecture
+in the world.
 
@@
 This satisfies one of the major components of
-improved CI testing laid out in a version of our roadmap
-\cite{numpy-roadmap}---specifically, ``CI for more exotic
+improved CI testing laid out in a version of our
+roadmap---specifically, ``CI for more exotic
 platforms."
 
-PEP 599\cite{PEP599} lays out a plan for new Python binary wheel
+PEP 599 (\url{https://www.python.org/dev/peps/pep-0599/})
+lays out a plan for new Python binary wheel
 distribution support, \texttt{manylinux2014}, that adds

### 2020-05-29 scripts/fetch-pr-data.py
Remove old files

@@ -1,80 +1,0 @@
-#!/usr/bin/env python3
-
-import sys
-
-import github
-import pickle
-
-
-token = sys.argv[1]
-
-# Using a token/login is required to download these statistics:
-g = github.Github(token)
-numpy = g.get_repo('numpy/numpy')
-
-# Closed will ignore PRs that are not yet merged...
-merged_pulls = numpy.get_pulls(state="closed")
-
-NUMPY_ORG_MEMBER_IDS = {u.id for u in numpy.get_collaborators()}
-
-try:
-    with open("pr_data.pkl", "rb") as pickle_file:
-        pr_list = pickle.load(pickle_file)
-
-    if "number" not in pr_list[0]:
-        for pr in pr_list:
-            pr["number"] = int(pr["url"].replace("https://github.com/numpy/numpy/pull/", ""))
-
-except Exception as e:
-    print("no data loaded:", e)
-    pr_list = []
-
-known = {pr["number"] for pr in pr_list}
-
-
-for pr in merged_pulls:
-    if pr.number in known:
-        print("Already knew PR number:", pr.number)
-        continue
-
-    if not pr.base.label == "numpy:master":
-        # skip non-master (usually not interesting w.r.t. review)
-        continue
-
-    if not pr.merged:
-        continue
-
-    user = pr.user
-    pr_author_id = user.id
-    pr_author = user.login
-
-    merger = pr.merged_by
-    try:
-
-        merged_by = merger.login
-        merged_by_id = merger.id
-    except:
-        merged_by = merger
-        merged_by_id = None
-
-    info = {
-        "number": pr.number,
-        "title": pr.title,
-        # There is probably a better way to get the URL...
-        "url": pr.url.replace("//api.", "//").replace("/pulls/", "/pull/").replace("/repos/", "/"),
-        "created_at": pr.created_at,
-        "merged_at": pr.merged_at,
-        "merged_by": merged_by,
-        "merged_by_id": merged_by_id,
-        "author": pr_author,
-        "author_id": pr_author_id,
-        "is_collaborator": pr_author_id in NUMPY_ORG_MEMBER_IDS,
-        "merge_commit_sha": pr.merge_commit_sha,
-        "labels": [l.name for l in pr.labels],
-        }
-    print(info)
-    pr_list.append(info)
-
-
-with open("pr_data.pkl", "wb") as pickle_file:
-    pickle.dump(pr_list, pickle_file)

### 2020-05-29 scripts/plot_merged_prs.py

@@ -1,181 +1,0 @@
-#!/usr/bin/env python3
-
-import numpy as np
-import pickle
-import datetime
-import pandas
-
-from matplotlib import pyplot as plt
-
-from pandas.plotting import register_matplotlib_converters
-register_matplotlib_converters()
-
-# Choose either "CURRENT_MAINTAINERS" (list of those with commit rights
-# and users who had commit rights based on having done a merge),
-# or "HAS_MERGED_A_PR" which defines a "Maintainer" as anyone who has
-# merged a PR _before_ their PR was merged.
-DEFINE_MAINTAINERS_BY = "CURRENT_MAINTAINERS"  #  "HAS_MERGED_A_PR"
-
-
-# List of maintainers (NumPy users with commit rights)
-CURRENT_MAINTAINERS = """
-    ahaldane
-    certik
-    charris
-    cournape
-    eric-wieser
-    FrancescAlted
-    jaimefrio
-    jjhelmus
-    juliantaylor
-    mhvk
-    mwiebe
-    njsmith
-    pv
-    rgommers
-    rkern
-    seberg
-    shoyer
-    stefanv
-    teoliphant
-    matthew-brett
-    mattip
-    tylerjereddy
-    WarrenWeckesser
-    """.split()
-assert(len(CURRENT_MAINTAINERS) == 23)
-
-# Users who have merged a PR and/or served on the Steering Council
-OLDER_MAINTAINERS = """
-    argriffing
-    timcera
-    pletnes
-    jennystone
-""".split()
-CURRENT_MAINTAINERS.extend(OLDER_MAINTAINERS)
-
-
-def is_current_maintainer(user_login):
-    return user_login in CURRENT_MAINTAINERS
-
-# Filled using the date they first merged a PR.
-MAINTAINERS_SINCE = {}
-
-with open("pr_data.pkl", "rb") as pickle_file:
-    pr_list = pickle.load(pickle_file)
-
-
-def maintainer_by_first_merge(pr_list):
-    maintainer_since = MAINTAINERS_SINCE
-    names = {}
-    for pr in pr_list:
-        merger = pr["merged_by_id"]
-        if merger is None:
-            print("PR merged by unknown user:", pr["number"])
-        if pr["merged_by"] == "homu":
-            continue
-
-        if merger not in maintainer_since:
-            maintainer_since[merger] = pr["merged_at"]
-            names[merger] = pr["merged_by"]
-            continue
-
-
[…]

### 2020-05-29 move_figures.py
Move figures to end of submitted document

@@ -1,0 +1,33 @@
+#!/usr/bin/env python3
+
+"""
+Move figures to the end of the file.
+"""
+
+import sys
+import os
+import re
+import bs4
+
+import subprocess
+from subprocess import PIPE
+
+
+if len(sys.argv) != 2:
+    print("Usage: move_figures.py document.tex")
+    sys.exit(0)
+
+tex_file = sys.argv[1]
+
+with open(tex_file, 'rb') as f:
+    data = f.read().decode('utf-8')
+
+regexp = re.compile(r'(\\begin\{figure[*]?}.*?\\end\{figure[*]?})',
+                    re.DOTALL|re.MULTILINE)
+
+figures = re.findall(regexp, data)
+data = re.sub(regexp, '', data)
+data = data.replace('%% FIGURES HERE %%', '\n\n'.join(figures))
+
+with open(tex_file, 'w') as f:
+    f.write(data)

### 2020-06-01 body.tex
Remove old files
Fix typos

@@
   \centering
-  \includegraphics[width=\textwidth]{static/sketches/concepts}
+  \includegraphics[width=\textwidth]{concepts}
   \caption{\textbf{The NumPy array incorporates several fundamental array concepts.}
@@
   \centering
-  \includegraphics[width=.5\textwidth]{static/sketches/ecosystem}
+  \includegraphics[width=.5\textwidth]{ecosystem}
   \caption{\textbf{NumPy is the base of the scientific Python ecosystem.}
@@
 this foundation to provide several, widely used \emph{technique specific}
-libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
+libraries \cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
 that in turn underlay numerous \emph{domain specific} projects
@@
   \centering
-  \includegraphics[width=\textwidth]{static/sketches/protocol}
+  \includegraphics[width=\textwidth]{protocol}
 

### 2020-06-01 methods.tex
Fix typo
Remove old files
Fix typos

--- 2020-05-25 methods.tex
@@
 More details about the data structure are given in ``The NumPy array:
-a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.
+a structure for efficient numerical computation'' \cite{vanderwalt2011numpy}.
 
@@
 NumPy adopted an official Governance Document on October~5,
-2015\cite{NumPyProjectGovernance}.
+2015 \cite{NumPyProjectGovernance}.
 Project decisions are usually made by consensus of interested contributors.
@@
 % https://github.com/numpy/numpy/pull/11865
-NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
+NumPy's official Code of Conduct was approved on September~1, 2018 \cite{NumPyCodeofConduct}.
 In brief, we strive to:
@@
     \centering
-    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
+    \includegraphics[width=0.9\linewidth]{PRs-using-CURRENT_MAINTAINERS.pdf}
     \caption{\textbf{Number of pull requests merged into the NumPy master branch for each
@@
 
-A Python \emph{wheel}\cite{PEP427} is a standard file format for
+A Python \emph{wheel} \cite{PEP427} is a standard file format for
 distributing Python libraries.  In addition to Python code, a
@@
 \texttt{multibuild} suite of tools developed by Matthew Brett and
-other developers\cite{multibuild}.  Currently, scripts using
+other developers \cite{multibuild}.  Currently, scripts using
 \texttt{multibuild} are written for the continuous integration
@@
 \texttt{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
-for most mathematical functions.\cite{NEP13}
+for most mathematical functions \cite{NEP13}.
 In 2019 this was expanded more generally with the inclusion of the
@@
 These two protocols allow providers of array objects to be interoperable
-with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
+with the NumPy API: their arrays work correctly with almost all NumPy functions \cite{NEP18}.
 For the users relying on specialized array projects it means that even though
@@
 \cite{pcg64} (\texttt{PCG64}). The three other bit generators are a 64-bit version
-of the Philox generator\cite{random123} (\texttt{Philox}), Chris Doty-Humphrey's
-Small Fast Chaotic generator\cite{practrand} (\texttt{SFC64}), and the 32-bit
-Mersenne Twister\cite{mt19937} (\texttt{MT19937}) which has been used in older
+of the Philox generator \cite{random123} (\texttt{Philox}), Chris Doty-Humphrey's
+Small Fast Chaotic generator \cite{practrand} (\texttt{SFC64}), and the 32-bit
+Mersenne Twister \cite{mt19937} (\texttt{MT19937}) which has been used in older
 versions of NumPy.\footnote{The
@@
 generators (\texttt{AESCtr}) and generators that expose hardware random number
-generators (\texttt{RDRAND})\cite{randomgen}.} Bit generators provide
+generators (\texttt{RDRAND}) \cite{randomgen}.} Bit generators provide
 functions, exposed both in Python and C, for generating random integer
@@
 variates from common distributions were implemented, including the Ziggurat
-method for normal, exponential and gamma variates\cite{ziggurat}, and Lemire's
-method for bounded random integer generation\cite{lemire}. The \texttt{Generator}
+method for normal, exponential and gamma variates \cite{ziggurat}, and Lemire's
+method for bounded random integer generation \cite{lemire}. The \texttt{Generator}
 is more similar to the legacy \texttt{RandomState}, and its API is substantially
@@
 \cite{top500nov2019}. In late 2018, Astra, the first ARM-based
-supercomputer to enter the TOP500 list, went into production\cite{
-astra-wiki}. Furthermore, over 100 billion ARM processors have been
-produced as of 2017\cite{arm-architecture}, making it the most 
+supercomputer to enter the TOP500 list, went into production
+\cite{astra-wiki}. Furthermore, over 100 billion ARM processors have been
+produced as of 2017 \cite{arm-architecture}, making it the most 
 widely used instruction set architecture in the world.
@@
 
-PEP 599\cite{PEP599} lays out a plan for new Python binary wheel
+PEP 599 \cite{PEP599} lays out a plan for new Python binary wheel
 distribution support, \texttt{manylinux2014}, that adds

### 2020-06-01 methods.tex
Mention lapack history and note it is an external tool
This tries to address gh-72. I expanded the list of functionality
a bit, since it seemed strange to quite as concise if the paragraph
is longer. Hoewver, now the first paragraph might be very concise
since it uses the old style.

--- 2020-05-26 methods.tex
@@
 
-NumPy's linear algebra library includes functions for: solving linear
-systems of equations; computing various functions of a matrix, including
-the determinant, the norm, the inverse, and the pseudo-inverse;
-computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
-and more.
+NumPy's linear algebra library provides an interface to established
+linear algebra methods. These include solving linear systems of equations
+and computing various matrix properties, such as the determinant,
+the norm, the inverse, and the pseudo-inverse.
+As well as computing the Cholesky, eigenvalue, and singular value decompositions
+of a matrix and further functions.
+For best performance and reliability the algorithms are provided by leveraging
+an external implemention of the LAPACK interface \cite{LAPACK}.
+LAPACK itself was first released in 1992 and is based on the EISPACK and LINPACK
+open-source libraries. It is part of Netlib, a repository of
+mathematical software, papers and databases and has a long history of open,
+community-wide development. \cite{dongarra1990lapack,dongarra2008netlib}
 
@@
 
-A Python \emph{wheel} ({https://www.python.org/dev/peps/pep-0427/})
+A Python \emph{wheel} (\url{https://www.python.org/dev/peps/pep-0427/})
 is a standard file format for distributing Python libraries.

### 2020-06-02 paper.tex
Remove URLs from references
Revert default build to 1 file
Fix typo
Remove old files
Merge tex files

--- 2020-05-25 paper.tex
@@
-\documentclass[fleqn,10pt,twocolumn]{wlscirep}
+\documentclass[twocolumn]{article}
 
-% Packages
-\usepackage{moreverb} % for verbatim ouput (for wordcount)
+\usepackage{listings, textcomp}
+\usepackage{graphicx,xcolor}
+\bibliographystyle{ieeetr}
+\RequirePackage[colorlinks=true, allcolors=blue]{hyperref}
+
+\lstset{language=Python}
+
+\lstset{
+   basicstyle=\ttfamily\footnotesize,
+   xleftmargin=\parindent,
+   breaklines=true,
+   captionpos=b,
+   commentstyle=\color{gray},
+   keywordstyle=\color{blue},
+   stringstyle=\color{red},
+   upquote=true
+}
 
 \usepackage{caption}
-\captionsetup{%
+\captionsetup{
    figurename=Fig.,
@@
 
-\usepackage[switch]{lineno}
-\linenumbers
-
+\usepackage{titlesec}
+\titleformat*{\section}{\large\bfseries}
 \title{Array Programming with NumPy}
 
-\input{authors}
+\RequirePackage{authblk}
+\setlength{\affilsep}{1.5em}
+\renewcommand\Authfont{\fontsize{12}{12}\usefont{OT1}{phv}{b}{n}}
+\renewcommand\Affilfont{\fontsize{8}{10}\usefont{OT1}{phv}{m}{n}}
 
-\keywords{Scientific computing, Python, Mathematics, Array Programming}
-
-\begin{abstract}
-\input{summary} 
-\end{abstract}
+\author[1]{Charles R. Harris}
+\author[2,3,4,*]{K. Jarrod Millman}
+\author[5,2,4,*]{St\'efan J. van der Walt}
+\author[6,*]{Ralf Gommers}
+\author[7]{Pauli Virtanen}
+\author[8]{David Cournapeau}
+\author[9]{Eric Wieser}
+\author[10]{Julian Taylor}
+\author[4]{Sebastian Berg}
+\author[11]{Nathaniel J. Smith}
+\author[12]{Robert Kern}
+\author[4]{Matti Picus}
+\author[13]{Stephan Hoyer}
+\author[14]{Marten H. van Kerkwijk}
+\author[2,15]{Matthew Brett}
+\author[16]{Allan Haldane}
+\author[17]{Jaime Fern\'andez del R\'io}
+\author[18,19]{Mark Wiebe}
+\author[6,20,21]{Pearu Peterson}
+\author[22,23]{Pierre G\'erard-Marchant}
+\author[24]{Kevin Sheppard}
+\author[25]{Tyler Reddy}
+\author[4]{Warren Weckesser}
+\author[6]{Hameer Abbasi}
+\author[26]{Christoph Gohlke}
+\author[6]{Travis E. Oliphant}
+\affil[1]{Independent Researcher, Logan, Utah, USA}
+\affil[2]{Brain Imaging Center, University of California, Berkeley, Berkeley, CA, USA}
+\affil[3]{Division of Biostatistics, University of California, Berkeley, Berkeley, CA, USA}
+\affil[4]{Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA, USA}
+\affil[5]{Applied Mathematics, Stellenbosch University, Stellenbosch, South Africa}
+\affil[6]{Quansight LLC, Austin, TX, USA}
+\affil[7]{Department of Physics and Nanoscience Center, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
+\affil[8]{Mercari JP, Tokyo, Japan}
+\affil[9]{Department of Engineering, University of Cambridge, Cambridge, UK}
+\affil[10]{Independent Researcher, Karlsruhe, Germany}
+\affil[11]{Independent Researcher, Berkeley, CA, USA}
+\affil[12]{Enthought, Inc., Austin, TX, USA}
+\affil[13]{Google Research, Mountain View, CA, USA}
+\affil[14]{Department of Astronomy \& Astrophysics, University of Toronto, Toronto, ON, Canada}
+\affil[15]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
+\affil[16]{Department of Physics, Temple University, Philadelphia, PA, USA}
+\affil[17]{Google, Zurich, Switzerland}
+\affil[18]{Department of Physics and Astronomy, The University of British Columbia, Vancouver, BC, Canada}
+\affil[19]{Amazon, Seattle, Washington, USA}
+\affil[20]{Independent Researcher, Saue, Estonia}
+\affil[21]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
+\affil[22]{Department of Biological and Agricultural Engineering, University of Georgia, Athens, GA}
+\affil[23]{France-IX Services, Paris, France}
+\affil[24]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[25]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[26]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
+\affil[*]{millman@berkeley.edu, stefanv@berkeley.edu, ralf.gommers@gmail.com}
 
@@
 
-% Count of words
-%\verbatiminput{wordcount.tex}
+\twocolumn[
+  \begin{@twocolumnfalse}
+    \begin{abstract}
+     Array programming provides a powerful, compact, expressive syntax for accessing,
+     manipulating, and operating on data in vectors, matrices, and
+     higher-dimensional arrays \cite{iverson1980notation}.
+     NumPy is the primary array programming library for the Python language
+     \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
+     It plays an essential role in research analysis pipelines in fields as
+     diverse as physics, chemistry, astronomy, geoscience, biology, psychology,
+     material science, engineering, finance, and economics.
+     For example, in astronomy, NumPy was an important part of the software stack used
+     in the discovery of gravitational waves \cite{abbott2016observation}
+     and the first imaging of a black hole \cite{eht-imaging}.
+     Here we show how a few fundamental array concepts lead to a simple and
+     powerful programming paradigm for organizing, exploring, and analyzing
+     scientific data.
+     NumPy is the foundation upon which the entire scientific Python
+     universe is constructed. It is so pervasive that several projects,
+     targeting audiences with specialized needs, have developed their own
+     NumPy-like interfaces and array objects. Because of its central position in the
+     ecosystem, NumPy increasingly plays the role of an interoperability layer
+     between these new array computation libraries.
+     \end{abstract}
+    \vspace{1cm}
+  \end{@twocolumnfalse}
+]
 
-\input{body}
+\begin{figure*}[h]
+  \centering
+  \includegraphics[width=\textwidth]{concepts}
+  \caption{\textbf{The NumPy array incorporates several fundamental array concepts.}
+    \textbf{a,} The NumPy array data structure and its associated metadata fields.
+    \textbf{b,} Indexing an array with slices and steps. These
+  operations return a \emph{view} of the original data.
+    \textbf{c,} Indexing an array with masks, scalar coordinates, or
+  other arrays, so that it returns a copy of the original data. In the
+  bottom example, an array is indexed with other arrays; this broadcasts the indexing arguments before performing the lookup.
+    \textbf{d,} Vectorization efficiently applies operations to groups
+  of elements.
+    \textbf{e,} Broadcasting in the multiplication of two-dimensional arrays.
+    \textbf{f,} Reduction operations act along one or more axes. In this
+    example, an array is summed along select axes to produce a vector, or along two axes consecutively to
+    produce a scalar.
+    \textbf{g,} Example NumPy code, illustrating some of these concepts.
+   }
+  \label{fig:array-concepts}
+\end{figure*}
 
-\bibliography{references}
+Two Python array packages existed before NumPy.
+The Numeric package began in the mid-1990s and provided an array object and array-aware functions
+in Python, written in C, and linking to standard fast implementations of linear
+algebra \cite{dubois1996numerical,Numericmanual}.
+One of its earliest uses was to steer C++ applications for
+inertial confinement fusion research at
+Lawrence Livermore National Laboratory \cite{yang1997steering}.
+To handle large astronomical images coming from the Hubble Space Telescope,
+a reimplementation of Numeric, called Numarray, added
+support for structured arrays, flexible indexing, memory mapping, byte-order variants,
+more efficient memory use, flexible IEEE error handling capabilities, and
+better type casting rules \cite{greenfield2003numarray}.
+While Numarray was highly compatible with Numeric, the two packages had enough
+differences that it divided the community, until 2005, when NumPy emerged as a
+``best of both worlds'' unification \cite{oliphant2006guide}---combining
+Numarray's features with Numeric's performance on small arrays and its rich C
+\emph{Application Programming Interface} (API).
+
+Now, fifteen years later, NumPy underpins almost every Python library that does scientific or
+numerical computation including SciPy \cite{virtanen2020scipy},
+Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
+scikit-learn \cite{pedregosa2011scikit}, and
+scikit-image \cite{vanderwalt2014scikit}.
+It is a community-developed, open-source library, which provides a
+multidimensional Python array object along with array-aware functions
+that operate on it.
+Because of its inherent simplicity, the NumPy array is
+the \emph{de facto} exchange format for array data in Python.
+
+NumPy operates on in-memory arrays using the CPU. To utilize modern,
+specialized storage and hardware, there has been a recent
+proliferation of Python array packages. Unlike with the Numarray and
+Numeric divide, it is now much harder for these new libraries to
+fracture the user community---given how much work already builds
+on top of NumPy.  However, to provide the ecosystem with access to
+new and exploratory technologies, NumPy is transitioning into a
+central coordinating mechanism that specifies a well-defined array
+programming API and
+dispatches it, as appropriate, to specialized array implementations.
+
+\section*{NumPy arrays}
+
+The NumPy array is a data structure that efficiently stores and accesses
+multidimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, and
+enables a wide variety of scientific computation.
+It consists of a pointer to memory, along with metadata used to interpret the
+data stored there, notably {\em data type}, {\em shape}, and {\em strides}
+(Fig.~\ref{fig:array-concepts}a).
+
+The \emph{data type} describes the nature of elements stored in an array.
+An array has a single data type, and each array element occupies the same
+number of bytes in memory.
+Examples of data types include real and complex numbers (of lower and higher
+precision), strings, timestamps, and pointers to Python objects.
+
+The \emph{shape} of an array determines the number of elements along each axis,
+and the number of axes is the array's dimensionality.
+For example, a vector of numbers can be stored as a one-dimensional array of
+shape $N$, while color videos are four-dimensional arrays of shape
+$(T, M, N, 3)$.
+
+\emph{Strides} are necessary to interpret computer memory, which stores elements
+linearly, as multidimensional arrays.
+It describes the number of bytes to move forward in memory to jump from row to
+row, column to column, and so forth.
+Consider, for example, a 2-D array of floating-point numbers with shape
+$(4, 3)$, where each element occupies 8 bytes in memory.
+To move between consecutive columns, we need to jump forward 8 bytes in memory,
+and to access the next row $3 \times 8 = 24$ bytes.
+The strides of that array are therefore $(24, 8)$.  NumPy can
+store arrays in either C or Fortran memory order, iterating
+first over either rows or columns.  This allows external libraries
+written in those languages to access NumPy array data in memory directly.
+
+Users interact with NumPy arrays using {\em indexing} (to access
+subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$
+for vectorized operations and $@$ for matrix multiplication), as well as {\em array-aware functions};
+together, these provide an easily readable, expressive, high-level API for
+array programming, while NumPy
+deals with the underlying mechanics of making operations fast.
+
+\emph{Indexing} an array returns single elements, subarrays, or elements that satisfy
+a specific condition (Fig.~\ref{fig:array-concepts}b).
+Arrays can even be indexed using other arrays (Fig.~\ref{fig:array-concepts}c).
+Wherever possible, indexing that retrieves a subarray returns a {\em view} on
+the original array, such that data is shared between the two arrays.
+This provides a powerful way to operate on subsets of array data while
+limiting memory usage.
+
+To complement the array syntax, NumPy includes functions that perform
+\emph{vectorized} calculations on arrays, including arithmetic, statistics, and
+trigonometry (Fig.~\ref{fig:array-concepts}d).
+Vectorization---operating on whole arrays rather than their individual
+elements---is essential to array programming.
+This means that operations that would take many tens of lines to express in
+languages such as C can often be implemented as a single, clear Python
+expression.
+This results in concise code and frees users to focus on the details of
+their analysis, while NumPy handles looping over array elements near-optimally,
+taking into consideration, for example, strides, to best utilize the
+computer's fast cache memory.
+
+When performing a vectorized operation (such as addition) on two arrays with
+the same shape, it is clear what should happen.
+Through \emph{broadcasting}, NumPy allows the dimensions to differ, while
+still producing results that appeal to intuition.
+A trivial example is the addition of a scalar value to an array, but broadcasting also
+generalizes to more complex examples such as scaling each column of an array
+or generating a grid of coordinates.
+In broadcasting, one or both arrays are virtually duplicated (that is, without
+copying any data in memory), so that the shapes of the operands match
+(Fig.~\ref{fig:array-concepts}d).
+Broadcasting is also applied when an array is indexed using arrays of
+indices (Fig.~\ref{fig:array-concepts}c).
+
+Other array-aware functions, such as \texttt{sum}, \texttt{mean}, and \texttt{maximum}, perform
+element-by-element \emph{reductions}, aggregating results across one,
+multiple, or all axes of a single array.
+For example, summing an $n$-dimensional array over $d$ axes results in a
+$(n-d)$-dimensional array (Fig.~\ref{fig:array-concepts}f).
+
+NumPy also includes array-aware functions for creating, reshaping, concatenating, and padding
+arrays; searching, sorting, and counting data; and reading and writing files.
+It provides extensive support for generating pseudorandom numbers,
+includes an assortment of probability distributions, and
+performs accelerated linear algebra, utilizing one of several backends
+such as OpenBLAS \cite{wang2013augem,xianyi2012model} or Intel MKL optimized
+for the CPUs at hand.
+
+Altogether, the combination of a simple in-memory array
+representation, a syntax that closely mimics mathematics, and a
+variety of array-aware utility functions forms a productive and
+powerfully expressive array programming language.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.5\textwidth]{ecosystem}
+  \caption{\textbf{NumPy is the base of the scientific Python ecosystem.}
+   Essential libraries and projects that depend on NumPy's API gain access to new array
+   implementations that support NumPy's array protocols (Fig.~\ref{fig:array-protocol}).
+  }
+  \label{fig:ecosystem}
+\end{figure}
+
+\section*{Scientific Python ecosystem}
+
+Python is an open-source, general-purpose, interpreted programming language
+well-suited to standard programming tasks such as cleaning data,
+interacting with web resources, and parsing text.
+Adding fast array operations and linear algebra allows scientists to do all
+their work within a single language---and one that has the advantage of
+being famously easy to learn and teach, as witnessed by its adoption
+as a primary learning language in many universities.
+
+Even though NumPy is not part of Python's standard library,
+it benefits from a good relationship with the Python developers.
+Over the years, the Python language has added new features and
+special syntax so that NumPy would have a more succinct and 
+easier to read array notation.
+Since it is not part of the standard library, NumPy is able to
+dictate its own release policies and development patterns.
+
+SciPy and Matplotlib are tightly coupled with NumPy---in terms of
+history, development, and use.
+SciPy provides fundamental algorithms for scientific computing,
+including mathematical, scientific, and engineering routines.
+Matplotlib generates publication-ready figures and visualizations.
+The combination of NumPy, SciPy, and Matplotlib, together with
+an advanced interactive environment like IPython \cite{perez2007ipython},
+or Jupyter \cite{Kluyver:2016aa}, provides a solid foundation for array
+programming in Python.
+The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
+this foundation to provide several, widely used \emph{technique specific}
+libraries \cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
+that in turn underlay numerous \emph{domain specific} projects
+\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
+NumPy, at the base of the ecosystem of array-aware libraries,
+sets documentation standards, provides array testing infrastructure,
+and adds build support for Fortran and other compilers.
+
+Many research groups have designed large,
+complex scientific libraries, which add \emph{application specific} functionality
+to the ecosystem.
+For example, the \texttt{eht-imaging} library \cite{chael2019ehtim} developed by
+the Event Horizon Telescope collaboration for radio interferometry imaging,
+analysis, and simulation, relies on many lower-level components of the scientific Python
+ecosystem.
+NumPy arrays are used to store and manipulate numerical data at every step
+in the processing chain: from raw data through calibration and image
+reconstruction.
+SciPy supplies tools for general image processing tasks such as
+filtering and image alignment, while scikit-image, an image processing
+library that extends SciPy, provides higher-level functionality such as
+edge filters and Hough transforms.
+The \texttt{scipy.optimize} module performs mathematical optimization.
+NetworkX \cite{SciPyProceedings_11}, a package for complex
+network analysis, is used to verify image comparison consistency.
+Astropy \cite{astropy:2013, astropy:2018} handles standard
+astronomical file formats and computes time/coordinate transformations.
+Matplotlib is used to visualize data and to generate the final image of the black hole.
+
+The interactive environment created by the array programming
+foundation along with the surrounding ecosystem of tools---inside of
+IPython or Jupyter---is ideally suited to exploratory data analysis.
+Users fluidly inspect, manipulate, and visualize their data, and
+rapidly iterate to refine programming statements. These statements are
+then stitched together into imperative or functional programs, or
+notebooks containing both computation and narrative.
+Scientific computing beyond exploratory work is often done in a text editor
+or an integrated development environment (IDEs) such as Spyder.
+This rich and productive environment has made Python popular
+for scientific research.
+
+To complement this facility for exploratory work and rapid
+prototyping, NumPy has developed a culture of
+employing time-tested software engineering practices to improve collaboration and
+reduce error \cite{millman2014developing}.  This culture is not only
+adopted by leaders in the project but also enthusiastically taught to
+newcomers. The NumPy team was early in adopting distributed revision
+control and code review to improve collaboration on code, and
+continuous testing that runs an extensive battery of automated tests for
+every proposed change to NumPy.  The project also has comprehensive,
+high-quality documentation, integrated with the source
+code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
+
+This culture of using best practices for producing reliable scientific software
+has been adopted by the ecosystem of libraries that build on NumPy.
+For example, in a recent award given by the Royal Astronomical Society to
+Astropy, they state:
+\begin{quotation}
+\noindent\emph{The Astropy Project has provided hundreds of junior scientists
+with experience in professional-standard software development practices
+including use of version control, unit testing, code review and issue tracking
+procedures. This is a vital skill set for modern researchers that is often
+missing from formal university education in physics or astronomy.}
+\end{quotation}
+Community members explicitly work to address this lack of formal education
+through courses and workshops
+\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
+
+The recent rapid growth of data science, machine learning, and
+artificial intelligence has further and dramatically boosted the usage of
+scientific Python.  Examples of its significant application, such as the
+\texttt{eht-imaging} library, now exist in almost every discipline in the natural and social
+sciences.  These tools have become \emph{the primary}
+software environment in many fields.  NumPy and its ecosystem are commonly
+taught in university courses, boot camps, and summer schools, and are
+at the focus of community conferences and workshops worldwide.
+
+NumPy and its API have become truly ubiquitous.
+
+\section*{Array proliferation and interoperability}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=\textwidth]{protocol}
+
+  \caption{\textbf{NumPy's API and array protocols expose new arrays to the ecosystem.}
+     In this example, NumPy's \texttt{mean} function is called on a Dask
+     array.  The call succeeds by dispatching to the appropriate library implementation
+     (i.e., Dask in this case) and results in a new Dask array.  Compare this
+     code to the example code in Fig.~\ref{fig:array-concepts}g.
+  }\label{fig:array-protocol}
+
+\end{figure*}
+
+NumPy provides in-memory, multidimensional, homogeneously typed
+(i.e., single pointer and strided) arrays on CPUs.  It runs on machines
+ranging from embedded devices to the world's largest supercomputers,
+with performance approaching that of compiled languages.
+For most its existence, NumPy addressed the vast majority of
+array computation use cases.
+
+However, scientific data sets now routinely exceed the memory capacity of a single machine and may
+be stored on multiple machines or in the cloud.
+In addition, the recent need to accelerate deep learning and artificial intelligence applications
+has led to the emergence of specialized accelerator hardware,
+including graphics processing units (GPUs), tensor processing units (TPUs),
+and field-programmable gate arrays (FPGAs).
+Due to its in-memory data model, NumPy is currently unable to
+utilize such storage and specialized hardware directly.  However, both
+distributed data and the parallel execution of GPUs, TPUs, and FPGAs map well
+to the \emph{paradigm} of array programming: a gap, therefore, existed between
+available modern hardware architectures and the tools necessary to
+leverage their computational power.
+
+The community's efforts to fill this gap led to a
+proliferation of new array implementations. For example, each deep learning framework created
+its own arrays; PyTorch \cite{NEURIPS2019_9015},
+Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
+and JAX arrays all have the
+capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
+to allow for additional performance optimizations.  SciPy and PyData/Sparse both
+provide sparse arrays---which typically contain few non-zero values and store
+only those in memory for efficiency.
+In addition, there are projects that build on top of NumPy arrays as a data
+container and \textit{extend} its capabilities.  Distributed arrays are
+made possible that way by Dask, and labeled arrays---referring to dimensions of
+an array by name rather than by index for clarity, compare \texttt{x[:,~1]} vs.
+\texttt{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.
+
+Such libraries often mimic the NumPy API, because it lowers the
+barrier to entry for newcomers and provides the wider community with a
+stable array programming interface. This, in turn, prevents disruptive
+schisms like the divergence of Numeric and Numarray.
+But exploring new ways of working with arrays is experimental by nature
+and, in fact, several promising libraries---such as Theano and Caffe---have
+already ceased development. And each time that a user
+decides to try a new technology, they must
+change import statements and ensure that the new library implements
+all the parts of the NumPy API they currently use.
+
+Ideally, operating on specialized arrays using NumPy functions or semantics would
+simply work, so that users could write code once, and would then benefit
+from switching between NumPy arrays, GPU arrays, distributed arrays,
+and so forth, as appropriate.
+To support array operations between external array objects, NumPy
+therefore added the capability to act as a central coordination
+mechanism with a well-specified API (Fig.~\ref{fig:ecosystem}).
+
+To facilitate this \emph{interoperability}, NumPy provides
+``protocols'' (or contracts of operation), that allow for specialized arrays to be
+passed to NumPy functions (Fig.~\ref{fig:array-protocol}).
+NumPy, in turn, dispatches operations to the originating library, as required.
+Over four hundred of the most popular
+NumPy functions are supported.
+The protocols are implemented by widely used libraries such as Dask, CuPy,
+xarray, and PyData/Sparse.
+Thanks to these developments, users can now, for example, scale
+their computation from a single machine to distributed systems using Dask.
+The protocols also \textit{compose} well, allowing users to redeploy NumPy
+code at scale on distributed, multi-GPU systems via, for instance, CuPy arrays embedded in Dask
+arrays. Using NumPy's high-level API, users can leverage highly parallel code
+execution on multiple systems with millions of cores, all with minimal code
+changes \cite{entschev2019}.
+
+These array protocols are now a key feature of NumPy, and are expected to only
+increase in importance.  As with the rest of NumPy, we iteratively refine and
+add protocol designs to improve utility and simplify adoption.
+
+\section*{Discussion}
+
+NumPy combines the expressive power of \emph{array programming},
+the performance of C, and
+the readability, usability, and versatility of Python in a mature,
+well-tested, well-documented, and community-developed library.
+Libraries in the scientific Python ecosystem provide fast implementations of most important algorithms.
+Where extreme optimization is warranted, compiled languages such as
+Cython \cite{behnel2011cython}, Numba \cite{Lam:2015:NLP:2833157.2833162},
+and Pythran \cite{guelton2015pythran}, that
+extend Python and transparently accelerate bottlenecks, can be
+used.
+Because of NumPy's simple memory model, it is easy to write low-level, hand-optimized code, usually in C
+or Fortran, to manipulate NumPy arrays and pass them back to
+Python.
+Furthermore, using array protocols, it is possible to utilize the full
+spectrum of specialized hardware acceleration with minimal changes to
+existing code.
+
+NumPy was initially developed by students, faculty, and researchers to
+provide an advanced, open-source array programming library for Python,
+which was free to use and unencumbered by license servers, dongles, and the like.
+There was a sense of building something consequential together,
+for the benefit of many others.  Participating in
+such an endeavor, within a welcoming community of like-minded
+individuals, held a powerful attraction for many early contributors.
+
+These user-developers frequently had to write code from scratch to solve
+their own or their colleagues' problems---often in low-level languages
+that precede Python, like Fortran \cite{dongarra2008netlib} and C.
+To them, the advantages of an interactive, high-level array library
+were evident. The design of this new tool was informed by other
+powerful interactive programming languages for scientific computing
+such as Basis \cite{dubois1989basis}, Yorick \cite{munro1995using}, R \cite{ihaka1996r},
+and APL \cite{iverson1962programming},
+as well as commercial languages and environments like IDL and {MATLAB}.
+
+What began as an attempt to add an array object to Python became the
+foundation of a vibrant ecosystem of tools.  Now, a large amount of
+scientific work depends on NumPy being correct, fast, and stable.  It
+is no longer a small community project, but is core scientific
+infrastructure.
+
+The developer culture has matured: while initial development was
+highly informal, NumPy now has a roadmap and a process for proposing
+and discussing large changes.  The project has formal governance
+structures and is fiscally sponsored by NumFOCUS, a nonprofit that
+promotes open practices in research, data, and scientific computing.
+Over the past few years, the project attracted its first funded
+development, sponsored by the Moore and Sloan Foundations, and
+received an award as part of the Chan Zuckerberg Initiative's
+Essentials of Open Source Software program.  With this funding, the
+project was (and is) able to have sustained focus over multiple months
+to implement substantial new features and improvements.  That said, it
+still depends heavily on contributions made by graduate students and
+researchers in their free time.
+
+NumPy is no longer \emph{just} the foundational array library underlying
+the scientific Python ecosystem, but has also become the standard API
+for tensor computation and a central coordinating mechanism between
+array types and technologies in Python. Work continues to expand on and
+improve these interoperability features.
+
+Over the next decade, we will face several challenges.  New devices will be
+developed, and existing specialized hardware will evolve, to meet diminishing
+returns on Moore's law.  There will be more, and a wider variety of, data
+science practitioners, a significant proportion of whom will be using NumPy.
+The scale of scientific data gathering will continue to expand, with the
+adoption of devices and instruments such as light sheet microscopes and the
+Large Synoptic Survey Telescope (LSST) \cite{jenness2018lsst}.  New generation
+languages, interpreters, and compilers, such as Rust
+\cite{10.1145/2692956.2663188}, Julia \cite{Julia-2017}, and LLVM
+\cite{LLVM:CGO04}, will invent and determine the viability of new concepts and
+data structures.
+
+Through various mechanisms described in this paper, NumPy is poised to
+embrace such a changing landscape, and to continue playing a leading
+role in interactive scientific computation.  To do so will require
+sustained funding from government, academia, and industry.  But,
+importantly, it will also need a new generation of graduate students
+and other developers to engage, to build a NumPy that meets the needs
+of the next decade of data science.
+
+\begin{thebibliography}{10}
+
+\bibitem{iverson1980notation}
+K.~E. Iverson, ``Notation as a tool of thought,'' {\em Communications of the
+  ACM}, vol.~23, p.~444–465, Aug. 1980.
+
+\bibitem{dubois2007guest}
+P.~F. Dubois, ``{P}ython: Batteries included,'' {\em Computing in Science \&
+  Engineering}, vol.~9, no.~3, pp.~7--9, 2007.
+
+\bibitem{oliphant2007python}
+T.~E. Oliphant, ``{P}ython for scientific computing,'' {\em Computing in
+  Science \& Engineering}, vol.~9, pp.~10--20, May-June 2007.
+
+\bibitem{millman2011python}
+K.~J. Millman and M.~Aivazis, ``{P}ython for scientists and engineers,'' {\em
+  Computing in Science \& Engineering}, vol.~13, no.~2, pp.~9--12, 2011.
+
+\bibitem{perez2011python}
+F.~P\'{e}rez, B.~E. Granger, and J.~D. Hunter, ``{P}ython: an ecosystem for
+  scientific computing,'' {\em Computing in Science \& Engineering}, vol.~13,
+  no.~2, pp.~13--21, 2011.
+
+\bibitem{abbott2016observation}
+B.~P. Abbott, R.~Abbott, T.~Abbott, M.~Abernathy, F.~Acernese, K.~Ackley,
+  C.~Adams, T.~Adams, P.~Addesso, R.~Adhikari, {\em et~al.}, ``Observation of
+  gravitational waves from a binary black hole merger,'' {\em Physical Review
+  Letters}, vol.~116, no.~6, p.~061102, 2016.
+
+\bibitem{eht-imaging}
+A.~A. Chael, M.~D. Johnson, R.~Narayan, S.~S. Doeleman, J.~F. Wardle, and K.~L.
+  Bouman, ``High-resolution linear polarimetric imaging for the event horizon
+  telescope,'' {\em The Astrophysical Journal}, vol.~829, no.~1, p.~11, 2016.
+
+\bibitem{dubois1996numerical}
+P.~F. Dubois, K.~Hinsen, and J.~Hugunin, ``Numerical {P}ython,'' {\em Computers
+  in Physics}, vol.~10, no.~3, pp.~262--267, 1996.
+
+\bibitem{Numericmanual}
+D.~Ascher, P.~F. Dubois, K.~Hinsen, J.~Hugunin, and T.~E. Oliphant, ``An open
+  source project: Numerical {P}ython,'' 2001.
+
+\bibitem{yang1997steering}
+T.-Y. Yang, G.~Furnish, and P.~F. Dubois, ``Steering object-oriented scientific
+  computations,'' in {\em Proceedings of TOOLS USA 97. International Conference
+  on Technology of Object Oriented Systems and Languages}, pp.~112--119, IEEE,
+  1997.
+
+\bibitem{greenfield2003numarray}
+P.~Greenfield, J.~T. Miller, J.~Hsu, and R.~L. White, ``numarray: A new
+  scientific array package for {Python},'' {\em PyCon DC}, 2003.
+
+\bibitem{oliphant2006guide}
+T.~E. Oliphant, {\em Guide to {NumPy}}.
+\newblock Trelgol Publishing USA, 1st~ed., 2006.
+
+\bibitem{virtanen2020scipy}
+P.~Virtanen, R.~Gommers, T.~E. Oliphant, M.~Haberland, T.~Reddy, D.~Cournapeau,
+  E.~Burovski, P.~Peterson, W.~Weckesser, J.~Bright, S.~J. van~der Walt,
+  M.~Brett, J.~Wilson, K.~J. Millman, N.~Mayorov, A.~R.~J. Nelson, E.~Jones,
+  R.~Kern, E.~Larson, C.~J. Carey, I.~Polat, Y.~Feng, E.~W. Moore,
+  J.~VanderPlas, D.~Laxalde, J.~Perktold, R.~Cimrman, I.~Henriksen, E.~A.
+  Quintero, C.~R. Harris, A.~M. Archibald, A.~H. Ribeiro, F.~Pedregosa, P.~van
+  Mulbregt, and {SciPy 1.0 Contributors}, ``{SciPy} 1.0---fundamental
+  algorithms for scientific computing in {Python},'' {\em Nature Methods},
+  vol.~17, pp.~261--272, 2020.
+
+\bibitem{hunter2007matplotlib}
+J.~D. Hunter, ``Matplotlib: A {2D} graphics environment,'' {\em Computing in
+  Science \& Engineering}, vol.~9, no.~3, pp.~90--95, 2007.
+
+\bibitem{mckinney-proc-scipy-2010}
+W.~McKinney, ``Data structures for statistical computing in {Python},'' in {\em
+  Proceedings of the 9th Python in Science Conference} (S.~van~der Walt and
+  J.~Millman, eds.), pp.~51 -- 56, 2010.
+
+\bibitem{pedregosa2011scikit}
+F.~Pedregosa, G.~Varoquaux, A.~Gramfort, V.~Michel, B.~Thirion, O.~Grisel,
+  M.~Blondel, P.~Prettenhofer, R.~Weiss, V.~Dubourg, J.~Vanderplas, A.~Passos,
+  D.~Cournapeau, M.~Brucher, M.~Perrot, and {\'E}.~Duchesnay, ``Scikit-learn:
+  Machine learning in {P}ython,'' {\em Journal of Machine Learning Research},
+  vol.~12, no.~Oct, pp.~2825--2830, 2011.
+
+\bibitem{vanderwalt2014scikit}
+S.~van~der Walt, J.~L. Sch{\"o}nberger, J.~Nunez-Iglesias, F.~Boulogne, J.~D.
+  Warner, N.~Yager, E.~Gouillart, T.~Yu, and {the scikit-image contributors},
+  ``{scikit-image}: image processing in {P}ython,'' {\em PeerJ}, vol.~2,
+  p.~e453, 2014.
+
+\bibitem{vanderwalt2011numpy}
+S.~van~der Walt, S.~C. Colbert, and G.~Varoquaux, ``The {NumPy} array: a
+  structure for efficient numerical computation,'' {\em Computing in Science \&
+  Engineering}, vol.~13, no.~2, pp.~22--30, 2011.
+
+\bibitem{wang2013augem}
+Q.~Wang, X.~Zhang, Y.~Zhang, and Q.~Yi, ``Augem: automatically generate high
+  performance dense linear algebra kernels on x86 cpus,'' in {\em SC'13:
+  Proceedings of the International Conference on High Performance Computing,
+  Networking, Storage and Analysis}, pp.~1--12, IEEE, 2013.
+
+\bibitem{xianyi2012model}
+Z.~Xianyi, W.~Qian, and Z.~Yunquan, ``Model-driven level 3 blas performance
+  optimization on loongson 3a processor,'' in {\em 2012 IEEE 18th International
+  Conference on Parallel and Distributed Systems}, pp.~684--691, IEEE, 2012.
+
+\bibitem{perez2007ipython}
+F.~P{\'e}rez and B.~E. Granger, ``{IP}ython: a system for interactive
+  scientific computing,'' {\em Computing in Science \& Engineering}, vol.~9,
+  no.~3, pp.~21--29, 2007.
+
+\bibitem{Kluyver:2016aa}
+T.~Kluyver, B.~Ragan-Kelley, F.~P{\'e}rez, B.~Granger, M.~Bussonnier,
+  J.~Frederic, K.~Kelley, J.~Hamrick, J.~Grout, S.~Corlay, P.~Ivanov, D.~Avila,
+  S.~Abdalla, and C.~Willing, ``{Jupyter Notebooks}---a publishing format for
+  reproducible computational workflows,'' in {\em Positioning and Power in
+  Academic Publishing: Players, Agents and Agendas} (F.~Loizides and
+  B.~Schmidt, eds.), pp.~87--90, IOS Press, 2016.
+
+\bibitem{SciPyProceedings_11}
+A.~A. Hagberg, D.~A. Schult, and P.~J. Swart, ``Exploring network structure,
+  dynamics, and function using {NetworkX},'' in {\em Proceedings of the 7th
+  Python in Science Conference} (G.~Varoquaux, T.~Vaught, and K.~J. Millman,
+  eds.), (Pasadena, CA USA), pp.~11--15, 2008.
+
+\bibitem{astropy:2013}
+{Astropy Collaboration}, T.~P. {Robitaille}, E.~J. {Tollerud}, P.~{Greenfield},
+  M.~{Droettboom}, E.~{Bray}, T.~{Aldcroft}, M.~{Davis}, A.~{Ginsburg}, A.~M.
+  {Price-Whelan}, W.~E. {Kerzendorf}, A.~{Conley}, N.~{Crighton}, K.~{Barbary},
+  D.~{Muna}, H.~{Ferguson}, F.~{Grollier}, M.~M. {Parikh}, P.~H. {Nair}, H.~M.
+  {Unther}, C.~{Deil}, J.~{Woillez}, S.~{Conseil}, R.~{Kramer}, J.~E.~H.
+  {Turner}, L.~{Singer}, R.~{Fox}, B.~A. {Weaver}, V.~{Zabalza}, Z.~I.
+  {Edwards}, K.~{Azalee Bostroem}, D.~J. {Burke}, A.~R. {Casey}, S.~M.
+  {Crawford}, N.~{Dencheva}, J.~{Ely}, T.~{Jenness}, K.~{Labrie}, P.~L. {Lim},
+  F.~{Pierfederici}, A.~{Pontzen}, A.~{Ptak}, B.~{Refsdal}, M.~{Servillat}, and
+  O.~{Streicher}, ``{Astropy: A community Python package for astronomy},'' {\em
+  Astronomy \& Astrophysics}, vol.~558, p.~A33, Oct. 2013.
+
+\bibitem{astropy:2018}
+A.~M. {Price-Whelan}, B.~M. {Sip{\H{o}}cz}, H.~M. {G{\"u}nther}, P.~L. {Lim},
+  S.~M. {Crawford}, S.~{Conseil}, D.~L. {Shupe}, M.~W. {Craig}, N.~{Dencheva},
+  A.~{Ginsburg}, J.~T. {VanderPlas}, L.~D. {Bradley},
+  D.~{P{\'e}rez-Su{\'a}rez}, M.~{de Val-Borro}, P.~{Paper Contributors}, T.~L.
+  {Aldcroft}, K.~L. {Cruz}, T.~P. {Robitaille}, E.~J. {Tollerud},
+  A.~{Coordination Committee}, C.~{Ardelean}, T.~{Babej}, Y.~P. {Bach},
+  M.~{Bachetti}, A.~V. {Bakanov}, S.~P. {Bamford}, G.~{Barentsen}, P.~{Barmby},
+  A.~{Baumbach}, K.~L. {Berry}, F.~{Biscani}, M.~{Boquien}, K.~A. {Bostroem},
+  L.~G. {Bouma}, G.~B. {Brammer}, E.~M. {Bray}, H.~{Breytenbach},
+  H.~{Buddelmeijer}, D.~J. {Burke}, G.~{Calderone}, J.~L. {Cano
+  Rodr{\'\i}guez}, M.~{Cara}, J.~V.~M. {Cardoso}, S.~{Cheedella}, Y.~{Copin},
+  L.~{Corrales}, D.~{Crichton}, D.~{D{\textquoteright}Avella}, C.~{Deil},
+  {\'E}.~{Depagne}, J.~P. {Dietrich}, A.~{Donath}, M.~{Droettboom}, N.~{Earl},
+  T.~{Erben}, S.~{Fabbro}, L.~A. {Ferreira}, T.~{Finethy}, R.~T. {Fox}, L.~H.
+  {Garrison}, S.~L.~J. {Gibbons}, D.~A. {Goldstein}, R.~{Gommers}, J.~P.
+  {Greco}, P.~{Greenfield}, A.~M. {Groener}, F.~{Grollier}, A.~{Hagen},
+  P.~{Hirst}, D.~{Homeier}, A.~J. {Horton}, G.~{Hosseinzadeh}, L.~{Hu}, J.~S.
+  {Hunkeler}, {\v{Z}}.~{Ivezi{\'c}}, A.~{Jain}, T.~{Jenness}, G.~{Kanarek},
+  S.~{Kendrew}, N.~S. {Kern}, W.~E. {Kerzendorf}, A.~{Khvalko}, J.~{King},
+  D.~{Kirkby}, A.~M. {Kulkarni}, A.~{Kumar}, A.~{Lee}, D.~{Lenz}, S.~P.
+  {Littlefair}, Z.~{Ma}, D.~M. {Macleod}, M.~{Mastropietro}, C.~{McCully},
+  S.~{Montagnac}, B.~M. {Morris}, M.~{Mueller}, S.~J. {Mumford}, D.~{Muna},
+  N.~A. {Murphy}, S.~{Nelson}, G.~H. {Nguyen}, J.~P. {Ninan}, M.~{N{\"o}the},
+  S.~{Ogaz}, S.~{Oh}, J.~K. {Parejko}, N.~{Parley}, S.~{Pascual}, R.~{Patil},
+  A.~A. {Patil}, A.~L. {Plunkett}, J.~X. {Prochaska}, T.~{Rastogi}, V.~{Reddy
+  Janga}, J.~{Sabater}, P.~{Sakurikar}, M.~{Seifert}, L.~E. {Sherbert},
+  H.~{Sherwood-Taylor}, A.~Y. {Shih}, J.~{Sick}, M.~T. {Silbiger},
+  S.~{Singanamalla}, L.~P. {Singer}, P.~H. {Sladen}, K.~A. {Sooley},
+  S.~{Sornarajah}, O.~{Streicher}, P.~{Teuben}, S.~W. {Thomas}, G.~R.
+  {Tremblay}, J.~E.~H. {Turner}, V.~{Terr{\'o}n}, M.~H. {van Kerkwijk}, A.~{de
+  la Vega}, L.~L. {Watkins}, B.~A. {Weaver}, J.~B. {Whitmore}, J.~{Woillez},
+  V.~{Zabalza}, and A.~{Contributors}, ``{The Astropy Project: Building an
+  Open-science Project and Status of the v2.0 Core Package},'' {\em The
+  Astronomical Journal}, vol.~156, p.~123, Sept. 2018.
+
+\bibitem{cock2009biopython}
+P.~J. Cock, T.~Antao, J.~T. Chang, B.~A. Chapman, C.~J. Cox, A.~Dalke,
+  I.~Friedberg, T.~Hamelryck, F.~Kauff, B.~Wilczynski, and M.~J.~L. de~Hoon,
+  ``Biopython: freely available {Python} tools for computational molecular
+  biology and bioinformatics,'' {\em Bioinformatics}, vol.~25, no.~11,
+  pp.~1422--1423, 2009.
+
+\bibitem{millman2007analysis}
+K.~J. Millman and M.~Brett, ``Analysis of functional {M}agnetic {R}esonance
+  {I}maging in {P}ython,'' {\em Computing in Science \& Engineering}, vol.~9,
+  no.~3, pp.~52--55, 2007.
+
+\bibitem{sunpy2015}
+T.~{SunPy Community}, S.~J. {Mumford}, S.~{Christe}, D.~{P{\'e}rez-Su{\'a}rez},
+  J.~{Ireland}, A.~Y. {Shih}, A.~R. {Inglis}, S.~{Liedtke}, R.~J. {Hewett},
+  F.~{Mayer}, K.~{Hughitt}, N.~{Freij}, T.~{Meszaros}, S.~M. {Bennett},
+  M.~{Malocha}, J.~{Evans}, A.~{Agrawal}, A.~J. {Leonard}, T.~P. {Robitaille},
+  B.~{Mampaey}, J.~{Iv{\'a}n Campos-Rozo}, and M.~S. {Kirk}, ``{SunPy---Python
+  for solar physics},'' {\em Computational Science and Discovery}, vol.~8,
+  p.~014009, Jan. 2015.
+
+\bibitem{2018EGUGA..2012146H}
+J.~{Hamman}, M.~{Rocklin}, and R.~{Abernathy}, ``{Pangeo: A Big-data Ecosystem
+  for Scalable Earth System Science},'' in {\em EGU General Assembly Conference
+  Abstracts}, EGU General Assembly Conference Abstracts, p.~12146, Apr 2018.
+
+\bibitem{chael2019ehtim}
+A.~A. Chael, K.~L. Bouman, M.~D. Johnson, R.~Narayan, S.~S. Doeleman, J.~F.
+  Wardle, L.~L. Blackburn, K.~Akiyama, M.~Wielgus, C.-k. Chan, {\em et~al.},
+  ``ehtim: Imaging, analysis, and simulation software for radio
+  interferometry,'' {\em Astrophysics Source Code Library}, 2019.
+
+\bibitem{millman2014developing}
+K.~J. Millman and F.~P{\'e}rez, ``Developing open-source scientific practice,''
+  {\em Implementing Reproducible Research. CRC Press, Boca Raton, FL},
+  pp.~149--183, 2014.
+
+\bibitem{vanderwalt2008scipy}
+S.~van~der Walt, ``The {SciPy} documentation project (technical overview),'' in
+  {\em Proceedings of the 7th {P}ython in Science Conference (SciPy 2008)}
+  (G.~Varoquaux, T.~Vaught, and K.~J. Millman, eds.), pp.~27--28, 2008.
+
+\bibitem{harrington2008scipy}
+J.~Harrington, ``The {SciPy} documentation project,'' in {\em Proceedings of
+  the 7th {P}ython in Science Conference (SciPy 2008)} (G.~Varoquaux,
+  T.~Vaught, and K.~J. Millman, eds.), pp.~33--35, 2008.
+
+\bibitem{harrington2009scipy}
+J.~Harrington and D.~Goldsmith, ``Progress report: {NumPy} and {SciPy}
+  documentation in 2009,'' in {\em Proceedings of the 8th {P}ython in Science
+  Conference (SciPy 2009)} (G.~Varoquaux, S.~van~der Walt, and K.~J. Millman,
+  eds.), pp.~84--87, 2009.
+
+\bibitem{wilson-software-carpentry}
+G.~Wilson, ``Software carpentry: Getting scientists to write better code by
+  making them more productive,'' {\em Computing in Science \& Engineering},
+  November--December 2006.
+
+\bibitem{hannay-scientific-software-survey}
+J.~E. Hannay, H.~P. Langtangen, C.~MacLeod, D.~Pfahl, J.~Singer, and G.~Wilson,
+  ``How do scientists develop and use scientific software?,'' in {\em Proc.
+  2009 ICSE Workshop on Software Engineering for Computational Science and
+  Engineering}, 2009.
+
+\bibitem{millman2018teaching}
+K.~J. Millman, M.~Brett, R.~Barnowski, and J.-B. Poline, ``Teaching
+  computational reproducibility for neuroimaging,'' {\em Frontiers in
+  Neuroscience}, vol.~12, p.~727, 2018.
+
+\bibitem{NEURIPS2019_9015}
+A.~Paszke, S.~Gross, F.~Massa, A.~Lerer, J.~Bradbury, G.~Chanan, T.~Killeen,
+  Z.~Lin, N.~Gimelshein, L.~Antiga, A.~Desmaison, A.~Kopf, E.~Yang, Z.~DeVito,
+  M.~Raison, A.~Tejani, S.~Chilamkurthy, B.~Steiner, L.~Fang, J.~Bai, and
+  S.~Chintala, ``Pytorch: An imperative style, high-performance deep learning
+  library,'' in {\em Advances in Neural Information Processing Systems 32}
+  (H.~Wallach, H.~Larochelle, A.~Beygelzimer, F.~d\textquotesingle
+  Alch\'{e}-Buc, E.~Fox, and R.~Garnett, eds.), pp.~8024--8035, Curran
+  Associates, Inc., 2019.
+
+\bibitem{abadi2016tensorflow}
+M.~Abadi, A.~Agarwal, P.~Barham, E.~Brevdo, Z.~Chen, C.~Citro, G.~S. Corrado,
+  A.~Davis, J.~Dean, M.~Devin, {\em et~al.}, ``Tensorflow: Large-scale machine
+  learning on heterogeneous distributed systems,'' {\em arXiv preprint
+  arXiv:1603.04467}, 2016.
+
+\bibitem{chen2015mxnet}
+T.~Chen, M.~Li, Y.~Li, M.~Lin, N.~Wang, M.~Wang, T.~Xiao, B.~Xu, C.~Zhang, and
+  Z.~Zhang, ``Mxnet: A flexible and efficient machine learning library for
+  heterogeneous distributed systems,'' {\em arXiv preprint arXiv:1512.01274},
+  2015.
+
+\bibitem{hoyer2017xarray}
+S.~Hoyer and J.~Hamman, ``xarray: {N-D} labeled arrays and datasets in
+  {Python},'' {\em Journal of Open Research Software}, vol.~5, no.~1, 2017.
+
+\bibitem{entschev2019}
+P.~Entschev, ``{D}istributed multi-{GPU} computing with {D}ask, {C}u{P}y and
+  {RAPIDS}.'' EuroPython 2019, 2019.
+
+\bibitem{behnel2011cython}
+S.~Behnel, R.~Bradshaw, C.~Citro, L.~Dalcin, D.~S. Seljebotn, and K.~Smith,
+  ``Cython: The best of both worlds,'' {\em Computing in Science \&
+  Engineering}, vol.~13, no.~2, pp.~31--39, 2011.
+
+\bibitem{Lam:2015:NLP:2833157.2833162}
+S.~K. Lam, A.~Pitrou, and S.~Seibert, ``Numba: A {LLVM}-based {P}ython {JIT}
+  compiler,'' in {\em Proceedings of the Second Workshop on the LLVM Compiler
+  Infrastructure in HPC}, LLVM '15, (New York, NY, USA), pp.~7:1--7:6, ACM,
+  2015.
+
+\bibitem{guelton2015pythran}
+S.~Guelton, P.~Brunet, M.~Amini, A.~Merlini, X.~Corbillon, and A.~Raynaud,
+  ``Pythran: Enabling static optimization of scientific python programs,'' {\em
+  Computational Science \& Discovery}, vol.~8, no.~1, p.~014001, 2015.
+
+\bibitem{dongarra2008netlib}
+J.~Dongarra, G.~H. Golub, E.~Grosse, C.~Moler, and K.~Moore, ``Netlib and
+  na-net: Building a scientific computing community,'' {\em IEEE Annals of the
+  History of Computing}, vol.~30, no.~2, pp.~30--41, 2008.
+
+\bibitem{dubois1989basis}
+P.~F. Dubois, ``The basis system,'' tech. rep., Lawrence Livermore National
+  Laboratory, CA (USA), 1989.
+\newblock UCRL-MA-118543, Parts I-VI.
+
+\bibitem{munro1995using}
+D.~H. Munro and P.~F. Dubois, ``Using the yorick interpreted language,'' {\em
+  Computers in Physics}, vol.~9, no.~6, pp.~609--615, 1995.
+
+\bibitem{ihaka1996r}
+R.~Ihaka and R.~Gentleman, ``R: a language for data analysis and graphics,''
+  {\em Journal of Computational and Graphical Statistics}, vol.~5, no.~3,
+  pp.~299--314, 1996.
+
+\bibitem{iverson1962programming}
+K.~E. Iverson, ``A programming language,'' in {\em Proceedings of the May 1-3,
+  1962, Spring Joint Computer Conference}, pp.~345--351, 1962.
+
+\bibitem{jenness2018lsst}
+T.~Jenness, F.~Economou, K.~Findeisen, F.~Hernandez, J.~Hoblitt, K.~S.
+  Krughoff, K.~Lim, R.~H. Lupton, F.~Mueller, W.~O'Mullane, {\em et~al.},
+  ``Lsst data management software development practices and tools,'' in {\em
+  Software and Cyberinfrastructure for Astronomy V}, vol.~10707, p.~1070709,
+  International Society for Optics and Photonics, 2018.
+
+\bibitem{10.1145/2692956.2663188}
+N.~D. Matsakis and F.~S. Klock, ``The rust language,'' {\em Ada Letters},
+  vol.~34, pp.~103--104, Oct. 2014.
+
+\bibitem{Julia-2017}
+J.~Bezanson, A.~Edelman, S.~Karpinski, and V.~B. Shah, ``Julia: A fresh
+  approach to numerical computing,'' {\em SIAM {R}eview}, vol.~59, no.~1,
+  pp.~65--98, 2017.
+
+\bibitem{LLVM:CGO04}
+C.~Lattner and V.~Adve, ``{LLVM}: A compilation framework for lifelong program
+  analysis and transformation,'' (San Jose, CA, USA), pp.~75--88, Mar 2004.
+
+\bibitem{peterson2009f2py}
+P.~Peterson, ``{F2PY}: a tool for connecting {F}ortran and {P}ython programs,''
+  {\em International Journal of Computational Science and Engineering}, vol.~4,
+  no.~4, pp.~296--305, 2009.
+
+\bibitem{NumPyProjectGovernance}
+{The NumPy Project Community}, ``{NumPy} project governance,'' 2015.
+
+\bibitem{NumPyCodeofConduct}
+{The NumPy Project Community}, ``{NumPy} code of conduct,'' 2018.
+
+\bibitem{PEP427}
+D.~Holth, ``Pep 427 -- the wheel binary package format 1.0,'' 2012.
+
+\bibitem{multibuild}
+{Brett, M. et al}, ``{multibuild},'' 2016.
+
+\bibitem{NEP13}
+B.~Griffith, P.~Virtanen, N.~Smith, M.~{van Kerkwijk}, and S.~Hoyer, ``{NEP} 13
+  -- a mechanism for overriding ufuncs,'' 2013.
+
+\bibitem{NEP18}
+S.~Hoyer, M.~Rocklin, M.~{van Kerkwijk}, H.~Abbasi, and E.~Wieser, ``{NEP} 18
+  -- a dispatch mechanism for numpy's high level array functions,'' 2018.
+
+\bibitem{pcg64}
+M.~E. O'Neill, ``Pcg: A family of simple fast space-efficient statistically
+  good algorithms for random number generation,'' Tech. Rep. HMC-CS-2014-0905,
+  Harvey Mudd College, Claremont, CA, Sept. 2014.
+
+\bibitem{random123}
+J.~K. Salmon, M.~A. Moraes, R.~O. Dror, and D.~E. Shaw, ``Parallel random
+  numbers: As easy as 1, 2, 3,'' in {\em Proceedings of 2011 International
+  Conference for High Performance Computing, Networking, Storage and Analysis},
+  SC '11, (New York, NY, USA), pp.~16:1--16:12, ACM, 2011.
+
+\bibitem{practrand}
+C.~Doty-Humphrey, ``Practrand, version 0.94.''
+
+\bibitem{mt19937}
+M.~Matsumoto and T.~Nishimura, ``{Mersenne Twister}: A 623-dimensionally
+  equidistributed uniform pseudo-random number generator,'' {\em ACM
+  Transactions on Modeling and Computer Simulation}, vol.~8, pp.~3--30, Jan.
+  1998.
+
+\bibitem{randomgen}
+K.~Sheppard, B.~Duvenhage, P.~de~Buyl, and D.~A. Ham, ``bashtage/randomgen:
+  Release 1.16.2,'' Apr. 2019.
+
+\bibitem{ziggurat}
+G.~Marsaglia and W.~W. Tsang, ``The ziggurat method for generating random
+  variables,'' {\em Journal of Statistical Software, Articles}, vol.~5, no.~8,
+  pp.~1--7, 2000.
+
+\bibitem{lemire}
+D.~Lemire, ``Fast random integer generation in an interval,'' {\em ACM
+  Transactions on Modeling and Computer Simulation}, vol.~29, pp.~1--12, Jan
+  2019.
+
+\bibitem{top500nov2019}
+{top500}, ``Top 10 sites for november 2019,'' 2019.
+
+\bibitem{astra-wiki}
+{wikichip}, ``Astra - supercomputers,'' 2019.
+
+\bibitem{arm-architecture}
+{Wikipedia}, ``Arm architecture,'' 2019.
+
+\bibitem{numpy-roadmap}
+{NumPy Developers}, ``Numpy roadmap,'' 2019.
+
+\bibitem{PEP599}
+{Dustin Ingram}, ``Pep 599 -- the manylinux2014 platform tag,'' 2019.
+
+\end{thebibliography}
+
+\newpage
+
+\section*{Methods}
+
+We use Git for version control and GitHub as the public hosting service for our
+official \emph{upstream} repository (\url{https://github.com/numpy/numpy}).
+We each work in our own copy (or fork) of the project and use the
+upstream repository as our integration point.
+To get new code into the upstream repository, we use GitHub's
+pull request (PR) mechanism.
+This allows us to review code before integrating it as well as to run a
+large number of tests on the modified code to ensure that the changes
+do not break expected behavior.
+
+We also use GitHub's issue tracking system to collect and triage problems and
+proposed improvements.
+
+\subsection*{Library organization}
+
+Broadly, the NumPy library consists of the following parts:
+the NumPy array data structure \texttt{ndarray}; the so-called \emph{universal functions};
+a set of library functions for manipulating arrays and doing scientific
+computation; infrastructure libraries for unit tests and Python package
+building; and the program \texttt{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
+The \texttt{ndarray} and the universal functions are generally considered
+the core of the library.
+In the following, we give a brief summary of these components of the
+library.
+
+\paragraph{\emph{Core.}}  The \texttt{ndarray} data structure and the
+universal functions make up the core of NumPy.
+
+The \texttt{ndarray} is the data structure at the heart of NumPy.
+The data structure stores regularly strided homogeneous data types
+inside a contiguous block memory, allowing for the efficient representation
+of $n$-dimensional data.
+More details about the data structure are given in ``The NumPy array:
+a structure for efficient numerical computation'' \cite{vanderwalt2011numpy}.
+
+The \emph{universal functions}, or more concisely, \emph{ufuncs},
+are functions written in C that implement efficient looping over
+NumPy arrays. An important feature of ufuncs is the built-in
+implementation of \emph{broadcasting}.  For example, the function
+\texttt{arctan2(x, y)} is a ufunc that accepts two values and computes
+$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
+the ufunc will take care of looping over the dimensions of the inputs
+in such a way that if, say, \texttt{x} is a 1-D array with length 3, and
+\texttt{y} is a 2-D array with shape $2 \times 1$, the output will be
+an array with shape $2 \times 3$ (Fig.~\ref{fig:array-concepts}c).
+The ufunc machinery takes care
+of calling the function with all the appropriate combinations of
+input array elements to complete the output array.
+The elementary arithmetic operations of addition, multiplication, etc.,
+are implemented as ufuncs, so that broadcasting also applies to expressions
+such as \texttt{x + y * z}.
+
+\paragraph{\emph{Computing libraries.}}
+NumPy provides a large library of functions for array manipulation
+and scientific computing, including functions for: creating, reshaping,
+concatenating, and padding arrays; searching, sorting and counting data
+in arrays; computing elementary statistics, such as the mean, median,
+variance, and standard deviation; file I/O; and more.
+
+A suite of functions for computing the \emph{fast Fourier transform (FFT)}
+and its inverse is provided.
+
+NumPy's linear algebra library includes functions for: solving linear
+systems of equations; computing various functions of a matrix, including
+the determinant, the norm, the inverse, and the pseudo-inverse;
+computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
+and more.
+
+The random number generator library in NumPy provides alternative
+\emph{bit stream generators} that provide the core function of generating
+random integers.
+A higher-level generator class that implements an assortment of
+probability distributions is provided. It includes the beta, gamma
+and Weibull distributions, the univariate and multivariate normal
+distributions, and more.
+
+\paragraph{\emph{Infrastructure libraries.}} NumPy provides utilities
+for writing tests and for building Python packages.
+
+The \texttt{testing} subpackage provides functions such as
+\texttt{assert\_allclose(actual, desired)} that may be used in
+test suites for code that uses NumPy arrays.
+
+NumPy provides the subpackage \texttt{distutils} which includes functions and classes
+to facilitate configuration, installation, and packaging of libraries depending on NumPy.
+These can be used, for example, when publishing to the PyPI website.
+
+\paragraph{\emph{F2PY.}}  The program \texttt{f2py} is a tool for
+building NumPy-aware Python wrappers of Fortran functions.
+NumPy itself does not use any Fortran code;  F2PY is part of NumPy
+for historical reasons.
+
+\subsection*{Governance}
+
+NumPy adopted an official Governance Document on October~5,
+2015 \cite{NumPyProjectGovernance}.
+Project decisions are usually made by consensus of interested contributors.
+This means that, for most decisions, everyone is entrusted with veto power.
+A Steering Council, currently composed of 12~members, facilitates this
+process and oversees daily development of the project by contributing code
+and reviewing contributions from the community.
+
+NumPy's official Code of Conduct was approved on September~1, 2018 \cite{NumPyCodeofConduct}.
+In brief, we strive to:
+\emph{be open};
+\emph{be empathetic, welcoming, friendly, and patient};
+\emph{be collaborative};
+\emph{be inquisitive}; and
+\emph{be careful in the words that we choose}.
+The Code of Conduct also specifies how breaches can be reported and outlines
+the process for responding to such reports.
+
+\subsection*{Funding}
+
+In 2017, NumPy received its first large grants totaling 1.3M USD from the
+Gordon \& Betty Moore and the Alfred P. Sloan foundations.
+Stéfan van der Walt is the PI and manages four programmers working on the project.
+These two grants focus on addressing the technical debt accrued over the years and
+on setting in place standards and architecture to encourage more sustainable development.
+
+NumPy received a third grant for 195K USD from the Chan Zuckerberg
+Initiative at the end of 2019 with Ralf Gommers as the PI.
+This grant focuses on better serving NumPy's large number of beginning
+to intermediate level users and on growing the community of NumPy
+contributors.
+It will also provide support to OpenBLAS, on which NumPy depends for
+accelerated linear algebra.
+
+Finally, since May 2019 the project receives a small amount annually from
+Tidelift, which is used to fund things like documentation and website
+improvements.
+
+\subsection*{Developers}
+
+NumPy is currently maintained by a group of 23 contributors with commit rights
+to the NumPy code base. Out of these, 17 maintainers were active in
+2019, 4 of whom were paid to work on the project full-time.
+Additionally, there are a few long term developers who contributed and maintain
+specific parts of NumPy, but are not officially maintainers.
+
+Over the course of its history, NumPy has attracted PRs by 823 contributors.
+However, its development relies heavily on a small number
+of active maintainers, who share more than half of the contributions among
+themselves.
+
+At a release cycle of about every half year, the five recent releases in the years
+2018 and 2019 have averaged about 450~PRs each,\footnote{
+    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
+    All data provided here is based on the development which happened through GitHub
+    PRs. In some cases contributions by maintainers may not be categorized as such.}
+with each release attracting more than a hundred new contributors.
+Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
+master branch.
+Although the number of PRs being merged fluctuates,
+the plot indicates an increased number of contributions over the past
+years.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\linewidth]{PRs-using-CURRENT_MAINTAINERS.pdf}
+    \caption{\textbf{Number of pull requests merged into the NumPy master branch for each
+        quarter since 2012.} The total number of PRs is indicated with the
+        lower blue area showing the portion contributed by current or previous
+        maintainers.}\label{fig:prs-over-time}
+\end{figure}
+
+\subsection*{Community calls}
+
+The massive number of scientific Python packages that
+built on NumPy meant that it had an unusually high need for stability.
+So to guide our development we formalized the feature proposal process, and
+constructed a development roadmap with extensive input and feedback from the
+community.
+
+Weekly community calls alternate between triage and
+higher level discussion.  The calls not only involve developers from
+the community, but provide a venue for vendors and other external
+groups to provide input.  For example, after Intel produced a forked
+version of NumPy, one of their developers joined a call to discuss
+community concerns.
+
+\subsection*{NumPy enhancement proposals}
+
+Given the complexity of the codebase and the massive number of projects depending
+on it, large changes require careful planning and substantial work.
+NumPy Enhancement Proposals (NEPs) are modeled after
+Python Enhancement Proposals (PEPs) for ``proposing major new
+features, for collecting community input on an issue, and for
+documenting the design decisions that have gone into
+Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
+Since then there have been 19 proposed NEPS---6 have been implemented,
+4 have been accepted and are being implemented, 4 are under
+consideration, 3 have been deferred or superseded, and 2 have been rejected
+or withdrawn.
+
+\subsection*{Central role}
+
+NumPy plays a central role in building and standardizing much of the scientific
+Python community infrastructure.
+NumPy's docstring standard is now widely adopted.
+We are also now using the NEP system as a way to help coordinate the larger
+scientific Python community.
+For example, in NEP 29, we recommend, along with leaders from various other
+projects, that all projects across the Scientific Python ecosystem adopt a
+common ``time window-based'' policy for support of Python and NumPy versions.
+This standard will simplify downstream project and release planning.
+
+\subsection*{Wheels build system}
+
+A Python \emph{wheel} \cite{PEP427} is a standard file format for
+distributing Python libraries.  In addition to Python code, a
+wheel may include compiled C extensions and other binary data.
+This is important, because many libraries, including NumPy,
+require a C compiler and other build tools to build the software
+from the source code, making it difficult for many users to install
+the software on their own.  The introduction of wheels to the Python
+packaging system has made it much easier for users to install
+precompiled libraries.
+
+A GitHub repository containing scripts to build NumPy wheels has
+been configured so that a simple commit to the repository triggers
+an automated build system that creates NumPy wheels for several
+computer platforms, including Windows, Mac OSX and Linux.  The wheels
+are uploaded to a public server and made available for anyone to use.
+This system makes it easy for users to install precompiled versions
+of NumPy on these platforms.
+
+The technology that is used to build the wheels evolves continually.
+At the time this paper is being written, a key component is the
+\texttt{multibuild} suite of tools developed by Matthew Brett and
+other developers \cite{multibuild}.  Currently, scripts using
+\texttt{multibuild} are written for the continuous integration
+platforms Travis-CI (for Linux and Mac OSX) and Appveyor
+(for Windows).
+
+\subsection*{Recent technical improvements}
+
+With the recent infusion of funding and a clear process for coordinating with
+the developer community, we have been able to tackle a number of important
+large scale changes.
+We highlight two of those below, as well as changes made to our testing
+infrastructure to support hardware platforms used in large scale computing.
+
+\subsection*{Array function protocol}
+
+A vast number of projects are built on NumPy;
+these projects are consumers of the NumPy API.
+Over the last several years, a growing number of projects are providers of
+a \emph{NumPy-like API} and array objects targeting audiences with specialized
+needs beyond NumPy's capabilities.
+For example, the NumPy API is implemented by several popular tensor computation
+libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
+JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
+and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
+and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
+provide tensor APIs with NumPy-inspired semantics.
+It is also implemented in packages that support sparse arrays
+such as \texttt{scipy.sparse} and PyData/Sparse.
+Another notable example is Dask, a library for parallel computing in
+Python.  Dask adopts the NumPy API and therefore presents a familiar
+interface to existing NumPy users, while adding powerful abilities to
+parallelize and distribute tasks.
+
+The multitude of specialized projects creates the difficulty that consumers
+of these NumPy-like APIs write code specific to a single project and do not support
+all of the above array providers.
+This is a burden for users relying on the specialized array-like, since
+a tool they need may not work for them.
+It also creates challenges for end-users who need to transition
+from NumPy to a more specialized array.
+The growing multitude of specialized projects with NumPy-like APIs threatened
+to again fracture the scientific Python community.
+
+To address these issues NumPy has the goal of providing the fundamental
+API for \emph{interoperability} between the various NumPy-like APIs.
+An earlier step in this direction was the implementation of the
+\texttt{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
+for most mathematical functions \cite{NEP13}.
+In 2019 this was expanded more generally with the inclusion of the
+\texttt{\_\_array\_function\_\_} protocol into NumPy~1.17.
+These two protocols allow providers of array objects to be interoperable
+with the NumPy API: their arrays work correctly with almost all NumPy functions \cite{NEP18}.
+For the users relying on specialized array projects it means that even though
+much code is written specifically for NumPy arrays and uses the NumPy API as
+\texttt{import numpy as np}, it can nevertheless work for them.
+For example, here is how a CuPy GPU array can be passed through NumPy for
+processing, with all operations being dispatched back to CuPy:
+
+\begin{lstlisting}
+import numpy as np
+import cupy as cp
+
+x_gpu = cp.array([1, 2, 3])
+y = np.sum(x_gpu)  # Returns a GPU array
+\end{lstlisting}
+
+Similarly, user defined functions composed using NumPy can now be
+applied to, e.g., multi-node distributed Dask arrays:
+
+\begin{lstlisting}
+import numpy as np
+import dask.array as da
+
+def f(x):
+    """Function using NumPy API calls"""
+    y = np.tensordot(x, x.T)
+    return np.mean(np.log(y + 1))
+
+x_local = np.random.random([10000, 10000])  # random local array
+x_distr = da.random.random([10000, 10000])  # random distributed array
+
+f(x_local)  # returns a NumPy array
+f(x_distr)  # works, returns a Dask array
+\end{lstlisting}
+
+\subsection*{Random number generation}
+
+The NumPy \texttt{random} module provides pseudorandom numbers from a wide range of
+distributions. In legacy versions of NumPy, simulated random values are produced
+by a \texttt{RandomState} object that: handles seeding and state initialization;
+wraps the core pseudorandom number generator based on a Mersenne Twister
+implementation\footnote{to be precise, the standard 32-bit version of MT19937};
+interfaces with the underlying code that transforms random bits into
+variates from other distributions; and supplies a singleton instance exposed in
+the root of the random module.
+
+The \texttt{RandomState} object makes a compatibility guarantee so that a fixed
+seed and sequence of function calls produce the same set of values. This
+guarantee has slowed progress since improving the underlying code requires
+extending the API with additional keyword arguments. This guarantee continues to
+apply to \texttt{RandomState}.
+
+NumPy 1.17 introduced a new API for generating random numbers that use a more
+flexible structure that can be extended by libraries or end-users. The new API
+is built using components that separate the steps required to generate random
+variates. Pseudorandom bits are generated by a bit generator. These bits are
+then transformed into variates from complex distributions by a generator.
+Finally, seeding is handled by an object that produces sequences of high-quality
+initial values.
+
+Bit generators are simple classes that manage the state of an underlying
+pseudorandom number generator. NumPy ships with four bit generators. The default
+bit generator is a 64-bit implementation of the Permuted Congruential Generator
+\cite{pcg64} (\texttt{PCG64}). The three other bit generators are a 64-bit version
+of the Philox generator \cite{random123} (\texttt{Philox}), Chris Doty-Humphrey's
+Small Fast Chaotic generator \cite{practrand} (\texttt{SFC64}), and the 32-bit
+Mersenne Twister \cite{mt19937} (\texttt{MT19937}) which has been used in older
+versions of NumPy.\footnote{The
+\href{https://github.com/bashtage/randomgen}{randomgen project} supplies a wide
+range of alternative bit generators such as a cryptographic counter-based
+generators (\texttt{AESCtr}) and generators that expose hardware random number
+generators (\texttt{RDRAND}) \cite{randomgen}.} Bit generators provide
+functions, exposed both in Python and C, for generating random integer
+and floating point numbers.
+
+The \texttt{Generator} consumes one of the bit generators and produces variates
+from complicated distributions. Many improved methods for generating random
+variates from common distributions were implemented, including the Ziggurat
+method for normal, exponential and gamma variates \cite{ziggurat}, and Lemire's
+method for bounded random integer generation \cite{lemire}. The \texttt{Generator}
+is more similar to the legacy \texttt{RandomState}, and its API is substantially
+the same. The key differences all relate to state management, which has been
+delegated to the bit generator. The \texttt{Generator} does not make the same
+stream guarantee as the \texttt{RandomState} object, and so variates may differ
+across versions as improved generation algorithms are
+introduced.\footnote{Despite the removal of the compatibility guarantee, simple
+reproducibility across versions is encouraged, and minor changes that do not
+produce meaningful performance gains or fix underlying bug are not generally
+adopted.}
+
+Finally, a \texttt{SeedSequence} is used to initialize a bit generator. The seed
+sequence can be initialized with no arguments, in which case it reads entropy
+from a system-dependent provider, or with a user-provided seed. The seed
+sequence then transforms the initial set of entropy into a sequence of
+high-quality pseudorandom integers, which can be used to initialize multiple bit
+generators deterministically. The key feature of a seed sequence is that
+it can be used to spawn child \texttt{SeedSequence}s to initialize
+multiple distinct bit generators.
+This capability allows a seed sequence to facilitate large distributed applications
+where the number of workers required is not known. The sequences generated from
+the same initial entropy and spawns are fully deterministic to ensure
+reproducibility.
+
+The three components are combined to construct a complete random number
+generator.
+
+\begin{lstlisting}
+from numpy.random import (
+    Generator,
+    PCG64,
+    SeedSequence,
+)
+
+seq = SeedSequence(1030424547444117993331016959)
+pcg = PCG64(seq)
+gen = Generator(pcg)
+\end{lstlisting}
+
+This approach retains access to the seed sequence which can then be
+used to spawn additional generators.
+
+\begin{lstlisting}
+children = seq.spawn(2)
+gen_0 = Generator(PCG64(children[0]))
+gen_1 = Generator(PCG64(children[1]))
+\end{lstlisting}
+
+While this approach retains complete flexibility, the method
+\texttt{np.random.default\_rng} can be used to instantiate a \texttt{Generator} when
+reproducibility is not needed.
+
+The final goal of the new API is to improve extensibility. \texttt{RandomState} is
+a monolithic object that obscures all of the underlying state and functions. The
+component architecture is one part of the extensibility improvements. The
+underlying functions (written in C) which transform the output of a bit
+generator to other distributions are available for use in CFFI. This allows the
+same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
+Numba. Both the bit generators and the low-level functions can also be used in C
+or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
+NumPy source. Alternative approaches that avoid this extra step are being
+explored.}
+
+\subsection*{Testing on multiple architectures}
+
+At the time of writing the two fastest supercomputers in the
+world, Summit and Sierra, both have IBM POWER9 architectures
+\cite{top500nov2019}. In late 2018, Astra, the first ARM-based
+supercomputer to enter the TOP500 list, went into production
+\cite{astra-wiki}. Furthermore, over 100 billion ARM processors have been
+produced as of 2017 \cite{arm-architecture}, making it the most 
+widely used instruction set architecture in the world.
+
+Clearly there are motivations for a large scientific computing
+software library to support POWER and ARM architectures. We've extended
+our continuous integration (CI) testing to include \texttt{ppc64le}
+(POWER8 on Travis CI) and ARMv8 (on Shippable service). We also test
+with the s390x architecture (IBM Z CPUs on Travis CI) so that we
+can probe the behavior of our library on a big-endian machine.
+This satisfies one of the major components of
+improved CI testing laid out in a version of our roadmap
+\cite{numpy-roadmap}---specifically, ``CI for more exotic
+platforms."
+
+PEP 599 \cite{PEP599} lays out a plan for new Python binary wheel
+distribution support, \texttt{manylinux2014}, that adds
+support for a number of architectures supported by the CentOS
+Alternative Architecture Special Interest Group, including
+ARMv8, ppc64le, as well as s390x. We are thus well-positioned
+for a future where provision of binaries on these architectures
+will be expected for a library at the base of the ecosystem.
 

### 2020-06-02 methods.tex
Provide response to referee comment on random

@@
 
-A suite of functions for computing the \emph{fast Fourier transform (FFT)}
-and its inverse is provided.
-
 NumPy's linear algebra library provides an interface to established
@@
 mathematical software, papers and databases and has a long history of open,
-community-wide development. \cite{dongarra1990lapack,dongarra2008netlib}
+community-wide development \cite{dongarra1990lapack,dongarra2008netlib}.
 
@@
 distributions, and more.
+
+A suite of functions for computing the \emph{fast Fourier transform (FFT)}
+and its inverse is provided.
 
@@
 generators (\texttt{AESCtr}) and generators that expose hardware random number
-generators (\texttt{RDRAND})\cite{randomgen}.} Bit generators provide
+generators (\texttt{RDRAND}) \cite{randomgen}.} Bit generators provide
 functions, exposed both in Python and C, for generating random integer
-and floating point numbers.
+and floating point numbers. The three new bit generators were chosen for their
+combination of statistical soundness and performance. The PCG family of generators
+have been widely tested using state-of-the-art statistical tests and found to
+perform well (\cite{lemire_blog}). All three of the new generators were tested
+using the PractRand test suite (\cite{practrand}) with four TB of random values.
+Each was tested using a single generator and using a composite generator assembled
+from multiple copies of the same bit generator, each seeded from a shared
+\texttt{SeedSequence}.
 

### 2020-06-02 methods.tex
Fix references

@@
 functions, exposed both in Python and C, for generating random integer
-and floating point numbers. The three new bit generators were chosen for their
-combination of statistical soundness and performance. The PCG family of generators
-have been widely tested using state-of-the-art statistical tests and found to
-perform well (\cite{lemire_blog}). All three of the new generators were tested
-using the PractRand test suite (\cite{practrand}) with four TB of random values.
-Each was tested using a single generator and using a composite generator assembled
-from multiple copies of the same bit generator, each seeded from a shared
-\texttt{SeedSequence}.
+and floating point numbers.
 

### 2020-06-03 methods.tex
Minor cleanup

--- 2020-06-02 methods.tex
@@
 For best performance and reliability the algorithms are provided by leveraging
-an external implemention of the LAPACK interface \cite{LAPACK}.
+an external implementation of the LAPACK interface \cite{LAPACK}.
 LAPACK itself was first released in 1992 and is based on the EISPACK and LINPACK
@@
 have been widely tested using state-of-the-art statistical tests and found to
-perform well (\cite{lemire_blog}). All three of the new generators were tested
-using the PractRand test suite (\cite{practrand}) with four TB of random values.
+perform well \cite{lemire_blog}. All three of the new generators were tested
+using the PractRand test suite \cite{practrand} with four TB of random values.
 Each was tested using a single generator and using a composite generator assembled

### 2020-06-03 methods.tex
Edits to linalg paragraph

@@
 
-NumPy's linear algebra library provides an interface to established
-linear algebra methods. These include solving linear systems of equations
-and computing various matrix properties, such as the determinant,
-the norm, the inverse, and the pseudo-inverse.
-As well as computing the Cholesky, eigenvalue, and singular value decompositions
-of a matrix and further functions.
-For best performance and reliability the algorithms are provided by leveraging
-an external implementation of the LAPACK interface \cite{LAPACK}.
-LAPACK itself was first released in 1992 and is based on the EISPACK and LINPACK
-open-source libraries. It is part of Netlib, a repository of
-mathematical software, papers and databases and has a long history of open,
-community-wide development \cite{dongarra1990lapack,dongarra2008netlib}.
+NumPy's linear algebra library includes functionality for: solving
+linear systems of equations; calculating various matrix properties
+such as the determinant, the norm, the inverse, the pseudo-inverse;
+and computing the Cholesky, eigenvalue, and singular value
+decompositions of a matrix.
+
+The NumPy module mainly provides interfaces to established
+implementations of the LAPACK (Linear Algebra PACKage) \cite{LAPACK}
+interface.  LAPACK itself, which was first released in 1992, is, in
+turn, based on the EISPACK and LINPACK open-source libraries. It is
+part of Netlib, a repository of mathematical software, papers, and
+databases, that has a long history of open, community-wide
+development \cite{dongarra1990lapack,dongarra2008netlib}. For improved
+performance, NumPy links to an accelerated BLAS (Basic Linear Algebra
+Subprograms) implementation, most commonly OpenBLAS
+(http://www.openblas.net/).
 

### 2020-06-03 body.tex
Minor fixes to response

@@
 such as OpenBLAS \cite{wang2013augem,xianyi2012model} or Intel MKL optimized
-for the CPUs at hand.
+for the CPUs at hand (see Supplementary Methods for more details).
 
@@
 still depends heavily on contributions made by graduate students and
-researchers in their free time.
+researchers in their free time (see Supplementary Methods for more details).
 

### 2020-06-05 main.tex
Finalize resubmission

--- 2020-06-02 paper.tex
@@
 
-\usepackage{listings, textcomp}
+\usepackage{textcomp}
 \usepackage{graphicx,xcolor}
 \bibliographystyle{ieeetr}
+\usepackage{caption}
 \RequirePackage[colorlinks=true, allcolors=blue]{hyperref}
-
-\lstset{language=Python}
-
-\lstset{
-   basicstyle=\ttfamily\footnotesize,
-   xleftmargin=\parindent,
-   breaklines=true,
-   captionpos=b,
-   commentstyle=\color{gray},
-   keywordstyle=\color{blue},
-   stringstyle=\color{red},
-   upquote=true
-}
-
-\usepackage{caption}
-\captionsetup{
-   figurename=Fig.,
-   tablename=Tab.
-}
 
@@
 \titleformat*{\section}{\large\bfseries}
+
 \title{Array Programming with NumPy}
@@
     \begin{abstract}
+
 Array programming provides a powerful, compact, expressive syntax for accessing,
 manipulating, and operating on data in vectors, matrices, and
-     higher-dimensional arrays \cite{iverson1980notation}.
-     NumPy is the primary array programming library for the Python language
-     \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
+higher-dimensional arrays. NumPy is the primary array programming library for the Python language.
 It plays an essential role in research analysis pipelines in fields as
@@
 For example, in astronomy, NumPy was an important part of the software stack used
-     in the discovery of gravitational waves \cite{abbott2016observation}
-     and the first imaging of a black hole \cite{eht-imaging}.
-     Here we show how a few fundamental array concepts lead to a simple and
+in the discovery of gravitational waves and the first imaging of a black hole. Here we review how a few fundamental array concepts lead to a simple and
 powerful programming paradigm for organizing, exploring, and analyzing
@@
 between these new array computation libraries.
+NumPy's API and its role as interoperability layer provides a flexible
+framework for the next decade of scientific and industrial analysis.
      \end{abstract}
@@
 ]
-
-\begin{figure*}[h]
-  \centering
-  \includegraphics[width=\textwidth]{concepts}
-  \caption{\textbf{The NumPy array incorporates several fundamental array concepts.}
-    \textbf{a,} The NumPy array data structure and its associated metadata fields.
-    \textbf{b,} Indexing an array with slices and steps. These
-  operations return a \emph{view} of the original data.
-    \textbf{c,} Indexing an array with masks, scalar coordinates, or
-  other arrays, so that it returns a copy of the original data. In the
-  bottom example, an array is indexed with other arrays; this broadcasts the indexing arguments before performing the lookup.
-    \textbf{d,} Vectorization efficiently applies operations to groups
-  of elements.
-    \textbf{e,} Broadcasting in the multiplication of two-dimensional arrays.
-    \textbf{f,} Reduction operations act along one or more axes. In this
-    example, an array is summed along select axes to produce a vector, or along two axes consecutively to
-    produce a scalar.
-    \textbf{g,} Example NumPy code, illustrating some of these concepts.
-   }
-  \label{fig:array-concepts}
-\end{figure*}
 
@@
 powerfully expressive array programming language.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.5\textwidth]{ecosystem}
-  \caption{\textbf{NumPy is the base of the scientific Python ecosystem.}
-   Essential libraries and projects that depend on NumPy's API gain access to new array
-   implementations that support NumPy's array protocols (Fig.~\ref{fig:array-protocol}).
-  }
-  \label{fig:ecosystem}
-\end{figure}
 
@@
 
-\begin{figure*}
-  \centering
-  \includegraphics[width=\textwidth]{protocol}
-
-  \caption{\textbf{NumPy's API and array protocols expose new arrays to the ecosystem.}
-     In this example, NumPy's \texttt{mean} function is called on a Dask
-     array.  The call succeeds by dispatching to the appropriate library implementation
-     (i.e., Dask in this case) and results in a new Dask array.  Compare this
-     code to the example code in Fig.~\ref{fig:array-concepts}g.
-  }\label{fig:array-protocol}
-
-\end{figure*}
-
 NumPy provides in-memory, multidimensional, homogeneously typed
@@
 
-\bibitem{iverson1980notation}
-K.~E. Iverson, ``Notation as a tool of thought,'' {\em Communications of the
-  ACM}, vol.~23, p.~444–465, Aug. 1980.
-
-\bibitem{dubois2007guest}
-P.~F. Dubois, ``{P}ython: Batteries included,'' {\em Computing in Science \&
-  Engineering}, vol.~9, no.~3, pp.~7--9, 2007.
-
-\bibitem{oliphant2007python}
-T.~E. Oliphant, ``{P}ython for scientific computing,'' {\em Computing in
-  Science \& Engineering}, vol.~9, pp.~10--20, May-June 2007.
-
-\bibitem{millman2011python}
-K.~J. Millman and M.~Aivazis, ``{P}ython for scientists and engineers,'' {\em
-  Computing in Science \& Engineering}, vol.~13, no.~2, pp.~9--12, 2011.
-
-\bibitem{perez2011python}
-F.~P\'{e}rez, B.~E. Granger, and J.~D. Hunter, ``{P}ython: an ecosystem for
-  scientific computing,'' {\em Computing in Science \& Engineering}, vol.~13,
-  no.~2, pp.~13--21, 2011.
-
-\bibitem{abbott2016observation}
-B.~P. Abbott, R.~Abbott, T.~Abbott, M.~Abernathy, F.~Acernese, K.~Ackley,
-  C.~Adams, T.~Adams, P.~Addesso, R.~Adhikari, {\em et~al.}, ``Observation of
-  gravitational waves from a binary black hole merger,'' {\em Physical Review
-  Letters}, vol.~116, no.~6, p.~061102, 2016.
-
-\bibitem{eht-imaging}
-A.~A. Chael, M.~D. Johnson, R.~Narayan, S.~S. Doeleman, J.~F. Wardle, and K.~L.
-  Bouman, ``High-resolution linear polarimetric imaging for the event horizon
-  telescope,'' {\em The Astrophysical Journal}, vol.~829, no.~1, p.~11, 2016.
-
 \bibitem{dubois1996numerical}
@@
 \newblock Trelgol Publishing USA, 1st~ed., 2006.
+
+\bibitem{dubois2007guest}
+P.~F. Dubois, ``{P}ython: Batteries included,'' {\em Computing in Science \&
+  Engineering}, vol.~9, no.~3, pp.~7--9, 2007.
+
+\bibitem{oliphant2007python}
+T.~E. Oliphant, ``{P}ython for scientific computing,'' {\em Computing in
+  Science \& Engineering}, vol.~9, pp.~10--20, May-June 2007.
+
+\bibitem{millman2011python}
+K.~J. Millman and M.~Aivazis, ``{P}ython for scientists and engineers,'' {\em
+  Computing in Science \& Engineering}, vol.~13, no.~2, pp.~9--12, 2011.
+
+\bibitem{perez2011python}
+F.~P\'{e}rez, B.~E. Granger, and J.~D. Hunter, ``{P}ython: an ecosystem for
+  scientific computing,'' {\em Computing in Science \& Engineering}, vol.~13,
+  no.~2, pp.~13--21, 2011.
+\newblock \newline \textbf{Explains why the scientific Python ecosystem is a
+  highly productive environment for research.}
 
@@
   vol.~17, pp.~261--272, 2020.
+\newblock \newline \textbf{Introduces the SciPy library and includes a more
+  detailed history of NumPy and SciPy.}
 
@@
   Proceedings of the 9th Python in Science Conference} (S.~van~der Walt and
-  J.~Millman, eds.), pp.~51 -- 56, 2010.
+  J.~Millman, eds.), pp.~51--56, 2010.
 
@@
 F.~Pedregosa, G.~Varoquaux, A.~Gramfort, V.~Michel, B.~Thirion, O.~Grisel,
-  M.~Blondel, P.~Prettenhofer, R.~Weiss, V.~Dubourg, J.~Vanderplas, A.~Passos,
-  D.~Cournapeau, M.~Brucher, M.~Perrot, and {\'E}.~Duchesnay, ``Scikit-learn:
-  Machine learning in {P}ython,'' {\em Journal of Machine Learning Research},
-  vol.~12, no.~Oct, pp.~2825--2830, 2011.
+  M.~Blondel, P.~Prettenhofer, R.~Weiss, V.~Dubourg, R.~Weiss, V.~Dubourg,
+  J.~Vanderplas, A.~Passos, D.~Cournapeau, M.~Brucher, M.~Perrot, and
+  {\'E}.~Duchesnay, ``Scikit-learn: Machine learning in {P}ython,'' {\em
+  Journal of Machine Learning Research}, vol.~12, no.~Oct, pp.~2825--2830,
+  2011.
 
@@
   Engineering}, vol.~13, no.~2, pp.~22--30, 2011.
+\newblock \newline \textbf{Discusses the NumPy array data structure with a
+  focus on how it enables efficient computation.}
 
@@
 P.~J. Cock, T.~Antao, J.~T. Chang, B.~A. Chapman, C.~J. Cox, A.~Dalke,
-  I.~Friedberg, T.~Hamelryck, F.~Kauff, B.~Wilczynski, and M.~J.~L. de~Hoon,
-  ``Biopython: freely available {Python} tools for computational molecular
-  biology and bioinformatics,'' {\em Bioinformatics}, vol.~25, no.~11,
-  pp.~1422--1423, 2009.
+  I.~Friedberg, T.~Hamelryck, F.~Kauff, B.~Wilczynski, F.~Kauff, B.~Wilczynski,
+  and M.~J.~L. de~Hoon, ``Biopython: freely available {Python} tools for
+  computational molecular biology and bioinformatics,'' {\em Bioinformatics},
+  vol.~25, no.~11, pp.~1422--1423, 2009.
 
 \bibitem{millman2007analysis}
-K.~J. Millman and M.~Brett, ``Analysis of functional {M}agnetic {R}esonance
-  {I}maging in {P}ython,'' {\em Computing in Science \& Engineering}, vol.~9,
+K.~J. Millman and M.~Brett, ``Analysis of functional {Magnetic Resonance
+  Imaging} in {P}ython,'' {\em Computing in Science \& Engineering}, vol.~9,
   no.~3, pp.~52--55, 2007.
@@
   pp.~149--183, 2014.
+\newblock \newline \textbf{Describes the software engineering practices
+  embraced by the NumPy and SciPy communities with a focus on how these
+  practices improve research.}
 
@@
 
-\bibitem{peterson2009f2py}
-P.~Peterson, ``{F2PY}: a tool for connecting {F}ortran and {P}ython programs,''
-  {\em International Journal of Computational Science and Engineering}, vol.~4,
-  no.~4, pp.~296--305, 2009.
-
-\bibitem{NumPyProjectGovernance}
-{The NumPy Project Community}, ``{NumPy} project governance,'' 2015.
-
-\bibitem{NumPyCodeofConduct}
-{The NumPy Project Community}, ``{NumPy} code of conduct,'' 2018.
-
-\bibitem{PEP427}
-D.~Holth, ``Pep 427 -- the wheel binary package format 1.0,'' 2012.
-
-\bibitem{multibuild}
-{Brett, M. et al}, ``{multibuild},'' 2016.
-
-\bibitem{NEP13}
-B.~Griffith, P.~Virtanen, N.~Smith, M.~{van Kerkwijk}, and S.~Hoyer, ``{NEP} 13
-  -- a mechanism for overriding ufuncs,'' 2013.
-
-\bibitem{NEP18}
-S.~Hoyer, M.~Rocklin, M.~{van Kerkwijk}, H.~Abbasi, and E.~Wieser, ``{NEP} 18
-  -- a dispatch mechanism for numpy's high level array functions,'' 2018.
-
-\bibitem{pcg64}
-M.~E. O'Neill, ``Pcg: A family of simple fast space-efficient statistically
-  good algorithms for random number generation,'' Tech. Rep. HMC-CS-2014-0905,
-  Harvey Mudd College, Claremont, CA, Sept. 2014.
-
-\bibitem{random123}
-J.~K. Salmon, M.~A. Moraes, R.~O. Dror, and D.~E. Shaw, ``Parallel random
-  numbers: As easy as 1, 2, 3,'' in {\em Proceedings of 2011 International
-  Conference for High Performance Computing, Networking, Storage and Analysis},
-  SC '11, (New York, NY, USA), pp.~16:1--16:12, ACM, 2011.
-
-\bibitem{practrand}
-C.~Doty-Humphrey, ``Practrand, version 0.94.''
-
-\bibitem{mt19937}
-M.~Matsumoto and T.~Nishimura, ``{Mersenne Twister}: A 623-dimensionally
-  equidistributed uniform pseudo-random number generator,'' {\em ACM
-  Transactions on Modeling and Computer Simulation}, vol.~8, pp.~3--30, Jan.
-  1998.
-
-\bibitem{randomgen}
-K.~Sheppard, B.~Duvenhage, P.~de~Buyl, and D.~A. Ham, ``bashtage/randomgen:
-  Release 1.16.2,'' Apr. 2019.
-
-\bibitem{ziggurat}
-G.~Marsaglia and W.~W. Tsang, ``The ziggurat method for generating random
-  variables,'' {\em Journal of Statistical Software, Articles}, vol.~5, no.~8,
-  pp.~1--7, 2000.
-
-\bibitem{lemire}
-D.~Lemire, ``Fast random integer generation in an interval,'' {\em ACM
-  Transactions on Modeling and Computer Simulation}, vol.~29, pp.~1--12, Jan
-  2019.
-
-\bibitem{top500nov2019}
-{top500}, ``Top 10 sites for november 2019,'' 2019.
-
-\bibitem{astra-wiki}
-{wikichip}, ``Astra - supercomputers,'' 2019.
-
-\bibitem{arm-architecture}
-{Wikipedia}, ``Arm architecture,'' 2019.
-
-\bibitem{numpy-roadmap}
-{NumPy Developers}, ``Numpy roadmap,'' 2019.
-
-\bibitem{PEP599}
-{Dustin Ingram}, ``Pep 599 -- the manylinux2014 platform tag,'' 2019.
-
 \end{thebibliography}
-
-\newpage
-
-\section*{Methods}
-
-We use Git for version control and GitHub as the public hosting service for our
-official \emph{upstream} repository (\url{https://github.com/numpy/numpy}).
-We each work in our own copy (or fork) of the project and use the
-upstream repository as our integration point.
-To get new code into the upstream repository, we use GitHub's
-pull request (PR) mechanism.
-This allows us to review code before integrating it as well as to run a
-large number of tests on the modified code to ensure that the changes
-do not break expected behavior.
-
-We also use GitHub's issue tracking system to collect and triage problems and
-proposed improvements.
-
-\subsection*{Library organization}
-
-Broadly, the NumPy library consists of the following parts:
-the NumPy array data structure \texttt{ndarray}; the so-called \emph{universal functions};
-a set of library functions for manipulating arrays and doing scientific
-computation; infrastructure libraries for unit tests and Python package
-building; and the program \texttt{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
-The \texttt{ndarray} and the universal functions are generally considered
-the core of the library.
-In the following, we give a brief summary of these components of the
-library.
-
-\paragraph{\emph{Core.}}  The \texttt{ndarray} data structure and the
-universal functions make up the core of NumPy.
-
-The \texttt{ndarray} is the data structure at the heart of NumPy.
-The data structure stores regularly strided homogeneous data types
-inside a contiguous block memory, allowing for the efficient representation
-of $n$-dimensional data.
-More details about the data structure are given in ``The NumPy array:
-a structure for efficient numerical computation'' \cite{vanderwalt2011numpy}.
-
-The \emph{universal functions}, or more concisely, \emph{ufuncs},
-are functions written in C that implement efficient looping over
-NumPy arrays. An important feature of ufuncs is the built-in
-implementation of \emph{broadcasting}.  For example, the function
-\texttt{arctan2(x, y)} is a ufunc that accepts two values and computes
-$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
-the ufunc will take care of looping over the dimensions of the inputs
-in such a way that if, say, \texttt{x} is a 1-D array with length 3, and
-\texttt{y} is a 2-D array with shape $2 \times 1$, the output will be
-an array with shape $2 \times 3$ (Fig.~\ref{fig:array-concepts}c).
-The ufunc machinery takes care
-of calling the function with all the appropriate combinations of
-input array elements to complete the output array.
-The elementary arithmetic operations of addition, multiplication, etc.,
-are implemented as ufuncs, so that broadcasting also applies to expressions
-such as \texttt{x + y * z}.
-
-\paragraph{\emph{Computing libraries.}}
-NumPy provides a large library of functions for array manipulation
-and scientific computing, including functions for: creating, reshaping,
-concatenating, and padding arrays; searching, sorting and counting data
-in arrays; computing elementary statistics, such as the mean, median,
-variance, and standard deviation; file I/O; and more.
-
-A suite of functions for computing the \emph{fast Fourier transform (FFT)}
-and its inverse is provided.
-
-NumPy's linear algebra library includes functions for: solving linear
-systems of equations; computing various functions of a matrix, including
-the determinant, the norm, the inverse, and the pseudo-inverse;
-computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
-and more.
-
-The random number generator library in NumPy provides alternative
-\emph{bit stream generators} that provide the core function of generating
-random integers.
-A higher-level generator class that implements an assortment of
-probability distributions is provided. It includes the beta, gamma
-and Weibull distributions, the univariate and multivariate normal
-distributions, and more.
-
-\paragraph{\emph{Infrastructure libraries.}} NumPy provides utilities
-for writing tests and for building Python packages.
-
-The \texttt{testing} subpackage provides functions such as
-\texttt{assert\_allclose(actual, desired)} that may be used in
-test suites for code that uses NumPy arrays.
-
-NumPy provides the subpackage \texttt{distutils} which includes functions and classes
-to facilitate configuration, installation, and packaging of libraries depending on NumPy.
-These can be used, for example, when publishing to the PyPI website.
-
-\paragraph{\emph{F2PY.}}  The program \texttt{f2py} is a tool for
-building NumPy-aware Python wrappers of Fortran functions.
-NumPy itself does not use any Fortran code;  F2PY is part of NumPy
-for historical reasons.
-
-\subsection*{Governance}
-
-NumPy adopted an official Governance Document on October~5,
-2015 \cite{NumPyProjectGovernance}.
-Project decisions are usually made by consensus of interested contributors.
-This means that, for most decisions, everyone is entrusted with veto power.
-A Steering Council, currently composed of 12~members, facilitates this
-process and oversees daily development of the project by contributing code
-and reviewing contributions from the community.
-
-NumPy's official Code of Conduct was approved on September~1, 2018 \cite{NumPyCodeofConduct}.
-In brief, we strive to:
-\emph{be open};
-\emph{be empathetic, welcoming, friendly, and patient};
-\emph{be collaborative};
-\emph{be inquisitive}; and
-\emph{be careful in the words that we choose}.
-The Code of Conduct also specifies how breaches can be reported and outlines
-the process for responding to such reports.
-
-\subsection*{Funding}
-
-In 2017, NumPy received its first large grants totaling 1.3M USD from the
-Gordon \& Betty Moore and the Alfred P. Sloan foundations.
-Stéfan van der Walt is the PI and manages four programmers working on the project.
-These two grants focus on addressing the technical debt accrued over the years and
-on setting in place standards and architecture to encourage more sustainable development.
-
-NumPy received a third grant for 195K USD from the Chan Zuckerberg
-Initiative at the end of 2019 with Ralf Gommers as the PI.
-This grant focuses on better serving NumPy's large number of beginning
-to intermediate level users and on growing the community of NumPy
-contributors.
-It will also provide support to OpenBLAS, on which NumPy depends for
-accelerated linear algebra.
-
-Finally, since May 2019 the project receives a small amount annually from
-Tidelift, which is used to fund things like documentation and website
-improvements.
-
-\subsection*{Developers}
-
-NumPy is currently maintained by a group of 23 contributors with commit rights
-to the NumPy code base. Out of these, 17 maintainers were active in
-2019, 4 of whom were paid to work on the project full-time.
-Additionally, there are a few long term developers who contributed and maintain
-specific parts of NumPy, but are not officially maintainers.
-
-Over the course of its history, NumPy has attracted PRs by 823 contributors.
-However, its development relies heavily on a small number
-of active maintainers, who share more than half of the contributions among
-themselves.
-
-At a release cycle of about every half year, the five recent releases in the years
-2018 and 2019 have averaged about 450~PRs each,\footnote{
-    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
-    All data provided here is based on the development which happened through GitHub
-    PRs. In some cases contributions by maintainers may not be categorized as such.}
-with each release attracting more than a hundred new contributors.
-Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
-master branch.
-Although the number of PRs being merged fluctuates,
-the plot indicates an increased number of contributions over the past
-years.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\linewidth]{PRs-using-CURRENT_MAINTAINERS.pdf}
-    \caption{\textbf{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012.} The total number of PRs is indicated with the
-        lower blue area showing the portion contributed by current or previous
-        maintainers.}\label{fig:prs-over-time}
-\end{figure}
-
-\subsection*{Community calls}
-
-The massive number of scientific Python packages that
-built on NumPy meant that it had an unusually high need for stability.
-So to guide our development we formalized the feature proposal process, and
-constructed a development roadmap with extensive input and feedback from the
-community.
-
-Weekly community calls alternate between triage and
-higher level discussion.  The calls not only involve developers from
-the community, but provide a venue for vendors and other external
-groups to provide input.  For example, after Intel produced a forked
-version of NumPy, one of their developers joined a call to discuss
-community concerns.
-
-\subsection*{NumPy enhancement proposals}
-
-Given the complexity of the codebase and the massive number of projects depending
-on it, large changes require careful planning and substantial work.
-NumPy Enhancement Proposals (NEPs) are modeled after
-Python Enhancement Proposals (PEPs) for ``proposing major new
-features, for collecting community input on an issue, and for
-documenting the design decisions that have gone into
-Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
-Since then there have been 19 proposed NEPS---6 have been implemented,
-4 have been accepted and are being implemented, 4 are under
-consideration, 3 have been deferred or superseded, and 2 have been rejected
-or withdrawn.
-
-\subsection*{Central role}
-
-NumPy plays a central role in building and standardizing much of the scientific
-Python community infrastructure.
-NumPy's docstring standard is now widely adopted.
-We are also now using the NEP system as a way to help coordinate the larger
-scientific Python community.
-For example, in NEP 29, we recommend, along with leaders from various other
-projects, that all projects across the Scientific Python ecosystem adopt a
-common ``time window-based'' policy for support of Python and NumPy versions.
-This standard will simplify downstream project and release planning.
-
-\subsection*{Wheels build system}
-
-A Python \emph{wheel} \cite{PEP427} is a standard file format for
-distributing Python libraries.  In addition to Python code, a
-wheel may include compiled C extensions and other binary data.
-This is important, because many libraries, including NumPy,
-require a C compiler and other build tools to build the software
-from the source code, making it difficult for many users to install
-the software on their own.  The introduction of wheels to the Python
-packaging system has made it much easier for users to install
-precompiled libraries.
-
-A GitHub repository containing scripts to build NumPy wheels has
-been configured so that a simple commit to the repository triggers
-an automated build system that creates NumPy wheels for several
-computer platforms, including Windows, Mac OSX and Linux.  The wheels
-are uploaded to a public server and made available for anyone to use.
-This system makes it easy for users to install precompiled versions
-of NumPy on these platforms.
-
-The technology that is used to build the wheels evolves continually.
-At the time this paper is being written, a key component is the
-\texttt{multibuild} suite of tools developed by Matthew Brett and
-other developers \cite{multibuild}.  Currently, scripts using
-\texttt{multibuild} are written for the continuous integration
-platforms Travis-CI (for Linux and Mac OSX) and Appveyor
-(for Windows).
-
-\subsection*{Recent technical improvements}
-
-With the recent infusion of funding and a clear process for coordinating with
-the developer community, we have been able to tackle a number of important
-large scale changes.
-We highlight two of those below, as well as changes made to our testing
-infrastructure to support hardware platforms used in large scale computing.
-
-\subsection*{Array function protocol}
-
-A vast number of projects are built on NumPy;
-these projects are consumers of the NumPy API.
-Over the last several years, a growing number of projects are providers of
-a \emph{NumPy-like API} and array objects targeting audiences with specialized
-needs beyond NumPy's capabilities.
-For example, the NumPy API is implemented by several popular tensor computation
-libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
-JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
-and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
-PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
-and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
-provide tensor APIs with NumPy-inspired semantics.
-It is also implemented in packages that support sparse arrays
-such as \texttt{scipy.sparse} and PyData/Sparse.
-Another notable example is Dask, a library for parallel computing in
-Python.  Dask adopts the NumPy API and therefore presents a familiar
-interface to existing NumPy users, while adding powerful abilities to
-parallelize and distribute tasks.
-
-The multitude of specialized projects creates the difficulty that consumers
-of these NumPy-like APIs write code specific to a single project and do not support
-all of the above array providers.
-This is a burden for users relying on the specialized array-like, since
-a tool they need may not work for them.
-It also creates challenges for end-users who need to transition
-from NumPy to a more specialized array.
-The growing multitude of specialized projects with NumPy-like APIs threatened
-to again fracture the scientific Python community.
-
-To address these issues NumPy has the goal of providing the fundamental
-API for \emph{interoperability} between the various NumPy-like APIs.
-An earlier step in this direction was the implementation of the
-\texttt{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
-for most mathematical functions \cite{NEP13}.
-In 2019 this was expanded more generally with the inclusion of the
-\texttt{\_\_array\_function\_\_} protocol into NumPy~1.17.
-These two protocols allow providers of array objects to be interoperable
-with the NumPy API: their arrays work correctly with almost all NumPy functions \cite{NEP18}.
-For the users relying on specialized array projects it means that even though
-much code is written specifically for NumPy arrays and uses the NumPy API as
-\texttt{import numpy as np}, it can nevertheless work for them.
-For example, here is how a CuPy GPU array can be passed through NumPy for
-processing, with all operations being dispatched back to CuPy:
-
-\begin{lstlisting}
-import numpy as np
-import cupy as cp
-
-x_gpu = cp.array([1, 2, 3])
-y = np.sum(x_gpu)  # Returns a GPU array
-\end{lstlisting}
-
-Similarly, user defined functions composed using NumPy can now be
-applied to, e.g., multi-node distributed Dask arrays:
-
-\begin{lstlisting}
-import numpy as np
-import dask.array as da
-
-def f(x):
-    """Function using NumPy API calls"""
-    y = np.tensordot(x, x.T)
-    return np.mean(np.log(y + 1))
-
-x_local = np.random.random([10000, 10000])  # random local array
-x_distr = da.random.random([10000, 10000])  # random distributed array
-
-f(x_local)  # returns a NumPy array
-f(x_distr)  # works, returns a Dask array
-\end{lstlisting}
-
-\subsection*{Random number generation}
-
-The NumPy \texttt{random} module provides pseudorandom numbers from a wide range of
-distributions. In legacy versions of NumPy, simulated random values are produced
-by a \texttt{RandomState} object that: handles seeding and state initialization;
-wraps the core pseudorandom number generator based on a Mersenne Twister
-implementation\footnote{to be precise, the standard 32-bit version of MT19937};
-interfaces with the underlying code that transforms random bits into
-variates from other distributions; and supplies a singleton instance exposed in
-the root of the random module.
-
-The \texttt{RandomState} object makes a compatibility guarantee so that a fixed
-seed and sequence of function calls produce the same set of values. This
-guarantee has slowed progress since improving the underlying code requires
-extending the API with additional keyword arguments. This guarantee continues to
-apply to \texttt{RandomState}.
-
-NumPy 1.17 introduced a new API for generating random numbers that use a more
-flexible structure that can be extended by libraries or end-users. The new API
-is built using components that separate the steps required to generate random
-variates. Pseudorandom bits are generated by a bit generator. These bits are
-then transformed into variates from complex distributions by a generator.
-Finally, seeding is handled by an object that produces sequences of high-quality
-initial values.
-
-Bit generators are simple classes that manage the state of an underlying
-pseudorandom number generator. NumPy ships with four bit generators. The default
-bit generator is a 64-bit implementation of the Permuted Congruential Generator
-\cite{pcg64} (\texttt{PCG64}). The three other bit generators are a 64-bit version
-of the Philox generator \cite{random123} (\texttt{Philox}), Chris Doty-Humphrey's
-Small Fast Chaotic generator \cite{practrand} (\texttt{SFC64}), and the 32-bit
-Mersenne Twister \cite{mt19937} (\texttt{MT19937}) which has been used in older
-versions of NumPy.\footnote{The
-\href{https://github.com/bashtage/randomgen}{randomgen project} supplies a wide
-range of alternative bit generators such as a cryptographic counter-based
-generators (\texttt{AESCtr}) and generators that expose hardware random number
-generators (\texttt{RDRAND}) \cite{randomgen}.} Bit generators provide
-functions, exposed both in Python and C, for generating random integer
-and floating point numbers.
-
-The \texttt{Generator} consumes one of the bit generators and produces variates
-from complicated distributions. Many improved methods for generating random
-variates from common distributions were implemented, including the Ziggurat
-method for normal, exponential and gamma variates \cite{ziggurat}, and Lemire's
-method for bounded random integer generation \cite{lemire}. The \texttt{Generator}
-is more similar to the legacy \texttt{RandomState}, and its API is substantially
-the same. The key differences all relate to state management, which has been
-delegated to the bit generator. The \texttt{Generator} does not make the same
-stream guarantee as the \texttt{RandomState} object, and so variates may differ
-across versions as improved generation algorithms are
-introduced.\footnote{Despite the removal of the compatibility guarantee, simple
-reproducibility across versions is encouraged, and minor changes that do not
-produce meaningful performance gains or fix underlying bug are not generally
-adopted.}
-
-Finally, a \texttt{SeedSequence} is used to initialize a bit generator. The seed
-sequence can be initialized with no arguments, in which case it reads entropy
-from a system-dependent provider, or with a user-provided seed. The seed
-sequence then transforms the initial set of entropy into a sequence of
-high-quality pseudorandom integers, which can be used to initialize multiple bit
-generators deterministically. The key feature of a seed sequence is that
-it can be used to spawn child \texttt{SeedSequence}s to initialize
-multiple distinct bit generators.
-This capability allows a seed sequence to facilitate large distributed applications
-where the number of workers required is not known. The sequences generated from
-the same initial entropy and spawns are fully deterministic to ensure
-reproducibility.
-
-The three components are combined to construct a complete random number
-generator.
-
-\begin{lstlisting}
-from numpy.random import (
-    Generator,
-    PCG64,
-    SeedSequence,
-)
-
-seq = SeedSequence(1030424547444117993331016959)
-pcg = PCG64(seq)
-gen = Generator(pcg)
-\end{lstlisting}
-
-This approach retains access to the seed sequence which can then be
-used to spawn additional generators.
-
-\begin{lstlisting}
-children = seq.spawn(2)
-gen_0 = Generator(PCG64(children[0]))
-gen_1 = Generator(PCG64(children[1]))
-\end{lstlisting}
-
-While this approach retains complete flexibility, the method
-\texttt{np.random.default\_rng} can be used to instantiate a \texttt{Generator} when
-reproducibility is not needed.
-
-The final goal of the new API is to improve extensibility. \texttt{RandomState} is
-a monolithic object that obscures all of the underlying state and functions. The
-component architecture is one part of the extensibility improvements. The
-underlying functions (written in C) which transform the output of a bit
-generator to other distributions are available for use in CFFI. This allows the
-same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
-Numba. Both the bit generators and the low-level functions can also be used in C
-or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
-NumPy source. Alternative approaches that avoid this extra step are being
-explored.}
-
-\subsection*{Testing on multiple architectures}
-
-At the time of writing the two fastest supercomputers in the
-world, Summit and Sierra, both have IBM POWER9 architectures
-\cite{top500nov2019}. In late 2018, Astra, the first ARM-based
-supercomputer to enter the TOP500 list, went into production
-\cite{astra-wiki}. Furthermore, over 100 billion ARM processors have been
-produced as of 2017 \cite{arm-architecture}, making it the most 
-widely used instruction set architecture in the world.
-
-Clearly there are motivations for a large scientific computing
-software library to support POWER and ARM architectures. We've extended
-our continuous integration (CI) testing to include \texttt{ppc64le}
-(POWER8 on Travis CI) and ARMv8 (on Shippable service). We also test
-with the s390x architecture (IBM Z CPUs on Travis CI) so that we
-can probe the behavior of our library on a big-endian machine.
-This satisfies one of the major components of
-improved CI testing laid out in a version of our roadmap
-\cite{numpy-roadmap}---specifically, ``CI for more exotic
-platforms."
-
-PEP 599 \cite{PEP599} lays out a plan for new Python binary wheel
-distribution support, \texttt{manylinux2014}, that adds
-support for a number of architectures supported by the CentOS
-Alternative Architecture Special Interest Group, including
-ARMv8, ppc64le, as well as s390x. We are thus well-positioned
-for a future where provision of binaries on these architectures
-will be expected for a library at the base of the ecosystem.
 
@@
 
+\section*{Figures}
+
+\begin{figure*}[h]
+  \centering
+
+  \caption{\textbf{The NumPy array incorporates several fundamental array concepts.}
+    \textbf{a,} The NumPy array data structure and its associated metadata fields.
+    \textbf{b,} Indexing an array with slices and steps. These
+  operations return a \emph{view} of the original data.
+    \textbf{c,} Indexing an array with masks, scalar coordinates, or
+  other arrays, so that it returns a copy of the original data. In the
+  bottom example, an array is indexed with other arrays; this broadcasts the indexing arguments before performing the lookup.
+    \textbf{d,} Vectorization efficiently applies operations to groups
+  of elements.
+    \textbf{e,} Broadcasting in the multiplication of two-dimensional arrays.
+    \textbf{f,} Reduction operations act along one or more axes. In this
+    example, an array is summed along select axes to produce a vector, or along two axes consecutively to
+    produce a scalar.
+    \textbf{g,} Example NumPy code, illustrating some of these concepts.
+   }
+  \label{fig:array-concepts}
+\end{figure*}
+
+\begin{figure}
+  \centering
+
+  \caption{\textbf{NumPy is the base of the scientific Python ecosystem.}
+   Essential libraries and projects that depend on NumPy's API gain access to new array
+   implementations that support NumPy's array protocols (Fig.~\ref{fig:array-protocol}).
+  }
+  \label{fig:ecosystem}
+\end{figure}
+
+\begin{figure*}
+  \centering
+
+  \caption{\textbf{NumPy's API and array protocols expose new arrays to the ecosystem.}
+     In this example, NumPy's \texttt{mean} function is called on a Dask
+     array.  The call succeeds by dispatching to the appropriate library implementation
+     (i.e., Dask in this case) and results in a new Dask array.  Compare this
+     code to the example code in Fig.~\ref{fig:array-concepts}g.
+  }\label{fig:array-protocol}
+
+\end{figure*}
+
 \end{document}

### 2020-06-05 supplementary.tex
Simplify tex for submission
Remove URLs from references
Finalize resubmission

@@
-\documentclass[fleqn,10pt,twocolumn]{wlscirep}
+\documentclass{article}
 
-% Packages
-\usepackage{pifont}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-\usepackage{float}
-% \usepackage[multiple]{footmisc}
-\usepackage{booktabs}
-\usepackage{csquotes}
-\usepackage{moreverb} % for verbatim ouput (for wordcount)
+\usepackage{graphicx,xcolor}
+\bibliographystyle{ieeetr}
+\usepackage{caption}
+\RequirePackage[colorlinks=true, allcolors=blue]{hyperref}
 \usepackage{listings, textcomp}
@@
 
-\lstset{ %
-  basicstyle=\ttfamily\footnotesize,  % size of fonts used for the code
+\lstset{
+   basicstyle=\ttfamily\footnotesize,
    xleftmargin=\parindent,
-  breaklines=true,   % automatic line breaking only at whitespace
-  captionpos=b,   % sets the caption-position to bottom
-  commentstyle=\color{gray},  % comment style
-  keywordstyle=\color{blue},  % keyword style
-  stringstyle=\color{red},  % string literal style
-  upquote=true  %straight single quotes (requires textcomp)
+   breaklines=true,
+   captionpos=b,
+   commentstyle=\color{gray},
+   keywordstyle=\color{blue},
+   stringstyle=\color{red},
+   upquote=true
 }
-
-\usepackage{caption}
-\captionsetup{%
-   figurename=Fig.,
-   tablename=Tab.
-}
-\usepackage{subcaption}
-
-% New Commands
-\newcommand{\cmark}{\ding{51}}%
-\newcommand{\xmark}{\ding{55}}%
-\newcommand{\code}[1]{\texttt{#1}}
-\newcommand{\fixme}[1]{\textcolor{red}{{#1}}}
-\newcommand{\inlinecite}[1]{\footnotesize\citen{#1}}
-\newcommand{\tightlist}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\usepackage[switch]{lineno}
-\linenumbers
 
@@
 
-\input{authors}
+\RequirePackage{authblk}
+\setlength{\affilsep}{1.5em}
+\renewcommand\Authfont{\fontsize{12}{12}\usefont{OT1}{phv}{b}{n}}
+\renewcommand\Affilfont{\fontsize{8}{10}\usefont{OT1}{phv}{m}{n}}
 
-\begin{abstract}
-\input{summary}
-\end{abstract}
+\author[1]{Charles R. Harris}
+\author[2,3,4,*]{K. Jarrod Millman}
+\author[5,2,4,*]{St\'efan J. van der Walt}
+\author[6,*]{Ralf Gommers}
+\author[7]{Pauli Virtanen}
+\author[8]{David Cournapeau}
+\author[9]{Eric Wieser}
+\author[10]{Julian Taylor}
+\author[4]{Sebastian Berg}
+\author[11]{Nathaniel J. Smith}
+\author[12]{Robert Kern}
+\author[4]{Matti Picus}
+\author[13]{Stephan Hoyer}
+\author[14]{Marten H. van Kerkwijk}
+\author[2,15]{Matthew Brett}
+\author[16]{Allan Haldane}
+\author[17]{Jaime Fern\'andez del R\'io}
+\author[18,19]{Mark Wiebe}
+\author[6,20,21]{Pearu Peterson}
+\author[22,23]{Pierre G\'erard-Marchant}
+\author[24]{Kevin Sheppard}
+\author[25]{Tyler Reddy}
+\author[4]{Warren Weckesser}
+\author[6]{Hameer Abbasi}
+\author[26]{Christoph Gohlke}
+\author[6]{Travis E. Oliphant}
+\affil[1]{Independent Researcher, Logan, Utah, USA}
+\affil[2]{Brain Imaging Center, University of California, Berkeley, Berkeley, CA, USA}
+\affil[3]{Division of Biostatistics, University of California, Berkeley, Berkeley, CA, USA}
+\affil[4]{Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA, USA}
+\affil[5]{Applied Mathematics, Stellenbosch University, Stellenbosch, South Africa}
+\affil[6]{Quansight LLC, Austin, TX, USA}
+\affil[7]{Department of Physics and Nanoscience Center, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
+\affil[8]{Mercari JP, Tokyo, Japan}
+\affil[9]{Department of Engineering, University of Cambridge, Cambridge, UK}
+\affil[10]{Independent Researcher, Karlsruhe, Germany}
+\affil[11]{Independent Researcher, Berkeley, CA, USA}
+\affil[12]{Enthought, Inc., Austin, TX, USA}
+\affil[13]{Google Research, Mountain View, CA, USA}
+\affil[14]{Department of Astronomy \& Astrophysics, University of Toronto, Toronto, ON, Canada}
+\affil[15]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
+\affil[16]{Department of Physics, Temple University, Philadelphia, PA, USA}
+\affil[17]{Google, Zurich, Switzerland}
+\affil[18]{Department of Physics and Astronomy, The University of British Columbia, Vancouver, BC, Canada}
+\affil[19]{Amazon, Seattle, Washington, USA}
+\affil[20]{Independent Researcher, Saue, Estonia}
+\affil[21]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
+\affil[22]{Department of Biological and Agricultural Engineering, University of Georgia, Athens, GA}
+\affil[23]{France-IX Services, Paris, France}
+\affil[24]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[25]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[26]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
+\affil[*]{millman@berkeley.edu, stefanv@berkeley.edu, ralf.gommers@gmail.com}
 
@@
 
-% Count of words
-%\verbatiminput{wordcount.tex}
+We use Git for version control and GitHub as the public hosting service for our
+official \emph{upstream} repository (\url{https://github.com/numpy/numpy}).
+We each work in our own copy (or fork) of the project and use the
+upstream repository as our integration point.
+To get new code into the upstream repository, we use GitHub's
+pull request (PR) mechanism.
+This allows us to review code before integrating it as well as to run a
+large number of tests on the modified code to ensure that the changes
+do not break expected behavior.
 
-\input{methods}
+We also use GitHub's issue tracking system to collect and triage problems and
+proposed improvements.
 
-\bibliography{references}
+\section*{Library organization}
+
+Broadly, the NumPy library consists of the following parts:
+the NumPy array data structure \texttt{ndarray}; the so-called \emph{universal functions};
+a set of library functions for manipulating arrays and doing scientific
+computation; infrastructure libraries for unit tests and Python package
+building; and the program \texttt{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
+The \texttt{ndarray} and the universal functions are generally considered
+the core of the library.
+In the following, we give a brief summary of these components of the
+library.
+
+\subsection*{Core}
+
+The \texttt{ndarray} data structure and the
+universal functions make up the core of NumPy.
+
+The \texttt{ndarray} is the data structure at the heart of NumPy.
+The data structure stores regularly strided homogeneous data types
+inside a contiguous block memory, allowing for the efficient representation
+of $n$-dimensional data.
+More details about the data structure are given in ``The NumPy array:
+a structure for efficient numerical computation'' \cite{vanderwalt2011numpy}.
+
+The \emph{universal functions}, or more concisely, \emph{ufuncs},
+are functions written in C that implement efficient looping over
+NumPy arrays. An important feature of ufuncs is the built-in
+implementation of \emph{broadcasting}.  For example, the function
+\texttt{arctan2(x, y)} is a ufunc that accepts two values and computes
+$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
+the ufunc will take care of looping over the dimensions of the inputs
+in such a way that if, say, \texttt{x} is a 1-D array with length 3, and
+\texttt{y} is a 2-D array with shape $2 \times 1$, the output will be
+an array with shape $2 \times 3$ (Fig.~\ref{fig:array-concepts}c).
+The ufunc machinery takes care
+of calling the function with all the appropriate combinations of
+input array elements to complete the output array.
+The elementary arithmetic operations of addition, multiplication, etc.,
+are implemented as ufuncs, so that broadcasting also applies to expressions
+such as \texttt{x + y * z}.
+
+\subsection*{Computing libraries.}
+
+NumPy provides a large library of functions for array manipulation
+and scientific computing, including functions for: creating, reshaping,
+concatenating, and padding arrays; searching, sorting and counting data
+in arrays; computing elementary statistics, such as the mean, median,
+variance, and standard deviation; file I/O; and more.
+
+NumPy's linear algebra library includes functionality for: solving
+linear systems of equations; calculating various matrix properties
+such as the determinant, the norm, the inverse, the pseudo-inverse;
+and computing the Cholesky, eigenvalue, and singular value
+decompositions of a matrix.
+
+The NumPy module mainly provides interfaces to established
+implementations of the LAPACK (Linear Algebra PACKage) \cite{LAPACK}
+interface.  LAPACK itself, which was first released in 1992, is, in
+turn, based on the EISPACK and LINPACK open-source libraries. It is
+part of Netlib, a repository of mathematical software, papers, and
+databases, that has a long history of open, community-wide
+development \cite{dongarra1990lapack,dongarra2008netlib}. For improved
+performance, NumPy links to an accelerated BLAS (Basic Linear Algebra
+Subprograms) implementation, most commonly OpenBLAS
+(http://www.openblas.net/).
+
+The random number generator library in NumPy provides alternative
+\emph{bit stream generators} that provide the core function of generating
+random integers.
+A higher-level generator class that implements an assortment of
+probability distributions is provided. It includes the beta, gamma
+and Weibull distributions, the univariate and multivariate normal
+distributions, and more.
+
+A suite of functions for computing the \emph{fast Fourier transform (FFT)}
+and its inverse is provided.
+
+\subsection*{Infrastructure libraries}
+
+NumPy provides utilities for writing tests and for building Python packages.
+
+The \texttt{testing} subpackage provides functions such as
+\texttt{assert\_allclose(actual, desired)} that may be used in
+test suites for code that uses NumPy arrays.
+
+NumPy provides the subpackage \texttt{distutils} which includes functions and classes
+to facilitate configuration, installation, and packaging of libraries depending on NumPy.
+These can be used, for example, when publishing to the PyPI website.
+
+\subsection*{F2PY}  The program \texttt{f2py} is a tool for
+building NumPy-aware Python wrappers of Fortran functions.
+NumPy itself does not use any Fortran code;  F2PY is part of NumPy
+for historical reasons.
+
+\section*{Governance}
+
+NumPy adopted an official Governance Document on October~5,
+2015 (\url{https://numpy.org/devdocs/dev/governance/governance.html}).
+Project decisions are usually made by consensus of interested contributors.
+This means that, for most decisions, everyone is entrusted with veto power.
+A Steering Council, currently composed of 12~members, facilitates this
+process and oversees daily development of the project by contributing code
+and reviewing contributions from the community.
+
+NumPy's official Code of Conduct was approved on September~1, 2018
+(\url{https://numpy.org/devdocs/dev/conduct/code_of_conduct.html}).
+In brief, we strive to:
+\emph{be open};
+\emph{be empathetic, welcoming, friendly, and patient};
+\emph{be collaborative};
+\emph{be inquisitive}; and
+\emph{be careful in the words that we choose}.
+The Code of Conduct also specifies how breaches can be reported and outlines
+the process for responding to such reports.
+
+\section*{Funding}
+
+In 2017, NumPy received its first large grants totaling 1.3M USD from the
+Gordon \& Betty Moore and the Alfred P. Sloan foundations.
+Stéfan van der Walt is the PI and manages four programmers working on the project.
+These two grants focus on addressing the technical debt accrued over the years and
+on setting in place standards and architecture to encourage more sustainable development.
+
+NumPy received a third grant for 195K USD from the Chan Zuckerberg
+Initiative at the end of 2019 with Ralf Gommers as the PI.
+This grant focuses on better serving NumPy's large number of beginning
+to intermediate level users and on growing the community of NumPy
+contributors.
+It will also provide support to OpenBLAS, on which NumPy depends for
+accelerated linear algebra.
+
+Finally, since May 2019 the project receives a small amount annually from
+Tidelift, which is used to fund things like documentation and website
+improvements.
+
+\section*{Developers}
+
+NumPy is currently maintained by a group of 23 contributors with commit rights
+to the NumPy code base. Out of these, 17 maintainers were active in
+2019, 4 of whom were paid to work on the project full-time.
+Additionally, there are a few long term developers who contributed and maintain
+specific parts of NumPy, but are not officially maintainers.
+
+Over the course of its history, NumPy has attracted PRs by 823 contributors.
+However, its development relies heavily on a small number
+of active maintainers, who share more than half of the contributions among
+themselves.
+
+At a release cycle of about every half year, the five recent releases in the years
+2018 and 2019 have averaged about 450~PRs each,\footnote{
+    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
+    All data provided here is based on the development which happened through GitHub
+    PRs. In some cases contributions by maintainers may not be categorized as such.}
+with each release attracting more than a hundred new contributors.
+Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
+master branch.
+Although the number of PRs being merged fluctuates,
+the plot indicates an increased number of contributions over the past
+years.
+
+\section*{Community calls}
+
+The massive number of scientific Python packages that
+built on NumPy meant that it had an unusually high need for stability.
+So to guide our development we formalized the feature proposal process, and
+constructed a development roadmap with extensive input and feedback from the
+community.
+
+Weekly community calls alternate between triage and
+higher level discussion.  The calls not only involve developers from
+the community, but provide a venue for vendors and other external
+groups to provide input.  For example, after Intel produced a forked
+version of NumPy, one of their developers joined a call to discuss
+community concerns.
+
+\section*{NumPy enhancement proposals}
+
+Given the complexity of the codebase and the massive number of projects depending
+on it, large changes require careful planning and substantial work.
+NumPy Enhancement Proposals (NEPs) are modeled after
+Python Enhancement Proposals (PEPs) for ``proposing major new
+features, for collecting community input on an issue, and for
+documenting the design decisions that have gone into
+Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
+Since then there have been 19 proposed NEPS---6 have been implemented,
+4 have been accepted and are being implemented, 4 are under
+consideration, 3 have been deferred or superseded, and 2 have been rejected
+or withdrawn.
+
+\section*{Central role}
+
+NumPy plays a central role in building and standardizing much of the scientific
+Python community infrastructure.
+NumPy's docstring standard is now widely adopted.
+We are also now using the NEP system as a way to help coordinate the larger
+scientific Python community.
+For example, in NEP 29, we recommend, along with leaders from various other
+projects, that all projects across the Scientific Python ecosystem adopt a
+common ``time window-based'' policy for support of Python and NumPy versions.
+This standard will simplify downstream project and release planning.
+
+\section*{Wheels build system}
+
+A Python \emph{wheel} (\url{https://www.python.org/dev/peps/pep-0427/})
+is a standard file format for distributing Python libraries.
+In addition to Python code, a wheel may include compiled
+C extensions and other binary data.
+This is important, because many libraries, including NumPy,
+require a C compiler and other build tools to build the software
+from the source code, making it difficult for many users to install
+the software on their own.  The introduction of wheels to the Python
+packaging system has made it much easier for users to install
+precompiled libraries.
+
+A GitHub repository containing scripts to build NumPy wheels has
+been configured so that a simple commit to the repository triggers
+an automated build system that creates NumPy wheels for several
+computer platforms, including Windows, Mac OSX and Linux.  The wheels
+are uploaded to a public server and made available for anyone to use.
+This system makes it easy for users to install precompiled versions
+of NumPy on these platforms.
+
+The technology that is used to build the wheels evolves continually.
+At the time this paper is being written, a key component is the
+\texttt{multibuild} suite of tools developed by Matthew Brett and
+other developers (\url{https://github.com/matthew-brett/multibuild}).
+Currently, scripts using \texttt{multibuild} are written for
+the continuous integration platforms Travis-CI (for Linux and Mac OSX)
+and Appveyor (for Windows).
+
+\section*{Recent technical improvements}
+
+With the recent infusion of funding and a clear process for coordinating with
+the developer community, we have been able to tackle a number of important
+large scale changes.
+We highlight two of those below, as well as changes made to our testing
+infrastructure to support hardware platforms used in large scale computing.
+
+\section*{Array function protocol}
+
+A vast number of projects are built on NumPy;
+these projects are consumers of the NumPy API.
+Over the last several years, a growing number of projects are providers of
+a \emph{NumPy-like API} and array objects targeting audiences with specialized
+needs beyond NumPy's capabilities.
+For example, the NumPy API is implemented by several popular tensor computation
+libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
+JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
+and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
+PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
+and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
+provide tensor APIs with NumPy-inspired semantics.
+It is also implemented in packages that support sparse arrays
+such as \texttt{scipy.sparse} and PyData/Sparse.
+Another notable example is Dask, a library for parallel computing in
+Python.  Dask adopts the NumPy API and therefore presents a familiar
+interface to existing NumPy users, while adding powerful abilities to
+parallelize and distribute tasks.
+
+The multitude of specialized projects creates the difficulty that consumers
+of these NumPy-like APIs write code specific to a single project and do not support
+all of the above array providers.
+This is a burden for users relying on the specialized array-like, since
+a tool they need may not work for them.
+It also creates challenges for end-users who need to transition
+from NumPy to a more specialized array.
+The growing multitude of specialized projects with NumPy-like APIs threatened
+to again fracture the scientific Python community.
+
+To address these issues NumPy has the goal of providing the fundamental
+API for \emph{interoperability} between the various NumPy-like APIs.
+An earlier step in this direction was the implementation of the
+\texttt{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
+for most mathematical functions (\url{https://numpy.org/neps/nep-0013-ufunc-overrides.html}).
+In 2019 this was expanded more generally with the inclusion of the
+\texttt{\_\_array\_function\_\_} protocol into NumPy~1.17.
+These two protocols allow providers of array objects to be interoperable
+with the NumPy API: their arrays work correctly with almost all NumPy
+functions (\url{https://numpy.org/neps/nep-0018-array-function-protocol.html}).
+For the users relying on specialized array projects it means that even though
+much code is written specifically for NumPy arrays and uses the NumPy API as
+\texttt{import numpy as np}, it can nevertheless work for them.
+For example, here is how a CuPy GPU array can be passed through NumPy for
+processing, with all operations being dispatched back to CuPy:
+
+\begin{lstlisting}
+import numpy as np
+import cupy as cp
+
+x_gpu = cp.array([1, 2, 3])
+y = np.sum(x_gpu)  # Returns a GPU array
+\end{lstlisting}
+
+Similarly, user defined functions composed using NumPy can now be
+applied to, e.g., multi-node distributed Dask arrays:
+
+\begin{lstlisting}
+import numpy as np
+import dask.array as da
+
+def f(x):
+    """Function using NumPy API calls"""
+    y = np.tensordot(x, x.T)
+    return np.mean(np.log(y + 1))
+
+x_local = np.random.random([10000, 10000])  # random local array
+x_distr = da.random.random([10000, 10000])  # random distributed array
+
+f(x_local)  # returns a NumPy array
+f(x_distr)  # works, returns a Dask array
+\end{lstlisting}
+
+\section*{Random number generation}
+
+The NumPy \texttt{random} module provides pseudorandom numbers from a wide range of
+distributions. In legacy versions of NumPy, simulated random values are produced
+by a \texttt{RandomState} object that: handles seeding and state initialization;
+wraps the core pseudorandom number generator based on a Mersenne Twister
+implementation\footnote{to be precise, the standard 32-bit version of MT19937};
+interfaces with the underlying code that transforms random bits into
+variates from other distributions; and supplies a singleton instance exposed in
+the root of the random module.
+
+The \texttt{RandomState} object makes a compatibility guarantee so that a fixed
+seed and sequence of function calls produce the same set of values. This
+guarantee has slowed progress since improving the underlying code requires
+extending the API with additional keyword arguments. This guarantee continues to
+apply to \texttt{RandomState}.
+
+NumPy 1.17 introduced a new API for generating random numbers that use a more
+flexible structure that can be extended by libraries or end-users. The new API
+is built using components that separate the steps required to generate random
+variates. Pseudorandom bits are generated by a bit generator. These bits are
+then transformed into variates from complex distributions by a generator.
+Finally, seeding is handled by an object that produces sequences of high-quality
+initial values.
+
+Bit generators are simple classes that manage the state of an underlying
+pseudorandom number generator. NumPy ships with four bit generators. The default
+bit generator is a 64-bit implementation of the Permuted Congruential Generator
+\cite{pcg64} (\texttt{PCG64}). The three other bit generators are a 64-bit version
+of the Philox generator \cite{random123} (\texttt{Philox}), Chris Doty-Humphrey's
+Small Fast Chaotic generator \cite{practrand} (\texttt{SFC64}), and the 32-bit
+Mersenne Twister \cite{mt19937} (\texttt{MT19937}) which has been used in older
+versions of NumPy.\footnote{The
+\href{https://github.com/bashtage/randomgen}{randomgen project} supplies a wide
+range of alternative bit generators such as a cryptographic counter-based
+generators (\texttt{AESCtr}) and generators that expose hardware random number
+generators (\texttt{RDRAND}) \cite{randomgen}.} Bit generators provide
+functions, exposed both in Python and C, for generating random integer
+and floating point numbers. The three new bit generators were chosen for their
+combination of statistical soundness and performance. The PCG family of generators
+have been widely tested using state-of-the-art statistical tests and found to
+perform well \cite{lemire_blog}. All three of the new generators were tested
+using the PractRand test suite \cite{practrand} with four TB of random values.
+Each was tested using a single generator and using a composite generator assembled
+from multiple copies of the same bit generator, each seeded from a shared
+\texttt{SeedSequence}.
+
+The \texttt{Generator} consumes one of the bit generators and produces variates
+from complicated distributions. Many improved methods for generating random
+variates from common distributions were implemented, including the Ziggurat
+method for normal, exponential and gamma variates \cite{ziggurat}, and Lemire's
+method for bounded random integer generation \cite{lemire}. The \texttt{Generator}
+is more similar to the legacy \texttt{RandomState}, and its API is substantially
+the same. The key differences all relate to state management, which has been
+delegated to the bit generator. The \texttt{Generator} does not make the same
+stream guarantee as the \texttt{RandomState} object, and so variates may differ
+across versions as improved generation algorithms are
+introduced.\footnote{Despite the removal of the compatibility guarantee, simple
+reproducibility across versions is encouraged, and minor changes that do not
+produce meaningful performance gains or fix underlying bug are not generally
+adopted.}
+
+Finally, a \texttt{SeedSequence} is used to initialize a bit generator. The seed
+sequence can be initialized with no arguments, in which case it reads entropy
+from a system-dependent provider, or with a user-provided seed. The seed
+sequence then transforms the initial set of entropy into a sequence of
+high-quality pseudorandom integers, which can be used to initialize multiple bit
+generators deterministically. The key feature of a seed sequence is that
+it can be used to spawn child \texttt{SeedSequence}s to initialize
+multiple distinct bit generators.
+This capability allows a seed sequence to facilitate large distributed applications
+where the number of workers required is not known. The sequences generated from
+the same initial entropy and spawns are fully deterministic to ensure
+reproducibility.
+
+The three components are combined to construct a complete random number
+generator.
+
+\begin{lstlisting}
+from numpy.random import (
+    Generator,
+    PCG64,
+    SeedSequence,
+)
+
+seq = SeedSequence(1030424547444117993331016959)
+pcg = PCG64(seq)
+gen = Generator(pcg)
+\end{lstlisting}
+
+This approach retains access to the seed sequence which can then be
+used to spawn additional generators.
+
+\begin{lstlisting}
+children = seq.spawn(2)
+gen_0 = Generator(PCG64(children[0]))
+gen_1 = Generator(PCG64(children[1]))
+\end{lstlisting}
+
+While this approach retains complete flexibility, the method
+\texttt{np.random.default\_rng} can be used to instantiate a \texttt{Generator} when
+reproducibility is not needed.
+
+The final goal of the new API is to improve extensibility. \texttt{RandomState} is
+a monolithic object that obscures all of the underlying state and functions. The
+component architecture is one part of the extensibility improvements. The
+underlying functions (written in C) which transform the output of a bit
+generator to other distributions are available for use in CFFI. This allows the
+same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
+Numba. Both the bit generators and the low-level functions can also be used in C
+or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
+NumPy source. Alternative approaches that avoid this extra step are being
+explored.}
+
+\section*{Testing on multiple architectures}
+
+At the time of writing the two fastest supercomputers in the
+world, Summit and Sierra, both have IBM POWER9 architectures
+(\url{https://www.top500.org/lists/2019/11/}).
+In late 2018, Astra, the first ARM-based supercomputer
+to enter the TOP500 list, went into production
+(\url{https://en.wikichip.org/wiki/supercomputers/astra}).
+Furthermore, over 100 billion ARM processors have been produced as
+of 2017 (\url{https://en.wikipedia.org/wiki/ARM_architecture}),
+making it the most widely used instruction set architecture
+in the world.
+
+Clearly there are motivations for a large scientific computing
+software library to support POWER and ARM architectures. We've extended
+our continuous integration (CI) testing to include \texttt{ppc64le}
+(POWER8 on Travis CI) and ARMv8 (on Shippable service). We also test
+with the s390x architecture (IBM Z CPUs on Travis CI) so that we
+can probe the behavior of our library on a big-endian machine.
+This satisfies one of the major components of
+improved CI testing laid out in a version of our
+roadmap---specifically, ``CI for more exotic
+platforms."
+
+PEP 599 (\url{https://www.python.org/dev/peps/pep-0599/})
+lays out a plan for new Python binary wheel
+distribution support, \texttt{manylinux2014}, that adds
+support for a number of architectures supported by the CentOS
+Alternative Architecture Special Interest Group, including
+ARMv8, ppc64le, as well as s390x. We are thus well-positioned
+for a future where provision of binaries on these architectures
+will be expected for a library at the base of the ecosystem.
+
+\begin{thebibliography}{10}
+
+\bibitem{peterson2009f2py}
+P.~Peterson, ``{F2PY}: a tool for connecting {F}ortran and {P}ython programs,''
+  {\em International Journal of Computational Science and Engineering}, vol.~4,
+  no.~4, pp.~296--305, 2009.
+
+\bibitem{vanderwalt2011numpy}
+S.~van~der Walt, S.~C. Colbert, and G.~Varoquaux, ``The {NumPy} array: a
+  structure for efficient numerical computation,'' {\em Computing in Science \&
+  Engineering}, vol.~13, no.~2, pp.~22--30, 2011.
+
+\bibitem{LAPACK}
+E.~Anderson, Z.~Bai, C.~Bischof, S.~Blackford, J.~Demmel, J.~Dongarra,
+  J.~Du~Croz, A.~Greenbaum, S.~Hammarling, A.~McKenney, and D.~Sorensen, {\em
+  {LAPACK} Users' Guide}.
+\newblock Philadelphia, PA: Society for Industrial and Applied Mathematics,
+  3rd~ed., 1999.
+
+\bibitem{dongarra1990lapack}
+J.~Dongarra and S.~Hammarling, ``Evolution of numerical software for dense
+  linear algebra,'' in {\em Reliable Numerical Computation} (M.~G. Cox and
+  S.~Hammarling, eds.), pp.~297--327, Clarendon Press, 1990.
+
+\bibitem{dongarra2008netlib}
+J.~Dongarra, G.~H. Golub, E.~Grosse, C.~Moler, and K.~Moore, ``Netlib and
+  na-net: Building a scientific computing community,'' {\em IEEE Annals of the
+  History of Computing}, vol.~30, no.~2, pp.~30--41, 2008.
+
+\bibitem{pcg64}
+M.~E. O'Neill, ``Pcg: A family of simple fast space-efficient statistically
+  good algorithms for random number generation,'' Tech. Rep. HMC-CS-2014-0905,
+  Harvey Mudd College, Claremont, CA, Sept. 2014.
+
+\bibitem{random123}
+J.~K. Salmon, M.~A. Moraes, R.~O. Dror, and D.~E. Shaw, ``Parallel random
+  numbers: As easy as 1, 2, 3,'' in {\em Proceedings of 2011 International
+  Conference for High Performance Computing, Networking, Storage and Analysis},
+  SC '11, (New York, NY, USA), pp.~16:1--16:12, ACM, 2011.
+
+\bibitem{practrand}
+C.~Doty-Humphrey, ``Practrand, version 0.94.''
+\newblock [Online; accessed 1-June-2020].
+
+\bibitem{mt19937}
+M.~Matsumoto and T.~Nishimura, ``{Mersenne Twister}: A 623-dimensionally
+  equidistributed uniform pseudo-random number generator,'' {\em ACM
+  Transactions on Modeling and Computer Simulation}, vol.~8, pp.~3--30, Jan.
+  1998.
+
+\bibitem{randomgen}
+K.~Sheppard, B.~Duvenhage, P.~de~Buyl, and D.~A. Ham, ``bashtage/randomgen:
+  Release 1.16.2,'' Apr. 2019.
+
+\bibitem{lemire_blog}
+D.~Lemire, ``{Testing non-cryptographic random number generators: my
+  results},'' 2017.
+\newblock [Online; accessed 1-June-2020].
+
+\bibitem{ziggurat}
+G.~Marsaglia and W.~W. Tsang, ``The ziggurat method for generating random
+  variables,'' {\em Journal of Statistical Software, Articles}, vol.~5, no.~8,
+  pp.~1--7, 2000.
+
+\bibitem{lemire}
+D.~Lemire, ``Fast random integer generation in an interval,'' {\em ACM
+  Transactions on Modeling and Computer Simulation}, vol.~29, pp.~1--12, Jan
+  2019.
+
+\end{thebibliography}
+
+\begin{figure}
+    \centering
+
+    \caption{\textbf{Number of pull requests merged into the NumPy master branch for each
+        quarter since 2012.} The total number of PRs is indicated with the
+        lower blue area showing the portion contributed by current or previous
+        maintainers.}\label{fig:prs-over-time}
+\end{figure}
 

### 2020-06-05 mkauth.py

@@ -1,30 +1,0 @@
-import yaml
-
-with open("authors.yml") as f:
-    d = yaml.load(f, Loader=yaml.FullLoader)
-
-auths = d["auth"]
-
-star = len(d["inst"])
-affil_dict = {}
-affil_print = []
-for n, a in enumerate(d["inst"]):
-    for k, v in a.items():
-        num = n + 1
-        if num == star:
-            num = "*"
-        affil_dict.update({k: num})
-        affil_print.append(f"\\affil[{num}]{{{v}}}")
-
-with open("authors.tex", "w") as f:
-    for auth in auths:
-        for name, i in auth.items():
-            insts = ",".join(
-                [str(item) for item in [affil_dict[ins] for ins in i["institute"]]]
-            )
-            insts = f"[{insts}]"
-            f.write(f"\\author{insts}{{{name}}}")
-            f.write("\n")
-    for affil in affil_print:
-        f.write(affil)
-        f.write("\n")

### 2020-06-05 move_figures.py

@@ -1,33 +1,0 @@
-#!/usr/bin/env python3
-
-"""
-Move figures to the end of the file.
-"""
-
-import sys
-import os
-import re
-import bs4
-
-import subprocess
-from subprocess import PIPE
-
-
-if len(sys.argv) != 2:
-    print("Usage: move_figures.py document.tex")
-    sys.exit(0)
-
-tex_file = sys.argv[1]
-
-with open(tex_file, 'rb') as f:
-    data = f.read().decode('utf-8')
-
-regexp = re.compile(r'(\\begin\{figure[*]?}.*?\\end\{figure[*]?})',
-                    re.DOTALL|re.MULTILINE)
-
-figures = re.findall(regexp, data)
-data = re.sub(regexp, '', data)
-data = data.replace('%% FIGURES HERE %%', '\n\n'.join(figures))
-
-with open(tex_file, 'w') as f:
-    f.write(data)

### 2020-06-05 static/sketches/fig1.py

@@ -1,32 +1,0 @@
-import numpy as np
-x = np.arange(12)
-x = x.reshape(4, 3)
-x
-np.mean(x, axis=0)
-x = x - np.mean(x, axis=0)
-x
-
-# In [1]: import numpy as np
-
-# In [2]: x = np.arange(12)
-
-# In [3]: x = x.reshape(4, 3)
-
-# In [4]: x
-# Out[4]:
-# array([[ 0,  1,  2],
-#        [ 3,  4,  5],
-#        [ 6,  7,  8],
-#        [ 9, 10, 11]])
-
-# In [5]: np.mean(x, axis=0)
-# Out[5]: array([4.5, 5.5, 6.5])
-
-# In [6]: x = x - np.mean(x, axis=0)
-
-# In [7]: x
-# Out[7]:
-# array([[-4.5, -4.5, -4.5],
-#        [-1.5, -1.5, -1.5],
-#        [ 1.5,  1.5,  1.5],
-#        [ 4.5,  4.5,  4.5]])

### 2020-06-05 static/sketches/fig3.py

@@ -1,49 +1,0 @@
-DejaVu Sans Mono
-Size: 62
-
-# Fig. 3
-
-In [1]: import numpy as np
-
-In [2]: import dask.array as da
-
-In [3]: x = da.arange(12)
-
-In [4]: x = np.reshape(x, (4, 3))
-
-In [5]: x
-Out[5]: dask.array<..., shape=(4, 3), ...>
-
-In [6]: np.mean(x, axis=0)
-Out[6]: dask.array<..., shape=(3,), ...>
-
-In [7]: x = x - np.mean(x, axis=0)
-
-In [8]: x = x / np.std(x, axis=0)
-
-In [9]: x
-Out[9]: dask.array<..., shape=(4, 3), ...>
-
-
----------
-
-In [1]: import numpy as np
-
-In [2]: import dask.array as da
-
-In [3]: x = da.arange(12)
-
-In [4]: x = np.reshape(x, (4, 3))
-
-In [5]: x
-Out[5]: dask.array<reshape, shape=(4, 3), dtype=int64, chunksize=(4, 3), chunktype=numpy.ndarray>
-
-In [6]: np.mean(x, axis=0)
-Out[6]: dask.array<mean_agg-aggregate, shape=(3,), dtype=float64, chunksize=(3,), chunktype=numpy.ndarray>
-
-In [7]: x = x - np.mean(x, axis=0)
-
-In [8]: x = x / np.std(x, axis=0)
-
-In [9]: x
-Out[9]: dask.array<truediv, shape=(4, 3), dtype=float64, chunksize=(4, 3), chunktype=numpy.ndarray>

### 2020-08-12 methods.tex
Move history to supplement

@@
 will be expected for a library at the base of the ecosystem.
+
+\section*{Conclusion}
+
+We thank the many members of the community who provided
+feedback, submitted bug reports, made improvements to the documentation,
+code, or website, promoted NumPy's use in their scientific fields, and built
+the vast ecosystem of tools and libraries around NumPy.
+We also gratefully acknowledge the Numeric and Numarray developers
+on whose work we built.
+
+Jim Hugunin wrote Numeric in 1995, while a graduate student at MIT.
+Hugunin based his package on previous work by Jim Fulton, then working at the
+US Geological Survey, with input from many others.
+After he graduated, Paul Dubois at the Lawrence Livermore National Laboratory
+became the maintainer.
+Many people contributed to the project including T.E.O. (a co-author
+of this paper), David Ascher, Tim Peters, and Konrad Hinsen.
+
+In 1998 the Space Telescope Science Institute started using Python
+and in 2000 began developing a new array package called Numarray, written
+almost entirely by Jay Todd Miller, starting from a prototype developed by
+Perry Greenfield.  Other contributors included Richard L. White, J. C. Hsu,
+Jochen Krupper, and Phil Hodge.
+The Numeric/Numarray split divided the community, yet ultimately pushed
+progress much further and faster than would otherwise have been possible.
+
+Shortly after Numarray development started, T.E.O. took over maintenance of
+Numeric. In 2005, he led the effort and did most of the work to unify Numeric
+and Numarray, and produce the first version of NumPy.
+
+Eric Jones co-founded (along with T.E.O. and P.P.) the SciPy community, gave early feedback on array
+implementations, and provided funding and travel support to several
+community members.
+Numerous people contributed to the creation and
+growth of the larger SciPy ecosystem, which gives NumPy much of its
+value. Others injected new energy and ideas by creating experimental
+array packages.

### 2020-08-12 paper.tex
Move history to supplement

--- 2020-05-25 paper.tex
@@
-\documentclass[fleqn,10pt,twocolumn]{wlscirep}
+\documentclass[twocolumn]{wlscirep}
 
 % Packages
+\usepackage{moreverb} % for verbatim ouput (for wordcount)
 \usepackage{listings, textcomp}
@@
 
-\usepackage{caption}
-\captionsetup{%
-   figurename=Fig.,
-   tablename=Tab.
-}
+\usepackage[switch]{lineno}
+\linenumbers
 
@@
 
+% Count of words
+%\verbatiminput{wordcount.tex}
+
 \input{body}
@@
 suggested text and provided helpful feedback on the manuscript.
-
-We also thank the many members of the community who provided
-feedback, submitted bug reports, made improvements to the documentation,
-code, or website, promoted NumPy's use in their scientific fields, and built
-the vast ecosystem of tools and libraries around NumPy.
-We also gratefully acknowledge the Numeric and Numarray developers
-on whose work we built.
-
-Jim Hugunin wrote Numeric in 1995, while a graduate student at MIT.
-Hugunin based his package on previous work by Jim Fulton, then working at the
-US Geological Survey, with input from many others.
-After he graduated, Paul Dubois at the Lawrence Livermore National Laboratory
-became the maintainer.
-Many people contributed to the project including T.E.O. (a co-author
-of this paper), David Ascher, Tim Peters, and Konrad Hinsen.
-
-In 1998 the Space Telescope Science Institute started using Python
-and in 2000 began developing a new array package called Numarray, written
-almost entirely by Jay Todd Miller, starting from a prototype developed by
-Perry Greenfield.  Other contributors included Richard L. White, J. C. Hsu,
-Jochen Krupper, and Phil Hodge.
-The Numeric/Numarray split divided the community, yet ultimately pushed
-progress much further and faster than would otherwise have been possible.
-
-Shortly after Numarray development started, T.E.O. took over maintenance of
-Numeric. In 2005, he led the effort and did most of the work to unify Numeric
-and Numarray, and produce the first version of NumPy.
-
-Eric Jones co-founded (along with T.E.O. and P.P.) the SciPy community, gave early feedback on array
-implementations, and provided funding and travel support to several
-community members.
-Numerous people contributed to the creation and
-growth of the larger SciPy ecosystem, which gives NumPy much of its
-value. Others injected new energy and ideas by creating experimental
-array packages.
-
 K.J.M. and S.J.v.d.W. were funded in part by the Gordon and Betty Moore

### 2020-08-12 authors.tex
Update corr email
Update KJM affil
Update pv affil

--- 2020-02-15 authors.tex
@@
 \author[6,*]{Ralf Gommers}
-\author[7]{Pauli Virtanen}
-\author[8]{David Cournapeau}
-\author[9]{Eric Wieser}
-\author[10]{Julian Taylor}
+\author[7,8]{Pauli Virtanen}
+\author[9]{David Cournapeau}
+\author[10]{Eric Wieser}
+\author[11]{Julian Taylor}
 \author[4]{Sebastian Berg}
-\author[11]{Nathaniel J. Smith}
-\author[12]{Robert Kern}
+\author[12]{Nathaniel J. Smith}
+\author[13]{Robert Kern}
 \author[4]{Matti Picus}
-\author[13]{Stephan Hoyer}
-\author[14]{Marten H. van Kerkwijk}
-\author[2,15]{Matthew Brett}
-\author[16]{Allan Haldane}
-\author[17]{Jaime Fern\'andez del R\'io}
-\author[18,19]{Mark Wiebe}
-\author[6,20,21]{Pearu Peterson}
-\author[22]{Pierre G\'erard-Marchant}
-\author[23]{Kevin Sheppard}
-\author[24]{Tyler Reddy}
+\author[14]{Stephan Hoyer}
+\author[15]{Marten H. van Kerkwijk}
+\author[2,16]{Matthew Brett}
+\author[17]{Allan Haldane}
+\author[18]{Jaime Fern\'andez del R\'io}
+\author[19,20]{Mark Wiebe}
+\author[6,21,22]{Pearu Peterson}
+\author[23,24]{Pierre G\'erard-Marchant}
+\author[25]{Kevin Sheppard}
+\author[26]{Tyler Reddy}
 \author[4]{Warren Weckesser}
 \author[6]{Hameer Abbasi}
-\author[25]{Christoph Gohlke}
+\author[27]{Christoph Gohlke}
 \author[6]{Travis E. Oliphant}
 \affil[1]{Independent Researcher, Logan, Utah, USA}
-\affil[2]{Helen Wills Neuroscience Institute, University of California, Berkeley, Berkeley, CA, USA}
+\affil[2]{Brain Imaging Center, University of California, Berkeley, Berkeley, CA, USA}
 \affil[3]{Division of Biostatistics, University of California, Berkeley, Berkeley, CA, USA}
@@
 \affil[6]{Quansight LLC, Austin, TX, USA}
-\affil[7]{Department of Physics and Nanoscience Center, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
-\affil[8]{Mercari JP, Tokyo, Japan}
-\affil[9]{Department of Engineering, University of Cambridge, Cambridge, UK}
-\affil[10]{Independent Researcher, Karlsruhe, Germany}
-\affil[11]{Independent Researcher, Berkeley, CA, USA}
-\affil[12]{Enthought, Inc., Austin, TX, USA}
-\affil[13]{Google Research, Mountain View, CA, USA}
-\affil[14]{Department of Astronomy \& Astrophysics, University of Toronto, Toronto, ON, Canada}
-\affil[15]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
-\affil[16]{Department of Physics, Temple University, Philadelphia, PA, USA}
-\affil[17]{Google, Zurich, Switzerland}
-\affil[18]{Department of Physics and Astronomy, The University of British Columbia, Vancouver, BC, Canada}
-\affil[19]{Amazon, Seattle, Washington, USA}
-\affil[20]{Independent Researcher, Saue, Estonia}
-\affil[21]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
-\affil[22]{France-IX Services, Paris, France}
-\affil[23]{Department of Economics, University of Oxford, Oxford, UK}
-\affil[24]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
-\affil[25]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
-\affil[*]{numpy.articles@gmail.com}
+\affil[7]{Department of Physics, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
+\affil[8]{Nanoscience Center, University of Jyv\"askyl\"a, Jyv\"askyl\"a, Finland}
+\affil[9]{Mercari JP, Tokyo, Japan}
+\affil[10]{Department of Engineering, University of Cambridge, Cambridge, UK}
+\affil[11]{Independent Researcher, Karlsruhe, Germany}
+\affil[12]{Independent Researcher, Berkeley, CA, USA}
+\affil[13]{Enthought, Inc., Austin, TX, USA}
+\affil[14]{Google Research, Mountain View, CA, USA}
+\affil[15]{Department of Astronomy \& Astrophysics, University of Toronto, Toronto, ON, Canada}
+\affil[16]{School of Psychology, University of Birmingham, Edgbaston, Birmigham, UK}
+\affil[17]{Department of Physics, Temple University, Philadelphia, PA, USA}
+\affil[18]{Google, Zurich, Switzerland}
+\affil[19]{Department of Physics and Astronomy, The University of British Columbia, Vancouver, BC, Canada}
+\affil[20]{Amazon, Seattle, Washington, USA}
+\affil[21]{Independent Researcher, Saue, Estonia}
+\affil[22]{Department of Mechanics and Applied Mathematics, Institute of Cybernetics at Tallinn Technical University, Tallinn, Estonia}
+\affil[23]{Department of Biological and Agricultural Engineering, University of Georgia, Athens, GA}
+\affil[24]{France-IX Services, Paris, France}
+\affil[25]{Department of Economics, University of Oxford, Oxford, UK}
+\affil[26]{CCS-7, Los Alamos National Laboratory, Los Alamos, NM, USA}
+\affil[27]{Laboratory for Fluorescence Dynamics, Biomedical Engineering Department, University of California, Irvine, Irvine, CA, USA}
+\affil[*]{millman@berkeley.edu, stefanv@berkeley.edu, ralf.gommers@gmail.com}

### 2020-08-12 methods.tex
Remove fig reference
Rearrange methods
Typo
More typos

@@
+We thank the many members of the community who provided
+feedback, submitted bug reports, made improvements to the documentation,
+code, or website, promoted NumPy's use in their scientific fields, and built
+the vast ecosystem of tools and libraries around NumPy.
+We also gratefully acknowledge the Numeric and Numarray developers
+on whose work we built.
+
+\section*{History}
+
+Jim Hugunin wrote Numeric in 1995, while a graduate student at MIT.
+Hugunin based his package on previous work by Jim Fulton, then working at the
+US Geological Survey, with input from many others.
+After he graduated, Paul Dubois at the Lawrence Livermore National Laboratory
+became the maintainer.
+Many people contributed to the project including T.E.O. (a co-author
+of this paper), David Ascher, Tim Peters, and Konrad Hinsen.
+
+In 1998 the Space Telescope Science Institute started using Python
+and in 2000 began developing a new array package called Numarray, written
+almost entirely by Jay Todd Miller, starting from a prototype developed by
+Perry Greenfield.  Other contributors included Richard L. White, J. C. Hsu,
+Jochen Krupper, and Phil Hodge.
+The Numeric/Numarray split divided the community, yet ultimately pushed
+progress much further and faster than would otherwise have been possible.
+
+Shortly after Numarray development started, T.E.O. took over maintenance of
+Numeric. In 2005, he led the effort and did most of the work to unify Numeric
+and Numarray, and produce the first version of NumPy.
+
+Eric Jones co-founded (along with T.E.O. and P.P.) the SciPy community, gave early feedback on array
+implementations, and provided funding and travel support to several
+community members.
+Numerous people contributed to the creation and
+growth of the larger SciPy ecosystem, which gives NumPy much of its
+value. Others injected new energy and ideas by creating experimental
+array packages.
+
+\section*{Version control and collaboration}
+
 We use Git for version control and GitHub as the public hosting service for our
@@
 \texttt{y} is a 2-D array with shape $2 \times 1$, the output will be
-an array with shape $2 \times 3$ (Fig.~\ref{fig:array-concepts}c).
+an array with shape $2 \times 3$.
 The ufunc machinery takes care
@@
 
-\subsection*{Computing libraries.}
+\subsection*{Computing libraries}
 
@@
 Subprograms) implementation, most commonly OpenBLAS
-(http://www.openblas.net/).
+(\url{http://www.openblas.net/}).
 
@@
     \caption{\textbf{Number of pull requests merged into the NumPy master branch for each
-        quarter since 2012.} The total number of PRs is indicated with the
+        quarter since 2012.} The total number of PRs is indicated, with the
         lower blue area showing the portion contributed by current or previous
@@
 will be expected for a library at the base of the ecosystem.
-
-\section*{Conclusion}
-
-We thank the many members of the community who provided
-feedback, submitted bug reports, made improvements to the documentation,
-code, or website, promoted NumPy's use in their scientific fields, and built
-the vast ecosystem of tools and libraries around NumPy.
-We also gratefully acknowledge the Numeric and Numarray developers
-on whose work we built.
-
-Jim Hugunin wrote Numeric in 1995, while a graduate student at MIT.
-Hugunin based his package on previous work by Jim Fulton, then working at the
-US Geological Survey, with input from many others.
-After he graduated, Paul Dubois at the Lawrence Livermore National Laboratory
-became the maintainer.
-Many people contributed to the project including T.E.O. (a co-author
-of this paper), David Ascher, Tim Peters, and Konrad Hinsen.
-
-In 1998 the Space Telescope Science Institute started using Python
-and in 2000 began developing a new array package called Numarray, written
-almost entirely by Jay Todd Miller, starting from a prototype developed by
-Perry Greenfield.  Other contributors included Richard L. White, J. C. Hsu,
-Jochen Krupper, and Phil Hodge.
-The Numeric/Numarray split divided the community, yet ultimately pushed
-progress much further and faster than would otherwise have been possible.
-
-Shortly after Numarray development started, T.E.O. took over maintenance of
-Numeric. In 2005, he led the effort and did most of the work to unify Numeric
-and Numarray, and produce the first version of NumPy.
-
-Eric Jones co-founded (along with T.E.O. and P.P.) the SciPy community, gave early feedback on array
-implementations, and provided funding and travel support to several
-community members.
-Numerous people contributed to the creation and
-growth of the larger SciPy ecosystem, which gives NumPy much of its
-value. Others injected new energy and ideas by creating experimental
-array packages.
